### 2014-02-05 paper/appendix.tex
paper start

\newpage
\appendix
\section{Injection Measurements}
We injected the copper a distance of 1.16m (according to the spiral length calculator at http://www.giangrandi.ch/soft/spiral/spiral.shtml ) in a spiral whose cross-section was a square of area $9mm^2$.  I suspect that is about as far as we can go without using a vacuum at the end.

\section{Objet 260 Connex Digital Materials}
Softer materials (concentration of > 65% Tango-series material) do not easily accept the copper paint (it cracks when bent and is easy to wash off).  This may preclude flex sensors made of interior copper paint (conductive thread or other materials may be able to be used for some parts of this).

\section{Material Properties}

\begin{table*}[t]
\begin{tabular}{ l || l | p{3cm} | l | p{4cm} }
  \hline
  Name & Material & Resistance & Drying Time & Application Notes\\ \hline
  CuPro-Cote Coating & Copper & 2$\Omega$/inch & O(1 day) & Syringe \\ \hline
  Spectra 360 Electrode Gel & Liquid/Electrolytes & 125k$\Omega$/inch & O(hours) & Does not conduct dry \\ \hline
  Wire Glue & Carbon/Graphite &  23.6k$\Omega$/inch & O(minutes) & Syringe, very runny \\ \hline
  Bare Conductive Electric Paint & Carbon/Graphite & 110$\Omega$/inch & O(days) & Syringe \\ \hline
  Homemade Conductive Paint & Carbon/Graphite & 120$\Omega$/inch & O(hours) & \parbox{4cm}{Too thick for syringe,\\ apply externally with brush} \\ \hline
  Conductive Thread & Steel & \parbox{3cm}{1.8$\Omega$/inch taut \\ 2.5$\Omega$/inch loose} & N/A & \parbox{4cm}{Difficult to feed through\\turns} \\ \hline
  Solder Paste & Lead & 2$\Omega$/inch & N/A & \parbox{4cm}{Too thick for syringe,\\ must bake to conduct} \\ \hline
\end{tabular}
\end{table*}

### 2014-02-05 paper/limitations.tex
paper start

\section{Limitations}
Issues of drying time for fluids in long tubes.  Flexible material doesn't work forever.  Water in Objet prints discolors them.  Can be difficult to get certain materials through certain kinds of tubes.

### 2014-02-05 paper/relatedwork.tex
paper start

\section{Related Work}
This should be thought about soon!

\subsection{Fabrication}

\subsection{Printing Electronics}

\subsection{Fluids For Interaction}

\subsection{Routing}

\subsection{The Design Space}

\begin{table*}[t]
\begin{tabular}{ r || p{3cm} | p{3cm} | p{3cm} | p{3cm} }
  \hline
   & Gas & Liquid & Particles & Threadables\\ \hline \hline
  Visual & \parbox{3cm}{PneUIs, \\Harrison latex thingy, \\ \emph{smoke display}} & \parbox{3cm}{Splash Controllers, \\ \emph{paint mixer}} &\parbox{3cm}{\emph{clear thing with popcorn}} &\parbox{3cm}{\emph{neon sign}} \\ \hline
  Aural & \parbox{3cm}{\emph{resonance}} & \parbox{3cm}{} &\parbox{3cm}{\emph{CNC maracas}} &\parbox{3cm}{} \\ \hline
  Tactile/Haptic & \parbox{3cm}{PneUIs,\\ \emph{haptic textures}} & \parbox{3cm}{Splash Controllers,\\ \emph{warm/cold liquid}} &\parbox{3cm}{Jamming UIs, \\ \emph{sparse particle haptic textures}} &\parbox{3cm}{Otherlab robots} \\ \hline
  Olfactory/Gustatory & \parbox{3cm}{\emph{scents}} & \parbox{3cm}{\emph{smoothie mixer!}} &\parbox{3cm}{} &\parbox{3cm}{} \\ \hline
  Touch Input & \parbox{3cm}{Harrison latex thingy} & \parbox{3cm}{Harrison SFCS, \\ \emph{injectable capacitive sensors}} &\parbox{3cm}{Jamming UIs} &\parbox{3cm}{\emph{capacitance on wires}} \\ \hline
  Pressure Input & \parbox{3cm}{Slyper printed doodads} & \parbox{3cm}{\emph{capacitance, flow meter}} &\parbox{3cm}{} &\parbox{3cm}{} \\ \hline
  Other Input & \parbox{3cm}{Slyper (twist, bend, flex, etc.), \\ \emph{printed flex sensors}} & \parbox{3cm}{} &\parbox{3cm}{} &\parbox{3cm}{\emph{use traditional components}} \\ \hline
\end{tabular}
\end{table*}

### 2014-02-05 paper/tubes.tex
paper start

\section{A Series of Tubes}

\subsection{Types of Tubes}

\begin{itemize}
\item open - system to user, both ends open
\item return - system to system, both ends open
\item semi-closed - system to user, system side open, user side covered (e.g. elastic material)
\item fully closed - no openings on system or user side (e.g. air bubbles, resonance chambers)
\end{itemize}

\subsection{Features of Tubes}

\begin{itemize}
\item emphasize exterior connection points
\item emphasize interior design/path of tubes (for display)
\end{itemize}

\subsection{Media in Tubes}

\begin{itemize}
\item Gas (compressible).  Air with smells.  Air with fog.
\item Liquid (incompressible).  Clear, colored, conductive, heated, flavored (smoothie!).  Fill or coat.
\item Particles.  Sparse for haptic feedback.  Dense for jamming.
\item Threadables.  EL wire, conductive wire, fiberoptic cable, muscle wire.
\end{itemize}

\subsection{Design, Fabrication, Construction}

\subsubsection{Two Tools for Design}

\textbf{designing interior paths}

\textbf{designing exterior connection points}

\subsection{Inputs}

\subsubsection{Flexing}

Much like \cite{Slyper-shape}, we can sense flexing and bending of prints made on the Objet.  We can make prints on the Objet and tunnel through them, though, without making crazytastic silicone molds.

\subsubsection{Touch}
Capacitance (digital).

\subsubsection{Pressure}
Pressure (capacitance (ish) or fluid pressure).

\subsubsection{Tapping}
 Tapping (audio/resonance?) carried through particular tubes, like we talked about hard tubes in a soft thing.

\subsubsection{Other stuff?}
Could be.

\subsubsection{Traditional Components}
Obviously you can hook up traditional buttons, etc., the same way as always.  With copper paint instead of traditional wiring, we can share grounds, and we don't have to solder.

\subsection{Outputs}

\subsubsection{Visual}
EL wire.  Colored liquids.  Mechanical motion by pushing light stuff with air.

\subsubsection{Aural}
Resonance chambers.  Air cavity design for sound/amplification (see passive iPhone speakers).  I mean, this is basically just 3D printing instruments, which we know has been done.

\subsubsection{Haptic}
Compressible and incompressible fluids for actuation.  Recreation of PneUIs.  Tactile output.  Adding particles to add extra feedback.

\subsubsection{Olfactory/Gustatory}
Different chambers filled with different scented/flavored fluids.  We can mix them using pipes and pressure.

\subsection{Identity}
Resonance chambers for identification.

### 2014-02-10 paper/abstract.tex
added more bib, fleshed out a few things, put some tables

\begin{abstract}
Using different types of tubes with different topologies and different meda inside, we can make cool printable interfaces wtih a variety of input and output mechanisms.
\end{abstract}

### 2014-02-10 paper/appendix.tex
added more bib, fleshed out a few things, put some tables

@@
 \section{Objet 260 Connex Digital Materials}
-Softer materials (concentration of > 65% Tango-series material) do not easily accept the copper paint (it cracks when bent and is easy to wash off).
+Softer materials (concentration of $> 65\%$ Tango-series material) do not easily accept the copper paint (it cracks when bent and is easy to wash off).
 This may preclude flex sensors made of interior copper paint (conductive thread or other materials may be able to be used for some parts of this).
+
+\section{Bend Radius Measurements}
+\begin{itemize}
+\item EL wire diameter 2.3mm: minimum bend radius .35in = 8.89mm
+\item Water: uh.  None?
+\item Muscle Wire: ?
+\item Fiber Optic Cable: ?
+\end{itemize}
 

### 2014-02-10 paper/conclusion.tex
added more bib, fleshed out a few things, put some tables

\section{Conclusion}
In conclusion, tubes are cool.  You can do lots of things with them.  There are different kinds and topologies of tubes.  There are different things you can put in them.  This is cool for makers and also people with really expensive machines.

### 2014-02-10 paper/introduction.tex
added more bib, fleshed out a few things, put some tables

\section{Introduction}
3D printers on the rise, different kinds of materials, don't want to make knick knacks, personalizable interfaces, cheap components, etc.

### 2014-02-10 paper/relatedwork.tex
added more bib, fleshed out a few things, put some tables

@@
 \section{Related Work}
-This should be thought about soon!
+
+\begin{itemize}
+\item \cite{Sarik-tracebrush} - this uses an airspray machine to add conductive paths to surface of 3D prints
+\item \cite{Harrison-buttons} - fabricate latex + acrylic buttons and pressurize with air
+\item \cite{Hashimoto-squirming} - hold a speaker in your hands, and air pressure changes make it feel like you're holidng a living, squirming thing
+\item \cite{Kim-inflatablemouse} - a basic mouse, but it inflates so you can store it and also use it more reasonably than a flat mouse
+\item \cite{Ono-touchandactivate} - the Touch\'{e} with sound paper from last year's UIST
+\item \cite{Iwata-volflex} - a display made up of many balloons that inflate and deflate to change the shape
+\item \cite{Sells-reprap} - the RepRap people using a syringe of hot solder to squirt flat circuits into flat channels
+\item \cite{Majidi-curvature} - inject liquid metal into really thin tubes in a soft substrate, sense stretching by changing resistance
+\item \cite{Navarrette-gps} - I'm a little unclear on what they did, but they fabbed something with a "3D circuitboard" that has a bypass that goes into 3D.
+they don't offer a routing algorithm or anything, though.
+\item \cite{Optomec} - fancy machines that spray conductive film
+\item \cite{Slyper-pressure} - robots that are made of squishy stuff where air pressure changes are sensed.
+input components designed to react to different manipulations (pushing, squeezing, twisting, etc.)
+\item \cite{icecats-conductivepaint} - an instructable I followed to make my first batch of conductive paint
+\item \cite{Follmer-jamming} - Jamming User Interfaces
+\item \cite{Park-microchannels} - similar to injection of liquid metal, above
+\item \cite{Slyper-shape} - creating silicone bendy things with embedded electronics to sense flexing, stretching, etc., supported by those shapes.
+\item \cite{Yao-pneui} - PneUIs, creating interfaces with pneumatics
+\item \cite{Willis-infrastructs} - InfraStructs, identification of 3d printed objects using Terahertz imaging
+\item \cite{Willis-printedoptics} - Printed Optics.  doing cool stuff with clear material (touch sensing and display)
+\item \cite{Sodhi-aireal} - air vortex generation in free space.  air haptics.
+\item \cite{Savage-midas} - Midas.  routing in 2D to connect up capacitive sensors
+\item \cite{Savage-sauron} - Sauron.  computer vision of mechanical components obviates electronics installation.
+\item \cite{Sato-touche} - Touch\'{e}.  It's like Touch\'{e} with sound, but without sound.  SFCS.
+\end{itemize}
 
@@
 \begin{table*}[t]
+\caption{Existing systems are written in regular font.
+What's in \emph{italic} are things that I think we could make.
+I don't know what else might go in the blank spaces...
+looking for suggestions!
+Also, I'm not confident this breakdown is the most clear: for example, the "liquids" category includes the copper paint, which functions as wires ("threadables"), but takes the form of a liquid...}
 \begin{tabular}{ r || p{3cm} | p{3cm} | p{3cm} | p{3cm} }
@@
   Visual & \parbox{3cm}{PneUIs, \\Harrison latex thingy, \\ \emph{smoke display}} & \parbox{3cm}{Splash Controllers, \\ \emph{paint mixer}} &\parbox{3cm}{\emph{clear thing with popcorn}} &\parbox{3cm}{\emph{neon sign}} \\ \hline
-  Aural & \parbox{3cm}{\emph{resonance}} & \parbox{3cm}{} &\parbox{3cm}{\emph{CNC maracas}} &\parbox{3cm}{} \\ \hline
+  Aural & \parbox{3cm}{\emph{resonance}} & \parbox{3cm}{\emph{bubbling/splashing noises}} &\parbox{3cm}{\emph{CNC maracas}} &\parbox{3cm}{} \\ \hline
   Tactile/Haptic & \parbox{3cm}{PneUIs,\\ \emph{haptic textures}} & \parbox{3cm}{Splash Controllers,\\ \emph{warm/cold liquid}} &\parbox{3cm}{Jamming UIs, \\ \emph{sparse particle haptic textures}} &\parbox{3cm}{Otherlab robots} \\ \hline
@@
   Pressure Input & \parbox{3cm}{Slyper printed doodads} & \parbox{3cm}{\emph{capacitance, flow meter}} &\parbox{3cm}{} &\parbox{3cm}{} \\ \hline
-  Other Input & \parbox{3cm}{Slyper (twist, bend, flex, etc.), \\ \emph{printed flex sensors}} & \parbox{3cm}{} &\parbox{3cm}{} &\parbox{3cm}{\emph{use traditional components}} \\ \hline
+  Other Input & \parbox{3cm}{Slyper (twist, bend, flex, etc.), \\ \emph{printed flex sensors}} & \parbox{3cm}{\emph{use traditional components}} &\parbox{3cm}{} &\parbox{3cm}{\emph{use traditional components}} \\ \hline
 \end{tabular}
 \end{table*}
+\

### 2014-02-10 paper/tubes.tex
added more bib, fleshed out a few things, put some tables

@@
 \item fully closed - no openings on system or user side (e.g. air bubbles, resonance chambers)
+\end{itemize}
+
+\subsection{Topology of Tubes}
+
+\begin{itemize}
+\item splitting - one tube becomes two (e.g., for grounding several components)
+\item mixing - two tubes become one (e.g., for mixing liquids)
+\item star - several tubes meet at a point (good for mixing I guess?)
+\item tree - a.. tree.  easy to inject with goo.
 \end{itemize}
@@
 \item Liquid (incompressible).  Clear, colored, conductive, heated, flavored (smoothie!).  Fill or coat.
-\item Particles.  Sparse for haptic feedback.  Dense for jamming.
+\item Particles.  Single for games/display.  Sparse for haptic feedback.  Dense for jamming.
 \item Threadables.  EL wire, conductive wire, fiberoptic cable, muscle wire.
@@
 
-\textbf{designing interior paths}
+\textbf{designing interior paths} -- take into account bend radius of desired material
 
-\textbf{designing exterior connection points}
+\textbf{designing exterior connection points} -- focus on either \emph{shape} or \emph{location}
+
+\subsubsection{Fabrication Techniques}
+
+\textbf{printing} - different strategies with Objet (all print-in-place) and Makerbot (may need to add things like balloons afterwards).
+Ryan just got flexible material, we should see how stretchy it is...!
+We could also consider assembleable things that are easier to create using parts that clip together...
+probably out of scope.
+
+\textbf{hand tools} - post-fabrication modification is possible using hand tools.  We can mark the surface to show where conduits are and how deep.  I can also use this to test things beforehand.
 
@@
 \subsubsection{Touch}
-Capacitance (digital).
+Capacitance (digital).  SFCS (possible with mixed resistance things).
 
@@
 \subsubsection{Tapping}
- Tapping (audio/resonance?) carried through particular tubes, like we talked about hard tubes in a soft thing.
+ Tapping (audio/resonance?) carried through particular tubes, like we talked about hard tubes in a soft thing.  We use hard tubes in soft things, anyway, for the conductive goo.
 
@@
 \subsubsection{Visual}
-EL wire.  Colored liquids.  Mechanical motion by pushing light stuff with air.
+EL wire.  Colored liquids.  Mechanical motion by pushing light or plugging (like a big ball) stuff with fluid.
 

### 2014-02-21 paper/abstract.tex
moar draft
  * figure of tube design space.  text about it.  some text about
    the tool, some intro.  abstract.  placeholders for most figures.
    basically just more text.

@@
 \begin{abstract}
-Using different types of tubes with different topologies and different meda inside, we can make cool printable interfaces wtih a variety of input and output mechanisms.
+3D printers offer extraordinary flexibilty in prototyping the shape and mechanical function of objects.
+However, in spite of recent work, a 3D printer's role in prototyping \emph{interactivity} is still not clear.
+While macro-scale digital fabrication devices cannot manufacture electronics in-place, there have been many clever approaches to redirecting interaction onto 3D prints.
+We describe the design space of tubes and hollow chambers for interaction design: there are a variety of types, topologies, and inserted media for tubes that can be leveraged to create diverse inputs and outputs.
+We present a technique and design tool for routing tubes of various topologies through the interior of 3D printed parts.
+Our design tool is integrated into a 3D model manipulation program.
+It allows users to select begin and end points for tubes, then uses A* path routing and physics-based simulation to minimize the bending energy of routed paths.
+We present several totally tubular prototypes we created using this tool to show its flexibility and potential, as well as to explore new points in the tube design space.
 \end{abstract}

### 2014-02-21 paper/designspace.tex
moar draft
  * figure of tube design space.  text about it.  some text about
    the tool, some intro.  abstract.  placeholders for most figures.
    basically just more text.

\section{The Design Space of Tubes}

\begin{figure}[h]
\centering
    \includegraphics[width=3.4in]{figures/tubespace.png}
\caption{The design space of tubes.  Tube types, media, topologies, and design are discussed more fully in the text.}
\label{fig:tubespace}
\end{figure}

\subsection{Types of Tubes}

Tubes can come in four types: open, return, semi-closed, and fully enclosed.  Each of these types offers distinct interaction capabilities (see Figure \ref{fig:tubespace}).

Open tubes originate from the system side and connect the user side, with both ends of the tube open.  This type of tube may be used to create, for example, capacitive sensors: an open tube filled with conductive paint can be connected to a sensing platform (e.g., Arduino) on the system side, while a user can touch the uncapped other end of the tube.  Using Swept Frequency Capacitive Sensing \cite{Sato-touche} or other techniques, a user's touch of the open end can be sensed.  An open tube can also be used for output, for example by creating in-air vortices as in \cite{Sodhi-aireal}.

A return tube originates at the system and returns back to it.  By threading an electroluminescent (EL) wire through a clear return tube, a maker can create a custom piece of neon art.  If a return tube passes very close to the surface of a 3D printed model, warm or cold water passed through the tube could be used for temperature-based haptic feedback.

Semi-closed tubes are open at the system end (for control of the enclosed medium) and closed at the user end.  We believe this closed interface is most interesting when it is fabricated from a mobile material: for example, a series of tubes terminated in thin rubber membranes on the user side can be actuated by an air pump to create haptic feedback.  Without a printer capable of fabricating flexible material, a maker could affix a balloon to an open tube's end to behave similarly; another possibility is to use semi-closed tubes as audio-generating resonance chambers.

While possible to create, a tube which is open on the user side and closed on the system side is outside our focus is on computer-mediated interaction, and we therefore do not discuss these tubes.

A fully enclosed tube has no openings on either the system side or the user side.  Fully enclosed tubes can be used as resonance chambers (e.g., for object identification), or as air bubbles (e.g., as used in \cite{Willis-printedoptics} for internal display).  Their physical design space is very limited, as any support material required to create their internal geometry cannot be accessed for removal.

\subsection{Media in Tubes}

Tubes can be filled with a variety of media to create different interface affordances and capabilities.

``Gas'' comprises all compressible fluids.  Use of fluid pressure inside tubes can create haptic feedback at semi-closed interfaces, or the gases can be used as carriers for scents or fog.  As in \cite{Slyper-pressure}, structures can be engineered to change in air pressure when manipulated correctly (e.g., a spiral that changes pressure when twisted, but not pressed), and thus fluid pressure can also be used as an input.

Incompressible fluids (``liquids'') can perform many of the same interface tasks as gases.  One opportunity with liquids is to fill the interior of tubes with them and cap the ends.  In addition, one can use driable conductive fluids, such as copper paint, to coat the interior of tubes and allow them to function as arbitrarily-shaped wires.  This is especially helpful for the creation of a shared ground, or for creating single-wire capacitive interfaces amenable to sensing with SFCS \cite{Sato-touche}.

Tubes need not have hollow centers: in the case where routed tubes are filled with solid material---in particular, a solid material different from the model material---, interactions such as those in \cite{Willis-printedoptics} are possible.

Particulates, either printed in-place or inserted, can be of varying densities.  A single particle can be used for display.  Sparse particles in a stream of fluid can provide haptic feedback.  Dense particles in a semi-closed tube allow for jamming-based interactions at any point on the surface of an object \cite{Follmer-jamming}.

Threadable inserted elements, such as electroluminescent (EL) wire or fiberoptic cables, are those that can be threaded through tubes post-printing.  This allows overcoming limitations of printers: for example, a Printed Optics-style interface can be created on an inexpensive consumer-grade 3D printer using tubes and inserted fiber optic cable.

\subsection{Topological variations}

Tube topology enables different types of interactions.

Splitting or mixing tubes offer flexibility in output.  If a maker wished to create a painting device, she might wish to have two system-side tubes feeding in primary-colored red and blue paints which mix in varying ratios, allowing their pigments to combine before purple exits from the device (see Figure \ref{fig:tubespace}).  Splitting can also be useful, for example if our maker wants red paint output in two locations from her painting device, she could have one system-side tube, but split the tube into two (see Figure \ref{fig:tubespace}).

Star and tree topologies are extensions of the splitting and mixing primitives.  Using a star topology in which the tubes were filled with conductive paint, we created a toy with several touch-sensitive areas, see Figure \ref{fig:toy}.

\subsection{Features of Tubes}

Tubes may emphasize either their exterior features (connection points) or their interior paths.  These two features lead to different kinds of interfaces.  An example interface that focuses on the exterior connection points is the touch sensitive toy in Figure \ref{fig:toy}, where tubes must exit the toy at the eyes, ears, tail, etc.  An interface focused on the internal path of the tube is the neon sign in Figure \ref{fig:neon}: output is based on the shape of the tube.

\subsection{Inputs and Outputs}

\valkyrie{we can just include related work in this section?  we basically cover it in the table, anyway.}

\begin{figure*}[t]
\centering
    \includegraphics[width=\textwidth]{figures/designspace.png}
\caption{The design space of tube-based interactions.  Existing systems are written in regular font.  Those created with fabrication are in {\color{blue}blue}. In \emph{{\color{red}red italic}} are unexplored interactions creatable with custom-fabricated tubes.  Darker grey is output.  Lighter grey is input.  \valkyrie{I don't know what else might go in the blank spaces... looking for suggestions!  Also, I'm not confident this breakdown is the most clear: for example, the "liquids" category includes the copper paint, which functions as wires ("threadables"), but takes the form of a liquid...}}
\label{fig-designspace}
\end{figure*}

\valkyrie{optimally, we can just describe this thoroughly in the chart, and avoid spending a bazillion paragraphs talking about each thing individually here}

\subsection{Inputs}

\subsubsection{Flexing}

Much like \cite{Slyper-shape}, we can sense flexing and bending of prints made on the Objet.  We can make prints on the Objet and tunnel through them, though, without making crazytastic silicone molds.  We can make things flex using muscle wire and cleverly-placed expandable air pockets.

\subsubsection{Touch}
Capacitance (digital).  SFCS.

\subsubsection{Pressure}
Pressure (via SFCS or fluid pressure).

\subsubsection{Tapping}
 Tapping (audio/resonance?) carried through particular tubes, like we talked about hard tubes in a soft thing.  We use hard tubes in soft things, anyway, for the conductive goo.

\subsubsection{Other stuff?}
Could be.

\subsubsection{Traditional Components}
Obviously you can hook up traditional buttons, etc., the same way as always.  With copper paint instead of traditional wiring, we can share grounds, and we don't have to solder.

\subsection{Outputs}

\subsubsection{Visual}
EL wire.  Colored liquids.  Mechanical motion by pushing light or tube-plugging (like a big ball) objects with fluid.

\subsubsection{Aural}
Resonance chambers.  Air cavity design for sound/amplification (see passive iPhone speakers).  I mean, this is basically just 3D printing instruments, which we know has been done.

\subsubsection{Haptic}
Compressible and incompressible fluids for actuation.  Recreation of PneUIs.  Tactile output.  Adding particles to add extra feedback.

\subsubsection{Olfactory/Gustatory}
Different chambers filled with different scented/flavored fluids.  We can mix them using pipes and pressure.

\subsection{Identity}

Tubes included in printed objects allow for their identification.  While the tubes need not be visible (especially in the case of fully-enclosed tubes), their presence, location, and length change the acoustic resonance properties of a printed object.

Both identification by recall and intentional encoding are possible.  As seen in \cite{Ono-touchandactivate}, different objects have different acoustic signatures.  The presence of tubes changes this, thus allowing for recall of a shape once its acoustic signature has been recorded.  In addition, semi-closed tubes can function as resonance chambers: the first resonant frequency $F_1$ Hz of a semi-closed tube length $L$ meters can be found by $F_1 = \frac{c}{4L}$ (where $c$ is the speed of sound).  All odd harmonics (i.e., $3*F_1$, $5*F_1$, $7*F_1$, ...) of this frequency are also resonant frequencies of the tube.

An object's resonant frequency can be measured by attaching a speaker and microphone to it, sweeping frequencies with the speaker, and performing a Fourier transform on the resultant signal from the microphone.  Peaks in the transformed data correspond to stronger returned impulses: the resonant frequencies of the object.

\valkyrie{in my experiments so far, it seems that the material printed does conduct sound fairly well, so this is feasible.  I think we should get some equipment and do a small test, though, since I don't know how the resonance will change with the big hunk of plastic around the semi-open tube.  hopefully discussing this with Alex soon.}

### 2014-02-21 paper/exampleobjects.tex
moar draft
  * figure of tube design space.  text about it.  some text about
    the tool, some intro.  abstract.  placeholders for most figures.
    basically just more text.

\section{Example Objects}

Totally tubular examples.  We are making

\subsection{Touch-sensitive Toy}

\begin{figure}[h]
\centering
    \includegraphics[width=3.4in]{figures/series-of-tubes.jpg}
\caption{A touch-sensitive rabbit whose tubes are filled with conductive paint.  Sensing is done on a single wire via SFCS.  Inset shows the internal structure of the tubes generated by our design tool.}
\label{fig:toy}
\end{figure}

We created a touch-sensitive toy and an app that goes along with it, reminiscent of the boat application in \cite{Harrison-acoustic}.  This toy has an interior star

\subsection{Braille Learning Tool}

\begin{figure}[h]
\centering
    \includegraphics[width=3.4in]{figures/series-of-tubes.jpg}
\caption{This box has 6 individually-actuated semi-closed tubes capped with rubberlike material.  When air pressure is changed in these tubes, the caps inflate or deflate.  By actuating these tubes in particular sets, we can create braille letters.  Inset shows the internal structure of the tubes generated by our design tool.}
\label{fig:braille}
\end{figure}

\subsection{Custom Radio}

\begin{figure}[h]
\centering
    \includegraphics[width=3.4in]{figures/series-of-tubes.jpg}
\caption{This radio is assembled from traditional electronic components connected by copper-filled tubes.  The case was designed to allow the components to recess into it slightly.  Inset shows the internal structure of the tubes generated by our design tool.}
\label{fig:radio}
\end{figure}

\subsection{Presence-aware Pen Holder}

\begin{figure}[h]
\centering
    \includegraphics[width=3.4in]{figures/series-of-tubes.jpg}
\caption{This pen holder (a) uses Wimmer's FlyEye technique \valkyrie{can't add direct citation here??} (b) to sense the presence or absence of an object in each of its tubes.  Inset shows the internal structure of the tubes generated by our design tool.}
\label{fig:pens}
\end{figure}

\subsection{Animated Neon Sign}

\begin{figure}[h]
\centering
    \includegraphics[width=3.4in]{figures/series-of-tubes.jpg}
\caption{A neon sign designed using our tool.  The tubes contains four separate electroluminescent wires, lit in sequence to create animation (a).  (b) shows the selections we made to generate our tube structures.  (c) shows the internal structure of the tubes generated by our design tool.}
\label{fig:neon}
\end{figure}

### 2014-02-21 paper/introduction.tex
moar draft
  * figure of tube design space.  text about it.  some text about
    the tool, some intro.  abstract.  placeholders for most figures.
    basically just more text.

@@
+\begin{figure}[h]
+\centering
+    \includegraphics[width=3.4in]{figures/series-of-tubes.jpg}
+\caption{Here's a teaser figure of some cool stuff we did. Probably it should show one of our example objects, or we can make it really big (figure*) and show them all!}
+\label{fig:teaser}
+\end{figure}
+
 \section{Introduction}
-3D printers on the rise, different kinds of materials, don't want to make knick knacks, personalizable interfaces, cheap components, etc.
+Makers, as well as professional designers, leverage 3D printers as tools for design work.
+A wide array of objects, ranging from bicycle helmets to jewelry to video game controllers, are now prototyped or even manufactured using these machines.
+However, most devices fabricated by 3D printers are passive: accessible printers are not yet capable of creating integrated active systems.
+
+Willis, et al., ``envision a future world where interactive devices can be printed rather than assembled; a world where a device with active components is created as a single object, rather than a case enclosing circuit boards and individually assembled parts" \cite{Willis-printedoptics}.
+This is a vision we eagerly share: 3D printers are capable of greating arbitrary geometries not feasible to manufacture using traditional processes, and we see these capabilities being underutilized by makers and designers.
+We see many opportunitities to increase the interactivity of 3D prints using today's printer technology.
+
+Using specially modified printers or extra machinery, it is possible to create electronics on the surface of 3D prints \cite{optomec} \cite{Sells-reprap}.
+These techniques, however, require a high capital investment and technical expertise.
+In addition, they lack flexibility: the printed circuits must be routed in 2D to be created on the objects' manifold exteriors, and they can only be used to create electronic circuits.
+\valkyrie{what?}
+
+In addition to electronics, many other means of interactivity exist.
+
+\valkyrie{Points to hit: there is more than electronics to worry about, like fluids and the organic haptic feedback they afford; tangible interfaces are important (do I have to say this every time??); assembly sucks}

### 2014-02-21 paper/relatedwork.tex
moar draft
  * figure of tube design space.  text about it.  some text about
    the tool, some intro.  abstract.  placeholders for most figures.
    basically just more text.

@@
 
-\begin{itemize}
-\item \cite{Sarik-tracebrush} - this uses an airspray machine to add conductive paths to surface of 3D prints
-\item \cite{Harrison-buttons} - fabricate latex + acrylic buttons and pressurize with air
-\item \cite{Hashimoto-squirming} - hold a speaker in your hands, and air pressure changes make it feel like you're holidng a living, squirming thing
-\item \cite{Kim-inflatablemouse} - a basic mouse, but it inflates so you can store it and also use it more reasonably than a flat mouse
-\item \cite{Ono-touchandactivate} - the Touch\'{e} with sound paper from last year's UIST
-\item \cite{Iwata-volflex} - a display made up of many balloons that inflate and deflate to change the shape
-\item \cite{Sells-reprap} - the RepRap people using a syringe of hot solder to squirt flat circuits into flat channels
-\item \cite{Majidi-curvature} - inject liquid metal into really thin tubes in a soft substrate, sense stretching by changing resistance
-\item \cite{Navarrette-gps} - I'm a little unclear on what they did, but they fabbed something with a "3D circuitboard" that has a bypass that goes into 3D.
-they don't offer a routing algorithm or anything, though.
-\item \cite{Optomec} - fancy machines that spray conductive film
-\item \cite{Slyper-pressure} - robots that are made of squishy stuff where air pressure changes are sensed.
-input components designed to react to different manipulations (pushing, squeezing, twisting, etc.)
-\item \cite{icecats-conductivepaint} - an instructable I followed to make my first batch of conductive paint
-\item \cite{Follmer-jamming} - Jamming User Interfaces
-\item \cite{Park-microchannels} - similar to injection of liquid metal, above
-\item \cite{Slyper-shape} - creating silicone bendy things with embedded electronics to sense flexing, stretching, etc., supported by those shapes.
-\item \cite{Yao-pneui} - PneUIs, creating interfaces with pneumatics
-\item \cite{Willis-infrastructs} - InfraStructs, identification of 3d printed objects using Terahertz imaging
-\item \cite{Willis-printedoptics} - Printed Optics.  doing cool stuff with clear material (touch sensing and display)
-\item \cite{Sodhi-aireal} - air vortex generation in free space.  air haptics.
-\item \cite{Savage-midas} - Midas.  routing in 2D to connect up capacitive sensors
-\item \cite{Savage-sauron} - Sauron.  computer vision of mechanical components obviates electronics installation.
-\item \cite{Sato-touche} - Touch\'{e}.  It's like Touch\'{e} with sound, but without sound.  SFCS.
-\end{itemize}
+Our work is related to three existing research areas: fabrication, interaction techniques, and routing.
 
@@
 
-\subsection{Printing Electronics}
+Previous work has investigated the integration of interactivity into the fabrication process.
+\cite{Willis-printedoptics} used light pipes, solid clear tubes, to create integrated displays.
+\cite{Savage-sauron} investigated the use of computer vision to track interactions with physical mechanisms.
+\begin{itemize}
+\item \cite{Savage-sauron} - Sauron.  computer vision of mechanical components obviates electronics installation.
+\item \cite{Willis-infrastructs} - InfraStructs, identification of 3d printed objects using Terahertz imaging
+\item \cite{Willis-printedoptics} - Printed Optics.  doing cool stuff with clear material (touch sensing and display)
+\item \cite{Slyper-pressure} - robots that are made of squishy stuff where air pressure changes are sensed.
+input components designed to react to different manipulations (pushing, squeezing, twisting, etc.)
+\item \cite{Navarrette-gps} - I'm a little unclear on what they did, but they fabbed something with a "3D circuitboard" that has a bypass that goes into 3D.
+they don't offer a routing algorithm or anything, though.
+\end{itemize}
 
-\subsection{Fluids For Interaction}
+\subsubsection{Printing Electronics}
+\begin{itemize}
+\item \cite{Sarik-tracebrush} - this uses an airspray machine to add conductive paths to surface of 3D prints
+\item \cite{Optomec} - fancy machines that spray conductive film
+\item \cite{Sells-reprap} - the RepRap people using a syringe of hot solder to squirt flat circuits into flat channels
+\item \cite{icecats-conductivepaint} - an instructable I followed to make my first batch of conductive paint
+\end{itemize}
 
 \subsection{Routing}
+\begin{itemize}
+\item \cite{Savage-midas} - Midas.  routing in 2D to connect up capacitive sensors
+\item \cite{Park-microchannels} - similar to injection of liquid metal, above
+\item \cite{Majidi-curvature} - inject liquid metal into really thin tubes in a soft substrate, sense stretching by changing resistance
+\end{itemize}
 
-\subsection{The Design Space}
-
-\begin{table*}[t]
-\caption{Existing systems are written in regular font.
-What's in \emph{italic} are things that I think we could make.
-I don't know what else might go in the blank spaces...
-looking for suggestions!
-Also, I'm not confident this breakdown is the most clear: for example, the "liquids" category includes the copper paint, which functions as wires ("threadables"), but takes the form of a liquid...}
-\begin{tabular}{ r || p{3cm} | p{3cm} | p{3cm} | p{3cm} }
-  \hline
-   & Gas & Liquid & Particles & Threadables\\ \hline \hline
-  Visual & \parbox{3cm}{PneUIs, \\Harrison latex thingy, \\ \emph{smoke display}} & \parbox{3cm}{Splash Controllers, \\ \emph{paint mixer}} &\parbox{3cm}{\emph{clear thing with popcorn}} &\parbox{3cm}{\emph{neon sign}} \\ \hline
-  Aural & \parbox{3cm}{\emph{resonance}} & \parbox{3cm}{\emph{bubbling/splashing noises}} &\parbox{3cm}{\emph{CNC maracas}} &\parbox{3cm}{} \\ \hline
-  Tactile/Haptic & \parbox{3cm}{PneUIs,\\ \emph{haptic textures}} & \parbox{3cm}{Splash Controllers,\\ \emph{warm/cold liquid}} &\parbox{3cm}{Jamming UIs, \\ \emph{sparse particle haptic textures}} &\parbox{3cm}{Otherlab robots} \\ \hline
-  Olfactory/Gustatory & \parbox{3cm}{\emph{scents}} & \parbox{3cm}{\emph{smoothie mixer!}} &\parbox{3cm}{} &\parbox{3cm}{} \\ \hline
-  Touch Input & \parbox{3cm}{Harrison latex thingy} & \parbox{3cm}{Harrison SFCS, \\ \emph{injectable capacitive sensors}} &\parbox{3cm}{Jamming UIs} &\parbox{3cm}{\emph{capacitance on wires}} \\ \hline
-  Pressure Input & \parbox{3cm}{Slyper printed doodads} & \parbox{3cm}{\emph{capacitance, flow meter}} &\parbox{3cm}{} &\parbox{3cm}{} \\ \hline
-  Other Input & \parbox{3cm}{Slyper (twist, bend, flex, etc.), \\ \emph{printed flex sensors}} & \parbox{3cm}{\emph{use traditional components}} &\parbox{3cm}{} &\parbox{3cm}{\emph{use traditional components}} \\ \hline
-\end{tabular}
-\end{table*}
-\
+\subsection{Interaction Techniques}
+\begin{itemize}
+\item \cite{Sato-touche} - Touch\'{e}.  It's like Touch\'{e} with sound, but without sound.  SFCS.
+\item \cite{Harrison-buttons} - fabricate latex + acrylic buttons and pressurize with air
+\item \cite{Sodhi-aireal} - air vortex generation in free space.  air haptics.
+\item \cite{Yao-pneui} - PneUIs, creating interfaces with pneumatics
+\item \cite{Slyper-shape} - creating silicone bendy things with embedded electronics to sense flexing, stretching, etc., supported by those shapes.
+\item \cite{Iwata-volflex} - a display made up of many balloons that inflate and deflate to change the shape
+\item \cite{Ono-touchandactivate} - the Touch\'{e} with sound paper from last year's UIST
+\item \cite{Follmer-jamming} - Jamming User Interfaces
+\item \cite{Kim-inflatablemouse} - a basic mouse, but it inflates so you can store it and also use it more reasonably than a flat mouse
+\item \cite{Hashimoto-squirming} - hold a speaker in your hands, and air pressure changes make it feel like you're holidng a living, squirming thing
+\end{itemize}

### 2014-02-21 paper/tubes.tex
moar draft
  * figure of tube design space.  text about it.  some text about
    the tool, some intro.  abstract.  placeholders for most figures.
    basically just more text.

--- 2014-02-05 paper/tubes.tex
@@
 
-\subsection{Types of Tubes}
+\subsection{Tool for Tube Design}
 
-\begin{itemize}
-\item open - system to user, both ends open
-\item return - system to system, both ends open
-\item semi-closed - system to user, system side open, user side covered (e.g. elastic material)
-\item fully closed - no openings on system or user side (e.g. air bubbles, resonance chambers)
-\end{itemize}
+We created a tool which novice designers can use to create tube-powered interfaces in arbitrary 3D models.
+This tool allows users to brush over the surface of their model to either select exterior connection points of their tubes (see Figure \ref{fig:tool-process-interior}) or to author the tubes' interior paths (see Figure \ref{fig:tool-process-exterior}).
+Once the user's selections are made, our tool creates an initial shortest-path routing using A* to estimate the routed distance between points.
+This routing is used to create a rod; we run physics-based simulation steps on the rod to minimize its bending energy (and thereby minimize the bend radius of the tubes).
+The resultant routing is subtracted from the user's mesh.
+The modified mesh can then be 3D printed.
 
-\subsection{Features of Tubes}
+\subsubsection{Selection}
 
-\begin{itemize}
-\item emphasize exterior connection points
-\item emphasize interior design/path of tubes (for display)
-\end{itemize}
+\textbf{Exterior Connection Points}
 
-\subsection{Media in Tubes}
+\begin{figure}[h!]
+\centering
+    \includegraphics[width=3.4in]{figures/series-of-tubes.jpg}
+\caption{This figure has several sub-figures.
+a) shows a model with exterior connection points brushed.
+b) shows initial routing with A*.
+c) shows our physics-based, energy-minimizing rod/tube, d) shows the printed object with the tube (with something in it, copper paint presumably)}
+\label{fig:tool-process-exterior}
+\end{figure}
 
-\begin{itemize}
-\item Gas (compressible).  Air with smells.  Air with fog.
-\item Liquid (incompressible).  Clear, colored, conductive, heated, flavored (smoothie!).  Fill or coat.
-\item Particles.  Sparse for haptic feedback.  Dense for jamming.
-\item Threadables.  EL wire, conductive wire, fiberoptic cable, muscle wire.
-\end{itemize}
+Focus on shape of points and location of points.
 
-\subsection{Design, Fabrication, Construction}
+\textbf{Designing Interior Paths}
 
-\subsubsection{Two Tools for Design}
+\begin{figure}[h!]
+\centering
+    \includegraphics[width=3.4in]{figures/series-of-tubes.jpg}
+\caption{This figure has several sub-figures.
+a) shows a model with path shape and exterior connection point(s) brushed.
+b) highlights the points which can't be tubed as drawn (if they are not connected, too tight, etc.).
+c) shows our physics-based, energy-minimizing rod/tube and any path-smoothing that we do, d) shows the printed object with the tube (with something in it, EL wire presumably)}
+\label{fig:tool-process-interior}
+\end{figure}
 
-\textbf{designing interior paths}
+take into account bend radius of desired material - no need, we just do the minimum-bending path
 
-\textbf{designing exterior connection points}
+\subsubsection{Routing}
 
-\subsection{Inputs}
+Our basic first-pass routing algorithm uses the A* routing algorithm \cite{Hart-Astar}.
+The path cost in our implementation is based only on shortest distance between the starting and ending points, without weighting for distance from the surface.
 
-\subsubsection{Flexing}
+\subsubsection{Physical Simulation}
 
-Much like \cite{Slyper-shape}, we can sense flexing and bending of prints made on the Objet.
-We can make prints on the Objet and tunnel through them, though, without making crazytastic silicone molds.
+\subsubsection{Mesh Modification}
 
-\subsubsection{Touch}
-Capacitance (digital).
+\subsection{Fabrication Techniques}
 
-\subsubsection{Pressure}
-Pressure (capacitance (ish) or fluid pressure).
+\subsubsection{printing} - different strategies with Objet (all print-in-place) and Makerbot (may need to add things like balloons afterwards).
+Ryan just got flexible material, we should see how stretchy it is...!
+We could also consider assembleable things that are easier to create using parts that clip together...
+probably out of scope.
 
-\subsubsection{Tapping}
- Tapping (audio/resonance?) carried through particular tubes, like we talked about hard tubes in a soft thing.
-
-\subsubsection{Other stuff?}
-Could be.
-
-\subsubsection{Traditional Components}
-Obviously you can hook up traditional buttons, etc., the same way as always.  With copper paint instead of traditional wiring, we can share grounds, and we don't have to solder.
-
-\subsection{Outputs}
-
-\subsubsection{Visual}
-EL wire.  Colored liquids.  Mechanical motion by pushing light stuff with air.
-
-\subsubsection{Aural}
-Resonance chambers.  Air cavity design for sound/amplification (see passive iPhone speakers).  I mean, this is basically just 3D printing instruments, which we know has been done.
-
-\subsubsection{Haptic}
-Compressible and incompressible fluids for actuation.  Recreation of PneUIs.  Tactile output.  Adding particles to add extra feedback.
-
-\subsubsection{Olfactory/Gustatory}
-Different chambers filled with different scented/flavored fluids.  We can mix them using pipes and pressure.
-
-\subsection{Identity}
-Resonance chambers for identification.
+\subsubsection{hand tools} - post-fabrication modification is possible using hand tools.
+We can mark the surface to show where conduits are and how deep.
+I can also use this to test things beforehand.

### 2014-02-21 paper/conclusion.tex
more text, hooray

@@
 This is cool for makers and also people with really expensive machines.
+
+Ten movies streaming across that, that Internet, and what happens to your own personal Internet?
+I just the other day got… an Internet [that was] sent by my staff at 10 o'clock in the morning on Friday.
+I got it yesterday [Tuesday].
+Why?
+Because it got tangled up with all these things going on the Internet commercially.
+[…] They want to deliver vast amounts of information over the Internet.
+And again, the Internet is not something that you just dump something on.
+It's not a big truck.
+It's a series of tubes.
+And if you don't understand, those tubes can be filled and if they are filled, when you put your message in, it gets in line and it's going to be delayed by anyone that puts into that tube enormous amounts of material, enormous amounts of material.

### 2014-02-21 paper/exampleobjects.tex
more text, hooray

@@
 
-We created a touch-sensitive toy and an app that goes along with it, reminiscent of the boat application in \cite{Harrison-acoustic}.  This toy has an interior star
+We created a touch-sensitive toy and an app that goes along with it, reminiscent of the boat application in \cite{Harrison-acoustic}.
+This toy has an interior star topology where a single wire attached to an Arduino running an SFCS sketch splits to connect to sensors on the eyes, ears, tail, and nose.
+To accompany this toy, we built an app which prompts the user to find and touch certain parts of the animal (see Figure \ref{fig:toy}).
 
-\subsection{Braille Learning Tool}
+\subsection{Braille Haptic Display}
 
@@
     \includegraphics[width=3.4in]{figures/series-of-tubes.jpg}
-\caption{This box has 6 individually-actuated semi-closed tubes capped with rubberlike material.
-When air pressure is changed in these tubes, the caps inflate or deflate.
-By actuating these tubes in particular sets, we can create braille letters.
+\caption{When air pressure is changed in the six tubes, their caps inflate or deflate.
+Through controlled actuation, we can create Braille letters.
 Inset shows the internal structure of the tubes generated by our design tool.}
@@
 \end{figure}
+
+Using a $2\times3$ array of semi-closed tubes, we created a Braille output interface.
+The closed ends of the tubes are covered in a rubberlike material, while on the system side the tubes are attached to individually-controlled air pumps.
+These pumps can create positive or negative pressure in each tube, pushing the cap up or down and rendering a tactile Braille letter when used in concert (see Figure \ref{fig:braille}).
 
@@
 
+A custom ``radio'' built using tubes allows users to tune to different stations and play sound files related to those stations.
+This device uses a network of disconnected open tubes filled with conductive paint to connect traditional electronic components (a potentiometer, a piezo, and an LED) to a microcontroller.
+
 \subsection{Presence-aware Pen Holder}
@@
     \includegraphics[width=3.4in]{figures/series-of-tubes.jpg}
-\caption{This pen holder (a) uses Wimmer's FlyEye technique \valkyrie{can't add direct citation here??} (b) to sense the presence or absence of an object in each of its tubes.
+\caption{This pen holder (a) uses the FlyEye technique (b) to sense the presence or absence of an object in each of its tubes.
 Inset shows the internal structure of the tubes generated by our design tool.}
@@
 \end{figure}
+
+Our presence-aware pen holder can distinguish which tool or tools a user has picked up (see Figure \ref{fig:pens}).
+Such information was used by \cite{Mueller-constructable} to determine which physical laser pointer a designer was using to interact with their interface.
+Our pen holder uses the FlyEye technique described by Wimmer in \cite{Wimmer-flyeye} and contains open tubes filled with fiber optic cables, two per pen chamber.
+One tube from each chamber leads to an infrared LED, and the other leads to a camera.
+When a pen is in its appointed place, light from the LED is reflected off its bottom and travels through the other optical fiber into the camera, where it registers as a bright point.
+A dark point on the camera region appears when a pen has been removed from its place, and the position of the point indicates which pen it is.
 
@@
     \includegraphics[width=3.4in]{figures/series-of-tubes.jpg}
-\caption{A neon sign designed using our tool.
-The tubes contains four separate electroluminescent wires, lit in sequence to create animation (a).
-(b) shows the selections we made to generate our tube structures.
-(c) shows the internal structure of the tubes generated by our design tool.}
+\caption{A neon sign (a) and its partial duplicate designed using our tool (b).
+(c) shows the selections we made to generate our tube structures.
+(d) shows the internal structure of the tubes generated by our design tool.}
 \label{fig:neon}
 \end{figure}
+
+Neon art, perhaps best known for its association with Las Vegas, is traditionally made from hand-formed glass tubes containing neon gas.
+The tubes light up when a current is passed through them.
+For this type of art, the path of the tubes is of crucial importance, as it determines how the sign will look.
+We designed a partial duplicate of a renowned piece of neon art from Seattle (or was it Portland?).
+The piece features open tubes due to support-flushing constraints, however the tubes could also be semi-closed.
+They have been threaded with EL wire which is lit in sequence to create an animation.

### 2014-02-21 paper/limitations.tex
added more bib, fleshed out a few things, put some tables
more text, hooray

@@
-\section{Limitations}
-Issues of drying time for fluids in long tubes.
-Flexible material doesn't work forever.
-Water in Objet prints discolors them.
-Can be difficult to get certain materials through certain kinds of tubes.
+\section{Limitations and Future Work}
+Tubes, while a powerful tool for directing interaction to arbitrary locations on an object's surface, have their shortcomings.
+Most obviously, their fundamental shortfall is that they cannot replace the active elements required for a functioning interactive device: they simply redirect sensing to different locations on an object's surface.
+All tubes need to, at some point, connect to a sensor, microcontroller, or pump.
+
+Some media are difficult to insert in tubes, or may not be compatible with all types of tubes.
+Fluids in particular provide challenges.
+For example, the conductive paint we used in our designs can take multiple days to dry in tubes of sufficient length.
+Any liquid kept in tubes printed by the Objet can cause the exterior of the part to discolor due to seepage.
+Hobbyist machines do not guarantee watertightness of their prints, so leaking of water or air can occur.
+
+The physical fabrication process also leads to some considerations.
+To print without support on a hobbyist machine, internal tubes cannot have overhangs of greater than $45^{\circ}$.
+On the Objet, support-free prints cannot overhang more than $14^{\circ}$.
+Removal of support from printed tubes can be complicated, especially when their internal geometry and branching structures are complex.
+
+Our design tool is not yet fully capable of realizing all possibilities in the design space of tubes.
+It lacks X and Y \valkyrie{probably it will lack the network manipulation to make mixing or splitting tubes...?}.
+We hope to address this in future work.
+
+\valkyrie{what other future work?
+miniaturization doesn't seem right, it's possible to use on mobile phones already with their on-board sensors (mic), maybe talking about shape-changing tubes?
+post-fabrication modification with drills?
+that doesn't sound very convincing.}

### 2014-02-21 paper/tubes.tex
more text, hooray

@@
 
-\subsubsection{Physical Simulation}
+Physical Simulation :
+
+How does it actually work?
+
+Bend Minimization
+
+Collision detection
 
@@
 
+How we actually cut stuff and make tubes happen.
+
 \subsection{Fabrication Techniques}
 
-\subsubsection{printing} - different strategies with Objet (all print-in-place) and Makerbot (may need to add things like balloons afterwards).
+\subsubsection{Printing} - different strategies with Objet (all print-in-place) and Makerbot (may need to add things like balloons afterwards).
 Ryan just got flexible material, we should see how stretchy it is...!
@@
 
-\subsubsection{hand tools} - post-fabrication modification is possible using hand tools.
+\subsubsection{Hand Tools} - post-fabrication modification is possible using hand tools.
 We can mark the surface to show where conduits are and how deep.
-I can also use this to test things beforehand.
+I can also use this to test things before spending time printing them using a big ol' block o' plastic.

### 2014-02-27 paper/relatedwork.tex
removed .pdf from repo, added some references

@@
 \end{itemize}
+
+\subsection{Misc}
+\bjoern{These may not belong in related work, but I'm just collecting them here.}
+\begin{itemize}
+\item \cite{Wong-cgf11} - Wong presents a graph-based approach for generating continuous line illustrations from images.
+It's approach of linking edges into a continuous line based on Semi-Eulerization is similar to ours.
+Prior work used tours of nodes, rather than edges, and cast continuous line drawing as a traveling salesman problem~\cite{Bosch-tsp}.
+\item \cite{Schweikardt-tei09} Schweikardt et al.
+also made a connection between electroluminescent wire and graph theory, though their contribution is orthogonal: they provide a tangible construction kit with smart nodes for building physical graph examples.
+Edges are lit up by EL wire.
+\item \cite{strattman1997neon} - Strattman is the standard "text book" for neon lighting designers.
+\item In neon sign making, segments of a tube that should not appear illuminated are covered with opaque paint. This is called "blocking out"\footnote{\url{http://en.wikipedia.org/wiki/Neon_sign}}.
+\end{itemize}

### 2014-02-27 paper/abstract.tex
more paper text, algo and proof for neon

@@
 However, in spite of recent work, a 3D printer's role in prototyping \emph{interactivity} is still not clear.
-While macro-scale digital fabrication devices cannot manufacture electronics in-place, there have been many clever approaches to redirecting interaction onto 3D prints.
+While macro-scale digital fabrication devices cannot manufacture electronics in-place, we propose a general technique for redirecting active components through the core and onto the surface of 3D printed interactive objects using a series of tubes.
 We describe the design space of tubes and hollow chambers for interaction design: there are a variety of types, topologies, and inserted media for tubes that can be leveraged to create diverse inputs and outputs.
@@
 Our design tool is integrated into a 3D model manipulation program.
-It allows users to select begin and end points for tubes, then uses A* path routing and physics-based simulation to minimize the bending energy of routed paths.
+There are two distinct routing algorithms.
+One allows users to select begin and end points for tubes, then uses A* path routing and physics-based simulation to minimize the bending energy of routed paths.
+The second allows users to enter a description of paths to follow: for this we developed a novel neon sign routing algorithm and offer proof of its optimality.
 We present several totally tubular prototypes we created using this tool to show its flexibility and potential, as well as to explore new points in the tube design space.

### 2014-02-27 paper/designspace.tex
more paper text, algo and proof for neon

@@
 \section{The Design Space of Tubes}
+
+We propose the use of tubes to redirect active components through the core and onto the surface of 3D printed interactive objects.
+By using different types and topologies of tubes and inserting different media into the completed tubes, a variety of input and output types \valkyrie{need clearer word than ``types"} can be located at arbitrary points on the surface of a 3D print.
+We describe the design space of tubes, and then follow up with some of the input and output possibilities---both those that have been previously explored and those that are possible but remain unexplored.
 
@@
 
-\valkyrie{we can just include related work in this section?  we basically cover it in the table, anyway.}
+\subsection{Interaction Techniques}
+These didn't get discussed yet.
+\begin{itemize}
+\item \cite{Iwata-volflex} - a display made up of many balloons that inflate and deflate to change the shape
+\item \cite{Kim-inflatablemouse} - a basic mouse, but it inflates so you can store it and also use it more reasonably than a flat mouse
+\item \cite{Hashimoto-squirming} - hold a speaker in your hands, and air pressure changes make it feel like you're holidng a living, squirming thing
+\end{itemize}
+
+\valkyrie{this section includes related work.
+is that ok?
+it's in the table, and I think we should cover the related input techniques here rather than in the designated related work section (which I think is more suited to fabrication and routing stuff).}
+
+By crossing tube types, topologies, and inserted media, we create a space of possible inputs and outputs that can be tube-mediated.
+We offer a table (see Figure \ref{fig:designspace}) containing these possibilities, referencing previous work on input/output techniques where appropriate, noting that many of these techniques have not been attempted using digital fabrication.
+We suggest new points in the design space that have been unexplored or not yet explored with digital fabrication.
 
@@
 looking for suggestions!
-Also, I'm not confident this breakdown is the most clear: for example, the "liquids" category includes the copper paint, which functions as wires ("threadables"), but takes the form of a liquid...}}
-\label{fig-designspace}
+Also, I'm not confident this breakdown is the most clear: for example, the "liquids" category includes the copper paint, which functions as wires ("threadables"), but takes the form of a liquid...
+and the general presentation of the table could probably be better in some way.}}
+\label{fig:designspace}
 \end{figure*}
 
-\valkyrie{optimally, we can just describe this thoroughly in the chart, and avoid spending a bazillion paragraphs talking about each thing individually here}
+\subsubsection{Inputs}
 
-\subsection{Inputs}
+Tubes create opportunities for many types of input sensing across the surface of printed devices.
+We describe sensing touch, pressure, grasp, flexing, tapping, and manipulation of traditional electronic components.
+We recognize that there are likely more input sensing techniques, both existing and on the horizon, that are compatible with the use of tubes.
 
-\subsubsection{Flexing}
+Touch sensing is enabled through tubes filled with conductive media.
+Traditional wires (threadable) may be used, but we have had success with conductive paint (liquid).
+Using conductive paint and Swept Frequency Capacitve Sensing (SFCS \cite{Sato-touche}), an interior star tube topology can enable single-wire touch- and grasp-sensing at any set of points on a printed object's surface (see Figure \cite{fig:toy}).
+Simple capacitance measurements are possible using the same media, and custom-designed sensors like those created in Savage, et al.,'s Midas system are also possible \cite{Savage-midas}.
 
-Much like \cite{Slyper-shape}, we can sense flexing and bending of prints made on the Objet.
-We can make prints on the Objet and tunnel through them, though, without making crazytastic silicone molds.
-We can make things flex using muscle wire and cleverly-placed expandable air pockets.
+Pressure can be sensed in multiple ways.
+Slyper, et al., in \cite{Slyper-pressure} contribute a set of semi-closed fabricatable primitives that respond to particular actions, like twisting or pushing, with changes in internal air pressure.
+\valkyrie{that sentence is not clear}  These primitives can be attached at the terminus of any semi-closed tube; the user can manipulate the printed endpoint and the system can sense air pressure changes.
+Pressure can also be more simply sensed via capacitance using those techniques.
 
-\subsubsection{Touch}
-Capacitance (digital).  SFCS.
+Grasp sensing can also be enabled using Wimmer's FlyEye technique \cite{Wimmer-flyeye}.
+This technique involves a pair of optical links at each point of interest: one of each pair is connected to an infrared (IR) LED and the other to a camera.
+When some object nears or touches a pair of cables, IR light from the LED is reflected back to the camera and the touched point appears as a bright spot.
+Follmer, et al.,'s Jamming User Interfaces \cite{Follmer-jamming} rely dense particulates enclosed in a flexible volume, and can be sensed optically.
+These dense particulates can be printed in-place with desired characteristics (such as varying particle size or shape) using sacrificial support material, or added post-print.
+The optical links necessary for sensing with FlyEye or Jamming can be created either via clear solid cores as in \cite{Willis-printedoptics} or via fiber optic cables threaded through hollow tubes (see Figure \ref{fig:pens}).
 
-\subsubsection{Pressure}
-Pressure (via SFCS or fluid pressure).
+Some 3D printers, like our Objet Connex 260, can fabricate rubber-like materials.
+Much like Slyper, et al., in \cite{Slyper-shape}, we can sense flexing and bending of prints made on these machines.
+While Slyper, et al., created multi-level multi-stage silicone molds with embedded wires to build their sensors, by placing tubes in our rubberlike prints and inserting conductive media afterwards, we can perform the same flex sensing.
 
-\subsubsection{Tapping}
- Tapping (audio/resonance?) carried through particular tubes, like we talked about hard tubes in a soft thing.  We use hard tubes in soft things, anyway, for the conductive goo.
+Tapping is another input possibility.
+Due to differential sound conductivity in different printed materials (in particular, the rubberlike Objet material compared to the hard Objet material \valkyrie{presumably we should get or take some measurements of this and include it in the appendix?}), tubes of a greater conducting material can be embedded in a model of a lesser conducting material.
+A microphone or piezo placed at the system-side of these sound-conducting tubes can determine where the model was tapped and how it was tapped.
+Active acoustic sensing as described in Touch \& Activate \cite{Ono-touchandactivate} is also possible; this technique would allow for particular areas of interest (connected to sound-conducting tubes) to be touch sensitive, while other areas (made of non-conducting material) are not sensed.
 
-\subsubsection{Other stuff?}
-Could be.
+Sensing via traditional electronic components (e.g., potentiometers, alcohol gas sensors) can be accomplished by conductive inserted media.
+Components can be recessed into a print's surface, with their leads implanted into open tubes.
+By inserting liquid copper paint instead of threading traditional wires, all components in a model can share a ground line, and the paint's drying process obviates the use of solder or glue to affix the components in place (see Figure \ref{fig:embeddedcomponents}).
 
-\subsubsection{Traditional Components}
-Obviously you can hook up traditional buttons, etc., the same way as always.  With copper paint instead of traditional wiring, we can share grounds, and we don't have to solder.
+\begin{figure}[h]
+\centering
+    \includegraphics[width=3.4in]{figures/series-of-tubes.jpg}
+\caption{This sphere has a recessed space to hold a potentiometer.  The potentiometer and the LEDs share a ground, seen by the translucent rendered view in (b).}
+\label{fig:embeddedcomponents}
+\end{figure}
 
-\subsection{Outputs}
+\subsubsection{Outputs}
 
-\subsubsection{Visual}
-EL wire.  Colored liquids.  Mechanical motion by pushing light or tube-plugging (like a big ball) objects with fluid.
+While many input techniques require the use of conductive media inserted into tubes, output media are more varied.
+We describe potential tube-based outputs organized by the five senses: visual, aural, haptic, and olfactory/gustatory.
+In addition to outputs we mention explicitly, any outputs possible using traditional electronic components are possible as described above.
 
-\subsubsection{Aural}
-Resonance chambers.  Air cavity design for sound/amplification (see passive iPhone speakers).  I mean, this is basically just 3D printing instruments, which we know has been done.
+\emph{Visual} outputs can be mediated by gases, liquids, solid, particulates, or threadables.
+EL wire can be threaded through custom paths to create neon signs, as in Figure \ref{fig:neon}.
+Colored liquids can be mixed and split before exiting an opaque device, or in a transparent device their mixing, splitting, and paths could be used for dispaly.
+PneUIs \cite{Yao-pneui} can be fabricated as single parts: the rubberlike material offered by our Objet is inflatable when thin (200\% elongation at break), and a PneUI located at the terminus of a tube could be actuated by a system-side air pump (see Figure \ref{fig:braille}).
+Mechanical motion can be mediated through fluids (a light object which is partially contained in a tube or located at its end could be pushed by the fluid pressure of the tube) or threadables (e.g., muscle wire).
+Large particles trapped in tubes visible to the user can be used for, e.g., maze games (see Figure \ref{fig:maze}).
+Using techniques like those described by Harrison, et al., in \cite{Harrison-buttons}  \valkyrie{this paragraph seems terrible.
+I think all these paragraphs might seem terrible...}
 
-\subsubsection{Haptic}
-Compressible and incompressible fluids for actuation.  Recreation of PneUIs.  Tactile output.  Adding particles to add extra feedback.
+\emph{Aural} outputs can be created with gases.
+Zoran has successfully 3D printed a functional flute, which in essence is a long open open tube with semi-closed tubes coverable by keys \cite{Zoran-flute}.
+Helmholz resonance (the phenomenon that creates noise when you blow into the top of a glass bottle, which is also how some musical instruments, like ocarinas, work) can be leveraged to create computer-controllable sound chambers.
+All that is necessary for this is an enclosed chamber with a tube connected to it via a narrow neck-point (see Figure \ref{fig:ocarina}).
+Passive sound amplification (like a phonograph horn) and sound redirection are also possible using tubes.
 
-\subsubsection{Olfactory/Gustatory}
-Different chambers filled with different scented/flavored fluids.  We can mix them using pipes and pressure.
+\begin{figure}[h]
+\centering
+    \includegraphics[width=3.4in]{figures/series-of-tubes.jpg}
+\caption{This box (a) contains an enclosed Helmholz resonator, seen in the cutaway view (b).
+This resonator creates a sound when air is blown through the tube at the top, because air compresses at the small neck joint between the chamber and the tube.
+\valkyrie{need to get a test print of this that actually works}}
+\label{fig:embeddedcomponents}
+\end{figure}
+
+\emph{Haptic} outputs are possible through compressible and incompressible fluids.
+Fluids can actuate particulates and create programmable pliability (as discussed above and in \cite{Follmer-jamming}).
+Additionally, use of semi-closed tubes with rubberlike material as the caps allows for gases to actuate surface features (see Figure \ref{fig:braille}).
+Sodhi, et al., in \cite{Sodhi-aireal} created free-air haptic feedback using controlled vortex generation: this technique could also be reproduced using a series of tubes.
+
+\emph{Olfactory} and \emph{gustatory} outputs can be created through the mixing and splitting of tubes carrying scented or flavored fluids, respectively.
 
@@
 As seen in \cite{Ono-touchandactivate}, different objects have different acoustic signatures.
-The presence of tubes changes this, thus allowing for recall of a shape once its acoustic signature has been recorded.
-In addition, semi-closed tubes can function as resonance chambers: the first resonant frequency $F_1$ Hz of a semi-closed tube length $L$ meters can be found by $F_1 = \frac{c}{4L}$ (where $c$ is the speed of sound).
+Additionally, two objects that are visually identical but which have fully enclosed (or other types) of tubes on their interior can be distinguised acoustically.
+Thus we can recall an object's identity once its acoustic signature has been recorded.
+
+We have also experimented with intentional encoding.
+Semi-closed tubes can function as resonance chambers: the first resonant frequency $F_1$ Hz of a semi-closed tube length $L$ meters can be found by $F_1 = \frac{c}{4L}$ (where $c$ is the speed of sound).
 All odd harmonics (i.e., $3*F_1$, $5*F_1$, $7*F_1$, ...) of this frequency are also resonant frequencies of the tube.
+\valkyrie{need to provide some guidelines for the number of IDs we can create in this way, and where in the sound spectrum they are.
+my sense is that this will depend highly upon the i/o sensitivities of the mic and speaker selected.}
 
@@
 
+This is similar to work done by Willis, et al., in \cite{Willis-infrastructs}.
+While their technique requires access a terahertz imaging tool, audio resonance identity encoding requires only a microphone and a speaker.
+
 \valkyrie{in my experiments so far, it seems that the material printed does conduct sound fairly well, so this is feasible.

### 2014-02-27 paper/relatedwork.tex
more paper text, algo and proof for neon

--- 2014-02-21 paper/relatedwork.tex
@@
 
-Our work is related to three existing research areas: fabrication, interaction techniques, and routing.
+While using a series of tubes can replicate the interaction techniques discussed (and more), our work's \emph{contribution} is more generally related to two existing research areas: fabrication and routing.
 
@@
 \item \cite{Savage-sauron} - Sauron.  computer vision of mechanical components obviates electronics installation.
-\item \cite{Willis-infrastructs} - InfraStructs, identification of 3d printed objects using Terahertz imaging
 \item \cite{Willis-printedoptics} - Printed Optics.  doing cool stuff with clear material (touch sensing and display)
@@
 they don't offer a routing algorithm or anything, though.
-\end{itemize}
-
-\subsubsection{Printing Electronics}
-\begin{itemize}
 \item \cite{Sarik-tracebrush} - this uses an airspray machine to add conductive paths to surface of 3D prints
@@
 \item \cite{Majidi-curvature} - inject liquid metal into really thin tubes in a soft substrate, sense stretching by changing resistance
+
+Our neon routing algorithm is similar to that described by Wong, et al., in \cite{Wong-continuousline}.
+However, We are not confined to the plane to avoid line-crossings, and we also have additional freedom in creating lines where none existed.
+Because we can shield neon post-print with black tape or material (thus rendering it invisible), we don't have the same need to avoid drawing new line segments, and we are free to create much shorter paths.
+It's not quite a TSP
 \end{itemize}
-
-\subsection{Interaction Techniques}
-\begin{itemize}
-\item \cite{Sato-touche} - Touch\'{e}.  It's like Touch\'{e} with sound, but without sound.  SFCS.
-\item \cite{Harrison-buttons} - fabricate latex + acrylic buttons and pressurize with air
-\item \cite{Sodhi-aireal} - air vortex generation in free space.  air haptics.
-\item \cite{Yao-pneui} - PneUIs, creating interfaces with pneumatics
-\item \cite{Slyper-shape} - creating silicone bendy things with embedded electronics to sense flexing, stretching, etc., supported by those shapes.
-\item \cite{Iwata-volflex} - a display made up of many balloons that inflate and deflate to change the shape
-\item \cite{Ono-touchandactivate} - the Touch\'{e} with sound paper from last year's UIST
-\item \cite{Follmer-jamming} - Jamming User Interfaces
-\item \cite{Kim-inflatablemouse} - a basic mouse, but it inflates so you can store it and also use it more reasonably than a flat mouse
-\item \cite{Hashimoto-squirming} - hold a speaker in your hands, and air pressure changes make it feel like you're holidng a living, squirming thing
-\end{itemize}

### 2014-02-27 paper/tubes.tex
more paper text, algo and proof for neon

@@
 
-\subsection{Tool for Tube Design}
+\section{Tool for Tube Design}
 
@@
 
-\subsubsection{Selection}
-
-\textbf{Exterior Connection Points}
+\subsection{Exterior Connection Points}
 
@@
 
+\subsubsection{Routing}
+
 Focus on shape of points and location of points.
-
-\textbf{Designing Interior Paths}
-
-\begin{figure}[h!]
-\centering
-    \includegraphics[width=3.4in]{figures/series-of-tubes.jpg}
-\caption{This figure has several sub-figures.
-a) shows a model with path shape and exterior connection point(s) brushed.
-b) highlights the points which can't be tubed as drawn (if they are not connected, too tight, etc.).
-c) shows our physics-based, energy-minimizing rod/tube and any path-smoothing that we do, d) shows the printed object with the tube (with something in it, EL wire presumably)}
-\label{fig:tool-process-interior}
-\end{figure}
-
-take into account bend radius of desired material - no need, we just do the minimum-bending path
 
@@
 
-\subsubsection{Mesh Modification}
+\subsection{Designing Interior Paths}
+
+\begin{figure}[h!]
+\centering
+    \includegraphics[width=3.4in]{figures/series-of-tubes.jpg}
+\caption{This figure has several sub-figures.
+a) shows a model with path shape and exterior connection point(s) brushed.
+b) highlights the points which can't be tubed as drawn (if they are not connected, too tight, etc.).
+c) shows our physics-based, energy-minimizing rod/tube and any path-smoothing that we do, d) shows the printed object with the tube (with something in it, EL wire presumably)}
+\label{fig:tool-process-interior}
+\end{figure}
+
+take into account bend radius of desired material - no need, we just do the minimum-bending path
+
+Our problem here is a relaxed version of the Chinese Postman Problem$^{[\valkyrie{citation needed}]}$ in which we wish to traverse all edges of the graph described in the user's input (as the edges of the graph are the lines the user wishes to have lit) with a single circuit.
+It is a relaxation in that we allow the creation of non-existing paths: if components are disconnected (e.g., a user wishes to spell ``UIST" in neon lights), we must create edges that connect them; additionally we can create edges to connect odd-degree vertices rather than simply retracing existing edges.
+In the final design, all created edges will be shielded by dark material so the inserted medium is not visible.
+
+We want a semi-Eulerian graph (i.e., we want a graph in which every vertex but two have even degree) so that the desired medium can be inserted, traverse every edge, and exit the graph at a different vertex.
+
+Let $G=(V,E_0)$ s.t. $\forall e \in E_0, weight(e)=0, start \in V $ the start point $end \in V$ the end point.  Let $e_{temp} = (start, end)$, $E' = e_{temp} \cup E$.
+
+We need to connect disconnected subgraphs in $G$.
+Let $G_{dis} = \{G_1, G_2, ...
+G_n\} \in G$ s.t.
+$\cup{\{G_i \in G_{dis}\}} = G$ and $G_i \cap G_j  = \{ \} \forall G_i, G_j \in G_{dis}, i\neq j$ be disconnected subgraphs in G.
+Create $E_{conn} = \{e = (u, v), u \in G_i, v \in G_j, i \neq j, weight(e) = distance(u, v)\}$ be a set of edges.
+Sort $E_{conn}$ s.t.
+$E_{conn} = \{e_1, e_2, ...
+e_n\}, weight(e_1) \leq weight(e_2) \leq \ldots \leq weight(e_n)$.
+Let $E_{conn-min} = \{ \}$.
+
+Beginning with $e_1$, add the first edge $e_i \in E_{conn}$ to $E_{conn-min}$.
+Then let $E_{dupe} = \{ e = (u, v) \in E_{conn}$ s.t.
+$u$  is reachable from $v$ along edges $E' \cup E_{conn-min}$.
+Let $E_{conn} = E_{conn} \setminus E_{dupe}$.
+Repeat until $E_{conn} = \{ \}$.
+Let $E' = E' \cup E_{conn-min}$.
+
+At this stage, we want an Eulerian graph: i.e., we need every vertex to be of even degree.
+Let $V_{odd} = \{ v \in V s.t.
+degree(v) \% 2 = 1 \}$.
+Create $E_{circuit} = V_{odd} \times V_{odd}$, s.t.
+$\forall e=(u,v) \in E_{circuit}, weight(e) = distance(u, v)$.
+Find minimum matching $E_{circuit-min} \subseteq E_{circuit}$.
+Let $E' = (E\setminus e_{temp}) \cup E_{circuit-min}$.
+
+\begin{theorem}
+ $G' = (V, E')$ is a connected, semi-Eulerian graph for which $E \subseteq E'$.
+\end{theorem}
+\begin{proof}
+$E \subseteq E'$ : we never remove edges from $E$ in our algorithm.  $\therefore E \subseteq E'$.
+
+$G' = (V, E')$ is connected : if $G'$ is not connected, $\exists G_i = (V_i, E_i) \subset G', u \in V_i, v\in V \setminus V_i$ s.t.
+$u$ is not reachable from $v$.
+$G_i \not\in G_{dis}$, because we create edges connecting $G_j, G_k \forall G_j, G_k \in G_{dis}$ and only remove an edge $e_{dupe} = (u, v)$ from $E_{conn}$ once we determine that $u$ is reachable from $v$ along edges $E' \cup E_{conn-min}$.
+$\therefore G_i \not\in G_{dis}$.
+This implies that $G_i$ was not initially disconnected from $G$, because by definition $G_{dis}$ is the set of all disconnected subgraphs of $G$.
+We cannot have disconnected $G_i$ from $G$ because $E \subseteq E'$.
+Thus, a contradiction.
+$\therefore G' = (V, E')$ is connected.
+
+$G' = (V, E')$ is semi-Eulerian : Each edge $e$ in our minimum weight matching connects a unique pair of vertices $v_i, v_j \in V_{odd}$ by the definition of minimum weight matching.
+Each edge $e = (v_i, v_j) \in E_{circuit-min}$ adds one to the degree of $v_i$ and $v_j$, causing them to be of even degree.
+$|V_{odd}| \% 2 = 0$ by the handshake lemma, $\therefore$ all edges can be paired.
+When we remove $e_{temp} = (start, end)$ from $E'$, we cause those vertices, which were of even degree by the above process, to be of odd degree.
+Thus, all vertices except $start$ and $end$ (which are of odd degree) are of even degree.
+$\therefore G'$ is semi-Eulerian.
+\end{proof}
+
+\subsection{Mesh Modification}
 

### 2014-02-27 paper/relatedwork.tex

--- 2014-02-27 paper/relatedwork.tex
@@
 \end{itemize}
+
+Our neon routing algorithm is similar to that described by Wong, et al., in \cite{Wong-continuousline}.
+However, We are not confined to the plane to avoid line-crossings, and we also have additional freedom in creating lines where none existed.
+Because we can shield neon post-print with black tape or material (thus rendering it invisible), we don't have the same need to avoid drawing new line segments, and we are free to create much shorter paths.
+It's not quite a TSP
 

### 2014-02-27 paper/tubes.tex
update of algorithm text

@@
 This tool allows users to brush over the surface of their model to either select exterior connection points of their tubes (see Figure \ref{fig:tool-process-interior}) or to author the tubes' interior paths (see Figure \ref{fig:tool-process-exterior}).
-Once the user's selections are made, our tool creates an initial shortest-path routing using A* to estimate the routed distance between points.
-This routing is used to create a rod; we run physics-based simulation steps on the rod to minimize its bending energy (and thereby minimize the bend radius of the tubes).
+Once the user's selections are made, we create a complete routing using either the exterior connection point method (A*) or the interior path method (graph edge creation, Euler circuit generation).
+We thicken our routing and use a physics-based rod simulation to minimize bending energy of tubes.
+We also apply templates where appropriate (e.g., 3D cross-overs for tubes that intersect).
 The resultant routing is subtracted from the user's mesh.
@@
 \subsection{Exterior Connection Points}
+
+our tool creates an initial shortest-path routing using A* to estimate the routed distance between points.
+This routing is used to create a rod; we run physics-based simulation steps on the rod to minimize its bending energy (and thereby minimize the bend radius of the tubes).
 
@@
 
-\subsubsection{Routing}
-
 Our basic first-pass routing algorithm uses the A* routing algorithm \cite{Hart-Astar}.
@@
 
+take into account bend radius of desired material - no need, we just do the minimum-bending path
+
+ (as the edges of the graph are the lines the user wishes to have lit)
+
 \begin{figure}[h!]
@@
 \caption{This figure has several sub-figures.
-a) shows a model with path shape and exterior connection point(s) brushed.
-b) highlights the points which can't be tubed as drawn (if they are not connected, too tight, etc.).
-c) shows our physics-based, energy-minimizing rod/tube and any path-smoothing that we do, d) shows the printed object with the tube (with something in it, EL wire presumably)}
+a) shows an SVG file with some line art  b) highlights the points which can't be tubed as drawn (if they are not connected, of odd degree, etc.).
+c) shows our connected graph, d) shows our routing e) shows a printed thing with that kind of tube filled with EL wire}
 \label{fig:tool-process-interior}
@@
 
-take into account bend radius of desired material - no need, we just do the minimum-bending path
-
-Our problem here is a relaxed version of the Chinese Postman Problem$^{[\valkyrie{citation needed}]}$ in which we wish to traverse all edges of the graph described in the user's input (as the edges of the graph are the lines the user wishes to have lit) with a single circuit.
-It is a relaxation in that we allow the creation of non-existing paths: if components are disconnected (e.g., a user wishes to spell ``UIST" in neon lights), we must create edges that connect them; additionally we can create edges to connect odd-degree vertices rather than simply retracing existing edges.
-In the final design, all created edges will be shielded by dark material so the inserted medium is not visible.
+This is a relaxed version of the Chinese Postman Problem\footnote{\url{http://en.wikipedia.org/wiki/Route_inspection_problem}}, in which we wish to traverse all edges of the graph described in the user's input with a single Euler circuit, much as a postman needs to walk along every road at least once to deliver mail.
+We call our relaxation the Spiderman Postman Problem: we allow the creation of non-existing paths (i.e., the postman may traverse buildings in addition to roads).
+If inputted path components are disconnected, we must create edges that connect them; additionally we can create edges to connect odd-degree vertices rather than simply retracing existing edges.
+In the final artifact, all created edges will be shielded by dark material so the inserted medium is not visible (see Figure \ref{fig:tool-process-interior}).
 
 We want a semi-Eulerian graph (i.e., we want a graph in which every vertex but two have even degree) so that the desired medium can be inserted, traverse every edge, and exit the graph at a different vertex.
-
-Let $G=(V,E_0)$ s.t. $\forall e \in E_0, weight(e)=0, start \in V $ the start point $end \in V$ the end point.  Let $e_{temp} = (start, end)$, $E' = e_{temp} \cup E$.
+Let $G=(V,E_0)$ s.t.
+$\forall e \in E_0, weight(e)=0, start \in V $ the start point $end \in V$ the end point.
+Let $e_{temp} = (start, end)$, $E' = E + e_{temp}$.
 
@@
 Find minimum matching $E_{circuit-min} \subseteq E_{circuit}$.
-Let $E' = (E\setminus e_{temp}) \cup E_{circuit-min}$.
+Let $E' = E \cup E_{circuit-min}$.
+
+Now to remove our temporary edge and make the graph semi-Eulerian instead of Eulerian, let $E' = E' \setminus e_{temp}$.
 
@@
 
+We believe that a lower total weight matching is possible by connecting components and ensuring edge degree evenness together in a global process, however it this not crucial for our purposes.
+
 \subsection{Mesh Modification}

### 2014-03-05 paper/designspace.tex
minor modifications

@@
 
-While possible to create, a tube which is open on the user side and closed on the system side is outside our focus is on computer-mediated interaction, and we therefore do not discuss these tubes.
+While possible to create, a tube which is open on the user side and closed on the system side is outside our focus on computer-mediated interaction, and we therefore do not discuss these tubes.
 

### 2014-03-05 paper/tubes.tex
minor modifications

@@
 We call our relaxation the Spiderman Postman Problem: we allow the creation of non-existing paths (i.e., the postman may traverse buildings in addition to roads).
-If inputted path components are disconnected, we must create edges that connect them; additionally we can create edges to connect odd-degree vertices rather than simply retracing existing edges.
+If inputted path components are disconnected, we must create edges that connect them; additionally we can create new edges connecting odd-degree vertices rather than simply retracing existing edges.
 In the final artifact, all created edges will be shielded by dark material so the inserted medium is not visible (see Figure \ref{fig:tool-process-interior}).
@@
 Let $E_{conn-min} = \{ \}$.
+\valkyrie{another approach to this that actually would minimize it is to reduce each disconnected subgraph to a single point with one edge going to each other disconnected subgraph-point.
+these edges' weights would be the lowest weight of any edge connecting any points in the two subgraphs.
+by finding a minimum matching on these subgraph points, I think we will get the best answer.}
 
@@
 
+Once we have a semi-Eulerian graph, we need to create an Euler tour.
+An Euler tour is a path that touches every edge once, beginning and ending at the two nodes of odd degree: in this case, those are the user-selected start and end nodes.
+We use a sligh modification of Fleury's algorithm\footnote{\url{http://en.wikipedia.org/wiki/Eulerian_path\#Fleury.27s_algorithm}} for this, where instead of randomly selecting from non-bridge paths at each node, we select the path which turns the least from the most recent path (i.e., we prefer to pass straight through a node, if possible).
+This minimizes turns in the final artifact, which eases support material removal and assembly.
+
 \subsection{Mesh Modification}

### 2014-03-07 paper/introduction.tex

@@
 We see many opportunitities to increase the interactivity of 3D prints using today's printer technology.
+\tovi{Might want some examples of what possibilities this opens up}
 
@@
 In addition, they lack flexibility: the printed circuits must be routed in 2D to be created on the objects' manifold exteriors, and they can only be used to create electronic circuits.
-\valkyrie{what?}
+\valkyrie{what?}  \tovi{I dont follow this paragraph - I think you want  to briefly summarize what has been previousl done in interactive 3d prints, and hihglight the shortcommings and remianing challenges}
 
@@
 \valkyrie{Points to hit: there is more than electronics to worry about, like fluids and the organic haptic feedback they afford; tangible interfaces are important (do I have to say this every time??); assembly sucks}
+
+In this paper, we propose a new technique, where interior channels, such as hollow champers or solid, are integrated into 3D models.
+Such channels introduces a entire design space of opportunities for adding interactvity to 3D printed objects.
+For example, copper material can fill the channels, to allow for standard electronic compoents to be easily integrated after the printout; air can be pumped through to create tactile output; or, Electroluminescent wire can be passed through to create interactive visual output.
+
+Our work makes the following contributions to the field of interactive digitial fabircation.
+First, ...
+deisgn space...Second ...
+modeling tool and algorithms ...
+Third showcase a set of examples, enabled by our modeling tool.

### 2014-03-08 paper/abstract.tex
paper: change tovi color, add lots of text, update routing algorithm

@@
 One allows users to select begin and end points for tubes, then uses A* path routing and physics-based simulation to minimize the bending energy of routed paths.
-The second allows users to enter a description of paths to follow: for this we developed a novel neon sign routing algorithm and offer proof of its optimality.
+The second allows users to enter a description of paths to follow: for this we developed a novel neon sign routing algorithm and offer proof of several key traits.
 We present several totally tubular prototypes we created using this tool to show its flexibility and potential, as well as to explore new points in the tube design space.

### 2014-03-08 paper/conclusion.tex
paper: change tovi color, add lots of text, update routing algorithm

@@
 \section{Conclusion}
-In conclusion, tubes are cool.
-You can do lots of things with them.
-There are different kinds and topologies of tubes.
-There are different things you can put in them.
-This is cool for makers and also people with really expensive machines.
+While today's 3D printers are not yet able to fabricate active components in-place, we suggest that 3D printed interactive devices can be created with redirection of active input and output via tubes.
+In addition to describing the design space of tubes and connecting related works created through both traditional processes and rapid prototyping methods, we also presented a design tool for the creation of tubes inside arbitrary 3D models.
+We fabricated several example objects using this tool to explore new points in the design space.
+
+We are heartened by our success: it opens new opportunities both for makers with consumer-grade 3D printers and research and industrial labs with higher-end machines.
+We are excited to release our tool into the wild and see what people will build.
 

### 2014-03-08 paper/designspace.tex
note about reference
paper: change tovi color, add lots of text, update routing algorithm

@@
 
+\subsubsection{Dimensions of Tubes}
+
+\valkyrie{Reference those pages from Bjoern: http://ttb.com/recent-news/diagrams/ We can talk about wall thickness, tangent length, degree of bend, arg bend angle, inner and outer diameters, centerline radius, etc.}
+
 \subsection{Inputs and Outputs}
-
-\subsection{Interaction Techniques}
-These didn't get discussed yet.
-\begin{itemize}
-\item \cite{Iwata-volflex} - a display made up of many balloons that inflate and deflate to change the shape
-\item \cite{Kim-inflatablemouse} - a basic mouse, but it inflates so you can store it and also use it more reasonably than a flat mouse
-\item \cite{Hashimoto-squirming} - hold a speaker in your hands, and air pressure changes make it feel like you're holidng a living, squirming thing
-\end{itemize}
 
@@
 Traditional wires (threadable) may be used, but we have had success with conductive paint (liquid).
-Using conductive paint and Swept Frequency Capacitve Sensing (SFCS \cite{Sato-touche}), an interior star tube topology can enable single-wire touch- and grasp-sensing at any set of points on a printed object's surface (see Figure \cite{fig:toy}).
+Using conductive paint and Swept Frequency Capacitive Sensing (SFCS \cite{Sato-touche}), an interior star tube topology can enable single-wire touch- and grasp-sensing at any set of points on a printed object's surface (see Figure \cite{fig:toy}).
 Simple capacitance measurements are possible using the same media, and custom-designed sensors like those created in Savage, et al.,'s Midas system are also possible \cite{Savage-midas}.
@@
 EL wire can be threaded through custom paths to create neon signs, as in Figure \ref{fig:neon}.
-Colored liquids can be mixed and split before exiting an opaque device, or in a transparent device their mixing, splitting, and paths could be used for dispaly.
-PneUIs \cite{Yao-pneui} can be fabricated as single parts: the rubberlike material offered by our Objet is inflatable when thin (200\% elongation at break), and a PneUI located at the terminus of a tube could be actuated by a system-side air pump (see Figure \ref{fig:braille}).
+Colored liquids can be mixed and split before exiting an opaque device, or in a transparent device their mixing, splitting, and paths could be used for display.
+PneUIs \cite{Yao-pneui} can be fabricated as single parts: the rubberlike material offered by our Objet is inflatable when thin (200\% elongation at break), and a PneUI located at the terminus of a tube could be actuated by a system-side air pump (see Figure \ref{fig:breathe}).
 Mechanical motion can be mediated through fluids (a light object which is partially contained in a tube or located at its end could be pushed by the fluid pressure of the tube) or threadables (e.g., muscle wire).
@@
 \valkyrie{need to get a test print of this that actually works}}
-\label{fig:embeddedcomponents}
+\label{fig:ocarina}
 \end{figure}
@@
 Fluids can actuate particulates and create programmable pliability (as discussed above and in \cite{Follmer-jamming}).
-Additionally, use of semi-closed tubes with rubberlike material as the caps allows for gases to actuate surface features (see Figure \ref{fig:braille}).
+Additionally, use of semi-closed tubes with rubberlike material as the caps allows for gases to actuate surface features (see Figure \ref{fig:breathe}).
 Sodhi, et al., in \cite{Sodhi-aireal} created free-air haptic feedback using controlled vortex generation: this technique could also be reproduced using a series of tubes.
+Iwata, et al., explored volumetric fluid-based interactions with programmatically inflatable balloons and an overhead projector in \cite{Iwata-volflex}, and additionally haptic outputs can be used to increase ergonomics, as in the inflatable mouse by Kim, et al.
+\cite{Kim-inflatablemouse}.
 
@@
 As seen in \cite{Ono-touchandactivate}, different objects have different acoustic signatures.
-Additionally, two objects that are visually identical but which have fully enclosed (or other types) of tubes on their interior can be distinguised acoustically.
+Additionally, two objects that are visually identical but which have fully enclosed (or other types) of tubes on their interior can be distinguished acoustically.
 Thus we can recall an object's identity once its acoustic signature has been recorded.

### 2014-03-08 paper/exampleobjects.tex
paper: change tovi color, add lots of text, update routing algorithm

@@
 
-\subsection{Touch-sensitive Toy}
+\subsection{Touch-sensitive Toys}
 
@@
     \includegraphics[width=3.4in]{figures/series-of-tubes.jpg}
-\caption{A touch-sensitive rabbit whose tubes are filled with conductive paint.
+\caption{A touch-sensitive brain whose tubes are filled with conductive paint.
 Sensing is done on a single wire via SFCS.
@@
 
-We created a touch-sensitive toy and an app that goes along with it, reminiscent of the boat application in \cite{Harrison-acoustic}.
-This toy has an interior star topology where a single wire attached to an Arduino running an SFCS sketch splits to connect to sensors on the eyes, ears, tail, and nose.
-To accompany this toy, we built an app which prompts the user to find and touch certain parts of the animal (see Figure \ref{fig:toy}).
+We created a set of touch-sensitive toys and a companion app, reminiscent of the boat application in \cite{Harrison-acoustic}.
+The brain, rabbit, and boat each have several distinct touch points on their surfaces.
+These touch points are connected by an interior star topology of tubes filled with conductive copper paint, and touch sensing is performed via a single wire and SFCS (see Figure \ref{fig:toy}).
+We built a smart base which can distinguish between the toys and also determine which toy is mounted: since each toy and each gesture has a distinct capacitive signature, we use a simple classifier trained to detect both toy and gesture based on profile.
+These devices were all fabricated on a Makerbot Replicator 2.
 
-\subsection{Braille Haptic Display}
+\subsection{Breathing Bunny}
 
@@
     \includegraphics[width=3.4in]{figures/series-of-tubes.jpg}
-\caption{When air pressure is changed in the six tubes, their caps inflate or deflate.
-Through controlled actuation, we can create Braille letters.
-Inset shows the internal structure of the tubes generated by our design tool.}
-\label{fig:braille}
+\caption{This rabbit has an open tube leading to its mouth and a semi-closed tube capped with flexible material in its back.
+When air is sucked in through the open tube, the back becomes stiff as though the rabbit has breathed in.}
+\label{fig:breathe}
 \end{figure}
 
-Using a $2\times3$ array of semi-closed tubes, we created a Braille output interface.
-The closed ends of the tubes are covered in a rubberlike material, while on the system side the tubes are attached to individually-controlled air pumps.
-These pumps can create positive or negative pressure in each tube, pushing the cap up or down and rendering a tactile Braille letter when used in concert (see Figure \ref{fig:braille}).
+We created a rabbit with a pair of tubes that can simulate breathing (see Figure \ref{fig:breathe}).
+For this, we used a combination air/vacuum pump: one terminal creates a vacuum while the other creates positive pressure.
+Our rabbit has two tubes, one open tube exiting at its mouth and one semi-closed tube capped with rubberlike material in its abdomen.
+We connected one tube to each of our pump's terminals, and using a programmable power supply we can mimic a rabbit's breathing pattern.
+This example was printed on our Objet Connex 260.
+\valkyrie{this paragraph is really unclear}
 
@@
 
-A custom ``radio'' built using tubes allows users to tune to different stations and play sound files related to those stations.
-This device uses a network of disconnected open tubes filled with conductive paint to connect traditional electronic components (a potentiometer, a piezo, and an LED) to a microcontroller.
+A custom radio built using tubes allows users to tune to different stations and listen in.
+This device uses a network of disconnected open tubes filled with conductive paint to connect traditional electronic components (two potentiometers, a speaker, and an LED) to a microcontroller.
+We hid the microcontroller and a battery-driven power supply in the base of the radio to make it portable.
+We fabricated this device on our Makerbot Replicator 2.
 
@@
 A dark point on the camera region appears when a pen has been removed from its place, and the position of the point indicates which pen it is.
+This prototype was built by our Objet Connex 260.
+
+\subsection{Maze}
+
+\begin{figure}[h]
+\centering
+    \includegraphics[width=3.4in]{figures/series-of-tubes.jpg}
+\caption{Our maze game has a large single particle trapped inside a fully enclosed tube.  It offers hours of fun!}
+\label{fig:maze}
+\end{figure}
+
+We created a maze game with a single particle trapped in a fully enclosed tube.
+To fabricate the tube, we created it in two halves that were fastened together via glue, due to limitations in our ability to remove support material otherwise.
+We created this prototype on our Makerbot Replicator 2.
 
@@
     \includegraphics[width=3.4in]{figures/series-of-tubes.jpg}
-\caption{A neon sign (a) and its partial duplicate designed using our tool (b).
-(c) shows the selections we made to generate our tube structures.
-(d) shows the internal structure of the tubes generated by our design tool.}
+\caption{A two-state animated neon sign designed using our tool (a)(b) \valkyrie{Tovi: we could use your PDF has to make this animate in place :)}.
+(c) shows the SVG input files to create this sign, and (d) shows an isometric view of the tubes created by our tool.}
 \label{fig:neon}
@@
 For this type of art, the path of the tubes is of crucial importance, as it determines how the sign will look.
-We designed a partial duplicate of a renowned piece of neon art from Seattle (or was it Portland?).
+We designed a custom neon sign bearing the UIST logo and the waves of Hawaii.
 The piece features open tubes due to support-flushing constraints, however the tubes could also be semi-closed.
 They have been threaded with EL wire which is lit in sequence to create an animation.
+Our sign was fabricated on the Objet Connex 260.

### 2014-03-08 paper/introduction.tex
paper: change tovi color, add lots of text, update routing algorithm

@@
 Willis, et al., ``envision a future world where interactive devices can be printed rather than assembled; a world where a device with active components is created as a single object, rather than a case enclosing circuit boards and individually assembled parts" \cite{Willis-printedoptics}.
-This is a vision we eagerly share: 3D printers are capable of greating arbitrary geometries not feasible to manufacture using traditional processes, and we see these capabilities being underutilized by makers and designers.
-We see many opportunitities to increase the interactivity of 3D prints using today's printer technology.
+This is a vision we eagerly share: 3D printers are capable of creating arbitrary geometries not feasible to manufacture using traditional processes, and we see these capabilities being underutilized by makers and designers.
+We see many opportunities to increase the interactivity of 3D prints using today's printer technology.
 \tovi{Might want some examples of what possibilities this opens up}

### 2014-03-08 paper/limitations.tex
paper: change tovi color, add lots of text, update routing algorithm

@@
 On the Objet, support-free prints cannot overhang more than $14^{\circ}$.
-Removal of support from printed tubes can be complicated, especially when their internal geometry and branching structures are complex.
+Removal of support from printed tubes can be troublesome, especially when their internal geometry and branching structures are complex.
 
@@
 It lacks X and Y \valkyrie{probably it will lack the network manipulation to make mixing or splitting tubes...?}.
+Additionally, as mentioned there are opportunities to improve the algorithm for neon sign routing, as ours does not guarantee a shortest-path routing.
 We hope to address this in future work.

### 2014-03-08 paper/relatedwork.tex
update of algorithm text
paper: change tovi color, add lots of text, update routing algorithm

--- 2014-02-27 paper/relatedwork.tex
@@
 Previous work has investigated the integration of interactivity into the fabrication process.
-\cite{Willis-printedoptics} used light pipes, solid clear tubes, to create integrated displays.
+\cite{Willis-printedoptics} used light pipes, solid clear tubes, to create integrated displays: our work allows a superset of these capabilities, and Printed Optics is a part of our design space.
 \cite{Savage-sauron} investigated the use of computer vision to track interactions with physical mechanisms.
+Using tubes for redirection, it is possible to replicate these mechanically sensitive techniques (see Figure \ref{fig:pens}), however our work does not focus on computer vision alone.
+Slyper, et al., \cite{Slyper-pressure} 3D printed flexible robot armatures capable of sensing interaction via changes in air pressure.
+This technique is complementary to ours, as we allow for placing these interactive elements anywhere on an object's surface while they had to place them in locations accessible for electronics insertion.
+
+Other research has attempted to bring traditional electronics to 3D prints, either on the surface or via interior channels.
+Sells, et al., in \cite{Sells-reprap} create channels on the surface of a 3D printed object which are filled with a special computer-controlled syringe of liquid solder to create connections.
+Their focus is on creating self-replicating 3D printers, and their electronics are combined to traditional 2D routed circuitboards.
+Sarik, et al., \cite{Sarik-tracebrush} \valkyrie{I never found a real paper of this.
+The Microsoft website has a "print" of it with TEI boilerplate, but I couldn't find it in the digital library.
+What's up with that?} and the Optomec company \cite{optomec} both created techniques for taking an already-printed object and spraying conductive material onto its surface.
+In contrast, our technique can lead to more durable circuitry, since it is routed through the core of the object rather than on its surface.
+Navarette, et al., in \cite{Navarrette-gps} created a functioning GPS whose circuit routing extended into true 3D (as compared to the 2.5D boards created by stacking traditional 2D circuitboards).
+However, they created this one-off routing by hand and do not offer any general algorithms or techniques.
+In the area of robotics, softer-than-skin interactive surfaces have been created using tubed interfaces fabricated from silicone molds.
+\cite{Park-microchannels} and \cite{Majidi-curvature} use microchannels filled with liquid metal for pressure or curvature sensing: the metal in the channels changes resistance as the channels are stretched.
 \begin{itemize}
-\item \cite{Savage-sauron} - Sauron.  computer vision of mechanical components obviates electronics installation.
-\item \cite{Willis-printedoptics} - Printed Optics.  doing cool stuff with clear material (touch sensing and display)
-\item \cite{Slyper-pressure} - robots that are made of squishy stuff where air pressure changes are sensed.
-input components designed to react to different manipulations (pushing, squeezing, twisting, etc.)
-\item \cite{Navarrette-gps} - I'm a little unclear on what they did, but they fabbed something with a "3D circuitboard" that has a bypass that goes into 3D.
-they don't offer a routing algorithm or anything, though.
-\item \cite{Sarik-tracebrush} - this uses an airspray machine to add conductive paths to surface of 3D prints
-\item \cite{Optomec} - fancy machines that spray conductive film
-\item \cite{Sells-reprap} - the RepRap people using a syringe of hot solder to squirt flat circuits into flat channels
 \item \cite{icecats-conductivepaint} - an instructable I followed to make my first batch of conductive paint
@@
 \subsection{Routing}
+
+Part of our contribution is in the algorithms and techniques used for routing tubes through 3D objects.
+
+General-purpose routing for connecting 2D capacitive sensors to terminals is discussed in \cite{Savage-midas}.
+These techniques, based on circuitboard routing, are limited to planar and unfoldable shapes, and are intended for use on the surface of objects rather than through their cores.
+
+Similar to our neon-sign routing algorithm, Wong presents a graph-based approach for generating continuous line illustrations from images \cite{Wong-continuousline}.
+Its approach of linking edges into a continuous line based on semi-Eulerization is similar to ours.
+However, we are not confined to the plane to avoid line-crossings, and we have additional freedom in creating invisible lines where none existed.
+This is possible for us thanks to \emph{blocking out}\footnote{In neon sign making, segments of a tube that should not appear illuminated are covered with opaque paint.
+This is called "blocking out".\url{http://en.wikipedia.org/wiki/Neon_sign}}.
+Prior work used tours of nodes, rather than edges, and cast continuous line drawing as a traveling salesman problem~\cite{Bosch-tsp}.
+Schweikardt, et al., also made a connection between electroluminescent wire and graph theory, though their contribution is orthogonal to ours: they provide a tangible construction kit with smart nodes for building physical graph examples \cite{Schweikardt-tei09}.
+Edges are lit by EL wire.
+
+\subsection{Misc}
+\bjoern{These may not belong in related work, but I'm just collecting them here.}
 \begin{itemize}
-\item \cite{Savage-midas} - Midas.  routing in 2D to connect up capacitive sensors
-\item \cite{Park-microchannels} - similar to injection of liquid metal, above
-\item \cite{Majidi-curvature} - inject liquid metal into really thin tubes in a soft substrate, sense stretching by changing resistance
-
-Our neon routing algorithm is similar to that described by Wong, et al., in \cite{Wong-continuousline}.
-However, We are not confined to the plane to avoid line-crossings, and we also have additional freedom in creating lines where none existed.
-Because we can shield neon post-print with black tape or material (thus rendering it invisible), we don't have the same need to avoid drawing new line segments, and we are free to create much shorter paths.
-It's not quite a TSP
+\item \cite{strattman1997neon} - Strattman is the standard "text book" for neon lighting designers.
 \end{itemize}

### 2014-03-08 paper/tubes.tex
paper: change tovi color, add lots of text, update routing algorithm

@@
-\section{A Series of Tubes}
+\section{A Series of Tubes - A Tool for Tube Design}
 
-\section{Tool for Tube Design}
-
-We created a tool which novice designers can use to create tube-powered interfaces in arbitrary 3D models.
-This tool allows users to brush over the surface of their model to either select exterior connection points of their tubes (see Figure \ref{fig:tool-process-interior}) or to author the tubes' interior paths (see Figure \ref{fig:tool-process-exterior}).
+We created a tool which makers can use to create tube-powered interfaces in arbitrary 3D models.
+This tool allows designers to brush over the surface of their model to select exterior connection points of their tubes (see Figure \ref{fig:tool-process-interior}) or to import vector art describing the tubes' interior paths (see Figure \ref{fig:tool-process-exterior}).
 Once the user's selections are made, we create a complete routing using either the exterior connection point method (A*) or the interior path method (graph edge creation, Euler circuit generation).
 We thicken our routing and use a physics-based rod simulation to minimize bending energy of tubes.
-We also apply templates where appropriate (e.g., 3D cross-overs for tubes that intersect).
-The resultant routing is subtracted from the user's mesh.
+We also apply templates where appropriate (e.g., 3D cross-overs for tubes that intersect in the plane).
+The resultant routing is subtracted from the original mesh.
 The modified mesh can then be 3D printed.
+
+Our tool is implemented as a part of Meshmixer, a consumer mesh editing tool, in C++.
 
@@
 
-our tool creates an initial shortest-path routing using A* to estimate the routed distance between points.
+Our tool creates an initial shortest-path routing using A* to estimate the routed distance between points.
 This routing is used to create a rod; we run physics-based simulation steps on the rod to minimize its bending energy (and thereby minimize the bend radius of the tubes).
@@
     \includegraphics[width=3.4in]{figures/series-of-tubes.jpg}
-\caption{This figure has several sub-figures.
-a) shows a model with exterior connection points brushed.
-b) shows initial routing with A*.
-c) shows our physics-based, energy-minimizing rod/tube, d) shows the printed object with the tube (with something in it, copper paint presumably)}
+\caption{An example mesh with exterior connection points brushed (a), and the initial A* routing generated (b).
+Our physics-based, energy-minimizing rod/tube at rest (c) and after simulation (d).
+(e) shows the complete printed object.}
 \label{fig:tool-process-exterior}
@@
 
-How does it actually work?
+How does it actually work? \valkyrie{Ryan?}
 
@@
 
-Collision detection
+In order to force our simulated rods away from the surface of the mesh, we create ``wind'' pushing inward from the exterior of the mesh.  This wind pushes the tube away from the surface.
+
+\subsubsection{Tubes with Multiple Endpoints}
+
+To allow designers to create tubes with multiple endpoints (like the star topologies in Figure \ref{fig:toys}), we ....????
+\valkyrie{create 0-size mesh points in the center of the object, to which we attach one end of every tube?}
 
@@
 
-take into account bend radius of desired material - no need, we just do the minimum-bending path
-
- (as the edges of the graph are the lines the user wishes to have lit)
+A user can import a vector graphics file (such as SVG) describing the path she desires for her tubes.
+We create a graph based on this input data, then add edges to make it connected and semi-Eulerian.
+We create an Euler tour on the graph, and then thicken the path to create tubes.
+Templates are used to resolve tube crossings in the plane.
 
@@
     \includegraphics[width=3.4in]{figures/series-of-tubes.jpg}
-\caption{This figure has several sub-figures.
-a) shows an SVG file with some line art  b) highlights the points which can't be tubed as drawn (if they are not connected, of odd degree, etc.).
-c) shows our connected graph, d) shows our routing e) shows a printed thing with that kind of tube filled with EL wire}
+\caption{An input vector graphics file (a) and the points which cannot be tubed as drawn (b).
+The connected graph created by our software (c) and the resulting Euler circuit (d) permit creation of a novel neon sign (e).}
 \label{fig:tool-process-interior}
@@
 
-This is a relaxed version of the Chinese Postman Problem\footnote{\url{http://en.wikipedia.org/wiki/Route_inspection_problem}}, in which we wish to traverse all edges of the graph described in the user's input with a single Euler circuit, much as a postman needs to walk along every road at least once to deliver mail.
+The interior path routing problem is a version of the Chinese Postman Problem\footnote{\url{http://en.wikipedia.org/wiki/Route_inspection_problem}}, in which we wish to traverse all edges of the graph described in the user's input with a single Euler circuit, much as a postman needs to walk along every road at least once to deliver mail.
 We call our relaxation the Spiderman Postman Problem: we allow the creation of non-existing paths (i.e., the postman may traverse buildings in addition to roads).
 If inputted path components are disconnected, we must create edges that connect them; additionally we can create new edges connecting odd-degree vertices rather than simply retracing existing edges.
-In the final artifact, all created edges will be shielded by dark material so the inserted medium is not visible (see Figure \ref{fig:tool-process-interior}).
+In the final artifact, all created edges will be blocked out by dark material so the inserted medium is not visible (see Figure \ref{fig:tool-process-interior}).
 
@@
 $\cup{\{G_i \in G_{dis}\}} = G$ and $G_i \cap G_j  = \{ \} \forall G_i, G_j \in G_{dis}, i\neq j$ be disconnected subgraphs in G.
-Create $E_{conn} = \{e = (u, v), u \in G_i, v \in G_j, i \neq j, weight(e) = distance(u, v)\}$ be a set of edges.
+Create $E_{conn} = \{e = (u, v), u \in G_i, degree(u)\%2 = 1, v \in G_j, degree(v) \%2 = 1 i \neq j, weight(e) = distance(u, v)\}$ be the set of all edges connecting odd-degree points in two distinct disconnected subgraphs.
+To keep only the most desirable edges in $E_{conn}$, contract each subgraph $G_i$ to a single point $u_i$ with the shortest outgoing edges to each other subgraph: $E_{conn} = \cup_{G_i \in G_{dis}} \{e=(u_i, v), v \in G_j,$ s.t.
+$\forall E_{conn} \ni e_k = (u, v), u \in G_i, v \in G_j, weight(e_k) \geq weight(e)\}$.
 Sort $E_{conn}$ s.t.
@@
 e_n\}, weight(e_1) \leq weight(e_2) \leq \ldots \leq weight(e_n)$.
+
 Let $E_{conn-min} = \{ \}$.
-\valkyrie{another approach to this that actually would minimize it is to reduce each disconnected subgraph to a single point with one edge going to each other disconnected subgraph-point.
-these edges' weights would be the lowest weight of any edge connecting any points in the two subgraphs.
-by finding a minimum matching on these subgraph points, I think we will get the best answer.}
-
 Beginning with $e_1$, add the first edge $e_i \in E_{conn}$ to $E_{conn-min}$.
@@
 At this stage, we want an Eulerian graph: i.e., we need every vertex to be of even degree.
-Let $V_{odd} = \{ v \in V s.t.
-degree(v) \% 2 = 1 \}$.
+Let $V_{odd} = \{ v \in V$ s.t.
+$degree(v) \% 2 = 1 \}$.
 Create $E_{circuit} = V_{odd} \times V_{odd}$, s.t.
@@
 
-How we actually cut stuff and make tubes happen.
+How we actually cut stuff and make tubes happen. \valkyrie{Ryan?}
 

### 2014-03-08 paper/designspace.tex
more text, more and more and more

@@
 Also, I'm not confident this breakdown is the most clear: for example, the "liquids" category includes the copper paint, which functions as wires ("threadables"), but takes the form of a liquid...
-and the general presentation of the table could probably be better in some way.}}
+and the general presentation of the table could probably be better in some way.
+also, hello, make this wayyyyyy smaller.}}
 \label{fig:designspace}
@@
 Components can be recessed into a print's surface, with their leads implanted into open tubes.
-By inserting liquid copper paint instead of threading traditional wires, all components in a model can share a ground line, and the paint's drying process obviates the use of solder or glue to affix the components in place (see Figure \ref{fig:embeddedcomponents}).
-
-\begin{figure}[h]
-\centering
-    \includegraphics[width=3.4in]{figures/series-of-tubes.jpg}
-\caption{This sphere has a recessed space to hold a potentiometer.  The potentiometer and the LEDs share a ground, seen by the translucent rendered view in (b).}
-\label{fig:embeddedcomponents}
-\end{figure}
+By inserting liquid copper paint instead of threading traditional wires, all components in a model can share a ground line, and the paint's drying process obviates the use of solder or glue to affix the components in place (see Figure \ref{fig:radio}).
 

### 2014-03-08 paper/exampleobjects.tex
more text, more and more and more

@@
 Inset shows the internal structure of the tubes generated by our design tool.}
-\label{fig:toy}
+\label{fig:toys}
 \end{figure}
@@
 The brain, rabbit, and boat each have several distinct touch points on their surfaces.
-These touch points are connected by an interior star topology of tubes filled with conductive copper paint, and touch sensing is performed via a single wire and SFCS (see Figure \ref{fig:toy}).
+These touch points are connected by an interior star topology of tubes filled with conductive copper paint, and touch sensing is performed via a single wire and SFCS (see Figure \ref{fig:toys}).
 We built a smart base which can distinguish between the toys and also determine which toy is mounted: since each toy and each gesture has a distinct capacitive signature, we use a simple classifier trained to detect both toy and gesture based on profile.

### 2014-03-08 paper/tubes.tex
more text, more and more and more

@@
 
-Our tool creates an initial shortest-path routing using A* to estimate the routed distance between points.
+To create tubes in which the location and shape of exterior connection points matters, we first allow users to select the connection points on the mesh's surface using a brush.
+Our tool then creates an initial shortest-path routing using A* to estimate the routed distance between points.
 This routing is used to create a rod; we run physics-based simulation steps on the rod to minimize its bending energy (and thereby minimize the bend radius of the tubes).
@@
 
-Focus on shape of points and location of points.
-
 Our basic first-pass routing algorithm uses the A* routing algorithm \cite{Hart-Astar}.
 The path cost in our implementation is based only on shortest distance between the starting and ending points, without weighting for distance from the surface.
+We only permit the routing within the voxel grid of the mesh: i.e., the routing is not permitted to traverse through space outside the mesh itself.
+
+Given this initial routing, we create a rod of slightly longer proportional length.
+The rod must be longer, as the A* routing can follow the mesh boundaries and the rod requires extra length to be pushed away from the mesh surface.
+While we do not have a theoretical basis for this proportional increase, \valkyrie{yet.
+what might it be related to?
+we can test how much of the A* path is outside the model, and how much is inside; further we can consider the distance from the A* path that is outside the model to the surface.
+I think there's a good way to do this.
+It might even just mean iteratively changing the rod length until we get something reasonably short and also smooth.
+we can also return to the A* that I wrote which finds a routing through the mesh that respects its boundaries; this would get us closer to what we need for length in the first place.}, we have found 1.15 to be a reasonable multiplier in practice.
+A too-long rod is forced to kink and expand further than necessary inside the model, while a too-short rod must be stretched, leading to non-smooth bending.
 
@@
 We create a graph based on this input data, then add edges to make it connected and semi-Eulerian.
-We create an Euler tour on the graph, and then thicken the path to create tubes.
-Templates are used to resolve tube crossings in the plane.
+We create an Euler tour on the modified graph, and thicken the path to create tubes.
+3D templates are used to resolve tube crossings in the plane.
 
@@
 
+\valkyrie{does the proof of this theorem belong in the paper, or in the appendix?  we can push it back there if need be, although it's not super long.}
 \begin{theorem}
- $G' = (V, E')$ is a connected, semi-Eulerian graph for which $E \subseteq E'$.
+ $G' = (V, E')$ is a connected, semi-Eulerian graph for which $E \subseteq E'$. \valkyrie{why is there such a huge space after this paragraph?}
 \end{theorem}
@@
 An Euler tour is a path that touches every edge once, beginning and ending at the two nodes of odd degree: in this case, those are the user-selected start and end nodes.
-We use a sligh modification of Fleury's algorithm\footnote{\url{http://en.wikipedia.org/wiki/Eulerian_path\#Fleury.27s_algorithm}} for this, where instead of randomly selecting from non-bridge paths at each node, we select the path which turns the least from the most recent path (i.e., we prefer to pass straight through a node, if possible).
+We use a weighted modification of Fleury's algorithm\footnote{\url{http://en.wikipedia.org/wiki/Eulerian_path\#Fleury.27s_algorithm}} for this, where instead of randomly selecting from non-bridge paths at each node, we select the path which turns the least from the most recent path (i.e., we prefer to pass straight through a node, if possible).
 This minimizes turns in the final artifact, which eases support material removal and assembly.
+
+\begin{figure}[h!]
+\centering
+    \includegraphics[width=3.4in]{figures/series-of-tubes.jpg}
+\caption{When edges intersect in the plane (a), we match them to the nearest of our several intersection templates (b) to create 3D paths that do not interfere with each other.}
+\label{fig:templates}
+\end{figure}
+
+\subsubsection{Edges that Intersect in the Plane}
+Once we have a tour, we must resolve tubes that intersect in the plane.  Because we have a 3-dimensional canvas in which to work, we can push overlapping paths into the third dimension.
+
+To detect intersecting edges, we \valkyrie{I'm not sure what we do.}.
+We also consider edges at nodes with degree $>2$, as these edges are considered to cross.
+When intersections have been detected, we match one of several templates to the affected edges (see Figure \ref{fig:templates}) using a nearest neighbors technique.
+Our templates are parameterized, so the templates can be modified to accommodate precise intersection characteristics.
 
@@
 
-How we actually cut stuff and make tubes happen. \valkyrie{Ryan?}
+How we actually cut stuff and make tubes happen.
+\valkyrie{Ryan?}  We have to actually respect whatever shapes people drew on the surface in the case of the exterior connection tubes (I mean, ideally, we would do a loft operation here).
 
 \subsection{Fabrication Techniques}
+
+\valkyrie{this section may not be at all important.
+I was thinking to use it to talk about different ways you can actually build the tubed things, and also constraints set down by the Makerbot vs.
+the Objet, but maybe that discussion is best made really short and moved to elsewhere}
 

### 2014-03-12 paper/designspace.tex
integrate tovi comments

@@
 
-\subsection{Types of Tubes}
+\subsection{Types of Tubes}\tovi{Maybe rename types to 
+"openings"?}
 
-Tubes can come in four types: open, return, semi-closed, and fully enclosed.  Each of these types offers distinct interaction capabilities (see Figure \ref{fig:tubespace}).
+Tubes can come in four types: open, return, semi-closed, and fully enclosed.
+Each of these types offers distinct interaction capabilities (see Figure \ref{fig:tubespace}).\tovi{Need to clarify if the type is defined by the 3D model at the time it
+is printed, or by any alternations made after the print. For example, 
+if I print semiclosed, fill with water, then close it manually with a lit, 
+is it a semclsoed tube or a fully enclosed tube?}
 
-Open tubes originate from the system side and connect the user side, with both ends of the tube open.
-This type of tube may be used to create, for example, capacitive sensors: an open tube filled with conductive paint can be connected to a sensing platform (e.g., Arduino) on the system side, while a user can touch the uncapped other end of the tube.
+Open tubes originate from the system side and connect the user side, with both ends of the tube open. \tovi{Not clear what 
+"system side and user 
+site means.}  This type of tube may be used to create, for example, capacitive sensors: an open tube filled with conductive paint can be connected to a sensing platform (e.g., Arduino) on the system side, while a user can touch the uncapped other end of the tube.
 Using Swept Frequency Capacitive Sensing \cite{Sato-touche} or other techniques, a user's touch of the open end can be sensed.
@@
 Semi-closed tubes are open at the system end (for control of the enclosed medium) and closed at the user end.
-We believe this closed interface is most interesting when it is fabricated from a mobile material: for example, a series of tubes terminated in thin rubber membranes on the user side can be actuated by an air pump to create haptic feedback.
+We believe \tovi{Reword ("we believe" is strange)} this closed interface is most interesting when it is fabricated from a mobile material \tovi{Do you mean mobile materials or flexible material?}\valkyrie{I think mobile...
+we can cap them with solid pieces that can be blown around by air, for example.
+how do I make this clear?}: for example, a series of tubes terminated in thin rubber membranes on the user side can be actuated by an air pump to create haptic feedback.
 Without a printer capable of fabricating flexible material, a maker could affix a balloon to an open tube's end to behave similarly; another possibility is to use semi-closed tubes as audio-generating resonance chambers.
@@
 While possible to create, a tube which is open on the user side and closed on the system side is outside our focus on computer-mediated interaction, and we therefore do not discuss these tubes.
+\tovi{Reword, and may not be necceary}
 
@@
 Fully enclosed tubes can be used as resonance chambers (e.g., for object identification), or as air bubbles (e.g., as used in \cite{Willis-printedoptics} for internal display).
-Their physical design space is very limited, as any support material required to create their internal geometry cannot be accessed for removal.
+\tovi{used as a container for water
+or other particles that would 
+otherwise fall out?}  Their physical design space is very limited, as any support material required to create their internal geometry cannot be accessed for removal.
 
@@
 As in \cite{Slyper-pressure}, structures can be engineered to change in air pressure when manipulated correctly (e.g., a spiral that changes pressure when twisted, but not pressed), and thus fluid pressure can also be used as an input.
+\tovi{I realize "fluid" pressure is
+technically correct, but might 
+confuse readers, since this is 
+the gas pararaph.}
 
@@
 
-Tubes need not have hollow centers: in the case where routed tubes are filled with solid material---in particular, a solid material different from the model material---, interactions such as those in \cite{Willis-printedoptics} are possible.
+\tovi{Again, this raises the question of how we are applying the design 
+space - is it based on its properties when it is printed, or based on 
+any post-print modifications}Tubes need not have hollow centers: in the case where routed tubes are filled with solid material---in particular, a solid material different from the model material---, interactions such as those in \cite{Willis-printedoptics} are possible.
 
@@
 
-Tube topology enables different types of interactions.
+Tube topology enables different types of interactions. \tovi{How about "newwork 
+topologies", since this 
+applies to groups of tubes.
+}
 
@@
 If a maker wished to create a painting device, she might wish to have two system-side tubes feeding in primary-colored red and blue paints which mix in varying ratios, allowing their pigments to combine before purple exits from the device (see Figure \ref{fig:tubespace}).
-Splitting can also be useful, for example if our maker wants red paint output in two locations from her painting device, she could have one system-side tube, but split the tube into two (see Figure \ref{fig:tubespace}).
+Splitting can also be useful, for example if our maker wants red paint output in two locations from her painting device, she could have one system-side tube, but split the tube into two (see Figure \ref{fig:tubespace}).\tovi{How about multiple in/
+multiple out (in contrast to 
+branching/splitting which is 
+single in multiple out)
+} \tovi{This seciton should refer to the 
+possibilities, not to what we 
+actually did.}
 
@@
 Using a star topology in which the tubes were filled with conductive paint, we created a toy with several touch-sensitive areas, see Figure \ref{fig:toy}.
+\tovi{May want star, tree added to
+your figure
+}
 
@@
 An interface focused on the internal path of the tube is the neon sign in Figure \ref{fig:neon}: output is based on the shape of the tube.
+\tovi{This is not consistent with the
+fourth row of your table. Not sure 
+this paragraph is needed.
+}
 
@@
 is that ok?
-it's in the table, and I think we should cover the related input techniques here rather than in the designated related work section (which I think is more suited to fabrication and routing stuff).}
+it's in the table, and I think we should cover the related input techniques here rather than in the designated related work section (which I think is more suited to fabrication and routing stuff).} \tovi{Yes, related work here is ok, if it
+doesnt belong in the realted 
+work section. Agreed that 
+seciton should be highler level 
+discussion of fabrication and 
+routing.
+}
 

### 2014-03-12 paper/designspace.tex
fixing some tovi comments, intro rewrite

--- 2014-03-08 paper/designspace.tex
@@
     \includegraphics[width=3.4in]{figures/tubespace.png}
-\caption{The design space of tubes.  Tube types, media, topologies, and design are discussed more fully in the text.}
+\caption{The design space of tubes.  Tube types, media, topologies, and design are discussed more fully in the text.  \tovi{May want star, tree added to 
+your figure
+}
+}
 \label{fig:tubespace}
@@
 
-\subsection{Types of Tubes}
+\subsection{Types of Tubes}\tovi{Maybe rename types to 
+"openings"?}
 
-Tubes can come in four types: open, return, semi-closed, and fully enclosed.  Each of these types offers distinct interaction capabilities (see Figure \ref{fig:tubespace}).
+Tubes can come in four types: open, return, semi-closed, and fully enclosed.
+Each of these types offers distinct interaction capabilities (see Figure \ref{fig:tubespace}).
+Definitionally, in the following paragraphs we refer to a tube which is ``open on the user side'' to have its contents physically accessible to an interacting user at interaction time.
+This includes tubes which function as fountains, but it does not include tubes whose ends have been capped to keep water inside (whether this capping was done at print time, or later as a manual modification).
+Tubes whose contents are not physically accessible to the user are ``closed on the user side''.
+A tube which is ``open on the system side'' can have its contents manipulated by a machine, e.g., by resonating, pressurizing, or passing electricity through those contents.
+The system does not need to have physical access to the contents in order to manipulate them.
 
@@
 Semi-closed tubes are open at the system end (for control of the enclosed medium) and closed at the user end.
-We believe this closed interface is most interesting when it is fabricated from a mobile material: for example, a series of tubes terminated in thin rubber membranes on the user side can be actuated by an air pump to create haptic feedback.
+The closed interface is interesting when it fabricated from a mobile material: for example, a series of tubes terminated in thin rubber membranes on the user side can be actuated by an air pump to create haptic feedback, or a hinged cap which can be blown open by air pressure.
 Without a printer capable of fabricating flexible material, a maker could affix a balloon to an open tube's end to behave similarly; another possibility is to use semi-closed tubes as audio-generating resonance chambers.
-
-While possible to create, a tube which is open on the user side and closed on the system side is outside our focus on computer-mediated interaction, and we therefore do not discuss these tubes.
 
@@
 Fully enclosed tubes can be used as resonance chambers (e.g., for object identification), or as air bubbles (e.g., as used in \cite{Willis-printedoptics} for internal display).
-Their physical design space is very limited, as any support material required to create their internal geometry cannot be accessed for removal.
+\tovi{used as a container for water
+or other particles that would 
+otherwise fall out?}  \valkyrie{I re-worded above.
+please check.}  Their physical design space is very limited, as any support material required to create their internal geometry cannot be accessed for removal.
 
@@
 
-Tubes need not have hollow centers: in the case where routed tubes are filled with solid material---in particular, a solid material different from the model material---, interactions such as those in \cite{Willis-printedoptics} are possible.
+\tovi{Again, this raises the question of how we are applying the design 
+space - is it based on its properties when it is printed, or based on 
+any post-print modifications}\valkyrie{re-worded above.
+I think we are basing it on status at ``interaction time"}Tubes need not have hollow centers: in the case where routed tubes are filled with solid material---in particular, a solid material different from the model material---, interactions such as those in \cite{Willis-printedoptics} are possible.
 
@@
 
-\subsection{Topological variations}
+\subsection{Tube Topologies}
 
-Tube topology enables different types of interactions.
+Tube network topologies enable different types of interactions.
 
@@
 If a maker wished to create a painting device, she might wish to have two system-side tubes feeding in primary-colored red and blue paints which mix in varying ratios, allowing their pigments to combine before purple exits from the device (see Figure \ref{fig:tubespace}).
-Splitting can also be useful, for example if our maker wants red paint output in two locations from her painting device, she could have one system-side tube, but split the tube into two (see Figure \ref{fig:tubespace}).
+Splitting can also be useful, for example if our maker wants red paint output in two locations from her painting device, she could have one system-side tube, but split the tube into two (see Figure \ref{fig:tubespace}).\tovi{How about multiple in/
+multiple out (in contrast to 
+branching/splitting which is 
+single in multiple out)
+} \tovi{This seciton should refer to the 
+possibilities, not to what we 
+actually did.}
 
 Star and tree topologies are extensions of the splitting and mixing primitives.
-Using a star topology in which the tubes were filled with conductive paint, we created a toy with several touch-sensitive areas, see Figure \ref{fig:toy}.
+Using a star topology in which the tubes were filled with conductive paint, we created toys with several touch-sensitive areas, see Figure \ref{fig:toys}.
 
-\subsection{Features of Tubes}
+\subsection{Design of Tubes}
 
-Tubes may emphasize either their exterior features (connection points) or their interior paths.
+Tube designs may emphasize either their exterior features (connection points) or their interior paths.
 These two features lead to different kinds of interfaces.
@@
 \subsection{Inputs and Outputs}
-
-\valkyrie{this section includes related work.
-is that ok?
-it's in the table, and I think we should cover the related input techniques here rather than in the designated related work section (which I think is more suited to fabrication and routing stuff).}
 

### 2014-03-12 paper/introduction.tex
fixing some tovi comments, intro rewrite

--- 2014-02-21 paper/introduction.tex
@@
 \section{Introduction}
-Makers, as well as professional designers, leverage 3D printers as tools for design work.
-A wide array of objects, ranging from bicycle helmets to jewelry to video game controllers, are now prototyped or even manufactured using these machines.
-However, most devices fabricated by 3D printers are passive: accessible printers are not yet capable of creating integrated active systems.
+Redirection leads to reusable and iterable devices.
+Televisions have long been operating on this idea: before the recent switch to digital television signals, TV signals had remained unchanged since roughly the 1920s, and the only requirement to bigger and better viewing was an upgrade of the \emph{endpoint itself}.
+A larger television does not require an upgrade of the underlying broadcast system.
+Similarly, the Internet, famously described by US Senator Ted Stephens (R-Alaska) as a ``series of tubes'', works on the principle of routing and distributing information to endpoints; there is no need to reduplicate information in multiple locations as long as it can be retrieved and manipulated by users at their laptops, tablets, mobile phones, or swarm devices.
+
+We propose the application of this idea, simple endpoints redirecting information to and from complex and localized systems, to user interface design.
+In particular, we believe that it lends itself to design for 3D printed devices.
+Since today's 3D printers cannot fabricate the electronics required to enable most devices, their place in designing interactivity has remained unclear.
+Typically, when electronics are integrated with 3D printed objects it is through a manual process, where the design of the object is tightly coupled to the form factor of the electronics.
+By using channels for redirection, a simple 3D printed object can be plugged in to a complex base structure to provide interactivity.
+This allows more rapid iteration on the device's look and feel, while preserving its functionality.
+\valkyrie{I like this idea for the introduction much better, but the way to write it is not totally clear to me yet.
+suggestions welcome.}
 
 Willis, et al., ``envision a future world where interactive devices can be printed rather than assembled; a world where a device with active components is created as a single object, rather than a case enclosing circuit boards and individually assembled parts" \cite{Willis-printedoptics}.
-This is a vision we eagerly share: 3D printers are capable of greating arbitrary geometries not feasible to manufacture using traditional processes, and we see these capabilities being underutilized by makers and designers.
-We see many opportunitities to increase the interactivity of 3D prints using today's printer technology.
+This is a vision we eagerly share: 3D printers are capable of creating arbitrary geometries not feasible to manufacture using traditional processes, and makers and designers are underutilizing these capabilities.
 
-Using specially modified printers or extra machinery, it is possible to create electronics on the surface of 3D prints \cite{optomec} \cite{Sells-reprap}.
-These techniques, however, require a high capital investment and technical expertise.
-In addition, they lack flexibility: the printed circuits must be routed in 2D to be created on the objects' manifold exteriors, and they can only be used to create electronic circuits.
-\valkyrie{what?}
+In this paper, we propose a new technique in which interior hollow chambers and pipes are integrated into 3D printed devices to redirect interactive components from a central system to arbitrary points on a printed object's surface.
+Such channels introduce an entire design space of opportunities for adding interactivity to these objects.
+For example, copper material can fill the channels, to allow for standard electronic components to be easily integrated after the printout; air can be pumped through to create tactile output; or electroluminescent wire can be threaded through to create computer-controlled visual output.
 
-In addition to electronics, many other means of interactivity exist.
+Our work makes the following contributions to the field of interactive digital fabrication:
 
-\valkyrie{Points to hit: there is more than electronics to worry about, like fluids and the organic haptic feedback they afford; tangible interfaces are important (do I have to say this every time??); assembly sucks}
+\begin{itemize}
+\item We lay out a design space of tube-mediated interactive possibilities. 
+\item We offer algorithms and techniques for routing tubes, as well as a design tool implementing them.
+\item We showcase a set of examples, enabled by our modeling tool and exploring new points in the design space.
+\end{itemize}

### 2014-03-12 paper/relatedwork.tex
fixing some tovi comments, intro rewrite

@@
 \cite{Willis-printedoptics} used light pipes, solid clear tubes, to create integrated displays: our work allows a superset of these capabilities, and Printed Optics is a part of our design space.
+Willis, et al., also did not offer a design tool or techniques for creating light pipes automatically.
 \cite{Savage-sauron} investigated the use of computer vision to track interactions with physical mechanisms.
@@
 \cite{Park-microchannels} and \cite{Majidi-curvature} use microchannels filled with liquid metal for pressure or curvature sensing: the metal in the channels changes resistance as the channels are stretched.
-\begin{itemize}
-\item \cite{icecats-conductivepaint} - an instructable I followed to make my first batch of conductive paint
-\end{itemize}
+These techniques have not been used with fabrication, nor has a design tool to create such sensors been previously described.
 
@@
 This is possible for us thanks to \emph{blocking out}\footnote{In neon sign making, segments of a tube that should not appear illuminated are covered with opaque paint.
-This is called "blocking out".\url{http://en.wikipedia.org/wiki/Neon_sign}}.
+This is called "blocking out".
+\url{http://en.wikipedia.org/wiki/Neon_sign}}.
 Prior work used tours of nodes, rather than edges, and cast continuous line drawing as a traveling salesman problem~\cite{Bosch-tsp}.
-Schweikardt, et al., also made a connection between electroluminescent wire and graph theory, though their contribution is orthogonal to ours: they provide a tangible construction kit with smart nodes for building physical graph examples \cite{Schweikardt-tei09}.
-Edges are lit by EL wire.
+Schweikardt, et al., also made a connection between electroluminescent wire and graph theory, though their contribution is orthogonal to ours: they provide a tangible construction kit with smart nodes for building physical graph examples in which edges are lit by EL wire \cite{Schweikardt-tei09}.
 

### 2014-03-14 paper/designspace.tex
new figures!  well, still using placeholders, but templated now, anyway.  also, added star and tree to tubespace figure

@@
 We propose the use of tubes to redirect active components through the core and onto the surface of 3D printed interactive objects.
-By using different types and topologies of tubes and inserting different media into the completed tubes, a variety of input and output types \valkyrie{need clearer word than ``types"} can be located at arbitrary points on the surface of a 3D print.
+By using different types and topologies of tubes and inserting different media into the completed tubes, a variety of inputs and outputs can be located at arbitrary points on the surface of a 3D print.
 We describe the design space of tubes, and then follow up with some of the input and output possibilities---both those that have been previously explored and those that are possible but remain unexplored.
@@
 \centering
-    \includegraphics[width=3.4in]{figures/series-of-tubes.jpg}
+    \includegraphics[width=3.4in]{figures/helmholz.png}
 \caption{This box (a) contains an enclosed Helmholz resonator, seen in the cutaway view (b).

### 2014-03-14 paper/designspace.tex
figures for end of paper, still with placeholders but templated.  notes in identity section.

@@
     \includegraphics[width=3.4in]{figures/tubespace.png}
-\caption{The design space of tubes.  Tube types, media, topologies, and design are discussed more fully in the text.  \tovi{May want star, tree added to 
-your figure
-}
-}
+\caption{The design space of tubes.  Tube types, media, topologies, and design are discussed more fully in the text.}
 \label{fig:tubespace}
@@
 These two features lead to different kinds of interfaces.
-An example interface that focuses on the exterior connection points is the touch sensitive toy in Figure \ref{fig:toy}, where tubes must exit the toy at the eyes, ears, tail, etc.
+An example interface that focuses on the exterior connection points is the touch sensitive toy in Figure \ref{fig:toys}, where tubes must exit the toy at the eyes, ears, tail, etc.
 An interface focused on the internal path of the tube is the neon sign in Figure \ref{fig:neon}: output is based on the shape of the tube.
@@
 Those created with fabrication are in {\color{blue}blue}.
-In \emph{{\color{red}red italic}} are unexplored interactions creatable with custom-fabricated tubes.
-Darker grey is output.
-Lighter grey is input.
-\valkyrie{I don't know what else might go in the blank spaces...
-looking for suggestions!
-Also, I'm not confident this breakdown is the most clear: for example, the "liquids" category includes the copper paint, which functions as wires ("threadables"), but takes the form of a liquid...
-and the general presentation of the table could probably be better in some way.
-also, hello, make this wayyyyyy smaller.}}
+In \emph{{\color{red}red italic}} are unexplored interactions creatable with custom-fabricated tubes.}
 \label{fig:designspace}
@@
 Pressure can be sensed in multiple ways.
-Slyper, et al., in \cite{Slyper-pressure} contribute a set of semi-closed fabricatable primitives that respond to particular actions, like twisting or pushing, with changes in internal air pressure.
-\valkyrie{that sentence is not clear}  These primitives can be attached at the terminus of any semi-closed tube; the user can manipulate the printed endpoint and the system can sense air pressure changes.
+Slyper, et al., in \cite{Slyper-pressure} contribute a set of semi-closed 3D printable primitives that respond to particular actions, like twisting or pushing, with changes in internal air pressure.
+These primitives can be attached as the terminus of any semi-closed tube; the user can manipulate the printed endpoint and the system can sense air pressure changes.
 Pressure can also be more simply sensed via capacitance using those techniques.
@@
 When some object nears or touches a pair of cables, IR light from the LED is reflected back to the camera and the touched point appears as a bright spot.
-Follmer, et al.,'s Jamming User Interfaces \cite{Follmer-jamming} rely dense particulates enclosed in a flexible volume, and can be sensed optically.
+Follmer, et al.,'s Jamming User Interfaces \cite{Follmer-jamming} rely on dense particulates enclosed in a flexible volume, and can be sensed optically.
 These dense particulates can be printed in-place with desired characteristics (such as varying particle size or shape) using sacrificial support material, or added post-print.
@@
 Tubes included in printed objects allow for their identification.
-While the tubes need not be visible (especially in the case of fully-enclosed tubes), their presence, location, and length change the acoustic resonance properties of a printed object.
+While the tubes need not be visible (especially in the case of fully-enclosed tubes), their presence, location, network structure, and length change many properties of a printed object.
+This includes weight, acoustic resonance, and (if conductive material is present) capacitive signature.
+
+\valkyrie{need to do a rewrite of this that doesn't necessarily focus so deeply on audio only.
+we can talk about these other non-visible measurements like weight and capacitive signature.
+I also need to actually get an intentional encoding test up and running...
+we are going to have to be clever about how we stimulate the resonance tubes, though.
+using a speaker won't really work: we have to agitate the air in the chambers, so we might need to use something like a reed (as in reed instruments).}
 

### 2014-03-14 paper/exampleobjects.tex
figures for end of paper, still with placeholders but templated.  notes in identity section.

@@
 \centering
-    \includegraphics[width=3.4in]{figures/series-of-tubes.jpg}
+    \includegraphics[width=3.4in]{figures/brain.png}
 \caption{A touch-sensitive brain whose tubes are filled with conductive paint.
@@
 \centering
-    \includegraphics[width=3.4in]{figures/series-of-tubes.jpg}
+    \includegraphics[width=3.4in]{figures/bunny.png}
 \caption{This rabbit has an open tube leading to its mouth and a semi-closed tube capped with flexible material in its back.
-When air is sucked in through the open tube, the back becomes stiff as though the rabbit has breathed in.}
+When air is sucked in through the open tube, the back becomes stiff as though the rabbit has breathed in.
+Inset shows the interior structure.}
 \label{fig:breathe}
@@
 \centering
-    \includegraphics[width=3.4in]{figures/series-of-tubes.jpg}
+    \includegraphics[width=3.4in]{figures/radio.png}
 \caption{This radio is assembled from traditional electronic components connected by copper-filled tubes.
@@
 \centering
-    \includegraphics[width=3.4in]{figures/series-of-tubes.jpg}
-\caption{This pen holder (a) uses the FlyEye technique (b) to sense the presence or absence of an object in each of its tubes.
-Inset shows the internal structure of the tubes generated by our design tool.}
+    \includegraphics[width=3.4in]{figures/pens.png}
+\caption{This pen holder uses the FlyEye technique to sense the presence or absence of an object in each of its tubes.  Inset shows the internal structure of the tubes generated by our design tool.}
 \label{fig:pens}
@@
 \centering
-    \includegraphics[width=3.4in]{figures/series-of-tubes.jpg}
+    \includegraphics[width=3.4in]{figures/maze.png}
 \caption{Our maze game has a large single particle trapped inside a fully enclosed tube.  It offers hours of fun!}
@@
 \centering
-    \includegraphics[width=3.4in]{figures/series-of-tubes.jpg}
-\caption{A two-state animated neon sign designed using our tool (a)(b) \valkyrie{Tovi: we could use your PDF has to make this animate in place :)}.
+    \includegraphics[width=3.4in]{figures/sign.png}
+\caption{A two-state animated neon sign designed using our tool (a)(b) \valkyrie{Tovi: we could use your PDF hax to make this animate in place :)}.
 (c) shows the SVG input files to create this sign, and (d) shows an isometric view of the tubes created by our tool.}

### 2014-03-17 paper/designspace.tex
bjoern's comments

--- 2014-03-12 paper/designspace.tex
@@
 \section{The Design Space of Tubes}
+\bjoern{We should be careful with our terminology - apparently pipes and tubes have different connotations.
+Tubes are measured by outer diameter (OD) and tend to be structural, while pipes are measured by inside diameter (ID) and tend to be used to transport media inside them.
+At the least, we should mention that different communities have these different connotations.
+I think that in our case, we're thinking mostly of the 3D print as the structural object that we're then tunneling through with {\em pipes}.}
 
@@
 
-\subsection{Types of Tubes}\tovi{Maybe rename types to 
-"openings"?}
+\subsection{Types of Tubes}
 
-Tubes can come in four types: open, return, semi-closed, and fully enclosed.
-Each of these types offers distinct interaction capabilities (see Figure \ref{fig:tubespace}).
-Definitionally, in the following paragraphs we refer to a tube which is ``open on the user side'' to have its contents physically accessible to an interacting user at interaction time.
-This includes tubes which function as fountains, but it does not include tubes whose ends have been capped to keep water inside (whether this capping was done at print time, or later as a manual modification).
-Tubes whose contents are not physically accessible to the user are ``closed on the user side''.
-A tube which is ``open on the system side'' can have its contents manipulated by a machine, e.g., by resonating, pressurizing, or passing electricity through those contents.
-The system does not need to have physical access to the contents in order to manipulate them.
+Tubes can come in four types: open, return, semi-closed, and fully enclosed.  Each of these types offers distinct interaction capabilities (see Figure \ref{fig:tubespace}).
 
@@
 Semi-closed tubes are open at the system end (for control of the enclosed medium) and closed at the user end.
-The closed interface is interesting when it fabricated from a mobile material: for example, a series of tubes terminated in thin rubber membranes on the user side can be actuated by an air pump to create haptic feedback, or a hinged cap which can be blown open by air pressure.
+We believe this closed interface is most interesting when it is fabricated from a mobile material: for example, a series of tubes terminated in thin rubber membranes on the user side can be actuated by an air pump to create haptic feedback.
 Without a printer capable of fabricating flexible material, a maker could affix a balloon to an open tube's end to behave similarly; another possibility is to use semi-closed tubes as audio-generating resonance chambers.
+
+While possible to create, a tube which is open on the user side and closed on the system side is outside our focus on computer-mediated interaction, and we therefore do not discuss these tubes.
 
@@
 Fully enclosed tubes can be used as resonance chambers (e.g., for object identification), or as air bubbles (e.g., as used in \cite{Willis-printedoptics} for internal display).
-\tovi{used as a container for water
-or other particles that would 
-otherwise fall out?}  \valkyrie{I re-worded above.
-please check.}  Their physical design space is very limited, as any support material required to create their internal geometry cannot be accessed for removal.
+Their physical design space is very limited, as any support material required to create their internal geometry cannot be accessed for removal.
 
@@
 As in \cite{Slyper-pressure}, structures can be engineered to change in air pressure when manipulated correctly (e.g., a spiral that changes pressure when twisted, but not pressed), and thus fluid pressure can also be used as an input.
-\tovi{I realize "fluid" pressure is
-technically correct, but might 
-confuse readers, since this is 
-the gas pararaph.}
 
@@
 
-\tovi{Again, this raises the question of how we are applying the design 
-space - is it based on its properties when it is printed, or based on 
-any post-print modifications}\valkyrie{re-worded above.
-I think we are basing it on status at ``interaction time"}Tubes need not have hollow centers: in the case where routed tubes are filled with solid material---in particular, a solid material different from the model material---, interactions such as those in \cite{Willis-printedoptics} are possible.
+Tubes need not have hollow centers: in the case where routed tubes are filled with solid material---in particular, a solid material different from the model material---, interactions such as those in \cite{Willis-printedoptics} are possible.
 
@@
 
-\subsection{Tube Topologies}
+\subsection{Topological variations}
 
-Tube network topologies enable different types of interactions.
+Tube topology enables different types of interactions.
 
@@
 If a maker wished to create a painting device, she might wish to have two system-side tubes feeding in primary-colored red and blue paints which mix in varying ratios, allowing their pigments to combine before purple exits from the device (see Figure \ref{fig:tubespace}).
-Splitting can also be useful, for example if our maker wants red paint output in two locations from her painting device, she could have one system-side tube, but split the tube into two (see Figure \ref{fig:tubespace}).\tovi{How about multiple in/
-multiple out (in contrast to 
-branching/splitting which is 
-single in multiple out)
-} \tovi{This seciton should refer to the 
-possibilities, not to what we 
-actually did.}
+Splitting can also be useful, for example if our maker wants red paint output in two locations from her painting device, she could have one system-side tube, but split the tube into two (see Figure \ref{fig:tubespace}).
 
 Star and tree topologies are extensions of the splitting and mixing primitives.
-Using a star topology in which the tubes were filled with conductive paint, we created toys with several touch-sensitive areas, see Figure \ref{fig:toys}.
+Using a star topology in which the tubes were filled with conductive paint, we created a toy with several touch-sensitive areas, see Figure \ref{fig:toy}.
 
-\subsection{Design of Tubes}
+\subsection{Features of Tubes}
 
-Tube designs may emphasize either their exterior features (connection points) or their interior paths.
+Tubes may emphasize either their exterior features (connection points) or their interior paths.
 These two features lead to different kinds of interfaces.
@@
 \subsection{Inputs and Outputs}
+
+\valkyrie{this section includes related work.
+is that ok?
+it's in the table, and I think we should cover the related input techniques here rather than in the designated related work section (which I think is more suited to fabrication and routing stuff).}
 

### 2014-03-17 paper/introduction.tex
bjoern's comments

--- 2014-03-08 paper/introduction.tex
@@
 However, most devices fabricated by 3D printers are passive: accessible printers are not yet capable of creating integrated active systems.
+\bjoern{not sure I understand phrase after the colon.
+Rewrite and explain what you mean in more detail?}
 
@@
 \tovi{Might want some examples of what possibilities this opens up}
+\bjoern{I think we're doing something a little different from Karl's vision - in all of our examples, you need to hook up extra sensors, actuators, or lights to the 3D-printed object.
+So there's always still assembly required.
+We're not really printing interactive objects in a single pass, but we're offering new ways of adding interactivity to 3D printed objects.
+
+I'd argue the principal novelty of our work is to {\em remove material} from a model to form pipes and other cavities.
+We demonstrate how designers/makers can then use these pipes, e.g.
+by filling them with various media to enable input, display and tactile feedback.
+This {\em subtractive} approach is complementary to {\em additive} approaches that try to print different materials such as conductors.
+While our approach requires manual assembly afterwards, it opens up a large useful design space on commonly available 3D printers.}
 
@@
 In addition, they lack flexibility: the printed circuits must be routed in 2D to be created on the objects' manifold exteriors, and they can only be used to create electronic circuits.
-\valkyrie{what?}  \tovi{I dont follow this paragraph - I think you want  to briefly summarize what has been previousl done in interactive 3d prints, and hihglight the shortcommings and remianing challenges}
+\valkyrie{what?}  \tovi{I dont follow this paragraph - I think you want  to briefly summarize what has been previousl done in interactive 3d prints, and hihglight the shortcommings and remianing challenges} \bjoern{agree.
+Also, probably too much focus on electronics since it's only one of a series of examples we'll demonstrate.}
 
@@
 
-In this paper, we propose a new technique, where interior channels, such as hollow champers or solid, are integrated into 3D models.
+In this paper, we propose a new technique, where interior channels, such as hollow chambers or solid \bjoern{?}, are integrated into 3D models.
 Such channels introduces a entire design space of opportunities for adding interactvity to 3D printed objects.
-For example, copper material can fill the channels, to allow for standard electronic compoents to be easily integrated after the printout; air can be pumped through to create tactile output; or, Electroluminescent wire can be passed through to create interactive visual output.
+For example, copper material can fill the channels, to allow for standard electronic components to be easily integrated after the printout; air can be pumped through to create tactile output; or, Electroluminescent wire can be passed through to create interactive visual output.
+\bjoern{this is good - I'd make this the central argument of the intro and move it up earlier.}
 
-Our work makes the following contributions to the field of interactive digitial fabircation.
+Our work makes the following contributions to the field of interactive digitial fabrication.
 First, ...
-deisgn space...Second ...
+design space...Second ...
 modeling tool and algorithms ...
 Third showcase a set of examples, enabled by our modeling tool.
+\bjoern{I'd say more about the software contribution in the abstract.
+Maybe that just requires expanding these notes into a full paragraph.}

### 2014-03-17 paper/tubes.tex

@@
 \centering
-    \includegraphics[width=3.4in]{figures/series-of-tubes.jpg}
-\caption{An example mesh with exterior connection points brushed (a), and the initial A* routing generated (b).
-Our physics-based, energy-minimizing rod/tube at rest (c) and after simulation (d).
-(e) shows the complete printed object.}
+    \includegraphics[width=3.4in]{figures/exterior.png}
+\caption{An example mesh with exterior connection points brushed (a), and the initial A* routing generated (b).  Our physics-based, energy-minimizing rod/tube at rest (c) and after simulation (d).}
 \label{fig:tool-process-exterior}
@@
 \centering
-    \includegraphics[width=3.4in]{figures/series-of-tubes.jpg}
-\caption{An input vector graphics file (a) and the points which cannot be tubed as drawn (b).
-The connected graph created by our software (c) and the resulting Euler circuit (d) permit creation of a novel neon sign (e).}
+    \includegraphics[width=3.4in]{figures/interior.png}
+\caption{An input vector graphics file with the points which cannot be tubed as drawn highlighted in {\color{blue}blue} (a).
+The connected graph created by our software (b) and the resulting Euler circuit (c) permit creation of a novel neon sign (d).}
 \label{fig:tool-process-interior}
@@
 
-The interior path routing problem is a version of the Chinese Postman Problem\footnote{\url{http://en.wikipedia.org/wiki/Route_inspection_problem}}, in which we wish to traverse all edges of the graph described in the user's input with a single Euler circuit, much as a postman needs to walk along every road at least once to deliver mail.
+The interior path routing problem is a version of the Chinese Postman Problem\footnote{The CPP is also known as the route inspection problem: \url{http://en.wikipedia.org/wiki/Route_inspection_problem}}, in which we wish to traverse all edges of the graph described in the user's input with a single Euler circuit, much as a postman needs to walk along every road at least once to deliver mail.
 We call our relaxation the Spiderman Postman Problem: we allow the creation of non-existing paths (i.e., the postman may traverse buildings in addition to roads).
@@
 \centering
-    \includegraphics[width=3.4in]{figures/series-of-tubes.jpg}
+    \includegraphics[width=3.4in]{figures/intersection-templates.png}
 \caption{When edges intersect in the plane (a), we match them to the nearest of our several intersection templates (b) to create 3D paths that do not interfere with each other.}

### 2014-03-17 paper/abstract.tex

--- 2014-02-27 paper/abstract.tex
@@
 However, in spite of recent work, a 3D printer's role in prototyping \emph{interactivity} is still not clear.
-While macro-scale digital fabrication devices cannot manufacture electronics in-place, we propose a general technique for redirecting active components through the core and onto the surface of 3D printed interactive objects using a series of tubes.
-We describe the design space of tubes and hollow chambers for interaction design: there are a variety of types, topologies, and inserted media for tubes that can be leveraged to create diverse inputs and outputs.
+While macro-scale digital fabrication devices cannot manufacture electronics in-place..., \bjoern{Confusing from ``However..." through``in-place"}.
+We propose a general technique for redirecting active components through the core and onto the surface of 3D printed interactive objects using a series of tubes.
+\bjoern{``redirecting active components" unclear - maybe: ``a general technique for adding interactivity by removing interior material to form pipes and cavities.
+These can be used to sense input and generate haptic as well as visible output."}  We describe the design space of tubes and hollow chambers for interaction design: there are a variety of types, topologies, and inserted media for tubes that can be leveraged to create diverse inputs and outputs.
 We present a technique and design tool for routing tubes of various topologies through the interior of 3D printed parts.
@@
 One allows users to select begin and end points for tubes, then uses A* path routing and physics-based simulation to minimize the bending energy of routed paths.
-The second allows users to enter a description of paths to follow: for this we developed a novel neon sign routing algorithm and offer proof of its optimality.
+The second allows users to enter a description of paths to follow \bjoern{add more detail: why?}: for this we developed a novel neon sign routing algorithm and offer proof of several key traits.
 We present several totally tubular prototypes we created using this tool to show its flexibility and potential, as well as to explore new points in the tube design space.

### 2014-03-19 paper/exampleobjects.tex

@@
 \centering
-    \includegraphics[width=3.4in]{figures/brain.png}
+    \includegraphics[width=3.4in]{figures/placeholder/brain.jpg}
 \caption{A touch-sensitive brain whose tubes are filled with conductive paint.
@@
 \centering
-    \includegraphics[width=3.4in]{figures/bunny.png}
+    \includegraphics[width=3.4in]{figures/placeholder/bunny.jpg}
 \caption{This rabbit has an open tube leading to its mouth and a semi-closed tube capped with flexible material in its back.
@@
 \centering
-    \includegraphics[width=3.4in]{figures/radio.png}
+    \includegraphics[width=3.4in]{figures/placeholder/radio.jpg}
 \caption{This radio is assembled from traditional electronic components connected by copper-filled tubes.
@@
 \centering
-    \includegraphics[width=3.4in]{figures/pens.png}
+    \includegraphics[width=3.4in]{figures/placeholder/pens.jpg}
 \caption{This pen holder uses the FlyEye technique to sense the presence or absence of an object in each of its tubes.  Inset shows the internal structure of the tubes generated by our design tool.}
@@
 \centering
-    \includegraphics[width=3.4in]{figures/maze.png}
+    \includegraphics[width=3.4in]{figures/placeholder/maze.jpg}
 \caption{Our maze game has a large single particle trapped inside a fully enclosed tube.  It offers hours of fun!}
@@
 \centering
-    \includegraphics[width=3.4in]{figures/sign.png}
+    \includegraphics[width=3.4in]{figures/placeholder/sign.jpg}
 \caption{A two-state animated neon sign designed using our tool (a)(b) \valkyrie{Tovi: we could use your PDF hax to make this animate in place :)}.
-(c) shows the SVG input files to create this sign, and (d) shows an isometric view of the tubes created by our tool.}
+(c) shows the SVG input files to create this sign, and (d) shows an isometric view of the tubes created by our tool.
+\valkyrie{d) is a lot less interesting if we are making planar tubes.
+also, for c, should we show an overlay of the two states?
+maybe in different colors?}}
 \label{fig:neon}

### 2014-03-19 paper/introduction.tex

--- 2014-03-12 paper/introduction.tex
@@
 \section{Introduction}
+
+\bjoern{I liked the first sentence of the older intro: ``Makers, as well as professional designers, leverage 3D printers as tools for design work.
+A wide array of objects, ranging from bicycle helmets to jewelry to video game controllers, are now prototyped or even manufactured using these machines.
+However, most devices fabricated by 3D printers are passive.''}
+
 Redirection leads to reusable and iterable devices.
@@
 Similarly, the Internet, famously described by US Senator Ted Stephens (R-Alaska) as a ``series of tubes'', works on the principle of routing and distributing information to endpoints; there is no need to reduplicate information in multiple locations as long as it can be retrieved and manipulated by users at their laptops, tablets, mobile phones, or swarm devices.
+\bjoern{Not sure this works as an introduction.
+The first sentence requires a lot of explanation before the reader understands what you mean by ``redirection'' or ``reuse'' or ``iterable''.
+I don't follow the TV example.}
 
@@
 This is a vision we eagerly share: 3D printers are capable of creating arbitrary geometries not feasible to manufacture using traditional processes, and makers and designers are underutilizing these capabilities.
+\bjoern{I think we're doing something a little different from Karl's vision - in all of our examples, you need to hook up extra sensors, actuators, or lights to the 3D-printed object.
+So there's always still assembly required.
+We're not really printing interactive objects in a single pass, but we're offering new ways of adding interactivity to 3D printed objects.
+
+I'd argue the principal novelty of our work is to {\em remove material} from a model to form pipes and other cavities.
+We demonstrate how designers/makers can then use these pipes, e.g.
+by filling them with various media to enable input, display and tactile feedback.
+This {\em subtractive} approach is complementary to {\em additive} approaches that try to print different materials such as conductors.
+While our approach requires manual assembly afterwards, it opens up a large useful design space on commonly available 3D printers.
+
+It's a useful insight that many of our approaches work by ``redirection'' (but not all - EL wire???). But I'd mention it as a sub-point here, not as the driving story.}
 
@@
 Such channels introduce an entire design space of opportunities for adding interactivity to these objects.
-For example, copper material can fill the channels, to allow for standard electronic components to be easily integrated after the printout; air can be pumped through to create tactile output; or electroluminescent wire can be threaded through to create computer-controlled visual output.
+For example, copper material can fill the channels, to allow for standard electronic components to be easily integrated after the printout; air can be pumped through to create tactile output; or electroluminescent wire can be threaded through to create computer-controlled visual output.\bjoern{this is good - I'd make this the central argument of the intro and move it up earlier.}
 
@@
 \end{itemize}
+\bjoern{I'd say more about the software contribution in the abstract. Maybe that just requires expanding these bullets into a full paragraph.}

### 2014-03-19 paper/designspace.tex
moar pdf figures
some introduction rework started, more detailed explanations of example objects, addition of discussion about gate-based tubes

@@
 \centering
-    \includegraphics[width=3.4in]{figures/tubespace.png}
+    \includegraphics[width=3.4in]{figures/tubespace.pdf}
 \caption{The design space of tubes.  Tube types, media, topologies, and design are discussed more fully in the text.}
@@
 \centering
-    \includegraphics[width=\textwidth]{figures/designspace.png}
+    \includegraphics[width=\textwidth]{figures/designspace.pdf}
 \caption{The design space of tube-based interactions.
@@
 
+\begin{figure}[h]
+\centering
+    \includegraphics[width=3.4in]{figures/placeholder/direct.jpg}
+\caption{By attaching a servo to a gate structure, it is possible to direct fluid through that tube.
+The gate not only has binary positions (a) and (b), but can direct fractional flow, as well (c).
+This could be used to create logic gates, as in (d).
+These could be used, for example, in science exhibits.}
+\label{fig:direct}
+\end{figure}
+
 \emph{Visual} outputs can be mediated by gases, liquids, solid, particulates, or threadables.
@@
 Large particles trapped in tubes visible to the user can be used for, e.g., maze games (see Figure \ref{fig:maze}).
+We can even direct fluids through a tube network using a servo-controlled gate, as in Figure \ref{fig:direct}.
 Using techniques like those described by Harrison, et al., in \cite{Harrison-buttons}  \valkyrie{this paragraph seems terrible.
@@
 \centering
-    \includegraphics[width=3.4in]{figures/helmholz.png}
+    \includegraphics[width=3.4in]{figures/placeholder/helmholz.jpg}
 \caption{This box (a) contains an enclosed Helmholz resonator, seen in the cutaway view (b).

### 2014-03-19 paper/exampleobjects.tex
some introduction rework started, more detailed explanations of example objects, addition of discussion about gate-based tubes

@@
 
-Totally tubular examples.  We are making
+To evaluate our tool's capabilities and to explore new points in the design space, we fabricated a set of five prototype objects designed using the tool.
 
@@
 We built a smart base which can distinguish between the toys and also determine which toy is mounted: since each toy and each gesture has a distinct capacitive signature, we use a simple classifier trained to detect both toy and gesture based on profile.
-These devices were all fabricated on a Makerbot Replicator 2.
+All components were fabricated on a Makerbot Replicator 2.
 
@@
 A custom radio built using tubes allows users to tune to different stations and listen in.
-This device uses a network of disconnected open tubes filled with conductive paint to connect traditional electronic components (two potentiometers, a speaker, and an LED) to a microcontroller.
+This device uses a network of disconnected open tubes filled with conductive paint to connect traditional electronic components (two potentiometers and a speaker) to a microcontroller (in our case, an Arduino).
 We hid the microcontroller and a battery-driven power supply in the base of the radio to make it portable.
+For the radio functionality, we attached an Si4703 to the Arduino.
 We fabricated this device on our Makerbot Replicator 2.
@@
     \includegraphics[width=3.4in]{figures/placeholder/pens.jpg}
-\caption{This pen holder uses the FlyEye technique to sense the presence or absence of an object in each of its tubes.  Inset shows the internal structure of the tubes generated by our design tool.}
+\caption{This pen holder a modified FlyEye technique to sense the presence or absence of an object in each of its tubes.  Inset shows the internal structure of the tubes generated by our design tool.}
 \label{fig:pens}
@@
 Such information was used by \cite{Mueller-constructable} to determine which physical laser pointer a designer was using to interact with their interface.
-Our pen holder uses the FlyEye technique described by Wimmer in \cite{Wimmer-flyeye} and contains open tubes filled with fiber optic cables, two per pen chamber.
-One tube from each chamber leads to an infrared LED, and the other leads to a camera.
-When a pen is in its appointed place, light from the LED is reflected off its bottom and travels through the other optical fiber into the camera, where it registers as a bright point.
-A dark point on the camera region appears when a pen has been removed from its place, and the position of the point indicates which pen it is.
+Our pen holder uses a modification of the FlyEye technique described by Wimmer in \cite{Wimmer-flyeye} and contains open tubes filled with fiber optic cables, one per pen chamber.
+Our modification of using a single cable per chamber is possible because we use 6mm diameter fiber optic cables, in comparison to the angel hair tubes used in the original work, and can thus send and receive through the same cable.
+At the base of each tube is a QRE1113 line sensor digital breakout board, which has an integrated IR emitter and receiver.
+When a pen is in its appointed place, the emitted infrared light is reflected off its bottom and travels back to the receiver, where it registers as bright.
 This prototype was built by our Objet Connex 260.
+\valkyrie{this whole paragraph is unclear}
 
@@
 We designed a custom neon sign bearing the UIST logo and the waves of Hawaii.
-The piece features open tubes due to support-flushing constraints, however the tubes could also be semi-closed.
+The piece features open tubes due to support-flushing constraints, however the tubes could also be semi-closed (one end must remain open to connect the EL wire to its power source).
 They have been threaded with EL wire which is lit in sequence to create an animation.

### 2014-03-19 paper/introduction.tex
some introduction rework started, more detailed explanations of example objects, addition of discussion about gate-based tubes

@@
 
-\bjoern{I liked the first sentence of the older intro: ``Makers, as well as professional designers, leverage 3D printers as tools for design work.
+Makers, as well as professional designers, leverage 3D printers as tools for design work.
 A wide array of objects, ranging from bicycle helmets to jewelry to video game controllers, are now prototyped or even manufactured using these machines.
-However, most devices fabricated by 3D printers are passive.''}
+However, most devices fabricated by 3D printers are passive.
+The major challenge in creating interactive devices using 3D printing technology is the fundamental limitation that these machines cannot fabricate electronics in-place alongside plastic casings.
+This forces makers to link their printable designs with the form factor of the off-the-shelf electronics they will insert post-print, rather than giving them the freedom to create interactive components of the size, shape, and type they desire.
+
+Tools, for example Enclosed \cite{Weichel-enclosed}, have been developed to ease this integration step.  However, there remains an elementary gap between the
 

### 2014-03-19 paper/tubes.tex
add in placeholder figures that are less awful
some introduction rework started, more detailed explanations of example objects, addition of discussion about gate-based tubes

@@
 \centering
-    \includegraphics[width=3.4in]{figures/exterior.png}
-\caption{An example mesh with exterior connection points brushed (a), and the initial A* routing generated (b).  Our physics-based, energy-minimizing rod/tube at rest (c) and after simulation (d).}
+    \includegraphics[width=3.4in]{figures/placeholder/exterior.jpg}
+\caption{An example mesh with exterior connection points selected.  The initial A* routing is drawn in {\color{red}red}, and in {\color{tovi}green} is our physically-based rod after simulation.}
 \label{fig:tool-process-exterior}
@@
 \centering
-    \includegraphics[width=3.4in]{figures/interior.png}
+    \includegraphics[width=3.4in]{figures/interior.pdf}
 \caption{An input vector graphics file with the points which cannot be tubed as drawn highlighted in {\color{blue}blue} (a).
@@
 
+\subsubsection{Routing}
 The interior path routing problem is a version of the Chinese Postman Problem\footnote{The CPP is also known as the route inspection problem: \url{http://en.wikipedia.org/wiki/Route_inspection_problem}}, in which we wish to traverse all edges of the graph described in the user's input with a single Euler circuit, much as a postman needs to walk along every road at least once to deliver mail.
@@
 \centering
-    \includegraphics[width=3.4in]{figures/intersection-templates.png}
+    \includegraphics[width=3.4in]{figures/placeholder/templates.jpg}
 \caption{When edges intersect in the plane (a), we match them to the nearest of our several intersection templates (b) to create 3D paths that do not interfere with each other.}
@@
 Our templates are parameterized, so the templates can be modified to accommodate precise intersection characteristics.
+
+\valkyrie{For now, we just let them intersect.  The angel hair EL wire is thin enough that we don't actually care.}
 

### 2014-03-20 paper/abstract.tex
paper work, adding a new figure

--- 2014-02-27 paper/abstract.tex
@@
 3D printers offer extraordinary flexibilty in prototyping the shape and mechanical function of objects.
-However, in spite of recent work, a 3D printer's role in prototyping \emph{interactivity} is still not clear.
-While macro-scale digital fabrication devices cannot manufacture electronics in-place, we propose a general technique for redirecting active components through the core and onto the surface of 3D printed interactive objects using a series of tubes.
-We describe the design space of tubes and hollow chambers for interaction design: there are a variety of types, topologies, and inserted media for tubes that can be leveraged to create diverse inputs and outputs.
-We present a technique and design tool for routing tubes of various topologies through the interior of 3D printed parts.
+However, a 3D printer's role in prototyping interactive objects is still not clear.
+We propose a general technique for adding interactivity by removing interior material to form pipes and cavities.
+These can be used to sense input and generate haptic as well as visible output.
+We describe the design space of pipes and hollow chambers for interaction design: there are a variety of types, topologies, and inserted media for pipes that can be leveraged to create diverse inputs and outputs.
+We present a technique and design tool for routing pipes of various topologies through the interior of 3D printed parts.
 Our design tool is integrated into a 3D model manipulation program.
 There are two distinct routing algorithms.
-One allows users to select begin and end points for tubes, then uses A* path routing and physics-based simulation to minimize the bending energy of routed paths.
-The second allows users to enter a description of paths to follow: for this we developed a novel neon sign routing algorithm and offer proof of its optimality.
-We present several totally tubular prototypes we created using this tool to show its flexibility and potential, as well as to explore new points in the tube design space.
+One, for creating particular points of interaction or integrating with existing electronics, allows users to select begin and end points for pipes, then uses A* path routing and physics-based simulation to minimize the bending energy of routed paths.
+The second, for creating, e.g., novel neon signs, allows users to enter a description of paths to follow: for this we developed a novel routing algorithm and offer proof of key traits.
+We present several totally tubular prototypes we created using this tool to show its flexibility and potential, as well as to explore new points in the pipe design space.
 \end{abstract}

### 2014-03-20 paper/appendix.tex
paper work, adding a new figure

@@
 \appendix
+
+\section{Proof of Theorem 1}
+
+\begin{theorem} \label{thm:graphness}
+ $G' = (V, E')$ is a connected, semi-Eulerian graph for which $E \subseteq E'$.
+\end{theorem}
+
+\begin{proof}
+$E \subseteq E'$ : we never remove edges from $E$ in our algorithm.  $\therefore E \subseteq E'$.
+
+$G' = (V, E')$ is connected : if $G'$ is not connected, $\exists G_i = (V_i, E_i) \subset G', u \in V_i, v\in V \setminus V_i$ s.t.
+$u$ is not reachable from $v$.
+$G_i \not\in G_{dis}$, because we create edges connecting $G_j, G_k \forall G_j, G_k \in G_{dis}$ and only remove an edge $e_{dupe} = (u, v)$ from $E_{conn}$ once we determine that $u$ is reachable from $v$ along edges $E' \cup E_{conn-min}$.
+$\therefore G_i \not\in G_{dis}$.
+This implies that $G_i$ was not initially disconnected from $G$, because by definition $G_{dis}$ is the set of all disconnected subgraphs of $G$.
+We cannot have disconnected $G_i$ from $G$ because $E \subseteq E'$.
+Thus, a contradiction.
+$\therefore G' = (V, E')$ is connected.
+
+$G' = (V, E')$ is semi-Eulerian : Each edge $e$ in our minimum weight matching connects a unique pair of vertices $v_i, v_j \in V_{odd}$ by the definition of minimum weight matching.
+Each edge $e = (v_i, v_j) \in E_{circuit-min}$ adds one to the degree of $v_i$ and $v_j$, causing them to be of even degree.
+$|V_{odd}| \% 2 = 0$ by the handshake lemma, $\therefore$ all edges can be paired.
+When we remove $e_{temp} = (start, end)$ from $E'$, we cause those vertices, which were of even degree by the above process, to be of odd degree.
+Thus, all vertices except $start$ and $end$ (which are of odd degree) are of even degree.
+$\therefore G'$ is semi-Eulerian.
+\end{proof}
+
 \section{Injection Measurements}

### 2014-03-20 paper/designspace.tex
paper work, adding a new figure

@@
-\section{The Design Space of Tubes}
-\bjoern{We should be careful with our terminology - apparently pipes and tubes have different connotations.
-Tubes are measured by outer diameter (OD) and tend to be structural, while pipes are measured by inside diameter (ID) and tend to be used to transport media inside them.
+\section{The Design Space of Pipes}
+\bjoern{We should be careful with our terminology - apparently pipes and pipes have different connotations.
+Pipes are measured by outer diameter (OD) and tend to be structural, while pipes are measured by inside diameter (ID) and tend to be used to transport media inside them.
 At the least, we should mention that different communities have these different connotations.
@@
 
-We propose the use of tubes to redirect active components through the core and onto the surface of 3D printed interactive objects.
-By using different types and topologies of tubes and inserting different media into the completed tubes, a variety of inputs and outputs can be located at arbitrary points on the surface of a 3D print.
-We describe the design space of tubes, and then follow up with some of the input and output possibilities---both those that have been previously explored and those that are possible but remain unexplored.
+We propose the use of pipes to redirect active components through the core and onto the surface of 3D printed interactive objects, as well as to allow for the creation of designed paths.
+By using different types and topologies of pipes and inserting different media into the completed pipes, a variety of inputs and outputs can be located at arbitrary points on the surface of a 3D print.
+We describe the design space of pipes, and then follow up with some of the input and output possibilities---both those that have been previously explored and those that are possible but remain unexplored.
 
@@
     \includegraphics[width=3.4in]{figures/tubespace.pdf}
-\caption{The design space of tubes.  Tube types, media, topologies, and design are discussed more fully in the text.}
-\label{fig:tubespace}
+\caption{The design space of pipes.  Pipe types, media, topologies, and design are discussed more fully in the text.}
+\label{fig:pipespace}
 \end{figure}
 
-\subsection{Types of Tubes}\tovi{Maybe rename types to 
+\subsection{Types of Pipes}\tovi{Maybe rename types to 
 "openings"?}
 
-Tubes can come in four types: open, return, semi-closed, and fully enclosed.
-Each of these types offers distinct interaction capabilities (see Figure \ref{fig:tubespace}).
-Definitionally, in the following paragraphs we refer to a tube which is ``open on the user side'' to have its contents physically accessible to an interacting user at interaction time.
-This includes tubes which function as fountains, but it does not include tubes whose ends have been capped to keep water inside (whether this capping was done at print time, or later as a manual modification).
-Tubes whose contents are not physically accessible to the user are ``closed on the user side''.
-A tube which is ``open on the system side'' can have its contents manipulated by a machine, e.g., by resonating, pressurizing, or passing electricity through those contents.
+Pipes can come in four types: open, return, semi-closed, and fully enclosed.
+Each of these types offers distinct interaction capabilities (see Figure \ref{fig:pipespace}).
+Definitionally, in the following paragraphs we refer to a pipe which is ``open on the user side'' to have its contents physically accessible to an interacting user at interaction time.
+This includes pipes which function as fountains, but it does not include pipes whose ends have been capped to keep water inside (whether this capping was done at print time, or later as a manual modification).
+Pipes whose contents are not physically accessible to the user are ``closed on the user side''.
+A pipe which is ``open on the system side'' can have its contents manipulated by a machine, e.g., by resonating, pressurizing, or passing electricity through those contents.
 The system does not need to have physical access to the contents in order to manipulate them.
 
-Open tubes originate from the system side and connect the user side, with both ends of the tube open.
-This type of tube may be used to create, for example, capacitive sensors: an open tube filled with conductive paint can be connected to a sensing platform (e.g., Arduino) on the system side, while a user can touch the uncapped other end of the tube.
+Open pipes originate from the system side and connect the user side, with both ends of the pipe open.
+This type of pipe may be used to create, for example, capacitive sensors: an open pipe filled with conductive paint can be connected to a sensing platform (e.g., Arduino) on the system side, while a user can touch the uncapped other end of the pipe.
 Using Swept Frequency Capacitive Sensing \cite{Sato-touche} or other techniques, a user's touch of the open end can be sensed.
-An open tube can also be used for output, for example by creating in-air vortices as in \cite{Sodhi-aireal}.
+An open pipe can also be used for output, for example by creating in-air vortices as in \cite{Sodhi-aireal}.
 
-A return tube originates at the system and returns back to it.
-By threading an electroluminescent (EL) wire through a clear return tube, a maker can create a custom piece of neon art.
-If a return tube passes very close to the surface of a 3D printed model, warm or cold water passed through the tube could be used for temperature-based haptic feedback.
+A return pipe originates at the system and returns back to it.
+By threading an electroluminescent (EL) wire through a clear return pipe, a maker can create a custom piece of neon art.
+If a return pipe passes very close to the surface of a 3D printed model, warm or cold water passed through the pipe could be used for temperature-based haptic feedback.
 
-Semi-closed tubes are open at the system end (for control of the enclosed medium) and closed at the user end.
-The closed interface is interesting when it fabricated from a mobile material: for example, a series of tubes terminated in thin rubber membranes on the user side can be actuated by an air pump to create haptic feedback, or a hinged cap which can be blown open by air pressure.
-Without a printer capable of fabricating flexible material, a maker could affix a balloon to an open tube's end to behave similarly; another possibility is to use semi-closed tubes as audio-generating resonance chambers.
+Semi-closed pipes are open at the system end (for control of the enclosed medium) and closed at the user end.
+The closed interface is interesting when it fabricated from a mobile material: for example, a series of pipes terminated in thin rubber membranes on the user side can be actuated by an air pump to create haptic feedback, or a hinged cap which can be blown open by air pressure.
+Without a printer capable of fabricating flexible material, a maker could affix a balloon to an open pipe's end to behave similarly; another possibility is to use semi-closed pipes as audio-generating resonance chambers.
 
-A fully enclosed tube has no openings on either the system side or the user side.
-Fully enclosed tubes can be used as resonance chambers (e.g., for object identification), or as air bubbles (e.g., as used in \cite{Willis-printedoptics} for internal display).
+A fully enclosed pipe has no openings on either the system side or the user side.
+Fully enclosed pipes can be used as resonance chambers (e.g., for object identification), or as air bubbles (e.g., as used in \cite{Willis-printedoptics} for internal display).
 \tovi{used as a container for water
@@
 
-\subsection{Media in Tubes}
+\subsection{Media in Pipes}
 
-Tubes can be filled with a variety of media to create different interface affordances and capabilities.
+Pipes can be filled with a variety of media to create different interface affordances and capabilities.
 
 ``Gas'' comprises all compressible fluids.
-Use of fluid pressure inside tubes can create haptic feedback at semi-closed interfaces, or the gases can be used as carriers for scents or fog.
+Use of fluid pressure inside pipes can create haptic feedback at semi-closed interfaces, or the gases can be used as carriers for scents or fog.
 As in \cite{Slyper-pressure}, structures can be engineered to change in air pressure when manipulated correctly (e.g., a spiral that changes pressure when twisted, but not pressed), and thus fluid pressure can also be used as an input.
@@
 Incompressible fluids (``liquids'') can perform many of the same interface tasks as gases.
-One opportunity with liquids is to fill the interior of tubes with them and cap the ends.
-In addition, one can use driable conductive fluids, such as copper paint, to coat the interior of tubes and allow them to function as arbitrarily-shaped wires.
+One opportunity with liquids is to fill the interior of pipes with them and cap the ends.
+In addition, one can use driable conductive fluids, such as copper paint, to coat the interior of pipes and allow them to function as arbitrarily-shaped wires.
 This is especially helpful for the creation of a shared ground, or for creating single-wire capacitive interfaces amenable to sensing with SFCS \cite{Sato-touche}.
@@
 any post-print modifications}\valkyrie{re-worded above.
-I think we are basing it on status at ``interaction time"}Tubes need not have hollow centers: in the case where routed tubes are filled with solid material---in particular, a solid material different from the model material---, interactions such as those in \cite{Willis-printedoptics} are possible.
+I think we are basing it on status at ``interaction time"}Pipes need not have hollow centers: in the case where routed pipes are filled with solid material---in particular, a solid material different from the model material---, interactions such as those in \cite{Willis-printedoptics} are possible.
 
@@
 Sparse particles in a stream of fluid can provide haptic feedback.
-Dense particles in a semi-closed tube allow for jamming-based interactions at any point on the surface of an object \cite{Follmer-jamming}.
+Dense particles in a semi-closed pipe allow for jamming-based interactions at any point on the surface of an object \cite{Follmer-jamming}.
 
-Threadable inserted elements, such as electroluminescent (EL) wire or fiberoptic cables, are those that can be threaded through tubes post-printing.
-This allows overcoming limitations of printers: for example, a Printed Optics-style interface can be created on an inexpensive consumer-grade 3D printer using tubes and inserted fiber optic cable.
+Threadable inserted elements, such as electroluminescent (EL) wire or fiberoptic cables, are those that can be threaded through pipes post-printing.
+This allows overcoming limitations of printers: for example, a Printed Optics-style interface can be created on an inexpensive consumer-grade 3D printer using pipes and inserted fiber optic cable.
 
-\subsection{Tube Topologies}
+\subsection{Pipe Topologies}
 
-Tube network topologies enable different types of interactions.
+Pipe network topologies enable different types of interactions.
 
-Splitting or mixing tubes offer flexibility in output.
-If a maker wished to create a painting device, she might wish to have two system-side tubes feeding in primary-colored red and blue paints which mix in varying ratios, allowing their pigments to combine before purple exits from the device (see Figure \ref{fig:tubespace}).
-Splitting can also be useful, for example if our maker wants red paint output in two locations from her painting device, she could have one system-side tube, but split the tube into two (see Figure \ref{fig:tubespace}).\tovi{How about multiple in/
+Splitting or mixing pipes offer flexibility in output.
+If a maker wished to create a painting device, she might wish to have two system-side pipes feeding in primary-colored red and blue paints which mix in varying ratios, allowing their pigments to combine before purple exits from the device (see Figure \ref{fig:pipespace}).
+Splitting can also be useful, for example if our maker wants red paint output in two locations from her painting device, she could have one system-side pipe, but split the pipe into two (see Figure \ref{fig:pipespace}).\tovi{How about multiple in/
 multiple out (in contrast to 
@@
 Star and tree topologies are extensions of the splitting and mixing primitives.
-Using a star topology in which the tubes were filled with conductive paint, we created toys with several touch-sensitive areas, see Figure \ref{fig:toys}.
+Using a star topology in which the pipes were filled with conductive paint, we created toys with several touch-sensitive areas, see Figure \ref{fig:toys}.
 
-\subsection{Design of Tubes}
+\subsection{Design of Pipes}
 
-Tube designs may emphasize either their exterior features (connection points) or their interior paths.
+Pipe designs may emphasize either their exterior features (connection points) or their interior paths.
 These two features lead to different kinds of interfaces.
-An example interface that focuses on the exterior connection points is the touch sensitive toy in Figure \ref{fig:toys}, where tubes must exit the toy at the eyes, ears, tail, etc.
-An interface focused on the internal path of the tube is the neon sign in Figure \ref{fig:neon}: output is based on the shape of the tube.
+An example interface that focuses on the exterior connection points is the touch sensitive toy in Figure \ref{fig:toys}, where pipes must exit the toy at the eyes, ears, tail, etc.
+An interface focused on the internal path of the pipe is the neon sign in Figure \ref{fig:neon}: output is based on the shape of the pipe.
 
-\subsubsection{Dimensions of Tubes}
+\subsubsection{Dimensions of Pipes}
 
@@
 
-By crossing tube types, topologies, and inserted media, we create a space of possible inputs and outputs that can be tube-mediated.
+By crossing pipe types, topologies, and inserted media, we create a space of possible inputs and outputs that can be pipe-mediated.
 We offer a table (see Figure \ref{fig:designspace}) containing these possibilities, referencing previous work on input/output techniques where appropriate, noting that many of these techniques have not been attempted using digital fabrication.
@@
     \includegraphics[width=\textwidth]{figures/designspace.pdf}
-\caption{The design space of tube-based interactions.
+\caption{The design space of pipe-based interactions.
 Existing systems are written in regular font.
 Those created with fabrication are in {\color{blue}blue}.
-In \emph{{\color{red}red italic}} are unexplored interactions creatable with custom-fabricated tubes.}
+In \emph{{\color{red}red italic}} are unexplored interactions creatable with custom-fabricated pipes.}
 \label{fig:designspace}
@@
 
-Tubes create opportunities for many types of input sensing across the surface of printed devices.
+Pipes create opportunities for many types of input sensing across the surface of printed devices.
 We describe sensing touch, pressure, grasp, flexing, tapping, and manipulation of traditional electronic components.
-We recognize that there are likely more input sensing techniques, both existing and on the horizon, that are compatible with the use of tubes.
+We recognize that there are likely more input sensing techniques, both existing and on the horizon, that are compatible with the use of pipes.
 
-Touch sensing is enabled through tubes filled with conductive media.
+Touch sensing is enabled through pipes filled with conductive media.
 Traditional wires (threadable) may be used, but we have had success with conductive paint (liquid).
-Using conductive paint and Swept Frequency Capacitive Sensing (SFCS \cite{Sato-touche}), an interior star tube topology can enable single-wire touch- and grasp-sensing at any set of points on a printed object's surface (see Figure \cite{fig:toy}).
+Using conductive paint and Swept Frequency Capacitive Sensing (SFCS \cite{Sato-touche}), an interior star pipe topology can enable single-wire touch- and grasp-sensing at any set of points on a printed object's surface (see Figure \cite{fig:toy}).
 Simple capacitance measurements are possible using the same media, and custom-designed sensors like those created in Savage, et al.,'s Midas system are also possible \cite{Savage-midas}.
@@
 Slyper, et al., in \cite{Slyper-pressure} contribute a set of semi-closed 3D printable primitives that respond to particular actions, like twisting or pushing, with changes in internal air pressure.
-These primitives can be attached as the terminus of any semi-closed tube; the user can manipulate the printed endpoint and the system can sense air pressure changes.
+These primitives can be attached as the terminus of any semi-closed pipe; the user can manipulate the printed endpoint and the system can sense air pressure changes.
 Pressure can also be more simply sensed via capacitance using those techniques.
@@
 These dense particulates can be printed in-place with desired characteristics (such as varying particle size or shape) using sacrificial support material, or added post-print.
-The optical links necessary for sensing with FlyEye or Jamming can be created either via clear solid cores as in \cite{Willis-printedoptics} or via fiber optic cables threaded through hollow tubes (see Figure \ref{fig:pens}).
+The optical links necessary for sensing with FlyEye or Jamming can be created either via clear solid cores as in \cite{Willis-printedoptics} or via fiber optic cables threaded through hollow pipes (see Figure \ref{fig:pens}).
 
@@
 Much like Slyper, et al., in \cite{Slyper-shape}, we can sense flexing and bending of prints made on these machines.
-While Slyper, et al., created multi-level multi-stage silicone molds with embedded wires to build their sensors, by placing tubes in our rubberlike prints and inserting conductive media afterwards, we can perform the same flex sensing.
+While Slyper, et al., created multi-level multi-stage silicone molds with embedded wires to build their sensors, by placing pipes in our rubberlike prints and inserting conductive media afterwards, we can perform the same flex sensing.
 
 Tapping is another input possibility.
-Due to differential sound conductivity in different printed materials (in particular, the rubberlike Objet material compared to the hard Objet material \valkyrie{presumably we should get or take some measurements of this and include it in the appendix?}), tubes of a greater conducting material can be embedded in a model of a lesser conducting material.
-A microphone or piezo placed at the system-side of these sound-conducting tubes can determine where the model was tapped and how it was tapped.
-Active acoustic sensing as described in Touch \& Activate \cite{Ono-touchandactivate} is also possible; this technique would allow for particular areas of interest (connected to sound-conducting tubes) to be touch sensitive, while other areas (made of non-conducting material) are not sensed.
+Due to differential sound conductivity in different printed materials (in particular, the rubberlike Objet material compared to the hard Objet material \valkyrie{presumably we should get or take some measurements of this and include it in the appendix?}), pipes of a greater conducting material can be embedded in a model of a lesser conducting material.
+A microphone or piezo placed at the system-side of these sound-conducting pipes can determine where the model was tapped and how it was tapped.
+Active acoustic sensing as described in Touch \& Activate \cite{Ono-touchandactivate} is also possible; this technique would allow for particular areas of interest (connected to sound-conducting pipes) to be touch sensitive, while other areas (made of non-conducting material) are not sensed.
 
 Sensing via traditional electronic components (e.g., potentiometers, alcohol gas sensors) can be accomplished by conductive inserted media.
-Components can be recessed into a print's surface, with their leads implanted into open tubes.
+Components can be recessed into a print's surface, with their leads implanted into open pipes.
 By inserting liquid copper paint instead of threading traditional wires, all components in a model can share a ground line, and the paint's drying process obviates the use of solder or glue to affix the components in place (see Figure \ref{fig:radio}).
@@
 
-While many input techniques require the use of conductive media inserted into tubes, output media are more varied.
-We describe potential tube-based outputs organized by the five senses: visual, aural, haptic, and olfactory/gustatory.
+While many input techniques require the use of conductive media inserted into pipes, output media are more varied.
+We describe potential pipe-based outputs organized by the five senses: visual, aural, haptic, and olfactory/gustatory.
 In addition to outputs we mention explicitly, any outputs possible using traditional electronic components are possible as described above.
@@
     \includegraphics[width=3.4in]{figures/placeholder/direct.jpg}
-\caption{By attaching a servo to a gate structure, it is possible to direct fluid through that tube.
+\caption{By attaching a servo to a gate structure, it is possible to direct fluid through that pipe.
 The gate not only has binary positions (a) and (b), but can direct fractional flow, as well (c).
@@
 Colored liquids can be mixed and split before exiting an opaque device, or in a transparent device their mixing, splitting, and paths could be used for display.
-PneUIs \cite{Yao-pneui} can be fabricated as single parts: the rubberlike material offered by our Objet is inflatable when thin (200\% elongation at break), and a PneUI located at the terminus of a tube could be actuated by a system-side air pump (see Figure \ref{fig:breathe}).
-Mechanical motion can be mediated through fluids (a light object which is partially contained in a tube or located at its end could be pushed by the fluid pressure of the tube) or threadables (e.g., muscle wire).
-Large particles trapped in tubes visible to the user can be used for, e.g., maze games (see Figure \ref{fig:maze}).
-We can even direct fluids through a tube network using a servo-controlled gate, as in Figure \ref{fig:direct}.
+PneUIs \cite{Yao-pneui} can be fabricated as single parts: the rubberlike material offered by our Objet is inflatable when thin (200\% elongation at break), and a PneUI located at the terminus of a pipe could be actuated by a system-side air pump (see Figure \ref{fig:breathe}).
+Mechanical motion can be mediated through fluids (a light object which is partially contained in a pipe or located at its end could be pushed by the fluid pressure of the pipe) or threadables (e.g., muscle wire).
+Large particles trapped in pipes visible to the user can be used for, e.g., maze games (see Figure \ref{fig:maze}).
+We can even direct fluids through a pipe network using a servo-controlled gate, as in Figure \ref{fig:direct}.
 Using techniques like those described by Harrison, et al., in \cite{Harrison-buttons}  \valkyrie{this paragraph seems terrible.
@@
 \emph{Aural} outputs can be created with gases.
-Zoran has successfully 3D printed a functional flute, which in essence is a long open open tube with semi-closed tubes coverable by keys \cite{Zoran-flute}.
+Zoran has successfully 3D printed a functional flute, which in essence is a long open open pipe with semi-closed pipes coverable by keys \cite{Zoran-flute}.
 Helmholz resonance (the phenomenon that creates noise when you blow into the top of a glass bottle, which is also how some musical instruments, like ocarinas, work) can be leveraged to create computer-controllable sound chambers.
-All that is necessary for this is an enclosed chamber with a tube connected to it via a narrow neck-point (see Figure \ref{fig:ocarina}).
-Passive sound amplification (like a phonograph horn) and sound redirection are also possible using tubes.
+All that is necessary for this is an enclosed chamber with a pipe connected to it via a narrow neck-point (see Figure \ref{fig:ocarina}).
+Passive sound amplification (like a phonograph horn) and sound redirection are also possible using pipes.
 
@@
 \caption{This box (a) contains an enclosed Helmholz resonator, seen in the cutaway view (b).
-This resonator creates a sound when air is blown through the tube at the top, because air compresses at the small neck joint between the chamber and the tube.
+This resonator creates a sound when air is blown through the pipe at the top, because air compresses at the small neck joint between the chamber and the pipe.
 \valkyrie{need to get a test print of this that actually works}}
@@
 Fluids can actuate particulates and create programmable pliability (as discussed above and in \cite{Follmer-jamming}).
-Additionally, use of semi-closed tubes with rubberlike material as the caps allows for gases to actuate surface features (see Figure \ref{fig:breathe}).
-Sodhi, et al., in \cite{Sodhi-aireal} created free-air haptic feedback using controlled vortex generation: this technique could also be reproduced using a series of tubes.
+Additionally, use of semi-closed pipes with rubberlike material as the caps allows for gases to actuate surface features (see Figure \ref{fig:breathe}).
+Sodhi, et al., in \cite{Sodhi-aireal} created free-air haptic feedback using controlled vortex generation: this technique could also be reproduced using a series of pipes.
 Iwata, et al., explored volumetric fluid-based interactions with programmatically inflatable balloons and an overhead projector in \cite{Iwata-volflex}, and additionally haptic outputs can be used to increase ergonomics, as in the inflatable mouse by Kim, et al.
@@
 
-\emph{Olfactory} and \emph{gustatory} outputs can be created through the mixing and splitting of tubes carrying scented or flavored fluids, respectively.
+\emph{Olfactory} and \emph{gustatory} outputs can be created through the mixing and splitting of pipes carrying scented or flavored fluids, respectively.
 
@@
 
-Tubes included in printed objects allow for their identification.
-While the tubes need not be visible (especially in the case of fully-enclosed tubes), their presence, location, network structure, and length change many properties of a printed object.
+Pipes included in printed objects allow for their identification.
+While the pipes need not be visible (especially in the case of fully-enclosed pipes), their presence, location, network structure, and length change many properties of a printed object.
 This includes weight, acoustic resonance, and (if conductive material is present) capacitive signature.
@@
 I also need to actually get an intentional encoding test up and running...
-we are going to have to be clever about how we stimulate the resonance tubes, though.
+we are going to have to be clever about how we stimulate the resonance pipes, though.
 using a speaker won't really work: we have to agitate the air in the chambers, so we might need to use something like a reed (as in reed instruments).}
@@
 As seen in \cite{Ono-touchandactivate}, different objects have different acoustic signatures.
-Additionally, two objects that are visually identical but which have fully enclosed (or other types) of tubes on their interior can be distinguished acoustically.
+Additionally, two objects that are visually identical but which have fully enclosed (or other types) of pipes on their interior can be distinguished acoustically.
 Thus we can recall an object's identity once its acoustic signature has been recorded.
@@
 We have also experimented with intentional encoding.
-Semi-closed tubes can function as resonance chambers: the first resonant frequency $F_1$ Hz of a semi-closed tube length $L$ meters can be found by $F_1 = \frac{c}{4L}$ (where $c$ is the speed of sound).
-All odd harmonics (i.e., $3*F_1$, $5*F_1$, $7*F_1$, ...) of this frequency are also resonant frequencies of the tube.
+Semi-closed pipes can function as resonance chambers: the first resonant frequency $F_1$ Hz of a semi-closed pipe length $L$ meters can be found by $F_1 = \frac{c}{4L}$ (where $c$ is the speed of sound).
+All odd harmonics (i.e., $3*F_1$, $5*F_1$, $7*F_1$, ...) of this frequency are also resonant frequencies of the pipe.
 \valkyrie{need to provide some guidelines for the number of IDs we can create in this way, and where in the sound spectrum they are.
@@
 \valkyrie{in my experiments so far, it seems that the material printed does conduct sound fairly well, so this is feasible.
-I think we should get some equipment and do a small test, though, since I don't know how the resonance will change with the big hunk of plastic around the semi-open tube.
+I think we should get some equipment and do a small test, though, since I don't know how the resonance will change with the big hunk of plastic around the semi-open pipe.
 hopefully discussing this with Alex soon.}

### 2014-03-20 paper/tubes.tex
paper work, adding a new figure

@@
-\section{A Series of Tubes - A Tool for Tube Design}
+\section{A Series of Tubes - A Tool for Pipe Design}
 
-We created a tool which makers can use to create tube-powered interfaces in arbitrary 3D models.
-This tool allows designers to brush over the surface of their model to select exterior connection points of their tubes (see Figure \ref{fig:tool-process-interior}) or to import vector art describing the tubes' interior paths (see Figure \ref{fig:tool-process-exterior}).
-Once the user's selections are made, we create a complete routing using either the exterior connection point method (A*) or the interior path method (graph edge creation, Euler circuit generation).
-We thicken our routing and use a physics-based rod simulation to minimize bending energy of tubes.
-We also apply templates where appropriate (e.g., 3D cross-overs for tubes that intersect in the plane).
-The resultant routing is subtracted from the original mesh.
-The modified mesh can then be 3D printed.
+To allow makers to design novel objects with pipe-powered interfaces, we created a tool for use on arbitrary 3D models.
+This tool allows designers to select exterior connection points of their pipes (see Figure \ref{fig:tool-process-interior}) or to import vector art describing the pipes' interior paths (see Figure \ref{fig:tool-process-exterior}).
+Once the maker's selections are made, we create a complete routing using either the exterior connection point method (A*, physical simulation) or the interior path method (graph edge creation, Euler circuit generation).
+We thicken our routing to create pipes.
+\valkyrie{We also apply templates where appropriate (e.g., 3D cross-overs for tubes that intersect in the plane).}  The resultant pipes are subtracted from the original mesh, which can then be 3D printed.
 
-Our tool is implemented as a part of Meshmixer, a consumer mesh editing tool, in C++.
+Our tool is implemented as a part of Meshmixer, a consumer 3D mesh editing tool, in C++.
 
@@
 
-To create tubes in which the location and shape of exterior connection points matters, we first allow users to select the connection points on the mesh's surface using a brush.
+Designing objects which, for example, are touch sensitive in particular areas or need to integrate with existing electronics, requires precise location and sizing of pipe endpoints.
+The interior of these pipes should have as large a bending radius as possible so that post-print insertion of solid or viscous media (like wire or paint) is easier.
+
+To create pipes in which the location and shape of exterior connection points matters, we first allow makers to select the connection points on the mesh's surface using a brush.
 Our tool then creates an initial shortest-path routing using A* to estimate the routed distance between points.
@@
 
-\subsubsection{Routing}
+\subsubsection{Routing and Physical Simulation}
 
@@
 
-Physical Simulation :
-
-How does it actually work? \valkyrie{Ryan?}
-
-Bend Minimization
-
-In order to force our simulated rods away from the surface of the mesh, we create ``wind'' pushing inward from the exterior of the mesh.  This wind pushes the tube away from the surface.
+To allow for easiest insertion of media post-print, our physical simulation step minimizes the bend radius of the generated rods.
+In order to force our simulated rods away from the surface of the mesh, we create ``wind'' pushing inward from the exterior of the mesh.
+This wind keeps the resulting pipe away from the surface.
+\valkyrie{Ryan: How does it actually work?}
 
 \subsubsection{Tubes with Multiple Endpoints}
+For the creation of tubes with multiple endpoints, for example the star topologies in Figure \ref{fig:toys} or tree topologies, makers can select interior faces of existing pipes as endpoints, and thereby create pipe networks by hand.
+We do not currently offer any pipe network manipulation tools.
 
-To allow designers to create tubes with multiple endpoints (like the star topologies in Figure \ref{fig:toys}), we ....????
-\valkyrie{create 0-size mesh points in the center of the object, to which we attach one end of every tube?}
+\valkyrie{To allow designers to create tubes with multiple endpoints (like the star topologies in Figure \ref{fig:toys}), we create 0-size mesh points in the center of the object, to which we attach one end of every tube?}
+
+\subsubsection{Tubes with Non-Round Endpoints}
+Our tool does not explicitly allow makers to design the shape of their tubes' endpoints.
+However, the larger Meshmixer package does allow brushing arbitrary shapes onto the surface and extruding them inward, as in Figure \ref{fig:meshmixer-endpoint}.
+\valkyrie{should we have a figure about this?
+I  don't know how to describe it really, but we also didn't build it as a part of our tool.}
+
+\begin{figure}[h!]
+\centering
+    \includegraphics[width=3.4in]{figures/meshmixer-endpoint.png}
+\caption{After creating a tube, a maker can select a surface shape using a brush (a).  She can then extrude her selection inward (b).}
+\label{fig:meshmixer-endpoint}
+\end{figure}
 
@@
 
-A user can import a vector graphics file (such as SVG) describing the path she desires for her tubes.
+Neon signs, as in Figure \ref{fig:tool-process-interior}(d), require a single long path that passes through many desired edges: for example, in our UIST sign, the path must pass through all the edges that comprise the word ``UIST''.
+To allow makers to create novel neon signs, as well as similar objects that require design of the pipe's interior path, we built a tool which leverages graph theory to generate this single long path.
+
+A maker can import a vector graphics file (such as SVG) describing the path she desires for her pipes.
 We create a graph based on this input data, then add edges to make it connected and semi-Eulerian.
-We create an Euler tour on the modified graph, and thicken the path to create tubes.
-3D templates are used to resolve tube crossings in the plane.
+We create an Euler tour on the modified graph, and thicken the path to create pipes.
+\valkyrie{3D templates are used to resolve tube crossings in the plane.}  We cut these pipes from a generated block mesh large enough to hold them.
 
@@
 Now to remove our temporary edge and make the graph semi-Eulerian instead of Eulerian, let $E' = E' \setminus e_{temp}$.
-
-\valkyrie{does the proof of this theorem belong in the paper, or in the appendix?  we can push it back there if need be, although it's not super long.}
+{
+\renewcommand{\thetheorem}{\ref{thm:graphness}}
 \begin{theorem}
- $G' = (V, E')$ is a connected, semi-Eulerian graph for which $E \subseteq E'$. \valkyrie{why is there such a huge space after this paragraph?}
+ $G' = (V, E')$ is a connected, semi-Eulerian graph for which $E \subseteq E'$.
 \end{theorem}
+\addtocounter{theorem}{-1}
+}
 \begin{proof}
-$E \subseteq E'$ : we never remove edges from $E$ in our algorithm.  $\therefore E \subseteq E'$.
-
-$G' = (V, E')$ is connected : if $G'$ is not connected, $\exists G_i = (V_i, E_i) \subset G', u \in V_i, v\in V \setminus V_i$ s.t.
-$u$ is not reachable from $v$.
-$G_i \not\in G_{dis}$, because we create edges connecting $G_j, G_k \forall G_j, G_k \in G_{dis}$ and only remove an edge $e_{dupe} = (u, v)$ from $E_{conn}$ once we determine that $u$ is reachable from $v$ along edges $E' \cup E_{conn-min}$.
-$\therefore G_i \not\in G_{dis}$.
-This implies that $G_i$ was not initially disconnected from $G$, because by definition $G_{dis}$ is the set of all disconnected subgraphs of $G$.
-We cannot have disconnected $G_i$ from $G$ because $E \subseteq E'$.
-Thus, a contradiction.
-$\therefore G' = (V, E')$ is connected.
-
-$G' = (V, E')$ is semi-Eulerian : Each edge $e$ in our minimum weight matching connects a unique pair of vertices $v_i, v_j \in V_{odd}$ by the definition of minimum weight matching.
-Each edge $e = (v_i, v_j) \in E_{circuit-min}$ adds one to the degree of $v_i$ and $v_j$, causing them to be of even degree.
-$|V_{odd}| \% 2 = 0$ by the handshake lemma, $\therefore$ all edges can be paired.
-When we remove $e_{temp} = (start, end)$ from $E'$, we cause those vertices, which were of even degree by the above process, to be of odd degree.
-Thus, all vertices except $start$ and $end$ (which are of odd degree) are of even degree.
-$\therefore G'$ is semi-Eulerian.
+See appendix.
 \end{proof}
@@
     \includegraphics[width=3.4in]{figures/placeholder/templates.jpg}
-\caption{When edges intersect in the plane (a), we match them to the nearest of our several intersection templates (b) to create 3D paths that do not interfere with each other.}
+\caption{\valkyrie{When edges intersect in the plane (a), we match them to the nearest of our several intersection templates (b) to create 3D paths that do not interfere with each other.}}
 \label{fig:templates}
@@
 \subsubsection{Edges that Intersect in the Plane}
-Once we have a tour, we must resolve tubes that intersect in the plane.  Because we have a 3-dimensional canvas in which to work, we can push overlapping paths into the third dimension.
+We do not currently redirect edges that intersect in the plane.
+Because we have a 3-dimensional canvas in which to work, it would be possible to push overlapping paths into the third dimension, however the angel hair EL wire we have used for our prototypes is so thin as to make this unnecessary.
+
+\valkyrie{Once we have a tour, we must resolve tubes that intersect in the plane.  Because we have a 3-dimensional canvas in which to work, we can push overlapping paths into the third dimension.
 
@@
 When intersections have been detected, we match one of several templates to the affected edges (see Figure \ref{fig:templates}) using a nearest neighbors technique.
-Our templates are parameterized, so the templates can be modified to accommodate precise intersection characteristics.
-
-\valkyrie{For now, we just let them intersect.  The angel hair EL wire is thin enough that we don't actually care.}
+Our templates are parameterized, so the templates can be modified to accommodate precise intersection characteristics.}
 
@@
 
-How we actually cut stuff and make tubes happen.
-\valkyrie{Ryan?}  We have to actually respect whatever shapes people drew on the surface in the case of the exterior connection tubes (I mean, ideally, we would do a loft operation here).
+How we actually cut stuff and make tubes happen. \valkyrie{Ryan?}
 

### 2014-03-20 paper/designspace.tex
addressing bjoern and tovi comments

@@
 \section{The Design Space of Pipes}
-\bjoern{We should be careful with our terminology - apparently pipes and pipes have different connotations.
-Pipes are measured by outer diameter (OD) and tend to be structural, while pipes are measured by inside diameter (ID) and tend to be used to transport media inside them.
-At the least, we should mention that different communities have these different connotations.
-I think that in our case, we're thinking mostly of the 3D print as the structural object that we're then tunneling through with {\em pipes}.}
-
 We propose the use of pipes to redirect active components through the core and onto the surface of 3D printed interactive objects, as well as to allow for the creation of designed paths.

### 2014-03-20 paper/introduction.tex
addressing bjoern and tovi comments

--- 2014-03-17 paper/introduction.tex
@@
 \centering
-    \includegraphics[width=3.4in]{figures/series-of-tubes.jpg}
-\caption{Here's a teaser figure of some cool stuff we did. Probably it should show one of our example objects, or we can make it really big (figure*) and show them all!}
+    \includegraphics[width=3.4in]{figures/placeholder/teaser.png}
+\caption{A novel neon sign (a) and a touch-sensitive brain (b) designed with our pipe tool and fabricated on a consumer-grade 3D printer.
+(c) and (d) show the structure of the internal pipes generated by our tool.}
 \label{fig:teaser}
@@
 \section{Introduction}
+
 Makers, as well as professional designers, leverage 3D printers as tools for design work.
 A wide array of objects, ranging from bicycle helmets to jewelry to video game controllers, are now prototyped or even manufactured using these machines.
-However, most devices fabricated by 3D printers are passive: accessible printers are not yet capable of creating integrated active systems.
-\bjoern{not sure I understand phrase after the colon.
-Rewrite and explain what you mean in more detail?}
+However, most devices fabricated by 3D printers are passive.
+We propose a novel technique involving the removal of material from the interior of 3D models, prior to printing, to create pipes and other cavities.
+These pipes can be filled, post-print, by a variety of media that enable input, display, and tactile feedback.
+This {\em subtractive} approach is complementary to {\em additive} approaches that try to print different materials, such as conductors.
+While our approach requires manual assembly afterwards, it opens up a large useful design space on commonly available 3D printers, allowing makers to create objects like those in Figure \ref{fig:teaser}.
 
-Willis, et al., ``envision a future world where interactive devices can be printed rather than assembled; a world where a device with active components is created as a single object, rather than a case enclosing circuit boards and individually assembled parts" \cite{Willis-printedoptics}.
-This is a vision we eagerly share: 3D printers are capable of creating arbitrary geometries not feasible to manufacture using traditional processes, and we see these capabilities being underutilized by makers and designers.
-We see many opportunities to increase the interactivity of 3D prints using today's printer technology.
-\tovi{Might want some examples of what possibilities this opens up}
-\bjoern{I think we're doing something a little different from Karl's vision - in all of our examples, you need to hook up extra sensors, actuators, or lights to the 3D-printed object.
-So there's always still assembly required.
-We're not really printing interactive objects in a single pass, but we're offering new ways of adding interactivity to 3D printed objects.
+These channels introduce an entire design space of opportunities for adding interactivity to these objects.
+For example, copper material can fill the channels, to allow for standard electronic components to be easily integrated after the printout; air can be pumped through to create tactile output; or electroluminescent wire can be threaded through to create computer-controlled visual output.
 
-I'd argue the principal novelty of our work is to {\em remove material} from a model to form pipes and other cavities.
-We demonstrate how designers/makers can then use these pipes, e.g.
-by filling them with various media to enable input, display and tactile feedback.
-This {\em subtractive} approach is complementary to {\em additive} approaches that try to print different materials such as conductors.
-While our approach requires manual assembly afterwards, it opens up a large useful design space on commonly available 3D printers.}
+The use of pipes can allow for \emph{redirection} and \emph{centralization} of active components, which creates \emph{iterable} devices.
+For example, to create a touch-sensitive toy as in Figure \ref{fig:teaser}, a maker can select several locations of interest on the surface of the brain.
+After creating pipes to those locations and printing her object, she can fill the pipes with conductive paint.
+For sensing, she need only connect a single wire to a shared out of all the tubes, and can distinguish them via Swept Frequency Capacitve Sensing \cite{Sato-touche}.
+If she decides that another area of interest is important to distinguish, she can simply print a new model with a new pipe leading to that location: the electronics assembly remains the same and she need only connect the same single wire to the tube out.
+\valkyrie{not clear that this is our contribution.}
 
-Using specially modified printers or extra machinery, it is possible to create electronics on the surface of 3D prints \cite{optomec} \cite{Sells-reprap}.
-These techniques, however, require a high capital investment and technical expertise.
-In addition, they lack flexibility: the printed circuits must be routed in 2D to be created on the objects' manifold exteriors, and they can only be used to create electronic circuits.
-\valkyrie{what?}  \tovi{I dont follow this paragraph - I think you want  to briefly summarize what has been previousl done in interactive 3d prints, and hihglight the shortcommings and remianing challenges} \bjoern{agree.
-Also, probably too much focus on electronics since it's only one of a series of examples we'll demonstrate.}
+Pipes can also be used for sending media on a specified route through an object; for example, neon signs are created by evacuated glass tubes formed into a specific path.
+Electricity is run through the near-vacuum, and this makes the path visible.
+If our maker wants to build such a sign, she can substitute Electroluminescent (EL) wire for the evacuated glass and electricity: she need only design a path for the wire to be fed through and then create pipes along that path.
 
-In addition to electronics, many other means of interactivity exist.
+The generation of these pipes, however, is not trivial.
+Pipes need to be as straight as possible to ease the insertion of media post-print.
+Independent pipes should not intersect, as this could lead to electrical shorting or other problems.
+Pipes should not pass too close to the surface (unless they are being used for display purposes), especially since consumer-grade 3D printers do not guarantee air- or water-tightness of their prints, so more than one layer may be required to prevent leakage.
 
-\valkyrie{Points to hit: there is more than electronics to worry about, like fluids and the organic haptic feedback they afford; tangible interfaces are important (do I have to say this every time??); assembly sucks}
+Our work makes the following contributions to the field of digitally fabricated interactive objects:
 
-In this paper, we propose a new technique, where interior channels, such as hollow chambers or solid \bjoern{?}, are integrated into 3D models.
-Such channels introduces a entire design space of opportunities for adding interactvity to 3D printed objects.
-For example, copper material can fill the channels, to allow for standard electronic components to be easily integrated after the printout; air can be pumped through to create tactile output; or, Electroluminescent wire can be passed through to create interactive visual output.
-\bjoern{this is good - I'd make this the central argument of the intro and move it up earlier.}
-
-Our work makes the following contributions to the field of interactive digitial fabrication.
-First, ...
-design space...Second ...
-modeling tool and algorithms ...
-Third showcase a set of examples, enabled by our modeling tool.
-\bjoern{I'd say more about the software contribution in the abstract.
-Maybe that just requires expanding these notes into a full paragraph.}
+\begin{itemize}
+\item We lay out a design space of pipe-mediated interactive possibilities. 
+\item We offer algorithms and techniques for routing pipes, as well as a design tool implementing them.
+\item We showcase a set of examples, enabled by our modeling tool and exploring new points in the design space.
+\end{itemize}
+\bjoern{I'd say more about the software contribution in the abstract. Maybe that just requires expanding these bullets into a full paragraph.}

### 2014-03-20 paper/introduction.tex
integrate new temp figure

@@
 Independent pipes should not intersect, as this could lead to electrical shorting or other problems.
-Pipes should not pass too close to the surface (unless they are being used for display purposes), especially since consumer-grade 3D printers do not guarantee air- or water-tightness of their prints, so more than one layer may be required to prevent leakage.
+Pipes should not pass too close to the surface (unless they are being used for display purposes).
+Consumer-grade 3D printers do not guarantee air- or water-tightness of their prints, so more than one layer may be required to prevent leakage.
+Finally, pipes
 

### 2014-03-20 paper/designspace.tex
addressing some comments, but mostly just copying the tovi comments in

@@
 \section{The Design Space of Pipes}
-We propose the use of pipes to redirect active components through the core and onto the surface of 3D printed interactive objects, as well as to allow for the creation of designed paths.
+The design space of pipes is large.
 By using different types and topologies of pipes and inserting different media into the completed pipes, a variety of inputs and outputs can be located at arbitrary points on the surface of a 3D print.
-We describe the design space of pipes, and then follow up with some of the input and output possibilities---both those that have been previously explored and those that are possible but remain unexplored.
+We describe the  space, and then follow up with some of the input and output possibilities---both those that have been previously explored and those that are possible but remain unexplored.
+We also discuss some possible implications of pipes in 3D prints.
 
-\begin{figure}[h]
+\begin{figure}[h!]
 \centering
@@
 
+\tovi{The table indicates that media is the most
+prominent variable in the design space.
+Should justify that.
+}
+
+\tovi{There's lots of great ideas an examples in the inputs and outputs section, but the orgnization and coorespondence to the table is a little hard to follow. }
+
 By crossing pipe types, topologies, and inserted media, we create a space of possible inputs and outputs that can be pipe-mediated.
@@
 Those created with fabrication are in {\color{blue}blue}.
-In \emph{{\color{red}red italic}} are unexplored interactions creatable with custom-fabricated pipes.}
+In \emph{{\color{red}red italic}} are unexplored interactions creatable with custom-fabricated pipes.
+\tovi{Im not sure how well this table will work.
+Division between input and output should be more prominent.}}
 \label{fig:designspace}
@@
 
-Touch sensing is enabled through pipes filled with conductive media.
-Traditional wires (threadable) may be used, but we have had success with conductive paint (liquid).
+Touch sensing can be enabled through pipes filled with conductive media.
+Traditional wires (threadable) may be used, but conductive paint (liquid) may be more flexible.
 Using conductive paint and Swept Frequency Capacitive Sensing (SFCS \cite{Sato-touche}), an interior star pipe topology can enable single-wire touch- and grasp-sensing at any set of points on a printed object's surface (see Figure \cite{fig:toy}).
@@
 
-\subsection{Identity}
+\subsection{Implications}
+
+\tovi{Where does this fit in the design space? This seciton seems out of place.
+}
+
+\valkyrie{is this something we want to talk about at all?}
+
+Pipes included in objects can have additional effects beyond their designed intentions.
 
 Pipes included in printed objects allow for their identification.
-While the pipes need not be visible (especially in the case of fully-enclosed pipes), their presence, location, network structure, and length change many properties of a printed object.
+While the pipes need not be visible (especially in the case of fully-enclosed pipes), their presence, location, network structure, and length change the physical properties of a printed object.
 This includes weight, acoustic resonance, and (if conductive material is present) capacitive signature.
@@
 Thus we can recall an object's identity once its acoustic signature has been recorded.
+
+\tovi{I would save these details for later, 
+when we tlak about our actual 
+implementations.}
 

### 2014-03-20 paper/exampleobjects.tex
integrate new temp figure
addressing some comments, but mostly just copying the tovi comments in

@@
 To evaluate our tool's capabilities and to explore new points in the design space, we fabricated a set of five prototype objects designed using the tool.
+
+\tovi{Maybe a materials section 
+before this - to talk about the 
+specific materilas we used, 
+such as the copper paint 
+specifications, el lighting, etc.
+
+Need to highlight which aspect of the design space each example object relates to. 
+}
 
@@
 \end{figure}
+
+\tovi{go into a bit more detail about modelling , fabrication, and
+assembly process, and apparatus details (especially the 
+copper paint - unless discussed earlier in a materials 
+seciton.}
 
@@
 \centering
-    \includegraphics[width=3.4in]{figures/placeholder/radio.jpg}
+    \includegraphics[width=3.4in]{figures/radio.png}
 \caption{This radio is assembled from traditional electronic components connected by copper-filled tubes.
@@
 \centering
-    \includegraphics[width=3.4in]{figures/placeholder/sign.jpg}
+    \includegraphics[width=3.4in]{figures/sign.png}
 \caption{A two-state animated neon sign designed using our tool (a)(b) \valkyrie{Tovi: we could use your PDF hax to make this animate in place :)}.

### 2014-03-20 paper/introduction.tex
addressing some comments, but mostly just copying the tovi comments in

@@
 Pipes should not pass too close to the surface (unless they are being used for display purposes).
-Consumer-grade 3D printers do not guarantee air- or water-tightness of their prints, so more than one layer may be required to prevent leakage.
-Finally, pipes
+Consumer-grade 3D printers do not guarantee air- or water-tightness of their prints, so more than one layer may be required to prevent fluid leakage.
+We have developed techniques for minimizing the bending energy of, enforcing non-intersection with, and avoiding the surface for pipes.
+These techniques are embodied in our software tool to enable makers to create pipe-powered interfaces.
 
@@
 \begin{itemize}
-\item We lay out a design space of pipe-mediated interactive possibilities. 
+\item We lay out a design space of pipes and their types, contained media, topologies, and design focus.
+\item We describe the interaction possibilities of pipes, including prior work as well as uncharted possibilities.
 \item We offer algorithms and techniques for routing pipes, as well as a design tool implementing them.
@@
 \end{itemize}
-\bjoern{I'd say more about the software contribution in the abstract. Maybe that just requires expanding these bullets into a full paragraph.}

### 2014-03-20 paper/relatedwork.tex
addressing some comments, but mostly just copying the tovi comments in

@@
 
-While using a series of tubes can replicate the interaction techniques discussed (and more), our work's \emph{contribution} is more generally related to two existing research areas: fabrication and routing.
+While using a series of tubes can replicate a variety of interaction techniques, our work's \emph{contribution} is more generally related to two existing research areas: fabrication and routing.
 
-\subsection{Fabrication}
+\subsection{Fabrication of Interactive Devices}
 
-Previous work has investigated the integration of interactivity into the fabrication process.
+Previous work has investigated the integration of interactivity with the fabrication process.
 \cite{Willis-printedoptics} used light pipes, solid clear tubes, to create integrated displays: our work allows a superset of these capabilities, and Printed Optics is a part of our design space.
@@
 This technique is complementary to ours, as we allow for placing these interactive elements anywhere on an object's surface while they had to place them in locations accessible for electronics insertion.
+
+\subsubsection{Fabricating Channels}
 
@@
 
-\subsection{Routing}
+\subsection{Routing and Internal Modeling}
 
@@
 
+Prior work on modeling the interior of objects (as opposed to their exterior) has been limited.
+\valkyrie{ryan, can you give citations here?}  Sauron automatically modified the interior of existing mechanical objects to make them amenable to vision-based sensing \cite{Savage-sauron}, while Willis, et al.,'s Infrastructs project created encoded interior cavities in 3D printed objects designed to be sensed with a terahertz imaging tool.
+Our work differs in that we want to allow users to design pipes and integrate interactive objects with them.
+
 \subsection{Misc}

### 2014-03-20 paper/tubes.tex
addressing some comments, but mostly just copying the tovi comments in

@@
 
+\tovi{discussion points 
+- remove support material
+-evenutally medium could be printed into tube}
+
 \begin{figure}[h!]
@@
 \subsection{Designing Interior Paths}
+
+\tovi{Is this useful for anything other 
+than the neon sign usage 
+scenario? It's probably too 
+detailed for the purpose which it 
+serves}
 
@@
 
+\tovi{I think this seciton is actually 
+important. Should talk about 
+support material, and 
+feasability of printing hollow 
+channels (vs channels with 
+support material) on different 
+printer models. 
+Also talk about process of 
+removing support material.
+And finally, process of 
+printing models in multiple 
+pieces, to aid support material 
+removal and threading.
+Would be nice if the software 
+automatically provided the 
+slicing and snap together 
+parts based on the channel 
+geometry}
+
 \subsubsection{Printing} - different strategies with Objet (all print-in-place) and Makerbot (may need to add things like balloons afterwards).

### 2014-03-24 paper/appendix.tex
addressing comments

@@
 
-\section{Proof of Theorem 1}
+\section{Routing Algorithm}
+
+We want a semi-Eulerian graph (i.e., we want a graph in which every vertex but two have even degree) so that the desired medium can be inserted, traverse every edge, and exit the graph at a different vertex.
+Let $G=(V,E_0)$ s.t.
+$\forall e \in E_0, weight(e)=0, start \in V $ the start point $end \in V$ the end point.
+Let $e_{temp} = (start, end)$, $E' = E + e_{temp}$.
+
+We need to connect disconnected subgraphs in $G$.
+Let $G_{dis} = \{G_1, G_2, ...
+G_n\} \in G$ s.t.
+$\cup{\{G_i \in G_{dis}\}} = G$ and $G_i \cap G_j  = \{ \} \forall G_i, G_j \in G_{dis}, i\neq j$ be disconnected subgraphs in G.
+Create $E_{conn} = \{e = (u, v), u \in G_i, degree(u)\%2 = 1, v \in G_j, degree(v) \%2 = 1 i \neq j, weight(e) = distance(u, v)\}$ be the set of all edges connecting odd-degree points in two distinct disconnected subgraphs.
+To keep only the most desirable edges in $E_{conn}$, contract each subgraph $G_i$ to a single point $u_i$ with the shortest outgoing edges to each other subgraph: $E_{conn} = \cup_{G_i \in G_{dis}} \{e=(u_i, v), v \in G_j,$ s.t.
+$\forall E_{conn} \ni e_k = (u, v), u \in G_i, v \in G_j, weight(e_k) \geq weight(e)\}$.
+Sort $E_{conn}$ s.t.
+$E_{conn} = \{e_1, e_2, ...
+e_n\}, weight(e_1) \leq weight(e_2) \leq \ldots \leq weight(e_n)$.
+
+Let $E_{conn-min} = \{ \}$.
+Beginning with $e_1$, add the first edge $e_i \in E_{conn}$ to $E_{conn-min}$.
+Then let $E_{dupe} = \{ e = (u, v) \in E_{conn}$ s.t.
+$u$  is reachable from $v$ along edges $E' \cup E_{conn-min}$.
+Let $E_{conn} = E_{conn} \setminus E_{dupe}$.
+Repeat until $E_{conn} = \{ \}$.
+Let $E' = E' \cup E_{conn-min}$.
+
+At this stage, we want an Eulerian graph: i.e., we need every vertex to be of even degree.
+Let $V_{odd} = \{ v \in V$ s.t.
+$degree(v) \% 2 = 1 \}$.
+Create $E_{circuit} = V_{odd} \times V_{odd}$, s.t.
+$\forall e=(u,v) \in E_{circuit}, weight(e) = distance(u, v)$.
+Find minimum matching $E_{circuit-min} \subseteq E_{circuit}$.
+Let $E' = E \cup E_{circuit-min}$.
+
+Now to remove our temporary edge and make the graph semi-Eulerian instead of Eulerian, let $E' = E' \setminus e_{temp}$.
 

### 2014-03-24 paper/exampleobjects.tex
addressing comments

@@
 To evaluate our tool's capabilities and to explore new points in the design space, we fabricated a set of five prototype objects designed using the tool.
+All prototypes were fabricated in a single piece, unless described otherwise.
 
@@
 specifications, el lighting, etc.
+} \valkyrie{I'm not clear on what you are asking for here}
 
-Need to highlight which aspect of the design space each example object relates to. 
-}
+\valkyrie{Bj\"{o}rn, I can't figure out what happened to my figures?  and why they got thrown down into a weird state with my limitations section, which appears now in two places???}
 
-\subsection{Touch-sensitive Toys}
+\subsection{Touch-sensitive Toys (open, liquid, star, exterior)}
 
@@
 Sensing is done on a single wire via SFCS.
-Inset shows the internal structure of the tubes generated by our design tool.}
+Inset shows the internal structure of the tubes generated by our design tool.}\input{limitations.tex}
+
 \label{fig:toys}
@@
 We built a smart base which can distinguish between the toys and also determine which toy is mounted: since each toy and each gesture has a distinct capacitive signature, we use a simple classifier trained to detect both toy and gesture based on profile.
-All components were fabricated on a Makerbot Replicator 2.
+All components were fabricated on a Makerbot, support-free.
 
-\subsection{Breathing Bunny}
+\subsection{Breathing Bunny (semi-closed, gas, exterior)}
 
@@
 We connected one tube to each of our pump's terminals, and using a programmable power supply we can mimic a rabbit's breathing pattern.
-This example was printed on our Objet Connex 260.
+This example was printed on our Objet in rubberlike material, with support flushed post-print.
 \valkyrie{this paragraph is really unclear}
 
-\subsection{Custom Radio}
+\subsection{Custom Radio (open, threadable, exterior)}
 
@@
 For the radio functionality, we attached an Si4703 to the Arduino.
-We fabricated this device on our Makerbot Replicator 2.
+We fabricated this device on our Makerbot, support-free but cut in two halves to allow the Arduino to be inserted.
 
-\subsection{Presence-aware Pen Holder}
+\subsection{Presence-aware Pen Holder (open, solid, exterior)}
 
@@
 Our presence-aware pen holder can distinguish which tool or tools a user has picked up (see Figure \ref{fig:pens}).
-Such information was used by \cite{Mueller-constructable} to determine which physical laser pointer a designer was using to interact with their interface.
 Our pen holder uses a modification of the FlyEye technique described by Wimmer in \cite{Wimmer-flyeye} and contains open tubes filled with fiber optic cables, one per pen chamber.
-Our modification of using a single cable per chamber is possible because we use 6mm diameter fiber optic cables, in comparison to the angel hair tubes used in the original work, and can thus send and receive through the same cable.
+We use a single cable per chamber; our 6mm diameter fiber optic cables, in comparison to the angel hair tubes used in the original work, can send and receive through the same cable.
 At the base of each tube is a QRE1113 line sensor digital breakout board, which has an integrated IR emitter and receiver.
 When a pen is in its appointed place, the emitted infrared light is reflected off its bottom and travels back to the receiver, where it registers as bright.
-This prototype was built by our Objet Connex 260.
+This prototype was built by our Objet and support was flushed post-print.
 \valkyrie{this whole paragraph is unclear}
 
-\subsection{Maze}
+\subsection{Maze (fully enclosed, particulate, tree, interior)}
 
@@
 To fabricate the tube, we created it in two halves that were fastened together via glue, due to limitations in our ability to remove support material otherwise.
-We created this prototype on our Makerbot Replicator 2.
+We created this prototype on our Makerbot in two halves to allow for insertion of the marble.
 
-\subsection{Animated Neon Sign}
+\subsection{Animated Neon Sign (semi-closed, threadable, interior)}
 
@@
 They have been threaded with EL wire which is lit in sequence to create an animation.
-Our sign was fabricated on the Objet Connex 260.
+Our sign was fabricated on the Objet in two halves, cutting through the plane of the sign itself, with clear material for the sign around the input design and opaque everywhere else.
+We manually flushed support from the non-planar connecting tubes.
+The EL wire was threaded post-print, and is controlled by an EL wire sequencer.

### 2014-03-24 paper/io.tex
addressing comments

--- 2014-03-20 paper/designspace.tex
@@
-\section{The Design Space of Pipes}
-The design space of pipes is large.
-By using different types and topologies of pipes and inserting different media into the completed pipes, a variety of inputs and outputs can be located at arbitrary points on the surface of a 3D print.
-We describe the  space, and then follow up with some of the input and output possibilities---both those that have been previously explored and those that are possible but remain unexplored.
-We also discuss some possible implications of pipes in 3D prints.
+\section{Tubular Inputs and Outputs}
 
-\begin{figure}[h!]
-\centering
-    \includegraphics[width=3.4in]{figures/tubespace.pdf}
-\caption{The design space of pipes.  Pipe types, media, topologies, and design are discussed more fully in the text.}
-\label{fig:pipespace}
-\end{figure}
+\valkyrie{What do you guys think about trying to merge this with the example objects section?  It would cut down the text a lot, but it might seem weird to just shunt those in here.}
 
-\subsection{Types of Pipes}\tovi{Maybe rename types to 
-"openings"?}
-
-Pipes can come in four types: open, return, semi-closed, and fully enclosed.
-Each of these types offers distinct interaction capabilities (see Figure \ref{fig:pipespace}).
-Definitionally, in the following paragraphs we refer to a pipe which is ``open on the user side'' to have its contents physically accessible to an interacting user at interaction time.
-This includes pipes which function as fountains, but it does not include pipes whose ends have been capped to keep water inside (whether this capping was done at print time, or later as a manual modification).
-Pipes whose contents are not physically accessible to the user are ``closed on the user side''.
-A pipe which is ``open on the system side'' can have its contents manipulated by a machine, e.g., by resonating, pressurizing, or passing electricity through those contents.
-The system does not need to have physical access to the contents in order to manipulate them.
-
-Open pipes originate from the system side and connect the user side, with both ends of the pipe open.
-This type of pipe may be used to create, for example, capacitive sensors: an open pipe filled with conductive paint can be connected to a sensing platform (e.g., Arduino) on the system side, while a user can touch the uncapped other end of the pipe.
-Using Swept Frequency Capacitive Sensing \cite{Sato-touche} or other techniques, a user's touch of the open end can be sensed.
-An open pipe can also be used for output, for example by creating in-air vortices as in \cite{Sodhi-aireal}.
-
-A return pipe originates at the system and returns back to it.
-By threading an electroluminescent (EL) wire through a clear return pipe, a maker can create a custom piece of neon art.
-If a return pipe passes very close to the surface of a 3D printed model, warm or cold water passed through the pipe could be used for temperature-based haptic feedback.
-
-Semi-closed pipes are open at the system end (for control of the enclosed medium) and closed at the user end.
-The closed interface is interesting when it fabricated from a mobile material: for example, a series of pipes terminated in thin rubber membranes on the user side can be actuated by an air pump to create haptic feedback, or a hinged cap which can be blown open by air pressure.
-Without a printer capable of fabricating flexible material, a maker could affix a balloon to an open pipe's end to behave similarly; another possibility is to use semi-closed pipes as audio-generating resonance chambers.
-
-A fully enclosed pipe has no openings on either the system side or the user side.
-Fully enclosed pipes can be used as resonance chambers (e.g., for object identification), or as air bubbles (e.g., as used in \cite{Willis-printedoptics} for internal display).
-\tovi{used as a container for water
-or other particles that would 
-otherwise fall out?}  \valkyrie{I re-worded above.
-please check.}  Their physical design space is very limited, as any support material required to create their internal geometry cannot be accessed for removal.
-
-\subsection{Media in Pipes}
-
-Pipes can be filled with a variety of media to create different interface affordances and capabilities.
-
-``Gas'' comprises all compressible fluids.
-Use of fluid pressure inside pipes can create haptic feedback at semi-closed interfaces, or the gases can be used as carriers for scents or fog.
-As in \cite{Slyper-pressure}, structures can be engineered to change in air pressure when manipulated correctly (e.g., a spiral that changes pressure when twisted, but not pressed), and thus fluid pressure can also be used as an input.
-\tovi{I realize "fluid" pressure is
-technically correct, but might 
-confuse readers, since this is 
-the gas pararaph.}
-
-Incompressible fluids (``liquids'') can perform many of the same interface tasks as gases.
-One opportunity with liquids is to fill the interior of pipes with them and cap the ends.
-In addition, one can use driable conductive fluids, such as copper paint, to coat the interior of pipes and allow them to function as arbitrarily-shaped wires.
-This is especially helpful for the creation of a shared ground, or for creating single-wire capacitive interfaces amenable to sensing with SFCS \cite{Sato-touche}.
-
-\tovi{Again, this raises the question of how we are applying the design 
-space - is it based on its properties when it is printed, or based on 
-any post-print modifications}\valkyrie{re-worded above.
-I think we are basing it on status at ``interaction time"}Pipes need not have hollow centers: in the case where routed pipes are filled with solid material---in particular, a solid material different from the model material---, interactions such as those in \cite{Willis-printedoptics} are possible.
-
-Particulates, either printed in-place or inserted, can be of varying densities.
-A single particle can be used for display.
-Sparse particles in a stream of fluid can provide haptic feedback.
-Dense particles in a semi-closed pipe allow for jamming-based interactions at any point on the surface of an object \cite{Follmer-jamming}.
-
-Threadable inserted elements, such as electroluminescent (EL) wire or fiberoptic cables, are those that can be threaded through pipes post-printing.
-This allows overcoming limitations of printers: for example, a Printed Optics-style interface can be created on an inexpensive consumer-grade 3D printer using pipes and inserted fiber optic cable.
-
-\subsection{Pipe Topologies}
-
-Pipe network topologies enable different types of interactions.
-
-Splitting or mixing pipes offer flexibility in output.
-If a maker wished to create a painting device, she might wish to have two system-side pipes feeding in primary-colored red and blue paints which mix in varying ratios, allowing their pigments to combine before purple exits from the device (see Figure \ref{fig:pipespace}).
-Splitting can also be useful, for example if our maker wants red paint output in two locations from her painting device, she could have one system-side pipe, but split the pipe into two (see Figure \ref{fig:pipespace}).\tovi{How about multiple in/
-multiple out (in contrast to 
-branching/splitting which is 
-single in multiple out)
-} \tovi{This seciton should refer to the 
-possibilities, not to what we 
-actually did.}
-
-Star and tree topologies are extensions of the splitting and mixing primitives.
-Using a star topology in which the pipes were filled with conductive paint, we created toys with several touch-sensitive areas, see Figure \ref{fig:toys}.
-
-\subsection{Design of Pipes}
-
-Pipe designs may emphasize either their exterior features (connection points) or their interior paths.
-These two features lead to different kinds of interfaces.
-An example interface that focuses on the exterior connection points is the touch sensitive toy in Figure \ref{fig:toys}, where pipes must exit the toy at the eyes, ears, tail, etc.
-An interface focused on the internal path of the pipe is the neon sign in Figure \ref{fig:neon}: output is based on the shape of the pipe.
-
-\subsubsection{Dimensions of Pipes}
-
-\valkyrie{Reference those pages from Bjoern: http://ttb.com/recent-news/diagrams/ We can talk about wall thickness, tangent length, degree of bend, arg bend angle, inner and outer diameters, centerline radius, etc.}
-
-\subsection{Inputs and Outputs}
-
-\tovi{The table indicates that media is the most
-prominent variable in the design space.
-Should justify that.
-}
+Once objects have been designed and fabricated, the major flexibility for prototyping with pipes stems from the media inserted in them.
+The five types of media--gas, liquid, solid, particulate, and threadable--give rise to unique possibilities in output (visual, aural, tactile/haptic, olfactory/gustatory) as well as input (touch/pressure and others).
 
@@
 \tovi{Im not sure how well this table will work.
-Division between input and output should be more prominent.}}
+Division between input and output should be more prominent.} \valkyrie{one possibility is making the table larger and adding some short descriptions.
+we could throw out a lot of words from the paper that way}}
 \label{fig:designspace}
@@
 
-\subsubsection{Inputs}
+\subsection{Outputs}
 
-Pipes create opportunities for many types of input sensing across the surface of printed devices.
-We describe sensing touch, pressure, grasp, flexing, tapping, and manipulation of traditional electronic components.
-We recognize that there are likely more input sensing techniques, both existing and on the horizon, that are compatible with the use of pipes.
-
-Touch sensing can be enabled through pipes filled with conductive media.
-Traditional wires (threadable) may be used, but conductive paint (liquid) may be more flexible.
-Using conductive paint and Swept Frequency Capacitive Sensing (SFCS \cite{Sato-touche}), an interior star pipe topology can enable single-wire touch- and grasp-sensing at any set of points on a printed object's surface (see Figure \cite{fig:toy}).
-Simple capacitance measurements are possible using the same media, and custom-designed sensors like those created in Savage, et al.,'s Midas system are also possible \cite{Savage-midas}.
-
-Pressure can be sensed in multiple ways.
-Slyper, et al., in \cite{Slyper-pressure} contribute a set of semi-closed 3D printable primitives that respond to particular actions, like twisting or pushing, with changes in internal air pressure.
-These primitives can be attached as the terminus of any semi-closed pipe; the user can manipulate the printed endpoint and the system can sense air pressure changes.
-Pressure can also be more simply sensed via capacitance using those techniques.
-
-Grasp sensing can also be enabled using Wimmer's FlyEye technique \cite{Wimmer-flyeye}.
-This technique involves a pair of optical links at each point of interest: one of each pair is connected to an infrared (IR) LED and the other to a camera.
-When some object nears or touches a pair of cables, IR light from the LED is reflected back to the camera and the touched point appears as a bright spot.
-Follmer, et al.,'s Jamming User Interfaces \cite{Follmer-jamming} rely on dense particulates enclosed in a flexible volume, and can be sensed optically.
-These dense particulates can be printed in-place with desired characteristics (such as varying particle size or shape) using sacrificial support material, or added post-print.
-The optical links necessary for sensing with FlyEye or Jamming can be created either via clear solid cores as in \cite{Willis-printedoptics} or via fiber optic cables threaded through hollow pipes (see Figure \ref{fig:pens}).
-
-Some 3D printers, like our Objet Connex 260, can fabricate rubber-like materials.
-Much like Slyper, et al., in \cite{Slyper-shape}, we can sense flexing and bending of prints made on these machines.
-While Slyper, et al., created multi-level multi-stage silicone molds with embedded wires to build their sensors, by placing pipes in our rubberlike prints and inserting conductive media afterwards, we can perform the same flex sensing.
-
-Tapping is another input possibility.
-Due to differential sound conductivity in different printed materials (in particular, the rubberlike Objet material compared to the hard Objet material \valkyrie{presumably we should get or take some measurements of this and include it in the appendix?}), pipes of a greater conducting material can be embedded in a model of a lesser conducting material.
-A microphone or piezo placed at the system-side of these sound-conducting pipes can determine where the model was tapped and how it was tapped.
-Active acoustic sensing as described in Touch \& Activate \cite{Ono-touchandactivate} is also possible; this technique would allow for particular areas of interest (connected to sound-conducting pipes) to be touch sensitive, while other areas (made of non-conducting material) are not sensed.
-
-Sensing via traditional electronic components (e.g., potentiometers, alcohol gas sensors) can be accomplished by conductive inserted media.
-Components can be recessed into a print's surface, with their leads implanted into open pipes.
-By inserting liquid copper paint instead of threading traditional wires, all components in a model can share a ground line, and the paint's drying process obviates the use of solder or glue to affix the components in place (see Figure \ref{fig:radio}).
-
-\subsubsection{Outputs}
-
-While many input techniques require the use of conductive media inserted into pipes, output media are more varied.
 We describe potential pipe-based outputs organized by the five senses: visual, aural, haptic, and olfactory/gustatory.
@@
 \emph{Visual} outputs can be mediated by gases, liquids, solid, particulates, or threadables.
-EL wire can be threaded through custom paths to create neon signs, as in Figure \ref{fig:neon}.
-Colored liquids can be mixed and split before exiting an opaque device, or in a transparent device their mixing, splitting, and paths could be used for display.
-PneUIs \cite{Yao-pneui} can be fabricated as single parts: the rubberlike material offered by our Objet is inflatable when thin (200\% elongation at break), and a PneUI located at the terminus of a pipe could be actuated by a system-side air pump (see Figure \ref{fig:breathe}).
-Mechanical motion can be mediated through fluids (a light object which is partially contained in a pipe or located at its end could be pushed by the fluid pressure of the pipe) or threadables (e.g., muscle wire).
-Large particles trapped in pipes visible to the user can be used for, e.g., maze games (see Figure \ref{fig:maze}).
-We can even direct fluids through a pipe network using a servo-controlled gate, as in Figure \ref{fig:direct}.
-Using techniques like those described by Harrison, et al., in \cite{Harrison-buttons}  \valkyrie{this paragraph seems terrible.
-I think all these paragraphs might seem terrible...}
+Pneumatic actuation (as in \cite{Yao-pneui} or \cite{Harrison-buttons}) provides output, but gases carrying smoke (as in \cite{Sodhi-aireal} demo) also offers visual feedback.
+Splitting and mixing of colored liquids can be used as a display tactic; additionally mechanical motion can be fluid-mediated.
+Liquids can also be mixed in specific concentrations by servo-controlled gate structures as in Figure \ref{fig:direct}.
+Printed optics \cite{Willis-printedoptics} explored visual output at the ends of pipes, and   EL wire can be threaded through custom paths to create neon signs, as in Figure \ref{fig:neon}.
 
 \emph{Aural} outputs can be created with gases.
-Zoran has successfully 3D printed a functional flute, which in essence is a long open open pipe with semi-closed pipes coverable by keys \cite{Zoran-flute}.
-Helmholz resonance (the phenomenon that creates noise when you blow into the top of a glass bottle, which is also how some musical instruments, like ocarinas, work) can be leveraged to create computer-controllable sound chambers.
+Helmholz resonance can be leveraged to create computer-controllable sound chambers, as in Zoran's flute \cite{Zoran-flute}.
 All that is necessary for this is an enclosed chamber with a pipe connected to it via a narrow neck-point (see Figure \ref{fig:ocarina}).
-Passive sound amplification (like a phonograph horn) and sound redirection are also possible using pipes.
+Passive sound amplification and sound redirection are also possible using pipes.
 
@@
 \emph{Haptic} outputs are possible through compressible and incompressible fluids.
-Fluids can actuate particulates and create programmable pliability (as discussed above and in \cite{Follmer-jamming}).
+Fluids can actuate particulates and create programmable pliability \cite{Follmer-jamming}.
 Additionally, use of semi-closed pipes with rubberlike material as the caps allows for gases to actuate surface features (see Figure \ref{fig:breathe}).
-Sodhi, et al., in \cite{Sodhi-aireal} created free-air haptic feedback using controlled vortex generation: this technique could also be reproduced using a series of pipes.
-Iwata, et al., explored volumetric fluid-based interactions with programmatically inflatable balloons and an overhead projector in \cite{Iwata-volflex}, and additionally haptic outputs can be used to increase ergonomics, as in the inflatable mouse by Kim, et al.
-\cite{Kim-inflatablemouse}.
+Free-air haptic feedback \cite{Sodhi-aireal} and volumetric interactions \cite{Iwata-volflex} can be pipe-mediated.
 
@@
 
-\subsection{Implications}
+\subsection{Inputs}
 
-\tovi{Where does this fit in the design space? This seciton seems out of place.
-}
+Pipes create opportunities for many types of input sensing across the surface of printed devices.
+We describe sensing touch, pressure, grasp, flexing, tapping, and manipulation of traditional electronic components.
+There are likely more input sensing techniques, both existing and on the horizon, that are compatible with the use of pipes.
 
-\valkyrie{is this something we want to talk about at all?}
+\emph{Touch} sensing can be enabled through pipes filled with conductive media.
+Traditional wires (threadable) may be used, but conductive paint (liquid) may be more flexible.
+Using conductive paint and Swept Frequency Capacitive Sensing (SFCS \cite{Sato-touche}), an interior star pipe topology can enable single-wire touch- and grasp-sensing at any set of points on a printed object's surface (see Figure \cite{fig:toys}).
+Simple capacitance measurements are possible using the same media, and custom-designed sensors like those created in Savage, et al.,'s Midas system are also possible \cite{Savage-midas}.
 
-Pipes included in objects can have additional effects beyond their designed intentions.
+\emph{Pressure} can be sensed in multiple ways.
+Slyper, et al.,'s primitives can be attached as the terminus of any semi-closed pipe; the user can manipulate the printed endpoint and the system can sense air pressure changes \cite{Slyper-pressure}.
+Pressure can also be sensed via capacitance.
 
-Pipes included in printed objects allow for their identification.
-While the pipes need not be visible (especially in the case of fully-enclosed pipes), their presence, location, network structure, and length change the physical properties of a printed object.
-This includes weight, acoustic resonance, and (if conductive material is present) capacitive signature.
+\emph{Grasp} sensing can be enabled using Wimmer's FlyEye techniques \cite{Wimmer-flyeye} or those described in Jamming User Interfaces \cite{Follmer-jamming}.
+The optical links necessary for sensing can be created either via clear solid cores as in \cite{Willis-printedoptics} or via fiber optic cables threaded through hollow pipes (see Figure \ref{fig:pens}).
 
-\valkyrie{need to do a rewrite of this that doesn't necessarily focus so deeply on audio only.
-we can talk about these other non-visible measurements like weight and capacitive signature.
-I also need to actually get an intentional encoding test up and running...
-we are going to have to be clever about how we stimulate the resonance pipes, though.
-using a speaker won't really work: we have to agitate the air in the chambers, so we might need to use something like a reed (as in reed instruments).}
+Some 3D printers can fabricate rubber-like materials.
+Similar to Slyper, et al., in \cite{Slyper-shape}, we can sense \emph{flexing} and \emph{bending} of prints made on these machines: the ability to 3D print devices to sensing bending and flexing significantly cuts molding and assembly time.
 
-Both identification by recall and intentional encoding are possible.
-As seen in \cite{Ono-touchandactivate}, different objects have different acoustic signatures.
-Additionally, two objects that are visually identical but which have fully enclosed (or other types) of pipes on their interior can be distinguished acoustically.
-Thus we can recall an object's identity once its acoustic signature has been recorded.
+\emph{Tapping} is another input possibility.
+Due to differential sound conductivity in different printed materials, pipes of a greater conducting material can be embedded in a model of a lesser conducting material.
+A microphone or piezo placed at the system-side of these sound-conducting pipes can determine where the model was tapped and how it was tapped.
+Active acoustic sensing as described in Touch \& Activate \cite{Ono-touchandactivate} is also possible; this technique would allow for particular areas of interest (connected to sound-conducting pipes) to be touch sensitive, while other areas (made of non-conducting material) are not sensed.
 
-\tovi{I would save these details for later, 
-when we tlak about our actual 
-implementations.}
+Sensing via traditional electronic components (e.g., potentiometers, alcohol gas sensors) can be accomplished by conductive inserted media.
+Components can be recessed into a print's surface, with their leads implanted into open pipes.
+By inserting liquid copper paint instead of threading traditional wires, all components in a model can share a ground line, and the paint's drying process obviates the use of solder or glue to affix the components in place (see Figure \ref{fig:radio}).
 
-We have also experimented with intentional encoding.
-Semi-closed pipes can function as resonance chambers: the first resonant frequency $F_1$ Hz of a semi-closed pipe length $L$ meters can be found by $F_1 = \frac{c}{4L}$ (where $c$ is the speed of sound).
-All odd harmonics (i.e., $3*F_1$, $5*F_1$, $7*F_1$, ...) of this frequency are also resonant frequencies of the pipe.
-\valkyrie{need to provide some guidelines for the number of IDs we can create in this way, and where in the sound spectrum they are.
-my sense is that this will depend highly upon the i/o sensitivities of the mic and speaker selected.}
+\subsection{Fabrication Techniques}
 
-An object's resonant frequency can be measured by attaching a speaker and microphone to it, sweeping frequencies with the speaker, and performing a Fourier transform on the resultant signal from the microphone.
-Peaks in the transformed data correspond to stronger returned impulses: the resonant frequencies of the object.
+The physical fabrication process used by different machines gives rise to differing design guidelines.
+The two machines we used to create our example objects were the Objet Connex 260 and the Makerbot Replicator 2.
 
-This is similar to work done by Willis, et al., in \cite{Willis-infrastructs}.
-While their technique requires access a terahertz imaging tool, audio resonance identity encoding requires only a microphone and a speaker.
+The Objet is an inkjet-based 3D printer which jets drops of polymer in very fine layers, curing each layer with an ultraviolet lamp.
+It offers a wide range of material types, including clear/opaque, flexible/hard, and black/white.
+Additionally, it permits the mixing of two different materials in a single print: a mixed material could include 70\% flexible and 30\% hard, for example, to be firm but somewhat flexible.
+The flexible materials can be inflated when they are thin: they elongate 200\% before breaking.
+We have found $.05$inches to be a good thickness for inflatable materials.
+Although this machine has many extraordinary capabilities, it unfortunately cannot print overhangs of greater than $14^{\circ}$ without support material.
+The support structure used by this machine is a loose matrix of material that can be removed by hand from the surface or blasted out with a water jet from the interior.
+The support is also soluble in lye.
 
-\valkyrie{in my experiments so far, it seems that the material printed does conduct sound fairly well, so this is feasible.
-I think we should get some equipment and do a small test, though, since I don't know how the resonance will change with the big hunk of plastic around the semi-open pipe.
-hopefully discussing this with Alex soon.}
+The Makerbot Replicator 2 is a fused-deposition modeling (FDM) printer, in which we use exclusively Polylactic Acid (PLA) bioplastics.
+More material properties are being explored in PLA: there are now transparent and flexible materials.
+They cannot be blended like those in the Objet.
+The Makerbot has a fairly coarse resolution compared to the Objet, however it permits overhangs up to $45^{\circ}$ without support and also allows bridging (short spans of material that bridge the gap between two structures with no support underneath).
+Due to the FDM process it can even recover from greater overhangs and longer bridges, as the material builds up and can reorganize into a structure \valkyrie{how do I say this?
+do I need a photo?
+print recovery is pretty awesome-looking.}.
+The support used by the Makerbot is made of the same material as the model itself, just printed in a different pattern and attached over as small a surface area as possible to ease removal.
+Support inside of a pipe may require pliers, or be infeasible to remove dependent upon geometry.
+
+Assembly is easiest on both machines when the model does not require support material.
+Models can be ``cut up'' into assembleable pieces to evade support requirements or ease material removal.
+We believe that an automatic tool for this task would be highly beneficial, however it is outside the scope of this work.

### 2014-03-24 paper/limitations.tex
addressing some comments, but mostly just copying the tovi comments in
addressing comments

@@
 
-The physical fabrication process also leads to some considerations.
-To print without support on a hobbyist machine, internal tubes cannot have overhangs of greater than $45^{\circ}$.
-On the Objet, support-free prints cannot overhang more than $14^{\circ}$.
-Removal of support from printed tubes can be troublesome, especially when their internal geometry and branching structures are complex.
+\valkyrie{Our design tool is not yet fully capable of realizing all possibilities in the design space of tubes.
+It lacks the network manipulation to make mixing or splitting tubes, unless they are designed in an input vector graphics file.
+Future work should also include assisting makers in the fabrication process itself by automatically cutting objects to reduce support material removal issues.
+Additionally, as mentioned there are opportunities to improve the algorithm for neon sign routing, as ours does not guarantee a shortest-path routing.
+--------- this paragraph breaks the latex for some reason?}
 
-Our design tool is not yet fully capable of realizing all possibilities in the design space of tubes.
-It lacks X and Y \valkyrie{probably it will lack the network manipulation to make mixing or splitting tubes...?}.
-Additionally, as mentioned there are opportunities to improve the algorithm for neon sign routing, as ours does not guarantee a shortest-path routing.
-We hope to address this in future work.
-
-\valkyrie{what other future work?
-miniaturization doesn't seem right, it's possible to use on mobile phones already with their on-board sensors (mic), maybe talking about shape-changing tubes?
-post-fabrication modification with drills?
-that doesn't sound very convincing.}
+We are also eager to test our tool with users and get feedback from makers who use it in the wild.

### 2014-03-24 paper/relatedwork.tex
addressing comments

@@
 
-While using a series of tubes can replicate a variety of interaction techniques, our work's \emph{contribution} is more generally related to two existing research areas: fabrication and routing.
+While using a series of tubes can replicate a variety of interaction techniques (discussed later), our work's \emph{contribution} is more generally related to two existing research areas: fabrication and routing.
 
@@
 Prior work on modeling the interior of objects (as opposed to their exterior) has been limited.
-\valkyrie{ryan, can you give citations here?}  Sauron automatically modified the interior of existing mechanical objects to make them amenable to vision-based sensing \cite{Savage-sauron}, while Willis, et al.,'s Infrastructs project created encoded interior cavities in 3D printed objects designed to be sensed with a terahertz imaging tool.
+\valkyrie{ryan, can you give citations here?
+I know there is that one about balancing, but Tovi mentioned a ``fruit slicing" one?}  Sauron automatically modified the interior of existing mechanical objects to make them amenable to vision-based sensing \cite{Savage-sauron}, while Willis, et al.,'s Infrastructs project created encoded interior cavities in 3D printed objects designed to be sensed with a terahertz imaging tool.
 Our work differs in that we want to allow users to design pipes and integrate interactive objects with them.

### 2014-03-24 paper/tubes.tex
addressing comments

--- 2014-03-20 paper/tubes.tex
@@
 
-\subsubsection{Tubes with Multiple Endpoints}
-For the creation of tubes with multiple endpoints, for example the star topologies in Figure \ref{fig:toys} or tree topologies, makers can select interior faces of existing pipes as endpoints, and thereby create pipe networks by hand.
+\subsubsection{Pipes with Multiple Endpoints}
+For the creation of pipes with multiple endpoints, for example the star topologies in Figure \ref{fig:toys} or tree topologies, makers can select interior faces of existing pipes as endpoints, and thereby create pipe networks by hand.
 We do not currently offer any pipe network manipulation tools.
 
-\valkyrie{To allow designers to create tubes with multiple endpoints (like the star topologies in Figure \ref{fig:toys}), we create 0-size mesh points in the center of the object, to which we attach one end of every tube?}
+\valkyrie{To allow designers to create pipes with multiple endpoints (like the star topologies in Figure \ref{fig:toys}), we create 0-size mesh points in the center of the object, to which we attach one end of every pipe?}
 
-\subsubsection{Tubes with Non-Round Endpoints}
-Our tool does not explicitly allow makers to design the shape of their tubes' endpoints.
+\subsubsection{Pipes with Non-Round Endpoints}
+Our tool does not explicitly allow makers to design the shape of their pipes' endpoints.
 However, the larger Meshmixer package does allow brushing arbitrary shapes onto the surface and extruding them inward, as in Figure \ref{fig:meshmixer-endpoint}.
-\valkyrie{should we have a figure about this?
-I  don't know how to describe it really, but we also didn't build it as a part of our tool.}
+
+\subsubsection{Semi-Closed and Fully Enclosed Pipes}
+Users can select via a checkbox if they wish to cap one end, the other end, or both ends of their pipe.
+The software simply creates an offset from the surface in these cases, so that material remains between the pipe and the surface after mesh modification.
 
@@
 In the final artifact, all created edges will be blocked out by dark material so the inserted medium is not visible (see Figure \ref{fig:tool-process-interior}).
+We offer a short algorithm here, with a more mathematically precise definition and associated proof in the appendix.
 
-We want a semi-Eulerian graph (i.e., we want a graph in which every vertex but two have even degree) so that the desired medium can be inserted, traverse every edge, and exit the graph at a different vertex.
-Let $G=(V,E_0)$ s.t.
-$\forall e \in E_0, weight(e)=0, start \in V $ the start point $end \in V$ the end point.
-Let $e_{temp} = (start, end)$, $E' = E + e_{temp}$.
+We need a graph which is semi-Eulerian (i.e., the inserted media can enter at one point, traverse every edge exactly once, and exit at a different point).
+In order to create a semi-Eulerian graph, we first add a temporary edge to the start and end points.
+This is removed after the graph is made fully Eulerian.
 
-We need to connect disconnected subgraphs in $G$.
-Let $G_{dis} = \{G_1, G_2, ...
-G_n\} \in G$ s.t.
-$\cup{\{G_i \in G_{dis}\}} = G$ and $G_i \cap G_j  = \{ \} \forall G_i, G_j \in G_{dis}, i\neq j$ be disconnected subgraphs in G.
-Create $E_{conn} = \{e = (u, v), u \in G_i, degree(u)\%2 = 1, v \in G_j, degree(v) \%2 = 1 i \neq j, weight(e) = distance(u, v)\}$ be the set of all edges connecting odd-degree points in two distinct disconnected subgraphs.
-To keep only the most desirable edges in $E_{conn}$, contract each subgraph $G_i$ to a single point $u_i$ with the shortest outgoing edges to each other subgraph: $E_{conn} = \cup_{G_i \in G_{dis}} \{e=(u_i, v), v \in G_j,$ s.t.
-$\forall E_{conn} \ni e_k = (u, v), u \in G_i, v \in G_j, weight(e_k) \geq weight(e)\}$.
-Sort $E_{conn}$ s.t.
-$E_{conn} = \{e_1, e_2, ...
-e_n\}, weight(e_1) \leq weight(e_2) \leq \ldots \leq weight(e_n)$.
+In order to connect disconnected subgraphs in the input, we reduce each disconnected subgraph to a point.
+We add edges from each subgraph-point to each other subgraph-point that represent the minimum distance between any two points in the expanded subgraphs (where distance is Euclidean distance).
+We greedily select the smallest weight edges until all subgraphs are joined into a single graph.
+We re-expand the subgraphs.
 
-Let $E_{conn-min} = \{ \}$.
-Beginning with $e_1$, add the first edge $e_i \in E_{conn}$ to $E_{conn-min}$.
-Then let $E_{dupe} = \{ e = (u, v) \in E_{conn}$ s.t.
-$u$  is reachable from $v$ along edges $E' \cup E_{conn-min}$.
-Let $E_{conn} = E_{conn} \setminus E_{dupe}$.
-Repeat until $E_{conn} = \{ \}$.
-Let $E' = E' \cup E_{conn-min}$.
-
-At this stage, we want an Eulerian graph: i.e., we need every vertex to be of even degree.
-Let $V_{odd} = \{ v \in V$ s.t.
-$degree(v) \% 2 = 1 \}$.
-Create $E_{circuit} = V_{odd} \times V_{odd}$, s.t.
-$\forall e=(u,v) \in E_{circuit}, weight(e) = distance(u, v)$.
-Find minimum matching $E_{circuit-min} \subseteq E_{circuit}$.
-Let $E' = E \cup E_{circuit-min}$.
-
-Now to remove our temporary edge and make the graph semi-Eulerian instead of Eulerian, let $E' = E' \setminus e_{temp}$.
-{
-\renewcommand{\thetheorem}{\ref{thm:graphness}}
-\begin{theorem}
- $G' = (V, E')$ is a connected, semi-Eulerian graph for which $E \subseteq E'$.
-\end{theorem}
-\addtocounter{theorem}{-1}
-}
-\begin{proof}
-See appendix.
-\end{proof}
+In order to make our graph Eulerian, we consider all vertices of odd degree in the connected graph and make a clique of edges once more based on distance.
+We again greedily select edges between the odd nodes until no odd nodes remain.
+Finally, we remove our temporary edge, which changes the graph from Eulerian to semi-Eulerian.
+This graph is a connected, semi-Eulerian graph which contains all edges in the input SVG (see Appendix).
 
@@
 How we actually cut stuff and make tubes happen. \valkyrie{Ryan?}
-
-\subsection{Fabrication Techniques}
-
-\valkyrie{this section may not be at all important.
-I was thinking to use it to talk about different ways you can actually build the tubed things, and also constraints set down by the Makerbot vs.
-the Objet, but maybe that discussion is best made really short and moved to elsewhere}
-
-\subsubsection{Printing} - different strategies with Objet (all print-in-place) and Makerbot (may need to add things like balloons afterwards).
-Ryan just got flexible material, we should see how stretchy it is...!
-We could also consider assembleable things that are easier to create using parts that clip together...
-probably out of scope.
-
-\subsubsection{Hand Tools} - post-fabrication modification is possible using hand tools.
-We can mark the surface to show where conduits are and how deep.
-I can also use this to test things before spending time printing them using a big ol' block o' plastic.

### 2014-03-24 paper/abstract.tex

@@
 These can be used to sense input and generate haptic as well as visible output.
-We describe the design space of pipes and hollow chambers for interaction design: there are a variety of types, topologies, and inserted media for pipes that can be leveraged to create diverse inputs and outputs.
+We describe the design space of pipes and hollow chambers for interaction design: a variety of openings, topologies, and inserted media for pipes create diverse inputs and outputs.
 We present a technique and design tool for routing pipes of various topologies through the interior of 3D printed parts.
 Our design tool is integrated into a 3D model manipulation program.
-There are two distinct routing algorithms.
+We use two distinct routing algorithms.
 One, for creating particular points of interaction or integrating with existing electronics, allows users to select begin and end points for pipes, then uses A* path routing and physics-based simulation to minimize the bending energy of routed paths.

### 2014-03-24 paper/conclusion.tex

--- 2014-02-10 paper/conclusion.tex
@@
 \section{Conclusion}
-In conclusion, tubes are cool.
-You can do lots of things with them.
-There are different kinds and topologies of tubes.
-There are different things you can put in them.
-This is cool for makers and also people with really expensive machines.
+While today's 3D printers are not yet able to fabricate active components in-place, we suggest that 3D printed interactive devices can be created with redirection of active input and output via tubes.
+In addition to describing the design space of tubes and connecting related works created through both traditional processes and rapid prototyping methods, we also presented a design tool for the creation of tubes inside arbitrary 3D models.
+We fabricated several example objects using this tool to explore new points in the design space.
+
+We are heartened by our success: it opens new interactive device prototyping opportunities both for makers with consumer-grade 3D printers and research and industrial labs with higher-end machines.
+We are excited to release our tool into the wild and see what people will build.

### 2014-03-24 paper/designspace.tex

--- 2014-03-20 paper/designspace.tex
@@
 \section{The Design Space of Pipes}
-We propose the use of pipes to redirect active components through the core and onto the surface of 3D printed interactive objects, as well as to allow for the creation of designed paths.
+The design space of pipes is large.
 By using different types and topologies of pipes and inserting different media into the completed pipes, a variety of inputs and outputs can be located at arbitrary points on the surface of a 3D print.
-We describe the design space of pipes, and then follow up with some of the input and output possibilities---both those that have been previously explored and those that are possible but remain unexplored.
+We describe the space below.
 
-\begin{figure}[h]
+\begin{figure}[h!]
 \centering
@@
 
-\subsection{Types of Pipes}\tovi{Maybe rename types to 
-"openings"?}
+\subsection{Openings of Pipes}
 
@@
 Definitionally, in the following paragraphs we refer to a pipe which is ``open on the user side'' to have its contents physically accessible to an interacting user at interaction time.
-This includes pipes which function as fountains, but it does not include pipes whose ends have been capped to keep water inside (whether this capping was done at print time, or later as a manual modification).
+This includes, for example, pipes which function as fountains, but it does not include pipes whose ends have been capped to keep water inside (whether this capping was done at print time, or later as a manual modification).
 Pipes whose contents are not physically accessible to the user are ``closed on the user side''.
-A pipe which is ``open on the system side'' can have its contents manipulated by a machine, e.g., by resonating, pressurizing, or passing electricity through those contents.
-The system does not need to have physical access to the contents in order to manipulate them.
+A pipe which is ``open on the system side'' is physically open to the machine, and one which is ``closed on the system side'' has its contents physically cut off from the machine.
+User and system manipulation of contents are possible with or without physical access, e.g., by resonating, pressurizing, or passing electricity through those contents.
 
@@
 
-A return pipe originates at the system and returns back to it.
+A return pipe originates at the system and returns back to it, without offering physical access to the user.
 By threading an electroluminescent (EL) wire through a clear return pipe, a maker can create a custom piece of neon art.
@@
 Semi-closed pipes are open at the system end (for control of the enclosed medium) and closed at the user end.
-The closed interface is interesting when it fabricated from a mobile material: for example, a series of pipes terminated in thin rubber membranes on the user side can be actuated by an air pump to create haptic feedback, or a hinged cap which can be blown open by air pressure.
-Without a printer capable of fabricating flexible material, a maker could affix a balloon to an open pipe's end to behave similarly; another possibility is to use semi-closed pipes as audio-generating resonance chambers.
+The closed interface  may be fabricated from a mobile material: for example, a series of pipes terminated in thin rubber membranes on the user side can be actuated by an air pump to create haptic feedback, or a hinged cap can be blown open by air pressure.
+Semi-closed pipes can also function as audio-generating resonance chambers.
 
@@
 Fully enclosed pipes can be used as resonance chambers (e.g., for object identification), or as air bubbles (e.g., as used in \cite{Willis-printedoptics} for internal display).
-\tovi{used as a container for water
-or other particles that would 
-otherwise fall out?}  \valkyrie{I re-worded above.
-please check.}  Their physical design space is very limited, as any support material required to create their internal geometry cannot be accessed for removal.
+It can also be used as a container for water or other particles that would otherwise fall out.
 
@@
 
-\tovi{Again, this raises the question of how we are applying the design 
-space - is it based on its properties when it is printed, or based on 
-any post-print modifications}\valkyrie{re-worded above.
-I think we are basing it on status at ``interaction time"}Pipes need not have hollow centers: in the case where routed pipes are filled with solid material---in particular, a solid material different from the model material---, interactions such as those in \cite{Willis-printedoptics} are possible.
+Pipes need not have hollow centers: in the case where routed pipes are filled with solid material---in particular, a solid material different from the model material---, interactions such as those in \cite{Willis-printedoptics} are possible.
 
@@
 If a maker wished to create a painting device, she might wish to have two system-side pipes feeding in primary-colored red and blue paints which mix in varying ratios, allowing their pigments to combine before purple exits from the device (see Figure \ref{fig:pipespace}).
-Splitting can also be useful, for example if our maker wants red paint output in two locations from her painting device, she could have one system-side pipe, but split the pipe into two (see Figure \ref{fig:pipespace}).\tovi{How about multiple in/
-multiple out (in contrast to 
-branching/splitting which is 
-single in multiple out)
-} \tovi{This seciton should refer to the 
-possibilities, not to what we 
-actually did.}
+Splitting can also be useful, for example if our maker wants red paint output in two locations from her painting device, she could have one system-side pipe, but split the pipe into two (see Figure \ref{fig:pipespace}).
 
-Star and tree topologies are extensions of the splitting and mixing primitives.
+Star and tree topologies are extensions of the splitting and mixing primitives, encompassing multiple-in and multiple-out in contrast to single-in/single-out.
 Using a star topology in which the pipes were filled with conductive paint, we created toys with several touch-sensitive areas, see Figure \ref{fig:toys}.
@@
 These two features lead to different kinds of interfaces.
-An example interface that focuses on the exterior connection points is the touch sensitive toy in Figure \ref{fig:toys}, where pipes must exit the toy at the eyes, ears, tail, etc.
+An example interface that focuses on the exterior connection points is the touch sensitive toy in Figure \ref{fig:toys}, where pipes must exit the toy at the frontal lobe, parietal lobes, etc.
 An interface focused on the internal path of the pipe is the neon sign in Figure \ref{fig:neon}: output is based on the shape of the pipe.
-
-\subsubsection{Dimensions of Pipes}
-
-\valkyrie{Reference those pages from Bjoern: http://ttb.com/recent-news/diagrams/ We can talk about wall thickness, tangent length, degree of bend, arg bend angle, inner and outer diameters, centerline radius, etc.}
-
-\subsection{Inputs and Outputs}
-
-By crossing pipe types, topologies, and inserted media, we create a space of possible inputs and outputs that can be pipe-mediated.
-We offer a table (see Figure \ref{fig:designspace}) containing these possibilities, referencing previous work on input/output techniques where appropriate, noting that many of these techniques have not been attempted using digital fabrication.
-We suggest new points in the design space that have been unexplored or not yet explored with digital fabrication.
-
-\begin{figure*}[t]
-\centering
-    \includegraphics[width=\textwidth]{figures/designspace.pdf}
-\caption{The design space of pipe-based interactions.
-Existing systems are written in regular font.
-Those created with fabrication are in {\color{blue}blue}.
-In \emph{{\color{red}red italic}} are unexplored interactions creatable with custom-fabricated pipes.}
-\label{fig:designspace}
-\end{figure*}
-
-\subsubsection{Inputs}
-
-Pipes create opportunities for many types of input sensing across the surface of printed devices.
-We describe sensing touch, pressure, grasp, flexing, tapping, and manipulation of traditional electronic components.
-We recognize that there are likely more input sensing techniques, both existing and on the horizon, that are compatible with the use of pipes.
-
-Touch sensing is enabled through pipes filled with conductive media.
-Traditional wires (threadable) may be used, but we have had success with conductive paint (liquid).
-Using conductive paint and Swept Frequency Capacitive Sensing (SFCS \cite{Sato-touche}), an interior star pipe topology can enable single-wire touch- and grasp-sensing at any set of points on a printed object's surface (see Figure \cite{fig:toy}).
-Simple capacitance measurements are possible using the same media, and custom-designed sensors like those created in Savage, et al.,'s Midas system are also possible \cite{Savage-midas}.
-
-Pressure can be sensed in multiple ways.
-Slyper, et al., in \cite{Slyper-pressure} contribute a set of semi-closed 3D printable primitives that respond to particular actions, like twisting or pushing, with changes in internal air pressure.
-These primitives can be attached as the terminus of any semi-closed pipe; the user can manipulate the printed endpoint and the system can sense air pressure changes.
-Pressure can also be more simply sensed via capacitance using those techniques.
-
-Grasp sensing can also be enabled using Wimmer's FlyEye technique \cite{Wimmer-flyeye}.
-This technique involves a pair of optical links at each point of interest: one of each pair is connected to an infrared (IR) LED and the other to a camera.
-When some object nears or touches a pair of cables, IR light from the LED is reflected back to the camera and the touched point appears as a bright spot.
-Follmer, et al.,'s Jamming User Interfaces \cite{Follmer-jamming} rely on dense particulates enclosed in a flexible volume, and can be sensed optically.
-These dense particulates can be printed in-place with desired characteristics (such as varying particle size or shape) using sacrificial support material, or added post-print.
-The optical links necessary for sensing with FlyEye or Jamming can be created either via clear solid cores as in \cite{Willis-printedoptics} or via fiber optic cables threaded through hollow pipes (see Figure \ref{fig:pens}).
-
-Some 3D printers, like our Objet Connex 260, can fabricate rubber-like materials.
-Much like Slyper, et al., in \cite{Slyper-shape}, we can sense flexing and bending of prints made on these machines.
-While Slyper, et al., created multi-level multi-stage silicone molds with embedded wires to build their sensors, by placing pipes in our rubberlike prints and inserting conductive media afterwards, we can perform the same flex sensing.
-
-Tapping is another input possibility.
-Due to differential sound conductivity in different printed materials (in particular, the rubberlike Objet material compared to the hard Objet material \valkyrie{presumably we should get or take some measurements of this and include it in the appendix?}), pipes of a greater conducting material can be embedded in a model of a lesser conducting material.
-A microphone or piezo placed at the system-side of these sound-conducting pipes can determine where the model was tapped and how it was tapped.
-Active acoustic sensing as described in Touch \& Activate \cite{Ono-touchandactivate} is also possible; this technique would allow for particular areas of interest (connected to sound-conducting pipes) to be touch sensitive, while other areas (made of non-conducting material) are not sensed.
-
-Sensing via traditional electronic components (e.g., potentiometers, alcohol gas sensors) can be accomplished by conductive inserted media.
-Components can be recessed into a print's surface, with their leads implanted into open pipes.
-By inserting liquid copper paint instead of threading traditional wires, all components in a model can share a ground line, and the paint's drying process obviates the use of solder or glue to affix the components in place (see Figure \ref{fig:radio}).
-
-\subsubsection{Outputs}
-
-While many input techniques require the use of conductive media inserted into pipes, output media are more varied.
-We describe potential pipe-based outputs organized by the five senses: visual, aural, haptic, and olfactory/gustatory.
-In addition to outputs we mention explicitly, any outputs possible using traditional electronic components are possible as described above.
-
-\begin{figure}[h]
-\centering
-    \includegraphics[width=3.4in]{figures/placeholder/direct.jpg}
-\caption{By attaching a servo to a gate structure, it is possible to direct fluid through that pipe.
-The gate not only has binary positions (a) and (b), but can direct fractional flow, as well (c).
-This could be used to create logic gates, as in (d).
-These could be used, for example, in science exhibits.}
-\label{fig:direct}
-\end{figure}
-
-\emph{Visual} outputs can be mediated by gases, liquids, solid, particulates, or threadables.
-EL wire can be threaded through custom paths to create neon signs, as in Figure \ref{fig:neon}.
-Colored liquids can be mixed and split before exiting an opaque device, or in a transparent device their mixing, splitting, and paths could be used for display.
-PneUIs \cite{Yao-pneui} can be fabricated as single parts: the rubberlike material offered by our Objet is inflatable when thin (200\% elongation at break), and a PneUI located at the terminus of a pipe could be actuated by a system-side air pump (see Figure \ref{fig:breathe}).
-Mechanical motion can be mediated through fluids (a light object which is partially contained in a pipe or located at its end could be pushed by the fluid pressure of the pipe) or threadables (e.g., muscle wire).
-Large particles trapped in pipes visible to the user can be used for, e.g., maze games (see Figure \ref{fig:maze}).
-We can even direct fluids through a pipe network using a servo-controlled gate, as in Figure \ref{fig:direct}.
-Using techniques like those described by Harrison, et al., in \cite{Harrison-buttons}  \valkyrie{this paragraph seems terrible.
-I think all these paragraphs might seem terrible...}
-
-\emph{Aural} outputs can be created with gases.
-Zoran has successfully 3D printed a functional flute, which in essence is a long open open pipe with semi-closed pipes coverable by keys \cite{Zoran-flute}.
-Helmholz resonance (the phenomenon that creates noise when you blow into the top of a glass bottle, which is also how some musical instruments, like ocarinas, work) can be leveraged to create computer-controllable sound chambers.
-All that is necessary for this is an enclosed chamber with a pipe connected to it via a narrow neck-point (see Figure \ref{fig:ocarina}).
-Passive sound amplification (like a phonograph horn) and sound redirection are also possible using pipes.
-
-\begin{figure}[h]
-\centering
-    \includegraphics[width=3.4in]{figures/placeholder/helmholz.jpg}
-\caption{This box (a) contains an enclosed Helmholz resonator, seen in the cutaway view (b).
-This resonator creates a sound when air is blown through the pipe at the top, because air compresses at the small neck joint between the chamber and the pipe.
-\valkyrie{need to get a test print of this that actually works}}
-\label{fig:ocarina}
-\end{figure}
-
-\emph{Haptic} outputs are possible through compressible and incompressible fluids.
-Fluids can actuate particulates and create programmable pliability (as discussed above and in \cite{Follmer-jamming}).
-Additionally, use of semi-closed pipes with rubberlike material as the caps allows for gases to actuate surface features (see Figure \ref{fig:breathe}).
-Sodhi, et al., in \cite{Sodhi-aireal} created free-air haptic feedback using controlled vortex generation: this technique could also be reproduced using a series of pipes.
-Iwata, et al., explored volumetric fluid-based interactions with programmatically inflatable balloons and an overhead projector in \cite{Iwata-volflex}, and additionally haptic outputs can be used to increase ergonomics, as in the inflatable mouse by Kim, et al.
-\cite{Kim-inflatablemouse}.
-
-\emph{Olfactory} and \emph{gustatory} outputs can be created through the mixing and splitting of pipes carrying scented or flavored fluids, respectively.
-
-\subsection{Identity}
-
-Pipes included in printed objects allow for their identification.
-While the pipes need not be visible (especially in the case of fully-enclosed pipes), their presence, location, network structure, and length change many properties of a printed object.
-This includes weight, acoustic resonance, and (if conductive material is present) capacitive signature.
-
-\valkyrie{need to do a rewrite of this that doesn't necessarily focus so deeply on audio only.
-we can talk about these other non-visible measurements like weight and capacitive signature.
-I also need to actually get an intentional encoding test up and running...
-we are going to have to be clever about how we stimulate the resonance pipes, though.
-using a speaker won't really work: we have to agitate the air in the chambers, so we might need to use something like a reed (as in reed instruments).}
-
-Both identification by recall and intentional encoding are possible.
-As seen in \cite{Ono-touchandactivate}, different objects have different acoustic signatures.
-Additionally, two objects that are visually identical but which have fully enclosed (or other types) of pipes on their interior can be distinguished acoustically.
-Thus we can recall an object's identity once its acoustic signature has been recorded.
-
-We have also experimented with intentional encoding.
-Semi-closed pipes can function as resonance chambers: the first resonant frequency $F_1$ Hz of a semi-closed pipe length $L$ meters can be found by $F_1 = \frac{c}{4L}$ (where $c$ is the speed of sound).
-All odd harmonics (i.e., $3*F_1$, $5*F_1$, $7*F_1$, ...) of this frequency are also resonant frequencies of the pipe.
-\valkyrie{need to provide some guidelines for the number of IDs we can create in this way, and where in the sound spectrum they are.
-my sense is that this will depend highly upon the i/o sensitivities of the mic and speaker selected.}
-
-An object's resonant frequency can be measured by attaching a speaker and microphone to it, sweeping frequencies with the speaker, and performing a Fourier transform on the resultant signal from the microphone.
-Peaks in the transformed data correspond to stronger returned impulses: the resonant frequencies of the object.
-
-This is similar to work done by Willis, et al., in \cite{Willis-infrastructs}.
-While their technique requires access a terahertz imaging tool, audio resonance identity encoding requires only a microphone and a speaker.
-
-\valkyrie{in my experiments so far, it seems that the material printed does conduct sound fairly well, so this is feasible.
-I think we should get some equipment and do a small test, though, since I don't know how the resonance will change with the big hunk of plastic around the semi-open pipe.
-hopefully discussing this with Alex soon.}

### 2014-03-24 paper/discussion.tex

\section{Discussion}

Pipes included in objects can have additional effects beyond their designed intentions.

Pipes included in printed objects allow for their identification.  While the pipes need not be visible (especially in the case of fully-enclosed pipes), their presence, location, network structure, and length change the physical properties of a printed object.  This includes weight, acoustic resonance, and (if conductive material is present) capacitive signature.

Both identification by recall and intentional encoding are possible.  As seen in \cite{Ono-touchandactivate}, different objects have different acoustic signatures.  Additionally, two objects that are visually identical but which have fully enclosed (or other types) of pipes on their interior can be distinguished acoustically.  Thus we can recall an object's identity once its acoustic signature has been recorded.  Similarly, the Touch\'{e} system in \cite{Sato-touche} relies on distinguishing capacitive signatures of objects, both in static states and as their boundary conditions are changed by human interaction.  Weight is a simple metric of object identity, but weights are less distinctive than acoustic or capacitive signatures, as they are represented by a single number.

We have also experimented with intentional encoding.  Semi-closed pipes can function as resonance chambers.  An object's resonant frequency can be measured by attaching a speaker and microphone to it, sweeping frequencies with the speaker, and performing a Fourier transform on the resultant signal from the microphone.  Peaks in the transformed data correspond to stronger returned impulses: the resonant frequencies of the object.  Intentional encoding in the case of capacitance is significantly more difficult, and we leave it to future work.  Encoding of weight is trivial by creating fully-enclosed chambers inside an object.

This is similar to work done by Willis, et al., in \cite{Willis-infrastructs}.  While their technique requires access a terahertz imaging tool, audio resonance identity encoding requires only a microphone and a speaker.

\valkyrie{To discuss: printing media into tubes?  Design of interiors of objects more generally?  Need to cite that paper about balancing printed objects from SIGGRAPH.}

### 2014-03-25 paper/abstract.tex
start of video draft, figure images, pen holder files

@@
 We use two distinct routing algorithms.
-One, for creating particular points of interaction or integrating with existing electronics, allows users to select begin and end points for pipes, then uses A* path routing and physics-based simulation to minimize the bending energy of routed paths.
-The second, for creating, e.g., novel neon signs, allows users to enter a description of paths to follow: for this we developed a novel routing algorithm and offer proof of key traits.
+One, for creating particular points of interaction or integrating with existing electronics on the model's surface, allows users to select begin and end points for pipes, then uses A* path routing and physics-based simulation to minimize the bending energy of routed paths.
+The second, for creating, e.g., novel neon signs, has users supply a desired internal shape which we fit a pipe route to: for this we developed a novel routing algorithm and offer proof of key traits.
 We present several totally tubular prototypes we created using this tool to show its flexibility and potential, as well as to explore new points in the pipe design space.

### 2014-03-25 paper/designspace.tex
start of video draft, figure images, pen holder files

@@
 Each of these types offers distinct interaction capabilities (see Figure \ref{fig:pipespace}).
+Since we are interested in interactive devices, we distinguish a ``user side" - part of the surface or interior of an object facing its user; and a ``system side" - part of the object (either interior or exterior) where other hardware such as electronics will be connected.
 Definitionally, in the following paragraphs we refer to a pipe which is ``open on the user side'' to have its contents physically accessible to an interacting user at interaction time.
@@
 User and system manipulation of contents are possible with or without physical access, e.g., by resonating, pressurizing, or passing electricity through those contents.
+\bjoern{wordy paragraph}
 
@@
 Semi-closed pipes are open at the system end (for control of the enclosed medium) and closed at the user end.
-The closed interface  may be fabricated from a mobile material: for example, a series of pipes terminated in thin rubber membranes on the user side can be actuated by an air pump to create haptic feedback, or a hinged cap can be blown open by air pressure.
+The closed interface  may be fabricated from a malleable material: for example, a series of pipes terminated in thin rubber membranes on the user side can be actuated by an air pump to create haptic feedback, or a hinged cap can be blown open by air pressure.
 Semi-closed pipes can also function as audio-generating resonance chambers.
@@
 It can also be used as a container for water or other particles that would otherwise fall out.
+
+\bjoern{This section in general may need a small multipart figure that shows example photos or at least illustrations of these points in the design space.
+(Fig 2 is the abstract representation of the dimensions; here we have some concrete examples of points in the space).
+
+Or maybe this comes later? it's a bit hard to follow these descriptions in the abstract.
+}
 
@@
 As in \cite{Slyper-pressure}, structures can be engineered to change in air pressure when manipulated correctly (e.g., a spiral that changes pressure when twisted, but not pressed), and thus fluid pressure can also be used as an input.
+\bjoern{Isn't this the same example as two paragraph above?}
 

### 2014-03-25 paper/introduction.tex
start of video draft, figure images, pen holder files

@@
 
-The use of pipes can allow for \emph{redirection} and \emph{centralization} of active components, which creates \emph{iterable} devices.
-For example, to create a touch-sensitive toy as in Figure \ref{fig:teaser}, a maker can select several locations of interest on the surface of the brain.
+The use of pipes can allow for \emph{redirection} and \emph{centralization} of active components to various locations on a printed object's surface.
+For example, to create a touch-sensitive toy as in Figure \ref{fig:teaser}, a maker can select several locations of interest on the brain.
 After creating pipes to those locations and printing her object, she can fill the pipes with conductive paint.
 For sensing, she need only connect a single wire to a shared out of all the tubes, and can distinguish them via Swept Frequency Capacitve Sensing \cite{Sato-touche}.
-If she decides that another area of interest is important to distinguish, she can simply print a new model with a new pipe leading to that location: the electronics assembly remains the same and she need only connect the same single wire to the tube out.
-\valkyrie{not clear that this is our contribution.}
+In this way, she has only a single active element (the microcontroller setup), but her otherwise-inert 3D printed object (the brain) has several active sensing locations.
+This can also be used for rapid iteration of designs.
+Our maker need only add a digital tube, reprint her brain, and re-fill it with copper paint to have a new version; no complicated electromechanical assembly required.
 
@@
 The generation of these pipes, however, is not trivial.
-Pipes need to be as straight as possible to ease the insertion of media post-print.
+Pipes need to have maximum possible bend radius to ease the insertion of media post-print.
 Independent pipes should not intersect, as this could lead to electrical shorting or other problems.
@@
 Consumer-grade 3D printers do not guarantee air- or water-tightness of their prints, so more than one layer may be required to prevent fluid leakage.
-We have developed techniques for minimizing the bending energy of, enforcing non-intersection with, and avoiding the surface for pipes.
+Pipes for applications like neon signs must have an Euler tour through them: this allows a single EL wire to pass through every edge on the pipe graph.
+We have developed techniques for minimizing the bending energy of, enforcing non-intersection with, and avoiding the surface for pipes, as well as expanding user-specified paths and generating Euler tours.
 These techniques are embodied in our software tool to enable makers to create pipe-powered interfaces.
@@
 \begin{itemize}
-\item We lay out a design space of pipes and their types, contained media, topologies, and design focus.
+\item We lay out a design space of pipes and their openings, contained media, topologies, and design focus.
+\item We offer algorithms and techniques for routing pipes, as well as a design tool implementing them.
 \item We describe the interaction possibilities of pipes, including prior work as well as uncharted possibilities.
-\item We offer algorithms and techniques for routing pipes, as well as a design tool implementing them.
 \item We showcase a set of examples, enabled by our modeling tool and exploring new points in the design space.

### 2014-03-25 paper/relatedwork.tex
start of video draft, figure images, pen holder files

--- 2014-03-20 paper/relatedwork.tex
@@
 
-While using a series of tubes can replicate a variety of interaction techniques, our work's \emph{contribution} is more generally related to two existing research areas: fabrication and routing.
+Our contributions relate to other efforts to add interactivity to digitally fabricated objects; and to techniques for routing and mode modification.
+\bjoern{The related work section is taking shape but still has a bit of a "missing the forest for the trees" feeling.)}
+
+\bjoern{Tactically, to communicate who would be is a qualified reviewer for this paper in the UIST community, we should make sure to have appropriate references to Baudisch, Poupyrev, Harrison in here}
 
@@
 Previous work has investigated the integration of interactivity with the fabrication process.
-\cite{Willis-printedoptics} used light pipes, solid clear tubes, to create integrated displays: our work allows a superset of these capabilities, and Printed Optics is a part of our design space.
+Printed Optics (\cite{Willis-printedoptics}) used light pipes, solid clear tubes, to create integrated displays: our work is inspired by this and goes towards a design tool for these capabilities \valkyrie{bad phrasing; don't de-fang the tool at the front of the paper!}; Printed Optics is a part of our design space, which includes other uses of pipes.
 Willis, et al., also did not offer a design tool or techniques for creating light pipes automatically.
-\cite{Savage-sauron} investigated the use of computer vision to track interactions with physical mechanisms.
-Using tubes for redirection, it is possible to replicate these mechanically sensitive techniques (see Figure \ref{fig:pens}), however our work does not focus on computer vision alone.
+Sauron (\cite{Savage-sauron}) investigated the use of computer vision to track interactions with physical mechanisms.
+While Savage, et al., used mirrors for redirection, we believe that the required line-of-sight operation for their computer vision techniques may be possible using printed optics or inserted optical fibers.
 Slyper, et al., \cite{Slyper-pressure} 3D printed flexible robot armatures capable of sensing interaction via changes in air pressure.
@@
 Sells, et al., in \cite{Sells-reprap} create channels on the surface of a 3D printed object which are filled with a special computer-controlled syringe of liquid solder to create connections.
-Their focus is on creating self-replicating 3D printers, and their electronics are combined to traditional 2D routed circuitboards.
+Their focus is on creating self-replicating 3D printers, and their electronics are confined to traditional 2D routed circuitboards.
 Sarik, et al., \cite{Sarik-tracebrush} \valkyrie{I never found a real paper of this.
@@
 
-General-purpose routing for connecting 2D capacitive sensors to terminals is discussed in \cite{Savage-midas}.
+General-purpose routing for connecting 2D capacitive sensors to terminals is discussed in Midas (\cite{Savage-midas}) \bjoern{Maybe move discussion of Midas to prior section since it really is about fabricating interactive devices.
+And then here just quickly review that lots of work has gone into routing algorithms for PCBs but that they assume multiple stacked planes connected with vertical vias, not general 3d routing.}.
 These techniques, based on circuitboard routing, are limited to planar and unfoldable shapes, and are intended for use on the surface of objects rather than through their cores.
@@
 Prior work on modeling the interior of objects (as opposed to their exterior) has been limited.
-\valkyrie{ryan, can you give citations here?}  Sauron automatically modified the interior of existing mechanical objects to make them amenable to vision-based sensing \cite{Savage-sauron}, while Willis, et al.,'s Infrastructs project created encoded interior cavities in 3D printed objects designed to be sensed with a terahertz imaging tool.
+\valkyrie{ryan, can you give citations here?
+I know there is that one about balancing, but Tovi mentioned a ``fruit slicing" one?}  In Make it Stand, material is removed from the interior of objects with the goal of changing their center of gravity \cite{Prevost-makeitstand}.
+Bickel, et al., explored enforcing deformation behavior with material blending \cite{Bickel-deformation}, while B\"{a}cher, et al., built an algorithm for automatically placing 3D fabricatable joints into meshes \cite{Bacher-articulated}.
+Sauron automatically modified the interior of existing mechanical objects to make them amenable to vision-based sensing \cite{Savage-sauron}, and Infrastructs created encoded interior cavities in 3D printed objects designed to be sensed with a terahertz imaging tool.
 Our work differs in that we want to allow users to design pipes and integrate interactive objects with them.

### 2014-03-26 paper/conclusion.tex
adding bjoern comments

@@
 \section{Conclusion}
+
+\bjoern{We should be more upfront in the paper that the particular sensing and actuation strategies we use are taken from prior work (for example wimmer's), but that there is still novelty in how we modify the models or redirect the input and output to different parts of the model with our pipe techniques.
+Otherwise I can see reviewers criticizing that we only show a collection of sensing techniques that have already been published.
+
+it's important to frame that our techniques allow us to amplify and extend the utility of existing sensing and actuation approaches}
+
 While today's 3D printers are not yet able to fabricate active components in-place, we suggest that 3D printed interactive devices can be created with redirection of active input and output via tubes.

### 2014-03-26 paper/designspace.tex
add new figure to paper to illustrate types of pipes
adding bjoern comments

@@
 \bjoern{wordy paragraph}
+
+\begin{figure}[h!]
+\centering
+    \includegraphics[width=3.4in]{figures/types.png}
+\caption{Types of pipes.
+A shows a pair of open tubes filled with copper paint to power an LED.
+B shows a return pipe with EL wire threaded through it.
+C shows two semi-closed pipes with rubber membranes at their ends.
+In D, a rabbit has a fully enclosed chamber to modify its weight.}
+\label{fig:pipespace}
+\end{figure}
 
@@
 
-Pipes need not have hollow centers: in the case where routed pipes are filled with solid material---in particular, a solid material different from the model material---, interactions such as those in \cite{Willis-printedoptics} are possible.
+Pipes need not have hollow centers: in the case where routed pipes are filled with solid material---in particular, a solid material different from the model material---, interactions such as those in Printed Optics (\cite{Willis-printedoptics}) are possible.
 
@@
 Sparse particles in a stream of fluid can provide haptic feedback.
-Dense particles in a semi-closed pipe allow for jamming-based interactions at any point on the surface of an object \cite{Follmer-jamming}.
+Dense particles in a semi-closed pipe allow for jamming-based interactions at any point on the surface of an object \cite{Follmer-jamming} \bjoern{may need more of an explanation.
+you need jamming plus flexible surface material, right?}.
 
@@
 
+\bjoern{Think of the ordering of dimensions is it a conscious choice to list this dimension last?}
 Pipe designs may emphasize either their exterior features (connection points) or their interior paths.

### 2014-03-26 paper/exampleobjects.tex
start of video draft, figure images, pen holder files

@@
 
-\valkyrie{Bj\"{o}rn, I can't figure out what happened to my figures?  and why they got thrown down into a weird state with my limitations section, which appears now in two places???}
-
 \subsection{Touch-sensitive Toys (open, liquid, star, exterior)}
@@
 Sensing is done on a single wire via SFCS.
-Inset shows the internal structure of the tubes generated by our design tool.}\input{limitations.tex}
-
+Inset shows the internal structure of the tubes generated by our design tool.}
 \label{fig:toys}
@@
 Our pen holder uses a modification of the FlyEye technique described by Wimmer in \cite{Wimmer-flyeye} and contains open tubes filled with fiber optic cables, one per pen chamber.
-We use a single cable per chamber; our 6mm diameter fiber optic cables, in comparison to the angel hair tubes used in the original work, can send and receive through the same cable.
+We use a single cable per chamber; our 6mm diameter fiber optic cables, in comparison to the angel hair tubes used in the original work \bjoern{Mixed terminology between el wire and fiber optics?
+"angel hair" is a specific product name, not a general technical term (I think)}, can send and receive through the same cable.
 At the base of each tube is a QRE1113 line sensor digital breakout board, which has an integrated IR emitter and receiver.
@@
 We created this prototype on our Makerbot in two halves to allow for insertion of the marble.
+\bjoern{You are giving two different reasons here for printing in pieces.}
 

### 2014-03-26 paper/tubes.tex

@@
 
+\bjoern{Lead with the general goals of tools to help users leverage pipes. E.g. List design goals or requirements. Them describe our tool and how it satisfies these requirements.}
 To allow makers to design novel objects with pipe-powered interfaces, we created a tool for use on arbitrary 3D models.
@@
 
-Our tool is implemented as a part of Meshmixer, a consumer 3D mesh editing tool, in C++.
+Our tool is implemented as a part of Meshmixer, a consumer 3D mesh editing tool, in C++.  \bjoern{add a reference to meshmixer}
 
@@
 
-To create pipes in which the location and shape of exterior connection points matters, we first allow makers to select the connection points on the mesh's surface using a brush.
+To create pipes in which the location and shape of exterior connection points matters, we first allow makers to select the connection points on the mesh's surface using a brush  \bjoern{Need to add detail that you voxelize the mesh and route on the voxel grid.
+Add justification why this is a good way to do this and why a* is the right algorithm.
+(I guess this is oping in the next paragraph)}.
 Our tool then creates an initial shortest-path routing using A* to estimate the routed distance between points.
@@
     \includegraphics[width=3.4in]{figures/placeholder/exterior.jpg}
-\caption{An example mesh with exterior connection points selected.  The initial A* routing is drawn in {\color{red}red}, and in {\color{tovi}green} is our physically-based rod after simulation.}
+\caption{An example mesh with exterior connection points selected.
+The initial A* routing is drawn in {\color{red}red}, and in {\color{tovi}green} is our physically-based rod after simulation.
+\bjoern{Tis example is strange.
+Why wouldn't be a good idea to route through the air outside the model during A*?
+How is A* in this case more useful than just drawing a straight line form start to end, regardless of the model geometry?}}
 \label{fig:tool-process-exterior}
@@
 
+\bjoern{Need to say something in this section about avoiding intersection with previously routed pipes - how does that work? Is it because pipe interiors are now "outside"? Say this explicitly.
+
+Also, mention somewhere that we don't yet solve the general routing problem with many routes - we only make greedy pair wise decisions.}
 Our basic first-pass routing algorithm uses the A* routing algorithm \cite{Hart-Astar}.
@@
 A too-long rod is forced to kink and expand further than necessary inside the model, while a too-short rod must be stretched, leading to non-smooth bending.
+\bjoern{Once you start describing ad hoc changes, I start to wonder if there are alternative approaches to this routing problem that would feel more principled.
+This suggests we need a stronger argument for rod bending.}
 
@@
 This wind keeps the resulting pipe away from the surface.
-\valkyrie{Ryan: How does it actually work?}
+\valkyrie{Ryan: How does it actually work?}  \bjoern{Unrelated?
+Note from our call today: argue somewhere in the paper that the geometries we create would be very tricky to impossible to create with other manufacturing technologies like injection molding.}
 
@@
 However, the larger Meshmixer package does allow brushing arbitrary shapes onto the surface and extruding them inward, as in Figure \ref{fig:meshmixer-endpoint}.
+\bjoern{Need to motivate the "inward" phrase here?
+I assume  to helps with brushing on of conductive paint, but do you need this extrusion in general?}
 
@@
 
+\bjoern{A general point - even though we focus on uses that require manual post print assembly, we may want to think about how useful our software is once fancier printers do come along that can do more.
+I think some of the contributions will still hold, which would be great.}
+
+\bjoern{Do we have a second use case for paths beyond neon signs? Would be nice.}
+
 Neon signs, as in Figure \ref{fig:tool-process-interior}(d), require a single long path that passes through many desired edges: for example, in our UIST sign, the path must pass through all the edges that comprise the word ``UIST''.
@@
 \caption{An input vector graphics file with the points which cannot be tubed as drawn highlighted in {\color{blue}blue} (a).
-The connected graph created by our software (b) and the resulting Euler circuit (c) permit creation of a novel neon sign (d).}
+The connected graph created by our software (b) and the resulting Euler circuit (c) permit creation of a novel neon sign (d).
+\bjoern{Shouldnt the right arrow on the top of the t point in the other direction?}}
 \label{fig:tool-process-interior}
@@
 \subsubsection{Routing}
+\bjoern{You may want to include a primer on graphs and Euler tours here to remind the reader...}
 The interior path routing problem is a version of the Chinese Postman Problem\footnote{The CPP is also known as the route inspection problem: \url{http://en.wikipedia.org/wiki/Route_inspection_problem}}, in which we wish to traverse all edges of the graph described in the user's input with a single Euler circuit, much as a postman needs to walk along every road at least once to deliver mail.
 We call our relaxation the Spiderman Postman Problem: we allow the creation of non-existing paths (i.e., the postman may traverse buildings in addition to roads).
-If inputted path components are disconnected, we must create edges that connect them; additionally we can create new edges connecting odd-degree vertices rather than simply retracing existing edges.
+If input path components are disconnected, we must create edges that connect them; additionally we can create new edges connecting odd-degree vertices rather than simply retracing existing edges.
 In the final artifact, all created edges will be blocked out by dark material so the inserted medium is not visible (see Figure \ref{fig:tool-process-interior}).
@@
 
-In order to connect disconnected subgraphs in the input, we reduce each disconnected subgraph to a point.
+In order to connect disconnected subgraphs in the input, we reduce each disconnected subgraph to a point \bjoern{Hw do you pick that point?
+Don't you reduce it to two points for inbound and outbound connections?
+I guess I don't follow this paragraph.}.
 We add edges from each subgraph-point to each other subgraph-point that represent the minimum distance between any two points in the expanded subgraphs (where distance is Euclidean distance).
@@
 
-Once we have a semi-Eulerian graph, we need to create an Euler tour.
+Once we have a semi-Eulerian graph, we need to create an Euler tour  \bjoern{Good explanation of terminology here, but you probably need this earlier.}.
 An Euler tour is a path that touches every edge once, beginning and ending at the two nodes of odd degree: in this case, those are the user-selected start and end nodes.
-We use a weighted modification of Fleury's algorithm\footnote{\url{http://en.wikipedia.org/wiki/Eulerian_path\#Fleury.27s_algorithm}} for this, where instead of randomly selecting from non-bridge paths at each node, we select the path which turns the least from the most recent path (i.e., we prefer to pass straight through a node, if possible).
+We use a weighted modification of Fleury's algorithm for finding Euler tours\footnote{\url{http://en.wikipedia.org/wiki/Eulerian_path\#Fleury.27s_algorithm}} for this,  \bjoern{This may bee too low level a detail if you don't explain how fleury works at a conceptual level.
+}where instead of randomly selecting from non-bridge paths at each node, we select the path which turns the least from the most recent path (i.e., we prefer to pass straight through a node, if possible).
 This minimizes turns in the final artifact, which eases support material removal and assembly.
@@
 
+\bjoern{Too specific. This section is not just about el wire.
+
+A most, say, this is required in the general case, but not for the specific applications we have tested so far, so we leave it to future work.
+
+Of course it'd be great to get it to run.}
+
 \valkyrie{Once we have a tour, we must resolve tubes that intersect in the plane.  Because we have a 3-dimensional canvas in which to work, we can push overlapping paths into the third dimension.

### 2014-03-26 paper/io.tex

@@
 Once objects have been designed and fabricated, the major flexibility for prototyping with pipes stems from the media inserted in them.
-The five types of media--gas, liquid, solid, particulate, and threadable--give rise to unique possibilities in output (visual, aural, tactile/haptic, olfactory/gustatory) as well as input (touch/pressure and others).
+The five types of media--gas, liquid, solid, particulate, and threadable--\bjoern{Not just the media in them, but also the systems attached to them to sense changes to chose media or actuate them.
+
+I wonder if you could still push this table yet further back, essentially into the discussion part.
+So you first show the objects, and then the discussion puts these into context of other work thats been done and provides pointers to future work in other areas of this design space.
+
+}give rise to unique possibilities in output (visual, aural, tactile/haptic, olfactory/gustatory) as well as input (touch/pressure and others).
 
@@
 By crossing pipe types, topologies, and inserted media, we create a space of possible inputs and outputs that can be pipe-mediated.
-We offer a table (see Figure \ref{fig:designspace}) containing these possibilities, referencing previous work on input/output techniques where appropriate, noting that many of these techniques have not been attempted using digital fabrication.
+We offer a table (see Figure \ref{fig:designspace}) containing these possibilities, referencing previous work on input/output techniques where appropriate, noting that many of these techniques have not been attempted using digital fabrication \bjoern{It may be a bit tricky to define what "using digital fab" means here.
+For example, Harrison 's pneumatically actuated buttons were done on a laser cutter, but also required manual gluing of other layers.
+Does that count or not?}.
 We suggest new points in the design space that have been unexplored or not yet explored with digital fabrication.
@@
 
+\bjoern{The biggest question wrt fabrication for me is - how do I print these cavities? So general support material avoidance or removal strategies should be more prominent.
+
+(I think the lye bath would probably work better than the high pressure water for our objet prints)}
 The physical fabrication process used by different machines gives rise to differing design guidelines.
@@
 Assembly is easiest on both machines when the model does not require support material.
-Models can be ``cut up'' into assembleable pieces to evade support requirements or ease material removal.
+Models can be ``cut up'' \bjoern{This negates some of the benefits compared to other manufacturing techniques.} into assembleable pieces to evade support requirements or ease material removal.
 We believe that an automatic tool for this task would be highly beneficial, however it is outside the scope of this work.

### 2014-03-26 paper/limitations.tex

@@
 
-\valkyrie{Our design tool is not yet fully capable of realizing all possibilities in the design space of tubes.
+Our design tool is not yet fully capable of realizing all possibilities in the design space of tubes.
 It lacks the network manipulation to make mixing or splitting tubes, unless they are designed in an input vector graphics file.
@@
 Additionally, as mentioned there are opportunities to improve the algorithm for neon sign routing, as ours does not guarantee a shortest-path routing.
---------- this paragraph breaks the latex for some reason?}
+
+\bjoern{Mention need for a better general 3d routing algorithm for multiple traces }
 

### 2014-03-27 paper/conclusion.tex
paper workin

--- 2014-03-24 paper/conclusion.tex
@@
 \section{Conclusion}
-While today's 3D printers are not yet able to fabricate active components in-place, we suggest that 3D printed interactive devices can be created with redirection of active input and output via tubes.
-In addition to describing the design space of tubes and connecting related works created through both traditional processes and rapid prototyping methods, we also presented a design tool for the creation of tubes inside arbitrary 3D models.
+
+While today's 3D printers are not yet able to fabricate active components in-place, we suggest that 3D printed interactive devices can be created with redirection of active input and output via pipes.
+This amplifies and extends the utility of existing sensing and actuation approaches.
+In addition to describing the design space of pipes, we also presented a design tool for the creation of tubes inside arbitrary 3D models and discuss many techniques that combine nicely wtih pipes.
 We fabricated several example objects using this tool to explore new points in the design space.
@@
 We are heartened by our success: it opens new interactive device prototyping opportunities both for makers with consumer-grade 3D printers and research and industrial labs with higher-end machines.
-We are excited to release our tool into the wild and see what people will build.
+We are excited to release our tool into the wild and see what people build.

### 2014-03-27 paper/designspace.tex
paper workin

@@
 Since we are interested in interactive devices, we distinguish a ``user side" - part of the surface or interior of an object facing its user; and a ``system side" - part of the object (either interior or exterior) where other hardware such as electronics will be connected.
-Definitionally, in the following paragraphs we refer to a pipe which is ``open on the user side'' to have its contents physically accessible to an interacting user at interaction time.
-This includes, for example, pipes which function as fountains, but it does not include pipes whose ends have been capped to keep water inside (whether this capping was done at print time, or later as a manual modification).
-Pipes whose contents are not physically accessible to the user are ``closed on the user side''.
-A pipe which is ``open on the system side'' is physically open to the machine, and one which is ``closed on the system side'' has its contents physically cut off from the machine.
+Definitionally, in the following paragraphs we refer to a pipe which is ``open on the user side'' or ``open on the system side'' to have its contents \emph{physically accessible} to an interacting user or the system \emph{at interaction time}.
+A pipe which is ``open on the user side'' or ``closed on the system side'' is physically cut off from the user or machine.
 User and system manipulation of contents are possible with or without physical access, e.g., by resonating, pressurizing, or passing electricity through those contents.
-\bjoern{wordy paragraph}
 
@@
 
-Open pipes originate from the system side and connect the user side, with both ends of the pipe open.
+Open pipes (see Figure \ref{fig:pipespace}A) originate from the system side and connect the user side, with both ends of the pipe open.
 This type of pipe may be used to create, for example, capacitive sensors: an open pipe filled with conductive paint can be connected to a sensing platform (e.g., Arduino) on the system side, while a user can touch the uncapped other end of the pipe.
@@
 
-A return pipe originates at the system and returns back to it, without offering physical access to the user.
+A return pipe (see Figure \ref{fig:pipespace}B) originates at the system and returns back to it, without offering physical access to the user.
 By threading an electroluminescent (EL) wire through a clear return pipe, a maker can create a custom piece of neon art.
@@
 
-Semi-closed pipes are open at the system end (for control of the enclosed medium) and closed at the user end.
+Semi-closed pipes (see Figure \ref{fig:pipespace}C) are open at the system end (for control of the enclosed medium) and closed at the user end.
 The closed interface  may be fabricated from a malleable material: for example, a series of pipes terminated in thin rubber membranes on the user side can be actuated by an air pump to create haptic feedback, or a hinged cap can be blown open by air pressure.
@@
 
-A fully enclosed pipe has no openings on either the system side or the user side.
-Fully enclosed pipes can be used as resonance chambers (e.g., for object identification), or as air bubbles (e.g., as used in \cite{Willis-printedoptics} for internal display).
-It can also be used as a container for water or other particles that would otherwise fall out.
+A fully enclosed pipe (see Figure \ref{fig:pipespace}D) is closed to both the system side and the user side.
+Fully enclosed pipes can be used as resonance chambers (e.g., for object identification), or as air bubbles (e.g., as used in Printed Optics (\cite{Willis-printedoptics}) for internal display).
+They can also be used as containers for water or other particles that would otherwise fall out.
 
-\bjoern{This section in general may need a small multipart figure that shows example photos or at least illustrations of these points in the design space.
-(Fig 2 is the abstract representation of the dimensions; here we have some concrete examples of points in the space).
+\subsection{Design of Pipes}
 
-Or maybe this comes later? it's a bit hard to follow these descriptions in the abstract.
-}
-
-\subsection{Media in Pipes}
-
-Pipes can be filled with a variety of media to create different interface affordances and capabilities.
-
-``Gas'' comprises all compressible fluids.
-Use of fluid pressure inside pipes can create haptic feedback at semi-closed interfaces, or the gases can be used as carriers for scents or fog.
-As in \cite{Slyper-pressure}, structures can be engineered to change in air pressure when manipulated correctly (e.g., a spiral that changes pressure when twisted, but not pressed), and thus fluid pressure can also be used as an input.
-\bjoern{Isn't this the same example as two paragraph above?}
-
-Incompressible fluids (``liquids'') can perform many of the same interface tasks as gases.
-One opportunity with liquids is to fill the interior of pipes with them and cap the ends.
-In addition, one can use driable conductive fluids, such as copper paint, to coat the interior of pipes and allow them to function as arbitrarily-shaped wires.
-This is especially helpful for the creation of a shared ground, or for creating single-wire capacitive interfaces amenable to sensing with SFCS \cite{Sato-touche}.
-
-Pipes need not have hollow centers: in the case where routed pipes are filled with solid material---in particular, a solid material different from the model material---, interactions such as those in Printed Optics (\cite{Willis-printedoptics}) are possible.
-
-Particulates, either printed in-place or inserted, can be of varying densities.
-A single particle can be used for display.
-Sparse particles in a stream of fluid can provide haptic feedback.
-Dense particles in a semi-closed pipe allow for jamming-based interactions at any point on the surface of an object \cite{Follmer-jamming} \bjoern{may need more of an explanation.
-you need jamming plus flexible surface material, right?}.
-
-Threadable inserted elements, such as electroluminescent (EL) wire or fiberoptic cables, are those that can be threaded through pipes post-printing.
-This allows overcoming limitations of printers: for example, a Printed Optics-style interface can be created on an inexpensive consumer-grade 3D printer using pipes and inserted fiber optic cable.
+Pipe designs may emphasize either their exterior features (connection points) or their interior paths.
+These two features lead to different kinds of interfaces.
+An example interface that focuses on the exterior connection points is the touch sensitive toy in Figure \ref{fig:toys}, where pipes must exit the toy at the frontal lobe, parietal lobes, etc.
+An interface focused on the internal path of the pipe is the neon sign in Figure \ref{fig:neon}: output is based on the shape of the pipe.
 
@@
 
-\subsection{Design of Pipes}
+\subsection{Media in Pipes}
 
-\bjoern{Think of the ordering of dimensions is it a conscious choice to list this dimension last?}
-Pipe designs may emphasize either their exterior features (connection points) or their interior paths.
-These two features lead to different kinds of interfaces.
-An example interface that focuses on the exterior connection points is the touch sensitive toy in Figure \ref{fig:toys}, where pipes must exit the toy at the frontal lobe, parietal lobes, etc.
-An interface focused on the internal path of the pipe is the neon sign in Figure \ref{fig:neon}: output is based on the shape of the pipe.
+Pipes can be filled with a variety of media to create different interface affordances and capabilities.
+
+``Gas'' comprises all compressible fluids.
+Fluid pressure inside semi-closed pipes can create haptic feedback, or gases can be used as carriers for scents or fog.
+Slyper, et al.,  in \cite{Slyper-pressure} engineered structures to change in air pressure when manipulated correctly (e.g., a spiral that changes pressure when twisted, but not when pressed); fluid pressure can also be used as an input.
+
+Incompressible fluids (``liquids'') can perform many of the same interface tasks as gases.
+One opportunity with liquids is to fill the interior of pipes with them and cap the ends.
+In addition, one can use driable conductive fluids, such as copper paint, to coat the interior of pipes and allow them to function as arbitrarily-shaped wires.
+This is especially helpful for the creation of a shared ground, or for creating single-wire capacitive interfaces amenable to sensing with SFCS \cite{Sato-touche}.
+
+Pipes need not have hollow centers: in the case where routed pipes are filled with solid material---in particular, a solid material different from the model material---, interactions such as those in Printed Optics (\cite{Willis-printedoptics}) are possible.
+
+Particulates, either printed in-place or inserted, can be of varying densities.
+A single particle can be used for display.
+Sparse particles in a stream of fluid can provide haptic feedback.
+Dense particles in a semi-closed pipe capped with flexible surface material allow for jamming-based interactions at any point on the surface of an object \cite{Follmer-jamming}; when fluid pressure in the pipe is reduced, the trapped particles jam to form a solid mass the same shape as their enclosing flexible membrane.
+
+Threadable inserted elements, such as electroluminescent (EL) wire or fiberoptic cables, are those that can be threaded through pipes post-printing.
+This allows overcoming limitations of printers: for example, a Printed Optics-style interface can be created on an inexpensive consumer-grade 3D printer using pipes and inserted fiber optic cable.

### 2014-03-27 paper/exampleobjects.tex
paper workin

@@
 
-To evaluate our tool's capabilities and to explore new points in the design space, we fabricated a set of five prototype objects designed using the tool.
-All prototypes were fabricated in a single piece, unless described otherwise.
+To evaluate and highlight our tool's capabilities, we fabricated a set of five prototype objects designed using the tool.  All prototypes were fabricated in a single piece, unless described otherwise.
+
+\valkyrie{one way of combatting the enormousness of this paper is to put all these example objects figures into a single figure* instead}
 
@@
 
-We created a set of touch-sensitive toys and a companion app, reminiscent of the boat application in \cite{Harrison-acoustic}.
+We created a set of touch-sensitive toys and a companion app, reminiscent of the boat application in Acoustic Barcodes \cite{Harrison-acoustic}.
 The brain, rabbit, and boat each have several distinct touch points on their surfaces.
@@
 We built a smart base which can distinguish between the toys and also determine which toy is mounted: since each toy and each gesture has a distinct capacitive signature, we use a simple classifier trained to detect both toy and gesture based on profile.
+
+The models we used for the toys' exteriors were all downloaded from the free website Thingiverse.
 All components were fabricated on a Makerbot, support-free.
+After printing, we injected copper paint (CuPro-Cote) into the interior pipes using a craft syringe.
+This paint requires approximately 2 days to fully dry in this configuration (3mm diameter tubes, maximum tube length 10cm).
+Our smart base is powered by an Arduino Uno running code by Instructables users DZL and madshobye.
 
@@
 A custom radio built using tubes allows users to tune to different stations and listen in.
-This device uses a network of disconnected open tubes filled with conductive paint to connect traditional electronic components (two potentiometers and a speaker) to a microcontroller (in our case, an Arduino).
+This device uses a network of disconnected open tubes filled with conductive paint to connect traditional electronic components (two potentiometers and a speaker) to a microcontroller (in our case, an Arduino Pro Micro).
 We hid the microcontroller and a battery-driven power supply in the base of the radio to make it portable.
-For the radio functionality, we attached an Si4703 to the Arduino.
-We fabricated this device on our Makerbot, support-free but cut in two halves to allow the Arduino to be inserted.
+For the radio functionality, we attached an Si4703 to the Arduino, which offers headphone-jack output.
+We fabricated the radio itself on our Makerbot, support-free but cut in two pieces to allow the Arduino and battery pack to be inserted.
 
@@
 Our pen holder uses a modification of the FlyEye technique described by Wimmer in \cite{Wimmer-flyeye} and contains open tubes filled with fiber optic cables, one per pen chamber.
-We use a single cable per chamber; our 6mm diameter fiber optic cables, in comparison to the angel hair tubes used in the original work \bjoern{Mixed terminology between el wire and fiber optics?
-"angel hair" is a specific product name, not a general technical term (I think)}, can send and receive through the same cable.
+We use a single cable per chamber; our 6mm diameter fiber optic cables, in comparison to the fine tubes used in the original work , can send and receive through the same cable.
 At the base of each tube is a QRE1113 line sensor digital breakout board, which has an integrated IR emitter and receiver.
@@
 We created a maze game with a single particle trapped in a fully enclosed tube.
+This maze is based on an SVG we created by hand and processed with our internal path tube tool.
 To fabricate the tube, we created it in two halves that were fastened together via glue, due to limitations in our ability to remove support material otherwise.
-We created this prototype on our Makerbot in two halves to allow for insertion of the marble.
-\bjoern{You are giving two different reasons here for printing in pieces.}
+We created this prototype on our Makerbot in two halves, inserting a marble before closing.
 
@@
 
-\begin{figure}[h]
+\begin{figure}[h!]
 \centering
@@
 For this type of art, the path of the tubes is of crucial importance, as it determines how the sign will look.
-We designed a custom neon sign bearing the UIST logo and the waves of Hawaii.
-The piece features open tubes due to support-flushing constraints, however the tubes could also be semi-closed (one end must remain open to connect the EL wire to its power source).
-They have been threaded with EL wire which is lit in sequence to create an animation.
+We designed a custom neon robot head which can be animated to ``talk''.
+The piece features semi-open pipes: one end must remain open to connect the EL wire to its power source.
+The pipes have been threaded with EL wire which is lit in sequence to create an animation.
 Our sign was fabricated on the Objet in two halves, cutting through the plane of the sign itself, with clear material for the sign around the input design and opaque everywhere else.
-We manually flushed support from the non-planar connecting tubes.
+We manually flushed support from the non-planar connecting pipes.
 The EL wire was threaded post-print, and is controlled by an EL wire sequencer.

### 2014-03-27 paper/limitations.tex
paper workin

--- 2014-03-24 paper/limitations.tex
@@
 \section{Limitations and Future Work}
-Tubes, while a powerful tool for directing interaction to arbitrary locations on an object's surface, have their shortcomings.
+Pipes, while a powerful tool for directing interaction to arbitrary locations on an object's surface, have their shortcomings.
 Most obviously, their fundamental shortfall is that they cannot replace the active elements required for a functioning interactive device: they simply redirect sensing to different locations on an object's surface.
-All tubes need to, at some point, connect to a sensor, microcontroller, or pump.
+All pipes need to, at some point, connect to a sensor, microcontroller, or pump.
 
-Some media are difficult to insert in tubes, or may not be compatible with all types of tubes.
+Some media are difficult to insert in pipes, or may not be compatible with all types of pipes.
 Fluids in particular provide challenges.
@@
 
-\valkyrie{Our design tool is not yet fully capable of realizing all possibilities in the design space of tubes.
-It lacks the network manipulation to make mixing or splitting tubes, unless they are designed in an input vector graphics file.
+Our design tool is not yet fully capable of realizing all possibilities in the design space of pipes.
+It lacks the network manipulation to make mixing or splitting pipes, unless they are designed in an input vector graphics file.
+Since pipes are committed to the mesh one at a time, our tool does not currently optimize multiple pipes at a time: we simply make greedy pairwise decisions.
+This suggests the need for a better general 3D routing algorithm with multiple traces.
 Future work should also include assisting makers in the fabrication process itself by automatically cutting objects to reduce support material removal issues.
 Additionally, as mentioned there are opportunities to improve the algorithm for neon sign routing, as ours does not guarantee a shortest-path routing.
---------- this paragraph breaks the latex for some reason?}
 

### 2014-03-27 paper/discussion.tex
adding bjoern comments
paper workin

@@
 
+The promise of 3D printing is that it allows for the manufacture of geometries that are impossible to create using traditional techniques like injection molding.
+Pipes leverage that promise: designing the internal structure of an object was not previously feasible.
+As printers at the hobbyist and corporate levels become more sophisticated and able to fabricate intricate geometries of ever more materials, we believe that our tool will still assist in design tasks.
+Even if printers are capable of laying down conductive, optically transparent, or luminescent materials in-place, the process of designing their paths is still crucial.
+
+Once objects have been designed and fabricated, the major flexibility for prototyping with pipes stems from the media inserted in them and the systems attached which sense changes in, or actuate, those media.
+The five types of media--gas, liquid, solid, particulate, and threadable--give rise to unique possibilities in output (visual, aural, tactile/haptic, olfactory/gustatory) as well as input (touch/pressure and others).
+
+\valkyrie{this section is probably my project for the weekend, or something.  I can't figure out how to improve it right now.  Tovi, we should talk about this in our meeting this afternoon.}
+
+\tovi{There's lots of great ideas an examples in the inputs and outputs section, but the orgnization and coorespondence to the table is a little hard to follow. }
+
+By crossing pipe types, topologies, and inserted media, we create a space of possible inputs and outputs that can be pipe-mediated.
+We offer a table (see Figure \ref{fig:designspace}) containing these possibilities, referencing previous work on input/output techniques where appropriate, noting that many of these techniques have not been attempted using digital fabrication \bjoern{It may be a bit tricky to define what "using digital fab" means here.
+For example, Harrison 's pneumatically actuated buttons were done on a laser cutter, but also required manual gluing of other layers.
+Does that count or not?}.
+We suggest new points in the design space that have been unexplored or not yet explored with digital fabrication.
+
+\begin{figure*}[t]
+\centering
+    \includegraphics[width=\textwidth]{figures/designspace.pdf}
+\caption{The design space of pipe-based interactions.
+Existing systems are written in regular font.
+Those created with fabrication are in {\color{blue}blue}.
+In \emph{{\color{red}red italic}} are unexplored interactions creatable with custom-fabricated pipes.
+\tovi{Im not sure how well this table will work.
+Division between input and output should be more prominent.} \valkyrie{one possibility is making the table larger and adding some short descriptions.
+we could throw out a lot of words from the paper that way}}
+\label{fig:designspace}
+\end{figure*}
+
+\subsection{Outputs}
+
+We describe potential pipe-based outputs organized by the five senses: visual, aural, haptic, and olfactory/gustatory.
+In addition to outputs we mention explicitly, any outputs possible using traditional electronic components are possible as described above.
+
+\begin{figure}[h]
+\centering
+    \includegraphics[width=3.4in]{figures/placeholder/direct.jpg}
+\caption{By attaching a servo to a gate structure, it is possible to direct fluid through that pipe.
+The gate not only has binary positions (a) and (b), but can direct fractional flow, as well (c).
+This could be used to create logic gates, as in (d).
+These could be used, for example, in science exhibits.}
+\label{fig:direct}
+\end{figure}
+
+\emph{Visual} outputs can be mediated by gases, liquids, solid, particulates, or threadables.
+Pneumatic actuation (as in \cite{Yao-pneui} or \cite{Harrison-buttons}) provides output, but gases carrying smoke (as in \cite{Sodhi-aireal} demo) also offers visual feedback.
+Splitting and mixing of colored liquids can be used as a display tactic; additionally mechanical motion can be fluid-mediated.
+Liquids can also be mixed in specific concentrations by servo-controlled gate structures as in Figure \ref{fig:direct}.
+Printed optics \cite{Willis-printedoptics} explored visual output at the ends of pipes, and   EL wire can be threaded through custom paths to create neon signs, as in Figure \ref{fig:neon}.
+
+\emph{Aural} outputs can be created with gases.
+Helmholz resonance can be leveraged to create computer-controllable sound chambers, as in Zoran's flute \cite{Zoran-flute}.
+All that is necessary for this is an enclosed chamber with a pipe connected to it via a narrow neck-point (see Figure \ref{fig:ocarina}).
+Passive sound amplification and sound redirection are also possible using pipes.
+
+\begin{figure}[h]
+\centering
+    \includegraphics[width=3.4in]{figures/placeholder/helmholz.jpg}
+\caption{This box (a) contains an enclosed Helmholz resonator, seen in the cutaway view (b).
+This resonator creates a sound when air is blown through the pipe at the top, because air compresses at the small neck joint between the chamber and the pipe.
+\valkyrie{need to get a test print of this that actually works}}
+\label{fig:ocarina}
+\end{figure}
+
+\emph{Haptic} outputs are possible through compressible and incompressible fluids.
+Fluids can actuate particulates and create programmable pliability \cite{Follmer-jamming}.
+Additionally, use of semi-closed pipes with rubberlike material as the caps allows for gases to actuate surface features (see Figure \ref{fig:breathe}).
+Free-air haptic feedback \cite{Sodhi-aireal} and volumetric interactions \cite{Iwata-volflex} can be pipe-mediated.
+
+\emph{Olfactory} and \emph{gustatory} outputs can be created through the mixing and splitting of pipes carrying scented or flavored fluids, respectively.
+
+\subsection{Inputs}
+
+Pipes create opportunities for many types of input sensing across the surface of printed devices.
+We describe sensing touch, pressure, grasp, flexing, tapping, and manipulation of traditional electronic components.
+There are likely more input sensing techniques, both existing and on the horizon, that are compatible with the use of pipes.
+
+\emph{Touch} sensing can be enabled through pipes filled with conductive media.
+Traditional wires (threadable) may be used, but conductive paint (liquid) may be more flexible.
+Using conductive paint and Swept Frequency Capacitive Sensing (SFCS \cite{Sato-touche}), an interior star pipe topology can enable single-wire touch- and grasp-sensing at any set of points on a printed object's surface (see Figure \cite{fig:toys}).
+Simple capacitance measurements are possible using the same media, and custom-designed sensors like those created in Savage, et al.,'s Midas system are also possible \cite{Savage-midas}.
+
+\emph{Pressure} can be sensed in multiple ways.
+Slyper, et al.,'s primitives can be attached as the terminus of any semi-closed pipe; the user can manipulate the printed endpoint and the system can sense air pressure changes \cite{Slyper-pressure}.
+Pressure can also be sensed via capacitance.
+
+\emph{Grasp} sensing can be enabled using Wimmer's FlyEye techniques \cite{Wimmer-flyeye} or those described in Jamming User Interfaces \cite{Follmer-jamming}.
+The optical links necessary for sensing can be created either via clear solid cores as in \cite{Willis-printedoptics} or via fiber optic cables threaded through hollow pipes (see Figure \ref{fig:pens}).
+
+Some 3D printers can fabricate rubber-like materials.
+Similar to Slyper, et al., in \cite{Slyper-shape}, we can sense \emph{flexing} and \emph{bending} of prints made on these machines: the ability to 3D print devices to sensing bending and flexing significantly cuts molding and assembly time.
+
+\emph{Tapping} is another input possibility.
+Due to differential sound conductivity in different printed materials, pipes of a greater conducting material can be embedded in a model of a lesser conducting material.
+A microphone or piezo placed at the system-side of these sound-conducting pipes can determine where the model was tapped and how it was tapped.
+Active acoustic sensing as described in Touch \& Activate \cite{Ono-touchandactivate} is also possible; this technique would allow for particular areas of interest (connected to sound-conducting pipes) to be touch sensitive, while other areas (made of non-conducting material) are not sensed.
+
+Sensing via traditional electronic components (e.g., potentiometers, alcohol gas sensors) can be accomplished by conductive inserted media.
+Components can be recessed into a print's surface, with their leads implanted into open pipes.
+By inserting liquid copper paint instead of threading traditional wires, all components in a model can share a ground line, and the paint's drying process obviates the use of solder or glue to affix the components in place (see Figure \ref{fig:radio}).
+
+\subsection{Fabrication Techniques}
+
+\bjoern{The biggest question wrt fabrication for me is - how do I print these cavities? So general support material avoidance or removal strategies should be more prominent.
+
+(I think the lye bath would probably work better than the high pressure water for our objet prints)}
+The physical fabrication process used by different machines gives rise to differing design guidelines.
+The two machines we used to create our example objects were the Objet Connex 260 and the Makerbot Replicator 2.
+
+The Objet is an inkjet-based 3D printer which jets drops of polymer in very fine layers, curing each layer with an ultraviolet lamp.
+It offers a wide range of material types, including clear/opaque, flexible/hard, and black/white.
+Additionally, it permits the mixing of two different materials in a single print: a mixed material could include 70\% flexible and 30\% hard, for example, to be firm but somewhat flexible.
+The flexible materials can be inflated when they are thin: they elongate 200\% before breaking.
+We have found $.05$inches to be a good thickness for inflatable materials.
+Although this machine has many extraordinary capabilities, it unfortunately cannot print overhangs of greater than $14^{\circ}$ without support material.
+The support structure used by this machine is a loose matrix of material that can be removed by hand from the surface or blasted out with a water jet from the interior.
+The support is also soluble in lye.
+
+The Makerbot Replicator 2 is a fused-deposition modeling (FDM) printer, in which we use exclusively Polylactic Acid (PLA) bioplastics.
+More material properties are being explored in PLA: there are now transparent and flexible materials.
+They cannot be blended like those in the Objet.
+The Makerbot has a fairly coarse resolution compared to the Objet, however it permits overhangs up to $45^{\circ}$ without support and also allows bridging (short spans of material that bridge the gap between two structures with no support underneath).
+Due to the FDM process it can even recover from greater overhangs and longer bridges, as the material builds up and can reorganize into a structure \valkyrie{how do I say this?
+do I need a photo?
+print recovery is pretty awesome-looking.}.
+The support used by the Makerbot is made of the same material as the model itself, just printed in a different pattern and attached over as small a surface area as possible to ease removal.
+Support inside of a pipe may require pliers, or be infeasible to remove dependent upon geometry.
+
+Assembly is easiest on both machines when the model does not require support material.
+Models can be ``cut up'' \bjoern{This negates some of the benefits compared to other manufacturing techniques.} into assembleable pieces to evade support requirements or ease material removal.
+We believe that an automatic tool for this task would be highly beneficial, however it is outside the scope of this work.
+
+\subsection{Identity}
 Pipes included in objects can have additional effects beyond their designed intentions.
@@
 While their technique requires access a terahertz imaging tool, audio resonance identity encoding requires only a microphone and a speaker.
-
-\valkyrie{To discuss: printing media into tubes?  Design of interiors of objects more generally?  Need to cite that paper about balancing printed objects from SIGGRAPH.}

### 2014-03-27 paper/introduction.tex
paper workin

@@
 
+Many existing sensing and actuation approaches, such as FlyEye \cite{Wimmer-flyeye}, Jamming User Interfaces \cite{Follmer-jamming}, and Touch\'{e} \cite{Sato-touche}, can be enhanced by the use of pipes.
+Pipes allow makers utilizing these sensing and actuation strategies to locate input and output at arbitrary points on an object's surface.
+Thus, we leverage these existing techniques for sensing and actuation while our work's novelty is in \emph{internal modeling} for 3D printing and the \emph{redirection} of IO to arbitrary locations on a device.
+
 Our work makes the following contributions to the field of digitally fabricated interactive objects:
@@
 \item We offer algorithms and techniques for routing pipes, as well as a design tool implementing them.
+\item We showcase a set of examples, enabled by our modeling tool and exploring new points in the design space.
 \item We describe the interaction possibilities of pipes, including prior work as well as uncharted possibilities.
-\item We showcase a set of examples, enabled by our modeling tool and exploring new points in the design space.
 \end{itemize}

### 2014-03-27 paper/relatedwork.tex
paper workin

@@
 
+\valkyrie{can you offer suggestions?
+I can imagine the latex buttons paper, touche, acoustic barcodes, capacitive fingerprinting \cite{Harrison-capfingerprint}, printed optics for harrison/poupyrev, and for baudisch maybe....
+TDR \cite{Wimmer-TDR}?
+fiberio \cite{Holz-fiberio}?
+fabrickation?
+I'm not sure on this.}
+
 \subsection{Fabrication of Interactive Devices}
@@
 Previous work has investigated the integration of interactivity with the fabrication process.
-Printed Optics (\cite{Willis-printedoptics}) used light pipes, solid clear tubes, to create integrated displays: our work is inspired by this and goes towards a design tool for these capabilities \valkyrie{bad phrasing; don't de-fang the tool at the front of the paper!}; Printed Optics is a part of our design space, which includes other uses of pipes.
+Printed Optics (\cite{Willis-printedoptics}) used light pipes, solid clear tubes, to create integrated displays: our work is inspired by this and goes towards a design tool for these capabilities; Printed Optics is a part of our design space, which includes other uses of pipes.
 Willis, et al., also did not offer a design tool or techniques for creating light pipes automatically.
-Sauron (\cite{Savage-sauron}) investigated the use of computer vision to track interactions with physical mechanisms.
+Sauron investigated the use of computer vision to track interactions with physical mechanisms \cite{Savage-sauron}.
 While Savage, et al., used mirrors for redirection, we believe that the required line-of-sight operation for their computer vision techniques may be possible using printed optics or inserted optical fibers.
@@
 This technique is complementary to ours, as we allow for placing these interactive elements anywhere on an object's surface while they had to place them in locations accessible for electronics insertion.
+General-purpose routing for connecting 2D capacitive sensors to terminals is discussed in Midas \cite{Savage-midas}.
+This technique, based on circuitboard routing, are limited to planar and unfoldable shapes, and are intended for use on the surface of objects rather than through their cores.
+Acoustic Barcodes explored integrating ridges into digitally fabricated objects, whose acoustic signatures when scraped provided an ID code \cite{Harrison-acoustic}.
+Fabrickation works towards rapid fabrication of functional objects by integrating construction kit building blocks \cite{Mueller-fabrickation}.
 
@@
 
-General-purpose routing for connecting 2D capacitive sensors to terminals is discussed in Midas (\cite{Savage-midas}) \bjoern{Maybe move discussion of Midas to prior section since it really is about fabricating interactive devices.
-And then here just quickly review that lots of work has gone into routing algorithms for PCBs but that they assume multiple stacked planes connected with vertical vias, not general 3d routing.}.
-These techniques, based on circuitboard routing, are limited to planar and unfoldable shapes, and are intended for use on the surface of objects rather than through their cores.
+Circuitboard routing algorithms, such as Lee's maze router algorithm, work only on a 2D surface.
+So-called ``3D circuitboards'' are typically 2.5D, in that they comprise a stack of 2D circuitboards connected by vias \valkyrie{citation needed}.
 
@@
 Schweikardt, et al., also made a connection between electroluminescent wire and graph theory, though their contribution is orthogonal to ours: they provide a tangible construction kit with smart nodes for building physical graph examples in which edges are lit by EL wire \cite{Schweikardt-tei09}.
+In traditional neon sign making, ``routing'' is a conversation between a graphic designer and a glass bender on what is feasible and comfortable for the latter and what will help him avoid burning his hands, wasting glass, and save bending time \cite{strattman1997neon}.
 
 Prior work on modeling the interior of objects (as opposed to their exterior) has been limited.
-\valkyrie{ryan, can you give citations here?
-I know there is that one about balancing, but Tovi mentioned a ``fruit slicing" one?}  In Make it Stand, material is removed from the interior of objects with the goal of changing their center of gravity \cite{Prevost-makeitstand}.
+In Make it Stand, material is removed from the interior of objects with the goal of changing their center of gravity \cite{Prevost-makeitstand}.
 Bickel, et al., explored enforcing deformation behavior with material blending \cite{Bickel-deformation}, while B\"{a}cher, et al., built an algorithm for automatically placing 3D fabricatable joints into meshes \cite{Bacher-articulated}.
@@
 Our work differs in that we want to allow users to design pipes and integrate interactive objects with them.
-
-\subsection{Misc}
-\bjoern{These may not belong in related work, but I'm just collecting them here.}
-\begin{itemize}
-\item \cite{strattman1997neon} - Strattman is the standard "text book" for neon lighting designers.
-\end{itemize}

### 2014-03-27 paper/tubes.tex
paper workin

--- 2014-03-24 paper/tubes.tex
@@
 \section{A Series of Tubes - A Tool for Pipe Design}
+
+As mentioned, pipe design is not trivial.
+Independent pipes cannot intersect lest their contents be unintentionally mixed.
+Pipes' bending energy needs to be minimized to allow for easy insertion of media post-print.
+Pipes in general need to avoid the surface of objects to prevent fluid leakage when printed by hobbyist-grade machines.
+Finally, for designing pipes that follow a user-specified path, we must ensure certain characteristics of that path (e.g., that the path is connected).
 
@@
 This tool allows designers to select exterior connection points of their pipes (see Figure \ref{fig:tool-process-interior}) or to import vector art describing the pipes' interior paths (see Figure \ref{fig:tool-process-exterior}).
-Once the maker's selections are made, we create a complete routing using either the exterior connection point method (A*, physical simulation) or the interior path method (graph edge creation, Euler circuit generation).
+Once the maker's selections are made, we create a complete routing using either the exterior connection point method (A*, physical simulation) or the interior path method (graph edge creation, Euler tour generation).
 We thicken our routing to create pipes.
@@
 
-Our tool is implemented as a part of Meshmixer, a consumer 3D mesh editing tool, in C++.
+Our tool is implemented as a part of Meshmixer, a consumer 3D mesh editing tool, in C++ \cite{Schmidt-meshmixer}.
 
@@
 
-To create pipes in which the location and shape of exterior connection points matters, we first allow makers to select the connection points on the mesh's surface using a brush.
+To create pipes in which the location and shape of exterior connection points matters, we first allow makers to select the connection points on the mesh's surface.
 Our tool then creates an initial shortest-path routing using A* to estimate the routed distance between points.
@@
 \centering
-    \includegraphics[width=3.4in]{figures/placeholder/exterior.jpg}
-\caption{An example mesh with exterior connection points selected.  The initial A* routing is drawn in {\color{red}red}, and in {\color{tovi}green} is our physically-based rod after simulation.}
+    \includegraphics[width=3.4in]{figures/exterior.png}
+\caption{An example mesh with exterior connection points selected.  The smoothed A* routing is drawn in {\color{red}red}, and in {\color{tovi}green} is our physically-based rod after simulation.}
 \label{fig:tool-process-exterior}
@@
 Our basic first-pass routing algorithm uses the A* routing algorithm \cite{Hart-Astar}.
-The path cost in our implementation is based only on shortest distance between the starting and ending points, without weighting for distance from the surface.
-We only permit the routing within the voxel grid of the mesh: i.e., the routing is not permitted to traverse through space outside the mesh itself.
+This algorithm keeps a priority queue of points to check for routing based on a cost function; for each point it tags the distance from the start (e.g., startPoint = 0, startNeighbors=1, ...).
+Once the end point is reached, we backtrack through the tagged points, always moving towards a lower-valued point.
+The path cost in our implementation is based only on shortest distance between the starting and ending points, without weighting for distance from the surface, however the routed path is forced to stay inside the mesh's boundaries.
+We route on the voxelized 3D grid of the mesh, and we use a resolution of 128x128x128 voxels.
 
-Given this initial routing, we create a rod of slightly longer proportional length.
-The rod must be longer, as the A* routing can follow the mesh boundaries and the rod requires extra length to be pushed away from the mesh surface.
-While we do not have a theoretical basis for this proportional increase, \valkyrie{yet.
-what might it be related to?
-we can test how much of the A* path is outside the model, and how much is inside; further we can consider the distance from the A* path that is outside the model to the surface.
-I think there's a good way to do this.
-It might even just mean iteratively changing the rod length until we get something reasonably short and also smooth.
-we can also return to the A* that I wrote which finds a routing through the mesh that respects its boundaries; this would get us closer to what we need for length in the first place.}, we have found 1.15 to be a reasonable multiplier in practice.
-A too-long rod is forced to kink and expand further than necessary inside the model, while a too-short rod must be stretched, leading to non-smooth bending.
+Note that pipes, once cut, become ``outside the mesh''.
+This means that pipes cut in sequence cannot intersect each other.
+However, we do not currently track previously-cut pipes for global optimization: we simply greedily select the best routing per pair of points (i.e., per pipe).
 
-To allow for easiest insertion of media post-print, our physical simulation step minimizes the bend radius of the generated rods.
-In order to force our simulated rods away from the surface of the mesh, we create ``wind'' pushing inward from the exterior of the mesh.
+Given our initial routing, we create a rod of the same length.
+To allow for easiest insertion of media post-print, our physical simulation step minimizes the bend radius of the generated rod.
+In order to force our simulated rod away from the surface of the mesh, we create ``wind'' pushing inward from the exterior of the mesh.
 This wind keeps the resulting pipe away from the surface.
@@
 \subsubsection{Pipes with Non-Round Endpoints}
-Our tool does not explicitly allow makers to design the shape of their pipes' endpoints.
+Although start and end radius of pipes is configurable, our tool does not explicitly allow makers to design the \emph{shape} of their pipes' endpoints.
 However, the larger Meshmixer package does allow brushing arbitrary shapes onto the surface and extruding them inward, as in Figure \ref{fig:meshmixer-endpoint}.
+This technique can be used to create specially-shaped touch pads or to recess the footprint of an electronic component for it to sit flush with the object's surface.
 
@@
 Neon signs, as in Figure \ref{fig:tool-process-interior}(d), require a single long path that passes through many desired edges: for example, in our UIST sign, the path must pass through all the edges that comprise the word ``UIST''.
-To allow makers to create novel neon signs, as well as similar objects that require design of the pipe's interior path, we built a tool which leverages graph theory to generate this single long path.
+To allow makers to create novel neon signs, as well as similar objects that require design of the pipe's interior path (for example the maze in Figure \ref{fig:maze}), we built a tool which leverages graph theory to generate this single long path.
 
@@
     \includegraphics[width=3.4in]{figures/interior.pdf}
-\caption{An input vector graphics file with the points which cannot be tubed as drawn highlighted in {\color{blue}blue} (a).
+\caption{An input vector graphics file with the points which cannot be tubed as drawn highlighted in {\color{blue}blue} and the start and end points highlighted in {\color{gray}gray} (a).
 The connected graph created by our software (b) and the resulting Euler circuit (c) permit creation of a novel neon sign (d).}
@@
 \subsubsection{Routing}
+In graph theory, a semi-Eulerian graph is one which has exactly two edges of odd degree.
+This kind of graph supports an Euler tour: a path that touches every edge exactly once, beginning and ending at the two nodes of odd degree (in this case, the user-selected start and end nodes).
+To allow for a single path through a pipe for post-print media insertion, we need to find a semi-Eulerian graph for which a user's input graphics are a subset, and subsequently find an Euler tour on that graph.
+
 The interior path routing problem is a version of the Chinese Postman Problem\footnote{The CPP is also known as the route inspection problem: \url{http://en.wikipedia.org/wiki/Route_inspection_problem}}, in which we wish to traverse all edges of the graph described in the user's input with a single Euler circuit, much as a postman needs to walk along every road at least once to deliver mail.
 We call our relaxation the Spiderman Postman Problem: we allow the creation of non-existing paths (i.e., the postman may traverse buildings in addition to roads).
-If inputted path components are disconnected, we must create edges that connect them; additionally we can create new edges connecting odd-degree vertices rather than simply retracing existing edges.
+If input path components are disconnected, we must create edges that connect them; additionally we can create new edges connecting odd-degree vertices rather than simply retracing existing edges.
 In the final artifact, all created edges will be blocked out by dark material so the inserted medium is not visible (see Figure \ref{fig:tool-process-interior}).
@@
 
-In order to connect disconnected subgraphs in the input, we reduce each disconnected subgraph to a point.
-We add edges from each subgraph-point to each other subgraph-point that represent the minimum distance between any two points in the expanded subgraphs (where distance is Euclidean distance).
+In order to connect disconnected subgraphs in the input, each disconnected subgraph's vertices and edges are contracted to a single vertex.
+Each of these subgraph-vertices has no outgoing edges, as the expanded subgraphs are disconnected.
+We then add one edge from each subgraph-vertex to each other subgraph-vertex, whose weight is the minimum distance between any two vertices in the expanded subgraphs (where distance is Euclidean distance).
 We greedily select the smallest weight edges until all subgraphs are joined into a single graph.
@@
 
-We believe that a lower total weight matching is possible by connecting components and ensuring edge degree evenness together in a global process, however it this not crucial for our purposes.
+We believe that a lower total weight matching is possible by connecting components and ensuring edge degree evenness together in a global process, as well as by using minimum-weight matching rather than greedy selection, however this optimization is not crucial for our purposes.
 
 Once we have a semi-Eulerian graph, we need to create an Euler tour.
-An Euler tour is a path that touches every edge once, beginning and ending at the two nodes of odd degree: in this case, those are the user-selected start and end nodes.
-We use a weighted modification of Fleury's algorithm\footnote{\url{http://en.wikipedia.org/wiki/Eulerian_path\#Fleury.27s_algorithm}} for this, where instead of randomly selecting from non-bridge paths at each node, we select the path which turns the least from the most recent path (i.e., we prefer to pass straight through a node, if possible).
+We use a weighted modification of Fleury's algorithm for finding Euler tours\footnote{\url{http://en.wikipedia.org/wiki/Eulerian_path\#Fleury.27s_algorithm}} for this.
+In the classic algorithm, each selected edge is removed to create a reduced graph.
+To select the next edge while at a node, the algorithm randomly chooses from all incident non-bridge edges (where a bridge edge would create disconnected non-trivial subgraphs if it were removed).
+This ensures that every edge is both reachable and traversed.
+In our modification, instead of randomly selecting from non-bridge paths at each node, we select the non-bridge edge which turns the least from the most recent path (i.e., we prefer to pass straight through a node, if possible).
 This minimizes turns in the final artifact, which eases support material removal and assembly.
@@
 We do not currently redirect edges that intersect in the plane.
-Because we have a 3-dimensional canvas in which to work, it would be possible to push overlapping paths into the third dimension, however the angel hair EL wire we have used for our prototypes is so thin as to make this unnecessary.
+Because we have a 3-dimensional canvas in which to work, it would be possible to push overlapping paths into the third dimension.
+However, we leave this to future work: for our specific applications so far, it has been unnecessary.
 
@@
 
-How we actually cut stuff and make tubes happen. \valkyrie{Ryan?}
+Given either our rod-based or image-based routing, we create cylinders which use the routings as a sweep path.
+At each point in the routing, we create a cylinder tall enough to reach to the next point which is oriented along the line segment connecting its origin point to its destination point.
+This cylinder's radius is set by the user.
+For the rod-based routing, we allow independent start and end radii; in this case, the cylinder's radius is $\frac{k}{n}*\frac{r_1}{r_2}$, where $k$ is the point's index in the routing, $n$ is the total number of points in the routing, $r_1$ is the starting radius, and $r_2$ is the ending radius.
+
+All cylinders generated are joined into a large rod.  This rod is voxelized, along with the original mesh.  These two voxel grids are subtracted, and the result is re-meshed and smoothed.

### 2014-03-28 paper/tubes.tex
add figure about voxelization error, reorganize a whole bunch of stuff, basically this is what you will need to work from this weekend!

@@
     \includegraphics[width=3.4in]{figures/exterior.png}
-\caption{An example mesh with exterior connection points selected.  The smoothed A* routing is drawn in {\color{red}red}, and in {\color{tovi}green} is our physically-based rod after simulation.}
+\caption{An example mesh with exterior connection points selected.
+The smoothed A* routing is drawn in {\color{red}red}, and in {\color{tovi}green} is our physically-based rod after simulation.
+\valkyrie{make a figure that is less...
+tall}}
 \label{fig:tool-process-exterior}
@@
 \subsubsection{Pipes with Non-Round Endpoints}
+\valkyrie{this section can probably be scrapped, along with Figure \ref{fig:meshmixer-endpoint}.}
 Although start and end radius of pipes is configurable, our tool does not explicitly allow makers to design the \emph{shape} of their pipes' endpoints.
@@
 \caption{An input vector graphics file with the points which cannot be tubed as drawn highlighted in {\color{blue}blue} and the start and end points highlighted in {\color{gray}gray} (a).
-The connected graph created by our software (b) and the resulting Euler circuit (c) permit creation of a novel neon sign (d).}
+The connected graph created by our software (b) and the resulting Euler circuit (c) permit creation of a novel neon sign (d).
+\valkyrie{for this sign, embed it in a fun hawaiian block}}
 \label{fig:tool-process-interior}
@@
 
-All cylinders generated are joined into a large rod.  This rod is voxelized, along with the original mesh.  These two voxel grids are subtracted, and the result is re-meshed and smoothed.
+All cylinders generated are joined into a large rod.
+This rod is voxelized, along with the original mesh.
+These two voxel grids are subtracted, and the result is re-meshed and smoothed.
+When a user desires to cap the ends of her pipes, we create a sphere of the same radius as each endpoint whose center is the same as the endpoint.
+We intersect this sphere with the original mesh, then intersect this partial-sphere mesh with a shell of the original mesh.
+This gives us a voxel subgrid of the thickness of cap we desire (the same as the shell's thickness) which is located only in the region of the tube end.
+During subtraction, we preserve this subgrid.
+\valkyrie{maybe need a figure?}
+
+Unfortunately, because we use voxelization as a part of our subtraction operation, our models can accumulate error when many pipes are created in sequence (see Figure \ref{fig:voxelize}).
+This problem could be solved by performing a multi-pipe operation; fewer rounds of voxelization and re-meshing lead to lower error.
+This multi-pipe operation would also lead to a more globally optimal pipe routing.
+
+\begin{figure}[h!]
+\centering
+    \includegraphics[width=3.4in]{figures/voxelize-fail.png}
+\caption{After repeated pipe creation, the mesh containing these pipes, which were not originally intersecting, accumulated sufficient error to open the pipes into each other.}
+\label{fig:voxelize}
+\end{figure}

### 2014-03-28 paper/discussion.tex
notes from tovi
add figure about voxelization error, reorganize a whole bunch of stuff, basically this is what you will need to work from this weekend!

@@
 
-Once objects have been designed and fabricated, the major flexibility for prototyping with pipes stems from the media inserted in them and the systems attached which sense changes in, or actuate, those media.
-The five types of media--gas, liquid, solid, particulate, and threadable--give rise to unique possibilities in output (visual, aural, tactile/haptic, olfactory/gustatory) as well as input (touch/pressure and others).
-
-\valkyrie{this section is probably my project for the weekend, or something.  I can't figure out how to improve it right now.  Tovi, we should talk about this in our meeting this afternoon.}
+\tovi{we should make this into a "1D table", just base it on the I/O we can enable.
+scrap the table illustration itself; it's not needed.
+also, focus more on future work in these areas, and work on relating back to what we've discussed already (and the examples we've built)}
 
@@
 By crossing pipe types, topologies, and inserted media, we create a space of possible inputs and outputs that can be pipe-mediated.
-We offer a table (see Figure \ref{fig:designspace}) containing these possibilities, referencing previous work on input/output techniques where appropriate, noting that many of these techniques have not been attempted using digital fabrication \bjoern{It may be a bit tricky to define what "using digital fab" means here.
-For example, Harrison 's pneumatically actuated buttons were done on a laser cutter, but also required manual gluing of other layers.
-Does that count or not?}.
-We suggest new points in the design space that have been unexplored or not yet explored with digital fabrication.
-
-\begin{figure*}[t]
-\centering
-    \includegraphics[width=\textwidth]{figures/designspace.pdf}
-\caption{The design space of pipe-based interactions.
-Existing systems are written in regular font.
-Those created with fabrication are in {\color{blue}blue}.
-In \emph{{\color{red}red italic}} are unexplored interactions creatable with custom-fabricated pipes.
-\tovi{Im not sure how well this table will work.
-Division between input and output should be more prominent.} \valkyrie{one possibility is making the table larger and adding some short descriptions.
-we could throw out a lot of words from the paper that way}}
-\label{fig:designspace}
-\end{figure*}
+We arrange our I/O discussion around the modes of input or output (the five senses, touch/tapping/pressure/etc.).
+We also suggest new points in the design space of pipe-mediated interaction that remain unexplored.
 
@@
 We describe potential pipe-based outputs organized by the five senses: visual, aural, haptic, and olfactory/gustatory.
-In addition to outputs we mention explicitly, any outputs possible using traditional electronic components are possible as described above.
+In addition to outputs we mention explicitly, any outputs possible using traditional electronic components are possible using conductive material in pipes.
 
@@
 \emph{Visual} outputs can be mediated by gases, liquids, solid, particulates, or threadables.
-Pneumatic actuation (as in \cite{Yao-pneui} or \cite{Harrison-buttons}) provides output, but gases carrying smoke (as in \cite{Sodhi-aireal} demo) also offers visual feedback.
+Pneumatic actuation (as in PneUIs \cite{Yao-pneui} or Harrison's latex buttons \cite{Harrison-buttons}) provides output, but gases carrying smoke (as in the Aireal \cite{Sodhi-aireal} demo) also offers visual feedback.
 Splitting and mixing of colored liquids can be used as a display tactic; additionally mechanical motion can be fluid-mediated.
 Liquids can also be mixed in specific concentrations by servo-controlled gate structures as in Figure \ref{fig:direct}.
-Printed optics \cite{Willis-printedoptics} explored visual output at the ends of pipes, and   EL wire can be threaded through custom paths to create neon signs, as in Figure \ref{fig:neon}.
+Printed optics \cite{Willis-printedoptics} explored visual output at the ends of pipes, and  EL wire can be threaded through custom paths to create neon signs, as in our neon robot (Figure \ref{fig:neon}).
 
@@
 Fluids can actuate particulates and create programmable pliability \cite{Follmer-jamming}.
-Additionally, use of semi-closed pipes with rubberlike material as the caps allows for gases to actuate surface features (see Figure \ref{fig:breathe}).
-Free-air haptic feedback \cite{Sodhi-aireal} and volumetric interactions \cite{Iwata-volflex} can be pipe-mediated.
+Additionally, use of semi-closed pipes with rubberlike material as the caps allows for gases to actuate surface features, like the back of our breathing rabbit (see Figure \ref{fig:breathe}).
+Free-air haptic feedback as in Aireal \cite{Sodhi-aireal} and volumetric interactions as in Volflex \cite{Iwata-volflex} can also be pipe-mediated.
 
@@
 Traditional wires (threadable) may be used, but conductive paint (liquid) may be more flexible.
-Using conductive paint and Swept Frequency Capacitive Sensing (SFCS \cite{Sato-touche}), an interior star pipe topology can enable single-wire touch- and grasp-sensing at any set of points on a printed object's surface (see Figure \cite{fig:toys}).
+Using conductive paint and Swept Frequency Capacitive Sensing (SFCS \cite{Sato-touche}), an interior star pipe topology can enable single-wire touch- and grasp-sensing at any set of points on a printed object's surface, like our toys (see Figure \cite{fig:toys}).
 Simple capacitance measurements are possible using the same media, and custom-designed sensors like those created in Savage, et al.,'s Midas system are also possible \cite{Savage-midas}.
@@
 \emph{Grasp} sensing can be enabled using Wimmer's FlyEye techniques \cite{Wimmer-flyeye} or those described in Jamming User Interfaces \cite{Follmer-jamming}.
-The optical links necessary for sensing can be created either via clear solid cores as in \cite{Willis-printedoptics} or via fiber optic cables threaded through hollow pipes (see Figure \ref{fig:pens}).
+The optical links necessary for sensing can be created either via clear solid cores as in Printed Optics \cite{Willis-printedoptics} or via fiber optic cables threaded through hollow pipes (see Figure \ref{fig:pens}).
 
@@
 
-\subsection{Fabrication Techniques}
-
-\bjoern{The biggest question wrt fabrication for me is - how do I print these cavities? So general support material avoidance or removal strategies should be more prominent.
-
-(I think the lye bath would probably work better than the high pressure water for our objet prints)}
-The physical fabrication process used by different machines gives rise to differing design guidelines.
-The two machines we used to create our example objects were the Objet Connex 260 and the Makerbot Replicator 2.
-
-The Objet is an inkjet-based 3D printer which jets drops of polymer in very fine layers, curing each layer with an ultraviolet lamp.
-It offers a wide range of material types, including clear/opaque, flexible/hard, and black/white.
-Additionally, it permits the mixing of two different materials in a single print: a mixed material could include 70\% flexible and 30\% hard, for example, to be firm but somewhat flexible.
-The flexible materials can be inflated when they are thin: they elongate 200\% before breaking.
-We have found $.05$inches to be a good thickness for inflatable materials.
-Although this machine has many extraordinary capabilities, it unfortunately cannot print overhangs of greater than $14^{\circ}$ without support material.
-The support structure used by this machine is a loose matrix of material that can be removed by hand from the surface or blasted out with a water jet from the interior.
-The support is also soluble in lye.
-
-The Makerbot Replicator 2 is a fused-deposition modeling (FDM) printer, in which we use exclusively Polylactic Acid (PLA) bioplastics.
-More material properties are being explored in PLA: there are now transparent and flexible materials.
-They cannot be blended like those in the Objet.
-The Makerbot has a fairly coarse resolution compared to the Objet, however it permits overhangs up to $45^{\circ}$ without support and also allows bridging (short spans of material that bridge the gap between two structures with no support underneath).
-Due to the FDM process it can even recover from greater overhangs and longer bridges, as the material builds up and can reorganize into a structure \valkyrie{how do I say this?
-do I need a photo?
-print recovery is pretty awesome-looking.}.
-The support used by the Makerbot is made of the same material as the model itself, just printed in a different pattern and attached over as small a surface area as possible to ease removal.
-Support inside of a pipe may require pliers, or be infeasible to remove dependent upon geometry.
-
-Assembly is easiest on both machines when the model does not require support material.
-Models can be ``cut up'' \bjoern{This negates some of the benefits compared to other manufacturing techniques.} into assembleable pieces to evade support requirements or ease material removal.
-We believe that an automatic tool for this task would be highly beneficial, however it is outside the scope of this work.
-
-\subsection{Identity}
-Pipes included in objects can have additional effects beyond their designed intentions.
-
-Pipes included in printed objects allow for their identification.
+\subsubsection{Identification}
+In addition to active inputs from users, pipes can be used for passive \emph{identification}.
 While the pipes need not be visible (especially in the case of fully-enclosed pipes), their presence, location, network structure, and length change the physical properties of a printed object.
@@
 While their technique requires access a terahertz imaging tool, audio resonance identity encoding requires only a microphone and a speaker.
+
+\subsection{Fabrication Techniques}
+
+\bjoern{The biggest question wrt fabrication for me is - how do I print these cavities? So general support material avoidance or removal strategies should be more prominent.)}
+
+The physical fabrication process used by different machines gives rise to differing strategies for avoiding support material or easing its removal.
+The two machines we used to create our example objects were the Objet Connex 260 and the Makerbot Replicator 2.
+
+The Objet is an inkjet-based 3D printer which jets drops of polymer in very fine layers, curing each layer with an ultraviolet lamp.
+It offers a wide range of material types, including clear/opaque, flexible/hard, and black/white.
+Additionally, it permits the mixing of two different materials in a single print: a mixed material could include 70\% flexible and 30\% hard, for example, to be firm but somewhat flexible.
+The flexible materials can be inflated when they are thin: they elongate 200\% before breaking.
+We have found $.05$ inches to be a good thickness for inflatable materials.
+Although this machine has many extraordinary capabilities, it unfortunately cannot print overhangs of greater than $14^{\circ}$ without support material.
+Due to this, support avoidance is difficult; our main tactic for removing support material has been to make the \emph{openings} to pipes as large as possible.
+This allows us to  \valkyrie{you are here!}  The support structure used by this machine is a loose matrix of material that can be removed by hand from the surface, blasted out with a water jet from large cavities, or dissolved in lye (the enclosing model material does not dissolve).
+
+The Makerbot Replicator 2 is a fused-deposition modeling (FDM) printer, in which we use exclusively Polylactic Acid (PLA) bioplastics.
+More material properties are being explored in PLA: there are now transparent and flexible materials.
+They cannot be blended like those in the Objet.
+The Makerbot has a fairly coarse resolution compared to the Objet, however it permits overhangs up to $45^{\circ}$ without support and also allows bridging (short spans of material that bridge the gap between two structures with no support underneath).
+Due to the FDM process it can even recover from greater overhangs and longer bridges, as the material builds up and can reorganize into a structure \valkyrie{how do I say this?
+do I need a photo?
+print recovery is pretty awesome-looking.}.
+The support used by the Makerbot is made of the same material as the model itself, just printed in a different pattern and attached over as small a surface area as possible to ease removal.
+Support inside of a pipe may require pliers, or be infeasible to remove dependent upon geometry.
+
+Assembly is easiest on both machines when the model does not require support material.
+Models can be ``cut up'' \bjoern{This negates some of the benefits compared to other manufacturing techniques.} into assembleable pieces to evade support requirements or ease material removal.
+We believe that an automatic tool for this task would be highly beneficial, however it is outside the scope of this work.

### 2014-03-31 paper/introduction.tex
paper work, adding tovi comments

@@
 
+\tovi{condense}
 Makers, as well as professional designers, leverage 3D printers as tools for design work.
@@
 
+\tovi{Im not sure this emphasis on
+redirection/centralization is
+that important (probably
+does require this entire
+paragraph)}
 The use of pipes can allow for \emph{redirection} and \emph{centralization} of active components to various locations on a printed object's surface.
@@
 Pipes for applications like neon signs must have an Euler tour through them: this allows a single EL wire to pass through every edge on the pipe graph.
+No tools yet exist for this design task.
 We have developed techniques for minimizing the bending energy of, enforcing non-intersection with, and avoiding the surface for pipes, as well as expanding user-specified paths and generating Euler tours.

### 2014-03-31 paper/relatedwork.tex
notes from tovi
paper work, adding tovi comments

@@
 \section{Related Work}
+
+\tovi{make sure to have mini intros and conclusions for each subsection.
+it's ok to discuss several papers at once if they are similar.
+each paper needs a sentence to say what distinguishes us from it}
 
@@
 Willis, et al., also did not offer a design tool or techniques for creating light pipes automatically.
+Slyper, et al., in \cite{Slyper-pressure} 3D printed flexible robot armatures capable of sensing interaction via changes in air pressure.
+This technique is complementary to ours, as we allow for placing these interactive elements anywhere on an object's surface while they had to place them in locations accessible for electronics insertion.
+Acoustic Barcodes explored integrating ridges into digitally fabricated objects, whose acoustic signatures when scraped provided an ID code \cite{Harrison-acoustic}; while pipes can be used for identification and to create aural feedback, this is not our only goal.
 Sauron investigated the use of computer vision to track interactions with physical mechanisms \cite{Savage-sauron}.
 While Savage, et al., used mirrors for redirection, we believe that the required line-of-sight operation for their computer vision techniques may be possible using printed optics or inserted optical fibers.
-Slyper, et al., \cite{Slyper-pressure} 3D printed flexible robot armatures capable of sensing interaction via changes in air pressure.
-This technique is complementary to ours, as we allow for placing these interactive elements anywhere on an object's surface while they had to place them in locations accessible for electronics insertion.
 General-purpose routing for connecting 2D capacitive sensors to terminals is discussed in Midas \cite{Savage-midas}.
 This technique, based on circuitboard routing, are limited to planar and unfoldable shapes, and are intended for use on the surface of objects rather than through their cores.
-Acoustic Barcodes explored integrating ridges into digitally fabricated objects, whose acoustic signatures when scraped provided an ID code \cite{Harrison-acoustic}.
 Fabrickation works towards rapid fabrication of functional objects by integrating construction kit building blocks \cite{Mueller-fabrickation}.
+While we recognize that 3D printing requires a significant amount of time, we believe that creating precise internal structures in 3D printed objects is a strong case for digital fabrication over manual assembly or traditional methods.
 
@@
 In Make it Stand, material is removed from the interior of objects with the goal of changing their center of gravity \cite{Prevost-makeitstand}.
+The creation of such hollow chambers is a part of our design space, however our contribution is orthogonal as we do not focus on re-balancing objects using these chambers.
 Bickel, et al., explored enforcing deformation behavior with material blending \cite{Bickel-deformation}, while B\"{a}cher, et al., built an algorithm for automatically placing 3D fabricatable joints into meshes \cite{Bacher-articulated}.
+Our focus is not on articulated or deformable 3D prints, although we do perform interior modeling.
+Takayama, et al., built a system described in \cite{Takayama-fruit} which allowed users to create internal color maps of the interior of objects, allowing them to see ``slices'' of their ``oranges.''  This differs from our contribution, as our pipes are designed for fabrication, and focus on function rather than appearance.
 Sauron automatically modified the interior of existing mechanical objects to make them amenable to vision-based sensing \cite{Savage-sauron}, and Infrastructs created encoded interior cavities in 3D printed objects designed to be sensed with a terahertz imaging tool.

### 2014-03-31 paper/abstract.tex
paper work, adding tovi comments

@@
 \begin{abstract}
+\tovi{can be condensed}
 3D printers offer extraordinary flexibilty in prototyping the shape and mechanical function of objects.
@@
 We propose a general technique for adding interactivity by removing interior material to form pipes and cavities.
-These can be used to sense input and generate haptic as well as visible output.
 We describe the design space of pipes and hollow chambers for interaction design: a variety of openings, topologies, and inserted media for pipes create diverse inputs and outputs.

### 2014-03-31 paper/designspace.tex
paper work, adding tovi comments

@@
 
+\tovi{this seems like a forced and
+unnecessary definition. Do we really
+need to distibguish the two ends?}
 Pipes can come in four types: open, return, semi-closed, and fully enclosed.
@@
 
-\begin{figure}[h!]
+\begin{figure}[h]
 \centering
@@
 \subsection{Design of Pipes}
-
+\tovi{This definition is a little bit vague.}
 Pipe designs may emphasize either their exterior features (connection points) or their interior paths.

### 2014-03-31 paper/exampleobjects.tex
paper work, adding tovi comments

@@
 
-\valkyrie{one way of combatting the enormousness of this paper is to put all these example objects figures into a single figure* instead}
+\begin{figure*}[h!]
+\centering
+    \includegraphics[width=7in]{figures/examples.png}
+\caption{A series of example objects created using our system.
+A) shows a portable radio.
+B) shows a presence-aware pen holder.
+C) is a rabbit which ``breathes'' using wind pipes we built.
+D) shows a maze game we created.
+E) is a touch-sensitive brain toy.
+\valkyrie{finalize order of example objects.
+what order is best?}}
+\label{fig:examples}
+\end{figure*}
 
@@
 \subsection{Touch-sensitive Toys (open, liquid, star, exterior)}
-
-\begin{figure}[h]
-\centering
-    \includegraphics[width=3.4in]{figures/placeholder/brain.jpg}
-\caption{A touch-sensitive brain whose tubes are filled with conductive paint.
-Sensing is done on a single wire via SFCS.
-Inset shows the internal structure of the tubes generated by our design tool.}
-\label{fig:toys}
-\end{figure}
 
@@
 
-\begin{figure}[h]
-\centering
-    \includegraphics[width=3.4in]{figures/placeholder/bunny.jpg}
-\caption{This rabbit has an open tube leading to its mouth and a semi-closed tube capped with flexible material in its back.
-When air is sucked in through the open tube, the back becomes stiff as though the rabbit has breathed in.
-Inset shows the interior structure.}
-\label{fig:breathe}
-\end{figure}
-
 We created a rabbit with a pair of tubes that can simulate breathing (see Figure \ref{fig:breathe}).
@@
 
-\begin{figure}[h]
-\centering
-    \includegraphics[width=3.4in]{figures/radio.png}
-\caption{This radio is assembled from traditional electronic components connected by copper-filled tubes.
-The case was designed to allow the components to recess into it slightly.
-Inset shows the internal structure of the tubes generated by our design tool.}
-\label{fig:radio}
-\end{figure}
-
 A custom radio built using tubes allows users to tune to different stations and listen in.
@@
 \subsection{Presence-aware Pen Holder (open, solid, exterior)}
-
-\begin{figure}[h]
-\centering
-    \includegraphics[width=3.4in]{figures/placeholder/pens.jpg}
-\caption{This pen holder a modified FlyEye technique to sense the presence or absence of an object in each of its tubes.  Inset shows the internal structure of the tubes generated by our design tool.}
-\label{fig:pens}
-\end{figure}
 
@@
 \subsection{Maze (fully enclosed, particulate, tree, interior)}
-
-\begin{figure}[h]
-\centering
-    \includegraphics[width=3.4in]{figures/placeholder/maze.jpg}
-\caption{Our maze game has a large single particle trapped inside a fully enclosed tube.  It offers hours of fun!}
-\label{fig:maze}
-\end{figure}
 

### 2014-03-31 paper/tubes.tex
paper work, adding tovi comments

@@
 \subsubsection{Routing}
+\tovi{Still too long....}
 In graph theory, a semi-Eulerian graph is one which has exactly two edges of odd degree.
@@
 \end{figure}
+
+\subsection{Fabrication Techniques}
+
+The physical fabrication process used by different machines gives rise to differing strategies for avoiding support material or easing its removal.
+The two machines we used to create our example objects were the Objet Connex 260 and the Makerbot Replicator 2.
+
+The Objet is an inkjet-based 3D printer which jets drops of polymer in very fine layers, curing each layer with an ultraviolet lamp.
+It offers a wide range of material types, including clear/opaque, flexible/hard, and black/white.
+Additionally, it permits the mixing of two different materials in a single print: a mixed material could include 70\% flexible and 30\% hard, for example, to be firm but somewhat flexible.
+The flexible materials can be inflated when they are thin: they elongate 200\% before breaking.
+We have found $.05$ inches to be a good thickness for inflatable materials.
+Although this machine has many extraordinary capabilities, it unfortunately cannot print overhangs of greater than $14^{\circ}$ without support material.
+Due to this, support avoidance is difficult; our main tactic for removing support material has been to make the \emph{openings} to pipes as large as possible.
+This allows us to flush the material most easily.
+The support structure used by this machine is a loose matrix of material that can be removed by hand from the surface, blasted out with a water jet from cavities, or dissolved in lye (the enclosing model material does not dissolve).
+
+The Makerbot Replicator 2 is a fused-deposition modeling (FDM) printer, in which we use exclusively Polylactic Acid (PLA) bioplastics.
+More material properties are being explored in PLA: there are now transparent and flexible materials.
+They cannot be blended like those in the Objet.
+The Makerbot has a fairly coarse resolution compared to the Objet, however it permits overhangs up to $45^{\circ}$ without support and also allows bridging (short spans of material that bridge the gap between two structures with no support underneath).
+Although spheres and sideways cylinders have a $>45^{\circ}$ overhang in some places, they are also printable without support.
+Due to the FDM process it can even recover from greater overhangs and longer bridges, as the material builds up and can reorganize into a structure \valkyrie{how do I say this?
+do I need a photo?
+print recovery is pretty awesome-looking.}.
+The support used by the Makerbot is made of the same material as the model itself, just printed in a different pattern and attached over as small a surface area as possible to ease removal.
+Support inside of a pipe may require pliers, or be infeasible to remove dependent upon geometry.
+Meshmixer has a built-in tool for minimizing support for prints; it re-orients the part to keep as much surface area as possible within the 45 degree constraint.
+In future work we would like to explore a tool in which a user can select surfaces to prioritize support avoidance for.
+
+Assembly is easiest on both machines when the model does not require support material.
+For especially complex geometries, models can be ``cut up'' \bjoern{This negates some of the benefits compared to other manufacturing techniques.} into assembleable pieces that evade support requirements or ease material removal.
+We believe that an automatic tool for this task would be highly beneficial, however it is outside the scope of this work.

### 2014-03-31 paper/discussion.tex
paper work, adding tovi comments

@@
 While their technique requires access a terahertz imaging tool, audio resonance identity encoding requires only a microphone and a speaker.
-
-\subsection{Fabrication Techniques}
-
-\bjoern{The biggest question wrt fabrication for me is - how do I print these cavities? So general support material avoidance or removal strategies should be more prominent.)}
-
-The physical fabrication process used by different machines gives rise to differing strategies for avoiding support material or easing its removal.
-The two machines we used to create our example objects were the Objet Connex 260 and the Makerbot Replicator 2.
-
-The Objet is an inkjet-based 3D printer which jets drops of polymer in very fine layers, curing each layer with an ultraviolet lamp.
-It offers a wide range of material types, including clear/opaque, flexible/hard, and black/white.
-Additionally, it permits the mixing of two different materials in a single print: a mixed material could include 70\% flexible and 30\% hard, for example, to be firm but somewhat flexible.
-The flexible materials can be inflated when they are thin: they elongate 200\% before breaking.
-We have found $.05$ inches to be a good thickness for inflatable materials.
-Although this machine has many extraordinary capabilities, it unfortunately cannot print overhangs of greater than $14^{\circ}$ without support material.
-Due to this, support avoidance is difficult; our main tactic for removing support material has been to make the \emph{openings} to pipes as large as possible.
-This allows us to  \valkyrie{you are here!}  The support structure used by this machine is a loose matrix of material that can be removed by hand from the surface, blasted out with a water jet from large cavities, or dissolved in lye (the enclosing model material does not dissolve).
-
-The Makerbot Replicator 2 is a fused-deposition modeling (FDM) printer, in which we use exclusively Polylactic Acid (PLA) bioplastics.
-More material properties are being explored in PLA: there are now transparent and flexible materials.
-They cannot be blended like those in the Objet.
-The Makerbot has a fairly coarse resolution compared to the Objet, however it permits overhangs up to $45^{\circ}$ without support and also allows bridging (short spans of material that bridge the gap between two structures with no support underneath).
-Due to the FDM process it can even recover from greater overhangs and longer bridges, as the material builds up and can reorganize into a structure \valkyrie{how do I say this?
-do I need a photo?
-print recovery is pretty awesome-looking.}.
-The support used by the Makerbot is made of the same material as the model itself, just printed in a different pattern and attached over as small a surface area as possible to ease removal.
-Support inside of a pipe may require pliers, or be infeasible to remove dependent upon geometry.
-
-Assembly is easiest on both machines when the model does not require support material.
-Models can be ``cut up'' \bjoern{This negates some of the benefits compared to other manufacturing techniques.} into assembleable pieces to evade support requirements or ease material removal.
-We believe that an automatic tool for this task would be highly beneficial, however it is outside the scope of this work.

### 2014-03-31 paper/tubes.tex
put in bjoern comments

@@
 
-As mentioned, pipe design is not trivial.
+Pipe design is not trivial.
 Independent pipes cannot intersect lest their contents be unintentionally mixed.
@@
 
+\bjoern{Hits section mentions a* three times, ech time with a little more detail than before.
+An probably cut this to two times as a general strategy: once in the overview, then a second time in detail.}
 Designing objects which, for example, are touch sensitive in particular areas or need to integrate with existing electronics, requires precise location and sizing of pipe endpoints.
@@
 This wind keeps the resulting pipe away from the surface.
-\valkyrie{Ryan: How does it actually work?}
+\valkyrie{Ryan: How does it actually work?} \bjoern{Need much more detail here or a reference to a previously published technique}
 
@@
 \valkyrie{this section can probably be scrapped, along with Figure \ref{fig:meshmixer-endpoint}.}
+\bjoern{Replace with a single sentence: combining pipe routing with other existing mesh modification operations like extrusions enables users to create custom ...}
 Although start and end radius of pipes is configurable, our tool does not explicitly allow makers to design the \emph{shape} of their pipes' endpoints.
@@
 
+\bjoern{Replace specific with more general category: pipes where interior path layout matters}
 Neon signs, as in Figure \ref{fig:tool-process-interior}(d), require a single long path that passes through many desired edges: for example, in our UIST sign, the path must pass through all the edges that comprise the word ``UIST''.
@@
 \tovi{Still too long....}
+\bjoern{Need this explanation above so readers don't wonder what Euler tours are
+
+Agree with Tovi that this can be condensed}
 In graph theory, a semi-Eulerian graph is one which has exactly two edges of odd degree.
@@
 These two voxel grids are subtracted, and the result is re-meshed and smoothed.
-When a user desires to cap the ends of her pipes, we create a sphere of the same radius as each endpoint whose center is the same as the endpoint.
-We intersect this sphere with the original mesh, then intersect this partial-sphere mesh with a shell of the original mesh.
-This gives us a voxel subgrid of the thickness of cap we desire (the same as the shell's thickness) which is located only in the region of the tube end.
+\bjoern{ rewrite When a user desires to cap the ends of her pipes, we create a sphere of the same radius as each endpoint whose center is the same as the endpoint.
+We intersect this sphere with the original mesh, then intersect this partial-sphere mesh with a shell of the original mesh.}  This gives us a voxel subgrid of the thickness of cap we desire (the same as the shell's thickness) which is located only in the region of the tube end.
 During subtraction, we preserve this subgrid.
@@
 Unfortunately, because we use voxelization as a part of our subtraction operation, our models can accumulate error when many pipes are created in sequence (see Figure \ref{fig:voxelize}).
-This problem could be solved by performing a multi-pipe operation; fewer rounds of voxelization and re-meshing lead to lower error.
-This multi-pipe operation would also lead to a more globally optimal pipe routing.
+This problem could be solved with a higher voxel resolution, but it requires balancing computation time and precision.
 
@@
     \includegraphics[width=3.4in]{figures/voxelize-fail.png}
-\caption{After repeated pipe creation, the mesh containing these pipes, which were not originally intersecting, accumulated sufficient error to open the pipes into each other.}
+\caption{After repeated pipe creation, the mesh containing these pipes, which were not originally intersecting, accumulated sufficient error to open the pipes into each other.
+\bjoern{I don't understand this figure- probably requires multiple parts - ideal geometry and actual geometry}}
 \label{fig:voxelize}
@@
 
+\bjoern{Argue first that we need processes to fabricate these hollow pipes- can do this by removing support material or never printing such material in the first place.}  \bjoern{Is section needs to bee written from a general standpoint first, and then our specific details second.}
 The physical fabrication process used by different machines gives rise to differing strategies for avoiding support material or easing its removal.

### 2014-03-31 paper/relatedwork.tex
put in bjoern comments

@@
 
-Our contributions relate to other efforts to add interactivity to digitally fabricated objects; and to techniques for routing and mode modification.
+Our contributions relate to other efforts to add interactivity to digitally fabricated objects; and to techniques for routing and model modification.
 \bjoern{The related work section is taking shape but still has a bit of a "missing the forest for the trees" feeling.)}
@@
 Previous work has investigated the integration of interactivity with the fabrication process.
-Printed Optics (\cite{Willis-printedoptics}) used light pipes, solid clear tubes, to create integrated displays: our work is inspired by this and goes towards a design tool for these capabilities; Printed Optics is a part of our design space, which includes other uses of pipes.
-Willis, et al., also did not offer a design tool or techniques for creating light pipes automatically.
+Printed Optics (\cite{Willis-printedoptics}) used light pipes, solid clear tubes, to create integrated displays and touch sensors: our work is inspired by this and goes towards a design tool for these capabilities.
 Slyper, et al., in \cite{Slyper-pressure} 3D printed flexible robot armatures capable of sensing interaction via changes in air pressure.
@@
 
+\bjoern{May be able to condense this entire subsection into ~10 lines}
 Other research has attempted to bring traditional electronics to 3D prints, either on the surface or via interior channels.
@@
 
-Circuitboard routing algorithms, such as Lee's maze router algorithm, work only on a 2D surface.
-So-called ``3D circuitboards'' are typically 2.5D, in that they comprise a stack of 2D circuitboards connected by vias \valkyrie{citation needed}.
+Circuitboard routing algorithms, such as Lee's maze router algorithm \bjoern{ref?}, work only on a 2D surface.
+Multi-layer boards are typically 2.5D, in that they comprise a stack of 2D circuitboards connected by vias \valkyrie{citation needed}.
+Instead we offer true 3d point to point routing though the interior of solid models.
 
@@
 Prior work used tours of nodes, rather than edges, and cast continuous line drawing as a traveling salesman problem~\cite{Bosch-tsp}.
-Schweikardt, et al., also made a connection between electroluminescent wire and graph theory, though their contribution is orthogonal to ours: they provide a tangible construction kit with smart nodes for building physical graph examples in which edges are lit by EL wire \cite{Schweikardt-tei09}.
-In traditional neon sign making, ``routing'' is a conversation between a graphic designer and a glass bender on what is feasible and comfortable for the latter and what will help him avoid burning his hands, wasting glass, and save bending time \cite{strattman1997neon}.
+In traditional neon sign making, ``routing'' is a conversation between a graphic designer and a glass bender on what is feasible and comfortable for the latter and what will help him avoid burning his hands and wasting glass or bending time \cite{strattman1997neon}.
 
-Prior work on modeling the interior of objects (as opposed to their exterior) has been limited.
+There is prior work on modeling the interior of objects (as opposed to their exterior).
 In Make it Stand, material is removed from the interior of objects with the goal of changing their center of gravity \cite{Prevost-makeitstand}.
@@
 Our work differs in that we want to allow users to design pipes and integrate interactive objects with them.
+\bjoern{Can condense this - they all make orthogonal contributions, so don't need to repeat this statement of differentiation n times.
+Just do it once at the end.}

### 2014-03-31 paper/designspace.tex
put in bjoern comments

@@
 \section{The Design Space of Pipes}
-The design space of pipes is large.
+\bjoern{Budget at most 1.5 columns for this section
+
+ There still some redundancy with the related work section, too}
 By using different types and topologies of pipes and inserting different media into the completed pipes, a variety of inputs and outputs can be located at arbitrary points on the surface of a 3D print.
@@
 C shows two semi-closed pipes with rubber membranes at their ends.
-In D, a rabbit has a fully enclosed chamber to modify its weight.}
+In D, a rabbit has a fully enclosed chamber to modify its weight.
+\bjoern{Great, but make figure  full column wide}}
 \label{fig:pipespace}
@@
 Splitting can also be useful, for example if our maker wants red paint output in two locations from her painting device, she could have one system-side pipe, but split the pipe into two (see Figure \ref{fig:pipespace}).
+\bjoern{??}
 

### 2014-03-31 paper/introduction.tex

@@
 \caption{A novel neon sign (a) and a touch-sensitive brain (b) designed with our pipe tool and fabricated on a consumer-grade 3D printer.
-(c) and (d) show the structure of the internal pipes generated by our tool.}
+(c) and (d) show the structure of the internal pipes generated by our tool.
+\bjoern{will this be the robot or the UIST sign?}}
 \label{fig:teaser}
@@
 paragraph)}
+\bjoern{Maybe focus on this on expanding the utile and applicability of sensing techniques recently introduced in hci into types of objects not initially intended for.}
 The use of pipes can allow for \emph{redirection} and \emph{centralization} of active components to various locations on a printed object's surface.
@@
 In this way, she has only a single active element (the microcontroller setup), but her otherwise-inert 3D printed object (the brain) has several active sensing locations.
-This can also be used for rapid iteration of designs.
-Our maker need only add a digital tube, reprint her brain, and re-fill it with copper paint to have a new version; no complicated electromechanical assembly required.
 
@@
 
+\bjoern{A bit repetitive with col 2 of page 1. There's opportunity for consolidation and shortening.}
 Many existing sensing and actuation approaches, such as FlyEye \cite{Wimmer-flyeye}, Jamming User Interfaces \cite{Follmer-jamming}, and Touch\'{e} \cite{Sato-touche}, can be enhanced by the use of pipes.
@@
 \item We showcase a set of examples, enabled by our modeling tool and exploring new points in the design space.
-\item We describe the interaction possibilities of pipes, including prior work as well as uncharted possibilities.
+\item We describe the interaction possibilities of pipes, including prior work as well as uncharted possibilities. \bjoern{This last point sounds a t like the first.}
 \end{itemize}

### 2014-04-01 paper/relatedwork.tex
shortening

@@
 \section{Related Work}
-
-\tovi{make sure to have mini intros and conclusions for each subsection.
-it's ok to discuss several papers at once if they are similar.
-each paper needs a sentence to say what distinguishes us from it}
-
 Our contributions relate to other efforts to add interactivity to digitally fabricated objects; and to techniques for routing and model modification.
-\bjoern{The related work section is taking shape but still has a bit of a "missing the forest for the trees" feeling.)}
-
-\bjoern{Tactically, to communicate who would be is a qualified reviewer for this paper in the UIST community, we should make sure to have appropriate references to Baudisch, Poupyrev, Harrison in here}
-
-\valkyrie{can you offer suggestions?
-I can imagine the latex buttons paper, touche, acoustic barcodes, capacitive fingerprinting \cite{Harrison-capfingerprint}, printed optics for harrison/poupyrev, and for baudisch maybe....
-TDR \cite{Wimmer-TDR}?
-fiberio \cite{Holz-fiberio}?
-fabrickation?
-I'm not sure on this.}
 
@@
 
-\bjoern{May be able to condense this entire subsection into ~10 lines}
-Other research has attempted to bring traditional electronics to 3D prints, either on the surface or via interior channels.
-Sells, et al., in \cite{Sells-reprap} create channels on the surface of a 3D printed object which are filled with a special computer-controlled syringe of liquid solder to create connections.
-Their focus is on creating self-replicating 3D printers, and their electronics are confined to traditional 2D routed circuitboards.
+Other research has brought traditional electronics to 3D prints.
+Sells, et al., \cite{Sells-reprap}, describe using a heated syringe to deposit metal in printed 2D circuitboard channels on objects' surfaces.
 Sarik, et al., \cite{Sarik-tracebrush} \valkyrie{I never found a real paper of this.
 The Microsoft website has a "print" of it with TEI boilerplate, but I couldn't find it in the digital library.
-What's up with that?} and the Optomec company \cite{optomec} both created techniques for taking an already-printed object and spraying conductive material onto its surface.
-In contrast, our technique can lead to more durable circuitry, since it is routed through the core of the object rather than on its surface.
-Navarette, et al., in \cite{Navarrette-gps} created a functioning GPS whose circuit routing extended into true 3D (as compared to the 2.5D boards created by stacking traditional 2D circuitboards).
-However, they created this one-off routing by hand and do not offer any general algorithms or techniques.
-In the area of robotics, softer-than-skin interactive surfaces have been created using tubed interfaces fabricated from silicone molds.
-\cite{Park-microchannels} and \cite{Majidi-curvature} use microchannels filled with liquid metal for pressure or curvature sensing: the metal in the channels changes resistance as the channels are stretched.
-These techniques have not been used with fabrication, nor has a design tool to create such sensors been previously described.
+What's up with that?} and the Optomec company \cite{optomec} describe \emph{spraying} conductive material onto objects to create circuitry.
+Injection of liquid metals for sensing in softer-than-skin electronics was explored by Park, et al., \cite{Park-microchannels} and Majidi, et al., \cite{Majidi-curvature}.
+Navarette, et al., created a functioning GPS routed in true 3D (not multilayer boards' 2.5D), hand-built using a one-off routing.
+Our work differs from these because we offer \emph{algorithms and techniques} for \emph{internal}, \emph{true 3D} routing.
 
@@
 
-Circuitboard routing algorithms, such as Lee's maze router algorithm \bjoern{ref?}, work only on a 2D surface.
-Multi-layer boards are typically 2.5D, in that they comprise a stack of 2D circuitboards connected by vias \valkyrie{citation needed}.
+Circuitboard routing algorithms, such as Lee's maze router algorithm \cite{Lee-maze}, work only on a 2D surface.
+Multi-layer boards are typically 2.5D, in that they comprise a stack of 2D circuitboards connected by vias.
 Instead we offer true 3d point to point routing though the interior of solid models.
@@
 In Make it Stand, material is removed from the interior of objects with the goal of changing their center of gravity \cite{Prevost-makeitstand}.
-The creation of such hollow chambers is a part of our design space, however our contribution is orthogonal as we do not focus on re-balancing objects using these chambers.
 Bickel, et al., explored enforcing deformation behavior with material blending \cite{Bickel-deformation}, while B\"{a}cher, et al., built an algorithm for automatically placing 3D fabricatable joints into meshes \cite{Bacher-articulated}.
-Our focus is not on articulated or deformable 3D prints, although we do perform interior modeling.
-Takayama, et al., built a system described in \cite{Takayama-fruit} which allowed users to create internal color maps of the interior of objects, allowing them to see ``slices'' of their ``oranges.''  This differs from our contribution, as our pipes are designed for fabrication, and focus on function rather than appearance.
-Sauron automatically modified the interior of existing mechanical objects to make them amenable to vision-based sensing \cite{Savage-sauron}, and Infrastructs created encoded interior cavities in 3D printed objects designed to be sensed with a terahertz imaging tool.
-Our work differs in that we want to allow users to design pipes and integrate interactive objects with them.
-\bjoern{Can condense this - they all make orthogonal contributions, so don't need to repeat this statement of differentiation n times.
-Just do it once at the end.}
+Takayama, et al., built a system described in \cite{Takayama-fruit} which allowed users to create internal color maps of the interior of objects, allowing them to see digital ``slices'' of their ``oranges.''  Sauron automatically modified the interior of existing mechanical objects to make them amenable to vision-based sensing \cite{Savage-sauron}, and Infrastructs created encoded interior cavities in 3D printed objects designed to be sensed with a terahertz imaging tool.
+All these contributions are orthogonal to ours, as our focus is on modeling pipes and hollow chambers for fabricating interactive objects.

### 2014-04-01 paper/introduction.tex
shortening

@@
 (c) and (d) show the structure of the internal pipes generated by our tool.
-\bjoern{will this be the robot or the UIST sign?}}
+\bjoern{will this be the robot or the UIST sign?} \valkyrie{arguably, the robot will be nicer-looking.
+I think I'm manufacturing the UIST sign on the Makerbot.}}
 \label{fig:teaser}
@@
 Makers, as well as professional designers, leverage 3D printers as tools for design work.
-A wide array of objects, ranging from bicycle helmets to jewelry to video game controllers, are now prototyped or even manufactured using these machines.
+A wide array of objects, ranging from bicycle helmets to video game controllers, are now prototyped or even manufactured using these machines.
 However, most devices fabricated by 3D printers are passive.
@@
 These channels introduce an entire design space of opportunities for adding interactivity to these objects.
-For example, copper material can fill the channels, to allow for standard electronic components to be easily integrated after the printout; air can be pumped through to create tactile output; or electroluminescent wire can be threaded through to create computer-controlled visual output.
+For example, copper material can fill the channels, to allow for standard electronic components to be easily integrated after the printout; electroluminescent (EL) wire can be threaded through to create computer-controlled visual output; or air can be pumped through to control haptic responses.
 
-\tovi{Im not sure this emphasis on
-redirection/centralization is
-that important (probably
-does require this entire
-paragraph)}
-\bjoern{Maybe focus on this on expanding the utile and applicability of sensing techniques recently introduced in hci into types of objects not initially intended for.}
-The use of pipes can allow for \emph{redirection} and \emph{centralization} of active components to various locations on a printed object's surface.
+Pipes allow makers to build with and upon existing input/output techniques, often integrating these methods into objects they were not originally designed for.
 For example, to create a touch-sensitive toy as in Figure \ref{fig:teaser}, a maker can select several locations of interest on the brain.
@@
 For sensing, she need only connect a single wire to a shared out of all the tubes, and can distinguish them via Swept Frequency Capacitve Sensing \cite{Sato-touche}.
-In this way, she has only a single active element (the microcontroller setup), but her otherwise-inert 3D printed object (the brain) has several active sensing locations.
+In this way, she has only a single active element (the microcontroller setup), but her otherwise-inert 3D printed object (the brain) has several sensing locations.
+Other existing sensing and actuation approaches, such as FlyEye \cite{Wimmer-flyeye} and Jamming User Interfaces \cite{Follmer-jamming}, can also be enhanced by the use of pipes: pipes allow makers utilizing such I/O strategies to locate input and output at arbitrary points on an object's surface.
+Thus, we leverage these existing techniques for sensing and actuation while our work's novelty is in \emph{internal modeling} for 3D printing and the \emph{redirection} of I/O to arbitrary locations on a device.
 
@@
 Electricity is run through the near-vacuum, and this makes the path visible.
-If our maker wants to build such a sign, she can substitute Electroluminescent (EL) wire for the evacuated glass and electricity: she need only design a path for the wire to be fed through and then create pipes along that path.
+If our maker wants to build such a sign, she can substitute EL wire for the evacuated glass and electricity: she need only design a path for the wire to be fed through and then create pipes along that path.
 
-The generation of these pipes, however, is not trivial.
-Pipes need to have maximum possible bend radius to ease the insertion of media post-print.
-Independent pipes should not intersect, as this could lead to electrical shorting or other problems.
-Pipes should not pass too close to the surface (unless they are being used for display purposes).
-Consumer-grade 3D printers do not guarantee air- or water-tightness of their prints, so more than one layer may be required to prevent fluid leakage.
-Pipes for applications like neon signs must have an Euler tour through them: this allows a single EL wire to pass through every edge on the pipe graph.
-No tools yet exist for this design task.
+Generating these pipes, however, is not trivial.
+Pipes need maximum possible bend radius to ease the insertion of media post-print.
+Independent pipes cannot intersect, as this could lead to electrical shorting or other problems.
+Pipes should not pass too close to the surface: consumer-grade 3D printers do not guarantee air- or water-tightness of prints, so more than one printed layer may be required to prevent leakage.
+Pipes for applications like neon signs must have an Euler tour through them: this allows a single EL wire to light every edge on the pipe graph.
+No tools yet exist for these design tasks.
 We have developed techniques for minimizing the bending energy of, enforcing non-intersection with, and avoiding the surface for pipes, as well as expanding user-specified paths and generating Euler tours.
-These techniques are embodied in our software tool to enable makers to create pipe-powered interfaces.
-
-\bjoern{A bit repetitive with col 2 of page 1. There's opportunity for consolidation and shortening.}
-Many existing sensing and actuation approaches, such as FlyEye \cite{Wimmer-flyeye}, Jamming User Interfaces \cite{Follmer-jamming}, and Touch\'{e} \cite{Sato-touche}, can be enhanced by the use of pipes.
-Pipes allow makers utilizing these sensing and actuation strategies to locate input and output at arbitrary points on an object's surface.
-Thus, we leverage these existing techniques for sensing and actuation while our work's novelty is in \emph{internal modeling} for 3D printing and the \emph{redirection} of IO to arbitrary locations on a device.
+They are embodied in our software tool enabling makers to create pipe-powered interfaces.
 
@@
 \item We showcase a set of examples, enabled by our modeling tool and exploring new points in the design space.
-\item We describe the interaction possibilities of pipes, including prior work as well as uncharted possibilities. \bjoern{This last point sounds a t like the first.}
 \end{itemize}

### 2014-04-01 paper/abstract.tex
shortening

--- 2014-03-24 paper/abstract.tex
@@
 We propose a general technique for adding interactivity by removing interior material to form pipes and cavities.
-These can be used to sense input and generate haptic as well as visible output.
-We describe the design space of pipes and hollow chambers for interaction design: a variety of openings, topologies, and inserted media for pipes create diverse inputs and outputs.
-We present a technique and design tool for routing pipes of various topologies through the interior of 3D printed parts.
-Our design tool is integrated into a 3D model manipulation program.
+We describe the design space of pipes and hollow chambers for interaction design, touching on of openings, topologies, and inserted media.
+We present a technique and design tool for routing pipes through the interior of 3D models.
+We integrated our prototype with a 3D modeling program.
 We use two distinct routing algorithms.
-One, for creating particular points of interaction or integrating with existing electronics, allows users to select begin and end points for pipes, then uses A* path routing and physics-based simulation to minimize the bending energy of routed paths.
-The second, for creating, e.g., novel neon signs, allows users to enter a description of paths to follow: for this we developed a novel routing algorithm and offer proof of key traits.
-We present several totally tubular prototypes we created using this tool to show its flexibility and potential, as well as to explore new points in the pipe design space.
+One, for creating particular points of interaction or integrating with existing electronics on the model's surface, allows users to select pipes' terminals, then uses A* path routing and physics-based simulation to minimize the bending energy of routed paths.
+The second, for creating, e.g., novel neon signs, has users supply a desired internal shape which we fit a pipe route to: for this we developed a novel graph-based routing algorithm.
+We present several totally tubular prototypes we created using our tool to show its flexibility and potential.
 \end{abstract}

### 2014-04-01 paper/designspace.tex
shortening

@@
  There still some redundancy with the related work section, too}
-By using different types and topologies of pipes and inserting different media into the completed pipes, a variety of inputs and outputs can be located at arbitrary points on the surface of a 3D print.
+By using different openings, designs, and topologies of pipes, and inserting different media into the completed pipes, inputs and outputs can be located at arbitrary points on the surface of a 3D print.
 We describe the space below.
@@
     \includegraphics[width=3.4in]{figures/tubespace.pdf}
-\caption{The design space of pipes.  Pipe types, media, topologies, and design are discussed more fully in the text.}
+\caption{The design space of pipes.  Pipe openings, design, topologies, and media are discussed more fully in the text.}
 \label{fig:pipespace}
@@
 need to distibguish the two ends?}
-Pipes can come in four types: open, return, semi-closed, and fully enclosed.
+Pipes can have four opening styles: open, return, semi-closed, and fully enclosed.
 Each of these types offers distinct interaction capabilities (see Figure \ref{fig:pipespace}).
-Since we are interested in interactive devices, we distinguish a ``user side" - part of the surface or interior of an object facing its user; and a ``system side" - part of the object (either interior or exterior) where other hardware such as electronics will be connected.
+As we are interested in interactive devices, we distinguish a ``user side"---part of the surface or interior of an object facing its user; and a ``system side"---where other hardware such as electronics may be connected.
 Definitionally, in the following paragraphs we refer to a pipe which is ``open on the user side'' or ``open on the system side'' to have its contents \emph{physically accessible} to an interacting user or the system \emph{at interaction time}.
-A pipe which is ``open on the user side'' or ``closed on the system side'' is physically cut off from the user or machine.
+Similarly, pipes ``closed on the user side'' or ``closed on the system side'' are physically cut off from the user or machine.
 User and system manipulation of contents are possible with or without physical access, e.g., by resonating, pressurizing, or passing electricity through those contents.
@@
 
-Open pipes (see Figure \ref{fig:pipespace}A) originate from the system side and connect the user side, with both ends of the pipe open.
-This type of pipe may be used to create, for example, capacitive sensors: an open pipe filled with conductive paint can be connected to a sensing platform (e.g., Arduino) on the system side, while a user can touch the uncapped other end of the pipe.
-Using Swept Frequency Capacitive Sensing \cite{Sato-touche} or other techniques, a user's touch of the open end can be sensed.
-An open pipe can also be used for output, for example by creating in-air vortices as in \cite{Sodhi-aireal}.
+Open pipes originate from the system side and connect the user side, with both ends of the pipe open.
+This type of pipe may be used to create electronic circuitry.
+An open pipe filled with conductive paint can power an LED (see Figure \ref{fig:pipespace}A).
 
-A return pipe (see Figure \ref{fig:pipespace}B) originates at the system and returns back to it, without offering physical access to the user.
-By threading an electroluminescent (EL) wire through a clear return pipe, a maker can create a custom piece of neon art.
-If a return pipe passes very close to the surface of a 3D printed model, warm or cold water passed through the pipe could be used for temperature-based haptic feedback.
+A return pipe originates at the system and returns back to it, without offering physical access to the user.
+By threading an electroluminescent (EL) wire through a clear return pipe, a maker can create a custom piece of neon art (see Figure \ref{fig:pipespace}B).
 
-Semi-closed pipes (see Figure \ref{fig:pipespace}C) are open at the system end (for control of the enclosed medium) and closed at the user end.
-The closed interface  may be fabricated from a malleable material: for example, a series of pipes terminated in thin rubber membranes on the user side can be actuated by an air pump to create haptic feedback, or a hinged cap can be blown open by air pressure.
-Semi-closed pipes can also function as audio-generating resonance chambers.
+Semi-closed pipes are open at the system end and closed at the user end.
+The closed interface may be fabricated from a malleable material: for example, a series of pipes terminated in thin rubber membranes on the user side can be actuated by an air pump to create haptic feedback (see Figure \ref{fig:pipespace}C).
 
-A fully enclosed pipe (see Figure \ref{fig:pipespace}D) is closed to both the system side and the user side.
-Fully enclosed pipes can be used as resonance chambers (e.g., for object identification), or as air bubbles (e.g., as used in Printed Optics (\cite{Willis-printedoptics}) for internal display).
+A fully enclosed pipe is closed to both the system and the user.
+Fully enclosed pipes may be used as weight modifiers (e.g., for object identification), as in Figure \ref{fig:pipespace}D).
 They can also be used as containers for water or other particles that would otherwise fall out.
@@
 \subsection{Design of Pipes}
-\tovi{This definition is a little bit vague.}
-Pipe designs may emphasize either their exterior features (connection points) or their interior paths.
-These two features lead to different kinds of interfaces.
-An example interface that focuses on the exterior connection points is the touch sensitive toy in Figure \ref{fig:toys}, where pipes must exit the toy at the frontal lobe, parietal lobes, etc.
-An interface focused on the internal path of the pipe is the neon sign in Figure \ref{fig:neon}: output is based on the shape of the pipe.
+Pipe design may emphasize either exterior connection points or interior paths.
+These two foci lead to different kinds of interfaces.
+An example focused on the exterior features is the touch sensitive toy in Figure \ref{fig:toys}, where pipes must exit the toy at the frontal lobe and parietal lobes to make those \emph{specific locations} touch-sensitive.
+Focused on the internal features is the neon sign in Figure \ref{fig:neon}: the sign's output is based on the \emph{path} of the pipe.
 
@@
 
-Pipe network topologies enable different types of interactions.
+Pipe network topologies enable different interactions.
 

### 2014-04-01 paper/designspace.tex
shortening

@@
 \section{The Design Space of Pipes}
-\bjoern{Budget at most 1.5 columns for this section
-
- There still some redundancy with the related work section, too}
 By using different openings, designs, and topologies of pipes, and inserting different media into the completed pipes, inputs and outputs can be located at arbitrary points on the surface of a 3D print.
@@
 
-\tovi{this seems like a forced and
-unnecessary definition. Do we really
-need to distibguish the two ends?}
-Pipes can have four opening styles: open, return, semi-closed, and fully enclosed.
-Each of these types offers distinct interaction capabilities (see Figure \ref{fig:pipespace}).
+Pipes can have four opening styles: open, return, semi-closed, and fully enclosed (see Figure \ref{fig:pipespace}).
 As we are interested in interactive devices, we distinguish a ``user side"---part of the surface or interior of an object facing its user; and a ``system side"---where other hardware such as electronics may be connected.
-Definitionally, in the following paragraphs we refer to a pipe which is ``open on the user side'' or ``open on the system side'' to have its contents \emph{physically accessible} to an interacting user or the system \emph{at interaction time}.
-Similarly, pipes ``closed on the user side'' or ``closed on the system side'' are physically cut off from the user or machine.
-User and system manipulation of contents are possible with or without physical access, e.g., by resonating, pressurizing, or passing electricity through those contents.
+We refer to a pipe which is ``open on the user side'' to have its contents \emph{physically accessible} to a user \emph{at interaction time}.
+``Closed on the user side'' implies physically cut off from the user.
+Similarly with the system side.
+User or system manipulation of pipe contents are possible even without physical access, e.g., via soundwaves.
 
@@
     \includegraphics[width=3.4in]{figures/types.png}
-\caption{Types of pipes.
-A shows a pair of open tubes filled with copper paint to power an LED.
+\caption{Pipe openings: A shows a pair of open tubes filled with copper paint to power an LED.
 B shows a return pipe with EL wire threaded through it.
-C shows two semi-closed pipes with rubber membranes at their ends.
+C shows two semi-closed pipes capped by rubber membranes.
 In D, a rabbit has a fully enclosed chamber to modify its weight.
 \bjoern{Great, but make figure  full column wide}}
-\label{fig:pipespace}
+\label{fig:openings}
 \end{figure}
@@
 This type of pipe may be used to create electronic circuitry.
-An open pipe filled with conductive paint can power an LED (see Figure \ref{fig:pipespace}A).
+An open pipe filled with conductive paint can power an LED (see Figure \ref{fig:openings}A).
 
 A return pipe originates at the system and returns back to it, without offering physical access to the user.
-By threading an electroluminescent (EL) wire through a clear return pipe, a maker can create a custom piece of neon art (see Figure \ref{fig:pipespace}B).
+By threading an electroluminescent (EL) wire through a clear return pipe, a maker can create a custom piece of neon art (see Figure \ref{fig:openings}B).
 
-Semi-closed pipes are open at the system end and closed at the user end.
-The closed interface may be fabricated from a malleable material: for example, a series of pipes terminated in thin rubber membranes on the user side can be actuated by an air pump to create haptic feedback (see Figure \ref{fig:pipespace}C).
+Semi-closed pipes are open on the system side, closed on the user side.
+The closed interface may be fabricated from a malleable material, e.g., a rubber membrane, and attached to a fluid pump to create haptic feedback (see Figure \ref{fig:openings}C).
 
 A fully enclosed pipe is closed to both the system and the user.
-Fully enclosed pipes may be used as weight modifiers (e.g., for object identification), as in Figure \ref{fig:pipespace}D).
-They can also be used as containers for water or other particles that would otherwise fall out.
+Fully enclosed pipes may be used as weight modifiers (e.g., for object identification), as in Figure \ref{fig:openings}D.
+They can also contain for water or particles that would otherwise fall out.
 
@@
 Pipe design may emphasize either exterior connection points or interior paths.
-These two foci lead to different kinds of interfaces.
-An example focused on the exterior features is the touch sensitive toy in Figure \ref{fig:toys}, where pipes must exit the toy at the frontal lobe and parietal lobes to make those \emph{specific locations} touch-sensitive.
-Focused on the internal features is the neon sign in Figure \ref{fig:neon}: the sign's output is based on the \emph{path} of the pipe.
+Focused on the exterior features is the touch sensitive toy in Figure \ref{fig:toys}, where pipes must exit the toy at the frontal lobe and parietal lobes to make those \emph{specific locations} touch-sensitive.
+Alternatively, the neon sign's output in Figure \ref{fig:neon} is based on the \emph{path} of the pipe.
 
@@
 Pipe network topologies enable different interactions.
-
-Splitting or mixing pipes offer flexibility in output.
-If a maker wished to create a painting device, she might wish to have two system-side pipes feeding in primary-colored red and blue paints which mix in varying ratios, allowing their pigments to combine before purple exits from the device (see Figure \ref{fig:pipespace}).
-Splitting can also be useful, for example if our maker wants red paint output in two locations from her painting device, she could have one system-side pipe, but split the pipe into two (see Figure \ref{fig:pipespace}).
-\bjoern{??}
-
-Star and tree topologies are extensions of the splitting and mixing primitives, encompassing multiple-in and multiple-out in contrast to single-in/single-out.
-Using a star topology in which the pipes were filled with conductive paint, we created toys with several touch-sensitive areas, see Figure \ref{fig:toys}.
+Splitting or mixing pipes offer flexibility in output, allowing a single source to have multiple outputs or multiple sources to combine (see Figure \ref{fig:pipespace}).
+Star and tree topologies extend the splitting and mixing primitives, encompassing multiple-in and multiple-out in contrast to single-in/single-out.
 
@@
 
-``Gas'' comprises all compressible fluids.
-Fluid pressure inside semi-closed pipes can create haptic feedback, or gases can be used as carriers for scents or fog.
-Slyper, et al.,  in \cite{Slyper-pressure} engineered structures to change in air pressure when manipulated correctly (e.g., a spiral that changes pressure when twisted, but not when pressed); fluid pressure can also be used as an input.
+``Gas'' comprises compressible fluids.
+Fluid pressure inside semi-closed pipes can create haptic feedback, or gases can be used as carriers for scents or fog; fluid pressure can also be used as an input.
 
-Incompressible fluids (``liquids'') can perform many of the same interface tasks as gases.
-One opportunity with liquids is to fill the interior of pipes with them and cap the ends.
-In addition, one can use driable conductive fluids, such as copper paint, to coat the interior of pipes and allow them to function as arbitrarily-shaped wires.
-This is especially helpful for the creation of a shared ground, or for creating single-wire capacitive interfaces amenable to sensing with SFCS \cite{Sato-touche}.
+Incompressible fluids (``liquids'') perform similarly to gases.  Additionally, using driable conductive liquids like copper paint, pipes can become custom wires.
 
-Pipes need not have hollow centers: in the case where routed pipes are filled with solid material---in particular, a solid material different from the model material---, interactions such as those in Printed Optics (\cite{Willis-printedoptics}) are possible.
+Pipes need not have hollow centers: in the case where routed pipes are filled with solid material---in particular, a solid material different from the enclosing model material---, interactions such as those in Printed Optics \cite{Willis-printedoptics} are possible.
 
-Particulates, either printed in-place or inserted, can be of varying densities.
-A single particle can be used for display.
-Sparse particles in a stream of fluid can provide haptic feedback.
-Dense particles in a semi-closed pipe capped with flexible surface material allow for jamming-based interactions at any point on the surface of an object \cite{Follmer-jamming}; when fluid pressure in the pipe is reduced, the trapped particles jam to form a solid mass the same shape as their enclosing flexible membrane.
+Particulates, either printed in-place or inserted, can be of varying densities.  A single particle can be used for display.  Sparse particles in fluid can provide haptic feedback.
 
-Threadable inserted elements, such as electroluminescent (EL) wire or fiberoptic cables, are those that can be threaded through pipes post-printing.
-This allows overcoming limitations of printers: for example, a Printed Optics-style interface can be created on an inexpensive consumer-grade 3D printer using pipes and inserted fiber optic cable.
+Threadable media, such as EL wire or fiberoptic cables, are those that can be threaded through pipes post-printing.
+This empowers inexpensive printers: for example, a Printed Optics-style interface can be created on a consumer-grade 3D printer with pipes and fiber optic cable.

### 2014-04-01 paper/introduction.tex
shortening

@@
 
-\tovi{condense}
 Makers, as well as professional designers, leverage 3D printers as tools for design work.
@@
 In this way, she has only a single active element (the microcontroller setup), but her otherwise-inert 3D printed object (the brain) has several sensing locations.
-Other existing sensing and actuation approaches, such as FlyEye \cite{Wimmer-flyeye} and Jamming User Interfaces \cite{Follmer-jamming}, can also be enhanced by the use of pipes: pipes allow makers utilizing such I/O strategies to locate input and output at arbitrary points on an object's surface.
-Thus, we leverage these existing techniques for sensing and actuation while our work's novelty is in \emph{internal modeling} for 3D printing and the \emph{redirection} of I/O to arbitrary locations on a device.
+Other existing sensing and actuation approaches, such as FlyEye \cite{Wimmer-flyeye} and Jamming User Interfaces \cite{Follmer-jamming}, can also be enhanced by the use of pipes: makers utilizing such I/O strategies can locate input and output at arbitrary points on an print's surface.
+We leverage these existing techniques for sensing and actuation while our work's novelty is in \emph{internal modeling} for 3D printing and the \emph{redirection} of I/O to arbitrary locations on a device.
 

### 2014-04-01 paper/relatedwork.tex
moar related work

@@
 Slyper, et al., in \cite{Slyper-pressure} 3D printed flexible robot armatures capable of sensing interaction via changes in air pressure.
-This technique is complementary to ours, as we allow for placing these interactive elements anywhere on an object's surface while they had to place them in locations accessible for electronics insertion.
-Acoustic Barcodes explored integrating ridges into digitally fabricated objects, whose acoustic signatures when scraped provided an ID code \cite{Harrison-acoustic}; while pipes can be used for identification and to create aural feedback, this is not our only goal.
-Sauron investigated the use of computer vision to track interactions with physical mechanisms \cite{Savage-sauron}.
+This technique is complementary to ours, as we allow placing these interactive elements anywhere on an object's surface; they had to place them in locations accessible for electronics insertion.
+Acoustic Barcodes explored integrating ridges into fabricated objects, whose acoustic signatures when scraped provided an ID code \cite{Harrison-acoustic}; while pipes can be used for identification and to create aural feedback, this is not our only goal.
+Sauron investigated computer vision to track interactions with physical mechanisms \cite{Savage-sauron}.
 While Savage, et al., used mirrors for redirection, we believe that the required line-of-sight operation for their computer vision techniques may be possible using printed optics or inserted optical fibers.
 General-purpose routing for connecting 2D capacitive sensors to terminals is discussed in Midas \cite{Savage-midas}.
-This technique, based on circuitboard routing, are limited to planar and unfoldable shapes, and are intended for use on the surface of objects rather than through their cores.
+This technique, based on circuitboard routing, is limited to planar and unfoldable shapes, and intended for use on the surface of objects rather than through their cores.
 Fabrickation works towards rapid fabrication of functional objects by integrating construction kit building blocks \cite{Mueller-fabrickation}.
-While we recognize that 3D printing requires a significant amount of time, we believe that creating precise internal structures in 3D printed objects is a strong case for digital fabrication over manual assembly or traditional methods.
+While we recognize that 3D printing requires significant time, we believe creating precise internal structures in 3D printed objects is a strong case for digital fabrication over manual or traditional methods.
 
@@
 Sells, et al., \cite{Sells-reprap}, describe using a heated syringe to deposit metal in printed 2D circuitboard channels on objects' surfaces.
-Sarik, et al., \cite{Sarik-tracebrush} \valkyrie{I never found a real paper of this.
-The Microsoft website has a "print" of it with TEI boilerplate, but I couldn't find it in the digital library.
-What's up with that?} and the Optomec company \cite{optomec} describe \emph{spraying} conductive material onto objects to create circuitry.
+Sarik, et al., \cite{Sarik-tracebrush} and the Optomec company \cite{optomec} describe \emph{spraying} conductive material onto objects to create circuitry.
 Injection of liquid metals for sensing in softer-than-skin electronics was explored by Park, et al., \cite{Park-microchannels} and Majidi, et al., \cite{Majidi-curvature}.
@@
 Our work differs from these because we offer \emph{algorithms and techniques} for \emph{internal}, \emph{true 3D} routing.
+
+\subsubsection{Interaction for Fabrication}
+
+Other researchers are investigating interaction for fabrication (i.e., interaction techniques for controlling digital fabrication machines) rather than fabrication for interaction as we do.
+This includes work like Constructable \cite{Mueller-constructable}, KidCAD \cite{Follmer-kidcad}, and Willis, et al,'s interactive interfaces \cite{Willis-interactive}.
+Their contributions are orthogonal to ours.
 
@@
 Multi-layer boards are typically 2.5D, in that they comprise a stack of 2D circuitboards connected by vias.
-Instead we offer true 3d point to point routing though the interior of solid models.
+Instead we offer true 3D point to point routing though the interior of solid models.
 
-Similar to our neon-sign routing algorithm, Wong presents a graph-based approach for generating continuous line illustrations from images \cite{Wong-continuousline}.
+Similar to our path routing algorithm, Wong presents a graph-based approach for generating continuous line illustrations from images \cite{Wong-continuousline}.
 Its approach of linking edges into a continuous line based on semi-Eulerization is similar to ours.
-However, we are not confined to the plane to avoid line-crossings, and we have additional freedom in creating invisible lines where none existed.
+However, we are not confined to the plane to avoid line-crossings, and we can create ``invisible'' lines.
 This is possible for us thanks to \emph{blocking out}\footnote{In neon sign making, segments of a tube that should not appear illuminated are covered with opaque paint.
@@
 Prior work used tours of nodes, rather than edges, and cast continuous line drawing as a traveling salesman problem~\cite{Bosch-tsp}.
-In traditional neon sign making, ``routing'' is a conversation between a graphic designer and a glass bender on what is feasible and comfortable for the latter and what will help him avoid burning his hands and wasting glass or bending time \cite{strattman1997neon}.
+In traditional neon sign making, ``routing'' is a conversation between a graphic designer and a glass bender on what is feasible and comfortable for the latter and what will help him avoid burning his hands and wasting time or glass \cite{strattman1997neon}.
 

### 2014-04-01 paper/discussion.tex
shortening
moar related work

@@
 \emph{Haptic} outputs are possible through compressible and incompressible fluids.
-Fluids can actuate particulates and create programmable pliability \cite{Follmer-jamming}.
+Fluids in pipes capped with flexible membranes can actuate particulates and create programmable pliability as in Jamming User Interfaces \cite{Follmer-jamming}.
 Additionally, use of semi-closed pipes with rubberlike material as the caps allows for gases to actuate surface features, like the back of our breathing rabbit (see Figure \ref{fig:breathe}).
@@
 Simple capacitance measurements are possible using the same media, and custom-designed sensors like those created in Savage, et al.,'s Midas system are also possible \cite{Savage-midas}.
+Using a technique like Fiberio \cite{Holz-fiberio}, users could even be identified on touches.
 

### 2014-04-01 paper/introduction.tex
moar related work

@@
 Pipes can also be used for sending media on a specified route through an object; for example, neon signs are created by evacuated glass tubes formed into a specific path.
-Electricity is run through the near-vacuum, and this makes the path visible.
-If our maker wants to build such a sign, she can substitute EL wire for the evacuated glass and electricity: she need only design a path for the wire to be fed through and then create pipes along that path.
+Electricity is run through the near-vacuum, creating light.
+If our maker wants to build such a sign, she can substitute EL wire for the evacuated glass: she need only design a path for it to be fed through and then create pipes along that path.
 

### 2014-04-01 paper/tubes.tex
shortn

@@
 
-Pipe design is not trivial.
+Pipe design is non-trivial.
 Independent pipes cannot intersect lest their contents be unintentionally mixed.
-Pipes' bending energy needs to be minimized to allow for easy insertion of media post-print.
-Pipes in general need to avoid the surface of objects to prevent fluid leakage when printed by hobbyist-grade machines.
+Pipes' bending energy needs to be minimized to allow easy insertion of media post-print.
+Pipes in general need to avoid objects' surfaces to prevent fluid leakage when printed by hobbyist-grade machines.
 Finally, for designing pipes that follow a user-specified path, we must ensure certain characteristics of that path (e.g., that the path is connected).
@@
 We thicken our routing to create pipes.
-\valkyrie{We also apply templates where appropriate (e.g., 3D cross-overs for tubes that intersect in the plane).}  The resultant pipes are subtracted from the original mesh, which can then be 3D printed.
+The resultant pipes are subtracted from the original mesh, which can then be 3D printed.
 
@@
 
-\bjoern{Hits section mentions a* three times, ech time with a little more detail than before.
-An probably cut this to two times as a general strategy: once in the overview, then a second time in detail.}
-Designing objects which, for example, are touch sensitive in particular areas or need to integrate with existing electronics, requires precise location and sizing of pipe endpoints.
-The interior of these pipes should have as large a bending radius as possible so that post-print insertion of solid or viscous media (like wire or paint) is easier.
+Designing objects which, for example, are touch sensitive in particular areas or integrate with existing electronics, requires precise location and sizing of pipe endpoints.
+The interior of these pipes should have as large a bending radius as possible so that post-print insertion of solid or viscous media is easier.
 
 To create pipes in which the location and shape of exterior connection points matters, we first allow makers to select the connection points on the mesh's surface.
-Our tool then creates an initial shortest-path routing using A* to estimate the routed distance between points.
-This routing is used to create a rod; we run physics-based simulation steps on the rod to minimize its bending energy (and thereby minimize the bend radius of the tubes).
 
@@
     \includegraphics[width=3.4in]{figures/exterior.png}
-\caption{An example mesh with exterior connection points selected.
-The smoothed A* routing is drawn in {\color{red}red}, and in {\color{tovi}green} is our physically-based rod after simulation.
-\valkyrie{make a figure that is less...
-tall}}
+\caption{An example mesh with exterior connection points selected.  The smoothed A* routing is drawn in {\color{red}red}, and in {\color{tovi}green} is our physically-based rod after simulation.}
 \label{fig:tool-process-exterior}
@@
 This algorithm keeps a priority queue of points to check for routing based on a cost function; for each point it tags the distance from the start (e.g., startPoint = 0, startNeighbors=1, ...).
-Once the end point is reached, we backtrack through the tagged points, always moving towards a lower-valued point.
+Once the end point is reached, we backtrack through tagged points, always moving to a lower-valued point.
 The path cost in our implementation is based only on shortest distance between the starting and ending points, without weighting for distance from the surface, however the routed path is forced to stay inside the mesh's boundaries.
-We route on the voxelized 3D grid of the mesh, and we use a resolution of 128x128x128 voxels.
+We route on the voxelized 3D grid of the mesh, using a resolution of 128x128x128 voxels.
 
-Note that pipes, once cut, become ``outside the mesh''.
-This means that pipes cut in sequence cannot intersect each other.
-However, we do not currently track previously-cut pipes for global optimization: we simply greedily select the best routing per pair of points (i.e., per pipe).
+Note that pipes once cut are ``outside the mesh'', i.e., pipes cut in sequence cannot intersect each other.
+However, we do not currently track all pipes for global optimization: we greedily select the best routing per pipe.
 
@@
 
-\valkyrie{To allow designers to create pipes with multiple endpoints (like the star topologies in Figure \ref{fig:toys}), we create 0-size mesh points in the center of the object, to which we attach one end of every pipe?}
+\valkyrie{To allow designers to create pipes with multiple endpoints (e.g., star topologies in Figure \ref{fig:toys}), we initialize an interior mesh point which we treat as a non-movable end of each pipe created.
+We then route the pipes as normal.}
 
 \subsubsection{Pipes with Non-Round Endpoints}
-\valkyrie{this section can probably be scrapped, along with Figure \ref{fig:meshmixer-endpoint}.}
-\bjoern{Replace with a single sentence: combining pipe routing with other existing mesh modification operations like extrusions enables users to create custom ...}
-Although start and end radius of pipes is configurable, our tool does not explicitly allow makers to design the \emph{shape} of their pipes' endpoints.
-However, the larger Meshmixer package does allow brushing arbitrary shapes onto the surface and extruding them inward, as in Figure \ref{fig:meshmixer-endpoint}.
-This technique can be used to create specially-shaped touch pads or to recess the footprint of an electronic component for it to sit flush with the object's surface.
+Combining pipe routing with other existing mesh modification operations like extrusions enables users to create custom-shaped touch pads or to allow an electronic component to sit flush with the print's surface.
 
 \subsubsection{Semi-Closed and Fully Enclosed Pipes}
-Users can select via a checkbox if they wish to cap one end, the other end, or both ends of their pipe.
-The software simply creates an offset from the surface in these cases, so that material remains between the pipe and the surface after mesh modification.
+Users can cap one or both ends of their pipe.  The software simply creates an offset from the surface in these cases, so that material remains between the pipe and the surface after mesh modification.
 
-\begin{figure}[h!]
-\centering
-    \includegraphics[width=3.4in]{figures/meshmixer-endpoint.png}
-\caption{After creating a tube, a maker can select a surface shape using a brush (a).  She can then extrude her selection inward (b).}
-\label{fig:meshmixer-endpoint}
-\end{figure}
+\subsection{Interior Paths}
 
-\subsection{Designing Interior Paths}
+In pipes where interior path layout matters, such as the neon sign in Figure \ref{fig:tool-process-interior} or the maze in Figure \ref{fig:maze}, we must create a single long path which passes through all desired path segments (i.e., all letters in the word ``UIST'' or all paths in the maze).
+We built a tool which leverages graph theory to generate this single long path.
 
-\bjoern{Replace specific with more general category: pipes where interior path layout matters}
-Neon signs, as in Figure \ref{fig:tool-process-interior}(d), require a single long path that passes through many desired edges: for example, in our UIST sign, the path must pass through all the edges that comprise the word ``UIST''.
-To allow makers to create novel neon signs, as well as similar objects that require design of the pipe's interior path (for example the maze in Figure \ref{fig:maze}), we built a tool which leverages graph theory to generate this single long path.
-
+In graph theory, a semi-Eulerian graph is one which has exactly two edges of odd degree.
+This kind of graph supports an Euler tour: a path that touches every edge exactly once, beginning and ending at the two nodes of odd degree (in this case, the user-selected start and end nodes).
 A maker can import a vector graphics file (such as SVG) describing the path she desires for her pipes.
@@
 We create an Euler tour on the modified graph, and thicken the path to create pipes.
-\valkyrie{3D templates are used to resolve tube crossings in the plane.}  We cut these pipes from a generated block mesh large enough to hold them.
+We then cut these pipes from the mesh.
 
@@
 \caption{An input vector graphics file with the points which cannot be tubed as drawn highlighted in {\color{blue}blue} and the start and end points highlighted in {\color{gray}gray} (a).
-The connected graph created by our software (b) and the resulting Euler circuit (c) permit creation of a novel neon sign (d).
-\valkyrie{for this sign, embed it in a fun hawaiian block}}
+The connected graph created by our software (b) and the resulting Euler circuit (c) permit creation of a novel neon sign (d).}
 \label{fig:tool-process-interior}
@@
 \tovi{Still too long....}
-\bjoern{Need this explanation above so readers don't wonder what Euler tours are
-
-Agree with Tovi that this can be condensed}
-In graph theory, a semi-Eulerian graph is one which has exactly two edges of odd degree.
-This kind of graph supports an Euler tour: a path that touches every edge exactly once, beginning and ending at the two nodes of odd degree (in this case, the user-selected start and end nodes).
-To allow for a single path through a pipe for post-print media insertion, we need to find a semi-Eulerian graph for which a user's input graphics are a subset, and subsequently find an Euler tour on that graph.
+\bjoern{Agree with Tovi that this can be condensed}
+To allow for a single path through a pipe for post-print media insertion, we need to find a semi-Eulerian graph of which a user's input graphics are a subset, and subsequently find an Euler tour on that graph.
 

### 2014-04-01 paper/abstract.tex
abstract

@@
 \begin{abstract}
-3D printers offer extraordinary flexibilty in prototyping the shape and mechanical function of objects.
+3D printers offer extraordinary flexibilty for prototyping the shape and mechanical function of objects.
 However, a 3D printer's role in prototyping interactive objects is still not clear.
-We propose a general technique for adding interactivity by removing interior material to form pipes and cavities.
-We describe the design space of pipes and hollow chambers for interaction design, touching on of openings, topologies, and inserted media.
-We present a technique and design tool for routing pipes through the interior of 3D models.
-We integrated our prototype with a 3D modeling program.
+We propose a general technique for supporting the rapid prototyping of interactivity by removing interior material of 3D models to form pipes and cavities.
+We describe this new design space of pipes and hollow chambers for interaction design, where variable sinclude openings, topologies, and inserted media.
+We then present a  tool for routing such pipes through the interior of 3D models, integrated  within a 3D modeling program.
 We use two distinct routing algorithms.
-One, for creating particular points of interaction or integrating with existing electronics on the model's surface, allows users to select pipes' terminals, then uses A* path routing and physics-based simulation to minimize the bending energy of routed paths.
-The second, for creating, e.g., novel neon signs, has users supply a desired internal shape which we fit a pipe route to: for this we developed a novel graph-based routing algorithm.
+The first allows users to define the pipes' terminals, and then uses A* path routing and physics-based simulation to route the paths with minimized  bending energy.
+The second allows users to supply a desired internal shape which we fit a pipe route to: for this we developed a novel graph-based routing algorithm.
 We present several totally tubular prototypes we created using our tool to show its flexibility and potential.

### 2014-04-01 paper/abstract.tex
updated intro

@@
 We propose a general technique for supporting the rapid prototyping of interactivity by removing interior material of 3D models to form pipes and cavities.
-We describe this new design space of pipes and hollow chambers for interaction design, where variable sinclude openings, topologies, and inserted media.
+We describe this new design space of pipes and hollow chambers for interaction design, where variables include openings, topologies, and inserted media.
 We then present a  tool for routing such pipes through the interior of 3D models, integrated  within a 3D modeling program.

### 2014-04-01 paper/discussion.tex

@@
 Even if printers are capable of laying down conductive, optically transparent, or luminescent materials in-place, the process of designing their paths is still crucial.
-
-\tovi{we should make this into a "1D table", just base it on the I/O we can enable.
-scrap the table illustration itself; it's not needed.
-also, focus more on future work in these areas, and work on relating back to what we've discussed already (and the examples we've built)}
-
-\tovi{There's lots of great ideas an examples in the inputs and outputs section, but the orgnization and coorespondence to the table is a little hard to follow. }
 

### 2014-04-01 paper/tubes.tex

@@
 In graph theory, a semi-Eulerian graph is one which has exactly two edges of odd degree.
-This kind of graph supports an Euler tour: a path that touches every edge exactly once, beginning and ending at the two nodes of odd degree (in this case, the user-selected start and end nodes).
+This kind of graph supports an Euler tour: a path that travels every edge exactly once, beginning and ending at the two nodes of odd degree (in this case, the user-selected start and end nodes).
 A maker can import a vector graphics file (such as SVG) describing the path she desires for her pipes.
@@
 \subsubsection{Routing}
-\tovi{Still too long....}
-\bjoern{Agree with Tovi that this can be condensed}
 To allow for a single path through a pipe for post-print media insertion, we need to find a semi-Eulerian graph of which a user's input graphics are a subset, and subsequently find an Euler tour on that graph.
 
-The interior path routing problem is a version of the Chinese Postman Problem\footnote{The CPP is also known as the route inspection problem: \url{http://en.wikipedia.org/wiki/Route_inspection_problem}}, in which we wish to traverse all edges of the graph described in the user's input with a single Euler circuit, much as a postman needs to walk along every road at least once to deliver mail.
-We call our relaxation the Spiderman Postman Problem: we allow the creation of non-existing paths (i.e., the postman may traverse buildings in addition to roads).
-If input path components are disconnected, we must create edges that connect them; additionally we can create new edges connecting odd-degree vertices rather than simply retracing existing edges.
+The interior path routing problem is a version of the Chinese Postman Problem\footnote{The CPP is also known as the route inspection problem: \url{http://en.wikipedia.org/wiki/Route_inspection_problem}}, in which we wish to traverse all edges of the graph described in the user's input, much as a postman needs to walk along every road at least once to deliver mail.
+In our relaxation, we allow creation of new edges (i.e., the postman may traverse buildings in addition to roads).
+If path components are disconnected, we create edges that connect them; additionally we can create edges connecting odd-degree vertices rather than simply retracing existing edges.
 In the final artifact, all created edges will be blocked out by dark material so the inserted medium is not visible (see Figure \ref{fig:tool-process-interior}).
@@
 
-We need a graph which is semi-Eulerian (i.e., the inserted media can enter at one point, traverse every edge exactly once, and exit at a different point).
-In order to create a semi-Eulerian graph, we first add a temporary edge to the start and end points.
-This is removed after the graph is made fully Eulerian.
+In order to create a semi-Eulerian graph, we first add a temporary edge to the start and end points.  This is removed after the graph is made fully Eulerian.
 
-In order to connect disconnected subgraphs in the input, each disconnected subgraph's vertices and edges are contracted to a single vertex.
-Each of these subgraph-vertices has no outgoing edges, as the expanded subgraphs are disconnected.
-We then add one edge from each subgraph-vertex to each other subgraph-vertex, whose weight is the minimum distance between any two vertices in the expanded subgraphs (where distance is Euclidean distance).
+To connect disconnected subgraphs in the input, each disconnected subgraph's vertices and edges are contracted to a single vertex.
+Each subgraph-vertex has no outgoing edges, as the expanded subgraphs are disconnected.
+We then add one edge from each subgraph-vertex to each other subgraph-vertex, whose weight is the minimum Euclidean distance between any two vertices in the expanded subgraphs.
 We greedily select the smallest weight edges until all subgraphs are joined into a single graph.
@@
 
-In order to make our graph Eulerian, we consider all vertices of odd degree in the connected graph and make a clique of edges once more based on distance.
-We again greedily select edges between the odd nodes until no odd nodes remain.
+In order to make our graph Eulerian, we consider all odd degree vertices in the connected graph and make a clique of potential edges based on distance.
+We greedily add edges between the odd nodes until no odd nodes remain.
 Finally, we remove our temporary edge, which changes the graph from Eulerian to semi-Eulerian.
-This graph is a connected, semi-Eulerian graph which contains all edges in the input SVG (see Appendix).
+This is a connected, semi-Eulerian graph which contains all edges in the input (see Appendix).
 
-We believe that a lower total weight matching is possible by connecting components and ensuring edge degree evenness together in a global process, as well as by using minimum-weight matching rather than greedy selection, however this optimization is not crucial for our purposes.
+We believe that a lower total weight matching is possible by connecting components and ensuring node evenness together in a global process, as well as by using minimum-weight matching rather than greedy selection, however this optimization is not crucial for our purposes.
 
@@
 This minimizes turns in the final artifact, which eases support material removal and assembly.
-
-\begin{figure}[h!]
-\centering
-    \includegraphics[width=3.4in]{figures/placeholder/templates.jpg}
-\caption{\valkyrie{When edges intersect in the plane (a), we match them to the nearest of our several intersection templates (b) to create 3D paths that do not interfere with each other.}}
-\label{fig:templates}
-\end{figure}
 
@@
 We also consider edges at nodes with degree $>2$, as these edges are considered to cross.
-When intersections have been detected, we match one of several templates to the affected edges (see Figure \ref{fig:templates}) using a nearest neighbors technique.
-Our templates are parameterized, so the templates can be modified to accommodate precise intersection characteristics.}
+When intersections have been detected, we just push things around.
+No, really.}
 
@@
 
-Given either our rod-based or image-based routing, we create cylinders which use the routings as a sweep path.
-At each point in the routing, we create a cylinder tall enough to reach to the next point which is oriented along the line segment connecting its origin point to its destination point.
+Given either our rod-based or image-based routing, we create cylinders which use the routing as a sweep path.
+At each point in the routing, we create a cylinder tall enough to reach to the next point, oriented along the line segment connecting its origin and destination.
 This cylinder's radius is set by the user.
@@
 These two voxel grids are subtracted, and the result is re-meshed and smoothed.
-\bjoern{ rewrite When a user desires to cap the ends of her pipes, we create a sphere of the same radius as each endpoint whose center is the same as the endpoint.
-We intersect this sphere with the original mesh, then intersect this partial-sphere mesh with a shell of the original mesh.}  This gives us a voxel subgrid of the thickness of cap we desire (the same as the shell's thickness) which is located only in the region of the tube end.
-During subtraction, we preserve this subgrid.
-\valkyrie{maybe need a figure?}
+For creating semi-closed pipes, during the voxelization subtraction we create an offset surface of the mesh where it intersects the tube and preserve it.
 
 Unfortunately, because we use voxelization as a part of our subtraction operation, our models can accumulate error when many pipes are created in sequence (see Figure \ref{fig:voxelize}).
-This problem could be solved with a higher voxel resolution, but it requires balancing computation time and precision.
+This problem can be solved with a higher voxel resolution, but requires balancing computation time and precision.
 
@@
 \subsection{Fabrication Techniques}
+
+Beyond design, the tubes must actually be printable to be useful.
+The physical fabrication process used by different machines gives rise to differing strategies for avoiding support material or easing its removal.
+We will discuss techniques for fused deposition modeling (FDM) machines like the Makerbot and multi-jetting machines like the Objet.
 

### 2014-04-01 paper/introduction.tex

@@
 \section{Introduction}
-
+\tovi{Typical intro format: first pragraph should not introduce your own work, but summarize the problem and motivation.
+Second paragraph should summarize prior art and its limations.
+After that you can describe your work and contributions.
+I've made some revisions to the strucutre...}
 Makers, as well as professional designers, leverage 3D printers as tools for design work.
 A wide array of objects, ranging from bicycle helmets to video game controllers, are now prototyped or even manufactured using these machines.
-However, most devices fabricated by 3D printers are passive.
+However, most devices fabricated by 3D printers are passive, prototyping the form, but not the interactivity, of an object.
+
+Due to the growing popularity of 3D printing and Maker Communities [ref?], recent work has looked at the issue of rapidly prototyping interactive objects.
+Existing approaches include ...
+but their limitations are ...
+
 We propose a novel technique involving the removal of material from the interior of 3D models, prior to printing, to create pipes and other cavities.
 These pipes can be filled, post-print, by a variety of media that enable input, display, and tactile feedback.
-This {\em subtractive} approach is complementary to {\em additive} approaches that try to print different materials, such as conductors.
-While our approach requires manual assembly afterwards, it opens up a large useful design space on commonly available 3D printers, allowing makers to create objects like those in Figure \ref{fig:teaser}.
+This {\em subtractive} approach is complementary to {\em additive} approaches that inject custom print materials, such as conductors, during the actual printing process [ref].
+While our approach requires some manual assembly after a print completes, it gives makers the opprotunity to rapidly prototype a diverse range of interactive objects on commonly available 3D printers (Figure \ref{fig:teaser}).\tovi{I guess we need to be careful here since not all of our examples are on doen with commonly available 3D printers}
 
-These channels introduce an entire design space of opportunities for adding interactivity to these objects.
-For example, copper material can fill the channels, to allow for standard electronic components to be easily integrated after the printout; electroluminescent (EL) wire can be threaded through to create computer-controlled visual output; or air can be pumped through to control haptic responses.
+We describe a new design space of pipes and hollow chambers for the prupose of interaction design, where variables include openings, topologies, and inserted media.
+By exploring this design space new opportunities for adding interactivity to 3D printed objects can be realized.
+For example, copper material can fill the channels, to allow for standard electronic components to be easily integrated after the printout; electroluminescent (EL) wire can be threaded through to create computer-controlled visual output; or air can be pumped through to control haptic responses.\tovi{at some point explain why this is better than running wires on the exterior (looks better, for one)}
 
+\tovi{I would remove this entire paragraph}
 Pipes allow makers to build with and upon existing input/output techniques, often integrating these methods into objects they were not originally designed for.
@@
 
+\tovi{I would remove this entire paragraph}
 Pipes can also be used for sending media on a specified route through an object; for example, neon signs are created by evacuated glass tubes formed into a specific path.
@@
 Generating these pipes, however, is not trivial.
+For one, most 3D modelling tools do not provide extensive tools for modelling the  {\em interior} of 3D models.
+Furthermore, careful consideration needs to go into the design and routing of the pipes.
 Pipes need maximum possible bend radius to ease the insertion of media post-print.
@@
 Pipes should not pass too close to the surface: consumer-grade 3D printers do not guarantee air- or water-tightness of prints, so more than one printed layer may be required to prevent leakage.
-Pipes for applications like neon signs must have an Euler tour through them: this allows a single EL wire to light every edge on the pipe graph.
-No tools yet exist for these design tasks.
-We have developed techniques for minimizing the bending energy of, enforcing non-intersection with, and avoiding the surface for pipes, as well as expanding user-specified paths and generating Euler tours.
-They are embodied in our software tool enabling makers to create pipe-powered interfaces.
+Pipes for applications like neon signs must have a Euler tour through them: this allows a single EL wire to light every edge on the pipe graph.
+No tools currently exist for these design tasks.
 
-Our work makes the following contributions to the field of digitally fabricated interactive objects:
-
-\begin{itemize}
-\item We lay out a design space of pipes and their openings, contained media, topologies, and design focus.
-\item We offer algorithms and techniques for routing pipes, as well as a design tool implementing them.
-\item We showcase a set of examples, enabled by our modeling tool and exploring new points in the design space.
-\end{itemize}
+Our work makes several contributions to the field of digitally fabricated interactive objects.
+First, we describe a design space of how pipes can be integrated into 3D models.
+We then offer algorithms and techniques for routing the pipes, and integrate these tools into  a 3D modelling applicaiton.
+The system takes into considertation the design complexities of routing interior pipes which are described above.
+Finally, we showcase a set of examples, enabled by our modeling tool, which allows us to explore various points in the design space.
+We conlcude by summarizing the input and out modalities enabled by our work, and discussing limitations and areas for future exploration.

### 2014-04-01 paper/relatedwork.tex
related work design space

@@
-\section{Related Work}
-Our contributions relate to other efforts to add interactivity to digitally fabricated objects; and to techniques for routing and model modification.
+\section{Related Work}\tovi{related work section still needs some work}
+Our contributions relate to other efforts aiming to add interactivity to digitally fabricated objects; and to techniques for routing and model modification.
 
@@
 Previous work has investigated the integration of interactivity with the fabrication process.
-Printed Optics (\cite{Willis-printedoptics}) used light pipes, solid clear tubes, to create integrated displays and touch sensors: our work is inspired by this and goes towards a design tool for these capabilities.
+Printed Optics (\cite{Willis-printedoptics}) used light pipes, solid clear tubes, to create integrated displays and touch sensors.
+The describes a process, but does not provide a design tool for creating such objects.
+Our work is inspired by Printed Optics and introduces a design tool for modelling objects that can have these types of capabilities.
+
+\tovi{dont use citations as nouns}
 Slyper, et al., in \cite{Slyper-pressure} 3D printed flexible robot armatures capable of sensing interaction via changes in air pressure.
-This technique is complementary to ours, as we allow placing these interactive elements anywhere on an object's surface; they had to place them in locations accessible for electronics insertion.
-Acoustic Barcodes explored integrating ridges into fabricated objects, whose acoustic signatures when scraped provided an ID code \cite{Harrison-acoustic}; while pipes can be used for identification and to create aural feedback, this is not our only goal.
+\tovi{clean up writing - finish telling the reader what they did or didnt do, then respond with what we do differently}This technique is complementary to ours, as we allow placing these interactive elements anywhere on an object's surface; they had to place them in locations accessible for electronics insertion.
+\tovi{not sure this is relevant enough to cite here}Acoustic Barcodes explored integrating ridges into fabricated objects, whose acoustic signatures when scraped provided an ID code \cite{Harrison-acoustic}; while pipes can be used for identification and to create aural feedback, this is not our only goal.
 Sauron investigated computer vision to track interactions with physical mechanisms \cite{Savage-sauron}.
 While Savage, et al., used mirrors for redirection, we believe that the required line-of-sight operation for their computer vision techniques may be possible using printed optics or inserted optical fibers.
+
 General-purpose routing for connecting 2D capacitive sensors to terminals is discussed in Midas \cite{Savage-midas}.
@@
 
-Other research has brought traditional electronics to 3D prints.
+Other research has investigated the integration of traditional electronics into 3D prints.
 Sells, et al., \cite{Sells-reprap}, describe using a heated syringe to deposit metal in printed 2D circuitboard channels on objects' surfaces.
@@
 Our work differs from these because we offer \emph{algorithms and techniques} for \emph{internal}, \emph{true 3D} routing.
+\tovi{relation of this stuff to ours isn't so clear}
 
@@
 This includes work like Constructable \cite{Mueller-constructable}, KidCAD \cite{Follmer-kidcad}, and Willis, et al,'s interactive interfaces \cite{Willis-interactive}.
-Their contributions are orthogonal to ours.
+Their goals and contributions are orthogonal to ours.
 
@@
 All these contributions are orthogonal to ours, as our focus is on modeling pipes and hollow chambers for fabricating interactive objects.
+\tovi{need better summary statement instead of ``all of this is orthogonal''}

### 2014-04-01 paper/tubes.tex

@@
 Beyond design, the tubes must actually be printable to be useful.
+When geometries contain many overhangs, 3D printers lay down support material that is later removed from the model.
+This removal process can be time-intensive and challenging.
 The physical fabrication process used by different machines gives rise to differing strategies for avoiding support material or easing its removal.
 We will discuss techniques for fused deposition modeling (FDM) machines like the Makerbot and multi-jetting machines like the Objet.
+
+FDM machines lay down lines of plastic on top of each other: this allows ``stepover'', when a layer of plastic is laid slightly offset from the layer below it, and ``bridging'', when plastic is laid across a gap between two surfaces.
+For a layer to fix correctly to its supporting layer, roughly half of the filament needs to overlap.
+This means that overhangs without added support material can be approximately $45^{\circ}$, and although cylinders (e.g., pipes) and spheres violate this in some places they are printable support-free.
+When support is laid, it is made from the same material as the model in a heatsink-like wave pattern.
+It is straightforward to remove from exterior model surfaces with fingers or pliers, but nearly impossible to remove from internal cavities.
+
+On multi-jetting machines, bridging is not possible and stepover potential is much less, because the machine is laying down a liquid rather than a plastic.
+Unsupported overhangs can be approximately $14^{\circ}$.
 

### 2014-04-02 paper/discussion.tex
comments on latter half of the paper

@@
 Pipes leverage that promise: designing the internal structure of an object was not previously feasible.
-As printers at the hobbyist and corporate levels become more sophisticated and able to fabricate intricate geometries of ever more materials, we believe that our tool will still assist in design tasks.
-Even if printers are capable of laying down conductive, optically transparent, or luminescent materials in-place, the process of designing their paths is still crucial.
+\bjoern{overly broad claim.
+companies are designing the "interior " when they make complex shelled parts using injection molding.
+But that process has lots of limitations as to the geometries that are possible --- there's a whole industry of specialists concerned with ``design for manufacturability'' (DFM).} As printers at the hobbyist and corporate levels become more sophisticated and able to fabricate intricate geometries of ever more materials, we believe that our tool will still assist in design tasks.
+\bjoern{oblique, and could be phrased more positively.} Even if printers are capable of laying down conductive, optically transparent, or luminescent materials in-place, the process of designing their paths is still crucial.
 
@@
 We also suggest new points in the design space of pipe-mediated interaction that remain unexplored.
+\bjoern{need to motivate first why the discussion is going to focus on inputs and outputs - why is this the logical topic to return to?
+Maybe something like this: ``By crossing pipe types, topologies, and inserted media, many pipe-mediated interactions emerge.
+We demonstrated a few such possibilities by designing and fabricating example objects with the help of our pipe routing design tool.
+However, there are many additional possibilities for leveraging sensing and actuation techniques from the HCI literature and extending their use in 3D-printed objects.
+We briefly survey such possibilities for output and input.''}
 
@@
 
-We describe potential pipe-based outputs organized by the five senses: visual, aural, haptic, and olfactory/gustatory.
+Pipe-based outputs can address the five senses: visual, aural, haptic, and olfactory/gustatory.
 In addition to outputs we mention explicitly, any outputs possible using traditional electronic components are possible using conductive material in pipes.
@@
 \emph{Visual} outputs can be mediated by gases, liquids, solid, particulates, or threadables.
-Pneumatic actuation (as in PneUIs \cite{Yao-pneui} or Harrison's latex buttons \cite{Harrison-buttons}) provides output, but gases carrying smoke (as in the Aireal \cite{Sodhi-aireal} demo) also offers visual feedback.
+Pneumatic actuation (as in PneUIs \cite{Yao-pneui} or Harrison's latex buttons \cite{Harrison-buttons}) provides output, but gases carrying smoke (as in the Aireal \cite{Sodhi-aireal} demo) also offers visual feedback \bjoern{mischaracterization of aireal - they only added smoke to show what a vortex looks like; the technique is all about invisible, haptic sensation}.
 Splitting and mixing of colored liquids can be used as a display tactic; additionally mechanical motion can be fluid-mediated.
 Liquids can also be mixed in specific concentrations by servo-controlled gate structures as in Figure \ref{fig:direct}.
-Printed optics \cite{Willis-printedoptics} explored visual output at the ends of pipes, and  EL wire can be threaded through custom paths to create neon signs, as in our neon robot (Figure \ref{fig:neon}).
+Printed optics \cite{Willis-printedoptics} explored visual output at the ends of pipes, and  EL wire can be threaded through custom paths to create neon signs, as in our neon robot (Figure \ref{fig:neon}) \bjoern{are we going to repeat things here we already talked about?
+Maybe we should just focus on ideas one could build that we haven't built yet}.
 
@@
 While their technique requires access a terahertz imaging tool, audio resonance identity encoding requires only a microphone and a speaker.
+
+\bjoern{We might pull back on the scope of this section a bit and think about which directions here are most closely related to pipes and most likely successful;  and which are more tangential or speculative - for example, Aireal is about feedback in free space and we are all about designing tangible objects; so maybe that's a candidate for cutting.}

### 2014-04-02 paper/designspace.tex

@@
 \section{The Design Space of Pipes}
-By using different openings, designs, and topologies of pipes, and inserting different media into the completed pipes, inputs and outputs can be located at arbitrary points on the surface of a 3D print.
+\tovi{clean up writing}By using different openings, designs, and topologies of pipes, and inserting different media into the completed pipes, inputs and outputs can be located at arbitrary points on the surface of a 3D print.
 We describe the space below.
@@
 User or system manipulation of pipe contents are possible even without physical access, e.g., via soundwaves.
+\tovi{still not convinced we need this distinction}
 
@@
 This type of pipe may be used to create electronic circuitry.
-An open pipe filled with conductive paint can power an LED (see Figure \ref{fig:openings}A).
+For example, an open pipe filled with conductive paint can power an LED (see Figure \ref{fig:openings}A).
 
@@
 Semi-closed pipes are open on the system side, closed on the user side.
-The closed interface may be fabricated from a malleable material, e.g., a rubber membrane, and attached to a fluid pump to create haptic feedback (see Figure \ref{fig:openings}C).
+A potential use of this is to create tactile output: The closed interface may be fabricated from a malleable material, e.g., a rubber membrane, and the open end attached to a fluid pump to create haptic feedback (see Figure \ref{fig:openings}C).
 
@@
 Alternatively, the neon sign's output in Figure \ref{fig:neon} is based on the \emph{path} of the pipe.
+\tovi{still think we can remove this dimension}
 
@@
 Star and tree topologies extend the splitting and mixing primitives, encompassing multiple-in and multiple-out in contrast to single-in/single-out.
+\tovi{give simple examples of where this might be useful}
 
 \subsection{Media in Pipes}
-
-Pipes can be filled with a variety of media to create different interface affordances and capabilities.
+\tovi{clean up this writing}
+One of the most interesting aspects of the design space is that pipes can be filled with a variety of media to create different interface affordances and capabilities.
+We consider gas, liquids, solids, particulates, and threadables.
 
@@
 
-Incompressible fluids (``liquids'') perform similarly to gases.  Additionally, using driable conductive liquids like copper paint, pipes can become custom wires.
+Incompressible fluids (``liquids'') perform similarly to gases.
+Additionally, using driable conductive liquids like copper paint, pipes can become custom wires.
+\tovi{custom water jet devices (uist 2013 competition)}
 

### 2014-04-02 paper/exampleobjects.tex

@@
 
-To evaluate and highlight our tool's capabilities, we fabricated a set of five prototype objects designed using the tool.  All prototypes were fabricated in a single piece, unless described otherwise.
+To evaluate and highlight our tool's capabilities, we fabricated a set of five prototype objects designed using a Series of Tubes.
+All prototypes were fabricated in a single piece, unless described otherwise.
 
@@
 
-We created a set of touch-sensitive toys and a companion app, reminiscent of the boat application in Acoustic Barcodes \cite{Harrison-acoustic}.
-The brain, rabbit, and boat each have several distinct touch points on their surfaces.
-These touch points are connected by an interior star topology of tubes filled with conductive copper paint, and touch sensing is performed via a single wire and SFCS (see Figure \ref{fig:toys}).
+We created a set of touch-sensitive toys and a companion app, reminiscent of the boat application in Acoustic Barcodes \cite{Harrison-acoustic} \bjoern{People will not  know what this is; they shouldn't have to look up that paper to understand.
+I rewrote:}.
+The objects --- a brain, rabbit and boat, in our example --- can be set on a base, which announces the type of object, and also the names of special features of an object when that feature is touched.
+For example, touching a strip on top of the brain model yields the announcement ``primary motor cortex''.
+The distinct touch points on each object are connected by an interior star topology of tubes filled with conductive copper paint, and touch sensing is performed via a single wire and SFCS (see Figure \ref{fig:toys}).
 We built a smart base which can distinguish between the toys and also determine which toy is mounted: since each toy and each gesture has a distinct capacitive signature, we use a simple classifier trained to detect both toy and gesture based on profile.
@@
 This paint requires approximately 2 days to fully dry in this configuration (3mm diameter tubes, maximum tube length 10cm).
-Our smart base is powered by an Arduino Uno running code by Instructables users DZL and madshobye.
+Our smart base is powered by an Arduino Uno running code by Instructables users DZL and madshobye \bjoern{what is that code?
+Maybe just mention open source SFCS code and footnote a URL?}.
 
 \subsection{Breathing Bunny (semi-closed, gas, exterior)}
+\bjoern{Start each section with the larger point why we built each example.
+`` (semi-closed, gas, exterior)'' is likely to short.
+``To demonstrate how gases can be used to both provide sensation to the user through openings and to deform a model internally, we created a rabbit...''}
+\bjoern{why do we have two bunnies in this and the prior application? Maybe we should substitute a different model for the bunny above.}
 
-We created a rabbit with a pair of tubes that can simulate breathing (see Figure \ref{fig:breathe}).
+We created a rabbit with a pair of tubes that can simulate breathing (see Figure \ref{fig:breathe}) \bjoern{How does it simulate breathing?
+Be specific ``the bunny exhales and its abdominal area rises and falls with each breath'' or something like that}.
 For this, we used a combination air/vacuum pump: one terminal creates a vacuum while the other creates positive pressure.
@@
 To fabricate the tube, we created it in two halves that were fastened together via glue, due to limitations in our ability to remove support material otherwise.
-We created this prototype on our Makerbot in two halves, inserting a marble before closing.
+\bjoern{observant reviewers will call us out on this - that many of our techniques may not actually be workable in a single piece because you can't get the support material out.
+I think if you had a lye bath you could actually get more complicated models done, as long as you're patient and there's some entry and exit for the lye to reach all pipes.}  We created this prototype on our Makerbot in two halves, inserting a marble before closing.
 

### 2014-04-02 paper/limitations.tex

@@
 Fluids in particular provide challenges.
-For example, the conductive paint we used in our designs can take multiple days to dry in tubes of sufficient length.
+\bjoern{I'd think that solids would be harder - they can get stuck along the way.} For example, the conductive paint we used in our designs can take multiple days to dry in tubes of sufficient length.
 Any liquid kept in tubes printed by the Objet can cause the exterior of the part to discolor due to seepage.
 Hobbyist machines do not guarantee watertightness of their prints, so leaking of water or air can occur.
+
+\bjoern{also comment on the difficulty of getting support materials out?}
 

### 2014-04-02 paper/tubes.tex
comments on latter half of the paper
first set of glamour shots, paper work, redo of figures on new mm background, redo some figures due to new modeling

@@
 \section{A Series of Tubes - A Tool for Pipe Design}
+\bjoern{What kind of visualizations and diagrams can help explain the technical part other than photos and screenshots?
+The "UIST" pipe routing figure goes in this direction.
+Can we make similar conceptual figures, flowcharts, etc for the earlier parts?}
 
@@
 Beyond design, the tubes must actually be printable to be useful.
-When geometries contain many overhangs, 3D printers lay down support material that is later removed from the model.
+When geometries contain overhangs, as there may be in designed pipes and cavities, 3D printers lay down support material that can be later removed from the model.
 This removal process can be time-intensive and challenging.
@@
 This means that overhangs without added support material can be approximately $45^{\circ}$, and although cylinders (e.g., pipes) and spheres violate this in some places they are printable support-free.
-When support is laid, it is made from the same material as the model in a heatsink-like wave pattern.
-It is straightforward to remove from exterior model surfaces with fingers or pliers, but nearly impossible to remove from internal cavities.
-
-On multi-jetting machines, bridging is not possible and stepover potential is much less, because the machine is laying down a liquid rather than a plastic.
-Unsupported overhangs can be approximately $14^{\circ}$.
-
-\bjoern{Argue first that we need processes to fabricate these hollow pipes- can do this by removing support material or never printing such material in the first place.}  \bjoern{Is section needs to bee written from a general standpoint first, and then our specific details second.}
-The physical fabrication process used by different machines gives rise to differing strategies for avoiding support material or easing its removal.
-The two machines we used to create our example objects were the Objet Connex 260 and the Makerbot Replicator 2.
-
-The Objet is an inkjet-based 3D printer which jets drops of polymer in very fine layers, curing each layer with an ultraviolet lamp.
-It offers a wide range of material types, including clear/opaque, flexible/hard, and black/white.
-Additionally, it permits the mixing of two different materials in a single print: a mixed material could include 70\% flexible and 30\% hard, for example, to be firm but somewhat flexible.
-The flexible materials can be inflated when they are thin: they elongate 200\% before breaking.
-We have found $.05$ inches to be a good thickness for inflatable materials.
-Although this machine has many extraordinary capabilities, it unfortunately cannot print overhangs of greater than $14^{\circ}$ without support material.
-Due to this, support avoidance is difficult; our main tactic for removing support material has been to make the \emph{openings} to pipes as large as possible.
-This allows us to flush the material most easily.
-The support structure used by this machine is a loose matrix of material that can be removed by hand from the surface, blasted out with a water jet from cavities, or dissolved in lye (the enclosing model material does not dissolve).
-
-The Makerbot Replicator 2 is a fused-deposition modeling (FDM) printer, in which we use exclusively Polylactic Acid (PLA) bioplastics.
-More material properties are being explored in PLA: there are now transparent and flexible materials.
-They cannot be blended like those in the Objet.
-The Makerbot has a fairly coarse resolution compared to the Objet, however it permits overhangs up to $45^{\circ}$ without support and also allows bridging (short spans of material that bridge the gap between two structures with no support underneath).
-Although spheres and sideways cylinders have a $>45^{\circ}$ overhang in some places, they are also printable without support.
-Due to the FDM process it can even recover from greater overhangs and longer bridges, as the material builds up and can reorganize into a structure \valkyrie{how do I say this?
-do I need a photo?
-print recovery is pretty awesome-looking.}.
-The support used by the Makerbot is made of the same material as the model itself, just printed in a different pattern and attached over as small a surface area as possible to ease removal.
-Support inside of a pipe may require pliers, or be infeasible to remove dependent upon geometry.
-Meshmixer has a built-in tool for minimizing support for prints; it re-orients the part to keep as much surface area as possible within the 45 degree constraint.
+Support is made from the same material as the model, laid in a heatsink-like wave pattern.
+It is straightforward to remove from exterior surfaces with fingers or pliers, but nearly impossible to remove from internal cavities.
+Meshmixer has a built-in tool for minimizing support for FDM prints; it re-orients the part to keep as much surface area as possible within the 45 degree constraint.
 In future work we would like to explore a tool in which a user can select surfaces to prioritize support avoidance for.
 
-Assembly is easiest on both machines when the model does not require support material.
-For especially complex geometries, models can be ``cut up'' \bjoern{This negates some of the benefits compared to other manufacturing techniques.} into assembleable pieces that evade support requirements or ease material removal.
-We believe that an automatic tool for this task would be highly beneficial, however it is outside the scope of this work.
+On multi-jetting machines, bridging is not possible and stepover potential is much less, because the machine is laying down droplets of liquid rather than more structural plastic.
+Unsupported overhangs can be approximately $14^{\circ}$.
+The support structure used by this machine is a loose matrix of material that can be removed by hand from the surface, blasted out with a water jet from cavities, or dissolved in lye (the enclosing model material does not dissolve).
+Thus, any cavity to which there is some access can be cleared of support.
+
+For especially complex geometries, models can be ``cut up'' into assembleable pieces that evade support requirements or ease material removal.
+This is undesirable as it means more time spent in assembly post-print, and additionally a cut pipe may not be fluid-tight any longer.
+An automatic tool for this task might be beneficial, however it is outside the scope of this work.

### 2014-04-03 paper/tubes.tex
ryan text, new figures of star topo objects

@@
 
-Given our initial routing, we create a rod of the same length.
-To allow for easiest insertion of media post-print, our physical simulation step minimizes the bend radius of the generated rod.
-In order to force our simulated rod away from the surface of the mesh, we create ``wind'' pushing inward from the exterior of the mesh.
-This wind keeps the resulting pipe away from the surface.
-\valkyrie{Ryan: How does it actually work?} \bjoern{Need much more detail here or a reference to a previously published technique}
+Our A* pathfinding produces the \emph{shortest} path, but in many cases
+we require a smoother path through which we can actually push a physical wire.
+We solve this problem via physical simulation. We create a straight virtual 
+wire (a 3D poly-line or \emph{discrete rod}), and then set its initial position 
+to be deformed to lie along our A* path.
+Running the physical simulation allows the wire to attempt to return to
+its straight configuration, while satisfying endpoint position constraints.
+We also increase the length of the wire, to provide some slack in cases where 
+it would otherwise bend sharply to stay inside the shape.
+
+The wire is modelled as a \emph{discrete rod}, ie a 3D poly-line with a bending 
+constraint between adjacent segments. Our simulation is based on an
+implementation of Position-Based Dynamics \cite{Muller07}, with all constraints
+modelled as penalty forces. To keep the rod inside the
+mesh, we compute a discretized distance field, we can then constrain the
+rod to stay within some offset shell, again using a simple penalty force.
+In the case of multiple initial paths, we solve for each smoothed path sequentially.
+After solving for a path, we add it as a constraint to the system, generating
+a penalty force that pushes additional paths away from it.
+
+This basic approach does have various limitations, in particular it
+cannot recover if the discrete rod becomes tangled or unstable. 
+However we have observed these problems only when attempting to break
+the system; they did not occur in any of the cases shown in our figures.
+Another issue is parameters, as the penalty forces used in PBD do not
+correspond to physical units. We avoided serious issues here by uniformly
+scaling the initial mesh to a unit box, and tuning the parameters on several test cases.
+PBD simulations rarely converge to a steady state, 
+so we simply halt after a fixed number of timesteps (250). 
+The simulation is very fast so this only takes a second for most of our examples.
 
 \subsubsection{Pipes with Multiple Endpoints}
-For the creation of pipes with multiple endpoints, for example the star topologies in Figure \ref{fig:toys} or tree topologies, makers can select interior faces of existing pipes as endpoints, and thereby create pipe networks by hand.
-We do not currently offer any pipe network manipulation tools.
-
-\valkyrie{To allow designers to create pipes with multiple endpoints (e.g., star topologies in Figure \ref{fig:toys}), we initialize an interior mesh point which we treat as a non-movable end of each pipe created.
-We then route the pipes as normal.}
-
-\subsubsection{Pipes with Non-Round Endpoints}
-Combining pipe routing with other existing mesh modification operations like extrusions enables users to create custom-shaped touch pads or to allow an electronic component to sit flush with the print's surface.
-
-\subsubsection{Semi-Closed and Fully Enclosed Pipes}
-Users can cap one or both ends of their pipe.  The software simply creates an offset from the surface in these cases, so that material remains between the pipe and the surface after mesh modification.
+To allow designers to create pipes with multiple endpoints (e.g., star topologies in Figure \ref{fig:toys}), we use one surface position per pipe and a central fixed endpoint that the user can interactively position.
+The user can connect several  then route the pipes as normal.
 
@@
 
-\valkyrie{Once we have a tour, we must resolve tubes that intersect in the plane.  Because we have a 3-dimensional canvas in which to work, we can push overlapping paths into the third dimension.
-
-To detect intersecting edges, we \valkyrie{I'm not sure what we do.}.
-We also consider edges at nodes with degree $>2$, as these edges are considered to cross.
-When intersections have been detected, we just push things around.
-No, really.}
-
 \subsection{Mesh Modification}
 
-Given either our rod-based or image-based routing, we create cylinders which use the routing as a sweep path.
-At each point in the routing, we create a cylinder tall enough to reach to the next point, oriented along the line segment connecting its origin and destination.
-This cylinder's radius is set by the user.
-For the rod-based routing, we allow independent start and end radii; in this case, the cylinder's radius is $\frac{k}{n}*\frac{r_1}{r_2}$, where $k$ is the point's index in the routing, $n$ is the total number of points in the routing, $r_1$ is the starting radius, and $r_2$ is the ending radius.
+Once we have a set of authored paths, we must cut tubes out of the input mesh.
+Our basic strategy is to create a polygonal tube by sweeping a profile polyline
+along the path, and subtract the tube from the mesh. We only use circular profiles
+in our examples, but clearly any other profile could be used. Similarly our interface
+allows different start and end radii, which are linearly interpolated along the path.
 
-All cylinders generated are joined into a large rod.
-This rod is voxelized, along with the original mesh.
-These two voxel grids are subtracted, and the result is re-meshed and smoothed.
-For creating semi-closed pipes, during the voxelization subtraction we create an offset surface of the mesh where it intersects the tube and preserve it.
-
-Unfortunately, because we use voxelization as a part of our subtraction operation, our models can accumulate error when many pipes are created in sequence (see Figure \ref{fig:voxelize}).
-This problem can be solved with a higher voxel resolution, but requires balancing computation time and precision.
-
-\begin{figure}[h!]
-\centering
-    \includegraphics[width=3.4in]{figures/voxelize-fail.png}
-\caption{After repeated pipe creation, the mesh containing these pipes, which were not originally intersecting, accumulated sufficient error to open the pipes into each other.
-\bjoern{I don't understand this figure- probably requires multiple parts - ideal geometry and actual geometry}}
-\label{fig:voxelize}
-\end{figure}
+To avoid the complexities of direct mesh booleans, we construct a discretized volumetric
+(ie voxel-based) representation of the original mesh and the tube mesh. Boolean
+subtraction is trivial in such representations. We specifically use narrow-band
+level sets~\cite{Museth04}, which represent the surface as the zero iso-contour of an
+approximate distance field.
+We then use marching cubes~\cite{Lorensen87} to produce the final mesh for printing.
+This strategy does incur some resampling artifacts, however so does the 3D printing, and compared to the print process, the time required to voxelize at high resolution is inconsequential.
+Since we have a distance field, we can easily introduce useful effects, such as the
+ability to add a thin membrane "cap" over the end of a tube. 
+This is accomplished by applying a standard "thicken" operation to the level set version
+of the initial mesh, intersecting with a sphere placed at the tube endpoint, and then performing
+a boolean union with the initial shape-minus-tube result.
+A straightforward extension would be to add or subtract additional elements at
+the tube endpoints to make them non-round.
+ \ryan{Figure 8 is not necessary.
+This is a known problem with any voxelization method.
+Nobody is going to care that we took shortcuts in our implementation like repeatedly voxelizing and meshing instead of updating a single initial voxelization.}
 

### 2014-04-03 paper/exampleobjects.tex
working through comments, now time for george to read

--- 2014-03-31 paper/exampleobjects.tex
@@
 
-\tovi{Maybe a materials section 
-before this - to talk about the 
-specific materilas we used, 
-such as the copper paint 
-specifications, el lighting, etc.
-} \valkyrie{I'm not clear on what you are asking for here}
-
 \subsection{Touch-sensitive Toys (open, liquid, star, exterior)}
 
-\tovi{go into a bit more detail about modelling , fabrication, and
-assembly process, and apparatus details (especially the 
-copper paint - unless discussed earlier in a materials 
-seciton.}
-
-We created a set of touch-sensitive toys and a companion app, reminiscent of the boat application in Acoustic Barcodes \cite{Harrison-acoustic}.
-The brain, rabbit, and boat each have several distinct touch points on their surfaces.
-These touch points are connected by an interior star topology of tubes filled with conductive copper paint, and touch sensing is performed via a single wire and SFCS (see Figure \ref{fig:toys}).
+In order to demonstrate our star topology feature and show how conductive fluids can be used for touch sensing, we created a set of touch-sensitive toys and a companion app, reminiscent of the boat application in Acoustic Barcodes \cite{Harrison-acoustic}.
+The objects --- a brain and boat, in our example --- can be set on a base, which announces ``brain'' or ``boat'', and also the names of special features on the object when they are touched.
+For example, touching the brain model's front protrusion yields the announcement ``olfactory bulb''.
+The distinct touch points on each object are connected by an interior star topology of tubes filled with conductive copper paint, and touch sensing is performed via a single wire and SFCS (see Figure \ref{fig:toys}).
 We built a smart base which can distinguish between the toys and also determine which toy is mounted: since each toy and each gesture has a distinct capacitive signature, we use a simple classifier trained to detect both toy and gesture based on profile.
@@
 This paint requires approximately 2 days to fully dry in this configuration (3mm diameter tubes, maximum tube length 10cm).
-Our smart base is powered by an Arduino Uno running code by Instructables users DZL and madshobye.
+Our smart base is powered by an Arduino Uno running open-source SFCS code\footnote{http://www.instructables.com/id/Touche-for-Arduino-Advanced-touch-sensing/?ALLSTEPS}
 
@@
 
-We created a rabbit with a pair of tubes that can simulate breathing (see Figure \ref{fig:breathe}).
+To demonstrate how gases can be used to both provide sensation to the user through openings and to deform a model internally, we created a rabbit with a pair of tubes that can simulate breathing (see Figure \ref{fig:breathe}).
+When the rabbit inhales through its nose, its abdomen rises, and as it exhales its abdomen falls.
 For this, we used a combination air/vacuum pump: one terminal creates a vacuum while the other creates positive pressure.
-Our rabbit has two tubes, one open tube exiting at its mouth and one semi-closed tube capped with rubberlike material in its abdomen.
-We connected one tube to each of our pump's terminals, and using a programmable power supply we can mimic a rabbit's breathing pattern.
+Our rabbit has two pipes, one open tube exiting at its mouth and one semi-closed pipe capped with rubberlike material in its abdomen.
+We connected one pipe to each of our pump's terminals, and using a programmable power supply we mimic a rabbit's breathing pattern.
 This example was printed on our Objet in rubberlike material, with support flushed post-print.
-\valkyrie{this paragraph is really unclear}
 
@@
 
-A custom radio built using tubes allows users to tune to different stations and listen in.
-This device uses a network of disconnected open tubes filled with conductive paint to connect traditional electronic components (two potentiometers and a speaker) to a microcontroller (in our case, an Arduino Pro Micro).
+A custom radio, built using pipes integrated with traditional electronics, allows users to tune to different stations and listen in.
+This device uses a network of disconnected open tubes threaded with wires to connect our components (two potentiometers, an LED, and a speaker) to a microcontroller (Arduino Pro Micro).
 We hid the microcontroller and a battery-driven power supply in the base of the radio to make it portable.
-For the radio functionality, we attached an Si4703 to the Arduino, which offers headphone-jack output.
+For the radio functionality, we attached an Si4703 radio tuner breakout board to the Arduino, which offers headphone-jack output.
 We fabricated the radio itself on our Makerbot, support-free but cut in two pieces to allow the Arduino and battery pack to be inserted.
@@
 \valkyrie{this whole paragraph is unclear}
-
-\subsection{Maze (fully enclosed, particulate, tree, interior)}
-
-We created a maze game with a single particle trapped in a fully enclosed tube.
-This maze is based on an SVG we created by hand and processed with our internal path tube tool.
-To fabricate the tube, we created it in two halves that were fastened together via glue, due to limitations in our ability to remove support material otherwise.
-We created this prototype on our Makerbot in two halves, inserting a marble before closing.
 
@@
 The EL wire was threaded post-print, and is controlled by an EL wire sequencer.
+
+\subsection{Maze (fully enclosed, particulate, tree, interior)}
+
+Using a modification of our graph-based algorithm which only cuts the literal drawing (and does not perform Eulerization, etc.), we created a maze game with a single particle trapped in a fully enclosed tube.
+This maze is based on an SVG we created by hand and processed with our internal path tube tool.
+To fabricate the tube, we created it in two halves that were fastened together via glue, due to limitations in our ability to remove support material otherwise.
+\bjoern{observant reviewers will call us out on this - that many of our techniques may not actually be workable in a single piece because you can't get the support material out.
+I think if you had a lye bath you could actually get more complicated models done, as long as you're patient and there's some entry and exit for the lye to reach all pipes.}  \valkyrie{we should discuss.
+I will try to get a model of this ready ASAP}  We created this prototype on our Objet in two halves, inserting a marble before closing.

### 2014-04-03 paper/tubes.tex
working through comments, now time for george to read

@@
 \centering
-    \includegraphics[width=3.4in]{figures/interior.pdf}
+    \includegraphics[width=3.4in]{figures/interior.png}
 \caption{An input vector graphics file with the points which cannot be tubed as drawn highlighted in {\color{blue}blue} and the start and end points highlighted in {\color{gray}gray} (a).
@@
 of the initial mesh, intersecting with a sphere placed at the tube endpoint, and then performing
-a boolean union with the initial shape-minus-tube result.
+a boolean union with the initial shape-minus-tube result (see Figure \ref{fig:cap}).
 A straightforward extension would be to add or subtract additional elements at
@@
 Nobody is going to care that we took shortcuts in our implementation like repeatedly voxelizing and meshing instead of updating a single initial voxelization.}
+
+\begin{figure}[h!]
+\centering
+    \includegraphics[width=3.4in]{figures/cap.pdf}
+\caption{The process to ``cap'' the end of a pipe.
+In (a), a pipe end is placed on the surface of the mesh.
+In (b), we create a sphere whose radius is slightly larger than the radius of the pipe, centred at the tube's endpoint.
+In (c), we consider the intersection of the sphere and the pipe, and create an offset surface of the mesh.
+Finally, in (d) we intersect with the offset surface.
+This is kept as a cap for the pipe in mesh subtraction.}
+\label{fig:cap}
+\end{figure}
 

### 2014-04-03 paper/designspace.tex
working through comments, now time for george to read

@@
 \section{The Design Space of Pipes}
-\tovi{clean up writing}By using different openings, designs, and topologies of pipes, and inserting different media into the completed pipes, inputs and outputs can be located at arbitrary points on the surface of a 3D print.
-We describe the space below.
+The design space of pipes has four dimensions: openings, design focus, topologies, and inserted media.  We describe the space below.
 
@@
 Splitting or mixing pipes offer flexibility in output, allowing a single source to have multiple outputs or multiple sources to combine (see Figure \ref{fig:pipespace}).
-Star and tree topologies extend the splitting and mixing primitives, encompassing multiple-in and multiple-out in contrast to single-in/single-out.
-\tovi{give simple examples of where this might be useful}
+Star and tree topologies extend the splitting and mixing primitives, encompassing multiple-in and multiple-out in contrast to single-in/single-out; we used star topologies in our toys \ref{fig:toys}.
 
@@
 
-Incompressible fluids (``liquids'') perform similarly to gases.
-Additionally, using driable conductive liquids like copper paint, pipes can become custom wires.
-\tovi{custom water jet devices (uist 2013 competition)}
+Incompressible fluids (``liquids'') perform similarly to gases for display or haptics.  Additionally, using driable conductive liquids like copper paint, pipes can become custom wires.
 

### 2014-04-03 paper/discussion.tex
first set of glamour shots, paper work, redo of figures on new mm background, redo some figures due to new modeling
working through comments, now time for george to read

--- 2014-04-01 paper/discussion.tex
@@
 The promise of 3D printing is that it allows for the manufacture of geometries that are impossible to create using traditional techniques like injection molding.
-Pipes leverage that promise: designing the internal structure of an object was not previously feasible.
-As printers at the hobbyist and corporate levels become more sophisticated and able to fabricate intricate geometries of ever more materials, we believe that our tool will still assist in design tasks.
-Even if printers are capable of laying down conductive, optically transparent, or luminescent materials in-place, the process of designing their paths is still crucial.
+Pipes leverage that promise: they require a previously unmanufacturable complexity of interior geometry.
+Printers additionally continue to evolve in their capabilities to lay down more materials.
+As they become able to lay down conductive, optically clear, or luminescent materials, our tools and techniques will enable even more complex internal designs.
 
-By crossing pipe types, topologies, and inserted media, we create a space of possible inputs and outputs that can be pipe-mediated.
-We arrange our I/O discussion around the modes of input or output (the five senses, touch/tapping/pressure/etc.).
-We also suggest new points in the design space of pipe-mediated interaction that remain unexplored.
+By crossing pipe types, topologies, and inserted media, many pipe-mediated interactions emerge.
+We demonstrated a few such possibilities by designing and fabricating example objects with the help of our pipe routing design tool.
+However, there are many additional possibilities for leveraging sensing and actuation techniques from the HCI literature and extending their use in 3D-printed objects.
+We briefly survey such possibilities for output and input.
 
@@
 \emph{Visual} outputs can be mediated by gases, liquids, solid, particulates, or threadables.
-Pneumatic actuation (as in PneUIs \cite{Yao-pneui} or Harrison's latex buttons \cite{Harrison-buttons}) provides output, but gases carrying smoke (as in the Aireal \cite{Sodhi-aireal} demo) also offers visual feedback.
+Pneumatic actuation (as in PneUIs \cite{Yao-pneui} or Harrison's latex buttons \cite{Harrison-buttons}) provides output, but gases carrying smoke also offer visual feedback.
 Splitting and mixing of colored liquids can be used as a display tactic; additionally mechanical motion can be fluid-mediated.
 Liquids can also be mixed in specific concentrations by servo-controlled gate structures as in Figure \ref{fig:direct}.
-Printed optics \cite{Willis-printedoptics} explored visual output at the ends of pipes, and  EL wire can be threaded through custom paths to create neon signs, as in our neon robot (Figure \ref{fig:neon}).
 
@@
 Fluids in pipes capped with flexible membranes can actuate particulates and create programmable pliability as in Jamming User Interfaces \cite{Follmer-jamming}.
-Additionally, use of semi-closed pipes with rubberlike material as the caps allows for gases to actuate surface features, like the back of our breathing rabbit (see Figure \ref{fig:breathe}).
-Free-air haptic feedback as in Aireal \cite{Sodhi-aireal} and volumetric interactions as in Volflex \cite{Iwata-volflex} can also be pipe-mediated.
 
@@
 Traditional wires (threadable) may be used, but conductive paint (liquid) may be more flexible.
-Using conductive paint and Swept Frequency Capacitive Sensing (SFCS \cite{Sato-touche}), an interior star pipe topology can enable single-wire touch- and grasp-sensing at any set of points on a printed object's surface, like our toys (see Figure \cite{fig:toys}).
+Using conductive paint and SFCS \cite{Sato-touche}, an interior star pipe topology can enable single-wire touch- and grasp-sensing at any set of points on a printed object's surface, like our toys (see Figure \cite{fig:toys}).
 Simple capacitance measurements are possible using the same media, and custom-designed sensors like those created in Savage, et al.,'s Midas system are also possible \cite{Savage-midas}.
@@
 Components can be recessed into a print's surface, with their leads implanted into open pipes.
-By inserting liquid copper paint instead of threading traditional wires, all components in a model can share a ground line, and the paint's drying process obviates the use of solder or glue to affix the components in place (see Figure \ref{fig:radio}).
+By inserting liquid copper paint instead of threading traditional wires, all components in a model can share a ground line, and the paint's drying process obviates the use of solder or glue to affix the components in place.
 
@@
 
-We have also experimented with intentional encoding.
+We have experimented with intentional encoding.
 Semi-closed pipes can function as resonance chambers.
-An object's resonant frequency can be measured by attaching a speaker and microphone to it, sweeping frequencies with the speaker, and performing a Fourier transform on the resultant signal from the microphone.
+A pipe's resonant frequency can be measured by attaching a speaker and microphone to it, sweeping frequencies with the speaker, and performing a Fourier transform on the resultant signal from the microphone.
 Peaks in the transformed data correspond to stronger returned impulses: the resonant frequencies of the object.
@@
 While their technique requires access a terahertz imaging tool, audio resonance identity encoding requires only a microphone and a speaker.
+
+\bjoern{We might pull back on the scope of this section a bit and think about which directions here are most closely related to pipes and most likely successful;  and which are more tangential or speculative - for example, Aireal is about feedback in free space and we are all about designing tangible objects; so maybe that's a candidate for cutting.}
+
+\valkyrie{I cut a few things, and removed references to techniques we have already shown to work (as in the spirit of the intro).  LMK if this is better.}

### 2014-04-03 paper/introduction.tex
working through comments, now time for george to read

@@
 \section{Introduction}
-\tovi{Typical intro format: first pragraph should not introduce your own work, but summarize the problem and motivation.
-Second paragraph should summarize prior art and its limations.
-After that you can describe your work and contributions.
-I've made some revisions to the strucutre...}
 Makers, as well as professional designers, leverage 3D printers as tools for design work.
@@
 
-Due to the growing popularity of 3D printing and Maker Communities [ref?], recent work has looked at the issue of rapidly prototyping interactive objects.
-Existing approaches include ...
-but their limitations are ...
+Due to the growing popularity of 3D printing and maker communities like Thingiverse\footnote{http://thingiverse.com}, recent work has looked at the issue of rapidly prototyping interactive objects.
+Existing approaches include Printed Optics \cite{Willis-printedoptics}, which uses optically clear material for light redirection, and Sauron \cite{Savage-sauron}, which is based on computer vision, but they are limited to visible interactions only; they do not give designers freedom to integrate haptic or acoustic inputs or outputs.
 
@@
 This {\em subtractive} approach is complementary to {\em additive} approaches that inject custom print materials, such as conductors, during the actual printing process [ref].
-While our approach requires some manual assembly after a print completes, it gives makers the opprotunity to rapidly prototype a diverse range of interactive objects on commonly available 3D printers (Figure \ref{fig:teaser}).\tovi{I guess we need to be careful here since not all of our examples are on doen with commonly available 3D printers}
+While our approach requires some manual assembly after a print completes, it gives makers the opprotunity to rapidly prototype a diverse range of interactive objects on 3D printers, including hobbyist machines (Figure \ref{fig:teaser}).
 
@@
 By exploring this design space new opportunities for adding interactivity to 3D printed objects can be realized.
-For example, copper material can fill the channels, to allow for standard electronic components to be easily integrated after the printout; electroluminescent (EL) wire can be threaded through to create computer-controlled visual output; or air can be pumped through to control haptic responses.\tovi{at some point explain why this is better than running wires on the exterior (looks better, for one)}
+For example, copper material can fill the channels, to allow for standard electronic components to be easily integrated after the printout; electroluminescent (EL) wire can be threaded through to create computer-controlled visual output; or air can be pumped through to control haptic responses.
+In comparison to traditional exterior electronics integration, our process allows makers to preserve aesthetics by hiding wires inside the object; additionally it offers more freedom in selecting precise surface locations for, e.g., air mediated tactile output.
 
-\tovi{I would remove this entire paragraph}
+\tovi{I would remove this entire paragraph} \valkyrie{I think we at least need to preserve the sentiment from the end, that we don't have any novel I/O techniques but that pipes help extend existing techniques in useful ways}
 Pipes allow makers to build with and upon existing input/output techniques, often integrating these methods into objects they were not originally designed for.
@@
 
-\tovi{I would remove this entire paragraph}
+\tovi{I would remove this entire paragraph}  \valkyrie{if we scrap the first one, sure.  George, what do you think?}
 Pipes can also be used for sending media on a specified route through an object; for example, neon signs are created by evacuated glass tubes formed into a specific path.

### 2014-04-03 paper/limitations.tex
working through comments, now time for george to read

--- 2014-03-27 paper/limitations.tex
@@
 Some media are difficult to insert in pipes, or may not be compatible with all types of pipes.
-Fluids in particular provide challenges.
-For example, the conductive paint we used in our designs can take multiple days to dry in tubes of sufficient length.
+Threadable materials in particular can provide challenges, as they can get stuck while being fed into pipes.
+Fluids also  For example, the conductive paint we used in our designs can take multiple days to dry in tubes of sufficient length.
 Any liquid kept in tubes printed by the Objet can cause the exterior of the part to discolor due to seepage.
 Hobbyist machines do not guarantee watertightness of their prints, so leaking of water or air can occur.
+
+As discussed, support material removal can also be a challenge.
+Particularly on multi-jetting printers, printing support-free is not possible for most geometries, while for hobbyist-grade printers physically detaching printed support is not always trivial.
 

### 2014-04-03 paper/relatedwork.tex
working through comments, now time for george to read

@@
 
-\tovi{dont use citations as nouns}
-Slyper, et al., in \cite{Slyper-pressure} 3D printed flexible robot armatures capable of sensing interaction via changes in air pressure.
-\tovi{clean up writing - finish telling the reader what they did or didnt do, then respond with what we do differently}This technique is complementary to ours, as we allow placing these interactive elements anywhere on an object's surface; they had to place them in locations accessible for electronics insertion.
-\tovi{not sure this is relevant enough to cite here}Acoustic Barcodes explored integrating ridges into fabricated objects, whose acoustic signatures when scraped provided an ID code \cite{Harrison-acoustic}; while pipes can be used for identification and to create aural feedback, this is not our only goal.
+Slyper, et al., 3D printed flexible robot armatures capable of sensing interaction via changes in air pressure \cite{Slyper-pressure}.
+They created all their models by hand, and were forced to place interaction locations in areas accessible to electronics assembly: this technique is complementary to ours, as we allow placing these interactive elements anywhere on an object's surface.
 Sauron investigated computer vision to track interactions with physical mechanisms \cite{Savage-sauron}.
@@
 
-\subsubsection{Fabricating Channels}
+\subsubsection{Fabrication + Electronics}
 
@@
 Sells, et al., \cite{Sells-reprap}, describe using a heated syringe to deposit metal in printed 2D circuitboard channels on objects' surfaces.
-Sarik, et al., \cite{Sarik-tracebrush} and the Optomec company \cite{optomec} describe \emph{spraying} conductive material onto objects to create circuitry.
+Sarik, et al., \cite{Sarik-tracebrush} and the Optomec company \cite{optomec} describe \emph{spraying} conductive material onto printed objects to create circuitry.
 Injection of liquid metals for sensing in softer-than-skin electronics was explored by Park, et al., \cite{Park-microchannels} and Majidi, et al., \cite{Majidi-curvature}.
@@
 There is prior work on modeling the interior of objects (as opposed to their exterior).
-In Make it Stand, material is removed from the interior of objects with the goal of changing their center of gravity \cite{Prevost-makeitstand}.
+Takayama, et al., built a system described in \cite{Takayama-fruit} which allowed users to create internal color maps of the interior of objects, allowing them to see digital ``slices'' of their ``oranges.''  Our work focuses on 3D printable geometry rather than rendering.
+In Make it Stand, material is removed from the interior of objects with the goal of changing their center of gravity \cite{Prevost-makeitstand}, and Infrastructs created encoded interior cavities in 3D printed objects designed to be sensed with a terahertz imaging tool.
+Modeling such chambers is one aspect of our work, however our focus is on interaction rather than balance or identification.
 Bickel, et al., explored enforcing deformation behavior with material blending \cite{Bickel-deformation}, while B\"{a}cher, et al., built an algorithm for automatically placing 3D fabricatable joints into meshes \cite{Bacher-articulated}.
-Takayama, et al., built a system described in \cite{Takayama-fruit} which allowed users to create internal color maps of the interior of objects, allowing them to see digital ``slices'' of their ``oranges.''  Sauron automatically modified the interior of existing mechanical objects to make them amenable to vision-based sensing \cite{Savage-sauron}, and Infrastructs created encoded interior cavities in 3D printed objects designed to be sensed with a terahertz imaging tool.
-All these contributions are orthogonal to ours, as our focus is on modeling pipes and hollow chambers for fabricating interactive objects.
-\tovi{need better summary statement instead of ``all of this is orthogonal''}
+Sauron automatically modified the interior of existing mechanical objects to make them amenable to vision-based sensing \cite{Savage-sauron}.
+All these contributions are orthogonal to ours, as our focus is on precisely \emph{removing} material from printed objects, rather than blending materials or modifying components.

### 2014-04-03 paper/designspace.tex
fiddling, integrating tovi comments, update script

@@
     \includegraphics[width=3.4in]{figures/tubespace.pdf}
-\caption{The design space of pipes.  Pipe openings, design, topologies, and media are discussed more fully in the text.}
+\caption{The design space of pipes.  See text for details.}
 \label{fig:pipespace}
@@
 Pipes can have four opening styles: open, return, semi-closed, and fully enclosed (see Figure \ref{fig:pipespace}).
-As we are interested in interactive devices, we distinguish a ``user side"---part of the surface or interior of an object facing its user; and a ``system side"---where other hardware such as electronics may be connected.
+As we are interested in interactive devices, we distinguish a ``user side"---part of the surface or interior of an object facing its user for interaction; and a ``system side"---where other hardware such as electronics may be connected.
 We refer to a pipe which is ``open on the user side'' to have its contents \emph{physically accessible} to a user \emph{at interaction time}.
@@
 User or system manipulation of pipe contents are possible even without physical access, e.g., via soundwaves.
-\tovi{still not convinced we need this distinction}
 
@@
 \subsection{Design of Pipes}
-Pipe design may emphasize either exterior connection points or interior paths.
-Focused on the exterior features is the touch sensitive toy in Figure \ref{fig:toys}, where pipes must exit the toy at the frontal lobe and parietal lobes to make those \emph{specific locations} touch-sensitive.
-Alternatively, the neon sign's output in Figure \ref{fig:neon} is based on the \emph{path} of the pipe.
-\tovi{still think we can remove this dimension}
+Pipe design may emphasize exterior connection points and/or interior paths.
+Our maze (see Figure \ref{fig:maze}) requires specific terminals (the start and end of the maze) as well as a specific path for its tubes (the maze itself).
+Alternatively, the neon sign's output in Figure \ref{fig:neon} is based only on the \emph{path} of the pipe.
+Focused on just the exterior features is the touch sensitive toy in Figure \ref{fig:toys}, where pipes must exit the toy at the frontal lobe and parietal lobes to make those \emph{specific locations} touch-sensitive.
+A pipe for weight modification does require specific terminals or a specific path (see Figure \ref{fig:openings}D).
 

### 2014-04-04 paper/exampleobjects.tex
figure is back

@@
-\section{Example Objects}
-
-To evaluate and highlight our tool's capabilities, we fabricated a set of five prototype objects designed using a Series of Tubes.
-All prototypes were fabricated in a single piece, unless described otherwise.
-
-\begin{figure*}[h!]
+\begin{figure*}
 \centering
@@
 
+\section{Example Objects}
+
+To evaluate and highlight our tool's capabilities, we fabricated a set of five prototype objects designed using a Series of Tubes.
+All prototypes were fabricated in a single piece, unless described otherwise.
+
 \subsection{Touch-sensitive Toys (open, liquid, star, exterior)}
@@
 This paint requires approximately 2 days to fully dry in this configuration (3mm diameter tubes, maximum tube length 10cm).
-Our smart base is powered by an Arduino Uno running open-source SFCS code\footnote{http://www.instructables.com/id/Touche-for-Arduino-Advanced-touch-sensing/?ALLSTEPS}
+Our smart base is powered by an Arduino Uno running open-source SFCS code\footnote{http://www.instructables.com/id/Touche-for-Arduino-Advanced-touch-sensing/?ALLSTEPS}.
 
@@
 A custom radio, built using pipes integrated with traditional electronics, allows users to tune to different stations and listen in.
-This device uses a network of disconnected open tubes threaded with wires to connect our components (two potentiometers, an LED, and a speaker) to a microcontroller (Arduino Pro Micro).
+This device uses a network of disconnected open pipes threaded with wires to connect our components (two potentiometers, an LED, and a speaker) to a microcontroller (Arduino Pro Micro).
 We hid the microcontroller and a battery-driven power supply in the base of the radio to make it portable.
 For the radio functionality, we attached an Si4703 radio tuner breakout board to the Arduino, which offers headphone-jack output.
+
+In order to create so many pipes (3 per potentiometer, 1 for the LED, 1 for the speaker) and still preserve surface quality after voxelizing, we added a "High Quality" for pipe-cutting.
+This mode voxelizes at a resolution of 512x512x512 instead of our previous 128x128x128.
+It does preserve surface geometry, however it requires around 3 minutes to complete a single cut operation, compared to our nearly-instant smaller voxel grid (computation on a 3D grid increases by the cube of the 1D size increase).
 We fabricated the radio itself on our Makerbot, support-free but cut in two pieces to allow the Arduino and battery pack to be inserted.
@@
 
-\subsection{Animated Neon Sign (semi-closed, threadable, interior)}
+\subsection{Animated Neon Sign (return, threadable, interior)}
 
@@
 
-\subsection{Maze (fully enclosed, particulate, tree, interior)}
+\subsection{Maze (fully enclosed, particulate, tree, interior + exterior)}
 

### 2014-04-04 paper/discussion.tex

@@
 Both identification by recall and intentional encoding are possible.
-As seen in \cite{Ono-touchandactivate}, different objects have different acoustic signatures.
+As seen in Touch \& Activate \cite{Ono-touchandactivate}, different objects have different acoustic signatures.
 Additionally, two objects that are visually identical but which have fully enclosed (or other types) of pipes on their interior can be distinguished acoustically.
 Thus we can recall an object's identity once its acoustic signature has been recorded.
-Similarly, the Touch\'{e} system in \cite{Sato-touche} relies on distinguishing capacitive signatures of objects, both in static states and as their boundary conditions are changed by human interaction.
+Similarly, the Touch\'{e} system \cite{Sato-touche} relies on distinguishing capacitive signatures of objects, both in static states and as their boundary conditions are changed by human interaction.
 Weight is a simple metric of object identity, but weights are less distinctive than acoustic or capacitive signatures, as they are represented by a single number.
@@
 Encoding of weight is trivial by creating fully-enclosed chambers inside an object.
-
-This is similar to work done by Willis, et al., in \cite{Willis-infrastructs}.
+This is similar to Infrastructs \cite{Willis-infrastructs}.
 While their technique requires access a terahertz imaging tool, audio resonance identity encoding requires only a microphone and a speaker.
 
+=======
+>>>>>>> bd673f22cceca8b1cbc4a00aae3717eacd7fd9f2
 \bjoern{We might pull back on the scope of this section a bit and think about which directions here are most closely related to pipes and most likely successful;  and which are more tangential or speculative - for example, Aireal is about feedback in free space and we are all about designing tangible objects; so maybe that's a candidate for cutting.}

### 2014-04-04 paper/designspace.tex
clean up writing, re-add ryan text

@@
 C shows two semi-closed pipes capped by rubber membranes.
-In D, a rabbit has a fully enclosed chamber to modify its weight.
-\bjoern{Great, but make figure  full column wide}}
+In D, a rabbit has a fully enclosed chamber to modify its weight.}
 \label{fig:openings}
@@
 \subsection{Media in Pipes}
-\tovi{clean up this writing}
-One of the most interesting aspects of the design space is that pipes can be filled with a variety of media to create different interface affordances and capabilities.
-We consider gas, liquids, solids, particulates, and threadables.
+After printing, there is still flexibility in inserting media to create different interface affordances and capabilities. We consider gas, liquids, solids, particulates, and threadables.
 
@@
 
-Threadable media, such as EL wire or fiberoptic cables, are those that can be threaded through pipes post-printing.
+Threadable media, such as EL wire or fiberoptic cables, can be threaded through pipes post-printing.
 This empowers inexpensive printers: for example, a Printed Optics-style interface can be created on a consumer-grade 3D printer with pipes and fiber optic cable.

### 2014-04-04 paper/exampleobjects.tex
mmodz
photoshopize photos for nicer-ness
clean up writing, re-add ryan text

@@
 \caption{A series of example objects created using our system.
-A) shows a portable radio.
-B) shows a presence-aware pen holder.
-C) is a rabbit which ``breathes'' using wind pipes we built.
-D) shows a maze game we created.
-E) is a touch-sensitive brain toy.
-\valkyrie{finalize order of example objects.
-what order is best?}}
+(a) is a touch-sensitive brain toy.
+(b) is a rabbit which ``breathes'' using wind pipes we built.
+(c) shows a portable radio.
+(d) shows a presence-aware pen holder.
+(e) is a maze game.}
 \label{fig:examples}
@@
     \includegraphics[width=3.4in]{figures/sign.png}
-\caption{A two-state animated neon sign designed using our tool (a)(b) \valkyrie{Tovi: we could use your PDF hax to make this animate in place :)}.
-(c) shows the SVG input files to create this sign, and (d) shows an isometric view of the tubes created by our tool.
-\valkyrie{d) is a lot less interesting if we are making planar tubes.
-also, for c, should we show an overlay of the two states?
-maybe in different colors?}}
+\caption{A two-state animated neon sign designed using our tool.
+(a) shows the input SVG files, with the two states of the mouth drawn in different colors.
+(b) shows the resultant mesh when we subtracted the generated tubes.
+In (c), we show the completed sign.}
 \label{fig:neon}

### 2014-04-04 paper/tubes.tex
paper modifications, and new figures!
clean up writing, re-add ryan text

@@
 wire (a 3D poly-line or \emph{discrete rod}), and then set its initial position 
-to be deformed to lie along our A* path.
+as our A* path.
 Running the physical simulation allows the wire to attempt to return to
-its straight configuration, while satisfying endpoint position constraints.
+a straight configuration, while satisfying its endpoint position constraints.
 We also increase the length of the wire, to provide some slack in cases where 
@@
 
-The wire is modelled as a \emph{discrete rod}, ie a 3D poly-line with a bending 
+The wire is modelled as a \emph{discrete rod}, i.e., a 3D poly-line with a bending 
 constraint between adjacent segments. Our simulation is based on an
@@
 \centering
-    \includegraphics[width=3.4in]{figures/interior.png}
+    \includegraphics[width=3.4in]{figures/interior.pdf}
 \caption{An input vector graphics file with the points which cannot be tubed as drawn highlighted in {\color{blue}blue} and the start and end points highlighted in {\color{gray}gray} (a).
@@
 \label{fig:tool-process-interior}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
+    \includegraphics[width=3.4in]{figures/uistphotos.png}
+\caption{Our neon sign built using the process above.  (a) shows the unlit sign, and (b) shows it lit.}
+\label{fig:UIST}
 \end{figure}
@@
 This minimizes turns in the final artifact, which eases support material removal and assembly.
+Our EL wire in Figure \ref{fig:UIST} follows an Euler tour through the input nodes.
 
@@
 \subsection{Mesh Modification}
-
 Once we have a set of authored paths, we must cut tubes out of the input mesh.
@@
 To avoid the complexities of direct mesh booleans, we construct a discretized volumetric
-(ie voxel-based) representation of the original mesh and the tube mesh. Boolean
+(i.e., voxel-based) representation of the original mesh and the tube mesh. Boolean
 subtraction is trivial in such representations. We specifically use narrow-band
@@
 Since we have a distance field, we can easily introduce useful effects, such as the
-ability to add a thin membrane "cap" over the end of a tube. 
-This is accomplished by applying a standard "thicken" operation to the level set version
+ability to add a thin membrane ``cap'' over the end of a tube. 
+This is accomplished by applying a standard ``thicken'' operation to the level set version
 of the initial mesh, intersecting with a sphere placed at the tube endpoint, and then performing
@@
 A straightforward extension would be to add or subtract additional elements at
-the tube endpoints to make them non-round.
- \ryan{Figure 8 is not necessary.
-This is a known problem with any voxelization method.
-Nobody is going to care that we took shortcuts in our implementation like repeatedly voxelizing and meshing instead of updating a single initial voxelization.}
+the tube endpoints.
 

### 2014-04-05 paper/abstract.tex
spelling
transcribing some george comments

@@
 However, a 3D printer's role in prototyping interactive objects is still not clear.
-We propose a general technique for supporting the rapid prototyping of interactivity by removing interior material of 3D models to form pipes and cavities.
-We describe this new design space of pipes and hollow chambers for interaction design, where variables include openings, topologies, and inserted media.
+We propose a general technique for supporting the rapid prototyping of interactivity by removing interior material of 3D models to form pipes and chambers.
+We describe this new design space of pipes and hollow chambers for interaction design, where variables include openings, topologies, design focus, and inserted media \george{this needs more description or explanation}.
 We then present a  tool for routing such pipes through the interior of 3D models, integrated within a 3D modeling program.
 We use two distinct routing algorithms.
-The first allows users to define the pipes' terminals, and then uses A* path routing and physics-based simulation to route the paths with minimized  bending energy.
+The first allows users to define the pipes' terminals, and then uses A* path routing and physics-based simulation to route the paths with minimized bending energy, to allow easy insertion of media post-print.
 The second allows users to supply a desired internal shape which we fit a pipe route to: for this we developed a novel graph-based routing algorithm.

### 2014-04-05 paper/designspace.tex
transcribing some george comments

@@
 
-Pipes can have four opening styles: open, return, semi-closed, and fully enclosed (see Figure \ref{fig:pipespace}).
+\george{difference between open/return not clear}Pipes can have four opening styles: open, return, semi-closed, and fully enclosed (see Figure \ref{fig:pipespace}).
 As we are interested in interactive devices, we distinguish a ``user side"---part of the surface or interior of an object facing its user for interaction; and a ``system side"---where other hardware such as electronics may be connected.
@@
 User or system manipulation of pipe contents are possible even without physical access, e.g., via soundwaves.
+\george{confusing}
 
@@
 By threading an electroluminescent (EL) wire through a clear return pipe, a maker can create a custom piece of neon art (see Figure \ref{fig:openings}B).
+\george{not sure this is adding something useful}
 
@@
 
-Pipe network topologies enable different interactions.
+Pipe network topologies enable different interactions \george{how?}.
 Splitting or mixing pipes offer flexibility in output, allowing a single source to have multiple outputs or multiple sources to combine (see Figure \ref{fig:pipespace}).

### 2014-04-05 paper/introduction.tex
transcribing some george comments

@@
 Due to the growing popularity of 3D printing and maker communities like Thingiverse\footnote{http://thingiverse.com}, recent work has looked at the issue of rapidly prototyping interactive objects.
-Existing approaches include Printed Optics \cite{Willis-printedoptics}, which uses optically clear material for light redirection, and Sauron \cite{Savage-sauron}, which is based on computer vision, but they are limited to visible interactions only; they do not give designers freedom to integrate haptic or acoustic inputs or outputs.
+Existing approaches include Printed Optics \cite{Willis-printedoptics}, which uses optically clear material for light redirection, and Sauron \cite{Savage-sauron}, which is based on computer vision, but they are limited to visible interactions only \george{what does this mean?}; they do not give designers freedom to integrate haptic or acoustic inputs or outputs \george{huh?
+why not?}.
 
@@
 This {\em subtractive} approach is complementary to {\em additive} approaches that inject custom print materials, such as conductors, during the actual printing process [ref].
-While our approach requires some manual assembly after a print completes, it gives makers the opprotunity to rapidly prototype a diverse range of interactive objects on 3D printers, including hobbyist machines (Figure \ref{fig:teaser}).
+While our approach requires some manual assembly after a print completes, it gives makers the opprotunity to cheaply and rapidly prototype a diverse range of interactive objects on 3D printers, including hobbyist machines (Figure \ref{fig:teaser}).
 
-We describe a new design space of pipes and hollow chambers for the prupose of interaction design, where variables include openings, topologies, and inserted media.
+We describe a new design space of pipes and hollow chambers for the prupose of interaction design, where variables include openings, topologies, and inserted media \george{need more detail}.
 By exploring this design space new opportunities for adding interactivity to 3D printed objects can be realized.
@@
 
-\tovi{I would remove this entire paragraph} \valkyrie{I think we at least need to preserve the sentiment from the end, that we don't have any novel I/O techniques but that pipes help extend existing techniques in useful ways}
+\tovi{I would remove this entire paragraph} \valkyrie{I think we at least need to preserve the sentiment from the end, that we don't have any novel I/O techniques but that pipes help extend existing techniques in useful ways} \george{remove, repeats}
 Pipes allow makers to build with and upon existing input/output techniques, often integrating these methods into objects they were not originally designed for.
-For example, to create a touch-sensitive toy as in Figure \ref{fig:teaser}, a maker can select several locations of interest on the brain.
+For example, to create a touch-sensitive toy as in Figure \ref{fig:teaser}, a maker can select several locations of interest on the brain \george{?}.
 After creating pipes to those locations and printing her object, she can fill the pipes with conductive paint.
@@
 In this way, she has only a single active element (the microcontroller setup), but her otherwise-inert 3D printed object (the brain) has several sensing locations.
-Other existing sensing and actuation approaches, such as FlyEye \cite{Wimmer-flyeye} and Jamming User Interfaces \cite{Follmer-jamming}, can also be enhanced by the use of pipes: makers utilizing such I/O strategies can locate input and output at arbitrary points on an print's surface.
+Other existing sensing and actuation approaches, such as FlyEye \cite{Wimmer-flyeye} and Jamming User Interfaces \cite{Follmer-jamming}, can also be enhanced by the use of pipes: makers utilizing such I/O strategies can locate input and output at arbitrary points on a printed 3D object's surface.
 We leverage these existing techniques for sensing and actuation while our work's novelty is in \emph{internal modeling} for 3D printing and the \emph{redirection} of I/O to arbitrary locations on a device.
 
-\tovi{I would remove this entire paragraph}  \valkyrie{if we scrap the first one, sure.  George, what do you think?}
+\tovi{I would remove this entire paragraph}  \valkyrie{if we scrap the first one, sure.  George, what do you think?} \george{remove}
 Pipes can also be used for sending media on a specified route through an object; for example, neon signs are created by evacuated glass tubes formed into a specific path.
@@
 Pipes should not pass too close to the surface: consumer-grade 3D printers do not guarantee air- or water-tightness of prints, so more than one printed layer may be required to prevent leakage.
-Pipes for applications like neon signs must have a Euler tour through them: this allows a single EL wire to light every edge on the pipe graph.
+Pipes for applications like neon signs must have a Euler tour \george{explain or ref} through them: this allows a single EL wire to light every edge on the pipe graph.
 No tools currently exist for these design tasks.
@@
 The system takes into considertation the design complexities of routing interior pipes which are described above.
-Finally, we showcase a set of examples, enabled by our modeling tool, which allows us to explore various points in the design space.
-We conlcude by summarizing the input and out modalities enabled by our work, and discussing limitations and areas for future exploration.
+Finally, we showcase a set of examples, enabled by our modeling tool, which allows us to explore various points in the design space \george{what design space?}.
+We conclude by summarizing the input and output modalities enabled by our work, and discussing limitations and areas for future exploration.

### 2014-04-05 paper/relatedwork.tex
fiddling, integrating tovi comments, update script
transcribing some george comments

@@
 Printed Optics (\cite{Willis-printedoptics}) used light pipes, solid clear tubes, to create integrated displays and touch sensors.
-The describes a process, but does not provide a design tool for creating such objects.
+While the research describes a process, it does not provide a design tool for creating such objects.
 Our work is inspired by Printed Optics and introduces a design tool for modelling objects that can have these types of capabilities.
@@
 Slyper, et al., 3D printed flexible robot armatures capable of sensing interaction via changes in air pressure \cite{Slyper-pressure}.
-They created all their models by hand, and were forced to place interaction locations in areas accessible to electronics assembly: this technique is complementary to ours, as we allow placing these interactive elements anywhere on an object's surface.
+They created all their models by hand, and were forced to place interaction locations in areas accessible to electronics assembly: this technique is complementary to ours, as we allow placing interactive elements anywhere on an object's surface.
 Sauron investigated computer vision to track interactions with physical mechanisms \cite{Savage-sauron}.
-While Savage, et al., used mirrors for redirection, we believe that the required line-of-sight operation for their computer vision techniques may be possible using printed optics or inserted optical fibers.
+While Savage, et al., used mirrors for redirection, we believe that the required line-of-sight operation for their computer vision techniques may be possible using printed optics or inserted optical fibers \george{huh?}.
 
@@
 While we recognize that 3D printing requires significant time, we believe creating precise internal structures in 3D printed objects is a strong case for digital fabrication over manual or traditional methods.
+\george{wording}
 
@@
 
-Other research has investigated the integration of traditional electronics into 3D prints.
+Other research has investigated the integration of traditional electronics into 3D prints by depositing non-printable materials after printing.
 Sells, et al., \cite{Sells-reprap}, describe using a heated syringe to deposit metal in printed 2D circuitboard channels on objects' surfaces.
@@
 Our work differs from these because we offer \emph{algorithms and techniques} for \emph{internal}, \emph{true 3D} routing.
-\tovi{relation of this stuff to ours isn't so clear}
 
@@
 Its approach of linking edges into a continuous line based on semi-Eulerization is similar to ours.
-However, we are not confined to the plane to avoid line-crossings, and we can create ``invisible'' lines.
-This is possible for us thanks to \emph{blocking out}\footnote{In neon sign making, segments of a tube that should not appear illuminated are covered with opaque paint.
+However, we are not confined to the plane to avoid line-crossings, and we can create ``invisible'' lines using \emph{blocking out}\footnote{In neon sign making, segments of a tube that should not appear illuminated are covered with opaque paint.
 This is called "blocking out".
-\url{http://en.wikipedia.org/wiki/Neon_sign}}.
+\url{http://en.wikipedia.org/wiki/Neon_sign}} \george{too much detail and talk of our work}.
 Prior work used tours of nodes, rather than edges, and cast continuous line drawing as a traveling salesman problem~\cite{Bosch-tsp}.

### 2014-04-05 paper/abstract.tex
george comments

@@
 We propose a general technique for supporting the rapid prototyping of interactivity by removing interior material of 3D models to form pipes and chambers.
-We describe this new design space of pipes and hollow chambers for interaction design, where variables include openings, topologies, design focus, and inserted media \george{this needs more description or explanation}.
+We describe this new design space of pipes and hollow chambers for interaction design, where variables include openings, topologies, design focus, and inserted media.
 We then present a  tool for routing such pipes through the interior of 3D models, integrated within a 3D modeling program.
 We use two distinct routing algorithms.
-The first allows users to define the pipes' terminals, and then uses A* path routing and physics-based simulation to route the paths with minimized bending energy, to allow easy insertion of media post-print.
-The second allows users to supply a desired internal shape which we fit a pipe route to: for this we developed a novel graph-based routing algorithm.
+The first allows users to define the pipes' terminals, and then uses path routing and physics-based simulation to minimize pipe bending energy, allowing easy insertion of media post-print.
+The second allows users to supply a desired internal shape to which we fit a pipe route: for this we developed a novel graph-based routing algorithm.
 We present several totally tubular prototypes we created using our tool to show its flexibility and potential.

### 2014-04-05 paper/designspace.tex
george comments

@@
 Similarly with the system side.
-User or system manipulation of pipe contents are possible even without physical access, e.g., via soundwaves.
+User or system manipulation of pipe contents are possible even without physical access: e.g., via soundwaves.
 \george{confusing}
@@
 
-Pipe network topologies enable different interactions \george{how?}.
-Splitting or mixing pipes offer flexibility in output, allowing a single source to have multiple outputs or multiple sources to combine (see Figure \ref{fig:pipespace}).
+Pipe network topologies enable different interactions: splitting or mixing pipes offer flexibility in output, allowing a single source to have multiple outputs or multiple sources to combine (see Figure \ref{fig:pipespace}).
 Star and tree topologies extend the splitting and mixing primitives, encompassing multiple-in and multiple-out in contrast to single-in/single-out; we used star topologies in our toys \ref{fig:toys}.
@@
 
-Pipes need not have hollow centers: in the case where routed pipes are filled with solid material---in particular, a solid material different from the enclosing model material---, interactions such as those in Printed Optics \cite{Willis-printedoptics} are possible.
+Pipes need not have hollow centers: if routed pipes are filled with solid material---in particular, a solid material different from the enclosing model material---, interactions such as those in Printed Optics \cite{Willis-printedoptics} are possible.
 
-Particulates, either printed in-place or inserted, can be of varying densities.  A single particle can be used for display.  Sparse particles in fluid can provide haptic feedback.
+Particulates, either printed in-place or inserted, can be of varying densities and can contain particles of various and variable size.
+A single large particle can be used for display.
+Small particles in fluid can provide variable stiffness haptics via jamming \cite{Follmer-jamming}.
 

### 2014-04-05 paper/exampleobjects.tex
george comments

@@
 
-\subsection{Touch-sensitive Toys (open, liquid, star, exterior)}
+\subsection{Touch-sensitive Toys (open, liquid, star, terminals)}
 
@@
 
-\subsection{Breathing Bunny (semi-closed, gas, exterior)}
+\subsection{Breathing Bunny (semi-closed, gas, terminals)}
 
@@
 
-\subsection{Custom Radio (open, threadable, exterior)}
+\subsection{Custom Radio (open, threadable, terminals)}
 
@@
 
-\subsection{Presence-aware Pen Holder (open, solid, exterior)}
+\subsection{Presence-aware Pen Holder (open, solid, ternimals)}
 
@@
 
-\subsection{Animated Neon Sign (return, threadable, interior)}
+\subsection{Animated Neon Sign (return, threadable, path)}
 
@@
 We designed a custom neon robot head which can be animated to ``talk''.
-The piece features semi-open pipes: one end must remain open to connect the EL wire to its power source.
 The pipes have been threaded with EL wire which is lit in sequence to create an animation.
-Our sign was fabricated on the Objet in two halves, cutting through the plane of the sign itself, with clear material for the sign around the input design and opaque everywhere else.
-We manually flushed support from the non-planar connecting pipes.
+Our sign was fabricated on the Objet in two halves, cutting through the plane of the sign itself, and with pipes cut in the rear piece to allow the wire to exit backwards and hide the EL wire controls.
 The EL wire was threaded post-print, and is controlled by an EL wire sequencer.
 
-\subsection{Maze (fully enclosed, particulate, tree, interior + exterior)}
+\subsection{Maze (fully enclosed, particulate, tree, terminals + path)}
 

### 2014-04-05 paper/introduction.tex
george comments

@@
     \includegraphics[width=3.4in]{figures/placeholder/teaser.png}
-\caption{A novel neon sign (a) and a touch-sensitive brain (b) designed with our pipe tool and fabricated on a consumer-grade 3D printer.
-(c) and (d) show the structure of the internal pipes generated by our tool.
-\bjoern{will this be the robot or the UIST sign?} \valkyrie{arguably, the robot will be nicer-looking.
-I think I'm manufacturing the UIST sign on the Makerbot.}}
+\caption{A collection of interactive objects designed using our tool.
+All have electronic sensing or actuation components which are routed through the \emph{interior} of the prints to preserve aesthetics.
+In (a), objects manufactured on a hobbyist 3D printer.
+In (b), those manufactured on industrial-grade printers.}
 \label{fig:teaser}
@@
 Due to the growing popularity of 3D printing and maker communities like Thingiverse\footnote{http://thingiverse.com}, recent work has looked at the issue of rapidly prototyping interactive objects.
-Existing approaches include Printed Optics \cite{Willis-printedoptics}, which uses optically clear material for light redirection, and Sauron \cite{Savage-sauron}, which is based on computer vision, but they are limited to visible interactions only \george{what does this mean?}; they do not give designers freedom to integrate haptic or acoustic inputs or outputs \george{huh?
-why not?}.
+Existing approaches include Printed Optics \cite{Willis-printedoptics}, which uses optically clear material for light redirection, and Sauron \cite{Savage-sauron}, which is based on computer vision, but they are limited to visible (reflecting light off fingers, mechanical motion of components) input and visual (screen-like) output; they do not give makers freedom to integrate haptic feedback or other inputs or outputs.
 
@@
 
-We describe a new design space of pipes and hollow chambers for the prupose of interaction design, where variables include openings, topologies, and inserted media \george{need more detail}.
+We describe a new design space of pipes and hollow chambers for the purpose of interaction design, where variables include openings, topologies, and inserted media.
 By exploring this design space new opportunities for adding interactivity to 3D printed objects can be realized.
@@
 
-\tovi{I would remove this entire paragraph} \valkyrie{I think we at least need to preserve the sentiment from the end, that we don't have any novel I/O techniques but that pipes help extend existing techniques in useful ways} \george{remove, repeats}
-Pipes allow makers to build with and upon existing input/output techniques, often integrating these methods into objects they were not originally designed for.
-For example, to create a touch-sensitive toy as in Figure \ref{fig:teaser}, a maker can select several locations of interest on the brain \george{?}.
-After creating pipes to those locations and printing her object, she can fill the pipes with conductive paint.
-For sensing, she need only connect a single wire to a shared out of all the tubes, and can distinguish them via Swept Frequency Capacitve Sensing \cite{Sato-touche}.
-In this way, she has only a single active element (the microcontroller setup), but her otherwise-inert 3D printed object (the brain) has several sensing locations.
-Other existing sensing and actuation approaches, such as FlyEye \cite{Wimmer-flyeye} and Jamming User Interfaces \cite{Follmer-jamming}, can also be enhanced by the use of pipes: makers utilizing such I/O strategies can locate input and output at arbitrary points on a printed 3D object's surface.
+Existing sensing and actuation approaches, such as FlyEye \cite{Wimmer-flyeye}, Swept Frequency Capacitive Sensing \cite{Sato-touche}, and Jamming User Interfaces \cite{Follmer-jamming}, can be enhanced by the use of pipes: makers utilizing such I/O strategies can locate input and output at arbitrary points on a printed 3D object's surface.
 We leverage these existing techniques for sensing and actuation while our work's novelty is in \emph{internal modeling} for 3D printing and the \emph{redirection} of I/O to arbitrary locations on a device.
-
-\tovi{I would remove this entire paragraph}  \valkyrie{if we scrap the first one, sure.  George, what do you think?} \george{remove}
-Pipes can also be used for sending media on a specified route through an object; for example, neon signs are created by evacuated glass tubes formed into a specific path.
-Electricity is run through the near-vacuum, creating light.
-If our maker wants to build such a sign, she can substitute EL wire for the evacuated glass: she need only design a path for it to be fed through and then create pipes along that path.
 
@@
 Pipes should not pass too close to the surface: consumer-grade 3D printers do not guarantee air- or water-tightness of prints, so more than one printed layer may be required to prevent leakage.
-Pipes for applications like neon signs must have a Euler tour \george{explain or ref} through them: this allows a single EL wire to light every edge on the pipe graph.
+Pipes for applications like neon signs must allow a single EL wire to light every edge on the pipe graph (i.e., have an Euler tour).
 No tools currently exist for these design tasks.
@@
 The system takes into considertation the design complexities of routing interior pipes which are described above.
-Finally, we showcase a set of examples, enabled by our modeling tool, which allows us to explore various points in the design space \george{what design space?}.
+Finally, we showcase a set of examples, enabled by our modeling tool, which allows us to explore various points in the design space.
 We conclude by summarizing the input and output modalities enabled by our work, and discussing limitations and areas for future exploration.

### 2014-04-05 paper/tubes.tex
george comments

@@
-\section{A Series of Tubes - A Tool for Pipe Design}
+\section{A Series of Tubes - A Tool for Pipe Design} \george{not clear that this is the beginning of our system description}
 \bjoern{What kind of visualizations and diagrams can help explain the technical part other than photos and screenshots?
@@
 Finally, for designing pipes that follow a user-specified path, we must ensure certain characteristics of that path (e.g., that the path is connected).
+\george{pipes are hard.
+not sure we convinced readers by now that they are worth the effort.
+maybe this has to happen more in the intro.
+the example objects section is still a few pages away.}
 
@@
 
-\subsection{Exterior Connection Points}
+\subsection{Exterior Terminals}
 
@@
 
-Note that pipes once cut are ``outside the mesh'', i.e., pipes cut in sequence cannot intersect each other.
+Pipes avoid each other in this manner: pipes once cut are ``outside the mesh'' and therefore not part of valid routes.
 However, we do not currently track all pipes for global optimization: we greedily select the best routing per pipe.
@@
 Running the physical simulation allows the wire to attempt to return to
-a straight configuration, while satisfying its endpoint position constraints.
+a straight configuration, while satisfying the user-selected endpoint position constraints.
 We also increase the length of the wire, to provide some slack in cases where 
@@
 
-\subsubsection{Routing}
+\subsubsection{Routing} \george{condense or just point to appendix}
 To allow for a single path through a pipe for post-print media insertion, we need to find a semi-Eulerian graph of which a user's input graphics are a subset, and subsequently find an Euler tour on that graph.
@@
 
+/george{too complicated}
 Once we have a semi-Eulerian graph, we need to create an Euler tour.

### 2014-04-06 paper/discussion.tex
mmodz
george comments
minor tweaks

@@
 \section{Discussion}
+
+\valkyrie{there are a lot of things in here that we could focus on.  input/output/identity interactions mediated by tubes, the creation of antennae and other custom electronics...}
+
+\valkyrie{which figures/discussions do we think are important?}
 
@@
 
-By crossing pipe types, topologies, and inserted media, many pipe-mediated interactions emerge.
+\begin{figure}[h]
+\centering
+    \includegraphics[width=3.4in]{figures/electronics.png}
+\caption{(a) A capacitor we manufactured on a 3D printer and coated, post-print, with copper paint.
+This particular one holds a charge of xx.
+(b) Comparison of theoretical printed resistors.
+For a cylindrical wire, resistance = $\rho\frac{l}{\pi r^2}$, where $\rho$ is based on the conductivity of the medium.
+In our case, assuming $ r << l$, resistance(B) < resistance(C) < resistance(A).
+(c) A possible design for a cell phone antenna augmentation inside a cell phone case.}
+\label{fig:electronics}
+\end{figure}
+
+These internal designs can give rise to new opportunities in electronics design and fabrication: we printed a proof-of-concept printed capacitor which features a thin printed plate with hollow cones on either side (see Figure \ref{fig:electronics}A).
+After printing, we coated the inner surfaces of capacitor with copper paint.
+It now can hold a charge of xx.
+Similarly, tubes can be used to create \emph{resistors} in ``printed'' circuitry: a longer or thinner tube has a greater resistance than a shorter or wider one (see Figure \ref{fig:electronics}B).
+Internal \emph{antenna} design is also possible, which could lead to new flexibility in, e.g., cell phone design (see Figure \ref{fig:electronics}C).
+
+By crossing pipe types, topologies, and inserted media, many pipe-mediated \emph{interactions} also emerge.
 We demonstrated a few such possibilities by designing and fabricating example objects with the help of our pipe routing design tool.
@@
 
-=======
->>>>>>> bd673f22cceca8b1cbc4a00aae3717eacd7fd9f2
 \bjoern{We might pull back on the scope of this section a bit and think about which directions here are most closely related to pipes and most likely successful;  and which are more tangential or speculative - for example, Aireal is about feedback in free space and we are all about designing tangible objects; so maybe that's a candidate for cutting.}
@@
 \valkyrie{I cut a few things, and removed references to techniques we have already shown to work (as in the spirit of the intro).  LMK if this is better.}
+
+\valkyrie{do we want those figures of the helmholz thing and the servo redirection gate?}

### 2014-04-06 paper/introduction.tex
minor tweaks

@@
 In (a), objects manufactured on a hobbyist 3D printer.
-In (b), those manufactured on industrial-grade printers.}
+In (b), those manufactured on industrial-grade printers.
+\valkyrie{once these photos are taken, we could expand this into a full-page-span figure that hopefully makes everything look exciting and awesome}}
 \label{fig:teaser}
@@
 A wide array of objects, ranging from bicycle helmets to video game controllers, are now prototyped or even manufactured using these machines.
-However, most devices fabricated by 3D printers are passive, prototyping the form, but not the interactivity, of an object.
+However, most devices fabricated by 3D printers are passive: prototyping the form, but not the interactivity, of an object.
 
@@
 These pipes can be filled, post-print, by a variety of media that enable input, display, and tactile feedback.
-This {\em subtractive} approach is complementary to {\em additive} approaches that inject custom print materials, such as conductors, during the actual printing process [ref].
-While our approach requires some manual assembly after a print completes, it gives makers the opprotunity to cheaply and rapidly prototype a diverse range of interactive objects on 3D printers, including hobbyist machines (Figure \ref{fig:teaser}).
+This {\em subtractive} approach is complementary to {\em additive} approaches that inject custom print materials, such as conductors, during the actual printing process \cite{Sells-reprap}.
+While our approach requires some manual assembly after a print completes, it gives makers the opportunity to cheaply and rapidly prototype a diverse range of interactive objects on 3D printers, including hobbyist machines (Figure \ref{fig:teaser}).
 

### 2014-04-07 paper/introduction.tex
move figure1 into area above abstract... still a bug with the caption for some reason

@@
-\begin{figure}[h]
-\centering
-    \includegraphics[width=3.4in]{figures/placeholder/teaser.png}
-\caption{A collection of interactive objects designed using our tool.
-All have electronic sensing or actuation components which are routed through the \emph{interior} of the prints to preserve aesthetics.
-In (a), objects manufactured on a hobbyist 3D printer.
-In (b), those manufactured on industrial-grade printers.
-\valkyrie{once these photos are taken, we could expand this into a full-page-span figure that hopefully makes everything look exciting and awesome}}
-\label{fig:teaser}
-\end{figure}
-
 \section{Introduction}

### 2014-04-08 paper/relatedwork.tex
george comments
shuffle figure around, add in teaser, fix up glamour shots a bit

@@
 Previous work has investigated the integration of interactivity with the fabrication process.
-Printed Optics (\cite{Willis-printedoptics}) used light pipes, solid clear tubes, to create integrated displays and touch sensors.
+Printed Optics \cite{Willis-printedoptics} used light pipes, solid clear tubes, to create integrated displays and touch sensors.
 While the research describes a process, it does not provide a design tool for creating such objects.
@@
 Sauron investigated computer vision to track interactions with physical mechanisms \cite{Savage-sauron}.
-While Savage, et al., used mirrors for redirection, we believe that the required line-of-sight operation for their computer vision techniques may be possible using printed optics or inserted optical fibers \george{huh?}.
-
+While Savage, et al., used mirrors for redirection, we believe that the required line-of-sight for their computer vision techniques may be possible using printed optics or inserted optical fibers \george{huh?}.
 General-purpose routing for connecting 2D capacitive sensors to terminals is discussed in Midas \cite{Savage-midas}.
 This technique, based on circuitboard routing, is limited to planar and unfoldable shapes, and intended for use on the surface of objects rather than through their cores.
+
 Fabrickation works towards rapid fabrication of functional objects by integrating construction kit building blocks \cite{Mueller-fabrickation}.
-While we recognize that 3D printing requires significant time, we believe creating precise internal structures in 3D printed objects is a strong case for digital fabrication over manual or traditional methods.
-\george{wording}
+While we recognize that 3D printing requires significant time, we believe that precise internal structures in 3D printed objects provide a strong case for digital fabrication over manual or traditional methods.
 
@@
 Its approach of linking edges into a continuous line based on semi-Eulerization is similar to ours.
-However, we are not confined to the plane to avoid line-crossings, and we can create ``invisible'' lines using \emph{blocking out}\footnote{In neon sign making, segments of a tube that should not appear illuminated are covered with opaque paint.
+However, we are not confined to the plane to avoid line-crossings, and we can create ``invisible'' lines by \emph{blocking out}\footnote{In neon sign making, segments of a path that should not appear illuminated are covered with opaque paint.
 This is called "blocking out".
-\url{http://en.wikipedia.org/wiki/Neon_sign}} \george{too much detail and talk of our work}.
-Prior work used tours of nodes, rather than edges, and cast continuous line drawing as a traveling salesman problem~\cite{Bosch-tsp}.
+\url{http://en.wikipedia.org/wiki/Neon_sign}} the inserted media \george{too much detail and talk of our work}.
+Prior work used tours of nodes, rather than edges, and cast continuous line drawing as a traveling salesman problem \cite{Bosch-tsp}.
 In traditional neon sign making, ``routing'' is a conversation between a graphic designer and a glass bender on what is feasible and comfortable for the latter and what will help him avoid burning his hands and wasting time or glass \cite{strattman1997neon}.

### 2014-04-08 paper/relatedwork.tex
bjoern's pass through intro, related work (almost complete)

--- 2014-04-05 paper/relatedwork.tex
@@
+%!TEX root = series of tubes.tex
 \section{Related Work}\tovi{related work section still needs some work}
@@
 
-Previous work has investigated the integration of interactivity with the fabrication process.
-Printed Optics (\cite{Willis-printedoptics}) used light pipes, solid clear tubes, to create integrated displays and touch sensors.
-While the research describes a process, it does not provide a design tool for creating such objects.
-Our work is inspired by Printed Optics and introduces a design tool for modelling objects that can have these types of capabilities.
+Previous work has begun to investigate how to fabricate interactive objects using particular sensing and display technologies.
+Printed Optics (\cite{Willis-printedoptics}) uses light pipes to create integrated displays and sensed input optically through components added at print time; while Slyper, et al., 3D print flexible robot armatures capable of sensing interaction via changes in air pressure \cite{Slyper-pressure}.
+Our work is inspired by these projects, and contributes a general design tool for modelling objects that can have such capabilities.
 
-Slyper, et al., 3D printed flexible robot armatures capable of sensing interaction via changes in air pressure \cite{Slyper-pressure}.
-They created all their models by hand, and were forced to place interaction locations in areas accessible to electronics assembly: this technique is complementary to ours, as we allow placing interactive elements anywhere on an object's surface.
-Sauron investigated computer vision to track interactions with physical mechanisms \cite{Savage-sauron}.
-While Savage, et al., used mirrors for redirection, we believe that the required line-of-sight operation for their computer vision techniques may be possible using printed optics or inserted optical fibers \george{huh?}.
+Sauron tracks interactions with printed physical mechanisms by placing a camera inside hollow objects, where all components have to be visible to a camera~\cite{Savage-sauron}.
+Our technique instead can work with general, solid models and only carves out pipes for interaction where needed.
+Midas generates fabrication files custom capacitive touch sensors~\cite{Savage-midas}.
+This technique, based on circuitboard routing, is limited to planar and unfoldable shapes.
+Connections are routed on the surface of objects.
+\systemname can also be used for touch sensing, but routes conducters through the inside of objects.
 
-General-purpose routing for connecting 2D capacitive sensors to terminals is discussed in Midas \cite{Savage-midas}.
-This technique, based on circuitboard routing, is limited to planar and unfoldable shapes, and intended for use on the surface of objects rather than through their cores.
 Fabrickation works towards rapid fabrication of functional objects by integrating construction kit building blocks \cite{Mueller-fabrickation}.
 While we recognize that 3D printing requires significant time, we believe creating precise internal structures in 3D printed objects is a strong case for digital fabrication over manual or traditional methods.
-\george{wording}
+\george{wording}\bjoern{needs work}
 
-\subsubsection{Fabrication + Electronics}
-
-Other research has investigated the integration of traditional electronics into 3D prints by depositing non-printable materials after printing.
-Sells, et al., \cite{Sells-reprap}, describe using a heated syringe to deposit metal in printed 2D circuitboard channels on objects' surfaces.
-Sarik, et al., \cite{Sarik-tracebrush} and the Optomec company \cite{optomec} describe \emph{spraying} conductive material onto printed objects to create circuitry.
-Injection of liquid metals for sensing in softer-than-skin electronics was explored by Park, et al., \cite{Park-microchannels} and Majidi, et al., \cite{Majidi-curvature}.
-Navarette, et al., created a functioning GPS routed in true 3D (not multilayer boards' 2.5D), hand-built using a one-off routing.
-Our work differs from these because we offer \emph{algorithms and techniques} for \emph{internal}, \emph{true 3D} routing.
-
-\subsubsection{Interaction for Fabrication}
-
-Other researchers are investigating interaction for fabrication (i.e., interaction techniques for controlling digital fabrication machines) rather than fabrication for interaction as we do.
-This includes work like Constructable \cite{Mueller-constructable}, KidCAD \cite{Follmer-kidcad}, and Willis, et al,'s interactive interfaces \cite{Willis-interactive}.
-Their goals and contributions are orthogonal to ours.
+\subsubsection{Fabrication with Printed Electronics}
+Researchers are also investigating how to print conductors and place electronic components as part of additive manufacturing.
+Sells, et al., \cite{Sells-reprap}use a heated syringe to deposit metal in printed 2D circuitboard channels on objects' surfaces.
+Sarik, et al., \cite{Sarik-tracebrush} use aerosol jetting of conductive material onto printed objects.
+Park, et al., \cite{Park-microchannels} and Majidi, et al., \cite{Majidi-curvature} inject liquid metals.
+Our approach does not deposit such materials, but creates structures inside a 3D model that make manual post-print injection of conductive ink possible.
+%Navarette, et al., created a functioning GPS routed in true 3D (not multilayer boards' 2.5D), hand-built using a one-off routing.  \bjoern{need more context for this}
+%Our work differs from these because we offer \emph{algorithms and techniques} for \emph{internal}, \emph{true 3D} routing.
 
@@
 \url{http://en.wikipedia.org/wiki/Neon_sign}} \george{too much detail and talk of our work}.
-Prior work used tours of nodes, rather than edges, and cast continuous line drawing as a traveling salesman problem~\cite{Bosch-tsp}.
+Prior work used tours of nodes, rather than edges, and cast continuous line drawing as a traveling salesman problem \cite{Bosch-tsp}.
 In traditional neon sign making, ``routing'' is a conversation between a graphic designer and a glass bender on what is feasible and comfortable for the latter and what will help him avoid burning his hands and wasting time or glass \cite{strattman1997neon}.
@@
 All these contributions are orthogonal to ours, as our focus is on precisely \emph{removing} material from printed objects, rather than blending materials or modifying components.
+
+\subsubsection{Interaction for Fabrication vs. Fabrication for Interaction}
+HCI researchers are also developing novel interaction techniques for providing input to digital fabrication machines~\cite{Mueller-constructable,Willis-interactive,mixfab}, and new hybrid hand tools combine manual processes with digital fabrication~\cite{zoran2013freed,rivers2012position}.
+These projects are complementary as they target different aspects of the design process, though it is conceivable that our pipe design tools could be added to augmented reality modeling interfaces like MixFab~\cite{mixfab}.

### 2014-04-08 paper/relatedwork.tex

@@
 %!TEX root = series of tubes.tex
-\section{Related Work}\tovi{related work section still needs some work}
+\section{Related Work}
 Our contributions relate to other efforts aiming to add interactivity to digitally fabricated objects; and to techniques for routing and model modification.
@@
 Connections are routed on the surface of objects.
-\systemname can also be used for touch sensing, but routes conducters through the inside of objects.
+\systemname can also be used for touch sensing, but routes conductors through the inside of objects.
 
-Fabrickation works towards rapid fabrication of functional objects by integrating construction kit building blocks \cite{Mueller-fabrickation}.
+%Fabrickation works towards rapid fabrication of functional objects by integrating construction kit building blocks \cite{Mueller-fabrickation}.
 While we recognize that 3D printing requires significant time, we believe creating precise internal structures in 3D printed objects is a strong case for digital fabrication over manual or traditional methods.
-\george{wording}\bjoern{needs work}
+\george{wording}\bjoern{not quite sure how this fits in.}
 
@@
 Park, et al., \cite{Park-microchannels} and Majidi, et al., \cite{Majidi-curvature} inject liquid metals.
-Our approach does not deposit such materials, but creates structures inside a 3D model that make manual post-print injection of conductive ink possible.
+Our approach does not deposit such materials automatically, but creates structures inside a 3D model that make manual post-print injection of conductive ink possible.
 %Navarette, et al., created a functioning GPS routed in true 3D (not multilayer boards' 2.5D), hand-built using a one-off routing.  \bjoern{need more context for this}
@@
 \subsection{Routing and Internal Modeling}
+Part of our contribution comprises algorithms and techniques for routing tubes through 3D objects.
+Circuitboard routing algorithms, such as Lee's maze router \cite{Lee-maze}, assume 2D surfaces.
+Multi-layer boards consist of stacked of 2D layers connected by vias.
+We compute true 3D point to point routing though the interior of solid models with flexible paths.
+This approach has been used in one-off designs~\cite{Navarrette-gps} but we are not aware of design tools for this purpose.
 
-Part of our contribution is in the algorithms and techniques used for routing tubes through 3D objects.
+One of our path routing algorithms routes a single, continuous pipe through an object based that follows the appearance of an input shape.
+The technique is inspired by work in generating continuous line illustrations from images~\cite{Wong-continuousline,Bosch-tsp}.
+Our approach is also based on graph theory (Euler tours), but differs in that it generates 3D geometry instead of 2D illustrations.
 
-Circuitboard routing algorithms, such as Lee's maze router algorithm \cite{Lee-maze}, work only on a 2D surface.
-Multi-layer boards are typically 2.5D, in that they comprise a stack of 2D circuitboards connected by vias.
-Instead we offer true 3D point to point routing though the interior of solid models.
-
-Similar to our path routing algorithm, Wong presents a graph-based approach for generating continuous line illustrations from images \cite{Wong-continuousline}.
+%Similar to our path routing algorithm, Wong presents a graph-based approach for generating continuous line illustrations from images \cite{Wong-continuousline}.
 Its approach of linking edges into a continuous line based on semi-Eulerization is similar to ours.
@@
 
-There is prior work on modeling the interior of objects (as opposed to their exterior).
-Takayama, et al., built a system described in \cite{Takayama-fruit} which allowed users to create internal color maps of the interior of objects, allowing them to see digital ``slices'' of their ``oranges.''  Our work focuses on 3D printable geometry rather than rendering.
-In Make it Stand, material is removed from the interior of objects with the goal of changing their center of gravity \cite{Prevost-makeitstand}, and Infrastructs created encoded interior cavities in 3D printed objects designed to be sensed with a terahertz imaging tool.
-Modeling such chambers is one aspect of our work, however our focus is on interaction rather than balance or identification.
-Bickel, et al., explored enforcing deformation behavior with material blending \cite{Bickel-deformation}, while B\"{a}cher, et al., built an algorithm for automatically placing 3D fabricatable joints into meshes \cite{Bacher-articulated}.
-Sauron automatically modified the interior of existing mechanical objects to make them amenable to vision-based sensing \cite{Savage-sauron}.
+Some prior work has also focused on modifying interior parts of 3D objects. 
+%Takayama, et al., built a system described in \cite{Takayama-fruit} which allowed users to create internal color maps of the interior of objects, allowing them to see digital ``slices'' of their ``oranges.'' \bjoern{don't understand this} Our work focuses on 3D printable geometry rather than rendering.
+In Make it Stand, material is removed from the interior of objects with the goal of changing their center of gravity \cite{Prevost-makeitstand}.
+Infrastructs encodes information using interior cavities that can be sensed using teraherz imaging.
+Bickel, et al., explored enforcing deformation behavior with multi-material blending inside objects\cite{Bickel-deformation}.
+Modeling internal chambers is one aspect of our work, however our focus is on interaction rather than balance, identification or deformation.
+%B\"{a}cher, et al., built an algorithm for automatically placing 3D fabricatable joints into meshes \cite{Bacher-articulated}.
 All these contributions are orthogonal to ours, as our focus is on precisely \emph{removing} material from printed objects, rather than blending materials or modifying components.

### 2014-04-08 paper/abstract.tex

@@
+%!TEX root = series of tubes.tex
 \begin{abstract}
 3D printers offer extraordinary flexibilty for prototyping the shape and mechanical function of objects.
-However, a 3D printer's role in prototyping interactive objects is still not clear.
-We propose a general technique for supporting the rapid prototyping of interactivity by removing interior material of 3D models to form pipes and chambers.
+We investigate how 3D models can be modified to facilitate the creation of interactive objects that offer dynamic input and output.
+%However, a 3D printer's role in prototyping interactive objects is still not clear.  
+We introduce a general technique for supporting the rapid prototyping of interactivity by removing interior material of 3D models to form pipes and chambers.
 We describe this new design space of pipes and hollow chambers for interaction design, where variables include openings, topologies, design focus, and inserted media.
-We then present a  tool for routing such pipes through the interior of 3D models, integrated within a 3D modeling program.
+We then present \systemname, a tool for routing such pipes through the interior of 3D models, integrated within a 3D modeling program.
 We use two distinct routing algorithms.
@@
 The second allows users to supply a desired internal shape to which we fit a pipe route: for this we developed a novel graph-based routing algorithm.
-We present several totally tubular prototypes we created using our tool to show its flexibility and potential.
+We present several prototypes we created using our tool to show its flexibility and potential.
 \end{abstract}

### 2014-04-08 paper/introduction.tex

@@
+%!TEX root = series of tubes.tex
+%\bjoern{Leave this line in - it makes compiling easier in sublime editor}
+
 \section{Introduction}
-Makers, as well as professional designers, leverage 3D printers as tools for design work.
-A wide array of objects, ranging from bicycle helmets to video game controllers, are now prototyped or even manufactured using these machines.
-However, most devices fabricated by 3D printers are passive: prototyping the form, but not the interactivity, of an object.
+Makers, as well as professional designers, increasingly leverage 3D printers as tools for design work.
+A wide array of objects, ranging from toy figurines to video game controllers and jet engine blades, are now prototyped (and some times even manufactured) using these machines.
+Most models fabricated on 3D printers today are passive: they express the form, but not the function, of an object.
 
-Due to the growing popularity of 3D printing and maker communities like Thingiverse\footnote{http://thingiverse.com}, recent work has looked at the issue of rapidly prototyping interactive objects.
-Existing approaches include Printed Optics \cite{Willis-printedoptics}, which uses optically clear material for light redirection, and Sauron \cite{Savage-sauron}, which is based on computer vision, but they are limited to visible (reflecting light off fingers, mechanical motion of components) input and visual (screen-like) output; they do not give makers freedom to integrate haptic feedback or other inputs or outputs.
+Recently, human-computer interaction researchers have begun to explore how to prototype interactive objects using additive 3D printing processes.
+%and maker communities like Thingiverse\footnote{http://thingiverse.com}, recent work has looked at the issue of rapidly prototyping interactive objects. 
+Existing approaches include Printed Optics \cite{Willis-printedoptics}, which print light pipes using optically clear material, and Sauron \cite{Savage-sauron}, which uses computer vision to track the movment of components on the inside of a hollow object.
+These systems exclusively focus on optical sensing and visual displays --- we seek to grow the vocabulary of supported tangible input and output modalities.
+%is based on computer vision, but they are limited to visible (reflecting light off fingers, mechanical motion of components) input and visual (screen-like) output; they do not give makers freedom to integrate haptic feedback or other inputs or outputs.
 
-We propose a novel technique involving the removal of material from the interior of 3D models, prior to printing, to create pipes and other cavities.
-These pipes can be filled, post-print, by a variety of media that enable input, display, and tactile feedback.
+This paper introduces a novel technique involving the removal of material from the interior of 3D models, prior to printing, to create pipes and other cavities.
+These pipes can be filled, post-print, with a variety of media that enable input, display, and tactile feedback.
 This {\em subtractive} approach is complementary to {\em additive} approaches that inject custom print materials, such as conductors, during the actual printing process \cite{Sells-reprap}.
-While our approach requires some manual assembly after a print completes, it gives makers the opportunity to cheaply and rapidly prototype a diverse range of interactive objects on 3D printers, including hobbyist machines (Figure \ref{fig:teaser}).
+While our approach requires some manual assembly after a print completes, it gives makers the opportunity to cheaply and rapidly prototype a diverse range of interactive objects on 3D printers, including popular single-material fused-deposition modeling machines (Figure \ref{fig:teaser}).
 
@@
 By exploring this design space new opportunities for adding interactivity to 3D printed objects can be realized.
-For example, copper material can fill the channels, to allow for standard electronic components to be easily integrated after the printout; electroluminescent (EL) wire can be threaded through to create computer-controlled visual output; or air can be pumped through to control haptic responses.
-In comparison to traditional exterior electronics integration, our process allows makers to preserve aesthetics by hiding wires inside the object; additionally it offers more freedom in selecting precise surface locations for, e.g., air mediated tactile output.
-
+For example, injecting conductive paint can turn pipes into conductors, which can in turn be used to integrate electronic components into prints.
+Electroluminescent (EL) wire can be threaded through pipes to create computer-controlled visual output.
+And air can be pumped through pipes to produce haptic effects.
 Existing sensing and actuation approaches, such as FlyEye \cite{Wimmer-flyeye}, Swept Frequency Capacitive Sensing \cite{Sato-touche}, and Jamming User Interfaces \cite{Follmer-jamming}, can be enhanced by the use of pipes: makers utilizing such I/O strategies can locate input and output at arbitrary points on a printed 3D object's surface.
 We leverage these existing techniques for sensing and actuation while our work's novelty is in \emph{internal modeling} for 3D printing and the \emph{redirection} of I/O to arbitrary locations on a device.
+
+%Routing pipes through the inside of objects allows makers to preserve the aesthetics of their design.
+%additionally it offers more freedom in selecting precise surface locations for, e.g., air mediated tactile output.
+\bjoern{too oblique?}
 
@@
 No tools currently exist for these design tasks.
+To aid users in working with 3D-printed pipes, we contribute new techniques for interactively routing and generating internal structures, and integrate these tools into a 3D modelling application.
 
-Our work makes several contributions to the field of digitally fabricated interactive objects.
+We then present a set of example objects, fabricated with the help of our modeling tool, that demonstrate different interaction modalities.
+We conclude by discussing limitations and outlining areas for future exploration.
+%Our work makes several contributions to the field of digitally fabricated interactive objects.
 First, we describe a design space of how pipes can be integrated into 3D models.

### 2014-04-08 paper/designspace.tex
design space section rewrite

@@
+%!TEX root = series of tubes.tex
+
 \section{The Design Space of Pipes}
-The design space of pipes has four dimensions: openings, design focus, topologies, and inserted media.  We describe the space below.
+\bjoern{would be nice to have a bridge from related work to here.}
+For the purposes of interaction desing, we have identified four relevant dimensions of pipe design: openings, path constraints \bjoern{note the rename - replicate in figure}, topologies, and inserted media.
+We describe the space below.
 
-\begin{figure}[h!]
+\begin{figure}[t]
 \centering
-    \includegraphics[width=3.4in]{figures/tubespace.pdf}
+    \includegraphics[width=1.0\columnwidth]{figures/tubespace.pdf}
 \caption{The design space of pipes.  See text for details.}
@@
 
-\subsection{Openings of Pipes}
-
-\george{difference between open/return not clear}Pipes can have four opening styles: open, return, semi-closed, and fully enclosed (see Figure \ref{fig:pipespace}).
+\subsection{Openings}
+Pipes inside solid objects can be either open to the outside on both ends, semi-enclosed, or fully enclosed (see Figure \ref{fig:pipespace}).
+For interactive devices, openings can be either user-facing (in places a user would for example touch a model); or system-facing, where sensors and actuators may be connected.
+%\george{difference between open/return not clear}
+%Pipes can have four opening styles: open, return, semi-closed, and fully enclosed .
 As we are interested in interactive devices, we distinguish a ``user side"---part of the surface or interior of an object facing its user for interaction; and a ``system side"---where other hardware such as electronics may be connected.
@@
 
-\begin{figure}[h]
+\begin{figure}[t]
 \centering
-    \includegraphics[width=3.4in]{figures/types.png}
+    \includegraphics[width=1.0\columnwidth]{figures/types.png}
 \caption{Pipe openings: A shows a pair of open tubes filled with copper paint to power an LED.
@@
 
-Open pipes originate from the system side and connect the user side, with both ends of the pipe open.
-This type of pipe may be used to create electronic circuitry.
-For example, an open pipe filled with conductive paint can power an LED (see Figure \ref{fig:openings}A).
+Open pipes may be used to create electronic circuitry.  For example, an open pipe filled with conductive paint and connected to a battery can provide power to an LED (see Figure \ref{fig:openings}A).
+%
+Semi-enclosed pipes are capped on one side.
+Caps can for example be used to create tactile output.
+By fabricating a cap from a malleable material and attaching a pump to the open ended, haptic feedback can be generated by raising and depressing the flexible cap (see Figure \ref{fig:openings}C).
+%
+A fully enclosed pipe or cavities are entirely internal to the object.
+Fully enclosed pipes may be used as weight modifiers and for object identification, as in Figure \ref{fig:openings}D.
+They can also contain for water or particles that would otherwise fall out.\bjoern{how would you get the particles inside?}
 
-A return pipe originates at the system and returns back to it, without offering physical access to the user.
-By threading an electroluminescent (EL) wire through a clear return pipe, a maker can create a custom piece of neon art (see Figure \ref{fig:openings}B).
-\george{not sure this is adding something useful}
-
-Semi-closed pipes are open on the system side, closed on the user side.
-A potential use of this is to create tactile output: The closed interface may be fabricated from a malleable material, e.g., a rubber membrane, and the open end attached to a fluid pump to create haptic feedback (see Figure \ref{fig:openings}C).
-
-A fully enclosed pipe is closed to both the system and the user.
-Fully enclosed pipes may be used as weight modifiers (e.g., for object identification), as in Figure \ref{fig:openings}D.
-They can also contain for water or particles that would otherwise fall out.
-
-\subsection{Design of Pipes}
-Pipe design may emphasize exterior connection points and/or interior paths.
-Our maze (see Figure \ref{fig:maze}) requires specific terminals (the start and end of the maze) as well as a specific path for its tubes (the maze itself).
-Alternatively, the neon sign's output in Figure \ref{fig:neon} is based only on the \emph{path} of the pipe.
+\subsection{Path Constraints}
+Pipe designs may require specific start and end points, but leave the precise path flexible.
+For example, circuit schematics specify connectivity, but many routing solutions are valid.
+Alternatively, the shape of the path itself may be important.
+For example, in neon signs, single tubes are bent into letterforms.
+Threading
+ electroluminescent (EL) wire through a clear pipe enables makers to emulate neon art (see Figure \ref{fig:openings}B). 
+In a maze (see Figure \ref{fig:maze}), both specific terminals as well as a specific path (the maze itself) matter. 
+% Alternatively, the neon sign's output in Figure \ref{fig:neon} is based only on the \emph{path} of the pipe.
 Focused on just the exterior features is the touch sensitive toy in Figure \ref{fig:toys}, where pipes must exit the toy at the frontal lobe and parietal lobes to make those \emph{specific locations} touch-sensitive.
@@
 \subsection{Pipe Topologies}
-
-Pipe network topologies enable different interactions: splitting or mixing pipes offer flexibility in output, allowing a single source to have multiple outputs or multiple sources to combine (see Figure \ref{fig:pipespace}).
-Star and tree topologies extend the splitting and mixing primitives, encompassing multiple-in and multiple-out in contrast to single-in/single-out; we used star topologies in our toys \ref{fig:toys}.
+Pipe network topologies enable splitting or mixing of media (see Figure \ref{fig:pipespace}).
+Star and tree topologies extend the splitting and mixing primitives, encompassing multiple-in and multiple-out configurations; we use star topologies to create different conductive paths in touch-sensitive toys, with a single terminal for swept-frequency capacitive sensing(Figure~\ref{fig:toys}).
 
 \subsection{Media in Pipes}
-After printing, there is still flexibility in inserting media to create different interface affordances and capabilities. We consider gas, liquids, solids, particulates, and threadables.
-
-``Gas'' comprises compressible fluids.
-Fluid pressure inside semi-closed pipes can create haptic feedback, or gases can be used as carriers for scents or fog; fluid pressure can also be used as an input.
-
-Incompressible fluids (``liquids'') perform similarly to gases for display or haptics.  Additionally, using driable conductive liquids like copper paint, pipes can become custom wires.
-
-Pipes need not have hollow centers: if routed pipes are filled with solid material---in particular, a solid material different from the enclosing model material---, interactions such as those in Printed Optics \cite{Willis-printedoptics} are possible.
-
+After printing, different media can be inserted into pipes to change affordances and capabilities. We consider gas, liquids, solids, particulates, and threadables. 
+%
+Compressible fluids (``gas'') and incompressible fluids (``liquids'')  inside semi-enclosed pipes can create haptic feedback; pressure can also be sensed to act as an input.
+Gases can be used as carriers for scents or fog.
+Conductive liquids like copper paint, when dried, turn pipes into conductive paths and can also fix electronic components in place.
+%
+%got to here
+Filling pipes with solids at print time enables designers to exploit the difference in material properties between pipe and contents.
+For example, in Printed Optics~\cite{Willis-printedoptics}, different refractive indices between solid cladding material and transparent material inside pipes made light transport along pipes for input and output possible.
+%
 Particulates, either printed in-place or inserted, can be of varying densities and can contain particles of various and variable size.
@@
 Small particles in fluid can provide variable stiffness haptics via jamming \cite{Follmer-jamming}.
+%
+Threadable media, such as regular wire, electroluminescent wire, or commercial fiberoptic cables, can be threaded through pipes post-printing.
+Threading such media enables addition of conduction, illumination, or optical sensing.
+% This empowers inexpensive printers: for example, a Printed Optics-style interface can be created on a consumer-grade 3D printer with pipes and fiber optic cable.
 
-Threadable media, such as EL wire or fiberoptic cables, can be threaded through pipes post-printing.
-This empowers inexpensive printers: for example, a Printed Optics-style interface can be created on a consumer-grade 3D printer with pipes and fiber optic cable.
+We next show how different pipe openings, constraints and topologies can be created with the help of a new design tool, \systemname.

### 2014-04-08 paper/tubes.tex
script
bjoern's pass through intro, related work (almost complete)
design space section rewrite

@@
-\section{A Series of Tubes - A Tool for Pipe Design} \george{not clear that this is the beginning of our system description}
-\bjoern{What kind of visualizations and diagrams can help explain the technical part other than photos and screenshots?
-The "UIST" pipe routing figure goes in this direction.
-Can we make similar conceptual figures, flowcharts, etc for the earlier parts?}
+\section{\systemname - A Tool for Pipe Design}
 

### 2014-04-09 paper/designspace.tex
bjoern's edits through end of tools section

@@
 A fully enclosed pipe or cavities are entirely internal to the object.
-Fully enclosed pipes may be used as weight modifiers and for object identification, as in Figure \ref{fig:openings}D.
+Fully enclosed pipes may be used as weight modifiers~\cite{Prevost-makeitstand} and for object identification~\cite{Willis-infrastructs}, as in Figure \ref{fig:openings}D.
 They can also contain for water or particles that would otherwise fall out.\bjoern{how would you get the particles inside?}
@@
 Pipe network topologies enable splitting or mixing of media (see Figure \ref{fig:pipespace}).
-Star and tree topologies extend the splitting and mixing primitives, encompassing multiple-in and multiple-out configurations; we use star topologies to create different conductive paths in touch-sensitive toys, with a single terminal for swept-frequency capacitive sensing(Figure~\ref{fig:toys}).
+Star and tree topologies extend the splitting and mixing primitives, encompassing multiple-in and multiple-out configurations; we use star topologies to create different conductive paths in touch-sensitive toys, with a single terminal for swept-frequency capacitive sensing (Figure~\ref{fig:toys}).
 
@@
 %
-Compressible fluids (``gas'') and incompressible fluids (``liquids'')  inside semi-enclosed pipes can create haptic feedback; pressure can also be sensed to act as an input.
+Compressible fluids (``gas'') and incompressible fluids (``liquids'')  inside semi-enclosed pipes can create haptic feedback; pressure can also be sensed to act as an input~\cite{Slyper-shape}.
 Gases can be used as carriers for scents or fog.
@@
 Small particles in fluid can provide variable stiffness haptics via jamming \cite{Follmer-jamming}.
+\bjoern{I thought jamming is about particles in a vacuum?
+why liquid?}
 %
 Threadable media, such as regular wire, electroluminescent wire, or commercial fiberoptic cables, can be threaded through pipes post-printing.
-Threading such media enables addition of conduction, illumination, or optical sensing.
+Threading such media enables addition of conduction, illumination, or optical sensing using materials that can easily be procured, but cannot themselves be printed, e.g., on consumer printers with a single material.
 % This empowers inexpensive printers: for example, a Printed Optics-style interface can be created on a consumer-grade 3D printer with pipes and fiber optic cable.

### 2014-04-09 paper/introduction.tex
bjoern's edits through end of tools section

@@
 Pipes should not pass too close to the surface: consumer-grade 3D printers do not guarantee air- or water-tightness of prints, so more than one printed layer may be required to prevent leakage.
-Pipes for applications like neon signs must allow a single EL wire to light every edge on the pipe graph (i.e., have an Euler tour).
+Pipes for applications like illumunated EL wire signs must allow a single pipe to follow a specific path layout (e.g., spelling out text).
 No tools currently exist for these design tasks.
 To aid users in working with 3D-printed pipes, we contribute new techniques for interactively routing and generating internal structures, and integrate these tools into a 3D modelling application.
+\bjoern{what's missing here is how our tool achieves the goal.
+May want to flip this paragraph around - ``we contribute a design tool that addresses various challenges of pipe design makers are likely to encounter.
+The tool offers 3D point-to-point routing of flexible pipes based on a physical simulation of flexible rods.
+This process ensures maximum bend angles, which ...'' etc.}
 

### 2014-04-09 paper/relatedwork.tex
bjoern's edits through end of tools section

@@
 Our technique instead can work with general, solid models and only carves out pipes for interaction where needed.
-Midas generates fabrication files custom capacitive touch sensors~\cite{Savage-midas}.
-This technique, based on circuitboard routing, is limited to planar and unfoldable shapes.
-Connections are routed on the surface of objects.
+Midas generates fabrication files for custom capacitive touch sensors~\cite{Savage-midas}.
+Their technique, is limited to planar and unfoldable shapes as it routes connections on the surface of objects.
 \systemname can also be used for touch sensing, but routes conductors through the inside of objects.
@@
 Researchers are also investigating how to print conductors and place electronic components as part of additive manufacturing.
-Sells, et al., \cite{Sells-reprap}use a heated syringe to deposit metal in printed 2D circuitboard channels on objects' surfaces.
+Sells, et al., \cite{Sells-reprap} use a heated syringe to deposit metal in printed 2D circuitboard channels on objects' surfaces.
 Sarik, et al., \cite{Sarik-tracebrush} use aerosol jetting of conductive material onto printed objects.
@@
 
-One of our path routing algorithms routes a single, continuous pipe through an object based that follows the appearance of an input shape.
+One of our path routing algorithms routes a single, continuous pipe through an object that follows the appearance of an input shape.
 The technique is inspired by work in generating continuous line illustrations from images~\cite{Wong-continuousline,Bosch-tsp}.
@@
 In Make it Stand, material is removed from the interior of objects with the goal of changing their center of gravity \cite{Prevost-makeitstand}.
-Infrastructs encodes information using interior cavities that can be sensed using teraherz imaging.
+Infrastructs encodes information using interior cavities that can be sensed using teraherz imaging~\cite{Willis-infrastructs}.
 Bickel, et al., explored enforcing deformation behavior with multi-material blending inside objects\cite{Bickel-deformation}.

### 2014-04-09 paper/tubes.tex
bjoern's edits through end of tools section

@@
+%!TEX root = series of tubes.tex
 \section{\systemname - A Tool for Pipe Design}
 
+To allow makers to design novel objects with pipe-powered interfaces, we created a tool, \systemnamenospace, to add pipe geometries to arbitrary 3D models.
 Pipe design is non-trivial.
-Independent pipes cannot intersect lest their contents be unintentionally mixed.
-Pipes' bending energy needs to be minimized to allow easy insertion of media post-print.
+Independent pipes should not intersect inside an object lest their contents be unintentionally mixed.
+To permit easy insertion of media after printing, pipes should have smooth bends.
 Pipes in general need to avoid objects' surfaces to prevent fluid leakage when printed by hobbyist-grade machines.
 Finally, for designing pipes that follow a user-specified path, we must ensure certain characteristics of that path (e.g., that the path is connected).
-\george{pipes are hard.
-not sure we convinced readers by now that they are worth the effort.
-maybe this has to happen more in the intro.
-the example objects section is still a few pages away.}
+%\george{pipes are hard.  not sure we convinced readers by now that they are worth the effort.  maybe this has to happen more in the intro.  the example objects section is still a few pages away.}
 
-To allow makers to design novel objects with pipe-powered interfaces, we created a tool for use on arbitrary 3D models.
-This tool allows designers to select exterior connection points of their pipes (see Figure \ref{fig:tool-process-interior}) or to import vector art describing the pipes' interior paths (see Figure \ref{fig:tool-process-exterior}).
-Once the maker's selections are made, we create a complete routing using either the exterior connection point method (A*, physical simulation) or the interior path method (graph edge creation, Euler tour generation).
-We thicken our routing to create pipes.
-The resultant pipes are subtracted from the original mesh, which can then be 3D printed.
-
-Our tool is implemented as a part of Meshmixer, a consumer 3D mesh editing tool, in C++ \cite{Schmidt-meshmixer}.
+ \systemname allows users to design pipes in two ways.
+First, they can select exterior anchor points on their objects and internal connection points; the tool then creates complete point-to-point routings using A* path search and physical rod simulation (see Figure \ref{fig:tool-process-exterior}).
+Second, users can also import vector art describing desired interior paths (see Figure \ref{fig:tool-process-interior}).
+\systemname then creates a single path that follows the input shape using edge graph manipulation and Euler tour generation.
+We thicken these routes to create pipes.
+The resultant pipes are subtracted from the original mesh using voxelization and remeshing.
+Our tool is implemented in C++ as an extension to Meshmixer, a consumer 3D mesh editing tool~\cite{Schmidt-meshmixer}.
 
@@
 
-Designing objects which, for example, are touch sensitive in particular areas or integrate with existing electronics, requires precise location and sizing of pipe endpoints.
+Designing objects which, for example, are touch sensitive in particular areas or contain electronic components, requires precise location and sizing of pipe endpoints.
 The interior of these pipes should have as large a bending radius as possible so that post-print insertion of solid or viscous media is easier.
 
-To create pipes in which the location and shape of exterior connection points matters, we first allow makers to select the connection points on the mesh's surface.
+To create pipes in which the location and shape of exterior connection points matters, we first allow makers to select these points on the mesh's surface.
+\bjoern{feels a bit redundant with the last paragraph in the prior section.}
 
@@
     \includegraphics[width=3.4in]{figures/exterior.png}
-\caption{An example mesh with exterior connection points selected.  The smoothed A* routing is drawn in {\color{red}red}, and in {\color{tovi}green} is our physically-based rod after simulation.}
+\caption{An example mesh with exterior connection points selected.
+The smoothed A* routing is drawn in {\color{red}red}, and in {\color{tovi}green} is our physically-based rod after simulation.
+\bjoern{This figure is visually hard to understand.
+I think a step-by-step figure would be better - a) selecting two points.
+b) red a* path c) final routed pipe.
+}}
 \label{fig:tool-process-exterior}
@@
 \subsubsection{Routing and Physical Simulation}
-
-Our basic first-pass routing algorithm uses the A* routing algorithm \cite{Hart-Astar}.
-This algorithm keeps a priority queue of points to check for routing based on a cost function; for each point it tags the distance from the start (e.g., startPoint = 0, startNeighbors=1, ...).
+Given two points, we first find a path between them using the A* search algorithm on a voxel representation of the object~\cite{Hart-Astar}.
+A* finds a least-cost path through a graph of nodes given a cost function.
+Our graph comprises the grid of voxels as nodes, with edges between adjacent nodes; our path cost is the based on the Euclidean distance to the end point.
+In this formulation, the routed path is naturally constrained to stay within the object boundaries.
+%; for each point it tags the distance from the start (e.g., startPoint = 0, startNeighbors=1, ...).
 Once the end point is reached, we backtrack through tagged points, always moving to a lower-valued point.
-The path cost in our implementation is based only on shortest distance between the starting and ending points, without weighting for distance from the surface, however the routed path is forced to stay inside the mesh's boundaries.
-We route on the voxelized 3D grid of the mesh, using a resolution of 128x128x128 voxels.
-
-Pipes avoid each other in this manner: pipes once cut are ``outside the mesh'' and therefore not part of valid routes.
-However, we do not currently track all pipes for global optimization: we greedily select the best routing per pipe.
+\bjoern{find a clearer way to express this?}
 
@@
 we require a smoother path through which we can actually push a physical wire.
+In addition, distance from the surface is not taken into account in A* search and pipes may thus be routed along surface contours, which may cause leakage.
+
 We solve this problem via physical simulation. We create a straight virtual 
@@
 constraint between adjacent segments. Our simulation is based on an
-implementation of Position-Based Dynamics \cite{Muller07}, with all constraints
+implementation of Position-Based Dynamics (PBD)~\cite{Muller07}, with all constraints
 modelled as penalty forces. To keep the rod inside the
@@
 rod to stay within some offset shell, again using a simple penalty force.
-In the case of multiple initial paths, we solve for each smoothed path sequentially.
+In the case of multiple initial paths, we solve for each smoothed path sequentially.\bjoern{Why would we ever end up with multiple paths here?
+It thought we processed each path by running it through the (A*-PBD-Cutout) pipeline.
+So this step should only ever see one path, no?}
 After solving for a path, we add it as a constraint to the system, generating
 a penalty force that pushes additional paths away from it.
+\bjoern{are there any guarantees that two paths are more than $pipe1{radius}+pipe2{radius}$ away from each other?
+if not, we can't really claim that we always avoid intersections - it just usually happens.}
 
@@
 so we simply halt after a fixed number of timesteps (250). 
-The simulation is very fast so this only takes a second for most of our examples.
+The simulation takes about a second for most of our examples.
+
+Pipes are routed and cut out of the mesh serially.
+This prevents pipe intersections: once a pipe has been cut out of the mesh, it cannot be part of valid routes in any future pipe routing operation.
+However, we do not currently track all pipes for global optimization: we greedily select the best routing per pipe.
+Our voxel grid currently uses a resolution of 128x128x128 voxels, which enables interactive operations, but this parameter can be adjusted for different model requirements.
 
 \subsubsection{Pipes with Multiple Endpoints}
-To allow designers to create pipes with multiple endpoints (e.g., star topologies in Figure \ref{fig:toys}), we use one surface position per pipe and a central fixed endpoint that the user can interactively position.
+To allow designers to create pipes with multiple endpoints (e.g., star topologies in Figure \ref{fig:toys}), we use one surface position per pipe and a central fixed endpoint that the user can interactively position (see Figure~\ref{fig:boat-star}.
 The user can connect several then route the pipes as normal.
+
+\begin{figure}[b]
+\centering
+    \includegraphics[width=1.0\columnwidth]{figures/boat-star.png}
+\caption{Left: Users can interactively position connection points for pipe topologies; they can also adjust pipe diameter and rod parameters. Right: Resulting model with cut pipes.}
+\label{fig:boat-star}
+\end{figure}
 
@@
 In pipes where interior path layout matters, such as the neon sign in Figure \ref{fig:tool-process-interior} or the maze in Figure \ref{fig:maze}, we must create a single long path which passes through all desired path segments (i.e., all letters in the word ``UIST'' or all paths in the maze).
-We built a tool which leverages graph theory to generate this single long path.
+\systemname leverages graph theory to generate this single path.
 
@@
 A maker can import a vector graphics file (such as SVG) describing the path she desires for her pipes.
-We create a graph based on this input data, then add edges to make it connected and semi-Eulerian.
-We create an Euler tour on the modified graph, and thicken the path to create pipes.
+We create a graph based on this input data, with edges for all line segments; then add edges to connect graph components and make graph the semi-Eulerian.
+We find an Euler tour on the modified graph, and thicken the path to create pipes.
 We then cut these pipes from the mesh.
@@
 \caption{An input vector graphics file with the points which cannot be tubed as drawn highlighted in {\color{blue}blue} and the start and end points highlighted in {\color{gray}gray} (a).
-The connected graph created by our software (b) and the resulting Euler circuit (c) permit creation of a novel neon sign (d).}
+The connected graph created by our software (b) and the resulting Euler circuit (c) permit creation of a novel neon sign (d).
+\bjoern{replace sub-figure d?
+looks strange.}}
 \label{fig:tool-process-interior}
@@
 
-\subsubsection{Routing} \george{condense or just point to appendix}
-To allow for a single path through a pipe for post-print media insertion, we need to find a semi-Eulerian graph of which a user's input graphics are a subset, and subsequently find an Euler tour on that graph.
+\subsubsection{Routing}
+%To allow for a single path through a pipe for post-print media insertion, we need to find a semi-Eulerian graph of which a user's input graphics are a subset, and subsequently find an Euler tour on that graph.
+\bjoern{just said the same thing in the last section.}
+%
+The interior path routing problem is a version of the Chinese Postman Problem\footnote{The CPP is also known as the route inspection problem: \url{http://en.wikipedia.org/wiki/Route_inspection_problem}}, in which we wish to traverse all edges of the graph described in the user's input, much as a postman needs to walk along every road at least once to deliver mail.
+In our relaxation, we allow creation of new edges (i.e., the postman may jump from building to building in addition to using existing roads).
+%
+%If path components are disconnected, we create edges that connect them; additionally we can create edges connecting odd-degree vertices rather than simply retracing existing edges.
+In the final artifact, all created edges will be blocked out by dark material so the inserted medium is not visible (see Figure \ref{fig:tool-process-interior}).
+We offer a short algorithm here, with a more mathematically precise definition and associated proof in supplemental materials appendix.
 
-The interior path routing problem is a version of the Chinese Postman Problem\footnote{The CPP is also known as the route inspection problem: \url{http://en.wikipedia.org/wiki/Route_inspection_problem}}, in which we wish to traverse all edges of the graph described in the user's input, much as a postman needs to walk along every road at least once to deliver mail.
-In our relaxation, we allow creation of new edges (i.e., the postman may traverse buildings in addition to roads).
-If path components are disconnected, we create edges that connect them; additionally we can create edges connecting odd-degree vertices rather than simply retracing existing edges.
-In the final artifact, all created edges will be blocked out by dark material so the inserted medium is not visible (see Figure \ref{fig:tool-process-interior}).
-We offer a short algorithm here, with a more mathematically precise definition and associated proof in the appendix.
+{\bf Setup:}We first add a temporary edge that connects the user's desired start and end points, and create a full Eulerian graph.
+This edge is removed at the end to yield a semi-Eulerian graph that can be threaded after printing.
 
-In order to create a semi-Eulerian graph, we first add a temporary edge to the start and end points.  This is removed after the graph is made fully Eulerian.
-
-To connect disconnected subgraphs in the input, each disconnected subgraph's vertices and edges are contracted to a single vertex.
+{\bf Connect disconnected components:} To connect disconnected subgraphs, we find minimum Euclidean distance vertex pairs that span two subgraphs, and greedily add edges to such pairs until all subgraphs are joined into a single connected graph.
+%To connect disconnected subgraphs in the input, each disconnected subgraph's vertices and edges are contracted to a single vertex.
 Each subgraph-vertex has no outgoing edges, as the expanded subgraphs are disconnected.
@@
 
-In order to make our graph Eulerian, we consider all odd degree vertices in the connected graph and make a clique of potential edges based on distance.
+{\bf Eulerization:} In order to make our graph Eulerian, we consider all odd degree vertices in the connected graph and make a clique of potential edges based on distance.
 We greedily add edges between the odd nodes until no odd nodes remain.
 Finally, we remove our temporary edge, which changes the graph from Eulerian to semi-Eulerian.
-This is a connected, semi-Eulerian graph which contains all edges in the input (see Appendix).
+This is a connected, semi-Eulerian graph which contains all edges in the input.
+A lower total weight matching may be possible by connecting components and ensuring node evenness together in a global process, as well as by using minimum-weight matching rather than greedy selection, however this optimization is not crucial for our purposes.
 
-We believe that a lower total weight matching is possible by connecting components and ensuring node evenness together in a global process, as well as by using minimum-weight matching rather than greedy selection, however this optimization is not crucial for our purposes.
-
-/george{too complicated}
-Once we have a semi-Eulerian graph, we need to create an Euler tour.
+{\bf Euler Tour:} To find an Euler tour, we use Fleury's algorithm\footnote{\url{http://en.wikipedia.org/wiki/Eulerian_path\#Fleury.27s_algorithm}}, with a modification to preserve continuity: at a given node, out of multiple candidate edges, we prefer the edges with the smallest angular deviation from the incoming edge (i.e., we prefer to pass straight through a node, if possible).
+This minimizes turns in the final artifact, which eases support material removal and assembly.
+Our EL wire in Figure \ref{fig:UIST} follows an Euler tour through the input nodes.
+%/george{too complicated} \bjoern{shortened this entire section}
+%Once we have a semi-Eulerian graph, we need to create an Euler tour.
 We use a weighted modification of Fleury's algorithm for finding Euler tours\footnote{\url{http://en.wikipedia.org/wiki/Eulerian_path\#Fleury.27s_algorithm}} for this.
@@
 along the path, and subtract the tube from the mesh. We only use circular profiles
-in our examples, but clearly any other profile could be used. Similarly our interface
-allows different start and end radii, which are linearly interpolated along the path.
+in our examples, but clearly any other profile could be used.  Our interface
+allows different start and end radii, which are linearly interpolated along the path. \bjoern{why is this useful?}
 
@@
 
-\begin{figure}[h!]
+\begin{figure}[t]
 \centering
@@
 
-\subsection{Fabrication Techniques}
-
-Beyond design, the tubes must actually be printable to be useful.
-When geometries contain overhangs, as there may be in designed pipes and cavities, 3D printers lay down support material that can be later removed from the model.
+\bjoern{made this a top-level section since it's really not about the software.}
+\section{Fabrication Techniques}
+Several pragmatic concerns arise when fabricating models with hollow pipes and cavities on additive manufactuing machines.
+Pipes will frequently lead to overhangs (where an underlying supporting layer has a smaller footprint than the subsequent layer).
+%Beyond design, the tubes must actually be printable to be useful.  
+%When geometries contain overhangs, as there may be in designed pipes and cavities, 
+3D printers lay down support material that must be later removed from the model to allow overhangs.
 This removal process can be time-intensive and challenging.
@@
 
-FDM machines lay down lines of plastic on top of each other: this allows ``stepover'', when a layer of plastic is laid slightly offset from the layer below it, and ``bridging'', when plastic is laid across a gap between two surfaces.
+\subsection{Fused Deposition Modeling}
+FDM machines lay down lines of extruded plastic filament on top of each other: this allows {\em stepover}, when a layer of plastic is laid slightly offset from the layer below it, and {\em bridging}, when plastic is laid across a gap between two surfaces.
 For a layer to fix correctly to its supporting layer, roughly half of the filament needs to overlap.
-This means that overhangs without added support material can be approximately $45^{\circ}$, and although cylinders (e.g., pipes) and spheres violate this in some places they are printable support-free.
-Support is made from the same material as the model, laid in a heatsink-like wave pattern.
-It is straightforward to remove from exterior surfaces with fingers or pliers, but nearly impossible to remove from internal cavities.
+This means that overhangs without added support material can be approximately $45^{\circ}$, and although cylinders (e.g., pipes) and spheres violate this in some places they can be printed support-free.
+\bjoern{Add a sentence: The models in Figure 1a were produced on a Makerbot FDM without any support structures inside the pipes.}
+
+On single material hobbyist FDM printer, support structures are built using the same material as the model.
+Such support may be impossible to remove from internal cavities.
 Meshmixer has a built-in tool for minimizing support for FDM prints; it re-orients the part to keep as much surface area as possible within the 45 degree constraint.
-In future work we would like to explore a tool in which a user can select surfaces to prioritize support avoidance for.
+In future work we would like to explore a tool in which a user can select surfaces (pipes) that should be prioritized for support avoidance.
 
-On multi-jetting machines, bridging is not possible and stepover potential is much less, because the machine is laying down droplets of liquid rather than more structural plastic.
-Unsupported overhangs can be approximately $14^{\circ}$.
-The support structure used by this machine is a loose matrix of material that can be removed by hand from the surface, blasted out with a water jet from cavities, or dissolved in lye (the enclosing model material does not dissolve).
-Thus, any cavity to which there is some access can be cleared of support.
+\subsection{Multi-Jetting}
+On multi-jetting machines, bridging is not possible and stepover potential is much less, because the machine is laying down droplets of liquid that are later UV cured.
+Unsupported overhangs can be approximately $14^{\circ}$ on our Objet printer.
+Multi-jetting and higher-end FDM machines use a secondary support material that must be manually removed afterwards.
+Some support can be melted out (Projet wax material); other support can be dissolved in lye or blasted with high-pressure water.
+Complex pipe geometries can make water blasting impossible and other processes more labor- and time-intensive.
 
-For especially complex geometries, models can be ``cut up'' into assembleable pieces that evade support requirements or ease material removal.
+For especially complex geometries, models can be ``cut up'' into multiple pieces that can be assembled to avoid support deposition or at least ease material removal.
 This is undesirable as it means more time spent in assembly post-print, and additionally a cut pipe may not be fluid-tight any longer.
 An automatic tool for this task might be beneficial, however it is outside the scope of this work.
+\bjoern{again, add a sentence about what we did for our objet models.}

### 2014-04-09 paper/exampleobjects.tex
bjoern's edits through end of tools section
bjoern's comments through example objects

@@
+%!TEX root = series of tubes.tex
 \begin{figure*}
@@
 \section{Example Objects}
-
-To evaluate and highlight our tool's capabilities, we fabricated a set of five prototype objects designed using a Series of Tubes.
+\bjoern{subsection headings need to be updated in light of the simplification and label changes in the design space.}
+To evaluate and highlight our tool's capabilities, we fabricated a set of five prototype objects designed using \systemnamenospace.
 All prototypes were fabricated in a single piece, unless described otherwise.
+\bjoern{2 of 5 will have been fabricated otherwise -  Radio, Neon sign - both in pieces.}
 
@@
 
-In order to demonstrate our star topology feature and show how conductive fluids can be used for touch sensing, we created a set of touch-sensitive toys and a companion app, reminiscent of the boat application in Acoustic Barcodes \cite{Harrison-acoustic}.
-The objects --- a brain and boat, in our example --- can be set on a base, which announces ``brain'' or ``boat'', and also the names of special features on the object when they are touched.
-For example, touching the brain model's front protrusion yields the announcement ``olfactory bulb''.
-The distinct touch points on each object are connected by an interior star topology of tubes filled with conductive copper paint, and touch sensing is performed via a single wire and SFCS (see Figure \ref{fig:toys}).
-We built a smart base which can distinguish between the toys and also determine which toy is mounted: since each toy and each gesture has a distinct capacitive signature, we use a simple classifier trained to detect both toy and gesture based on profile.
+To show how conductive fluids can be used for touch sensing, we created a set of touch-sensitive toys and a companion app that identifies the objects as well as touched parts of the objects (inspired by \cite{Harrison-acoustic}).
+The objects --- a brain and boat, in our example --- can be set on a base, upon which a speaker announces ``brain'' or ``boat''.
+Touching the brain model's front protrusion yields the announcement ``olfactory bulb''.
+The distinct touch points on each object are connected by an interior star topology, and touch sensing is performed via a single conductive connection to the base, using swept-frequency capacitive sensing~\cite{Sato-touche} (see Figure \ref{fig:toys}).
+%We built a smart base which can distinguish between the toys and also determine which toy is mounted:
+Since each toy and each gesture has a distinct capacitive signature, we use a simple classifier trained to detect both toy and gesture based on profile.
 
@@
 This paint requires approximately 2 days to fully dry in this configuration (3mm diameter tubes, maximum tube length 10cm).
-Our smart base is powered by an Arduino Uno running open-source SFCS code\footnote{http://www.instructables.com/id/Touche-for-Arduino-Advanced-touch-sensing/?ALLSTEPS}.
+Our smart base is powered by an Arduino Uno running open-source SFCS code\footnote{http://www.instructables.com/id/Touche-for-Arduino-Advanced-touch-sensing}.
 
@@
 For this, we used a combination air/vacuum pump: one terminal creates a vacuum while the other creates positive pressure.
-Our rabbit has two pipes, one open tube exiting at its mouth and one semi-closed pipe capped with rubberlike material in its abdomen.
+Our rabbit has two pipes, one open tube exiting at its mouth and one semi-enclosed, capped pipe capped in its abdomen.
 We connected one pipe to each of our pump's terminals, and using a programmable power supply we mimic a rabbit's breathing pattern.
@@
 
-A custom radio, built using pipes integrated with traditional electronics, allows users to tune to different stations and listen in.
+A custom radio shows how pipes can be used to integrate electronic components into the user-facing surface of an object.
 This device uses a network of disconnected open pipes threaded with wires to connect our components (two potentiometers, an LED, and a speaker) to a microcontroller (Arduino Pro Micro).
-We hid the microcontroller and a battery-driven power supply in the base of the radio to make it portable.
-For the radio functionality, we attached an Si4703 radio tuner breakout board to the Arduino, which offers headphone-jack output.
+We placed the microcontroller, an Si4703 radio tuner breakout board and a battery-driven power supply in the base of the radio.
 
-In order to create so many pipes (3 per potentiometer, 1 for the LED, 1 for the speaker) and still preserve surface quality after voxelizing, we added a "High Quality" for pipe-cutting.
-This mode voxelizes at a resolution of 512x512x512 instead of our previous 128x128x128.
-It does preserve surface geometry, however it requires around 3 minutes to complete a single cut operation, compared to our nearly-instant smaller voxel grid (computation on a 3D grid increases by the cube of the 1D size increase).
+This design with 8 pipes (3 per potentiometer, 1 for the LED, 1 for the speaker) showed the limitations of our incremental routing strategy: since we voxelize and remesh for each pipe, surface detail is incrementally lost.
+The problem can be addressed by increasing the voxel resolution.
+At 512x512x512 voxels, surface geometry is not notably degraded.
+However, a single cut operation takes around 3 minutes \bjoern{on what hardware?
+mention earlier}, compared to near-instant performance on a $128^3$ voxel grid.
 We fabricated the radio itself on our Makerbot, support-free but cut in two pieces to allow the Arduino and battery pack to be inserted.
 
+\bjoern{I ran out of steam somewhere here...}
 \subsection{Presence-aware Pen Holder (open, solid, ternimals)}
-
+\bjoern{isn't this a threadable instead of a solid?} \bjoern{the model looks like the acutal pipes are really short and basically straight at the bottom of the model.
+This may raise the question if one couldn't just use the line sensors without any fiber optic cable in the middle.}
 Our presence-aware pen holder can distinguish which tool or tools a user has picked up (see Figure \ref{fig:pens}).
-Our pen holder uses a modification of the FlyEye technique described by Wimmer in \cite{Wimmer-flyeye} and contains open tubes filled with fiber optic cables, one per pen chamber.
+Our pen holder uses a modification of the FlyEye technique described by Wimmer in \cite{Wimmer-flyeye} in which \bjoern{describe...}.
+Four internal pipes are threaded with fiber optic cables, one per pen chamber.
 We use a single cable per chamber; our 6mm diameter fiber optic cables, in comparison to the fine tubes used in the original work , can send and receive through the same cable.
-At the base of each tube is a QRE1113 line sensor digital breakout board, which has an integrated IR emitter and receiver.
+\bjoern{is this because it has many strands?}  At the base of each tube is a QRE1113 line sensor digital breakout board, which has an integrated IR emitter and receiver.
 When a pen is in its appointed place, the emitted infrared light is reflected off its bottom and travels back to the receiver, where it registers as bright.
-This prototype was built by our Objet and support was flushed post-print.
-\valkyrie{this whole paragraph is unclear}
+This prototype was built by our Objet and support was flushed post-print.\bjoern{Incorrect.
+Looks like a Makerbot print.} \valkyrie{this whole paragraph is unclear}
 
@@
 (a) shows the input SVG files, with the two states of the mouth drawn in different colors.
-(b) shows the resultant mesh when we subtracted the generated tubes.
-In (c), we show the completed sign.}
+\bjoern{it's hard to tell these colors apart!}  (b) shows the resultant mesh when we subtracted the generated tubes.
+\bjoern{hard to see what's going on here, too.} In (c), we show the fabricated sign.}
 \label{fig:neon}
@@
 We designed a custom neon robot head which can be animated to ``talk''.
-The pipes have been threaded with EL wire which is lit in sequence to create an animation.
-Our sign was fabricated on the Objet in two halves, cutting through the plane of the sign itself, and with pipes cut in the rear piece to allow the wire to exit backwards and hide the EL wire controls.
+The pipes have been threaded with electroluminescent wire which is lit in sequence to create an animation.
+Our sign was fabricated on the Objet in two halves, cutting through the plane of pipe path to facilitate assemblhy.
+Pipes enter and exit through the rear to hide the EL wire controls.
 The EL wire was threaded post-print, and is controlled by an EL wire sequencer.
@@
 This maze is based on an SVG we created by hand and processed with our internal path tube tool.
-To fabricate the tube, we created it in two halves that were fastened together via glue, due to limitations in our ability to remove support material otherwise.
-\bjoern{observant reviewers will call us out on this - that many of our techniques may not actually be workable in a single piece because you can't get the support material out.
-I think if you had a lye bath you could actually get more complicated models done, as long as you're patient and there's some entry and exit for the lye to reach all pipes.}  \valkyrie{we should discuss.
-I will try to get a model of this ready ASAP}  We created this prototype on our Objet in two halves, inserting a marble before closing.
+To fabricate the tube, we created it in two halves that were fastened together via glue, due to limitations in our ability to remove support material otherwise (we only have access to a high-pressure water tool, not a lye bath).
+We created this prototype on our Objet in two halves, inserting a marble before closing.
+\bjoern{check - i think the latest maze is in fact done with our tool?
+Will need rewriting.}

### 2014-04-09 paper/discussion.tex
discussion

@@
+%!TEX root = series of tubes.tex
 \section{Discussion}
+%\valkyrie{there are a lot of things in here that we could focus on.  input/output/identity interactions mediated by tubes, the creation of antennae and other custom electronics...}
+%
+%\valkyrie{which figures/discussions do we think are important?}
 
-\valkyrie{there are a lot of things in here that we could focus on.  input/output/identity interactions mediated by tubes, the creation of antennae and other custom electronics...}
+The promise of 3D printing is two-fold: it enables the manufacturing of complex objects at low volumes (n=1), and it allows for the manufacturing of geometries that are impossible to create using traditional techniques like injection molding.
+Printing pipes to prototype interactive objects exploit both advantages.
+Prototypes by definition are low-volume parts, and pipes require complex interior geometries that are very difficult or impossible to produce using other processes.
+Our work has focused on routing pipes that are then later manually filled after a print has completed.
+However, the fundamental operations of 3D path routing and topology specification that our tool offers will remain relevant as 3D printing technology advances.
+Multi-material printers are already available; as they become able to lay down additional materials (e.g., conductive), \systemname can enable designers to efficiently specify where different functional materials should be placed at design time.
 
-\valkyrie{which figures/discussions do we think are important?}
-
-The promise of 3D printing is that it allows for the manufacture of geometries that are impossible to create using traditional techniques like injection molding.
-Pipes leverage that promise: they require a previously unmanufacturable complexity of interior geometry.
-Printers additionally continue to evolve in their capabilities to lay down more materials.
-As they become able to lay down conductive, optically clear, or luminescent materials, our tools and techniques will enable even more complex internal designs.
-
-\begin{figure}[h]
-\centering
-    \includegraphics[width=3.4in]{figures/electronics.png}
-\caption{(a) A capacitor we manufactured on a 3D printer and coated, post-print, with copper paint.
-This particular one holds a charge of xx.
-(b) Comparison of theoretical printed resistors.
-For a cylindrical wire, resistance = $\rho\frac{l}{\pi r^2}$, where $\rho$ is based on the conductivity of the medium.
-In our case, assuming $ r << l$, resistance(B) < resistance(C) < resistance(A).
-(c) A possible design for a cell phone antenna augmentation inside a cell phone case.}
-\label{fig:electronics}
-\end{figure}
-
-These internal designs can give rise to new opportunities in electronics design and fabrication: we printed a proof-of-concept printed capacitor which features a thin printed plate with hollow cones on either side (see Figure \ref{fig:electronics}A).
-After printing, we coated the inner surfaces of capacitor with copper paint.
-It now can hold a charge of xx.
-Similarly, tubes can be used to create \emph{resistors} in ``printed'' circuitry: a longer or thinner tube has a greater resistance than a shorter or wider one (see Figure \ref{fig:electronics}B).
-Internal \emph{antenna} design is also possible, which could lead to new flexibility in, e.g., cell phone design (see Figure \ref{fig:electronics}C).
-
-By crossing pipe types, topologies, and inserted media, many pipe-mediated \emph{interactions} also emerge.
-We demonstrated a few such possibilities by designing and fabricating example objects with the help of our pipe routing design tool.
-However, there are many additional possibilities for leveraging sensing and actuation techniques from the HCI literature and extending their use in 3D-printed objects.
-We briefly survey such possibilities for output and input.
+We described a design space and fabricated example objects to demonstrate some of the interactions that can be aided by pipe routing.
+There are many additional possibilities for leveraging sensing and actuation techniques from the HCI literature and extending their use in 3D-printed objects.
+We briefly survey some of them.
 
@@
 Pipe-based outputs can address the five senses: visual, aural, haptic, and olfactory/gustatory.
-In addition to outputs we mention explicitly, any outputs possible using traditional electronic components are possible using conductive material in pipes.
+%In addition to outputs we mention explicitly, any outputs possible using traditional electronic components are possible using conductive material in pipes.
 
@@
 This could be used to create logic gates, as in (d).
-These could be used, for example, in science exhibits.}
+These could be used, for example, in science exhibits.
+\bjoern{If we use this figure, do the top row only, not the bottom (too big).}}
 \label{fig:direct}
@@
 \emph{Visual} outputs can be mediated by gases, liquids, solid, particulates, or threadables.
-Pneumatic actuation (as in PneUIs \cite{Yao-pneui} or Harrison's latex buttons \cite{Harrison-buttons}) provides output, but gases carrying smoke also offer visual feedback.
-Splitting and mixing of colored liquids can be used as a display tactic; additionally mechanical motion can be fluid-mediated.
-Liquids can also be mixed in specific concentrations by servo-controlled gate structures as in Figure \ref{fig:direct}.
+Pneumatic actuation (as in PneUIs \cite{Yao-pneui} or Harrison's latex buttons \cite{Harrison-buttons}) can provides shape-changing interfaces if maelleable materials can be printed.
+%Gases carrying smoke also offer visual feedback. 
+Air bubbles in liquids can also convey information~\cite{Heiner1999Percolator}.
+Splitting and mixing of colored liquids can be used as a display tactic; liquids can also be mixed in specific concentrations by servo-controlled gate structures as in Figure \ref{fig:direct}.
+%additionally mechanical motion can be fluid-mediated.
 
 \emph{Aural} outputs can be created with gases.
-Helmholz resonance can be leveraged to create computer-controllable sound chambers, as in Zoran's flute \cite{Zoran-flute}.
-All that is necessary for this is an enclosed chamber with a pipe connected to it via a narrow neck-point (see Figure \ref{fig:ocarina}).
+Pipes and cavities can act as resonators can be leveraged to create 3D-printed sound chambers with explicitly designed sound characteristics~\cite{Zoran-flute} (see Figure \ref{fig:ocarina}).
+%All that is necessary for this is an enclosed chamber with a pipe connected to it via a narrow neck-point (see Figure \ref{fig:ocarina}).
 Passive sound amplification and sound redirection are also possible using pipes.
@@
 Fluids in pipes capped with flexible membranes can actuate particulates and create programmable pliability as in Jamming User Interfaces \cite{Follmer-jamming}.
+\bjoern{Didn't we describe the first part already?
+Also I'm still skeptical about jamming, maybe skip it.}
 
-\emph{Olfactory} and \emph{gustatory} outputs can be created through the mixing and splitting of pipes carrying scented or flavored fluids, respectively.
+\emph{Olfactory} and \emph{gustatory} outputs can be created through the mixing and splitting of pipes carrying scented or flavored fluids, respectively~\cite{kaye2004making}.
 
 \subsection{Inputs}
-
-Pipes create opportunities for many types of input sensing across the surface of printed devices.
+%In addition to the described ways for sensing touch and object presence, additional opportunities also exist for leveraging pipes to sense pressure, grasp, flexing, and tapping. 
+\bjoern{most of the beginning of this section described techniques we had already implemented or described earlier in the paper.
+They  referenced earlier figures and earlier citations.
+I commented them out as I think this section is about additional techniques nto already described.
+For example, we referenced slyper  in the design space.
+this feels empty now though.
+I'm not quite sure what the right thing to do is.}
+% and manipulation of traditional electronic components.
+%Pipes create opportunities for many types of input sensing across the surface of printed devices.
 We describe sensing touch, pressure, grasp, flexing, tapping, and manipulation of traditional electronic components.
@@
 
+%\bjoern{we alsready covered touch in the paper...
+think this section should be about different ways.
 \emph{Touch} sensing can be enabled through pipes filled with conductive media.
@@
 Using a technique like Fiberio \cite{Holz-fiberio}, users could even be identified on touches.
+\bjoern{Fiberio requires fiber optic plate - not sure it applies to pipes.}
 
-\emph{Pressure} can be sensed in multiple ways.
+%\emph{Pressure} can be sensed in multiple ways.
 Slyper, et al.,'s primitives can be attached as the terminus of any semi-closed pipe; the user can manipulate the printed endpoint and the system can sense air pressure changes \cite{Slyper-pressure}.
@@
 
-\emph{Grasp} sensing can be enabled using Wimmer's FlyEye techniques \cite{Wimmer-flyeye} or those described in Jamming User Interfaces \cite{Follmer-jamming}.
+%\emph{Grasp} sensing can be enabled using Wimmer's FlyEye techniques \cite{Wimmer-flyeye} or those described in Jamming User Interfaces \cite{Follmer-jamming}.
 The optical links necessary for sensing can be created either via clear solid cores as in Printed Optics \cite{Willis-printedoptics} or via fiber optic cables threaded through hollow pipes (see Figure \ref{fig:pens}).
 
-Some 3D printers can fabricate rubber-like materials.
+%Some 3D printers can fabricate rubber-like materials.
 Similar to Slyper, et al., in \cite{Slyper-shape}, we can sense \emph{flexing} and \emph{bending} of prints made on these machines: the ability to 3D print devices to sensing bending and flexing significantly cuts molding and assembly time.
 
-\emph{Tapping} is another input possibility.
-Due to differential sound conductivity in different printed materials, pipes of a greater conducting material can be embedded in a model of a lesser conducting material.
+\emph{Tapping and Grasping} can be sensed acoustically.
+Due to differential sound conductivity in different printed materials, solid pipes of a material with greater conductivity (i.e., rigid, high-density) can be embedded in a model of a lesser conductivity (i.e., softer).
 A microphone or piezo placed at the system-side of these sound-conducting pipes can determine where the model was tapped and how it was tapped.
-Active acoustic sensing as described in Touch \& Activate \cite{Ono-touchandactivate} is also possible; this technique would allow for particular areas of interest (connected to sound-conducting pipes) to be touch sensitive, while other areas (made of non-conducting material) are not sensed.
+Active acoustic sensing as described in Touch \& Activate \cite{Ono-touchandactivate} is also possible; this technique would allow for particular areas of interest (connected to rigid pipes) to be touch sensitive and enable grasp sensing.
 
-Sensing via traditional electronic components (e.g., potentiometers, alcohol gas sensors) can be accomplished by conductive inserted media.
-Components can be recessed into a print's surface, with their leads implanted into open pipes.
-By inserting liquid copper paint instead of threading traditional wires, all components in a model can share a ground line, and the paint's drying process obviates the use of solder or glue to affix the components in place.
-
-\subsubsection{Identification}
+\subsection{Identification}
+\bjoern{this section should be about half as long as it is right now.}
 In addition to active inputs from users, pipes can be used for passive \emph{identification}.
 While the pipes need not be visible (especially in the case of fully-enclosed pipes), their presence, location, network structure, and length change the physical properties of a printed object.
-This includes weight, acoustic resonance, and (if conductive material is present) capacitive signature.
+This includes weight, acoustic resonance, and capacitive signature.b
 
@@
 Weight is a simple metric of object identity, but weights are less distinctive than acoustic or capacitive signatures, as they are represented by a single number.
+\bjoern{second half of this paragraph meanders.
+focus on the key story.}
 
@@
 A pipe's resonant frequency can be measured by attaching a speaker and microphone to it, sweeping frequencies with the speaker, and performing a Fourier transform on the resultant signal from the microphone.
-Peaks in the transformed data correspond to stronger returned impulses: the resonant frequencies of the object.
+Peaks in the transformed data correspond to resonant frequencies of the object.
 Intentional encoding in the case of capacitance is significantly more difficult, and we leave it to future work.
@@
 
-\bjoern{We might pull back on the scope of this section a bit and think about which directions here are most closely related to pipes and most likely successful;  and which are more tangential or speculative - for example, Aireal is about feedback in free space and we are all about designing tangible objects; so maybe that's a candidate for cutting.}
+\begin{figure}[h]
+\centering
+    \includegraphics[width=3.4in]{figures/electronics.png}
+\caption{(a) A capacitor we manufactured on a 3D printer and coated, post-print, with copper paint.
+This particular one holds a charge of xx.
+(b) Comparison of theoretical printed resistors.
+For a cylindrical wire, resistance = $\rho\frac{l}{\pi r^2}$, where $\rho$ is based on the conductivity of the medium.
+In our case, assuming $ r << l$, resistance(B) < resistance(C) < resistance(A).
+(c) A possible design for a cell phone antenna augmentation inside a cell phone case.
+\bjoern{skip the antenna since we haven't investigated this.
+If you have the capacitor and you're confident you know how to measure its charge, add it.
+The different conductors with different measurements are easy to do and worthwhile.
+Maybe combine these two with the helmholtz resonator into one smaller figure.}}
+\label{fig:electronics}
+\end{figure}
 
-\valkyrie{I cut a few things, and removed references to techniques we have already shown to work (as in the spirit of the intro).  LMK if this is better.}
+\subsection{Printed Conductors and Passive Components}
+
+Sensing via traditional electronic components (e.g., p botentiometers, alcohol gas sensors) can be accomplished by conductive inserted media.
+Components can be recessed into a print's surface, with their leads implanted into open pipes.
+\bjoern{didn't we already do this in the radio?} By inserting liquid copper paint instead of threading traditional wires, all components in a model can share a ground line, and the paint's drying process obviates the use of solder or glue to affix the components in place.
+
+Printing or injecting conductive material can also be used to design passive components with particular electrical properties.
+For example, we printed a proof-of-concept  capacitor which features a thin printed plate with hollow cones on either side (see Figure \ref{fig:electronics}A).
+After printing, we coated the inner surfaces of capacitor with copper paint.
+It now can hold a charge of xx.
+Similarly, tubes can be used to create \emph{resistors} in ``printed'' circuitry: a longer or thinner tube has a greater resistance than a shorter or wider one (see Figure \ref{fig:electronics}B).
+Internal \emph{antenna} design is also possible, which could lead to new flexibility in, e.g., cell phone design (see Figure \ref{fig:electronics}C).
+Future design tools could help designers achieve desired electrical properties automatically.
 

### 2014-04-09 paper/limitations.tex
limitations

@@
 Pipes, while a powerful tool for directing interaction to arbitrary locations on an object's surface, have their shortcomings.
-Most obviously, their fundamental shortfall is that they cannot replace the active elements required for a functioning interactive device: they simply redirect sensing to different locations on an object's surface.
+They cannot replace the active elements required for a functioning interactive device: they simply redirect sensing to different locations on an object's surface.
 All pipes need to, at some point, connect to a sensor, microcontroller, or pump.
@@
 As discussed, support material removal can also be a challenge.
-Particularly on multi-jetting printers, printing support-free is not possible for most geometries, while for hobbyist-grade printers physically detaching printed support is not always trivial.
+Particularly on multi-jetting printers, printing support-free is not possible for most geometries, and detaching printed support is not always trivial.
 
@@
 Future work should also include assisting makers in the fabrication process itself by automatically cutting objects to reduce support material removal issues.
-Additionally, as mentioned there are opportunities to improve the algorithm for neon sign routing, as ours does not guarantee a shortest-path routing.
+%Additionally, as mentioned there are opportunities to improve the algorithm for neon sign routing, as ours does not guarantee a shortest-path routing.
 

### 2014-04-09 paper/designspace.tex
bjoern's comments through example objects
abstract to system updates

@@
 
-\section{The Design Space of Pipes}
-\bjoern{would be nice to have a bridge from related work to here.}
-For the purposes of interaction desing, we have identified four relevant dimensions of pipe design: openings, path constraints \bjoern{note the rename - replicate in figure}, topologies, and inserted media.
+\section{The Design Space of Pipes}\tovi{might want to italicize all words that are part of the design space, such as threadable, tree, star, etc}
+\bjoern{would be nice to have a bridge from related work to here.}\tovi{tried to do this}
+Our review of prior art indicates the potential benifits of using interior pipes for adding interactivity to 3D printed objects, although no general purpose tools currently exist for this purpose.
+To guide the design of such tool, we now proivde a deisgn space of the types of pipes that a maker may want to utlize.
+We have identified four relevant dimensions of pipe design: openings, path constraints \bjoern{note the rename - replicate in figure}\tovi{like the new name}, topologies, and inserted media.
 We describe the space below.
@@
     \includegraphics[width=1.0\columnwidth]{figures/tubespace.pdf}
-\caption{The design space of pipes.  See text for details.}
+\caption{The design space of pipes.
+See text for details.
+\bjoern{needs full caption.} \bjoern{The caption font is messed up.
+This may be from including some math package.
+Look up how to redefine the font to match the UIST template.}}
 \label{fig:pipespace}
@@
 %
-Semi-enclosed pipes are capped on one side.
-Caps can for example be used to create tactile output.
+Semi-enclosed pipes are capped on one side and could be used to create create tactile output.
 By fabricating a cap from a malleable material and attaching a pump to the open ended, haptic feedback can be generated by raising and depressing the flexible cap (see Figure \ref{fig:openings}C).
 %
-A fully enclosed pipe or cavities are entirely internal to the object.
+Fully enclosed pipes (cavities) are entirely internal to the object.
 Fully enclosed pipes may be used as weight modifiers~\cite{Prevost-makeitstand} and for object identification~\cite{Willis-infrastructs}, as in Figure \ref{fig:openings}D.
-They can also contain for water or particles that would otherwise fall out.\bjoern{how would you get the particles inside?}
+If printed in parts and assembled post-print, enclosed pipes could also contain water or particles that would otherwise fall out.\bjoern{how would you get the particles inside?}\tovi{added this info}
 
 \subsection{Path Constraints}
-Pipe designs may require specific start and end points, but leave the precise path flexible.
-For example, circuit schematics specify connectivity, but many routing solutions are valid.
-Alternatively, the shape of the path itself may be important.
-For example, in neon signs, single tubes are bent into letterforms.
-Threading
- electroluminescent (EL) wire through a clear pipe enables makers to emulate neon art (see Figure \ref{fig:openings}B). 
-In a maze (see Figure \ref{fig:maze}), both specific terminals as well as a specific path (the maze itself) matter. 
+Depnding on the purpose of the pipe, it's  geometry may be contrainted to sepcific locations within the 3D model.
+An end-point constrained pipe requires specific start and end points, but the secific route of the internal path is irrelevant.
+This owuld be the case for circuit schematics, where the electronic compoents have a fixed location, but many routing solutions are valid.
+In contrast, with a path-constrained pipe it is the shape of the path that is important.
+For example, threading electroluminescent (EL) wire through a clear pipe enables makers to emulate neon art (see Figure \ref{fig:openings}B).
+A fully contrained pipe requires both specific terminals as well as a specific path.
+This would be the case for an interactive  maze (Figure \ref{fig:maze}).
+Finally, a unconstrained pipe does not need a specific path or terminal points.
+Using a cavity to modify an object's wieght is an example of an uncontrained pipe.
 % Alternatively, the neon sign's output in Figure \ref{fig:neon} is based only on the \emph{path} of the pipe.
@@
 
-We next show how different pipe openings, constraints and topologies can be created with the help of a new design tool, \systemname.
+In the next section we how different pipes within this design space space can be created with the help of our new design tool, \systemname.

### 2014-04-09 paper/introduction.tex
abstract to system updates

--- 2014-04-08 paper/introduction.tex
@@
 Existing sensing and actuation approaches, such as FlyEye \cite{Wimmer-flyeye}, Swept Frequency Capacitive Sensing \cite{Sato-touche}, and Jamming User Interfaces \cite{Follmer-jamming}, can be enhanced by the use of pipes: makers utilizing such I/O strategies can locate input and output at arbitrary points on a printed 3D object's surface.
-We leverage these existing techniques for sensing and actuation while our work's novelty is in \emph{internal modeling} for 3D printing and the \emph{redirection} of I/O to arbitrary locations on a device.
+\tovi{Is enhanced the right word?
+Or would "enabled" be better - since these techniques can now be applied to 3D printed objects} We leverage these existing techniques for sensing and actuation while our work's novelty is in \emph{internal modeling} for 3D printing and the \emph{redirection} of I/O to arbitrary locations on a device.
 
@@
 
-Generating these pipes, however, is not trivial.
+Generating the appropriate pipes for these examples is not trivial.
 For one, most 3D modelling tools do not provide extensive tools for modelling the  {\em interior} of 3D models.
@@
 Pipes should not pass too close to the surface: consumer-grade 3D printers do not guarantee air- or water-tightness of prints, so more than one printed layer may be required to prevent leakage.
-Pipes for applications like neon signs must allow a single EL wire to light every edge on the pipe graph (i.e., have an Euler tour).
-No tools currently exist for these design tasks.
-To aid users in working with 3D-printed pipes, we contribute new techniques for interactively routing and generating internal structures, and integrate these tools into a 3D modelling application.
+Pipes for applications like illumunated EL wire signs must allow a single pipe to follow a specific path layout (e.g., spelling out text).
 
+We contribute PipeDream, an interactive design tool that offers 3D point-to-point routing of internal pipes based on a physical simulation of flexible rods.
+Our system addresses the various challenges that pipe design makers are likely to encounter, such as avoiding intersections and ensuring maximum bend angles.
 We then present a set of example objects, fabricated with the help of our modeling tool, that demonstrate different interaction modalities.
 We conclude by discussing limitations and outlining areas for future exploration.
+
 %Our work makes several contributions to the field of digitally fabricated interactive objects.

### 2014-04-09 paper/relatedwork.tex
abstract to system updates

@@
 Previous work has begun to investigate how to fabricate interactive objects using particular sensing and display technologies.
-Printed Optics (\cite{Willis-printedoptics}) uses light pipes to create integrated displays and sensed input optically through components added at print time; while Slyper, et al., 3D print flexible robot armatures capable of sensing interaction via changes in air pressure \cite{Slyper-pressure}.
+Printed Optics (\cite{Willis-printedoptics}) uses light pipes to create integrated displays and sensed input optically through components added at print time; while Slyper et al., fabricate flexible robot armatures capable of sensing interaction via changes in air pressure \cite{Slyper-pressure}.
 Our work is inspired by these projects, and contributes a general design tool for modelling objects that can have such capabilities.
@@
 %Our work differs from these because we offer \emph{algorithms and techniques} for \emph{internal}, \emph{true 3D} routing.
+\tovi{should we say somewhere that 3d print manufacturers are considering conductive materials, but such printers would be extremely expensive?}
 
@@
 The technique is inspired by work in generating continuous line illustrations from images~\cite{Wong-continuousline,Bosch-tsp}.
-Our approach is also based on graph theory (Euler tours), but differs in that it generates 3D geometry instead of 2D illustrations.
+Our approach is also based on graph theory (Euler tours), but differs in that it generates 3D geometry instead of 2D illustrations.\tovi{I think this paragraph could be omited and the citation inlcuded is hte system section}
 

### 2014-04-09 paper/tubes.tex
abstract to system updates

@@
 \section{\systemname - A Tool for Pipe Design}
-
+\tovi{I understand why we use ``makers'' instead of ``users'', but it might make it sound like the tool requires a certain level of existing hacking skills.
+For example, we are also allowing researchers to design novel objects}
 To allow makers to design novel objects with pipe-powered interfaces, we created a tool, \systemnamenospace, to add pipe geometries to arbitrary 3D models.
@@
 To permit easy insertion of media after printing, pipes should have smooth bends.
-Pipes in general need to avoid objects' surfaces to prevent fluid leakage when printed by hobbyist-grade machines.
-Finally, for designing pipes that follow a user-specified path, we must ensure certain characteristics of that path (e.g., that the path is connected).
+Pipes, in general, need to avoid objects' surfaces to prevent fluid leakage when printed by hobbyist-grade machines.
+Finally, for designing path-constrained pipes, that follow a user-specified path, we must ensure certain characteristics of that path (e.g., that the path is connected).
 %\george{pipes are hard.  not sure we convinced readers by now that they are worth the effort.  maybe this has to happen more in the intro.  the example objects section is still a few pages away.}
@@
 
-Designing objects which, for example, are touch sensitive in particular areas or contain electronic components, requires precise location and sizing of pipe endpoints.
+% Designing objects which, for example, are touch sensitive in particular areas or contain electronic components, requires precise location and sizing of pipe endpoints.
 The interior of these pipes should have as large a bending radius as possible so that post-print insertion of solid or viscous media is easier.
-
-To create pipes in which the location and shape of exterior connection points matters, we first allow makers to select these points on the mesh's surface.
+\tovi{removed the first paragraph as it seemed repetitive.
+Do a better job explaining how this works from the user's perspective.
+What is the UI?
+I added a bit.
+I think it would be best to explain all UI in this paragraph, and just focus on the algortihms below}
+To create endpoint-cosntrained pipes, in which the location and shape of exterior connection points matters, a brush tool is used to specify a pair of points on the mesh's surface.
+The user then clicks the [x] button which casues the system to route a pipe between the specificed points, and update the mesh accordingly by subtracting the pipes geometry fromt the model.
  \bjoern{feels a bit redundant with the last paragraph in the prior section.}
@@
 
-\subsubsection{Routing and Physical Simulation}
+\subsubsection{Routing Algorithm and Physical Simulation}
 Given two points, we first find a path between them using the A* search algorithm on a voxel representation of the object~\cite{Hart-Astar}.
@@
 Our A* pathfinding produces the \emph{shortest} path, but in many cases
-we require a smoother path through which we can actually push a physical wire.
-In addition, distance from the surface is not taken into account in A* search and pipes may thus be routed along surface contours, which may cause leakage.
+we require a smoother path to aid the insertion of the desired medium, such as a threadable wire.
+In addition, the distance from the surface is not taken into account in the A* search and pipes may thus be routed along surface contours, which may cause leakage.
 
@@
 rod to stay within some offset shell, again using a simple penalty force.
-In the case of multiple initial paths, we solve for each smoothed path sequentially.\bjoern{Why would we ever end up with multiple paths here?
+\tovi{might want to specify this penalty force - or just say here that the value of the parameters were tuned on several test cases, and provide the associated values.
+}
+
+%In the case of multiple initial paths, we solve for each smoothed path sequentially.\bjoern{Why would we ever end up with multiple paths here?
 It thought we processed each path by running it through the (A*-PBD-Cutout) pipeline.
-So this step should only ever see one path, no?}
-After solving for a path, we add it as a constraint to the system, generating
-a penalty force that pushes additional paths away from it.
+So this step should only ever see one path, no?}\tovi{agree this is misleading, I think it can be removed since we talk about multiple pipes in a later paragraph}
+
+PBD simulations rarely converge to a steady state, so we simply halt after a fixed number of timesteps (250).
+Our voxel grid currently uses a resolution of 128x128x128 voxels, which enables interactive operations, but this parameter can be adjusted for different model requirements.
+The simulation takes about a second for most of our examples.
 \bjoern{are there any guarantees that two paths are more than $pipe1{radius}+pipe2{radius}$ away from each other?
@@
 
+\tovi{move this paragraph to discussion section}
 This basic approach does have various limitations, in particular it
@@
 scaling the initial mesh to a unit box, and tuning the parameters on several test cases.
-PBD simulations rarely converge to a steady state, 
-so we simply halt after a fixed number of timesteps (250). 
-The simulation takes about a second for most of our examples.
 
+\subsubsection{Multiple Pipes}
 Pipes are routed and cut out of the mesh serially.
 This prevents pipe intersections: once a pipe has been cut out of the mesh, it cannot be part of valid routes in any future pipe routing operation.
-However, we do not currently track all pipes for global optimization: we greedily select the best routing per pipe.
-Our voxel grid currently uses a resolution of 128x128x128 voxels, which enables interactive operations, but this parameter can be adjusted for different model requirements.
+After solving for a path, we add it as a constraint to the system, generating
+a penalty force that pushes additional paths away from it.  However, we do not currently track all pipes for global optimization: we greedily select the best routing per pipe.
 
-\subsubsection{Pipes with Multiple Endpoints}
+\subsubsection{Pipes with Multiple Endpoints}\tovi{need to describe the UI for this}
 To allow designers to create pipes with multiple endpoints (e.g., star topologies in Figure \ref{fig:toys}), we use one surface position per pipe and a central fixed endpoint that the user can interactively position (see Figure~\ref{fig:boat-star}.
@@
 
-In pipes where interior path layout matters, such as the neon sign in Figure \ref{fig:tool-process-interior} or the maze in Figure \ref{fig:maze}, we must create a single long path which passes through all desired path segments (i.e., all letters in the word ``UIST'' or all paths in the maze).
-\systemname leverages graph theory to generate this single path.
+For path-constrained pipes, the route of the pipe must conform to a desired shape.
+A maker can import a vector graphics file (such as SVG) describing the path she desires for her pipes.
+\tovi{explain the ui - how does the user specify the path geometry, execute the command to create the path, etc.}.
+We then cut these pipes from the mesh.
 
-In graph theory, a semi-Eulerian graph is one which has exactly two edges of odd degree.
-This kind of graph supports an Euler tour: a path that travels every edge exactly once, beginning and ending at the two nodes of odd degree (in this case, the user-selected start and end nodes).
-A maker can import a vector graphics file (such as SVG) describing the path she desires for her pipes.
-We create a graph based on this input data, with edges for all line segments; then add edges to connect graph components and make graph the semi-Eulerian.
-We find an Euler tour on the modified graph, and thicken the path to create pipes.
-We then cut these pipes from the mesh.
+In some cases, such as for the neon sign in Figure \ref{fig:tool-process-interior}, a single path must passes through a set of disconnected path segments (i.e., all letters in the word ``UIST'' ).
+\systemname leverages graph theory algorithms to generate this single path, as described below.
 
@@
 %
+
+%In graph theory, a semi-Eulerian graph is one which has exactly two edges of odd degree.
+This kind of graph supports an Euler tour: a path that travels every edge exactly once, beginning and ending at the two nodes of odd degree (in this case, the user-selected start and end nodes).
+We create a graph based on the user-specified geometry data, with edges for all line segments; then add edges to connect graph components and make graph the semi-Eulerian.
+We find an Euler tour on the modified graph, and thicken the path to create pipes.
+
 The interior path routing problem is a version of the Chinese Postman Problem\footnote{The CPP is also known as the route inspection problem: \url{http://en.wikipedia.org/wiki/Route_inspection_problem}}, in which we wish to traverse all edges of the graph described in the user's input, much as a postman needs to walk along every road at least once to deliver mail.
@@
 in our examples, but clearly any other profile could be used.  Our interface
-allows different start and end radii, which are linearly interpolated along the path. \bjoern{why is this useful?}
+allows different start and end radii, which are linearly interpolated along the path. \bjoern{why is this useful?} The radius is specified [how?]
 
@@
 The physical fabrication process used by different machines gives rise to differing strategies for avoiding support material or easing its removal.
-We will discuss techniques for fused deposition modeling (FDM) machines like the Makerbot and multi-jetting machines like the Objet.
+We will discuss techniques for fused deposition modeling (FDM) machines, suchas a Makerbot [ref] and multi-jetting machines such as an Objet [ref].
 
@@
 
-On single material hobbyist FDM printer, support structures are built using the same material as the model.
+On single material hobbyist FDM printers, support structures are built using the same material as the model.
 Such support may be impossible to remove from internal cavities.
@@
 In future work we would like to explore a tool in which a user can select surfaces (pipes) that should be prioritized for support avoidance.
+\tovi{print in parts and then assemble those parts together?}
 
@@
 
+\subsection{Model Assembly}
 For especially complex geometries, models can be ``cut up'' into multiple pieces that can be assembled to avoid support deposition or at least ease material removal.

### 2014-04-09 paper/discussion.tex
final sections

@@
 
-The promise of 3D printing is two-fold: it enables the manufacturing of complex objects at low volumes (n=1), and it allows for the manufacturing of geometries that are impossible to create using traditional techniques like injection molding.
+The promise of 3D printing is two-fold: it enables the manufacturing of complex objects at low volumes, and it allows for the manufacturing of geometries that are impossible to create using traditional techniques like injection molding.
 Printing pipes to prototype interactive objects exploit both advantages.
@@
 However, the fundamental operations of 3D path routing and topology specification that our tool offers will remain relevant as 3D printing technology advances.
-Multi-material printers are already available; as they become able to lay down additional materials (e.g., conductive), \systemname can enable designers to efficiently specify where different functional materials should be placed at design time.
+Multi-material printers are already available; as they advance and support the embedding of additional materials (e.g., conductive), \systemname can enable designers to efficiently specify where different functional materials should be placed at design time.
 
@@
 Similar to Slyper, et al., in \cite{Slyper-shape}, we can sense \emph{flexing} and \emph{bending} of prints made on these machines: the ability to 3D print devices to sensing bending and flexing significantly cuts molding and assembly time.
+As demonstrated in our Custom Radio example object, most forms of input can be easily supported by wiring traditional electronic compents through open pipes that connect to a microcotnroller and power source.
+Components can be recessed into a print's surface, with their leads inserted into the open pipes.
+By inserting liquid copper paint instead of threading traditional wires, all components in a model can share a ground line, and the paint's drying process obviates the use of solder or glue to affix the components in place.
+We have also shown how touch input can be enabled using conductive paint and SFCS \cite{Sato-touche}.
 
-\emph{Tapping and Grasping} can be sensed acoustically.
+Alternatively, it may be possible to sense tapping and grasping acoustically.
 Due to differential sound conductivity in different printed materials, solid pipes of a material with greater conductivity (i.e., rigid, high-density) can be embedded in a model of a lesser conductivity (i.e., softer).
@@
 In addition to active inputs from users, pipes can be used for passive \emph{identification}.
-While the pipes need not be visible (especially in the case of fully-enclosed pipes), their presence, location, network structure, and length change the physical properties of a printed object.
-This includes weight, acoustic resonance, and capacitive signature.b
+Fully-enclosed pipes which ar enot visually apparent can change that weight, acoustic resonance, and capacitive signature of a 3D object.
 
-Both identification by recall and intentional encoding are possible.
+Each of these properties could be used for object identification.
 As seen in Touch \& Activate \cite{Ono-touchandactivate}, different objects have different acoustic signatures.
-Additionally, two objects that are visually identical but which have fully enclosed (or other types) of pipes on their interior can be distinguished acoustically.
+As such, two objects that are visually identical but have differing internal chamber geometry on their interior may be distinguished acoustically.
 Thus we can recall an object's identity once its acoustic signature has been recorded.
 Similarly, the Touch\'{e} system \cite{Sato-touche} relies on distinguishing capacitive signatures of objects, both in static states and as their boundary conditions are changed by human interaction.
-Weight is a simple metric of object identity, but weights are less distinctive than acoustic or capacitive signatures, as they are represented by a single number.
-\bjoern{second half of this paragraph meanders.
-focus on the key story.}
+Weight could also be used as a simple metric of object identity, but would be less distinctive than acoustic or capacitive signatures.
 
-We have experimented with intentional encoding.
+%We have experimented with intentional encoding.
 Semi-closed pipes can function as resonance chambers.
@@
 
-\subsection{Printed Conductors and Passive Components}
-
-Sensing via traditional electronic components (e.g., p botentiometers, alcohol gas sensors) can be accomplished by conductive inserted media.
-Components can be recessed into a print's surface, with their leads implanted into open pipes.
-\bjoern{didn't we already do this in the radio?} By inserting liquid copper paint instead of threading traditional wires, all components in a model can share a ground line, and the paint's drying process obviates the use of solder or glue to affix the components in place.
+\subsection{Passive Components}
 

### 2014-04-09 paper/exampleobjects.tex
final sections

@@
 \section{Example Objects}
+\tovi{it might be intersting to relate our examples to existing HCI papers - in order we have: Touche, Dynamically Changeable Physical Buttons/PneUI, Midas, Suaron.
+I guess the maze doesnt really relate to anything.}
+\tovi{UIST is someitmes cranky about providing enough details so other can reporduce the work. We might want to be more careful about providing specific part numbers of specifications.}
 \bjoern{subsection headings need to be updated in light of the simplification and label changes in the design space.}
 To evaluate and highlight our tool's capabilities, we fabricated a set of five prototype objects designed using \systemnamenospace.
-All prototypes were fabricated in a single piece, unless described otherwise.
-\bjoern{2 of 5 will have been fabricated otherwise -  Radio, Neon sign - both in pieces.}
+\tovi{might want to specify the exact printer models we used here, and say we used multipel printers to show that are approach is printer independant.}
+%All prototypes were fabricated in a single piece, unless described otherwise. \bjoern{2 of 5 will have been fabricated otherwise -  Radio, Neon sign - both in pieces.}
 
@@
 
-The models we used for the toys' exteriors were all downloaded from the free website Thingiverse.
+The models we used for the toys were downloaded from the  Thingiverse website.
+We then used \systemnamenospace to create the desired internal pipes.
 All components were fabricated on a Makerbot, support-free.
@@
 Our smart base is powered by an Arduino Uno running open-source SFCS code\footnote{http://www.instructables.com/id/Touche-for-Arduino-Advanced-touch-sensing}.
+An advatnage of using this paint is that it also serves as an adhesive, so no soldering was required.
+\tovi{confirm}
 
@@
 For this, we used a combination air/vacuum pump: one terminal creates a vacuum while the other creates positive pressure.
-Our rabbit has two pipes, one open tube exiting at its mouth and one semi-enclosed, capped pipe capped in its abdomen.
+The rabbit model was downloaded from the Thingiverse website.
+\tovi{confirm} We then used \systemnamenospace to add two pipes: one open pipe exiting at its mouth, and one semi-closed pipe capped at its abdomen.
 We connected one pipe to each of our pump's terminals, and using a programmable power supply we mimic a rabbit's breathing pattern.
-This example was printed on our Objet in rubberlike material, with support flushed post-print.
+This example was printed on the Objet using a flexible material and its support material was flushed post-print.\tovi{give pump part number/details}
 
 \subsection{Custom Radio (open, threadable, terminals)}
+\tovi{stick with the strucutre - describe what it does, then describe modeling and fabrication (how it was built)}
+A custom radio shows how pipes can be used to integrate electronic components into the user-facing surface of an object.
+The radio includes a power and volume dial, and LED which goes on when the radio is on, and a tuner dial.
+\tovi{confirm}.
+A Arduino Pro Micro microcontroller, Si4703 radio tuner, and  [x volt] power supply are embedded in the base of the radio.
+The sound is played through a single speaker.
 
-A custom radio shows how pipes can be used to integrate electronic components into the user-facing surface of an object.
-This device uses a network of disconnected open pipes threaded with wires to connect our components (two potentiometers, an LED, and a speaker) to a microcontroller (Arduino Pro Micro).
-We placed the microcontroller, an Si4703 radio tuner breakout board and a battery-driven power supply in the base of the radio.
+The radio was manually modelled in SolidWorks.
+We included slots on the 3D model matching the shape of the electronic components that were to be embedded.
+The model was then imported into \systemnamenospace, where we created a network of 8 individual open pipes (3 per potentiometer, 1 for the LED, 1 for the speaker) to connect the the comments to the base of the radio.
+We fabricated the radio itself on our Makerbot, support-free, but cut in two pieces to allow the Arduino and battery pack to be inserted.
+We then threaded the pipes with wires to connect the compoenents to the microcontroller at the base of the radio.
 
-This design with 8 pipes (3 per potentiometer, 1 for the LED, 1 for the speaker) showed the limitations of our incremental routing strategy: since we voxelize and remesh for each pipe, surface detail is incrementally lost.
+%This design with 8 pipes (3 per potentiometer, 1 for the LED, 1 for the speaker) showed the limitations of our incremental routing strategy: since we voxelize and remesh for each pipe, surface detail is incrementally lost.
 The problem can be addressed by increasing the voxel resolution.
@@
 We fabricated the radio itself on our Makerbot, support-free but cut in two pieces to allow the Arduino and battery pack to be inserted.
+TOVI: Move to discussion if you want to keep this text.
 
 \bjoern{I ran out of steam somewhere here...}
+\tovi{relate this to sauron as well}
 \subsection{Presence-aware Pen Holder (open, solid, ternimals)}
@@
 Our pen holder uses a modification of the FlyEye technique described by Wimmer in \cite{Wimmer-flyeye} in which \bjoern{describe...}.
-Four internal pipes are threaded with fiber optic cables, one per pen chamber.
-We use a single cable per chamber; our 6mm diameter fiber optic cables, in comparison to the fine tubes used in the original work , can send and receive through the same cable.
+
+The model was created/downloaded from [x].
+We used \systemnamenospace to add four internal open pipes connecting each pen chamber to the based on the pen holder.
+This prototype was built by our Objet and support was flushed post-print.\bjoern{Incorrect.
+Looks like a Makerbot print.} \valkyrie{this whole paragraph is unclear}
+
+Each pipe was then threaded with a fiber optic cable.
+We use a single cable per pipe; our 6mm diameter fiber optic cables, in comparison to the fine tubes used in the original work , can send and receive through the same cable.
 \bjoern{is this because it has many strands?}  At the base of each tube is a QRE1113 line sensor digital breakout board, which has an integrated IR emitter and receiver.
 When a pen is in its appointed place, the emitted infrared light is reflected off its bottom and travels back to the receiver, where it registers as bright.
-This prototype was built by our Objet and support was flushed post-print.\bjoern{Incorrect.
-Looks like a Makerbot print.} \valkyrie{this whole paragraph is unclear}
 
@@
 
-Neon art, perhaps best known for its association with Las Vegas, is traditionally made from hand-formed glass tubes containing neon gas.
+Neon art is traditionally made from hand-formed glass tubes containing neon gas.
 The tubes light up when a current is passed through them.
@@
 We designed a custom neon robot head which can be animated to ``talk''.
-The pipes have been threaded with electroluminescent wire which is lit in sequence to create an animation.
-Our sign was fabricated on the Objet in two halves, cutting through the plane of pipe path to facilitate assemblhy.
-Pipes enter and exit through the rear to hide the EL wire controls.
-The EL wire was threaded post-print, and is controlled by an EL wire sequencer.
+\tovi{in the system seciton do we ever discuss how entry and exit points are specified for path-constrained pipes?}
+The geomertry of the model was imported into \systemnamenospace, which created the assocaited Pipes using its internal path routing tool.
+We then printed the model on the Objet using a transparent material.
+The model was printed in two halves, cutting through the plane of pipe path, to facilitate assembly.
+
+The pipes were threaded with electroluminescent wire which is lit in sequence to create an animation using an EL wire sequencer. Pipes enter and exit through the rear to hide the EL wire controls.
 
@@
 
-Using a modification of our graph-based algorithm which only cuts the literal drawing (and does not perform Eulerization, etc.), we created a maze game with a single particle trapped in a fully enclosed tube.
-This maze is based on an SVG we created by hand and processed with our internal path tube tool.
-To fabricate the tube, we created it in two halves that were fastened together via glue, due to limitations in our ability to remove support material otherwise (we only have access to a high-pressure water tool, not a lye bath).
-We created this prototype on our Objet in two halves, inserting a marble before closing.
+Our final example object is a maze game with a marble that can be navigated through a maze-like path, with exits on either side of the model.
+
+The maze is based on an SVG file we manually created and processed with our internal path routing tool.
+The Maze was printed on the Objet using a transparent material.
+We created the maze in two halves that were fastened together via glue, to support the removal of support materials using a high-pressure water tool.
 \bjoern{check - i think the latest maze is in fact done with our tool?

### 2014-04-09 paper/discussion.tex
mods from bjoern, new maze image

--- 2014-04-09 paper/discussion.tex
@@
 \section{Discussion}
-%\valkyrie{there are a lot of things in here that we could focus on.  input/output/identity interactions mediated by tubes, the creation of antennae and other custom electronics...}
-%
-%\valkyrie{which figures/discussions do we think are important?}
 
@@
 
-\subsection{Outputs}
+\subsection{Inputs and Outputs}
 
 Pipe-based outputs can address the five senses: visual, aural, haptic, and olfactory/gustatory.
+As we have already discussed many input techniques through our example objects, we focus our discussion around output.
 %In addition to outputs we mention explicitly, any outputs possible using traditional electronic components are possible using conductive material in pipes.
@@
 The gate not only has binary positions (a) and (b), but can direct fractional flow, as well (c).
-This could be used to create logic gates, as in (d).
-These could be used, for example, in science exhibits.
-\bjoern{If we use this figure, do the top row only, not the bottom (too big).}}
+These could be used, for example, in science exhibits.}
 \label{fig:direct}
@@
 %All that is necessary for this is an enclosed chamber with a pipe connected to it via a narrow neck-point (see Figure \ref{fig:ocarina}).
-Passive sound amplification and sound redirection are also possible using pipes.
+Passive sound amplification and sound redirection are also possible using pipes, and sensing of some interactions, e.g., \emph{tapping and grasping}, can be sensed acoustically.
+Due to differential sound conductivity in different printed materials, solid pipes of a material with greater conductivity (i.e., rigid, high-density) can be embedded in a model of a lesser conductivity (i.e., softer).
+A microphone or piezo placed at the system-side of these sound-conducting pipes can determine where the model was tapped and how it was tapped.
+Active acoustic sensing as described in Touch \& Activate \cite{Ono-touchandactivate} is also possible; this technique would allow for particular areas of interest (connected to rigid pipes) to be touch sensitive and enable grasp sensing.
 
@@
 
-\subsection{Inputs}
+%\subsection{Inputs}
 %In addition to the described ways for sensing touch and object presence, additional opportunities also exist for leveraging pipes to sense pressure, grasp, flexing, and tapping. 
-\bjoern{most of the beginning of this section described techniques we had already implemented or described earlier in the paper.
-They  referenced earlier figures and earlier citations.
-I commented them out as I think this section is about additional techniques nto already described.
-For example, we referenced slyper  in the design space.
-this feels empty now though.
-I'm not quite sure what the right thing to do is.}
 % and manipulation of traditional electronic components.
@@
 Similar to Slyper, et al., in \cite{Slyper-shape}, we can sense \emph{flexing} and \emph{bending} of prints made on these machines: the ability to 3D print devices to sensing bending and flexing significantly cuts molding and assembly time.
-
-\emph{Tapping and Grasping} can be sensed acoustically.
-Due to differential sound conductivity in different printed materials, solid pipes of a material with greater conductivity (i.e., rigid, high-density) can be embedded in a model of a lesser conductivity (i.e., softer).
-A microphone or piezo placed at the system-side of these sound-conducting pipes can determine where the model was tapped and how it was tapped.
-Active acoustic sensing as described in Touch \& Activate \cite{Ono-touchandactivate} is also possible; this technique would allow for particular areas of interest (connected to rigid pipes) to be touch sensitive and enable grasp sensing.
 

### 2014-04-09 paper/discussion.tex

--- 2014-04-09 paper/discussion.tex
@@
 \emph{Olfactory} and \emph{gustatory} outputs can be created through the mixing and splitting of pipes carrying scented or flavored fluids, respectively~\cite{kaye2004making}.
-
-\subsection{Inputs}
-%In addition to the described ways for sensing touch and object presence, additional opportunities also exist for leveraging pipes to sense pressure, grasp, flexing, and tapping. 
-\bjoern{most of the beginning of this section described techniques we had already implemented or described earlier in the paper.
-They  referenced earlier figures and earlier citations.
-I commented them out as I think this section is about additional techniques nto already described.
-For example, we referenced slyper  in the design space.
-this feels empty now though.
-I'm not quite sure what the right thing to do is.}
-% and manipulation of traditional electronic components.
-%Pipes create opportunities for many types of input sensing across the surface of printed devices.
-We describe sensing touch, pressure, grasp, flexing, tapping, and manipulation of traditional electronic components.
-There are likely more input sensing techniques, both existing and on the horizon, that are compatible with the use of pipes.
-
-%\bjoern{we alsready covered touch in the paper...
-think this section should be about different ways.
-\emph{Touch} sensing can be enabled through pipes filled with conductive media.
-Traditional wires (threadable) may be used, but conductive paint (liquid) may be more flexible.
-Using conductive paint and SFCS \cite{Sato-touche}, an interior star pipe topology can enable single-wire touch- and grasp-sensing at any set of points on a printed object's surface, like our toys (see Figure \cite{fig:toys}).
-Simple capacitance measurements are possible using the same media, and custom-designed sensors like those created in Savage, et al.,'s Midas system are also possible \cite{Savage-midas}.
-Using a technique like Fiberio \cite{Holz-fiberio}, users could even be identified on touches.
-\bjoern{Fiberio requires fiber optic plate - not sure it applies to pipes.}
-
-%\emph{Pressure} can be sensed in multiple ways.
-Slyper, et al.,'s primitives can be attached as the terminus of any semi-closed pipe; the user can manipulate the printed endpoint and the system can sense air pressure changes \cite{Slyper-pressure}.
-Pressure can also be sensed via capacitance.
-
-%\emph{Grasp} sensing can be enabled using Wimmer's FlyEye techniques \cite{Wimmer-flyeye} or those described in Jamming User Interfaces \cite{Follmer-jamming}.
-The optical links necessary for sensing can be created either via clear solid cores as in Printed Optics \cite{Willis-printedoptics} or via fiber optic cables threaded through hollow pipes (see Figure \ref{fig:pens}).
-
-%Some 3D printers can fabricate rubber-like materials.
-Similar to Slyper, et al., in \cite{Slyper-shape}, we can sense \emph{flexing} and \emph{bending} of prints made on these machines: the ability to 3D print devices to sensing bending and flexing significantly cuts molding and assembly time.
-As demonstrated in our Custom Radio example object, most forms of input can be easily supported by wiring traditional electronic compents through open pipes that connect to a microcotnroller and power source.
-Components can be recessed into a print's surface, with their leads inserted into the open pipes.
-By inserting liquid copper paint instead of threading traditional wires, all components in a model can share a ground line, and the paint's drying process obviates the use of solder or glue to affix the components in place.
-We have also shown how touch input can be enabled using conductive paint and SFCS \cite{Sato-touche}.
-
-Alternatively, it may be possible to sense tapping and grasping acoustically.
-Due to differential sound conductivity in different printed materials, solid pipes of a material with greater conductivity (i.e., rigid, high-density) can be embedded in a model of a lesser conductivity (i.e., softer).
-A microphone or piezo placed at the system-side of these sound-conducting pipes can determine where the model was tapped and how it was tapped.
-Active acoustic sensing as described in Touch \& Activate \cite{Ono-touchandactivate} is also possible; this technique would allow for particular areas of interest (connected to rigid pipes) to be touch sensitive and enable grasp sensing.
 

### 2014-04-09 paper/limitations.tex

@@
 \section{Limitations and Future Work}
-Pipes, while a powerful tool for directing interaction to arbitrary locations on an object's surface, have their shortcomings.
-They cannot replace the active elements required for a functioning interactive device: they simply redirect sensing to different locations on an object's surface.
+While pipes provide a powerful mechanism for directing interaction to arbitrary locations on an object's surface, they do have shortcomings which warrant discussion.
+In paticular, they cannot replace the active elements required for a functioning interactive device: they simply redirect sensing to different locations on an object's surface.
 All pipes need to, at some point, connect to a sensor, microcontroller, or pump.
@@
 Threadable materials in particular can provide challenges, as they can get stuck while being fed into pipes.
-Fluids also  For example, the conductive paint we used in our designs can take multiple days to dry in tubes of sufficient length.
-Any liquid kept in tubes printed by the Objet can cause the exterior of the part to discolor due to seepage.
-Hobbyist machines do not guarantee watertightness of their prints, so leaking of water or air can occur.
+Our system mitagates this challenge by minimizng the bend energy of the interior pipes, but for complex pipe geometries threading could be a challenge.
 
-As discussed, support material removal can also be a challenge.
+Fluids also raise unique challenges.
+For example, the conductive paint we used in our designs can take multiple days to dry in tubes of sufficient length.
+Liquids kept in the tubes can also cause discoloration due to seepage.
+Furthermore, not all printers guarantee watertightness of their prints, so leaking of water can occur.
+
+As discussed, removal of support material can also be a challenge.
 Particularly on multi-jetting printers, printing support-free is not possible for most geometries, and detaching printed support is not always trivial.
+In some cases we printed our models in several pieces, to avoid this problem.
 
-Our design tool is not yet fully capable of realizing all possibilities in the design space of pipes.
-It lacks the network manipulation to make mixing or splitting pipes, unless they are designed in an input vector graphics file.
-Since pipes are committed to the mesh one at a time, our tool does not currently optimize multiple pipes at a time: we simply make greedy pairwise decisions.
+There are also area for improvement and future work with our software tool.
+he design of PipeDreams was based on our prposed design space of pipes.
+However, the design tool is not yet fully capable of realizing all possibilities in this design space.
+In paticaulr, we did not specificalyl explore the mixing or splitting pipes.
+Furthermore, since pipes are committed to the mesh one at a time, our tool does not currently optimize a network of multiple pipes.
+Instead, we utilize a greedy algorithm and route a single pipe at a time.
 This suggests the need for a better general 3D routing algorithm with multiple traces.
@@
 We are also eager to test our tool with users and get feedback from makers who use it in the wild.
+However, based on our own experiences, we believe our system could be readily utlized by makers and researchers, and serve as a useful tool for rapdily proptotyping interactive 3D objects.

### 2014-04-09 paper/designspace.tex
paper mod start

@@
 \section{The Design Space of Pipes}\tovi{might want to italicize all words that are part of the design space, such as threadable, tree, star, etc}
-\bjoern{would be nice to have a bridge from related work to here.}\tovi{tried to do this}
 Our review of prior art indicates the potential benifits of using interior pipes for adding interactivity to 3D printed objects, although no general purpose tools currently exist for this purpose.
 To guide the design of such tool, we now proivde a deisgn space of the types of pipes that a maker may want to utlize.
-We have identified four relevant dimensions of pipe design: openings, path constraints \bjoern{note the rename - replicate in figure}\tovi{like the new name}, topologies, and inserted media.
+We have identified four relevant dimensions of pipe design: openings, path constraints, topologies, and inserted media.
 We describe the space below.
@@
     \includegraphics[width=1.0\columnwidth]{figures/tubespace.pdf}
-\caption{The design space of pipes.
-See text for details.
-\bjoern{needs full caption.} \bjoern{The caption font is messed up.
-This may be from including some math package.
-Look up how to redefine the font to match the UIST template.}}
+\caption{The design space of pipes.  See text for details. \bjoern{needs full caption.}}
 \label{fig:pipespace}

### 2014-04-09 paper/introduction.tex
paper mod start

@@
 \section{Introduction}
-Makers, as well as professional designers, increasingly leverage 3D printers as tools for design work.
+Makers, researchers, and professional designers (jointly "makers" for the remainder of this paper) increasingly leverage 3D printers as tools for design work.
 A wide array of objects, ranging from toy figurines to video game controllers and jet engine blades, are now prototyped (and some times even manufactured) using these machines.
@@
 \tovi{Is enhanced the right word?
-Or would "enabled" be better - since these techniques can now be applied to 3D printed objects} We leverage these existing techniques for sensing and actuation while our work's novelty is in \emph{internal modeling} for 3D printing and the \emph{redirection} of I/O to arbitrary locations on a device.
+Or would "enabled" be better - since these techniques can now be applied to 3D printed objects} \valkyrie{I like enhanced; we are allowing them to be applied in new places}  We leverage these existing techniques for sensing and actuation while our work's novelty is in \emph{internal modeling} for 3D printing and the \emph{redirection} of I/O to arbitrary locations on a device.
 

### 2014-04-09 paper/relatedwork.tex
paper mod start

@@
 Previous work has begun to investigate how to fabricate interactive objects using particular sensing and display technologies.
-Printed Optics (\cite{Willis-printedoptics}) uses light pipes to create integrated displays and sensed input optically through components added at print time; while Slyper et al., fabricate flexible robot armatures capable of sensing interaction via changes in air pressure \cite{Slyper-pressure}.
+Printed Optics \cite{Willis-printedoptics} uses light pipes to create integrated displays and sensed input optically through components added at print time; while Slyper et al., fabricate flexible robot armatures capable of sensing interaction via changes in air pressure \cite{Slyper-pressure}.
 Our work is inspired by these projects, and contributes a general design tool for modelling objects that can have such capabilities.
@@
 Sells, et al., \cite{Sells-reprap} use a heated syringe to deposit metal in printed 2D circuitboard channels on objects' surfaces.
+Other experiments have been done by the RepRap project\footnote{\url{http://reprap.org}} using plastic filament blended with conductive materials, however the reistance is too high to be useful in circuitry.
 Sarik, et al., \cite{Sarik-tracebrush} use aerosol jetting of conductive material onto printed objects.
@@
 Our approach does not deposit such materials automatically, but creates structures inside a 3D model that make manual post-print injection of conductive ink possible.
-%Navarette, et al., created a functioning GPS routed in true 3D (not multilayer boards' 2.5D), hand-built using a one-off routing.  \bjoern{need more context for this}
-%Our work differs from these because we offer \emph{algorithms and techniques} for \emph{internal}, \emph{true 3D} routing.
-\tovi{should we say somewhere that 3d print manufacturers are considering conductive materials, but such printers would be extremely expensive?}
 

### 2014-04-10 paper/designspace.tex
many paper updates.  need to finish print of whistle and resistors, and take photos, then we are done????

--- 2014-04-09 paper/designspace.tex
@@
 \section{The Design Space of Pipes}
-\bjoern{would be nice to have a bridge from related work to here.}
-For the purposes of interaction desing, we have identified four relevant dimensions of pipe design: openings, path constraints \bjoern{note the rename - replicate in figure}, topologies, and inserted media.
+Our review of prior art indicates the potential benifits of using interior pipes for adding interactivity to 3D printed objects, although no general purpose tools currently exist for this purpose.
+To guide the design of such tool, we now provide a deisgn space of the types of pipes that a maker may want to utilize.
+We have identified four relevant dimensions of pipe design: \emph{openings}, \emph{path constraints}, \emph{topologies}, and inserted \emph{media}.
 We describe the space below.
@@
     \includegraphics[width=1.0\columnwidth]{figures/tubespace.pdf}
-\caption{The design space of pipes.  See text for details.}
+\caption{The design space of pipes.  \emph{Pipe openings}, \emph{path constraints}, \emph{topologies}, and inserted \emph{media} give rise to unique input/output possibilities.}
 \label{fig:pipespace}
@@
 \subsection{Openings}
-Pipes inside solid objects can be either open to the outside on both ends, semi-enclosed, or fully enclosed (see Figure \ref{fig:pipespace}).
+Pipes inside solid objects can be either \emph{open} to the outside on both ends, \emph{semi-enclosed}, or \emph{fully enclosed} (see Figure \ref{fig:pipespace}).
 For interactive devices, openings can be either user-facing (in places a user would for example touch a model); or system-facing, where sensors and actuators may be connected.
@@
     \includegraphics[width=1.0\columnwidth]{figures/types.png}
-\caption{Pipe openings: A shows a pair of open tubes filled with copper paint to power an LED.
-B shows a return pipe with EL wire threaded through it.
-C shows two semi-closed pipes capped by rubber membranes.
-In D, a rabbit has a fully enclosed chamber to modify its weight.}
+\caption{Pipe openings: A shows a pair of \emph{open} pipes filled with copper paint to power an LED.
+B shows an \emph{open} pipe with EL wire threaded through it.
+C shows two \emph{semi-closed} pipes capped by rubber membranes.
+In D, a rabbit has a \emph{fully enclosed} chamber to modify its weight.}
 \label{fig:openings}
@@
 
-Open pipes may be used to create electronic circuitry.  For example, an open pipe filled with conductive paint and connected to a battery can provide power to an LED (see Figure \ref{fig:openings}A).
+\emph{Open} pipes may be used to create electronic circuitry.
+For example, an open pipe filled with conductive paint and connected to a battery can provide power to an LED (see Figure \ref{fig:openings}A).
 %
-Semi-enclosed pipes are capped on one side.
-Caps can for example be used to create tactile output.
+\emph{Semi-enclosed} pipes are capped on one side and could be used to create create tactile output.
 By fabricating a cap from a malleable material and attaching a pump to the open ended, haptic feedback can be generated by raising and depressing the flexible cap (see Figure \ref{fig:openings}C).
 %
-A fully enclosed pipe or cavities are entirely internal to the object.
+\emph{Fully enclosed} pipes (cavities) are entirely internal to the object.
 Fully enclosed pipes may be used as weight modifiers~\cite{Prevost-makeitstand} and for object identification~\cite{Willis-infrastructs}, as in Figure \ref{fig:openings}D.
-They can also contain for water or particles that would otherwise fall out.\bjoern{how would you get the particles inside?}
+If printed in parts and assembled post-print, enclosed pipes could also contain water or particles that would otherwise fall out.
 
 \subsection{Path Constraints}
-Pipe designs may require specific start and end points, but leave the precise path flexible.
-For example, circuit schematics specify connectivity, but many routing solutions are valid.
-Alternatively, the shape of the path itself may be important.
-For example, in neon signs, single tubes are bent into letterforms.
-Threading
- electroluminescent (EL) wire through a clear pipe enables makers to emulate neon art (see Figure \ref{fig:openings}B). 
-In a maze (see Figure \ref{fig:maze}), both specific terminals as well as a specific path (the maze itself) matter. 
-% Alternatively, the neon sign's output in Figure \ref{fig:neon} is based only on the \emph{path} of the pipe.
-Focused on just the exterior features is the touch sensitive toy in Figure \ref{fig:toys}, where pipes must exit the toy at the frontal lobe and parietal lobes to make those \emph{specific locations} touch-sensitive.
-A pipe for weight modification does require specific terminals or a specific path (see Figure \ref{fig:openings}D).
+Depending on the purpose of the pipe, its geometry may be constrained to specific locations within the 3D model.
+An \emph{endpoint-constrained} pipe requires specific start and end points, but the specific route of the internal path is irrelevant.
+This would be the case for circuit schematics, where the electronic components have a fixed location, but many routing solutions are valid.
+In contrast, with a \emph{path-constrained} pipe the shape of the path is important.
+For example, threading electroluminescent (EL) wire through a clear pipe enables makers to emulate neon art (see Figure \ref{fig:openings}B).
+A \emph{fully constrained} pipe requires both specific terminals as well as a specific path.
+If a maker wishes the EL wire to exit the rear of the sign to hide the electronics, it would require a fully-constrained path (Figure \ref{fig:neon}).
+Finally, an \emph{unconstrained} pipe does not need a specific path or terminal points.
+Using a cavity to modify an object's weight is an example of an unconstrained pipe.
 
 \subsection{Pipe Topologies}
-Pipe network topologies enable splitting or mixing of media (see Figure \ref{fig:pipespace}).
-Star and tree topologies extend the splitting and mixing primitives, encompassing multiple-in and multiple-out configurations; we use star topologies to create different conductive paths in touch-sensitive toys, with a single terminal for swept-frequency capacitive sensing (Figure~\ref{fig:toys}).
+Pipe network topologies enable \emph{splitting} or \emph{mixing} of media (see Figure \ref{fig:pipespace}).
+\emph{Star} and \emph{tree} topologies extend the splitting and mixing primitives, encompassing multiple-in and multiple-out configurations; we use star topologies to create different conductive paths in touch-sensitive toys, with a single terminal for swept-frequency capacitive sensing (Figure~\ref{fig:examples}a).
 
 \subsection{Media in Pipes}
-After printing, different media can be inserted into pipes to change affordances and capabilities. We consider gas, liquids, solids, particulates, and threadables. 
+After printing, different media can be inserted into pipes to change affordances and capabilities. We consider \emph{gas}, \emph{liquids}, \emph{solids}, \emph{particulates}, and \emph{threadables}. 
 %
-Compressible fluids (``gas'') and incompressible fluids (``liquids'')  inside semi-enclosed pipes can create haptic feedback; pressure can also be sensed to act as an input~\cite{Slyper-shape}.
+Compressible fluids (\emph{``gas''}) and incompressible fluids (\emph{``liquids''})  inside semi-enclosed pipes can create haptic feedback; pressure can also be sensed to act as an input~\cite{Slyper-shape}.
 Gases can be used as carriers for scents or fog.
@@
 %
-%got to here
-Filling pipes with solids at print time enables designers to exploit the difference in material properties between pipe and contents.
-For example, in Printed Optics~\cite{Willis-printedoptics}, different refractive indices between solid cladding material and transparent material inside pipes made light transport along pipes for input and output possible.
+Filling pipes with \emph{solids} at print time enables designers to exploit the difference in material properties between pipe and contents.
+For example, in Printed Optics~\cite{Willis-printedoptics}, different refractive indices between solid cladding material and transparent material inside pipes allowed light transport along pipes for input and output.
 %
-Particulates, either printed in-place or inserted, can be of varying densities and can contain particles of various and variable size.
+\emph{Particulates}, either printed in-place or inserted, can be of varying densities and can contain particles of various and variable size.
 A single large particle can be used for display.
-Small particles in fluid can provide variable stiffness haptics via jamming \cite{Follmer-jamming}.
-\bjoern{I thought jamming is about particles in a vacuum?
-why liquid?}
+Small particles in variable amounts of fluid can provide variable stiffness haptics via jamming \cite{Follmer-jamming}.
 %
-Threadable media, such as regular wire, electroluminescent wire, or commercial fiberoptic cables, can be threaded through pipes post-printing.
+\emph{Threadable} media, such as regular wire, electroluminescent wire, or commercial fiberoptic cables, can be threaded through pipes post-printing.
 Threading such media enables addition of conduction, illumination, or optical sensing using materials that can easily be procured, but cannot themselves be printed, e.g., on consumer printers with a single material.
@@
 
-We next show how different pipe openings, constraints and topologies can be created with the help of a new design tool, \systemname.
+In the next section we how different pipes within this design space space can be created with the help of our new design tool, \systemnamenospace.

### 2014-04-10 paper/discussion.tex
many paper updates.  need to finish print of whistle and resistors, and take photos, then we are done????

@@
 \centering
-    \includegraphics[width=3.4in]{figures/placeholder/direct.jpg}
+    \includegraphics[width=3.4in]{figures/gates.png}
 \caption{By attaching a servo to a gate structure, it is possible to direct fluid through that pipe.
 The gate not only has binary positions (a) and (b), but can direct fractional flow, as well (c).
-These could be used, for example, in science exhibits.}
+This could be used, for example, in science exhibits.}
 \label{fig:direct}
@@
     \includegraphics[width=3.4in]{figures/placeholder/helmholz.jpg}
-\caption{This box (a) contains an enclosed Helmholz resonator, seen in the cutaway view (b).
-This resonator creates a sound when air is blown through the pipe at the top, because air compresses at the small neck joint between the chamber and the pipe.
-\valkyrie{need to get a test print of this that actually works}}
+\caption{Some possible pipe uses.
+The printed whistle in (a) is a Helmholz resonator: it creates a sound when air is blown through the pipe at the front, because air compresses at the small neck joint between the internal chamber and the pipe.
+In (b), a capacitor manufactured on a 3D printer coated post-print with copper paint.
+This particular one holds a charge of \valkyrie{XX}.
+(c) gives a comparison of pipe-based resistors, also filled with copper paint.
+For a cylindrical wire, resistance = $\rho\frac{l}{\pi r^2}$, where $\rho$ is based on the conductivity of the medium.
+In our case, because $r << l$, $resistance(B) < resistance(C) < resistance(A)$.}
 \label{fig:ocarina}
@@
 
-\emph{Haptic} outputs are possible through compressible and incompressible fluids.
-Fluids in pipes capped with flexible membranes can actuate particulates and create programmable pliability as in Jamming User Interfaces \cite{Follmer-jamming}.
-\bjoern{Didn't we describe the first part already?
-Also I'm still skeptical about jamming, maybe skip it.}
+\emph{Haptic} outputs are possible through compressible and incompressible fluids, as discussed, or through threadable inputs: e.g., muscle wire can be inserted in flexible pipes and actuated, similar to shape changing interfaces \cite{Coelho-shapechanging}.
 
@@
 \subsection{Identification}
-\bjoern{this section should be about half as long as it is right now.}
 In addition to active inputs from users, pipes can be used for passive \emph{identification}.
-Fully-enclosed pipes which ar enot visually apparent can change that weight, acoustic resonance, and capacitive signature of a 3D object.
+Fully-enclosed pipes which are not visually apparent can change the weight, acoustic resonance, or capacitive signature of a 3D object.
 
@@
 Thus we can recall an object's identity once its acoustic signature has been recorded.
-Similarly, the Touch\'{e} system \cite{Sato-touche} relies on distinguishing capacitive signatures of objects, both in static states and as their boundary conditions are changed by human interaction.
+Similarly, the Touch\'{e} system \cite{Sato-touche} relies on distinguishing \emph{capacitive} signatures of objects, both in static states and as their boundary conditions are changed by human interaction.
 Weight could also be used as a simple metric of object identity, but would be less distinctive than acoustic or capacitive signatures.
@@
 
-\begin{figure}[h]
-\centering
-    \includegraphics[width=3.4in]{figures/electronics.png}
-\caption{(a) A capacitor we manufactured on a 3D printer and coated, post-print, with copper paint.
-This particular one holds a charge of xx.
-(b) Comparison of theoretical printed resistors.
-For a cylindrical wire, resistance = $\rho\frac{l}{\pi r^2}$, where $\rho$ is based on the conductivity of the medium.
-In our case, assuming $ r << l$, resistance(B) < resistance(C) < resistance(A).
-(c) A possible design for a cell phone antenna augmentation inside a cell phone case.
-\bjoern{skip the antenna since we haven't investigated this.
-If you have the capacitor and you're confident you know how to measure its charge, add it.
-The different conductors with different measurements are easy to do and worthwhile.
-Maybe combine these two with the helmholtz resonator into one smaller figure.}}
-\label{fig:electronics}
-\end{figure}
-
 \subsection{Passive Components}
@@
 Future design tools could help designers achieve desired electrical properties automatically.
-
-\valkyrie{do we want those figures of the helmholz thing and the servo redirection gate?}

### 2014-04-10 paper/exampleobjects.tex
many paper updates.  need to finish print of whistle and resistors, and take photos, then we are done????

@@
 \section{Example Objects}
-\tovi{it might be intersting to relate our examples to existing HCI papers - in order we have: Touche, Dynamically Changeable Physical Buttons/PneUI, Midas, Suaron.
-I guess the maze doesnt really relate to anything.}
-\tovi{UIST is someitmes cranky about providing enough details so other can reporduce the work. We might want to be more careful about providing specific part numbers of specifications.}
-\bjoern{subsection headings need to be updated in light of the simplification and label changes in the design space.}
 To evaluate and highlight our tool's capabilities, we fabricated a set of five prototype objects designed using \systemnamenospace.
-\tovi{might want to specify the exact printer models we used here, and say we used multipel printers to show that are approach is printer independant.}
+To show that our approach is printer-independent, we fabricated some on a Makerbot Replicator 2 and the rest on an Objet Connex 260.
+Each prototype object explores the expansion of an existing sensing or actuation technique to the surface of 3D printed objects.
 %All prototypes were fabricated in a single piece, unless described otherwise. \bjoern{2 of 5 will have been fabricated otherwise -  Radio, Neon sign - both in pieces.}
 
-\subsection{Touch-sensitive Toys (open, liquid, star, terminals)}
+\subsection{Touch-sensitive Toys (open, liquid, star, endpoint)}
 
-To show how conductive fluids can be used for touch sensing, we created a set of touch-sensitive toys and a companion app that identifies the objects as well as touched parts of the objects (inspired by \cite{Harrison-acoustic}).
+To expand Touch\'{e} \cite{Sato-touche} to 3D printed forms, we created a set of touch-sensitive toys and a companion app that identifies the objects as well as touched parts of the objects (inspired by \cite{Harrison-acoustic}).
 The objects --- a brain and boat, in our example --- can be set on a base, upon which a speaker announces ``brain'' or ``boat''.
 Touching the brain model's front protrusion yields the announcement ``olfactory bulb''.
-The distinct touch points on each object are connected by an interior star topology, and touch sensing is performed via a single conductive connection to the base, using swept-frequency capacitive sensing~\cite{Sato-touche} (see Figure \ref{fig:toys}).
+The distinct touch points on each object are connected by an interior star topology, and touch sensing is performed via a single conductive connection to the base, using swept-frequency capacitive sensing (see Figure \ref{fig:examples}a).
 %We built a smart base which can distinguish between the toys and also determine which toy is mounted:
@@
 The models we used for the toys were downloaded from the  Thingiverse website.
-We then used \systemnamenospace to create the desired internal pipes.
+We then used \systemname to create the desired internal pipes.
 All components were fabricated on a Makerbot, support-free.
@@
 This paint requires approximately 2 days to fully dry in this configuration (3mm diameter tubes, maximum tube length 10cm).
-Our smart base is powered by an Arduino Uno running open-source SFCS code\footnote{http://www.instructables.com/id/Touche-for-Arduino-Advanced-touch-sensing}.
-An advatnage of using this paint is that it also serves as an adhesive, so no soldering was required.
+Our smart base is powered by an Arduino Uno running open-source SFCS code\footnote{\url{http://www.instructables.com/id/Touche-for-Arduino-Advanced-touch-sensing}}.
+\valkyrie{removed sentence about solderless paint: all active components are affixed to a breadboard, and the only thing actually attached to the paint is a wire (which we plug into the breadboard).
+granted, we did not have to solder the wire, but it seems weird to mention that...?} %An advantage of using this paint is that it also serves as an adhesive, so no soldering was required.
 \tovi{confirm}
 
-\subsection{Breathing Bunny (semi-closed, gas, terminals)}
+\subsection{Breathing Bunny (semi-closed, gas, endpoint)}
 
-To demonstrate how gases can be used to both provide sensation to the user through openings and to deform a model internally, we created a rabbit with a pair of tubes that can simulate breathing (see Figure \ref{fig:breathe}).
+With inspiration from PneUIs \cite{Yao-pneui} for using gases to both provide sensation to the user through openings and deform a model internally, we created a rabbit with a pair of tubes that can simulate breathing (see Figure \ref{fig:examples}b).
 When the rabbit inhales through its nose, its abdomen rises, and as it exhales its abdomen falls.
-For this, we used a combination air/vacuum pump: one terminal creates a vacuum while the other creates positive pressure.
-The rabbit model was downloaded from the Thingiverse website.
-\tovi{confirm} We then used \systemnamenospace to add two pipes: one open pipe exiting at its mouth, and one semi-closed pipe capped at its abdomen.
+For this, we used a combination air/vacuum pump (ROB-10398): one terminal creates a vacuum while the other creates positive pressure.
+The rabbit model, the well-known Stanford Bunny\footnote{\url{http://en.wikipedia.org/wiki/Stanford_bunny}}, was downloaded from the Thingiverse website.
+We then used \systemname to add two pipes: one open pipe exiting at its mouth, and one semi-closed pipe capped at its abdomen.
 We connected one pipe to each of our pump's terminals, and using a programmable power supply we mimic a rabbit's breathing pattern.
-This example was printed on the Objet using a flexible material and its support material was flushed post-print.\tovi{give pump part number/details}
+This example was printed on the Objet using a flexible material and its support material was flushed post-print.
 
-\subsection{Custom Radio (open, threadable, terminals)}
-\tovi{stick with the strucutre - describe what it does, then describe modeling and fabrication (how it was built)}
-A custom radio shows how pipes can be used to integrate electronic components into the user-facing surface of an object.
+\subsection{Custom Radio (open, threadable, endpoint)}
+A custom radio (figure \ref{fig:examples}c) shows how pipes can be used to integrate electronic components into the user-facing surface of an object.
 The radio includes a power and volume dial, and LED which goes on when the radio is on, and a tuner dial.
-\tovi{confirm}.
-A Arduino Pro Micro microcontroller, Si4703 radio tuner, and  [x volt] power supply are embedded in the base of the radio.
-The sound is played through a single speaker.
+An Arduino Pro Micro microcontroller, Si4703 FM tuner, and 6V (4xAA) power supply are embedded in the base of the radio.
+The sound is played through a single .5 Watt speaker.
+This shows a Midas-like \cite{Savage-midas} possibility: multiple terminals all routed back to a shared endpoint (in this case, the Arduino).
 
-The radio was manually modelled in SolidWorks.
+The radio shell was modelled in SolidWorks.
 We included slots on the 3D model matching the shape of the electronic components that were to be embedded.
@@
 
-%This design with 8 pipes (3 per potentiometer, 1 for the LED, 1 for the speaker) showed the limitations of our incremental routing strategy: since we voxelize and remesh for each pipe, surface detail is incrementally lost.
-The problem can be addressed by increasing the voxel resolution.
-At 512x512x512 voxels, surface geometry is not notably degraded.
-However, a single cut operation takes around 3 minutes \bjoern{on what hardware?
-mention earlier}, compared to near-instant performance on a $128^3$ voxel grid.
-We fabricated the radio itself on our Makerbot, support-free but cut in two pieces to allow the Arduino and battery pack to be inserted.
-TOVI: Move to discussion if you want to keep this text.
+\subsection{Presence-aware Pen Holder (open, threadable, endpoint)}
+\bjoern{the model looks like the acutal pipes are really short and basically straight at the bottom of the model.
+This may raise the question if one couldn't just use the line sensors without any fiber optic cable in the middle.}  \valkyrie{I suspect either would work; we don't have to use this as a threadable example if we don't want to, since the radio and the EL wire are both threadable already}
+Our presence-aware pen holder can distinguish which tool or tools a user has picked up (see Figure \ref{fig:examples}d).
+Our pen holder uses a modification of the FlyEye technique described by Wimmer \cite{Wimmer-flyeye} in which an IR camera senses a user's grasp on an object, getting data from IR light transmitted on fiber optic cables.
+We marry this technique here with the Sauron technique \cite{Savage-sauron}: instead of using mirrors to redirect light and motion sensing as Sauron does, we use fiber optic cables.
 
-\bjoern{I ran out of steam somewhere here...}
-\tovi{relate this to sauron as well}
-\subsection{Presence-aware Pen Holder (open, solid, ternimals)}
-\bjoern{isn't this a threadable instead of a solid?} \bjoern{the model looks like the acutal pipes are really short and basically straight at the bottom of the model.
-This may raise the question if one couldn't just use the line sensors without any fiber optic cable in the middle.}
-Our presence-aware pen holder can distinguish which tool or tools a user has picked up (see Figure \ref{fig:pens}).
-Our pen holder uses a modification of the FlyEye technique described by Wimmer in \cite{Wimmer-flyeye} in which \bjoern{describe...}.
-
-The model was created/downloaded from [x].
-We used \systemnamenospace to add four internal open pipes connecting each pen chamber to the based on the pen holder.
-This prototype was built by our Objet and support was flushed post-print.\bjoern{Incorrect.
-Looks like a Makerbot print.} \valkyrie{this whole paragraph is unclear}
-
+The penholder body was created in Solidworks.
+We used \systemname to add four internal open pipes connecting each pen chamber to the based on the pen holder.
+This prototype was built by our Makerbot support-free.
 Each pipe was then threaded with a fiber optic cable.
-We use a single cable per pipe; our 6mm diameter fiber optic cables, in comparison to the fine tubes used in the original work , can send and receive through the same cable.
-\bjoern{is this because it has many strands?}  At the base of each tube is a QRE1113 line sensor digital breakout board, which has an integrated IR emitter and receiver.
+We use a single cable per pipe; our 6mm diameter fiber optic cables, in comparison to the fine tubes used in the original work, can send and receive through the same cable because their cross-sectional area is large enough to accommodate both functions.
+At the base of each tube is a QRE1113 line sensor digital breakout board, which has an integrated IR emitter and receiver.
 When a pen is in its appointed place, the emitted infrared light is reflected off its bottom and travels back to the receiver, where it registers as bright.
 
-\subsection{Animated Neon Sign (return, threadable, path)}
+\subsection{Maze (fully enclosed, particulate, tree, path)}
 
-\begin{figure}[h!]
+We created a maze game with a marble that can be navigated through a maze-like path, with exits on either side of the model (Figure \ref{fig:examples}e).
+
+The maze is based on an SVG file we manually created and processed with our internal path routing tool.
+We printed the maze on the Objet using a transparent material.
+The final artifact was fabricated in two halves fastened together via glue, to allow removal of support material with a high-pressure water tool.
+We plugged the ends of the tubes post-print with additional material to ensure the ball does not escape.
+
+\subsection{Animated Neon Sign (open, threadable, fully constrained)}
+
+\begin{figure}[t]
 \centering
@@
 \caption{A two-state animated neon sign designed using our tool.
-(a) shows the input SVG files, with the two states of the mouth drawn in different colors.
-\bjoern{it's hard to tell these colors apart!}  (b) shows the resultant mesh when we subtracted the generated tubes.
-\bjoern{hard to see what's going on here, too.} In (c), we show the fabricated sign.}
+(a) shows the input SVG files, with the eyes (non-animated) in black and the two states of the mouth in {\color{blue}blue} and {\color{gray}gray}.
+(b) shows the resultant mesh when we subtracted the generated tubes, with rear exit tubes circled in {\color{blue}blue}.
+In (c), we show the fabricated sign.}
 \label{fig:neon}
@@
 We designed a custom neon robot head which can be animated to ``talk''.
-\tovi{in the system seciton do we ever discuss how entry and exit points are specified for path-constrained pipes?}
-The geomertry of the model was imported into \systemnamenospace, which created the assocaited Pipes using its internal path routing tool.
-We then printed the model on the Objet using a transparent material.
-The model was printed in two halves, cutting through the plane of pipe path, to facilitate assembly.
+The external geometry of the model was imported into \systemnamenospace, which created the associated pipes using its internal path routing tool.
+We added additional endpoint-constrained tubes as exits out the rear of the model (see Figure \ref{fig:neon}B).
+We then printed the model on the Objet using transparent material.
+The model was printed in two halves, cutting through the plane of internal path, to facilitate assembly.
 
-The pipes were threaded with electroluminescent wire which is lit in sequence to create an animation using an EL wire sequencer. Pipes enter and exit through the rear to hide the EL wire controls.
-
-\subsection{Maze (fully enclosed, particulate, tree, terminals + path)}
-
-Our final example object is a maze game with a marble that can be navigated through a maze-like path, with exits on either side of the model.
-
-The maze is based on an SVG file we manually created and processed with our internal path routing tool.
-The Maze was printed on the Objet using a transparent material.
-We created the maze in two halves that were fastened together via glue, to support the removal of support materials using a high-pressure water tool.
-\bjoern{check - i think the latest maze is in fact done with our tool?
-Will need rewriting.}
+The pipes were threaded with electroluminescent wire which is lit in sequence, using an EL wire sequencer, to create the animation. Pipes enter and exit through the rear to hide the EL wire controls.

### 2014-04-10 paper/introduction.tex
many paper updates.  need to finish print of whistle and resistors, and take photos, then we are done????

@@
 And air can be pumped through pipes to produce haptic effects.
-Existing sensing and actuation approaches, such as FlyEye \cite{Wimmer-flyeye}, Swept Frequency Capacitive Sensing \cite{Sato-touche}, and Jamming User Interfaces \cite{Follmer-jamming}, can be enhanced by the use of pipes: makers utilizing such I/O strategies can locate input and output at arbitrary points on a printed 3D object's surface.
-\tovi{Is enhanced the right word?
-Or would "enabled" be better - since these techniques can now be applied to 3D printed objects} \valkyrie{I like enhanced; we are allowing them to be applied in new places}  We leverage these existing techniques for sensing and actuation while our work's novelty is in \emph{internal modeling} for 3D printing and the \emph{redirection} of I/O to arbitrary locations on a device.
+Existing sensing and actuation approaches, such as FlyEye \cite{Wimmer-flyeye}, Swept Frequency Capacitive Sensing \cite{Sato-touche}, and Jamming User Interfaces \cite{Follmer-jamming}, can be expanded to new contexts by the use of pipes: makers utilizing such I/O strategies can locate input and output at arbitrary points on a printed 3D object's surface.
+We leverage these existing techniques for sensing and actuation while our work's novelty is in \emph{internal modeling} for 3D printing and the \emph{redirection} of I/O to arbitrary locations on a device.
 

### 2014-04-10 paper/limitations.tex
many paper updates.  need to finish print of whistle and resistors, and take photos, then we are done????

@@
 
+This basic approach of using a discrete rod for physical simulation does have various limitations, in particular it
+cannot recover if the rod becomes tangled or unstable. 
+However we have observed these problems only when attempting to break
+the system; they did not occur in any of the cases shown in our figures.
+Another issue is parameters, as the penalty forces used in PBD do not
+correspond to physical units. We avoided serious issues here by uniformly
+scaling the initial mesh to a unit box, and tuning the parameters on several test cases.
+
 We are also eager to test our tool with users and get feedback from makers who use it in the wild.

### 2014-04-10 paper/relatedwork.tex
many paper updates.  need to finish print of whistle and resistors, and take photos, then we are done????

@@
 
-One of our path routing algorithms routes a single, continuous pipe through an object that follows the appearance of an input shape.
-The technique is inspired by work in generating continuous line illustrations from images~\cite{Wong-continuousline,Bosch-tsp}.
-Our approach is also based on graph theory (Euler tours), but differs in that it generates 3D geometry instead of 2D illustrations.\tovi{I think this paragraph could be omited and the citation inlcuded is hte system section}
-
 %Similar to our path routing algorithm, Wong presents a graph-based approach for generating continuous line illustrations from images \cite{Wong-continuousline}.

### 2014-04-10 paper/tubes.tex
many paper updates.  need to finish print of whistle and resistors, and take photos, then we are done????

@@
 \section{\systemname - A Tool for Pipe Design}
-\tovi{I understand why we use ``makers'' instead of ``users'', but it might make it sound like the tool requires a certain level of existing hacking skills.
-For example, we are also allowing researchers to design novel objects}
 To allow makers to design novel objects with pipe-powered interfaces, we created a tool, \systemnamenospace, to add pipe geometries to arbitrary 3D models.
@@
  \systemname allows users to design pipes in two ways.
-First, they can select exterior anchor points on their objects and internal connection points; the tool then creates complete point-to-point routings using A* path search and physical rod simulation (see Figure \ref{fig:tool-process-exterior}).
-Second, users can also import vector art describing desired interior paths (see Figure \ref{fig:tool-process-interior}).
+First, they can select exterior anchor points on their objects; the tool then creates complete point-to-point routings using A* path search and physical rod simulation (see Figure \ref{fig:tool-process-exterior}).
+Second, users can import vector art describing desired interior paths (see Figure \ref{fig:tool-process-interior}).
 \systemname then creates a single path that follows the input shape using edge graph manipulation and Euler tour generation.
@@
 The interior of these pipes should have as large a bending radius as possible so that post-print insertion of solid or viscous media is easier.
-\tovi{removed the first paragraph as it seemed repetitive.
-Do a better job explaining how this works from the user's perspective.
-What is the UI?
-I added a bit.
-I think it would be best to explain all UI in this paragraph, and just focus on the algortihms below}
 To create endpoint-cosntrained pipes, in which the location and shape of exterior connection points matters, a brush tool is used to specify a pair of points on the mesh's surface.
-The user then clicks the [x] button which casues the system to route a pipe between the specificed points, and update the mesh accordingly by subtracting the pipes geometry fromt the model.
- \bjoern{feels a bit redundant with the last paragraph in the prior section.}
+The system automatically routes the pipe, and displays a preview.
+Using sliders, the user can independently adjust the starting and ending radii of the pipe.
+This is useful for, e.g, our breathing bunny (Figure \ref{fig:examples}b), whose pipes must fit a piece of hardware on one end, but on the other end should have as large an interactive surface as possible.
 
@@
     \includegraphics[width=3.4in]{figures/exterior.png}
-\caption{An example mesh with exterior connection points selected.
-The smoothed A* routing is drawn in {\color{red}red}, and in {\color{tovi}green} is our physically-based rod after simulation.
-\bjoern{This figure is visually hard to understand.
-I think a step-by-step figure would be better - a) selecting two points.
-b) red a* path c) final routed pipe.
-}}
+\caption{In a), an example mesh with exterior connection points selected.
+The smoothed A* routing (b) is drawn in {\color{red}red}.
+In {\color{tovi}green} is our physically-based rod after simulation (c).}
 \label{fig:tool-process-exterior}
@@
 a straight configuration, while satisfying the user-selected endpoint position constraints.
-We also increase the length of the wire, to provide some slack in cases where 
-it would otherwise bend sharply to stay inside the shape.
 
@@
 modelled as penalty forces. To keep the rod inside the
-mesh, we compute a discretized distance field, we can then constrain the
-rod to stay within some offset shell, again using a simple penalty force.
-\tovi{might want to specify this penalty force - or just say here that the value of the parameters were tuned on several test cases, and provide the associated values.
-}
+mesh, we compute a discretized distance field; we can then constrain the
+rod to stay within our offset shell, again using a simple penalty force.  Our penalty force is 0.3, however this does not, unfortunately, correspond to any real-world units (see Limitations).
 
@@
 \bjoern{are there any guarantees that two paths are more than $pipe1{radius}+pipe2{radius}$ away from each other?
-if not, we can't really claim that we always avoid intersections - it just usually happens.}
-
-\tovi{move this paragraph to discussion section}
-This basic approach does have various limitations, in particular it
-cannot recover if the discrete rod becomes tangled or unstable. 
-However we have observed these problems only when attempting to break
-the system; they did not occur in any of the cases shown in our figures.
-Another issue is parameters, as the penalty forces used in PBD do not
-correspond to physical units. We avoided serious issues here by uniformly
-scaling the initial mesh to a unit box, and tuning the parameters on several test cases.
+if not, we can't really claim that we always avoid intersections - it just usually happens.}  \valkyrie{the pipe basically tries to pass through the midpoint in space between surfaces.
+It is possible to make the pipe go outside the surface, e.g., by putting it close to a corner and making the radius large (we don't re-adjust the pipe simulation as the radius changes, only as the endpoint constraints change).}
 
@@
 
-\subsubsection{Pipes with Multiple Endpoints}\tovi{need to describe the UI for this}
-To allow designers to create pipes with multiple endpoints (e.g., star topologies in Figure \ref{fig:toys}), we use one surface position per pipe and a central fixed endpoint that the user can interactively position (see Figure~\ref{fig:boat-star}.
-The user can connect several then route the pipes as normal.
+\subsubsection{Pipes with Multiple Endpoints}
+To allow designers to create pipes with multiple endpoints (e.g., star topologies in Figure \ref{fig:examples}a), we use one surface position per pipe and a central fixed endpoint that the user can interactively position (see Figure~\ref{fig:boat-star}).
+The user can connect several pipes, hitting a key to add each one in sequence and then positioning them with the mouse.
+The pipes route as normal.
 
@@
 For path-constrained pipes, the route of the pipe must conform to a desired shape.
-A maker can import a vector graphics file (such as SVG) describing the path she desires for her pipes.
-\tovi{explain the ui - how does the user specify the path geometry, execute the command to create the path, etc.}.
-We then cut these pipes from the mesh.
+A maker can import a vector graphics file (such as SVG) describing the path she desires for her pipes: all lines in this input file are interpreted as desired pipe components for the final artifact.
+The maker aligns her 3D model to the desired pipe cutout location.
+Using sliders, she can again change the pipes' radius.
+A preview is provided to show the radius and location of the cut.
 
-In some cases, such as for the neon sign in Figure \ref{fig:tool-process-interior}, a single path must passes through a set of disconnected path segments (i.e., all letters in the word ``UIST'' ).
+In some cases, such as for the neon sign in Figure \ref{fig:tool-process-interior}, a single path must pass through a set of disconnected path segments (i.e., all letters in the word ``UIST'' ).
 \systemname leverages graph theory algorithms to generate this single path, as described below.
+Without endpoint constraints, the path's exit(s) from the model must are defined by the SVG: the maker can see where the path exits the model and translate the cut according to her desired outcome.
 
@@
     \includegraphics[width=3.4in]{figures/interior.pdf}
-\caption{An input vector graphics file with the points which cannot be tubed as drawn highlighted in {\color{blue}blue} and the start and end points highlighted in {\color{gray}gray} (a).
-The connected graph created by our software (b) and the resulting Euler circuit (c) permit creation of a novel neon sign (d).
-\bjoern{replace sub-figure d?
-looks strange.}}
+\caption{An input vector graphics file with the start and end points highlighted in {\color{gray}gray} (a) and the points which cannot be tubed as drawn highlighted in {\color{blue}blue} (b).
+The connected graph created by our software (c) and the resulting Euler tour (d) permit insertion of media post-print.}
 \label{fig:tool-process-interior}
@@
 We find an Euler tour on the modified graph, and thicken the path to create pipes.
-
+Our technique for routing this path is inspired by work in generating continuous line illustrations from images~\cite{Wong-continuousline,Bosch-tsp}.
 The interior path routing problem is a version of the Chinese Postman Problem\footnote{The CPP is also known as the route inspection problem: \url{http://en.wikipedia.org/wiki/Route_inspection_problem}}, in which we wish to traverse all edges of the graph described in the user's input, much as a postman needs to walk along every road at least once to deliver mail.
@@
 In the final artifact, all created edges will be blocked out by dark material so the inserted medium is not visible (see Figure \ref{fig:tool-process-interior}).
-We offer a short algorithm here, with a more mathematically precise definition and associated proof in supplemental materials appendix.
+%We offer a short algorithm here, with a more mathematically precise definition and associated proof in supplemental materials appendix.
 
-{\bf Setup:}We first add a temporary edge that connects the user's desired start and end points, and create a full Eulerian graph.
+{\bf Setup:} We first add a temporary edge that connects the user's desired start and end points, and create a full Eulerian graph.
 This edge is removed at the end to yield a semi-Eulerian graph that can be threaded after printing.
@@
 Our EL wire in Figure \ref{fig:UIST} follows an Euler tour through the input nodes.
-%/george{too complicated} \bjoern{shortened this entire section}
-%Once we have a semi-Eulerian graph, we need to create an Euler tour.
-We use a weighted modification of Fleury's algorithm for finding Euler tours\footnote{\url{http://en.wikipedia.org/wiki/Eulerian_path\#Fleury.27s_algorithm}} for this.
-In the classic algorithm, each selected edge is removed to create a reduced graph.
-To select the next edge while at a node, the algorithm randomly chooses from all incident non-bridge edges (where a bridge edge would create disconnected non-trivial subgraphs if it were removed).
-This ensures that every edge is both reachable and traversed.
-In our modification, instead of randomly selecting from non-bridge paths at each node, we select the non-bridge edge which turns the least from the most recent path (i.e., we prefer to pass straight through a node, if possible).
-This minimizes turns in the final artifact, which eases support material removal and assembly.
-Our EL wire in Figure \ref{fig:UIST} follows an Euler tour through the input nodes.
 
@@
 \subsection{Mesh Modification}
-Once we have a set of authored paths, we must cut tubes out of the input mesh.
+After the maker hits "Accept" on her preview, we must cut pipes out of the input mesh.
 Our basic strategy is to create a polygonal tube by sweeping a profile polyline
@@
 in our examples, but clearly any other profile could be used.  Our interface
-allows different start and end radii, which are linearly interpolated along the path. \bjoern{why is this useful?} The radius is specified [how?]
+We allow different start and end radii, selected by the maker in the previous step, which are linearly interpolated along the path.
 
@@
 subtraction is trivial in such representations. We specifically use narrow-band
-level sets~\cite{Museth04}, which represent the surface as the zero iso-contour of an
+To avoid the complexities of direct mesh booleans, we construct a discretized volumetric (i.e., voxel-based) representation of the original mesh and the tube mesh.
+Boolean subtraction is trivial in such representations.
+We specifically use narrow-band level sets~\cite{Museth04}, which represent the surface as the zero iso-contour of an
 approximate distance field.
@@
 
-\bjoern{made this a top-level section since it's really not about the software.}
 \section{Fabrication Techniques}
@@
 The physical fabrication process used by different machines gives rise to differing strategies for avoiding support material or easing its removal.
-We will discuss techniques for fused deposition modeling (FDM) machines, suchas a Makerbot [ref] and multi-jetting machines such as an Objet [ref].
+We will discuss techniques for fused deposition modeling (FDM) machines, such as the Makerbot\footnote{\url{http://makerbot.com}}, and multi-jetting machines such as an Objet\footnote{\url{http://stratasys.com}}.
 
@@
 This means that overhangs without added support material can be approximately $45^{\circ}$, and although cylinders (e.g., pipes) and spheres violate this in some places they can be printed support-free.
-\bjoern{Add a sentence: The models in Figure 1a were produced on a Makerbot FDM without any support structures inside the pipes.}
+The models in \ref{fig:teaser}a were produced on a Makerbot FDM without any support structures inside the pipes.
 
@@
 In future work we would like to explore a tool in which a user can select surfaces (pipes) that should be prioritized for support avoidance.
-\tovi{print in parts and then assemble those parts together?}
 
@@
 Complex pipe geometries can make water blasting impossible and other processes more labor- and time-intensive.
+All models in \ref{fig:teaser}b required support material removal post-print: the neon sign was additionally cut into two pieces to give better access to the support.
 
@@
 An automatic tool for this task might be beneficial, however it is outside the scope of this work.
-\bjoern{again, add a sentence about what we did for our objet models.}

### 2014-04-11 paper/abstract.tex
paper mod start
grammar check

@@
 3D printers offer extraordinary flexibilty for prototyping the shape and mechanical function of objects.
-We investigate how 3D models can be modified to facilitate the creation of interactive objects that offer dynamic input and output.
+We investigate how 3D models can be modified to facilitate the creation of \emph{interactive} objects that offer dynamic input and output.
 %However, a 3D printer's role in prototyping interactive objects is still not clear.  
-We introduce a general technique for supporting the rapid prototyping of interactivity by removing interior material of 3D models to form pipes and chambers.
-We describe this new design space of pipes and hollow chambers for interaction design, where variables include openings, topologies, design focus, and inserted media.
-We then present \systemname, a tool for routing such pipes through the interior of 3D models, integrated within a 3D modeling program.
+We introduce a general technique for supporting the rapid prototyping of interactivity by removing interior material from 3D models to form pipes and chambers.
+We describe this new design space of pipes and hollow chambers for interaction design, where variables include openings, topologies, path constraints, and inserted media.
+We then present \systemnamenospace, a tool for routing such pipes through the interior of 3D models, integrated within a 3D modeling program.
 We use two distinct routing algorithms.

### 2014-04-11 paper/conclusion.tex
grammar check

@@
 This amplifies and extends the utility of existing sensing and actuation approaches.
-In addition to describing the design space of pipes, we also presented a design tool for the creation of tubes inside arbitrary 3D models and discuss many techniques that combine nicely wtih pipes.
-We fabricated several example objects using this tool to explore new points in the design space.
+In addition to describing the design space of pipes, we also presented \systemname, a design tool for the creation of tubes inside arbitrary 3D models, and discussed many techniques that combine nicely wtih pipes.
+We fabricated several example objects using our tool to explore new points in the design space.
 

### 2014-04-11 paper/designspace.tex
grammar check

@@
 Our review of prior art indicates the potential benifits of using interior pipes for adding interactivity to 3D printed objects, although no general purpose tools currently exist for this purpose.
-To guide the design of such tool, we now provide a deisgn space of the types of pipes that a maker may want to utilize.
+To guide the design of such tool, we now provide a design space of the types of pipes that a maker may want to utilize.
 We have identified four relevant dimensions of pipe design: \emph{openings}, \emph{path constraints}, \emph{topologies}, and inserted \emph{media}.
-We describe the space below.
+We describe the space (see Figure \ref{fig:pipespace}) below.
 
@@
     \includegraphics[width=1.0\columnwidth]{figures/tubespace.pdf}
-\caption{The design space of pipes.  \emph{Pipe openings}, \emph{path constraints}, \emph{topologies}, and inserted \emph{media} give rise to unique input/output possibilities.}
+\caption{The design space of pipes.  Pipe \emph{openings}, \emph{path constraints}, \emph{topologies}, and inserted \emph{media} give rise to unique input/output possibilities.}
 \label{fig:pipespace}
@@
 \emph{Open} pipes may be used to create electronic circuitry.
-For example, an open pipe filled with conductive paint and connected to a battery can provide power to an LED (see Figure \ref{fig:openings}A).
+For example, an open pipe filled with conductive paint and connected to a battery can provide power to an LED (see Figure \ref{fig:openings}a).
 %
 \emph{Semi-enclosed} pipes are capped on one side and could be used to create create tactile output.
-By fabricating a cap from a malleable material and attaching a pump to the open ended, haptic feedback can be generated by raising and depressing the flexible cap (see Figure \ref{fig:openings}C).
+By fabricating a cap from a malleable material and attaching a pump to the open end, haptic feedback can be generated by raising and depressing the flexible cap (see Figure \ref{fig:openings}c).
 %
 \emph{Fully enclosed} pipes (cavities) are entirely internal to the object.
-Fully enclosed pipes may be used as weight modifiers~\cite{Prevost-makeitstand} and for object identification~\cite{Willis-infrastructs}, as in Figure \ref{fig:openings}D.
+Fully enclosed pipes may be used as weight modifiers~\cite{Prevost-makeitstand} and for object identification~\cite{Willis-infrastructs}, as in Figure \ref{fig:openings}d.
 If printed in parts and assembled post-print, enclosed pipes could also contain water or particles that would otherwise fall out.
@@
 Depending on the purpose of the pipe, its geometry may be constrained to specific locations within the 3D model.
-An \emph{endpoint-constrained} pipe requires specific start and end points, but the specific route of the internal path is irrelevant.
-This would be the case for circuit schematics, where the electronic components have a fixed location, but many routing solutions are valid.
-In contrast, with a \emph{path-constrained} pipe the shape of the path is important.
-For example, threading electroluminescent (EL) wire through a clear pipe enables makers to emulate neon art (see Figure \ref{fig:openings}B).
+An \emph{endpoint-constrained} pipe requires specific start and end points, but the particular route of the internal path is irrelevant.
+This is the case for circuit schematics, where the electronic components have a fixed location, but many routing solutions are valid.
+In contrast, for a \emph{path-constrained} pipe the shape of the path is important.
+For example, threading electroluminescent (EL) wire through a clear pipe enables makers to emulate neon art (see Figure \ref{fig:openings}b).
 A \emph{fully constrained} pipe requires both specific terminals as well as a specific path.
-If a maker wishes the EL wire to exit the rear of the sign to hide the electronics, it would require a fully-constrained path (Figure \ref{fig:neon}).
+If a maker wishes the EL wire to exit the rear of the sign to hide the electronics, it requires a fully-constrained path (Figure \ref{fig:neon}).
 Finally, an \emph{unconstrained} pipe does not need a specific path or terminal points.
-Using a cavity to modify an object's weight is an example of an unconstrained pipe.
+A cavity to modify an object's weight is an example of an unconstrained pipe.
 
@@
 Pipe network topologies enable \emph{splitting} or \emph{mixing} of media (see Figure \ref{fig:pipespace}).
-\emph{Star} and \emph{tree} topologies extend the splitting and mixing primitives, encompassing multiple-in and multiple-out configurations; we use star topologies to create different conductive paths in touch-sensitive toys, with a single terminal for swept-frequency capacitive sensing (Figure~\ref{fig:examples}a).
+\emph{Star} and \emph{tree} topologies extend the splitting and mixing primitives, encompassing multiple-in and multiple-out configurations; we use star topologies to create different sensing location for our touch-sensitive toys, with one terminal used for swept-frequency capacitive sensing (Figure~\ref{fig:examples}a).
 
@@
 
-In the next section we how different pipes within this design space space can be created with the help of our new design tool, \systemnamenospace.
+In the next section detail we how different pipes within this design space space can be created with the help of our new design tool, \systemnamenospace.

### 2014-04-11 paper/discussion.tex
grammar check

@@
 The promise of 3D printing is two-fold: it enables the manufacturing of complex objects at low volumes, and it allows for the manufacturing of geometries that are impossible to create using traditional techniques like injection molding.
-Printing pipes to prototype interactive objects exploit both advantages.
+Printing pipes to prototype interactive objects exploits both advantages.
 Prototypes by definition are low-volume parts, and pipes require complex interior geometries that are very difficult or impossible to produce using other processes.
-Our work has focused on routing pipes that are then later manually filled after a print has completed.
+Our work has focused on routing pipes that are manually filled after a print has completed.
 However, the fundamental operations of 3D path routing and topology specification that our tool offers will remain relevant as 3D printing technology advances.
@@
 As seen in Touch \& Activate \cite{Ono-touchandactivate}, different objects have different acoustic signatures.
-As such, two objects that are visually identical but have differing internal chamber geometry on their interior may be distinguished acoustically.
+As such, two objects that are visually identical but have differing internal chamber geometry may be distinguished acoustically.
 Thus we can recall an object's identity once its acoustic signature has been recorded.
@@
 Printing or injecting conductive material can also be used to design passive components with particular electrical properties.
-For example, we printed a proof-of-concept  capacitor which features a thin printed plate with hollow cones on either side (see Figure \ref{fig:electronics}A).
+For example, we printed a proof-of-concept  capacitor which features a thin printed plate with hollow cones on either side (see Figure \ref{fig:electronics}a).
 After printing, we coated the inner surfaces of capacitor with copper paint.
-It now can hold a charge of xx.
-Similarly, tubes can be used to create \emph{resistors} in ``printed'' circuitry: a longer or thinner tube has a greater resistance than a shorter or wider one (see Figure \ref{fig:electronics}B).
-Internal \emph{antenna} design is also possible, which could lead to new flexibility in, e.g., cell phone design (see Figure \ref{fig:electronics}C).
+It now can hold a charge of \valkyrie{xx}.
+Similarly, tubes can be used to create \emph{resistors} in ``printed'' circuitry: a longer or thinner tube has a greater resistance than a shorter or wider one (see Figure \ref{fig:electronics}b).
+Internal \emph{antenna} design is also possible, which could lead to new flexibility in, e.g., cell phone design (see Figure \ref{fig:electronics}c).
 Future design tools could help designers achieve desired electrical properties automatically.

### 2014-04-11 paper/exampleobjects.tex
grammar check

@@
 
-\section{Example Objects}
+\section{Totally Tubular Example Objects}
 To evaluate and highlight our tool's capabilities, we fabricated a set of five prototype objects designed using \systemnamenospace.
@@
 To expand Touch\'{e} \cite{Sato-touche} to 3D printed forms, we created a set of touch-sensitive toys and a companion app that identifies the objects as well as touched parts of the objects (inspired by \cite{Harrison-acoustic}).
-The objects --- a brain and boat, in our example --- can be set on a base, upon which a speaker announces ``brain'' or ``boat''.
+The objects --- a brain and boat, in our example --- can be set on a base, upon which the computer displays ``brain'' or ``boat''.
 Touching the brain model's front protrusion yields the announcement ``olfactory bulb''.
@@
 Our smart base is powered by an Arduino Uno running open-source SFCS code\footnote{\url{http://www.instructables.com/id/Touche-for-Arduino-Advanced-touch-sensing}}.
-\valkyrie{removed sentence about solderless paint: all active components are affixed to a breadboard, and the only thing actually attached to the paint is a wire (which we plug into the breadboard).
-granted, we did not have to solder the wire, but it seems weird to mention that...?} %An advantage of using this paint is that it also serves as an adhesive, so no soldering was required.
-\tovi{confirm}
 
@@
 For this, we used a combination air/vacuum pump (ROB-10398): one terminal creates a vacuum while the other creates positive pressure.
-The rabbit model, the well-known Stanford Bunny\footnote{\url{http://en.wikipedia.org/wiki/Stanford_bunny}}, was downloaded from the Thingiverse website.
+The rabbit model, the well-known Stanford Bunny\footnote{\url{http://en.wikipedia.org/wiki/Stanford_bunny}}, was downloaded from Thingiverse.
 We then used \systemname to add two pipes: one open pipe exiting at its mouth, and one semi-closed pipe capped at its abdomen.
@@
 We included slots on the 3D model matching the shape of the electronic components that were to be embedded.
-The model was then imported into \systemnamenospace, where we created a network of 8 individual open pipes (3 per potentiometer, 1 for the LED, 1 for the speaker) to connect the the comments to the base of the radio.
+The model was then imported into \systemnamenospace, where we created a network of 8 individual open pipes (3 per potentiometer, 1 for the LED, 1 for the speaker) to connect the components to the base of the radio.
 We fabricated the radio itself on our Makerbot, support-free, but cut in two pieces to allow the Arduino and battery pack to be inserted.
-We then threaded the pipes with wires to connect the compoenents to the microcontroller at the base of the radio.
+We then threaded the pipes with wires to connect the components to the microcontroller at the base of the radio.
 
 \subsection{Presence-aware Pen Holder (open, threadable, endpoint)}
-\bjoern{the model looks like the acutal pipes are really short and basically straight at the bottom of the model.
-This may raise the question if one couldn't just use the line sensors without any fiber optic cable in the middle.}  \valkyrie{I suspect either would work; we don't have to use this as a threadable example if we don't want to, since the radio and the EL wire are both threadable already}
 Our presence-aware pen holder can distinguish which tool or tools a user has picked up (see Figure \ref{fig:examples}d).
@@
 (a) shows the input SVG files, with the eyes (non-animated) in black and the two states of the mouth in {\color{blue}blue} and {\color{gray}gray}.
-(b) shows the resultant mesh when we subtracted the generated tubes, with rear exit tubes circled in {\color{blue}blue}.
+(b) shows the resultant mesh when we subtracted the generated tubes, with rear exit tubes circled in {\color{blue}blue} \valkyrie{fix this with boolean magic}.
 In (c), we show the fabricated sign.}
@@
 The external geometry of the model was imported into \systemnamenospace, which created the associated pipes using its internal path routing tool.
-We added additional endpoint-constrained tubes as exits out the rear of the model (see Figure \ref{fig:neon}B).
+We added additional endpoint-constrained tubes as exits out the rear of the model to allow us to hide the controls (see Figure \ref{fig:neon}B).
 We then printed the model on the Objet using transparent material.
@@
 
-The pipes were threaded with electroluminescent wire which is lit in sequence, using an EL wire sequencer, to create the animation. Pipes enter and exit through the rear to hide the EL wire controls.
+The pipes are threaded with electroluminescent wire which is lit in sequence, using an EL wire sequencer, to create the animation.

### 2014-04-11 paper/introduction.tex
grammar check

@@
 Makers, researchers, and professional designers (jointly "makers" for the remainder of this paper) increasingly leverage 3D printers as tools for design work.
-A wide array of objects, ranging from toy figurines to video game controllers and jet engine blades, are now prototyped (and some times even manufactured) using these machines.
+A wide array of objects, ranging from toy figurines to video game controllers and jet engine blades, are now prototyped (and sometimes even manufactured) using these machines.
 Most models fabricated on 3D printers today are passive: they express the form, but not the function, of an object.
@@
 
-We describe a new design space of pipes and hollow chambers for the purpose of interaction design, where variables include openings, topologies, and inserted media.
+We describe a new design space of pipes and hollow chambers for the purpose of interaction design, where variables include openings, path constraints, topologies, and inserted media.
 By exploring this design space new opportunities for adding interactivity to 3D printed objects can be realized.
@@
 Existing sensing and actuation approaches, such as FlyEye \cite{Wimmer-flyeye}, Swept Frequency Capacitive Sensing \cite{Sato-touche}, and Jamming User Interfaces \cite{Follmer-jamming}, can be expanded to new contexts by the use of pipes: makers utilizing such I/O strategies can locate input and output at arbitrary points on a printed 3D object's surface.
-We leverage these existing techniques for sensing and actuation while our work's novelty is in \emph{internal modeling} for 3D printing and the \emph{redirection} of I/O to arbitrary locations on a device.
+We leverage these existing techniques for sensing and actuation while our work's novelty is in \emph{internal modeling} for 3D printing and the \emph{redirection} of I/O to or through arbitrary locations on a device.
 
@@
 Pipes should not pass too close to the surface: consumer-grade 3D printers do not guarantee air- or water-tightness of prints, so more than one printed layer may be required to prevent leakage.
-Pipes for applications like illumunated EL wire signs must allow a single pipe to follow a specific path layout (e.g., spelling out text).
+Pipes for applications like illumunated EL wire signs must allow a single pipe to follow a specific path layout (e.g., for spelling out text).
 
 We contribute PipeDream, an interactive design tool that offers 3D point-to-point routing of internal pipes based on a physical simulation of flexible rods.
-Our system addresses the various challenges that pipe design makers are likely to encounter, such as avoiding intersections and ensuring maximum bend angles.
-We then present a set of example objects, fabricated with the help of our modeling tool, that demonstrate different interaction modalities.
+Our system addresses the various challenges that makers are likely to encounter, such as avoiding intersections and ensuring maximum bend angles.
+We then present a set of example objects, designed with the help of our modeling tool, that demonstrate different interaction modalities.
 We conclude by discussing limitations and outlining areas for future exploration.

### 2014-04-11 paper/limitations.tex
grammar check

--- 2014-04-09 paper/limitations.tex
@@
 While pipes provide a powerful mechanism for directing interaction to arbitrary locations on an object's surface, they do have shortcomings which warrant discussion.
-In paticular, they cannot replace the active elements required for a functioning interactive device: they simply redirect sensing to different locations on an object's surface.
+In particular, they cannot replace the active elements required for a functioning interactive device: they simply redirect sensing to different locations on an object's surface.
 All pipes need to, at some point, connect to a sensor, microcontroller, or pump.
@@
 Liquids kept in the tubes can also cause discoloration due to seepage.
-Furthermore, not all printers guarantee watertightness of their prints, so leaking of water can occur.
+Furthermore, not all printers guarantee watertightness of their prints, so leakage can occur.
 
@@
 
-There are also area for improvement and future work with our software tool.
-he design of PipeDreams was based on our prposed design space of pipes.
+There are also areas for improvement and future work with our software tool.
+The design of \systemname was based on our proposed design space of pipes.
 However, the design tool is not yet fully capable of realizing all possibilities in this design space.
-In paticaulr, we did not specificalyl explore the mixing or splitting pipes.
-Furthermore, since pipes are committed to the mesh one at a time, our tool does not currently optimize a network of multiple pipes.
+Specifically, we did not explicitly explore the mixing or splitting of pipes, and our path-constrained cuts are strictly 2D for now.
+A 2.5D SVG import using, e.g., line colors as height would be straightforward, and our described algorithm would continue to work properly.
+Furthermore, since pipes are committed to the mesh one at a time, our tool does not currently optimize networks of multiple pipes.
 Instead, we utilize a greedy algorithm and route a single pipe at a time.
@@
 
+The basic approach of using a discrete rod for physical simulation does have various limitations, in particular it
+cannot recover if the rod becomes tangled or unstable. 
+However we have observed these problems only when attempting to break
+the system; they did not occur in any of the cases shown in our figures.
+Another issue is parameters, as the penalty forces used in PBD do not
+correspond to physical units. We avoided serious issues by uniformly
+scaling the initial mesh to a unit box, and tuning the parameters on several test cases.
+
 We are also eager to test our tool with users and get feedback from makers who use it in the wild.
-However, based on our own experiences, we believe our system could be readily utlized by makers and researchers, and serve as a useful tool for rapdily proptotyping interactive 3D objects.
+However, based on our own experiences, we believe our system could be readily utilized by makers and researchers, and serve as a useful tool for rapdily proptotyping interactive 3D objects.

### 2014-04-11 paper/relatedwork.tex
grammar check

@@
 \section{Related Work}
-Our contributions relate to other efforts aiming to add interactivity to digitally fabricated objects; and to techniques for routing and model modification.
+Our contributions relate to other efforts aiming to add interactivity to digitally fabricated objects and to techniques for routing and model modification.
 
@@
 Previous work has begun to investigate how to fabricate interactive objects using particular sensing and display technologies.
-Printed Optics \cite{Willis-printedoptics} uses light pipes to create integrated displays and sensed input optically through components added at print time; while Slyper et al., fabricate flexible robot armatures capable of sensing interaction via changes in air pressure \cite{Slyper-pressure}.
+Printed Optics \cite{Willis-printedoptics} uses light pipes to create integrated displays and senses input optically through components added at print time, while Slyper et al., fabricate flexible robot armatures capable of sensing interaction via changes in air pressure \cite{Slyper-pressure}.
 Our work is inspired by these projects, and contributes a general design tool for modelling objects that can have such capabilities.
 
-Sauron tracks interactions with printed physical mechanisms by placing a camera inside hollow objects, where all components have to be visible to a camera~\cite{Savage-sauron}.
+Sauron tracks interactions with printed physical mechanisms by placing a camera inside hollow objects, where all components must be visible to the camera~\cite{Savage-sauron}.
 Our technique instead can work with general, solid models and only carves out pipes for interaction where needed.
 Midas generates fabrication files for custom capacitive touch sensors~\cite{Savage-midas}.
-Their technique, is limited to planar and unfoldable shapes as it routes connections on the surface of objects.
+Their technique is limited to planar and unfoldable shapes as it routes connections on the surface of objects.
 \systemname can also be used for touch sensing, but routes conductors through the inside of objects.
@@
 In Make it Stand, material is removed from the interior of objects with the goal of changing their center of gravity \cite{Prevost-makeitstand}.
-Infrastructs encodes information using interior cavities that can be sensed using teraherz imaging~\cite{Willis-infrastructs}.
+Infrastructs encodes information using interior cavities that can be sensed using terahertz imaging~\cite{Willis-infrastructs}.
 Bickel, et al., explored enforcing deformation behavior with multi-material blending inside objects\cite{Bickel-deformation}.
-Modeling internal chambers is one aspect of our work, however our focus is on interaction rather than balance, identification or deformation.
+Modeling internal chambers is one aspect of our work, however our focus is on interaction rather than balance, identification, or deformation.
 %B\"{a}cher, et al., built an algorithm for automatically placing 3D fabricatable joints into meshes \cite{Bacher-articulated}.

### 2014-04-11 paper/tubes.tex
grammar check

@@
 Pipes, in general, need to avoid objects' surfaces to prevent fluid leakage when printed by hobbyist-grade machines.
-Finally, for designing path-constrained pipes, that follow a user-specified path, we must ensure certain characteristics of that path (e.g., that the path is connected).
+Finally, for designing path-constrained pipes that follow a user-specified path, we must ensure certain characteristics of that path (e.g., that it is connected).
 %\george{pipes are hard.  not sure we convinced readers by now that they are worth the effort.  maybe this has to happen more in the intro.  the example objects section is still a few pages away.}
@@
 \systemname allows users to design pipes in two ways.
-First, they can select exterior anchor points on their objects; the tool then creates complete point-to-point routings using A* path search and physical rod simulation (see Figure \ref{fig:tool-process-exterior}).
-Second, users can import vector art describing desired interior paths (see Figure \ref{fig:tool-process-interior}).
+In one, they can select exterior anchor points on their objects; the tool then creates complete point-to-point routings using A* path search and physical rod simulation (see Figure \ref{fig:tool-process-exterior}).
+In the other, users can import vector art describing desired interior paths (see Figure \ref{fig:tool-process-interior}).
 \systemname then creates a single path that follows the input shape using edge graph manipulation and Euler tour generation.
@@
 The interior of these pipes should have as large a bending radius as possible so that post-print insertion of solid or viscous media is easier.
-To create endpoint-cosntrained pipes, in which the location and shape of exterior connection points matters, a brush tool is used to specify a pair of points on the mesh's surface.
+To create endpoint-constrained pipes, in which the location and shape of exterior connection points matters, a brush tool is used to specify a pair of points on the mesh's surface.
 The system automatically routes the pipe, and displays a preview.
@@
 A* finds a least-cost path through a graph of nodes given a cost function.
-Our graph comprises the grid of voxels as nodes, with edges between adjacent nodes; our path cost is the based on the Euclidean distance to the end point.
+Our graph comprises the grid of voxels as nodes, with edges between adjacent nodes; our path cost is the based on the Euclidean distance to the selected end point.
 In this formulation, the routed path is naturally constrained to stay within the object boundaries.
@@
 We solve this problem via physical simulation. We create a straight virtual 
-wire (a 3D poly-line or \emph{discrete rod}), and then set its initial position 
+wire (a 3D polyline or \emph{discrete rod}), and then set its initial position 
 as our A* path.
@@
 a straight configuration, while satisfying the user-selected endpoint position constraints.
+In addition to user constraints, we require that the rod be perpendicular to the surface at its ends to prevent the terminal \emph{shape} from deforming.
 
@@
 mesh, we compute a discretized distance field; we can then constrain the
-rod to stay within our offset shell, again using a simple penalty force.  Our penalty force is 0.3, however this does not, unfortunately, correspond to any real-world units (see Limitations).
+rod to stay within our offset shell, again using a simple penalty force.
 
@@
 The simulation takes about a second for most of our examples.
-\bjoern{are there any guarantees that two paths are more than $pipe1{radius}+pipe2{radius}$ away from each other?
-if not, we can't really claim that we always avoid intersections - it just usually happens.}  \valkyrie{the pipe basically tries to pass through the midpoint in space between surfaces.
-It is possible to make the pipe go outside the surface, e.g., by putting it close to a corner and making the radius large (we don't re-adjust the pipe simulation as the radius changes, only as the endpoint constraints change).}
 
@@
 \subsubsection{Pipes with Multiple Endpoints}
-To allow designers to create pipes with multiple endpoints (e.g., star topologies in Figure \ref{fig:examples}a), we use one surface position per pipe and a central fixed endpoint that the user can interactively position (see Figure~\ref{fig:boat-star}).
+To allow designers to create pipes with multiple endpoints (e.g., the star topologies in Figure \ref{fig:examples}a), we use one surface position per pipe and a central fixed endpoint that the user can interactively position (see Figure~\ref{fig:boat-star}).
 The user can connect several pipes, hitting a key to add each one in sequence and then positioning them with the mouse.
@@
 For path-constrained pipes, the route of the pipe must conform to a desired shape.
-A maker can import a vector graphics file (such as SVG) describing the path she desires for her pipes: all lines in this input file are interpreted as desired pipe components for the final artifact.
-The maker aligns her 3D model to the desired pipe cutout location.
-Using sliders, she can again change the pipes' radius.
-A preview is provided to show the radius and location of the cut.
+A maker can import a vector graphics file (such as SVG) describing the path she wants for her pipes: all lines in this input file are interpreted as desired pipe components for the final artifact.
+Using sliders, she can change the pipes' radius.
+A preview is provided to show the radius and location of the cut, and the maker can align her model as appropriate.
 
@@
 \systemname leverages graph theory algorithms to generate this single path, as described below.
-Without endpoint constraints, the path's exit(s) from the model must are defined by the SVG: the maker can see where the path exits the model and translate the cut according to her desired outcome.
+Without endpoint constraints, the path's exit(s) from the model are defined by the SVG: the maker can see where the path exits the model and translate the cut according to her desired outcome.
 
@@
 
-{\bf Euler Tour:} To find an Euler tour, we use Fleury's algorithm\footnote{\url{http://en.wikipedia.org/wiki/Eulerian_path\#Fleury.27s_algorithm}}, with a modification to preserve continuity: at a given node, out of multiple candidate edges, we prefer the edges with the smallest angular deviation from the incoming edge (i.e., we prefer to pass straight through a node, if possible).
+{\bf Euler Tour:} To find an Euler tour, we use Fleury's algorithm\footnote{\url{http://en.wikipedia.org/wiki/Eulerian_path\#Fleury.27s_algorithm}}, with a modification to preserve continuity: at a given node, from multiple candidate edges, we prefer the edge(s) with the smallest angular deviation from the incoming edge (i.e., we prefer to pass straight through a node, if possible).
 This minimizes turns in the final artifact, which eases support material removal and assembly.
@@
 \subsection{Mesh Modification}
-After the maker hits "Accept" on her preview, we must cut pipes out of the input mesh.
-Our basic strategy is to create a polygonal tube by sweeping a profile polyline
+After the maker hits ``Accept'' on her preview, we must cut pipes out of the input mesh.
+At this stage, we verify whether the new pipe will intersect the surface or an existing pipe, and if so display a warning message suggesting that she should consider a smaller radius.
+Our basic strategy for the cut is to create a polygonal tube by sweeping a profile polyline
 along the path, and subtract the tube from the mesh. We only use circular profiles
-in our examples, but clearly any other profile could be used.  Our interface
+in our examples, but clearly any other profile could be used. 
 We allow different start and end radii, selected by the maker in the previous step, which are linearly interpolated along the path.
 
-To avoid the complexities of direct mesh booleans, we construct a discretized volumetric
-(i.e., voxel-based) representation of the original mesh and the tube mesh. Boolean
-subtraction is trivial in such representations. We specifically use narrow-band
 To avoid the complexities of direct mesh booleans, we construct a discretized volumetric (i.e., voxel-based) representation of the original mesh and the tube mesh.
@@
 %When geometries contain overhangs, as there may be in designed pipes and cavities, 
-3D printers lay down support material that must be later removed from the model to allow overhangs.
+To allow overhangs, 3D printers lay down support material that must be later removed from the model.
 This removal process can be time-intensive and challenging.
@@
 This means that overhangs without added support material can be approximately $45^{\circ}$, and although cylinders (e.g., pipes) and spheres violate this in some places they can be printed support-free.
-The models in \ref{fig:teaser}a were produced on a Makerbot FDM without any support structures inside the pipes.
+The models in Figure \ref{fig:teaser}a were produced on a Makerbot FDM without any support structures inside the pipes.
 
-On single material hobbyist FDM printers, support structures are built using the same material as the model.
+On single-material hobbyist FDM printers, support structures are built using the same material as the model.
 Such support may be impossible to remove from internal cavities.
@@
 Complex pipe geometries can make water blasting impossible and other processes more labor- and time-intensive.
-All models in \ref{fig:teaser}b required support material removal post-print: the neon sign was additionally cut into two pieces to give better access to the support.
+All models in Figure \ref{fig:teaser}b required support material removal post-print: the neon sign was additionally cut into two pieces to give better access to the support.
 

### 2014-04-11 paper/discussion.tex
paper to send to gord

@@
 \emph{Aural} outputs can be created with gases.
-Pipes and cavities can act as resonators can be leveraged to create 3D-printed sound chambers with explicitly designed sound characteristics~\cite{Zoran-flute} (see Figure \ref{fig:ocarina}).
+Pipes and cavities can act as resonators can be leveraged to create 3D-printed sound chambers with explicitly designed sound characteristics~\cite{Zoran-flute} (see Figure \ref{fig:speculative}a).
 %All that is necessary for this is an enclosed chamber with a pipe connected to it via a narrow neck-point (see Figure \ref{fig:ocarina}).
@@
 \centering
-    \includegraphics[width=3.4in]{figures/placeholder/helmholz.jpg}
+    \includegraphics[width=3.4in]{figures/speculative.png}
 \caption{Some possible pipe uses.
@@
 For a cylindrical wire, resistance = $\rho\frac{l}{\pi r^2}$, where $\rho$ is based on the conductivity of the medium.
-In our case, because $r << l$, $resistance(B) < resistance(C) < resistance(A)$.}
-\label{fig:ocarina}
+In our case, because $r << l$, $resistance(2) < resistance(3) < resistance(1)$.}
+\label{fig:speculative}
 \end{figure}
@@
 Printing or injecting conductive material can also be used to design passive components with particular electrical properties.
-For example, we printed a proof-of-concept  capacitor which features a thin printed plate with hollow cones on either side (see Figure \ref{fig:electronics}a).
+For example, we printed a proof-of-concept  capacitor which features a thin printed plate with hollow cones on either side (see Figure \ref{fig:speculative}b).
 After printing, we coated the inner surfaces of capacitor with copper paint.
 It now can hold a charge of \valkyrie{xx}.
-Similarly, tubes can be used to create \emph{resistors} in ``printed'' circuitry: a longer or thinner tube has a greater resistance than a shorter or wider one (see Figure \ref{fig:electronics}b).
-Internal \emph{antenna} design is also possible, which could lead to new flexibility in, e.g., cell phone design (see Figure \ref{fig:electronics}c).
+Similarly, tubes can be used to create \emph{resistors} in ``printed'' circuitry: a longer or thinner tube has a greater resistance than a shorter or wider one (see Figure \ref{fig:speculative}c).
+Internal \emph{antenna} design is also possible, which could lead to new flexibility in, e.g., cell phone design.
 Future design tools could help designers achieve desired electrical properties automatically.

### 2014-04-12 paper/designspace.tex
bjoern's pass

@@
 \emph{Semi-enclosed} pipes are capped on one side and could be used to create create tactile output.
-By fabricating a cap from a malleable material and attaching a pump to the open end, haptic feedback can be generated by raising and depressing the flexible cap (see Figure \ref{fig:openings}c).
+By fabricating a cap from a malleable material and attaching a pump to the open end, haptic feedback can be generated by raising and depressing the flexible cap (see Figure \ref{fig:openings}b).
 %
 \emph{Fully enclosed} pipes (cavities) are entirely internal to the object.
-Fully enclosed pipes may be used as weight modifiers~\cite{Prevost-makeitstand} and for object identification~\cite{Willis-infrastructs}, as in Figure \ref{fig:openings}d.
+Fully enclosed pipes may be used as weight modifiers~\cite{Prevost-makeitstand} and for object identification~\cite{Willis-infrastructs}, as in Figure \ref{fig:openings}c.
 If printed in parts and assembled post-print, enclosed pipes could also contain water or particles that would otherwise fall out.
@@
 Depending on the purpose of the pipe, its geometry may be constrained to specific locations within the 3D model.
-An \emph{endpoint-constrained} pipe requires specific start and end points, but the particular route of the internal path is irrelevant.
+An \emph{endpoint-constrained} pipe requires specific start and end points, but the particular route of the internal path is flexible.
 This is the case for circuit schematics, where the electronic components have a fixed location, but many routing solutions are valid.
 In contrast, for a \emph{path-constrained} pipe the shape of the path is important.
-For example, threading electroluminescent (EL) wire through a clear pipe enables makers to emulate neon art (see Figure \ref{fig:openings}b).
+For example, threading electroluminescent (EL) wire through a clear pipe enables makers to emulate neon art (see Figure \ref{fig:openings}b) \bjoern{what happened to this subfigure?
+Kill reference if it's not coming back}.
 A \emph{fully constrained} pipe requires both specific terminals as well as a specific path.

### 2014-04-12 paper/discussion.tex
bjoern's pass

@@
 However, the fundamental operations of 3D path routing and topology specification that our tool offers will remain relevant as 3D printing technology advances.
-Multi-material printers are already available; as they advance and support the embedding of additional materials (e.g., conductive), \systemname can enable designers to efficiently specify where different functional materials should be placed at design time.
+Multi-material printers are already available; as they advance and support the embedding of additional materials (e.g., conductors), \systemname can enable designers to efficiently specify where different functional materials should be placed at design time.
 
@@
 \emph{Aural} outputs can be created with gases.
-Pipes and cavities can act as resonators can be leveraged to create 3D-printed sound chambers with explicitly designed sound characteristics~\cite{Zoran-flute} (see Figure \ref{fig:speculative}a).
+Pipes and cavities can act as resonators that can be leveraged to create 3D-printed sound chambers with explicitly designed sound characteristics~\cite{Zoran-flute} (see Figure \ref{fig:speculative}a).
 %All that is necessary for this is an enclosed chamber with a pipe connected to it via a narrow neck-point (see Figure \ref{fig:ocarina}).

### 2014-04-12 paper/exampleobjects.tex
bjoern's pass

@@
 The models we used for the toys were downloaded from the  Thingiverse website.
-We then used \systemname to create the desired internal pipes.
+\bjoern{I thought the brain model came from Berkeley?} We then used \systemname to create the desired internal pipes.
 All components were fabricated on a Makerbot, support-free.
@@
 Our smart base is powered by an Arduino Uno running open-source SFCS code\footnote{\url{http://www.instructables.com/id/Touche-for-Arduino-Advanced-touch-sensing}}.
+\bjoern{If we don't have appendices, insert resistance measurement for this paint here.}
 
@@
 \subsection{Custom Radio (open, threadable, endpoint)}
-A custom radio (figure \ref{fig:examples}c) shows how pipes can be used to integrate electronic components into the user-facing surface of an object.
+A custom radio (Figure \ref{fig:examples}c) shows how pipes can be used to integrate electronic components into the user-facing surface of an object.
 The radio includes a power and volume dial, and LED which goes on when the radio is on, and a tuner dial.
@@
 The sound is played through a single .5 Watt speaker.
-This shows a Midas-like \cite{Savage-midas} possibility: multiple terminals all routed back to a shared endpoint (in this case, the Arduino).
+This shows a Midas-like \cite{Savage-midas} possibility : multiple terminals all routed back to a shared endpoint (in this case, the Arduino).
+\bjoern{last sentence is confusing.
+May not need Midas reference.}
 
@@
 We marry this technique here with the Sauron technique \cite{Savage-sauron}: instead of using mirrors to redirect light and motion sensing as Sauron does, we use fiber optic cables.
+\bjoern{I think people remember Sauron as being all about hollow objects and we use ray-tracing to know how reflected light will travel; using fiberoptics seems not that closely related, other than its use of optical sensors.}
 
@@
 We used \systemname to add four internal open pipes connecting each pen chamber to the based on the pen holder.
-This prototype was built by our Makerbot support-free.
+This prototype was built on our Makerbot support-free.
 Each pipe was then threaded with a fiber optic cable.
-We use a single cable per pipe; our 6mm diameter fiber optic cables, in comparison to the fine tubes used in the original work, can send and receive through the same cable because their cross-sectional area is large enough to accommodate both functions.
+We use a single cable per pipe; our 6mm diameter fiber optic cables, in comparison to the fine tubes used in the original work, can emit and receive through the same cable because their cross-sectional area is large enough to accommodate both functions \bjoern{still unclear how this actually works; I'm a little worried about having any vague statements about fiber optics as Patrick could get this paper}.
 At the base of each tube is a QRE1113 line sensor digital breakout board, which has an integrated IR emitter and receiver.
@@
 The maze is based on an SVG file we manually created and processed with our internal path routing tool.
-We printed the maze on the Objet using a transparent material.
+We printed the maze on the Objet using a transparent material \bjoern{bottom doesn't look transparent.
 The final artifact was fabricated in two halves fastened together via glue, to allow removal of support material with a high-pressure water tool.
@@
 We then printed the model on the Objet using transparent material.
-The model was printed in two halves, cutting through the plane of internal path, to facilitate assembly.
+The model was printed in two halves, cutting through the plane of the internal path, to facilitate assembly.
 

### 2014-04-12 paper/relatedwork.tex
bjoern's pass

@@
 \subsubsection{Interaction for Fabrication vs. Fabrication for Interaction}
-HCI researchers are also developing novel interaction techniques for providing input to digital fabrication machines~\cite{Mueller-constructable,Willis-interactive,mixfab}, and new hybrid hand tools combine manual processes with digital fabrication~\cite{zoran2013freed,rivers2012position}.
+HCI researchers are also developing novel interaction techniques for providing input to digital fabrication machines~\cite{Mueller-constructable,Willis-interactive,mixfab}, and new hybrid hand tools that combine manual processes with digital fabrication~\cite{zoran2013freed,rivers2012position}.
 These projects are complementary as they target different aspects of the design process, though it is conceivable that our pipe design tools could be added to augmented reality modeling interfaces like MixFab~\cite{mixfab}.

### 2014-04-14 paper/designspace.tex
small changes

@@
 \caption{Pipe openings: A shows a pair of \emph{open} pipes filled with copper paint to power an LED.
-B shows an \emph{open} pipe with EL wire threaded through it.
-C shows two \emph{semi-closed} pipes capped by rubber membranes.
-In D, a rabbit has a \emph{fully enclosed} chamber to modify its weight.}
+B shows two \emph{semi-closed} pipes capped by rubber membranes.
+In C, a rabbit has a \emph{fully enclosed} chamber to modify its weight.}
 \label{fig:openings}
@@
 Depending on the purpose of the pipe, its geometry may be constrained to specific locations within the 3D model.
-An \emph{endpoint-constrained} pipe requires specific start and end points, but the particular route of the internal path is flexible.
+An \emph{endpoint-constrained} pipe requires specific start and end points, but the particular route of the internal path is irrelevant.
 This is the case for circuit schematics, where the electronic components have a fixed location, but many routing solutions are valid.
 In contrast, for a \emph{path-constrained} pipe the shape of the path is important.
-For example, threading electroluminescent (EL) wire through a clear pipe enables makers to emulate neon art (see Figure \ref{fig:openings}b) \bjoern{what happened to this subfigure?
-Kill reference if it's not coming back}.
+For example, threading electroluminescent (EL) wire through a clear pipe enables makers to emulate neon art (see Figure \ref{fig:teaser}, right).
 A \emph{fully constrained} pipe requires both specific terminals as well as a specific path.
-If a maker wishes the EL wire to exit the rear of the sign to hide the electronics, it requires a fully-constrained path (Figure \ref{fig:neon}).
+If a maker wishes the EL wire to exit the rear of the sign to hide the electronics, she can use a fully-constrained path (Figure \ref{fig:neon}).
 Finally, an \emph{unconstrained} pipe does not need a specific path or terminal points.

### 2014-04-14 paper/discussion.tex
small changes

@@
 \emph{Visual} outputs can be mediated by gases, liquids, solid, particulates, or threadables.
-Pneumatic actuation (as in PneUIs \cite{Yao-pneui} or Harrison's latex buttons \cite{Harrison-buttons}) can provides shape-changing interfaces if maelleable materials can be printed.
+Pneumatic actuation (as in PneUIs \cite{Yao-pneui} or Harrison's latex buttons \cite{Harrison-buttons}) provides shape-changing interfaces if maelleable materials can be printed.
 %Gases carrying smoke also offer visual feedback. 
@@
 Similarly, the Touch\'{e} system \cite{Sato-touche} relies on distinguishing \emph{capacitive} signatures of objects, both in static states and as their boundary conditions are changed by human interaction.
-Weight could also be used as a simple metric of object identity, but would be less distinctive than acoustic or capacitive signatures.
+Weight could also be used as a simple metric of identity, but is less distinctive than acoustic or capacitive signatures.
 

### 2014-04-14 paper/exampleobjects.tex
bjoern's pass
small changes

@@
 \section{Totally Tubular Example Objects}
-To evaluate and highlight our tool's capabilities, we fabricated a set of five prototype objects designed using \systemnamenospace.
+To evaluate and highlight our tool's capabilities, we fabricated a set of six prototype objects designed using \systemnamenospace.
 To show that our approach is printer-independent, we fabricated some on a Makerbot Replicator 2 and the rest on an Objet Connex 260.
@@
 The sound is played through a single .5 Watt speaker.
-This shows a Midas-like \cite{Savage-midas} possibility : multiple terminals all routed back to a shared endpoint (in this case, the Arduino).
-\bjoern{last sentence is confusing.
-May not need Midas reference.}
 
@@
 The maze is based on an SVG file we manually created and processed with our internal path routing tool.
-We printed the maze on the Objet using a transparent material \bjoern{bottom doesn't look transparent.
+We printed the maze on the Objet using a transparent material \bjoern{bottom doesn't look transparent}.
 The final artifact was fabricated in two halves fastened together via glue, to allow removal of support material with a high-pressure water tool.

### 2014-04-14 paper/introduction.tex
small changes

@@
 \section{Introduction}
-Makers, researchers, and professional designers (jointly "makers" for the remainder of this paper) increasingly leverage 3D printers as tools for design work.
+Makers, researchers, and professional designers (jointly ``makers" for the remainder of this paper) increasingly leverage 3D printers as tools for design work.
 A wide array of objects, ranging from toy figurines to video game controllers and jet engine blades, are now prototyped (and sometimes even manufactured) using these machines.
@@
 
-This paper introduces a novel technique involving the removal of material from the interior of 3D models, prior to printing, to create pipes and other cavities.
-These pipes can be filled, post-print, with a variety of media that enable input, display, and tactile feedback.
+This paper introduces a novel technique involving removing material from the interior of 3D models, prior to printing, to create pipes and other cavities.
+These pipes can be filled, post-print, with a variety of media to enable input, display, and tactile feedback.
 This {\em subtractive} approach is complementary to {\em additive} approaches that inject custom print materials, such as conductors, during the actual printing process \cite{Sells-reprap}.
-While our approach requires some manual assembly after a print completes, it gives makers the opportunity to cheaply and rapidly prototype a diverse range of interactive objects on 3D printers, including popular single-material fused-deposition modeling machines (Figure \ref{fig:teaser}).
+While our approach requires some manual assembly after a print completes, it gives makers the opportunity to cheaply and rapidly prototype a range of interactive objects on 3D printers, including popular single-material fused-deposition modeling machines (colored objects in Figure \ref{fig:teaser}).
 

### 2014-04-14 paper/exampleobjects.tex
fixing feedback

--- 2014-04-12 paper/exampleobjects.tex
@@
 
-The models we used for the toys were downloaded from the  Thingiverse website.
-\bjoern{I thought the brain model came from Berkeley?} We then used \systemname to create the desired internal pipes.
+The models we used for the boat was downloaded from Thingiverse, while the brain model came from researchers at one of our instutions.
+\valkyrie{how to phrase that?} We then used \systemname to create the desired internal pipes.
 All components were fabricated on a Makerbot, support-free.
@@
 This paint requires approximately 2 days to fully dry in this configuration (3mm diameter tubes, maximum tube length 10cm).
-Our smart base is powered by an Arduino Uno running open-source SFCS code\footnote{\url{http://www.instructables.com/id/Touche-for-Arduino-Advanced-touch-sensing}}.
-\bjoern{If we don't have appendices, insert resistance measurement for this paint here.}
+The resistance of the paint in this configuation is approximately $2\Omega / inch.$  Our smart base is powered by an Arduino Uno running open-source SFCS code\footnote{\url{http://www.instructables.com/id/Touche-for-Arduino-Advanced-touch-sensing}}.
 
@@
 The sound is played through a single .5 Watt speaker.
-This shows a Midas-like \cite{Savage-midas} possibility : multiple terminals all routed back to a shared endpoint (in this case, the Arduino).
-\bjoern{last sentence is confusing.
-May not need Midas reference.}
+\valkyrie{Tovi, if you are thinking of a Midas reference here...
+can you help write it?}
 
@@
 Our pen holder uses a modification of the FlyEye technique described by Wimmer \cite{Wimmer-flyeye} in which an IR camera senses a user's grasp on an object, getting data from IR light transmitted on fiber optic cables.
-We marry this technique here with the Sauron technique \cite{Savage-sauron}: instead of using mirrors to redirect light and motion sensing as Sauron does, we use fiber optic cables.
-\bjoern{I think people remember Sauron as being all about hollow objects and we use ray-tracing to know how reflected light will travel; using fiberoptics seems not that closely related, other than its use of optical sensors.}
 
@@
 The maze is based on an SVG file we manually created and processed with our internal path routing tool.
-We printed the maze on the Objet using a transparent material \bjoern{bottom doesn't look transparent.
+We printed the maze on the Objet using a transparent material for the top and a black material for the bottom.
 The final artifact was fabricated in two halves fastened together via glue, to allow removal of support material with a high-pressure water tool.

### 2014-04-14 paper/exampleobjects.tex

@@
 The sound is played through a single .5 Watt speaker.
-\valkyrie{Tovi, if you are thinking of a Midas reference here...
+%\valkyrie{Tovi, if you are thinking of a Midas reference here...
 can you help write it?}

### 2014-04-14 paper/tubes.tex
bjoern's pass

@@
 A* finds a least-cost path through a graph of nodes given a cost function.
-Our graph comprises the grid of voxels as nodes, with edges between adjacent nodes; our path cost is the based on the Euclidean distance to the selected end point.
+Our graph comprises the grid of voxels as nodes, with edges between adjacent nodes; our path cost is based on the Euclidean distance to the selected end point.
 In this formulation, the routed path is naturally constrained to stay within the object boundaries.
@@
 Our A* pathfinding produces the \emph{shortest} path, but in many cases
-we require a smoother path to aid the insertion of the desired medium, such as a threadable wire.
+we require a smoother path to aid insertion of the desired medium, such as a threadable wire.
 In addition, the distance from the surface is not taken into account in the A* search and pipes may thus be routed along surface contours, which may cause leakage.
@@
 a straight configuration, while satisfying the user-selected endpoint position constraints.
-In addition to user constraints, we require that the rod be perpendicular to the surface at its ends to prevent the terminal \emph{shape} from deforming.
+In addition to user constraints, we require that the rod be perpendicular to the surface at its ends to prevent the terminals' shape from deforming.
 
@@
 This means that overhangs without added support material can be approximately $45^{\circ}$, and although cylinders (e.g., pipes) and spheres violate this in some places they can be printed support-free.
-The models in Figure \ref{fig:teaser}a were produced on a Makerbot FDM without any support structures inside the pipes.
+The colored models in Figure \ref{fig:teaser} were produced on a Makerbot FDM without any support structures inside the pipes.
 
@@
 Complex pipe geometries can make water blasting impossible and other processes more labor- and time-intensive.
-All models in Figure \ref{fig:teaser}b required support material removal post-print: the neon sign was additionally cut into two pieces to give better access to the support.
+All models in Figure \ref{fig:teaser}b required support material removal post-print: the neon sign and maze were additionally cut into two pieces to give better access to the support.
+\bjoern{should we say something about stereolithography printers like the Form 1?
+we don't have one, but they could be ideal for our techniques since the liquid will just flow out of our tubes.
+However, they also generate support structure and I'm not sure how those supports differ from FDM.}
 

### 2014-04-14 paper/tubes.tex
working through stuff, adding discussion of powder-based printers

@@
 The physical fabrication process used by different machines gives rise to differing strategies for avoiding support material or easing its removal.
-We will discuss techniques for fused deposition modeling (FDM) machines, such as the Makerbot\footnote{\url{http://makerbot.com}}, and multi-jetting machines such as an Objet\footnote{\url{http://stratasys.com}}.
+We will discuss techniques for fused deposition modeling (FDM) machines, such as the Makerbot\footnote{\url{http://makerbot.com}}, multi-jetting machines such as an Objet\footnote{\url{http://stratasys.com}}, and powder-based machines such as the ZCorp\footnote{\url{http://3dsystems.com}}.
+Other varities of machines, e.g., stereolithography machines such as the Form 1\footnote{\url{http://formlabs.com}}, require similar techniques to these.
 
@@
 All models in Figure \ref{fig:teaser}b required support material removal post-print: the neon sign and maze were additionally cut into two pieces to give better access to the support.
-\bjoern{should we say something about stereolithography printers like the Form 1?
-we don't have one, but they could be ideal for our techniques since the liquid will just flow out of our tubes.
-However, they also generate support structure and I'm not sure how those supports differ from FDM.}
+
+\subsection{Powder-based}
+Powder-based machines have a bed of powder which is selectively fused together using glue droplets.
+As each layer is completed, the bed moves down and another layer of powder is added on top.
+The unfused powder functions as the support structure for the fused powder: this makes the support trivial to remove post-print, as the loose grains will simply fall out of pipes and cavities in the model.
+These machines unfortunately cannot fabricate flexible or transparent materials (yet), however they can print in full-color.
+We did not have access to a powder-based printer, though in future work we would like to experiment with one.
 

### 2014-04-14 paper/exampleobjects.tex
working through stuff, adding discussion of powder-based printers
discuss fiber optics/light pipes

@@
 The models we used for the boat was downloaded from Thingiverse, while the brain model came from researchers at one of our instutions.
-\valkyrie{how to phrase that?} We then used \systemname to create the desired internal pipes.
+We then used \systemname to create the desired internal pipes.
 All components were fabricated on a Makerbot, support-free.
@@
 This prototype was built on our Makerbot support-free.
-Each pipe was then threaded with a fiber optic cable.
-We use a single cable per pipe; our 6mm diameter fiber optic cables, in comparison to the fine tubes used in the original work, can emit and receive through the same cable because their cross-sectional area is large enough to accommodate both functions \bjoern{still unclear how this actually works; I'm a little worried about having any vague statements about fiber optics as Patrick could get this paper}.
-At the base of each tube is a QRE1113 line sensor digital breakout board, which has an integrated IR emitter and receiver.
-When a pen is in its appointed place, the emitted infrared light is reflected off its bottom and travels back to the receiver, where it registers as bright.
+Each pipe was then threaded with Sparkfun clear core light pipe (PRT-10697), similar to fiber optic cable.
+We use a single cable per pipe; our 6mm diameter light pipe, in comparison to the fine fiber optics used in the original work, can emit and receive through the same cable because the cross-sectional area is large enough to accommodate both functions.
+At the base of each pipe is a QRE1113 line sensor breakout board, with integrated IR emitter and receiver.
+When a pen is in its appointed place, the emitted infrared light is reflected off its base and travels back to the receiver, where it registers as bright.
 

### 2014-04-14 paper/exampleobjects.tex
roboto images

@@
 (a) shows the input SVG files, with the eyes (non-animated) in black and the two states of the mouth in {\color{blue}blue} and {\color{gray}gray}.
-(b) shows the resultant mesh when we subtracted the generated tubes, with rear exit tubes circled in {\color{blue}blue} \valkyrie{fix this with boolean magic}.
+(b) shows the tubes we subtracted from the mesh, rear exit tubes circled in {\color{blue}blue}.
 In (c), we show the fabricated sign.}
@@
 
-The pipes are threaded with electroluminescent wire which is lit in sequence, using an EL wire sequencer, to create the animation.
+The pipes are threaded with electroluminescent wire which is lit in sequence, using an sequencer, to create the animation.

### 2014-04-14 paper/exampleobjects.tex
need to merge

--- 2014-04-14 paper/exampleobjects.tex
@@
 
-\subsection{Touch-sensitive Toys (open, liquid, star, endpoint)}
+\subsection{Touch-sensitive Toys (open, endpoint, star, liquid)}
 
@@
 
-\subsection{Breathing Bunny (semi-closed, gas, endpoint)}
+\subsection{Breathing Bunny (semi-closed, endpoint, gas)}
 
@@
 
-\subsection{Custom Radio (open, threadable, endpoint)}
+\subsection{Custom Radio (open, endpoint, threadable)}
 A custom radio (Figure \ref{fig:examples}c) shows how pipes can be used to integrate electronic components into the user-facing surface of an object.
@@
 
-\subsection{Presence-aware Pen Holder (open, threadable, endpoint)}
+\subsection{Presence-aware Pen Holder (open, endpoint, threadable)}
 Our presence-aware pen holder can distinguish which tool or tools a user has picked up (see Figure \ref{fig:examples}d).
@@
 
-\subsection{Maze (fully enclosed, particulate, tree, path)}
+\subsection{Maze (fully enclosed, path, tree, particulate)}
 
@@
 
-\subsection{Animated Neon Sign (open, threadable, fully constrained)}
+\subsection{Animated Neon Sign (open, fully constrained, threadable)}
 

### 2014-04-14 paper/tubes.tex
need to merge

--- 2014-04-14 paper/tubes.tex
@@
 we require a smoother path to aid insertion of the desired medium, such as a threadable wire.
-In addition, the distance from the surface is not taken into account in the A* search and pipes may thus be routed along surface contours, which may cause leakage.
+In addition, the distance from the surface is not taken into account in the A* search and pipes may thus be routed along surface contours, which may cause leakage or structural deficiencies.
 
@@
 Because we have a 3-dimensional canvas in which to work, it would be possible to push overlapping paths into the third dimension.
-However, we leave this to future work: for our specific applications so far, it has been unnecessary.
+However, we leave this to future work.
+For our specific applications so far, it has been unnecessary.
 
@@
 The physical fabrication process used by different machines gives rise to differing strategies for avoiding support material or easing its removal.
-We will discuss techniques for fused deposition modeling (FDM) machines, such as the Makerbot\footnote{\url{http://makerbot.com}}, and multi-jetting machines such as an Objet\footnote{\url{http://stratasys.com}}.
+We next discuss techniques for fused deposition modeling (FDM) machines, such as the Makerbot\footnote{\url{http://makerbot.com}} and multi-jetting machines such as Objet\footnote{\url{http://stratasys.com}}, as well as powder-based (e.g., ZCorp\footnote{\url{http://3dsystems.com}}) and stereolithography (e.g., Form 1\footnote{\url{http://formlabs.com}}).
 
@@
 All models in Figure \ref{fig:teaser}b required support material removal post-print: the neon sign and maze were additionally cut into two pieces to give better access to the support.
-\bjoern{should we say something about stereolithography printers like the Form 1?
-we don't have one, but they could be ideal for our techniques since the liquid will just flow out of our tubes.
-However, they also generate support structure and I'm not sure how those supports differ from FDM.}
+
+\subsection{Powder-based and Stereolithography}
+Powder-based and stereolithography (SLA) machines work with a bed of powder or liquid, respectively.
+The material at the surface is fused together via glue droplets (for powder) or a UV laser (for SLA).
+The remaining powder or liquid serves as support for the model.
+These types of printers are perhaps even more suitable for pipe creation than either FDM or multi-jetting printers, as the supporting liquid or loose powder is simpler to remove than structural plastic or UV cured resin.
+We hope to experiment with them in the future.
 

### 2014-04-14 paper/conclusion.tex
paper sent to gord

@@
 This amplifies and extends the utility of existing sensing and actuation approaches.
-In addition to describing the design space of pipes, we also presented \systemname, a design tool for the creation of tubes inside arbitrary 3D models, and discussed many techniques that combine nicely wtih pipes.
+In addition to describing the design space of pipes, we also presented \systemnamenospace, a design tool for the creation of tubes inside arbitrary 3D models, and discussed many techniques that combine nicely wtih pipes.
 We fabricated several example objects using our tool to explore new points in the design space.

### 2014-04-14 paper/designspace.tex

@@
 %
-\emph{Semi-enclosed} pipes are capped on one side and could be used to create create tactile output.
+\emph{Semi-enclosed} pipes are capped on one side and could be used to create tactile output.
 By fabricating a cap from a malleable material and attaching a pump to the open end, haptic feedback can be generated by raising and depressing the flexible cap (see Figure \ref{fig:openings}b).
@@
 
-In the next section detail we how different pipes within this design space space can be created with the help of our new design tool, \systemnamenospace.
+In the next section we detail how different pipes within this design space space can be created with the help of our new design tool, \systemnamenospace.

### 2014-04-14 paper/discussion.tex

@@
 \emph{Aural} outputs can be created with gases.
-Pipes and cavities can act as resonators that can be leveraged to create 3D-printed sound chambers with explicitly designed sound characteristics~\cite{Zoran-flute} (see Figure \ref{fig:speculative}a).
+Pipes and cavities can act as resonators and be leveraged to create 3D-printed sound chambers with explicitly designed sound characteristics~\cite{Zoran-flute} (see Figure \ref{fig:speculative}a).
 %All that is necessary for this is an enclosed chamber with a pipe connected to it via a narrow neck-point (see Figure \ref{fig:ocarina}).
@@
 For a cylindrical wire, resistance = $\rho\frac{l}{\pi r^2}$, where $\rho$ is based on the conductivity of the medium.
-In our case, because $r << l$, $resistance(2) < resistance(3) < resistance(1)$.}
+Thus we can modify resistance based on pipe length and cross-sectional area.}
 \label{fig:speculative}

### 2014-04-14 paper/exampleobjects.tex

@@
 We connected one pipe to each of our pump's terminals, and using a programmable power supply we mimic a rabbit's breathing pattern.
-This example was printed on the Objet using a flexible material and its support material was flushed post-print.
+This example was printed on the Objet using flexible material; support was flushed post-print.
 
@@
 The radio includes a power and volume dial, and LED which goes on when the radio is on, and a tuner dial.
-An Arduino Pro Micro microcontroller, Si4703 FM tuner, and 6V (4xAA) power supply are embedded in the base of the radio.
+An Arduino Uno microcontroller, Si4703 FM tuner, and 6V (4xAA) power supply are embedded in the base of the radio.
 The sound is played through a single .5 Watt speaker.
@@
 We included slots on the 3D model matching the shape of the electronic components that were to be embedded.
-The model was then imported into \systemnamenospace, where we created a network of 8 individual open pipes (3 per potentiometer, 1 for the LED, 1 for the speaker) to connect the components to the base of the radio.
-We fabricated the radio itself on our Makerbot, support-free, but cut in two pieces to allow the Arduino and battery pack to be inserted.
-We then threaded the pipes with wires to connect the components to the microcontroller at the base of the radio.
+The model was then imported into \systemnamenospace, where we created a network of 8 individual open pipes (3 per potentiometer, 1 for the LED, 1 for the speaker) connecting the components to the radio's base.
+We fabricated the radio body on our Makerbot, support-free but cut in two pieces to allow the Arduino and battery pack to be inserted.
+We then threaded the pipes with wires to connect the components to the microcontroller in the base.
 
 \subsection{Presence-aware Pen Holder (open, endpoint, threadable)}
-Our presence-aware pen holder can distinguish which tool or tools a user has picked up (see Figure \ref{fig:examples}d).
-Our pen holder uses a modification of the FlyEye technique described by Wimmer \cite{Wimmer-flyeye} in which an IR camera senses a user's grasp on an object, getting data from IR light transmitted on fiber optic cables.
+Our presence-aware pen holder distinguishes which tool or tools a user has picked up (see Figure \ref{fig:examples}d).
+We use a modification of the FlyEye technique described by Wimmer \cite{Wimmer-flyeye} in which an IR camera senses a user's grasp on an object, getting data from light transmitted on fiber optic cables.
 
@@
 
-The pipes are threaded with electroluminescent wire which is lit in sequence, using an sequencer, to create the animation.
+The pipes are threaded with electroluminescent wire which is lit in sequence, using a sequencer, to create the animation.

### 2014-04-14 paper/limitations.tex

@@
 Fluids also raise unique challenges.
-For example, the conductive paint we used in our designs can take multiple days to dry in tubes of sufficient length.
-Liquids kept in the tubes can also cause discoloration due to seepage.
+For example, the conductive paint we used in our designs can take multiple days to dry in long pipes.
+Liquids kept in the pipes can also cause discoloration due to seepage.
 Furthermore, not all printers guarantee watertightness of their prints, so leakage can occur.

### 2014-04-14 paper/tubes.tex

@@
 
-The wire is modelled as a \emph{discrete rod}, i.e., a 3D poly-line with a bending 
+The wire is modelled as a \emph{discrete rod}: a 3D poly-line with a bending 
 constraint between adjacent segments. Our simulation is based on an
@@
 modelled as penalty forces. To keep the rod inside the
-mesh, we compute a discretized distance field; we can then constrain the
-rod to stay within our offset shell, again using a simple penalty force.
+mesh, we compute a discretized distance field, then constrain the
+rod to stay within our offset shell using a simple penalty force.
 
@@
 To allow designers to create pipes with multiple endpoints (e.g., the star topologies in Figure \ref{fig:examples}a), we use one surface position per pipe and a central fixed endpoint that the user can interactively position (see Figure~\ref{fig:boat-star}).
-The user can connect several pipes, hitting a key to add each one in sequence and then positioning them with the mouse.
+The user can connect several pipes, hitting a key to add each in sequence and positioning them with the mouse.
 The pipes route as normal.
@@
 Our technique for routing this path is inspired by work in generating continuous line illustrations from images~\cite{Wong-continuousline,Bosch-tsp}.
-The interior path routing problem is a version of the Chinese Postman Problem\footnote{The CPP is also known as the route inspection problem: \url{http://en.wikipedia.org/wiki/Route_inspection_problem}}, in which we wish to traverse all edges of the graph described in the user's input, much as a postman needs to walk along every road at least once to deliver mail.
-In our relaxation, we allow creation of new edges (i.e., the postman may jump from building to building in addition to using existing roads).
+Our routing problem is a version of the Chinese Postman Problem\footnote{The CPP is also known as the route inspection problem: \url{http://en.wikipedia.org/wiki/Route_inspection_problem}}, in which we wish to traverse all edges of the user's input graph, much as a postman needs to walk along every road at least once to deliver mail.
+In our relaxation, we allow creating new edges (i.e., the postman may jump from building to building in addition to using existing roads).
 %
@@
 
-{\bf Connect disconnected components:} To connect disconnected subgraphs, we find minimum Euclidean distance vertex pairs that span two subgraphs, and greedily add edges to such pairs until all subgraphs are joined into a single connected graph.
+{\bf Connect disconnected components:} To connect disconnected subgraphs, we find minimum Euclidean distance vertex pairs spanning two subgraphs, and greedily add edges to such pairs until all subgraphs are joined into a single graph.
 %To connect disconnected subgraphs in the input, each disconnected subgraph's vertices and edges are contracted to a single vertex.
@@
 We greedily add edges between the odd nodes until no odd nodes remain.
-Finally, we remove our temporary edge, which changes the graph from Eulerian to semi-Eulerian.
+Finally, we remove our temporary edge, changing the graph from Eulerian to semi-Eulerian.
 This is a connected, semi-Eulerian graph which contains all edges in the input.

### 2014-04-14 paper/designspace.tex
small mods

@@
 We have identified four relevant dimensions of pipe design: \emph{openings}, \emph{path constraints}, \emph{topologies}, and inserted \emph{media}.
-We describe the space (see Figure \ref{fig:pipespace}) below.
+We describe the space (see Figure \ref{fig:pipespace}).
 
@@
 Pipe network topologies enable \emph{splitting} or \emph{mixing} of media (see Figure \ref{fig:pipespace}).
-\emph{Star} and \emph{tree} topologies extend the splitting and mixing primitives, encompassing multiple-in and multiple-out configurations; we use star topologies to create different sensing location for our touch-sensitive toys, with one terminal used for swept-frequency capacitive sensing (Figure~\ref{fig:examples}a).
+\emph{Star} and \emph{tree} topologies extend the splitting and mixing primitives, encompassing multiple-in and multiple-out configurations; we use star topologies to create different sensing locations for our touch-sensitive toys, where one terminal is used for swept-frequency capacitive sensing (Figure~\ref{fig:examples}a).
 

### 2014-04-14 paper/relatedwork.tex
small mods

--- 2014-04-11 paper/relatedwork.tex
@@
 Sauron tracks interactions with printed physical mechanisms by placing a camera inside hollow objects, where all components must be visible to the camera~\cite{Savage-sauron}.
-Our technique instead can work with general, solid models and only carves out pipes for interaction where needed.
+Our technique instead can work with arbitrary, solid models and only carves out pipes for interaction where needed.
 Midas generates fabrication files for custom capacitive touch sensors~\cite{Savage-midas}.
@@
 Circuitboard routing algorithms, such as Lee's maze router \cite{Lee-maze}, assume 2D surfaces.
-Multi-layer boards consist of stacked of 2D layers connected by vias.
+Multi-layer boards consist of stacked 2D layers connected by vias.
 We compute true 3D point to point routing though the interior of solid models with flexible paths.
@@
 
-\subsubsection{Interaction for Fabrication vs. Fabrication for Interaction}
-HCI researchers are also developing novel interaction techniques for providing input to digital fabrication machines~\cite{Mueller-constructable,Willis-interactive,mixfab}, and new hybrid hand tools combine manual processes with digital fabrication~\cite{zoran2013freed,rivers2012position}.
+%\subsubsection{Interaction for Fabrication vs. Fabrication for Interaction}
+%HCI researchers are also developing novel interaction techniques for providing input to digital fabrication machines~\cite{Mueller-constructable,Willis-interactive,mixfab}, and new hybrid hand tools that combine manual processes with digital fabrication~\cite{zoran2013freed,rivers2012position}.
 These projects are complementary as they target different aspects of the design process, though it is conceivable that our pipe design tools could be added to augmented reality modeling interfaces like MixFab~\cite{mixfab}.

### 2014-04-15 paper/discussion.tex
small mods
integrate gord comments, smash discussion sections into example  objects

@@
 Multi-material printers are already available; as they advance and support the embedding of additional materials (e.g., conductors), \systemname can enable designers to efficiently specify where different functional materials should be placed at design time.
-
-We described a design space and fabricated example objects to demonstrate some of the interactions that can be aided by pipe routing.
-There are many additional possibilities for leveraging sensing and actuation techniques from the HCI literature and extending their use in 3D-printed objects.
-We briefly survey some of them.
-
-\subsection{Inputs and Outputs}
-
-Pipe-based outputs can address the five senses: visual, aural, haptic, and olfactory/gustatory.
-As we have already discussed many input techniques through our example objects, we focus our discussion around output.
-%In addition to outputs we mention explicitly, any outputs possible using traditional electronic components are possible using conductive material in pipes.
-
-\begin{figure}[h]
-\centering
-    \includegraphics[width=3.4in]{figures/gates.png}
-\caption{By attaching a servo to a gate structure, it is possible to direct fluid through that pipe.
-The gate not only has binary positions (a) and (b), but can direct fractional flow, as well (c).
-This could be used, for example, in science exhibits.}
-\label{fig:direct}
-\end{figure}
-
-\emph{Visual} outputs can be mediated by gases, liquids, solid, particulates, or threadables.
-Pneumatic actuation (as in PneUIs \cite{Yao-pneui} or Harrison's latex buttons \cite{Harrison-buttons}) provides shape-changing interfaces if maelleable materials can be printed.
-%Gases carrying smoke also offer visual feedback. 
-Air bubbles in liquids can also convey information~\cite{Heiner1999Percolator}.
-Splitting and mixing of colored liquids can be used as a display tactic; liquids can also be mixed in specific concentrations by servo-controlled gate structures as in Figure \ref{fig:direct}.
-%additionally mechanical motion can be fluid-mediated.
-
-\emph{Aural} outputs can be created with gases.
-Pipes and cavities can act as resonators and be leveraged to create 3D-printed sound chambers with explicitly designed sound characteristics~\cite{Zoran-flute} (see Figure \ref{fig:speculative}a).
-%All that is necessary for this is an enclosed chamber with a pipe connected to it via a narrow neck-point (see Figure \ref{fig:ocarina}).
-Passive sound amplification and sound redirection are also possible using pipes, and sensing of some interactions, e.g., \emph{tapping and grasping}, can be sensed acoustically.
-Due to differential sound conductivity in different printed materials, solid pipes of a material with greater conductivity (i.e., rigid, high-density) can be embedded in a model of a lesser conductivity (i.e., softer).
-A microphone or piezo placed at the system-side of these sound-conducting pipes can determine where the model was tapped and how it was tapped.
-Active acoustic sensing as described in Touch \& Activate \cite{Ono-touchandactivate} is also possible; this technique would allow for particular areas of interest (connected to rigid pipes) to be touch sensitive and enable grasp sensing.
-
-\begin{figure}[h]
-\centering
-    \includegraphics[width=3.4in]{figures/speculative.png}
-\caption{Some possible pipe uses.
-The printed whistle in (a) is a Helmholz resonator: it creates a sound when air is blown through the pipe at the front, because air compresses at the small neck joint between the internal chamber and the pipe.
-In (b), a capacitor manufactured on a 3D printer coated post-print with copper paint.
-This particular one holds a charge of \valkyrie{XX}.
-(c) gives a comparison of pipe-based resistors, also filled with copper paint.
-For a cylindrical wire, resistance = $\rho\frac{l}{\pi r^2}$, where $\rho$ is based on the conductivity of the medium.
-Thus we can modify resistance based on pipe length and cross-sectional area.}
-\label{fig:speculative}
-\end{figure}
-
-\emph{Haptic} outputs are possible through compressible and incompressible fluids, as discussed, or through threadable inputs: e.g., muscle wire can be inserted in flexible pipes and actuated, similar to shape changing interfaces \cite{Coelho-shapechanging}.
-
-\emph{Olfactory} and \emph{gustatory} outputs can be created through the mixing and splitting of pipes carrying scented or flavored fluids, respectively~\cite{kaye2004making}.
-
-\subsection{Identification}
-In addition to active inputs from users, pipes can be used for passive \emph{identification}.
-Fully-enclosed pipes which are not visually apparent can change the weight, acoustic resonance, or capacitive signature of a 3D object.
-
-Each of these properties could be used for object identification.
-As seen in Touch \& Activate \cite{Ono-touchandactivate}, different objects have different acoustic signatures.
-As such, two objects that are visually identical but have differing internal chamber geometry may be distinguished acoustically.
-Thus we can recall an object's identity once its acoustic signature has been recorded.
-Similarly, the Touch\'{e} system \cite{Sato-touche} relies on distinguishing \emph{capacitive} signatures of objects, both in static states and as their boundary conditions are changed by human interaction.
-Weight could also be used as a simple metric of identity, but is less distinctive than acoustic or capacitive signatures.
-
-%We have experimented with intentional encoding.
-Semi-closed pipes can function as resonance chambers.
-A pipe's resonant frequency can be measured by attaching a speaker and microphone to it, sweeping frequencies with the speaker, and performing a Fourier transform on the resultant signal from the microphone.
-Peaks in the transformed data correspond to resonant frequencies of the object.
-Intentional encoding in the case of capacitance is significantly more difficult, and we leave it to future work.
-Encoding of weight is trivial by creating fully-enclosed chambers inside an object.
-This is similar to Infrastructs \cite{Willis-infrastructs}.
-While their technique requires access a terahertz imaging tool, audio resonance identity encoding requires only a microphone and a speaker.
-
-\subsection{Passive Components}
-
-Printing or injecting conductive material can also be used to design passive components with particular electrical properties.
-For example, we printed a proof-of-concept  capacitor which features a thin printed plate with hollow cones on either side (see Figure \ref{fig:speculative}b).
-After printing, we coated the inner surfaces of capacitor with copper paint.
-It now can hold a charge of \valkyrie{xx}.
-Similarly, tubes can be used to create \emph{resistors} in ``printed'' circuitry: a longer or thinner tube has a greater resistance than a shorter or wider one (see Figure \ref{fig:speculative}c).
-Internal \emph{antenna} design is also possible, which could lead to new flexibility in, e.g., cell phone design.
-Future design tools could help designers achieve desired electrical properties automatically.

### 2014-04-15 paper/exampleobjects.tex
integrate gord comments, smash discussion sections into example  objects

@@
 
+Identification of these objects and gestures could also be performed using active acoustic sensing \cite{Ono-touchandactivate}: the internal pipes, although not externally visually distinctive, affect the acoustic signature of the object.
+Intentional identity encoding using pipes could also be accomplished through print weight modification using fully enclosed pipes.
+
 \subsection{Breathing Bunny (semi-closed, endpoint, gas)}
@@
 This example was printed on the Objet using flexible material; support was flushed post-print.
+
+Additional options for haptic feedback include the insertion of muscle wire into flexible printed pipes, creating shape changing interfaces \cite{Coelho-shapechanging}.
 
@@
 
+Creating aural outputs using the printed pipes directly is also a possibility: the whistle in Figure \ref{fig:speculative}a is a fully printable audio output which is based on a Helmholz resonator, and which can be activated with air pressure.
+
 \subsection{Presence-aware Pen Holder (open, endpoint, threadable)}
@@
 
+Some passive components required in the design of the recognition circuit could in the future be created using pipes and printed or injected conductive material.
+For example, we printed a proof-of-concept  capacitor which features a thin printed plate with hollow cones on either side (see Figure \ref{fig:speculative}b).
+After printing, we coated the inner surfaces of capacitor with copper paint.
+It now can hold a charge of \valkyrie{xx}.
+Similarly, tubes can be used to create \emph{resistors} in ``printed'' circuitry: a longer or thinner tube has a greater resistance than a shorter or wider one (see Figure \ref{fig:speculative}c).
+
 \subsection{Maze (fully enclosed, path, tree, particulate)}
@@
 We plugged the ends of the tubes post-print with additional material to ensure the ball does not escape.
+
+Integration of our maze with, e.g., PneUIs \cite{Yao-pneui} could result in a maze that helps a user along or blocks her path.
 
@@
 The pipes are threaded with electroluminescent wire which is lit in sequence, using a sequencer, to create the animation.
+
+A similar display could be created using colored liquids along with mixing and splitting topologies.
+Mixing liquids in specific concentrations (e.g., for color output) could be performed using a servo-controlled gate as in Figure \ref{fig:direct}.
+
+\begin{figure}
+\centering
+    \includegraphics[width=3.4in]{figures/speculative.png}
+\caption{Some possible pipe uses.
+The printed whistle in (a) is a Helmholz resonator: it creates a sound when air is blown through the pipe at the front, because air compresses at the small neck joint between the internal chamber and the pipe.
+In (b), a capacitor manufactured on a 3D printer coated post-print with copper paint.
+This particular one holds a charge of \valkyrie{XX}.
+(c) gives a comparison of pipe-based resistors, also filled with copper paint.
+For a cylindrical wire, resistance is based on its length and cross-sectional area.}
+\label{fig:speculative}
+\end{figure}
+
+\begin{figure}
+\centering
+    \includegraphics[width=3.4in]{figures/gates.png}
+\caption{By attaching a servo to a gate structure, it is possible to direct fluid through that pipe.  The gate not only has binary positions (a) and (b), but can direct fractional flow, as well (c).}
+\label{fig:direct}
+\end{figure}

### 2014-04-15 paper/introduction.tex
need to merge
integrate gord comments, smash discussion sections into example  objects

@@
 A wide array of objects, ranging from toy figurines to video game controllers and jet engine blades, are now prototyped (and sometimes even manufactured) using these machines.
-Most models fabricated on 3D printers today are passive: they express the form, but not the function, of an object.
+Most models fabricated on 3D printers today are passive: they express the form, but not the interactivity, of an object.
 
@@
 
-Generating the appropriate pipes for these examples is not trivial.
+Figure \ref{fig:teaser} shows a collection of interactive objects with various functionalities including haptic feedback, neon sign-like visual output, and selective touch-sensitivity.
+All these objects have had the electronic components required to function redirected onto their surfaces using pipes.
+But generating the appropriate pipes for these examples is not trivial.
 For one, most 3D modelling tools do not provide extensive tools for modelling the  {\em interior} of 3D models.

### 2014-04-15 paper/discussion.tex
gord feedback

@@
 Multi-material printers are already available; as they advance and support the embedding of additional materials (e.g., conductors), \systemname can enable designers to efficiently specify where different functional materials should be placed at design time.
+
+\subsection{Limitations and Future Work}
+While pipes provide a powerful mechanism for directing interaction to arbitrary locations on an object's surface, they do have shortcomings which warrant discussion.
+In particular, they cannot replace the active elements required for a functioning interactive device: they simply redirect sensing to different locations on an object's surface.
+All pipes need to, at some point, connect to a sensor, microcontroller, or pump.
+
+Some media are difficult to insert in pipes, or may not be compatible with all types of pipes.
+Threadable materials in particular can provide challenges, as they can get stuck while being fed into pipes.
+Our system mitagates this by minimizing the bend energy of the interior pipes, but for complex pipe geometries threading could be a challenge.
+
+Fluids also raise unique challenges.
+For example, the conductive paint we used in our designs can take multiple days to dry in long pipes.
+Liquids kept in the pipes can also cause discoloration due to seepage.
+Furthermore, not all printers guarantee watertightness of their prints, so leakage can occur.
+
+As discussed, removal of support material can also be a challenge.
+Particularly on multi-jetting printers, printing support-free is not possible for most geometries, and detaching printed support is not always trivial.
+In some cases we printed our models in several pieces, to avoid this problem.
+
+There are also areas for improvement and future work with our software tool.
+The design of \systemname was based on our proposed design space of pipes.
+However, the design tool is not yet fully capable of realizing all possibilities in this design space.
+Specifically, we did not explicitly explore the mixing or splitting of pipes, and our path-constrained cuts are strictly 2D for now.
+A 2.5D SVG import using, e.g., line colors as height would be straightforward, and our described algorithm would continue to work properly.
+Furthermore, since pipes are committed to the mesh one at a time, our tool does not currently optimize networks of multiple pipes.
+Instead, we utilize a greedy algorithm and route a single pipe at a time.
+This suggests the need for a better general 3D routing algorithm with multiple traces.
+Future work should also include assisting makers in the fabrication process itself by automatically cutting objects to reduce support material removal issues.
+%Additionally, as mentioned there are opportunities to improve the algorithm for neon sign routing, as ours does not guarantee a shortest-path routing.
+
+The basic approach of using a discrete rod for physical simulation does have various limitations, in particular it
+cannot recover if the rod becomes tangled or unstable. 
+However we have observed these problems only when attempting to break
+the system; they did not occur in any of the cases shown in our figures.
+Another issue is parameters, as the penalty forces used in PBD do not
+correspond to physical units. We avoided serious issues by uniformly
+scaling the initial mesh to a unit box, and tuning the parameters on several test cases.
+
+We are also eager to test our tool with users and get feedback from makers who use it in the wild.
+However, based on our own experiences, we believe our system could be readily utilized by makers and researchers, and serve as a useful tool for rapdily proptotyping interactive 3D objects.

### 2014-04-15 paper/exampleobjects.tex

--- 2014-04-14 paper/exampleobjects.tex
@@
 The pipes are threaded with electroluminescent wire which is lit in sequence, using a sequencer, to create the animation.
+
+\subsection{Speculative Objects and Explorations}
+
+\begin{figure}
+\centering
+    \includegraphics[width=3.4in]{figures/speculative.png}
+\caption{Some possible pipe uses.
+The printed whistle in (a) is a Helmholz resonator: it creates a sound when air is blown through the pipe at the front, because air compresses at the small neck joint between the internal chamber and the pipe.
+In (b), a capacitor manufactured on a 3D printer coated post-print with copper paint.
+This particular one holds an impractically large charge of $>1F$.
+(c) gives a comparison of pipe-based resistors, also filled with copper paint.
+For a cylindrical wire, resistance is based on its length and cross-sectional area.}
+\label{fig:speculative}
+\end{figure}
+
+\begin{figure}
+\centering
+    \includegraphics[width=3.4in]{figures/gates.png}
+\caption{By attaching a servo to a gate structure, it is possible to direct fluid through that pipe.  The gate not only has binary positions (a) and (b), but can direct fractional flow, as well (c).}
+\label{fig:direct}
+\end{figure}
+
+In addition to our six complete examples, we also created some smaller explorations of techniques and possibilities using pipes.
+
+Creating aural outputs using the printed pipes directly is one such possibility: the whistle in Figure \ref{fig:speculative}a is a fully printable audio output which is based on a Helmholz resonator, and which can be activated with air pressure.
+
+Some passive components required in the design of the above circuitry could in the future be created using pipes and printed or injected conductive material.
+For example, we printed a proof-of-concept  capacitor which features a thin printed plate with hollow cones on either side (see Figure \ref{fig:speculative}b).
+After printing, we coated the inner surfaces of capacitor with copper paint.
+It now holds an impractically large charge of $>1F$, but miniturization is possible.
+Similarly, tubes can be used to create \emph{resistors} in ``printed'' circuitry: a longer or thinner tube has a greater resistance than a shorter or wider one (see Figure \ref{fig:speculative}c).
+
+Liquids can be mixed in specific concentrations (e.g., for color output) using mixing or splitting topologies along with a servo-controlled gate structure as in Figure \ref{fig:direct}.
+
+Identification of objects is possible using active acoustic sensing \cite{Ono-touchandactivate}: the internal pipes, although not externally visually distinctive, affect the acoustic signature of the object.
+While this is an opportunity for record and recall identification, intentional identity encoding using pipes could also be accomplished, through print weight modification using fully enclosed pipes.

### 2014-04-15 paper/tubes.tex
minor updates and a new photo

@@
 Our technique for routing this path is inspired by work in generating continuous line illustrations from images~\cite{Wong-continuousline,Bosch-tsp}.
-Our routing problem is a version of the Chinese Postman Problem\footnote{The CPP is also known as the route inspection problem: \url{http://en.wikipedia.org/wiki/Route_inspection_problem}}, in which we wish to traverse all edges of the user's input graph, much as a postman needs to walk along every road at least once to deliver mail.
-In our relaxation, we allow creating new edges (i.e., the postman may jump from building to building in addition to using existing roads).
+Our routing problem is a version of the Chinese Postman Problem\footnote{The CPP is also known as the route inspection problem: \url{http://en.wikipedia.org/wiki/Route_inspection_problem}}, in which we wish to traverse all edges of the graph described by the user's input, much as a postman needs to walk along every road at least once to deliver mail.
+In our relaxation, we allow the creation of new edges (i.e., the postman may jump from building to building in addition to using existing roads).
 %
@@
 The physical fabrication process used by different machines gives rise to differing strategies for avoiding support material or easing its removal.
-We next discuss techniques for fused deposition modeling (FDM) machines, such as the Makerbot\footnote{\url{http://makerbot.com}} and multi-jetting machines such as Objet\footnote{\url{http://stratasys.com}}, as well as powder-based (e.g., ZCorp\footnote{\url{http://3dsystems.com}}) and stereolithography (e.g., Form 1\footnote{\url{http://formlabs.com}}).
+We next discuss techniques for fused deposition modeling (FDM) machines, such as the Makerbot\footnote{\url{http://makerbot.com}} and multi-jetting machines such as Objet\footnote{\url{http://stratasys.com}}, as well as powder-based machines (e.g., ZCorp\footnote{\url{http://3dsystems.com}}) and those that use stereolithography (e.g., Form 1\footnote{\url{http://formlabs.com}}).
 

### 2014-04-15 paper/abstract.tex
typographical fun

@@
 \begin{abstract}
-3D printers offer extraordinary flexibilty for prototyping the shape and mechanical function of objects.
+3D printers offer extraordinary flexibility for prototyping the shape and mechanical function of objects.
 We investigate how 3D models can be modified to facilitate the creation of \emph{interactive} objects that offer dynamic input and output.
@@
 We introduce a general technique for supporting the rapid prototyping of interactivity by removing interior material from 3D models to form pipes and chambers.
-We describe this new design space of pipes and hollow chambers for interaction design, where variables include openings, topologies, path constraints, and inserted media.
+We describe this new design space of pipes for interaction design, where variables include openings, path constraints, topologies, and inserted media.
 We then present \systemnamenospace, a tool for routing such pipes through the interior of 3D models, integrated within a 3D modeling program.

### 2014-04-15 paper/designspace.tex
typographical fun

@@
 \section{The Design Space of Pipes}
-Our review of prior art indicates the potential benifits of using interior pipes for adding interactivity to 3D printed objects, although no general purpose tools currently exist for this purpose.
-To guide the design of such tool, we now provide a design space of the types of pipes that a maker may want to utilize.
+Our review of prior art indicates the potential benifits of using interior pipes for adding interactivity to 3D printed objects, although no general-purpose tools currently exist for this purpose.
+To guide the design of such a tool, we now provide a design space of the types of pipes that a maker may want to utilize.
 We have identified four relevant dimensions of pipe design: \emph{openings}, \emph{path constraints}, \emph{topologies}, and inserted \emph{media}.
@@
 \subsection{Openings}
-Pipes inside solid objects can be either \emph{open} to the outside on both ends, \emph{semi-enclosed}, or \emph{fully enclosed} (see Figure \ref{fig:pipespace}).
-For interactive devices, openings can be either user-facing (in places a user would for example touch a model); or system-facing, where sensors and actuators may be connected.
-%\george{difference between open/return not clear}
-%Pipes can have four opening styles: open, return, semi-closed, and fully enclosed .
-As we are interested in interactive devices, we distinguish a ``user side"---part of the surface or interior of an object facing its user for interaction; and a ``system side"---where other hardware such as electronics may be connected.
-We refer to a pipe which is ``open on the user side'' to have its contents \emph{physically accessible} to a user \emph{at interaction time}.
-``Closed on the user side'' implies physically cut off from the user.
-Similarly with the system side.
-User or system manipulation of pipe contents are possible even without physical access: e.g., via soundwaves.
-\george{confusing}
+Pipes inside solid objects can be either \emph{open} to the outside on both ends, \emph{semi-closed}, or \emph{fully enclosed} (see Figure \ref{fig:pipespace}).
+For interactive devices, openings can be either user-facing, where a user would, for example, touch a model; or system-facing, where sensors and actuators may be connected.
 
@@
 %
-\emph{Semi-enclosed} pipes are capped on one side and could be used to create tactile output.
+\emph{Semi-closed} pipes are capped on one side and could be used to create tactile output.
 By fabricating a cap from a malleable material and attaching a pump to the open end, haptic feedback can be generated by raising and depressing the flexible cap (see Figure \ref{fig:openings}b).
@@
 An \emph{endpoint-constrained} pipe requires specific start and end points, but the particular route of the internal path is irrelevant.
-This is the case for circuit schematics, where the electronic components have a fixed location, but many routing solutions are valid.
+This is the case for circuit schematics, where the electronic components have fixed locations, but many routing solutions are valid.
 In contrast, for a \emph{path-constrained} pipe the shape of the path is important.
 For example, threading electroluminescent (EL) wire through a clear pipe enables makers to emulate neon art (see Figure \ref{fig:teaser}, right).
-A \emph{fully constrained} pipe requires both specific terminals as well as a specific path.
+A \emph{fully constrained} pipe requires specific terminals as well as a specific path.
 If a maker wishes the EL wire to exit the rear of the sign to hide the electronics, she can use a fully-constrained path (Figure \ref{fig:neon}).
@@
 Pipe network topologies enable \emph{splitting} or \emph{mixing} of media (see Figure \ref{fig:pipespace}).
-\emph{Star} and \emph{tree} topologies extend the splitting and mixing primitives, encompassing multiple-in and multiple-out configurations; we use star topologies to create different sensing locations for our touch-sensitive toys, where one terminal is used for swept-frequency capacitive sensing (Figure~\ref{fig:examples}a).
+\emph{Star} and \emph{tree} topologies extend the splitting and mixing primitives, encompassing multiple-in and multiple-out configurations; we use star topologies to create different sensing locations for our touch-sensitive toys, where one terminal is used for swept-frequency capacitive sensing (Figure~\ref{fig:teaser}, center).
 
@@
 %
-Compressible fluids (\emph{``gas''}) and incompressible fluids (\emph{``liquids''})  inside semi-enclosed pipes can create haptic feedback; pressure can also be sensed to act as an input~\cite{Slyper-shape}.
+Compressible fluids (\emph{``gas''}) and incompressible fluids (\emph{``liquids''})  inside semi-closed pipes can create haptic feedback; pressure can also be sensed to act as an input~\cite{Slyper-shape}.
 Gases can be used as carriers for scents or fog.
@@
 Filling pipes with \emph{solids} at print time enables designers to exploit the difference in material properties between pipe and contents.
-For example, in Printed Optics~\cite{Willis-printedoptics}, different refractive indices between solid cladding material and transparent material inside pipes allowed light transport along pipes for input and output.
+For example, in Printed Optics~\cite{Willis-printedoptics}, different refractive indices between solid cladding material and transparent material inside pipes allow light transport along pipes for input and output.
 %

### 2014-04-15 paper/introduction.tex
typographical fun

@@
 %and maker communities like Thingiverse\footnote{http://thingiverse.com}, recent work has looked at the issue of rapidly prototyping interactive objects. 
-Existing approaches include Printed Optics \cite{Willis-printedoptics}, which print light pipes using optically clear material, and Sauron \cite{Savage-sauron}, which uses computer vision to track the movment of components on the inside of a hollow object.
+Existing approaches include Printed Optics \cite{Willis-printedoptics}, which uses light pipes printed in optically clear material, and Sauron \cite{Savage-sauron}, which uses computer vision to track the movement of components on the inside of a hollow object.
 These systems exclusively focus on optical sensing and visual displays --- we seek to grow the vocabulary of supported tangible input and output modalities.
@@
 
-We describe a new design space of pipes and hollow chambers for the purpose of interaction design, where variables include openings, path constraints, topologies, and inserted media.
+We describe a new design space of pipes for the purpose of interaction design, where variables include openings, path constraints, topologies, and inserted media.
 By exploring this design space new opportunities for adding interactivity to 3D printed objects can be realized.
@@
 Figure \ref{fig:teaser} shows a collection of interactive objects with various functionalities including haptic feedback, neon sign-like visual output, and selective touch-sensitivity.
-All these objects have had the electronic components required to function redirected onto their surfaces using pipes.
+All these objects have had the electronic components required to function redirected using pipes.
 But generating the appropriate pipes for these examples is not trivial.
@@
 Pipes should not pass too close to the surface: consumer-grade 3D printers do not guarantee air- or water-tightness of prints, so more than one printed layer may be required to prevent leakage.
-Pipes for applications like illumunated EL wire signs must allow a single pipe to follow a specific path layout (e.g., for spelling out text).
+Pipes for applications like illuminated EL wire signs must allow a single pipe to follow a specific path layout (e.g., for spelling out text).
 
-We contribute PipeDream, an interactive design tool that offers 3D point-to-point routing of internal pipes based on a physical simulation of flexible rods.
+We contribute PipeDream, an interactive design tool that offers 3D point-to-point routing of internal pipes based on physical simulation of flexible rods.
 Our system addresses the various challenges that makers are likely to encounter, such as avoiding intersections and ensuring maximum bend angles.

### 2014-04-15 paper/tubes.tex
typographical fun

@@
 The interior of these pipes should have as large a bending radius as possible so that post-print insertion of solid or viscous media is easier.
-To create endpoint-constrained pipes, in which the location and shape of exterior connection points matters, a brush tool is used to specify a pair of points on the mesh's surface.
+To create endpoint-constrained pipes, in which the location of exterior connection points matters, a brush tool is used to specify a pair of points on the mesh's surface.
 The system automatically routes the pipe, and displays a preview.
@@
 Our voxel grid currently uses a resolution of 128x128x128 voxels, which enables interactive operations, but this parameter can be adjusted for different model requirements.
-The simulation takes about a second for most of our examples.
+The simulation takes about a second for our examples.
 
@@
 We find an Euler tour on the modified graph, and thicken the path to create pipes.
-Our technique for routing this path is inspired by work in generating continuous line illustrations from images~\cite{Wong-continuousline,Bosch-tsp}.
+Our technique for routing this path is inspired by work in generating continuous line illustrations from images~\cite{Bosch-tsp,Wong-continuousline}.
 Our routing problem is a version of the Chinese Postman Problem\footnote{The CPP is also known as the route inspection problem: \url{http://en.wikipedia.org/wiki/Route_inspection_problem}}, in which we wish to traverse all edges of the graph described by the user's input, much as a postman needs to walk along every road at least once to deliver mail.
@@
 
-{\bf Euler Tour:} To find an Euler tour, we use Fleury's algorithm\footnote{\url{http://en.wikipedia.org/wiki/Eulerian_path\#Fleury.27s_algorithm}}, with a modification to preserve continuity: at a given node, from multiple candidate edges, we prefer the edge(s) with the smallest angular deviation from the incoming edge (i.e., we prefer to pass straight through a node, if possible).
+{\bf Euler Tour:} To find an Euler tour, we use Fleury's algorithm\footnote{\url{http://en.wikipedia.org/wiki/Eulerian_path\#Fleury.27s_algorithm}} with a modification to preserve continuity: at a given node, from multiple candidate edges, we prefer the edge(s) with the smallest angular deviation from the incoming edge (i.e., we prefer to pass straight through a node, if possible).
 This minimizes turns in the final artifact, which eases support material removal and assembly.
@@
 In (a), a pipe end is placed on the surface of the mesh.
-In (b), we create a sphere whose radius is slightly larger than the radius of the pipe, centred at the tube's endpoint.
+In (b), we create a sphere whose radius is slightly larger than the radius of the pipe, centred at the pipe's endpoint.
 In (c), we consider the intersection of the sphere and the pipe, and create an offset surface of the mesh.
@@
 Complex pipe geometries can make water blasting impossible and other processes more labor- and time-intensive.
-All models in Figure \ref{fig:teaser}b required support material removal post-print: the neon sign and maze were additionally cut into two pieces to give better access to the support.
+All black and clear models in Figure \ref{fig:teaser} required support removal post-print: the neon sign and maze were additionally cut into two pieces to give better access to the support.
 
@@
 The remaining powder or liquid serves as support for the model.
-These types of printers are perhaps even more suitable for pipe creation than either FDM or multi-jetting printers, as the supporting liquid or loose powder is simpler to remove than structural plastic or UV cured resin.
+These types of printers are perhaps even more suitable for pipe creation than either FDM or multi-jetting printers, as the supporting liquid or loose powder is simpler to remove than structural plastic or UV cured material.
 We hope to experiment with them in the future.

### 2014-04-16 paper/conclusion.tex

--- 2014-04-11 paper/conclusion.tex
@@
 This amplifies and extends the utility of existing sensing and actuation approaches.
-In addition to describing the design space of pipes, we also presented \systemname, a design tool for the creation of tubes inside arbitrary 3D models, and discussed many techniques that combine nicely wtih pipes.
+In addition to describing the design space of pipes, we also presented \systemnamenospace, a design tool for the creation of tubes inside arbitrary 3D models, and discussed many techniques that combine nicely with pipes.
 We fabricated several example objects using our tool to explore new points in the design space.

### 2014-04-16 paper/exampleobjects.tex

@@
 
-The models we used for the boat was downloaded from Thingiverse, while the brain model came from researchers at one of our instutions.
+The model we used for the boat was downloaded from Thingiverse, while the brain model came from researchers at one of our instutions.
 We then used \systemname to create the desired internal pipes.
@@
 This paint requires approximately 2 days to fully dry in this configuration (3mm diameter tubes, maximum tube length 10cm).
-The resistance of the paint in this configuation is approximately $2\Omega / inch.$  Our smart base is powered by an Arduino Uno running open-source SFCS code\footnote{\url{http://www.instructables.com/id/Touche-for-Arduino-Advanced-touch-sensing}}.
+The resistance of the paint with this cross-section is approximately $2\Omega / inch.$  Our smart base is powered by an Arduino Uno running open-source SFCS code\footnote{\url{http://www.instructables.com/id/Touche-for-Arduino-Advanced-touch-sensing}}.
 
@@
 
-The radio shell was modelled in SolidWorks.
+The radio shell was modelled in a commercial CAD tool.
 We included slots on the 3D model matching the shape of the electronic components that were to be embedded.
@@
 
-The penholder body was created in Solidworks.
-We used \systemname to add four internal open pipes connecting each pen chamber to the based on the pen holder.
+The penholder body was created in a commercial CAD tool.
+We used \systemname to add four internal open pipes connecting each pen chamber to the base of the pen holder.
 This prototype was built on our Makerbot support-free.
@@
 The final artifact was fabricated in two halves fastened together via glue, to allow removal of support material with a high-pressure water tool.
-We plugged the ends of the tubes post-print with additional material to ensure the ball does not escape.
+We plugged the ends of the pipes post-print with additional material to ensure the ball does not escape.
 
@@
 The external geometry of the model was imported into \systemnamenospace, which created the associated pipes using its internal path routing tool.
-We added additional endpoint-constrained tubes as exits out the rear of the model to allow us to hide the controls (see Figure \ref{fig:neon}B).
+We added additional endpoint-constrained tubes as exits out the rear of the model to allow us to hide the controls (see Figure \ref{fig:neon}b).
 We then printed the model on the Objet using transparent material.
@@
 
-Creating aural outputs using the printed pipes directly is one such possibility: the whistle in Figure \ref{fig:speculative}a is a fully printable audio output which is based on a Helmholz resonator, and which can be activated with air pressure.
+Creating aural outputs using printed pipes is one such possibility: the whistle in Figure \ref{fig:speculative}a is a fully printable audio output based on a Helmholz resonator, which can be activated with high pressure air.
+Zoran \cite{Zoran-flute} previously explored printable musical instruments.
 
-Some passive components required in the design of the above circuitry could in the future be created using pipes and printed or injected conductive material.
+Some passive components required in the design of circuitry could in the future be created using pipes and printed or injected conductive material.
 For example, we printed a proof-of-concept capacitor which features a thin printed plate with hollow cones on either side (see Figure \ref{fig:speculative}b).
 After printing, we coated the inner surfaces of capacitor with copper paint.
-It now holds an impractically large charge of $>1F$, but miniturization is possible.
+It now holds an impractically large charge of $>1F$, but miniaturization is possible.
 Similarly, tubes can be used to create \emph{resistors} in ``printed'' circuitry: a longer or thinner tube has a greater resistance than a shorter or wider one (see Figure \ref{fig:speculative}c).

### 2014-04-16 paper/introduction.tex

@@
 
-Figure \ref{fig:teaser} shows a collection of interactive objects with various functionalities including haptic feedback, neon sign-like visual output, and selective touch-sensitivity.
-All these objects have had the electronic components required to function redirected using pipes.
-But generating the appropriate pipes for these examples is not trivial.
-For one, most 3D modelling tools do not provide extensive tools for modelling the  {\em interior} of 3D models.
-Furthermore, careful consideration needs to go into the design and routing of the pipes.
+Figure \ref{fig:teaser} shows a collection of interactive objects fabricated with the help of internal pipes.
+These objects can sense touch, detect object presence, generate haptic feedback, or show internal illumination.
+They can also contain other active and passive electronic components.
+But manually modeling appropriate pipes for such functionality is not trivial.
+For one, many 3D modelling tools do not provide extensive tools for modelling the  {\em interior} of 3D models.
+Furthermore, careful consideration needs to go into the design and routing of pipes when they are used to add interactivity.
 Pipes need maximum possible bend radius to ease the insertion of media post-print.
@@
 
-We contribute PipeDream, an interactive design tool that offers 3D point-to-point routing of internal pipes based on physical simulation of flexible rods.
+We contribute PipeDream, an interactive design tool for interior pipes and cavities.
 Our system addresses the various challenges that makers are likely to encounter, such as avoiding intersections and ensuring maximum bend angles.
-We then present a set of example objects, designed with the help of our modeling tool, that demonstrate different interaction modalities.
+PipeDream offers 3D point-to-point routing of internal pipes based on physical simulation of flexible rods, and a graph algorithm-based path tracing technique to create shape-constrained internal paths.
+We present a set of example objects, designed with the help of our modeling tool, that demonstrate different interaction modalities.
 We conclude by discussing limitations and outlining areas for future exploration.

### 2014-04-16 paper/relatedwork.tex

@@
 
-Sauron tracks interactions with printed physical mechanisms by placing a camera inside hollow objects, where all components must be visible to the camera~\cite{Savage-sauron}.
+Sauron~\cite{Savage-sauron} tracks interactions with printed physical mechanisms by placing a camera inside hollow objects, where all components must be visible to the camera.
 Our technique instead can work with arbitrary, solid models and only carves out pipes for interaction where needed.
-Midas generates fabrication files for custom capacitive touch sensors~\cite{Savage-midas}.
+Midas~\cite{Savage-midas} generates fabrication files for custom capacitive touch sensors.
 Their technique is limited to planar and unfoldable shapes as it routes connections on the surface of objects.
@@
 Sells, et al., \cite{Sells-reprap} use a heated syringe to deposit metal in printed 2D circuitboard channels on objects' surfaces.
-Other experiments have been done by the RepRap project\footnote{\url{http://reprap.org}} using plastic filament blended with conductive materials, however the reistance is too high to be useful in circuitry.
+Other experiments have been done by the RepRap project\footnote{\url{http://reprap.org}} using plastic filament blended with conductive materials, however the resistance is too high to be useful in circuitry.
 Sarik, et al., \cite{Sarik-tracebrush} use aerosol jetting of conductive material onto printed objects.
@@
 Multi-layer boards consist of stacked 2D layers connected by vias.
-We compute true 3D point to point routing though the interior of solid models with flexible paths.
+We compute true 3D point-to-point routing though the interior of solid models with flexible paths.
 This approach has been used in one-off designs~\cite{Navarrette-gps} but we are not aware of design tools for this purpose.
@@
 %Takayama, et al., built a system described in \cite{Takayama-fruit} which allowed users to create internal color maps of the interior of objects, allowing them to see digital ``slices'' of their ``oranges.'' \bjoern{don't understand this} Our work focuses on 3D printable geometry rather than rendering.
-In Make it Stand, material is removed from the interior of objects with the goal of changing their center of gravity \cite{Prevost-makeitstand}.
-Infrastructs encodes information using interior cavities that can be sensed using terahertz imaging~\cite{Willis-infrastructs}.
-Bickel, et al., explored enforcing deformation behavior with multi-material blending inside objects\cite{Bickel-deformation}.
+In Make it Stand~ \cite{Prevost-makeitstand}, material is removed from the interior of objects with the goal of changing their center of gravity.
+Infrastructs~\cite{Willis-infrastructs} encodes information using interior cavities that can be sensed using terahertz imaging.
+Bickel, et al.,~\cite{Bickel-deformation} explored enforcing deformation behavior with multi-material blending inside objects.
 Modeling internal chambers is one aspect of our work, however our focus is on interaction rather than balance, identification, or deformation.

### 2014-04-16 paper/abstract.tex
working through tovi suggestions

@@
 %However, a 3D printer's role in prototyping interactive objects is still not clear.  
-We introduce a general technique for supporting the rapid prototyping of interactivity by removing interior material from 3D models to form pipes and chambers.
+We introduce a general technique for supporting the rapid prototyping of interactivity by removing interior material from 3D models to form internal pipes.
 We describe this new design space of pipes for interaction design, where variables include openings, path constraints, topologies, and inserted media.

### 2014-04-16 paper/designspace.tex
working through tovi suggestions

@@
     \includegraphics[width=1.0\columnwidth]{figures/types.png}
-\caption{Pipe openings: A shows a pair of \emph{open} pipes filled with copper paint to power an LED.
-B shows two \emph{semi-closed} pipes capped by rubber membranes.
-In C, a rabbit has a \emph{fully enclosed} chamber to modify its weight.}
+\caption{Pipe openings: a) shows a pair of \emph{open} pipes filled with copper paint to power an LED.
+b) shows two \emph{semi-closed} pipes capped by rubber membranes.
+In c), a rabbit has a \emph{fully enclosed} chamber to modify its weight.}
 \label{fig:openings}
@@
 \emph{Open} pipes may be used to create electronic circuitry.
-For example, an open pipe filled with conductive paint and connected to a battery can provide power to an LED (see Figure \ref{fig:openings}a).
+For example, an open pipe filled with conductive paint and connected to a battery can provide power to an LED (Figure \ref{fig:openings}a).
 %
@@
 
+\begin{figure}
+\centering
+    \includegraphics[width=1.0\columnwidth]{figures/glamour/return.jpg}
+\caption{A path-constrained open pipe with EL wire as a ``neon'' design.}
+\label{fig:justneon}
+\end{figure}
+
 \subsection{Path Constraints}
@@
 In contrast, for a \emph{path-constrained} pipe the shape of the path is important.
-For example, threading electroluminescent (EL) wire through a clear pipe enables makers to emulate neon art (see Figure \ref{fig:teaser}, right).
+For example, threading electroluminescent (EL) wire through a clear pipe enables makers to emulate neon art (Figure \ref{fig:justneon}).
 A \emph{fully constrained} pipe requires specific terminals as well as a specific path.
@@
 Finally, an \emph{unconstrained} pipe does not need a specific path or terminal points.
-A cavity to modify an object's weight is an example of an unconstrained pipe.
+A cavity to modify an object's weight is an example of an unconstrained pipe (Figure \ref{fig:openings}c).
 

### 2014-04-16 paper/exampleobjects.tex
working through tovi suggestions

@@
 
-To expand Touch\'{e} \cite{Sato-touche} to 3D printed forms, we created a set of touch-sensitive toys and a companion app that identifies the objects as well as touched parts of the objects (inspired by \cite{Harrison-acoustic}).
+To expand Touch\'{e} \cite{Sato-touche} to 3D printed forms, we created a set of touch-sensitive toys and a companion app that identifies the objects as well as touched parts of the objects (inspired by the boat application in Acoustic Barcodes~\cite{Harrison-acoustic}).
 The objects --- a brain and boat, in our example --- can be set on a base, upon which the computer displays ``brain'' or ``boat''.
@@
 We connected one pipe to each of our pump's terminals, and using a programmable power supply we mimic a rabbit's breathing pattern.
-This example was printed on the Objet using flexible material; support was flushed post-print.
+This example was printed on the Objet using flexible material (100\% TangoBlackPlus); support was flushed post-print.
 
@@
 At the base of each pipe is a QRE1113 line sensor breakout board, with integrated IR emitter and receiver.
-When a pen is in its appointed place, the emitted infrared light is reflected off its base and travels back to the receiver, where it registers as bright.
+When a pen is in its appointed place, the emitted infrared light is reflected off its base and travels back to the receiver, where it appears bright.
 
@@
 The maze is based on an SVG file we manually created and processed with our internal path routing tool.
-We printed the maze on the Objet using a transparent material for the top and a black material for the bottom.
+We printed the maze on the Objet using a transparent material (100\% VeroClear) for the top and a black material (100\% TangoBlackPlus) for the bottom.
 The final artifact was fabricated in two halves fastened together via glue, to allow removal of support material with a high-pressure water tool.
@@
 The printed whistle in (a) is a Helmholz resonator: it creates a sound when air is blown through the pipe at the front, because air compresses at the small neck joint between the internal chamber and the pipe.
-In (b), a capacitor manufactured on a 3D printer coated post-print with copper paint.
-This particular one holds an impractically large charge of $>1F$.
-(c) gives a comparison of pipe-based resistors, also filled with copper paint.
+(b) gives a comparison of pipe-based resistors, filled with copper paint.
 For a cylindrical wire, resistance is based on its length and cross-sectional area.}
@@
 
-In addition to our six complete examples, we also created some smaller explorations of techniques and possibilities using pipes.
+Our six examples showcase how PipeDream can be used to design objects that support a wide range of input and output modalities.
+In addition to these six complete examples, we also created some smaller explorations of techniques and possibilities using pipes.
 
 Creating aural outputs using printed pipes is one such possibility: the whistle in Figure \ref{fig:speculative}a is a fully printable audio output based on a Helmholz resonator, which can be activated with high pressure air.
-Zoran \cite{Zoran-flute} previously explored printable musical instruments.
+Zoran \cite{Zoran-flute} previously explored printable musical instruments, and it would be interesting to extend PipeDream to aid in the design process of such instruments.
 
 Some passive components required in the design of circuitry could in the future be created using pipes and printed or injected conductive material.
-For example, we printed a proof-of-concept capacitor which features a thin printed plate with hollow cones on either side (see Figure \ref{fig:speculative}b).
-After printing, we coated the inner surfaces of capacitor with copper paint.
-It now holds an impractically large charge of $>1F$, but miniaturization is possible.
-Similarly, tubes can be used to create \emph{resistors} in ``printed'' circuitry: a longer or thinner tube has a greater resistance than a shorter or wider one (see Figure \ref{fig:speculative}c).
+For example, tubes can be used to create \emph{resistors} in ``printed'' circuitry: a longer or thinner tube has a greater resistance than a shorter or wider one (see Figure \ref{fig:speculative}b).
 
 Liquids can be mixed in specific concentrations (e.g., for color output) using mixing or splitting topologies along with a servo-controlled gate structure as in Figure \ref{fig:direct}.
+This could be used in creating interactive science exhibits about, e.g., logic gates.
 

### 2014-04-16 paper/introduction.tex
working through tovi suggestions

--- 2014-04-15 paper/introduction.tex
@@
 Makers, researchers, and professional designers (jointly ``makers" for the remainder of this paper) increasingly leverage 3D printers as tools for design work.
-A wide array of objects, ranging from toy figurines to video game controllers and jet engine blades, are now prototyped (and sometimes even manufactured) using these machines.
+A wide array of objects, ranging from toy figurines to video game controllers and jet engine blades, are now prototyped, and sometimes even manufactured, using these machines.
 Most models fabricated on 3D printers today are passive: they express the form, but not the interactivity, of an object.
@@
 We describe a new design space of pipes for the purpose of interaction design, where variables include openings, path constraints, topologies, and inserted media.
-By exploring this design space new opportunities for adding interactivity to 3D printed objects can be realized.
+By exploring this design space, new opportunities for adding interactivity to 3D printed objects can be realized.
 For example, injecting conductive paint can turn pipes into conductors, which can in turn be used to integrate electronic components into prints.
 Electroluminescent (EL) wire can be threaded through pipes to create computer-controlled visual output.
-And air can be pumped through pipes to produce haptic effects.
+Air can be pumped through pipes to produce haptic effects.
 Existing sensing and actuation approaches, such as FlyEye \cite{Wimmer-flyeye}, Swept Frequency Capacitive Sensing \cite{Sato-touche}, and Jamming User Interfaces \cite{Follmer-jamming}, can be expanded to new contexts by the use of pipes: makers utilizing such I/O strategies can locate input and output at arbitrary points on a printed 3D object's surface.
@@
 All these objects have had the electronic components required to function redirected using pipes.
-But generating the appropriate pipes for these examples is not trivial.
+However, generating the appropriate pipes for these examples is not trivial.
 For one, most 3D modelling tools do not provide extensive tools for modelling the  {\em interior} of 3D models.
@@
 Our system addresses the various challenges that makers are likely to encounter, such as avoiding intersections and ensuring maximum bend angles.
-We then present a set of example objects, designed with the help of our modeling tool, that demonstrate different interaction modalities.
+We then present a set of example objects, designed with the help of our modelling tool, that demonstrate different interaction modalities.
 We conclude by discussing limitations and outlining areas for future exploration.

### 2014-04-16 paper/relatedwork.tex
working through tovi suggestions

@@
 Sells, et al., \cite{Sells-reprap} use a heated syringe to deposit metal in printed 2D circuitboard channels on objects' surfaces.
-Other experiments have been done by the RepRap project\footnote{\url{http://reprap.org}} using plastic filament blended with conductive materials, however the resistance is too high to be useful in circuitry.
+Other experiments have been done by the RepRap project\footnote{\url{http://reprap.org}} using plastic filament blended with conductive materials, however the resistance is currently too high to be useful in circuitry.
 Sarik, et al., \cite{Sarik-tracebrush} use aerosol jetting of conductive material onto printed objects.

### 2014-04-16 paper/tubes.tex
working through tovi suggestions

@@
 \systemname allows users to design pipes in two ways.
-In one, they can select exterior anchor points on their objects; the tool then creates complete point-to-point routings using A* path search and physical rod simulation (see Figure \ref{fig:tool-process-exterior}).
-In the other, users can import vector art describing desired interior paths (see Figure \ref{fig:tool-process-interior}).
+In one, they can select exterior anchor points on their objects; the tool then creates complete point-to-point routings using A* path search and physical rod simulation (Figure \ref{fig:tool-process-exterior}).
+In the other, users can import vector art describing desired interior paths (Figure \ref{fig:tool-process-interior}).
 \systemname then creates a single path that follows the input shape using edge graph manipulation and Euler tour generation.
@@
 \subsubsection{Pipes with Multiple Endpoints}
-To allow designers to create pipes with multiple endpoints (e.g., the star topologies in Figure \ref{fig:examples}a), we use one surface position per pipe and a central fixed endpoint that the user can interactively position (see Figure~\ref{fig:boat-star}).
+To allow designers to create pipes with multiple endpoints (e.g., the star topologies as in Figure \ref{fig:designspace}), we use one surface position per pipe and a central fixed endpoint that the user can interactively position (Figure~\ref{fig:boat-star}).
 The user can connect several pipes, hitting a key to add each in sequence and positioning them with the mouse.
@@
 %If path components are disconnected, we create edges that connect them; additionally we can create edges connecting odd-degree vertices rather than simply retracing existing edges.
-In the final artifact, all created edges will be blocked out by dark material so the inserted medium is not visible (see Figure \ref{fig:tool-process-interior}).
+In the final artifact, all created edges will be blocked out by dark material so the inserted medium is not visible (Figure \ref{fig:tool-process-interior}).
 %We offer a short algorithm here, with a more mathematically precise definition and associated proof in supplemental materials appendix.
@@
 of the initial mesh, intersecting with a sphere placed at the tube endpoint, and then performing
-a boolean union with the initial shape-minus-tube result (see Figure \ref{fig:cap}).
+a boolean union with the initial shape-minus-tube result (Figure \ref{fig:cap}).
 A straightforward extension would be to add or subtract additional elements at
@@
 The remaining powder or liquid serves as support for the model.
-These types of printers are perhaps even more suitable for pipe creation than either FDM or multi-jetting printers, as the supporting liquid or loose powder is simpler to remove than structural plastic or UV cured material.
+These types of printers are perhaps even more suitable for pipe creation than either FDM or multi-jetting printers, as the supporting liquid or loose powder is easier to remove than structural plastic or UV cured material.
 We hope to experiment with them in the future.
@@
 \subsection{Model Assembly}
-For especially complex geometries, models can be ``cut up'' into multiple pieces that can be assembled to avoid support deposition or at least ease material removal.
-This is undesirable as it means more time spent in assembly post-print, and additionally a cut pipe may not be fluid-tight any longer.
-An automatic tool for this task might be beneficial, however it is outside the scope of this work.
+For especially complex geometries, models can be ``cut up'' into multiple pieces that can be assembled to avoid support deposition or at least ease material removal; it can also ease insertion of the desired medium.
+This is somewhat undesirable as it means more time spent in assembly post-print, and additionally a cut pipe may no longer be fluid-tight.
+A tool aiding in or automatically performing this cutting task might be beneficial, however it is outside the scope of this work.

### 2014-04-16 paper/introduction.tex

--- 2014-04-16 paper/introduction.tex
@@
 These objects can sense touch, detect object presence, generate haptic feedback, or show internal illumination.
-They can also contain other active and passive electronic components.
-But manually modeling appropriate pipes for such functionality is not trivial.
+They could also contain other active and passive electronic components.
+However, manually modeling appropriate pipes for such functionality is not trivial.
 For one, many 3D modelling tools do not provide extensive tools for modelling the  {\em interior} of 3D models.
@@
 
-We contribute PipeDream, an interactive design tool for interior pipes and cavities.
+We contribute PipeDream, an interactive design tool for interior pipes.
 Our system addresses the various challenges that makers are likely to encounter, such as avoiding intersections and ensuring maximum bend angles.

### 2014-04-16 paper/designspace.tex
radius change figures, remaining tovi comments

@@
 
-\begin{figure}
-\centering
-    \includegraphics[width=1.0\columnwidth]{figures/glamour/return.jpg}
-\caption{A path-constrained open pipe with EL wire as a ``neon'' design.}
-\label{fig:justneon}
-\end{figure}
+%\begin{figure}
+%\centering
+%    \includegraphics[width=1.0\columnwidth]{figures/glamour/return.jpg}
+%\caption{A path-constrained open pipe with EL wire as a ``neon'' design.}
+%\label{fig:justneon}
+%\end{figure}
 
@@
 A \emph{fully constrained} pipe requires specific terminals as well as a specific path.
-If a maker wishes the EL wire to exit the rear of the sign to hide the electronics, she can use a fully-constrained path (Figure \ref{fig:neon}).
+If a maker wishes the EL wire to exit the rear of the sign to hide the electronics, she can use a fully-constrained path (Figure \ref{fig:teaser}, right).
 Finally, an \emph{unconstrained} pipe does not need a specific path or terminal points.

### 2014-04-16 paper/discussion.tex
radius change figures, remaining tovi comments

@@
 This suggests the need for a better general 3D routing algorithm with multiple traces.
+For integration with common electronic components, it would be possible to build a library of ``footprints'' which could be used for recessing them into printed objects' surfaces; such footprints could also be used for automatically generating connecting pipes based on an electrical connection graph, as in circuitboard routing tools like Eagle\footnote{\url{http://www.cadsoftusa.com}}.
 Future work should also include assisting makers in the fabrication process itself by automatically cutting objects to reduce support material removal issues.

### 2014-04-16 paper/introduction.tex
radius change figures, remaining tovi comments

@@
 By exploring this design space, new opportunities for adding interactivity to 3D printed objects can be realized.
-For example, injecting conductive paint can turn pipes into conductors, which can in turn be used to integrate electronic components into prints.
-Electroluminescent (EL) wire can be threaded through pipes to create computer-controlled visual output.
-Air can be pumped through pipes to produce haptic effects.
-Existing sensing and actuation approaches, such as FlyEye \cite{Wimmer-flyeye}, Swept Frequency Capacitive Sensing \cite{Sato-touche}, and Jamming User Interfaces \cite{Follmer-jamming}, can be expanded to new contexts by the use of pipes: makers utilizing such I/O strategies can locate input and output at arbitrary points on a printed 3D object's surface.
+For example, injecting conductive paint can turn pipes into conductors, which can in turn integrate electronic components into prints; electroluminescent (EL) wire can be threaded through pipes to create computer-controlled visual output; or air can be pumped through pipes to produce haptic effects.
+Existing sensing and actuation approaches, such as FlyEye \cite{Wimmer-flyeye}, Swept Frequency Capacitive Sensing \cite{Sato-touche}, and Jamming User Interfaces \cite{Follmer-jamming}, can be expanded to new contexts using pipes: makers utilizing such I/O strategies can locate input and output at arbitrary points on a printed 3D object's surface.
 We leverage these existing techniques for sensing and actuation while our work's novelty is in \emph{internal modeling} for 3D printing and the \emph{redirection} of I/O to or through arbitrary locations on a device.

### 2014-04-16 paper/conclusion.tex
submission to UIST

@@
 
+Pipe mediation opens new interactive device prototyping opportunities both for makers with consumer-grade 3D printers and research and industrial labs with higher-end machines.
 While today's 3D printers are not yet able to fabricate active components in-place, we suggest that 3D printed interactive devices can be created with redirection of active input and output via pipes.
@@
 We fabricated several example objects using our tool to explore new points in the design space.
-
-We are heartened by our success: it opens new interactive device prototyping opportunities both for makers with consumer-grade 3D printers and research and industrial labs with higher-end machines.
-We are excited to release our tool into the wild and see what people build.
+We look forward to extending our software contribution into future multi-material 3D modelling tools.

### 2014-04-16 paper/discussion.tex
submission to UIST

@@
 We are also eager to test our tool with users and get feedback from makers who use it in the wild.
-However, based on our own experiences, we believe our system could be readily utilized by makers and researchers, and serve as a useful tool for rapdily proptotyping interactive 3D objects.
+Based on our own experiences, we believe our system could be readily utilized by makers and researchers, and serve as a useful tool for rapdily proptotyping interactive 3D objects.

### 2014-04-16 paper/exampleobjects.tex
submission to UIST

@@
 To evaluate and highlight our tool's capabilities, we fabricated a set of six prototype objects designed using \systemnamenospace.
-To show that our approach is printer-independent, we fabricated some on a Makerbot Replicator 2 and the rest on an Objet Connex 260.
+To show that our approach is printer-independent, we fabricated some on a Makerbot Replicator 2 and the rest on a Stratasys Connex 260.
 Each prototype object explores the expansion of an existing sensing or actuation technique to the surface of 3D printed objects.
@@
 We connected one pipe to each of our pump's terminals, and using a programmable power supply we mimic a rabbit's breathing pattern.
-This example was printed on the Objet using flexible material (100\% TangoBlackPlus); support was flushed post-print.
+This example was printed on the Connex using flexible material (100\% TangoBlackPlus); support was flushed post-print.
 
@@
 We fabricated the radio body on our Makerbot, support-free but cut in two pieces to allow the Arduino and battery pack to be inserted.
-We then threaded the pipes with wires to connect the components to the microcontroller in the base.
+We then threaded the pipes with wires to connect the components to the microcontroller contained in the base.
 
@@
 The maze is based on an SVG file we manually created and processed with our internal path routing tool.
-We printed the maze on the Objet using a transparent material (100\% VeroClear) for the top and a black material (100\% TangoBlackPlus) for the bottom.
+We printed the maze on the Connex using a transparent material (100\% VeroClear) for the top and a black material (100\% TangoBlackPlus) for the bottom.
 The final artifact was fabricated in two halves fastened together via glue, to allow removal of support material with a high-pressure water tool.
@@
 We added additional endpoint-constrained tubes as exits out the rear of the model to allow us to hide the controls (see Figure \ref{fig:neon}b).
-We then printed the model on the Objet using transparent material.
+We then printed the model on the Connex using transparent material.
 The model was printed in two halves, cutting through the plane of the internal path, to facilitate assembly.

### 2014-04-16 paper/relatedwork.tex
submission to UIST

@@
 Previous work has begun to investigate how to fabricate interactive objects using particular sensing and display technologies.
-Printed Optics \cite{Willis-printedoptics} uses light pipes to create integrated displays and senses input optically through components added at print time, while Slyper et al., fabricate flexible robot armatures capable of sensing interaction via changes in air pressure \cite{Slyper-pressure}.
+Printed Optics \cite{Willis-printedoptics} uses light pipes to create integrated displays and senses input optically through components added at print time, while Slyper et al.,~\cite{Slyper-pressure} fabricate flexible robot armatures capable of sensing interaction via changes in air pressure.
 Our work is inspired by these projects, and contributes a general design tool for modelling objects that can have such capabilities.
@@
 
+\subsubsection{Interaction for Fabrication vs. Fabrication for Interaction}
+HCI researchers are also developing novel interaction techniques for providing input to digital fabrication machines~\cite{Mueller-constructable,mixfab,Willis-interactive}, and new hybrid hand tools that combine manual processes with digital fabrication~\cite{rivers2012position,zoran2013freed}.
+These projects are complementary as they target different aspects of the design process, though it is conceivable that our pipe design tools could be added to augmented reality modeling interfaces like MixFab~\cite{mixfab}.
+
 \subsection{Routing and Internal Modeling}
@@
 %Takayama, et al., built a system described in \cite{Takayama-fruit} which allowed users to create internal color maps of the interior of objects, allowing them to see digital ``slices'' of their ``oranges.'' \bjoern{don't understand this} Our work focuses on 3D printable geometry rather than rendering.
-In Make it Stand~ \cite{Prevost-makeitstand}, material is removed from the interior of objects with the goal of changing their center of gravity.
+In Make it Stand~\cite{Prevost-makeitstand}, material is removed from the interior of objects with the goal of changing their center of gravity.
 Infrastructs~\cite{Willis-infrastructs} encodes information using interior cavities that can be sensed using terahertz imaging.
 Bickel, et al.,~\cite{Bickel-deformation} explored enforcing deformation behavior with multi-material blending inside objects.
-Modeling internal chambers is one aspect of our work, however our focus is on interaction rather than balance, identification, or deformation.
+Modeling internal chambers is one aspect of our work, however our focus is interaction rather than balance, identification, or deformation.
 %B\"{a}cher, et al., built an algorithm for automatically placing 3D fabricatable joints into meshes \cite{Bacher-articulated}.
 All these contributions are orthogonal to ours, as our focus is on precisely \emph{removing} material from printed objects, rather than blending materials or modifying components.
-
-\subsubsection{Interaction for Fabrication vs. Fabrication for Interaction}
-HCI researchers are also developing novel interaction techniques for providing input to digital fabrication machines~\cite{Mueller-constructable,Willis-interactive,mixfab}, and new hybrid hand tools that combine manual processes with digital fabrication~\cite{zoran2013freed,rivers2012position}.
-These projects are complementary as they target different aspects of the design process, though it is conceivable that our pipe design tools could be added to augmented reality modeling interfaces like MixFab~\cite{mixfab}.

### 2014-04-16 paper/tubes.tex
submission to UIST

@@
 \systemname allows users to design pipes in two ways.
-In one, they can select exterior anchor points on their objects; the tool then creates complete point-to-point routings using A* path search and physical rod simulation (Figure \ref{fig:tool-process-exterior}).
-In the other, users can import vector art describing desired interior paths (Figure \ref{fig:tool-process-interior}).
-\systemname then creates a single path that follows the input shape using edge graph manipulation and Euler tour generation.
+In one, they select exterior anchor points on their objects; the tool then creates complete point-to-point routings using A* path search and physical rod simulation (Figure \ref{fig:tool-process-exterior}).
+In the other, users import vector art describing desired interior paths (Figure \ref{fig:tool-process-interior}).
+\systemname, using edge graph manipulation and Euler tour generation, creates a single path that follows the input shape.
 We thicken these routes to create pipes.
-The resultant pipes are subtracted from the original mesh using voxelization and remeshing.
+The resultant pipes are subtracted from the user's mesh via voxelization and remeshing.
 Our tool is implemented in C++ as an extension to Meshmixer, a consumer 3D mesh editing tool~\cite{Schmidt-meshmixer}.
@@
 \subsubsection{Pipes with Multiple Endpoints}
-To allow designers to create pipes with multiple endpoints (e.g., the star topologies as in Figure \ref{fig:designspace}), we use one surface position per pipe and a central fixed endpoint that the user can interactively position (Figure~\ref{fig:boat-star}).
+To allow designers to create pipes with multiple endpoints (e.g., the star topologies as in Figure \ref{fig:pipespace}), we use one surface position per pipe and a central fixed endpoint that the user can interactively position (Figure~\ref{fig:boat-star}).
 The user can connect several pipes, hitting a key to add each in sequence and positioning them with the mouse.
@@
 
+\begin{figure}
+\centering
+    \includegraphics[width=3.4in]{figures/radius.png}
+\caption{A user whose pipe has too small a radius (a) can adjust it with a slider (b) to be larger (c).
+PipeDream's live preview shows the real size and location of the cut before it is committed to the mesh.}
+\label{fig:radius}
+\end{figure}
+
 \subsection{Interior Paths}
@@
 A maker can import a vector graphics file (such as SVG) describing the path she wants for her pipes: all lines in this input file are interpreted as desired pipe components for the final artifact.
-Using sliders, she can change the pipes' radius.
+Using sliders, she can change the pipes' radius (Figure \ref{fig:radius}).
 A preview is provided to show the radius and location of the cut, and the maker can align her model as appropriate.
@@
 We then use marching cubes~\cite{Lorensen87} to produce the final mesh for printing.
-This strategy does incur some resampling artifacts, however so does the 3D printing, and compared to the print process, the time required to voxelize at high resolution is inconsequential.
+This strategy does incur some resampling artifacts, however so does the 3D printing; compared to the print process, the time required to voxelize at high resolution is inconsequential.
 Since we have a distance field, we can easily introduce useful effects, such as the
@@
 Several pragmatic concerns arise when fabricating models with hollow pipes and cavities on additive manufactuing machines.
-Pipes will frequently lead to overhangs (where an underlying supporting layer has a smaller footprint than the subsequent layer).
+Pipes frequently lead to overhangs (where an underlying supporting layer has a smaller footprint than the subsequent layer).
 %Beyond design, the tubes must actually be printable to be useful.  
 %When geometries contain overhangs, as there may be in designed pipes and cavities, 
-To allow overhangs, 3D printers lay down support material that must be later removed from the model.
-This removal process can be time-intensive and challenging.
+To allow overhangs, 3D printers may lay down support material that must be later removed from the model.
+The removal process can be time-intensive and challenging.
 The physical fabrication process used by different machines gives rise to differing strategies for avoiding support material or easing its removal.
-We next discuss techniques for fused deposition modeling (FDM) machines, such as the Makerbot\footnote{\url{http://makerbot.com}} and multi-jetting machines such as Objet\footnote{\url{http://stratasys.com}}, as well as powder-based machines (e.g., ZCorp\footnote{\url{http://3dsystems.com}}) and those that use stereolithography (e.g., Form 1\footnote{\url{http://formlabs.com}}).
+We next discuss techniques for fused deposition modeling (FDM) machines, such as the Makerbot\footnote{\url{http://makerbot.com}} and polyjet machines such as Stratasys Connex\footnote{\url{http://stratasys.com}}, as well as powder-based machines (e.g., ZCorp\footnote{\url{http://3dsystems.com}}) and those that use stereolithography (e.g., Form 1\footnote{\url{http://formlabs.com}}).
 
@@
 Meshmixer has a built-in tool for minimizing support for FDM prints; it re-orients the part to keep as much surface area as possible within the 45 degree constraint.
-In future work we would like to explore a tool in which a user can select surfaces (pipes) that should be prioritized for support avoidance.
+In future work, we plan to explore a tool allowing prioritzing specific surfaces (i.e., pipes) for support avoidance, or optimizing pipes to avoid overhangs.
 
-\subsection{Multi-Jetting}
-On multi-jetting machines, bridging is not possible and stepover potential is much less, because the machine is laying down droplets of liquid that are later UV cured.
-Unsupported overhangs can be approximately $14^{\circ}$ on our Objet printer.
-Multi-jetting and higher-end FDM machines use a secondary support material that must be manually removed afterwards.
-Some support can be melted out (Projet wax material); other support can be dissolved in lye or blasted with high-pressure water.
+\subsection{Polyjet}
+On polyjet machines, bridging is not possible and stepover is more constrained, because the machine is laying down droplets of liquid that are later UV cured.
+Unsupported overhangs can be approximately $14^{\circ}$ on our Connex printer.
+Polyjet and higher-end FDM machines use a secondary support material that must be manually removed afterwards.
+Some support can be melted out (Projet wax material); other varieties can be dissolved in lye or blasted with high-pressure water.
 Complex pipe geometries can make water blasting impossible and other processes more labor- and time-intensive.
-All black and clear models in Figure \ref{fig:teaser} required support removal post-print: the neon sign and maze were additionally cut into two pieces to give better access to the support.
+All black and clear models in Figure \ref{fig:teaser} required support removal post-print: the neon sign and maze were additionally cut into two pieces along the plane of the pipes to give better access to the support.
 

### 2014-04-17 paper/abstract.tex
submission to UIST
smash author names in

@@
 We investigate how 3D models can be modified to facilitate the creation of \emph{interactive} objects that offer dynamic input and output.
-%However, a 3D printer's role in prototyping interactive objects is still not clear.  
 We introduce a general technique for supporting the rapid prototyping of interactivity by removing interior material from 3D models to form internal pipes.
@@
 We use two distinct routing algorithms.
-The first allows users to define the pipes' terminals, and then uses path routing and physics-based simulation to minimize pipe bending energy, allowing easy insertion of media post-print.
+The first has users to define pipes' terminals, and uses path routing and physics-based simulation to minimize pipe bending energy, allowing easy insertion of media post-print.
 The second allows users to supply a desired internal shape to which we fit a pipe route: for this we developed a novel graph-based routing algorithm.
-We present several prototypes we created using our tool to show its flexibility and potential.
+We present several prototypes created using our tool to show its flexibility and potential.
 \end{abstract}

### 2014-04-17 paper/relatedwork.tex
smash author names in

@@
 HCI researchers are also developing novel interaction techniques for providing input to digital fabrication machines~\cite{Mueller-constructable,mixfab,Willis-interactive}, and new hybrid hand tools that combine manual processes with digital fabrication~\cite{rivers2012position,zoran2013freed}.
-These projects are complementary as they target different aspects of the design process, though it is conceivable that our pipe design tools could be added to augmented reality modeling interfaces like MixFab~\cite{mixfab}.
+These projects are complementary as they target different aspects of the design process, though it is conceivable that our pipe design tools be added to AR modeling interfaces like MixFab~\cite{mixfab}.
 
@@
 Infrastructs~\cite{Willis-infrastructs} encodes information using interior cavities that can be sensed using terahertz imaging.
-Bickel, et al.,~\cite{Bickel-deformation} explored enforcing deformation behavior with multi-material blending inside objects.
+Bickel, et al.,~\cite{Bickel-deformation} enforced deformation behavior with multi-material blending inside objects.
 Modeling internal chambers is one aspect of our work, however our focus is interaction rather than balance, identification, or deformation.

### 2014-04-23 paper/introduction.tex
fix orphans, add text about automatic plane cut

@@
 \section{Introduction}
-Makers, researchers, and professional designers (jointly ``makers" for the remainder of this paper) increasingly leverage 3D printers as tools for design work.
-A wide array of objects, ranging from toy figurines to video game controllers and jet engine blades, are now prototyped, and sometimes even manufactured, using these machines.
+Makers, researchers, and professional designers (jointly ``makers" in this paper) increasingly leverage 3D printers as tools for design work.
+A wide array of objects, from toy figurines to video game controllers and jet engine blades, are now prototyped or even manufactured using these machines.
 Most models fabricated on 3D printers today are passive: they express the form, but not the interactivity, of an object.
@@
 Pipes should not pass too close to the surface: consumer-grade 3D printers do not guarantee air- or water-tightness of prints, so more than one printed layer may be required to prevent leakage.
-Pipes for applications like illuminated EL wire signs must allow a single pipe to follow a specific path layout (e.g., for spelling out text).
+Pipes for applications like illuminated EL wire signs must allow a single pipe to follow a specific path (e.g., for writing text).
 
 We contribute PipeDream, an interactive design tool for interior pipes.
-Our system addresses the various challenges that makers are likely to encounter, such as avoiding intersections and ensuring maximum bend angles.
-PipeDream offers 3D point-to-point routing of internal pipes based on physical simulation of flexible rods, and a graph algorithm-based path tracing technique to create shape-constrained internal paths.
-We present a set of example objects, designed with the help of our modeling tool, that demonstrate different interaction modalities.
+Our system addresses various challenges makers are likely to encounter, such as avoiding intersections and ensuring maximum bend angles.
+PipeDream offers 3D point-to-point routing of internal pipes based on physical simulation of flexible rods, and a graph algorithm-based path tracing technique for path-constrained pipes.
+We present a set of example objects, designed with the help of our tool, that demonstrate different interaction modalities.
 We conclude by discussing limitations and outlining areas for future exploration.

### 2014-04-23 paper/tubes.tex
resize figure abcs, change caption of radius figure
fix orphans, add text about automatic plane cut

@@
     \includegraphics[width=3.4in]{figures/radius.png}
-\caption{A user whose pipe has too small a radius (a) can adjust it with a slider (b) to be larger (c).
-PipeDream's live preview shows the real size and location of the cut before it is committed to the mesh.}
+\caption{Users can interactively change the radius of their pipes using a slider (b).  PipeDream's live preview shows the real size and location of the cut before it is committed to the mesh.}
 \label{fig:radius}
@@
 
-{\bf Eulerization:} In order to make our graph Eulerian, we consider all odd degree vertices in the connected graph and make a clique of potential edges based on distance.
-We greedily add edges between the odd nodes until no odd nodes remain.
+{\bf Eulerization:} In order to make our graph Eulerian, we consider all odd-degree vertices in the connected graph and make a clique of potential edges based on distance.
+We greedily add edges between odd nodes until no odd nodes remain.
 Finally, we remove our temporary edge, changing the graph from Eulerian to semi-Eulerian.
-This is a connected, semi-Eulerian graph which contains all edges in the input.
-A lower total weight matching may be possible by connecting components and ensuring node evenness together in a global process, as well as by using minimum-weight matching rather than greedy selection, however this optimization is not crucial for our purposes.
+This is a connected, semi-Eulerian graph that contains all edges in the input.
+A lower total weight matching is possible by connecting components and ensuring node evenness together in a global process, as well as by using minimum-weight matching rather than greedy selection, however this optimization is not crucial for our purposes.
 
@@
 For our specific applications so far, it has been unnecessary.
+
+\subsubsection{Separating at the Pipe Plane}
+We allow users the choice to separate their meshes at the pipe plane, which may be necessary for support material removal (discussed later) or insertion of media: e.g., EL wire can be bent much more sharply by hand in the plane than it can by pushing from one end into a pipe.
+To do this, we simply perform a plane cut after the mesh modification operation described below.
 
@@
 Pipes frequently lead to overhangs (where an underlying supporting layer has a smaller footprint than the subsequent layer).
-%Beyond design, the tubes must actually be printable to be useful.  
-%When geometries contain overhangs, as there may be in designed pipes and cavities, 
 To allow overhangs, 3D printers may lay down support material that must be later removed from the model.
@@
 This is somewhat undesirable as it means more time spent in assembly post-print, and additionally a cut pipe may no longer be fluid-tight.
-A tool aiding in or automatically performing this cutting task might be beneficial, however it is outside the scope of this work.
+Our plane cut option goes towards this, however a more advanced tool for this purpose is outside the scope of this work.

### 2014-04-24 paper/designspace.tex
submission to UIST

@@
 We have identified four relevant dimensions of pipe design: \emph{openings}, \emph{path constraints}, \emph{topologies}, and inserted \emph{media}.
-We describe the space (see Figure \ref{fig:pipespace}).
+We describe the space (Figure \ref{fig:pipespace}).
 
@@
 \subsection{Openings}
-Pipes inside solid objects can be either \emph{open} to the outside on both ends, \emph{semi-closed}, or \emph{fully enclosed} (see Figure \ref{fig:pipespace}).
+Pipes inside solid objects can be either \emph{open} to the outside on both ends, \emph{semi-closed}, or \emph{fully enclosed} (Figure \ref{fig:pipespace}).
 For interactive devices, openings can be either user-facing, where a user would, for example, touch a model; or system-facing, where sensors and actuators may be connected.
@@
 
-\emph{Open} pipes may be used to create electronic circuitry.
-For example, an open pipe filled with conductive paint and connected to a battery can provide power to an LED (Figure \ref{fig:openings}a).
+\emph{Open} pipes may be used to create electronic circuitry.  For example, an open pipe filled with conductive paint and connected to a battery can power an LED (Figure \ref{fig:openings}a).
 %
-\emph{Semi-closed} pipes are capped on one side and could be used to create tactile output.
-By fabricating a cap from a malleable material and attaching a pump to the open end, haptic feedback can be generated by raising and depressing the flexible cap (see Figure \ref{fig:openings}b).
+\emph{Semi-closed} pipes are capped on one end and can be used to create tactile output.
+By fabricating a cap from a malleable material and attaching a pump to the open end, the flexible cap can be raised and depressed (Figure \ref{fig:openings}b).
 %
-\emph{Fully enclosed} pipes (cavities) are entirely internal to the object.
-Fully enclosed pipes may be used as weight modifiers~\cite{Prevost-makeitstand} and for object identification~\cite{Willis-infrastructs}, as in Figure \ref{fig:openings}c.
-If printed in parts and assembled post-print, enclosed pipes could also contain water or particles that would otherwise fall out.
+\emph{Fully enclosed} pipes (cavities) are entirely internal.
+They may be used as weight modifiers~\cite{Prevost-makeitstand} and for object identification~\cite{Willis-infrastructs} (Figure \ref{fig:openings}c).
+If printed in parts and assembled post-print, enclosed pipes could hold water or particles that would otherwise fall out.
 
@@
 In contrast, for a \emph{path-constrained} pipe the shape of the path is important.
-For example, threading electroluminescent (EL) wire through a clear pipe enables makers to emulate neon art (Figure \ref{fig:justneon}).
+For example, threading electroluminescent (EL) wire through a clear pipe enables makers to emulate neon art (Figure \ref{fig:teaser}, right).
 A \emph{fully constrained} pipe requires specific terminals as well as a specific path.
@@
 \subsection{Pipe Topologies}
-Pipe network topologies enable \emph{splitting} or \emph{mixing} of media (see Figure \ref{fig:pipespace}).
-\emph{Star} and \emph{tree} topologies extend the splitting and mixing primitives, encompassing multiple-in and multiple-out configurations; we use star topologies to create different sensing locations for our touch-sensitive toys, where one terminal is used for swept-frequency capacitive sensing (Figure~\ref{fig:teaser}, center).
+Pipe network topologies enable \emph{splitting} or \emph{mixing} of media (Figure \ref{fig:pipespace}).
+\emph{Star} and \emph{tree} topologies extend splitting and mixing primitives, encompassing multiple-in and multiple-out configurations; we use star topologies to create sensing locations for our touch-sensitive toys, with one terminal used for swept-frequency capacitive sensing (Figure~\ref{fig:teaser}, center).
 

### 2014-04-24 paper/relatedwork.tex

@@
 \subsubsection{Fabrication with Printed Electronics}
-Researchers are also investigating how to print conductors and place electronic components as part of additive manufacturing.
+Researchers are investigating how to print conductors and place electronic components as part of additive manufacturing.
 Sells, et al., \cite{Sells-reprap} use a heated syringe to deposit metal in printed 2D circuitboard channels on objects' surfaces.
-Other experiments have been done by the RepRap project\footnote{\url{http://reprap.org}} using plastic filament blended with conductive materials, however the resistance is currently too high to be useful in circuitry.
+Other experiments done by the RepRap project\footnote{\url{http://reprap.org}} use plastic filament blended with conductive materials, however the resistance is currently too high to be useful in circuitry.
 Sarik, et al., \cite{Sarik-tracebrush} use aerosol jetting of conductive material onto printed objects.
 Park, et al., \cite{Park-microchannels} and Majidi, et al., \cite{Majidi-curvature} inject liquid metals.
-Our approach does not deposit such materials automatically, but creates structures inside a 3D model that make manual post-print injection of conductive ink possible.
+Our approach does not deposit such materials automatically, but creates structures inside a model that make manual post-print injection of conductive ink possible.
 

### 2014-06-15 paper/UIST2014-rebuttal.txt
reviews and rebuttals

We would first like to thank the reviewers for their many, varied, and thoughtful comments.

CONTRIBUTION CLAIM OVERREACH (R1,R2) - We appreciate that our claims may have sounded overly broad, and agree that instances of interior pipes have been used in certain applications before. We will update statements of contribution to clarify.  For example, on p1, “This paper introduces a novel technique involving removing material...” can be changed to “This paper introduces a novel tool that provides path planning and other high-level functionality to support the design of interior pipes and cavities within 3D models.”

ALGORITHMIC CONTRIBUTION (1AC,R1,R2) - We understand that our system has limitations, and tried to openly describe these in our discussion section. We hope reviewers appreciate that the task of pipe routing is non-trivial and that the work we present offers a contribution to the UIST community; additionally it can serve as a stepping stone to more sophisticated future algorithmic contributions in this area. We can update our discussion section to provide additional insights and possible directions related to the specific limitations mentioned by R1.

We have considered the effectively 2D path-constrained routing problem which R2 brings up; it could be solved using a simulation similar to our endpoint-constrained condition, in which overlapping segments of the connected, Fleury-ized graph are represented as physically-based rods and pushed into the third dimension.  A globally optimum pipe solution requires defining a global pipe energy which can be minimized (based presumably on bending, pipe length, etc.), which we believe is based more on personal preference than global requirements and is outside the scope of this work. However, we will add this discussion to the limitations and future work section.

COMPLEX PIPE GEOMETRIES (R1) - We realize that our focus on useful applications of the technique in the paper does not sufficiently demonstrate the capabilities of our algorithms for routing through objects with complex shapes or topologies. Our tool can indeed route pipes through complex meshes. For example, we can successfully route pipes through knots and around holes in objects (genus >0).  We are happy to add images of routing solutions for more complex geometries in our revision.

SPECULATIVE OBJECTS SECTION (1AC) - In order to strengthen the paper, we are happy to either add additional examples to this section or to remove it, at the discretion of 1AC.

SUPPORT MATERIAL DISCUSSION (1AC,R2) - We agree to shorten this section of the paper and more plainly identify practical difficulties for those who do follow-on work.

VOXELIZATION VS. MESH BOOLEANS (R1) - The use of level sets allowed us more flexibility in, e.g., creating caps on our pipes.  We consider our particular choice orthogonal to the contributions, but this benefit of our voxelization implementation can be made more clear in a final revision.

SUNDRY CORRECTIONS/SUGGESTIONS (R2) - We appreciate this careful list of small changes and will integrate into final submission.

### 2014-06-29 paper/conclusion.tex
making changes thanks to reviews

@@
 This amplifies and extends the utility of existing sensing and actuation approaches.
-In addition to describing the design space of pipes, we also presented \systemnamenospace, a design tool for the creation of tubes inside arbitrary 3D models, and discussed many techniques that combine nicely with pipes.
+In addition to describing the design space of pipes, we also presented \systemnamenospace, a design tool for the creation of a series of tubes \cite{series-of-tubes} inside arbitrary 3D models, and discussed many techniques that combine nicely with pipes.
 We fabricated several example objects using our tool to explore new points in the design space.

### 2014-06-29 paper/discussion.tex
making changes thanks to reviews

@@
 Threadable materials in particular can provide challenges, as they can get stuck while being fed into pipes.
-Our system mitagates this by minimizing the bend energy of the interior pipes, but for complex pipe geometries threading could be a challenge.
+Our system mitigates this by minimizing the bend energy of the interior pipes, but for complex pipe geometries threading could be a challenge.
 
@@
 A 2.5D SVG import using, e.g., line colors as height would be straightforward, and our described algorithm would continue to work properly.
+An extension to \systemname could include detection of where the 2D path intersects itself; intersecting pipe components could be ``pushed into 3D'' using the discrete rods from our endpoint-constrained routing algorithm.
 Furthermore, since pipes are committed to the mesh one at a time, our tool does not currently optimize networks of multiple pipes.
 Instead, we utilize a greedy algorithm and route a single pipe at a time.
-This suggests the need for a better general 3D routing algorithm with multiple traces.
+This suggests the need for a better general 3D routing algorithm with multiple traces, which could take into account global bending energy, global pipe length, or other task-specific requirements.
 For integration with common electronic components, it would be possible to build a library of ``footprints'' which could be used for recessing them into printed objects' surfaces; such footprints could also be used for automatically generating connecting pipes based on an electrical connection graph, as in circuitboard routing tools like Eagle\footnote{\url{http://www.cadsoftusa.com}}.

### 2014-06-29 paper/introduction.tex
making changes thanks to reviews

@@
 Recently, human-computer interaction researchers have begun to explore how to prototype interactive objects using additive 3D printing processes.
-%and maker communities like Thingiverse\footnote{http://thingiverse.com}, recent work has looked at the issue of rapidly prototyping interactive objects. 
-Existing approaches include Printed Optics \cite{Willis-printedoptics}, which uses light pipes printed in optically clear material, and Sauron \cite{Savage-sauron}, which uses computer vision to track the movement of components on the inside of a hollow object.
-These systems exclusively focus on optical sensing and visual displays --- we seek to grow the vocabulary of supported tangible input and output modalities.
-%is based on computer vision, but they are limited to visible (reflecting light off fingers, mechanical motion of components) input and visual (screen-like) output; they do not give makers freedom to integrate haptic feedback or other inputs or outputs.
+Existing approaches, including Printed Optics \cite{Willis-printedoptics} and Sauron \cite{Savage-sauron}, exclusively focus on optical sensing and visual displays --- we seek to grow the vocabulary of supported tangible input and output modalities.
 
-This paper introduces a novel technique involving removing material from the interior of 3D models, prior to printing, to create pipes and other cavities.
+This paper introduces a novel tool that provides path planning and other high-level functionality to support the design of interior pipes and cavities within 3D models.
 These pipes can be filled, post-print, with a variety of media to enable input, display, and tactile feedback.

### 2014-06-29 paper/relatedwork.tex
making changes thanks to reviews

@@
 Sarik, et al., \cite{Sarik-tracebrush} use aerosol jetting of conductive material onto printed objects.
-Park, et al., \cite{Park-microchannels} and Majidi, et al., \cite{Majidi-curvature} inject liquid metals.
-Our approach does not deposit such materials automatically, but creates structures inside a model that make manual post-print injection of conductive ink possible.
+Park, et al., \cite{Park-microchannels} and Majidi, et al., \cite{Majidi-curvature} inject liquid metals, and Hudson \cite{Hudson-bears} provides techniques for embedding in felted objects.
+Our approach does not deposit these materials automatically, but creates structures inside a model that make manual post-print injection of conductive ink possible.
 
 \subsubsection{Interaction for Fabrication vs. Fabrication for Interaction}
-HCI researchers are also developing novel interaction techniques for providing input to digital fabrication machines~\cite{Mueller-constructable,mixfab,Willis-interactive}, and new hybrid hand tools that combine manual processes with digital fabrication~\cite{rivers2012position,zoran2013freed}.
+HCI researchers are also developing novel interaction techniques for providing input to digital fabrication machines~\cite{Mueller-constructable,mixfab,Willis-interactive}, new applications to extend machine capabilities \cite{Mueller-laserorigami,Hudson-bears}, and new hybrid hand tools that combine manual processes with digital fabrication~\cite{rivers2012position,zoran2013freed}.
 These projects are complementary as they target different aspects of the design process, though it is conceivable that our pipe design tools be added to AR modeling interfaces like MixFab~\cite{mixfab}.

### 2014-06-29 paper/tubes.tex
making changes thanks to reviews

@@
 Our voxel grid currently uses a resolution of 128x128x128 voxels, which enables interactive operations, but this parameter can be adjusted for different model requirements.
-The simulation takes about a second for our examples.
+The simulation takes about a second for our examples, including those in Figure \ref{fig:complex}.
+
+\begin{figure}[h!]
+\centering
+    \includegraphics[width=3.4in]{figures/routing-tests.png}
+\caption{A set of example routings produced by our tool.
+Note that because the initial state of the discrete rod is based on an A* routing, it cannot jump around holes in a genus $>0$ mesh (a) even if that would reduce overall bending.
+Pipes successfully route through long, curving paths in knots (b), and multiple pipes placed in a single mesh in a tangled configuration successfully avoid each other (c).}
+\label{fig:complex}
+\end{figure}
 
@@
 Our technique for routing this path is inspired by work in generating continuous line illustrations from images~\cite{Bosch-tsp,Wong-continuousline}.
-Our routing problem is a version of the Chinese Postman Problem\footnote{The CPP is also known as the route inspection problem: \url{http://en.wikipedia.org/wiki/Route_inspection_problem}}, in which we wish to traverse all edges of the graph described by the user's input, much as a postman needs to walk along every road at least once to deliver mail.
+Our routing problem is a version of the Chinese Postman Problem~\cite{CPP}, in which we wish to traverse all edges of the graph described by the user's input, much as a postman needs to walk along every road at least once to deliver mail.
 In our relaxation, we allow the creation of new edges (i.e., the postman may jump from building to building in addition to using existing roads).
@@
 
-{\bf Euler Tour:} To find an Euler tour, we use Fleury's algorithm\footnote{\url{http://en.wikipedia.org/wiki/Eulerian_path\#Fleury.27s_algorithm}} with a modification to preserve continuity: at a given node, from multiple candidate edges, we prefer the edge(s) with the smallest angular deviation from the incoming edge (i.e., we prefer to pass straight through a node, if possible).
+{\bf Euler Tour:} To find an Euler tour, we use Fleury's algorithm \cite{Fleury} with a modification to preserve continuity: at a given node, from multiple candidate edges, we prefer the edge(s) with the smallest angular deviation from the incoming edge (i.e., we prefer to pass straight through a node, if possible).
 This minimizes turns in the final artifact, which eases support material removal and assembly.
@@
 
-To avoid the complexities of direct mesh booleans, we construct a discretized volumetric (i.e., voxel-based) representation of the original mesh and the tube mesh.
+To avoid the complexities of direct mesh booleans and give us additional flexibility, we construct a discretized volumetric (i.e., voxel-based) representation of the original mesh and the tube mesh.
 Boolean subtraction is trivial in such representations.
@@
 \section{Fabrication Techniques}
-Several pragmatic concerns arise when fabricating models with hollow pipes and cavities on additive manufactuing machines.
+Several pragmatic concerns arise when fabricating models with pipes and cavities.
 Pipes frequently lead to overhangs (where an underlying supporting layer has a smaller footprint than the subsequent layer).
@@
 The physical fabrication process used by different machines gives rise to differing strategies for avoiding support material or easing its removal.
-We next discuss techniques for fused deposition modeling (FDM) machines, such as the Makerbot\footnote{\url{http://makerbot.com}} and polyjet machines such as Stratasys Connex\footnote{\url{http://stratasys.com}}, as well as powder-based machines (e.g., ZCorp\footnote{\url{http://3dsystems.com}}) and those that use stereolithography (e.g., Form 1\footnote{\url{http://formlabs.com}}).
+Techniques differ for fused deposition modeling (FDM) machines, such as the Makerbot\footnote{\url{http://makerbot.com}}, polyjet machines such as Stratasys Connex\footnote{\url{http://stratasys.com}},  powder-based machines (e.g., ZCorp\footnote{\url{http://3dsystems.com}}), and those that use stereolithography (e.g., Form 1\footnote{\url{http://formlabs.com}}).
 
-\subsection{Fused Deposition Modeling}
-FDM machines lay down lines of extruded plastic filament on top of each other: this allows {\em stepover}, when a layer of plastic is laid slightly offset from the layer below it, and {\em bridging}, when plastic is laid across a gap between two surfaces.
-For a layer to fix correctly to its supporting layer, roughly half of the filament needs to overlap.
-This means that overhangs without added support material can be approximately $45^{\circ}$, and although cylinders (e.g., pipes) and spheres violate this in some places they can be printed support-free.
-The colored models in Figure \ref{fig:teaser} were produced on a Makerbot FDM without any support structures inside the pipes.
+\subsection{Support Types}
+The physical form of support material varies between printer types.
+On single-material hobbyist FDM machines, the support is created from the same type of material as the model, laid down in a different pattern to ease detachment.
+Multi-material FDM machines typically feature dissolvable support material: by submersing a part in lye, one can eventually remove support material while leaving the model material behind.
+Polyjet machines also use a secondary type of material for support: it can be melted out in some cases (e.g., ProJet wax material) or blasted out with high-pressure water.
+This blasting process requires fairly open access to the support material: the robot head and maze in Figure \ref{fig:teaser} were cut into two pieces along the plane of the pipes to give better access to the support.
+For powder-based and stereolithography machines, the material (powder or photo-cure liquid) that is not used for the model remains in the bed and forms the support structure --- this is uniquely well-suited to removal, as it can simply be poured out after a print is completed.
+We hope to experiment with these types of printers more in the future.
 
-On single-material hobbyist FDM printers, support structures are built using the same material as the model.
-Such support may be impossible to remove from internal cavities.
-Meshmixer has a built-in tool for minimizing support for FDM prints; it re-orients the part to keep as much surface area as possible within the 45 degree constraint.
-In future work, we plan to explore a tool allowing prioritzing specific surfaces (i.e., pipes) for support avoidance, or optimizing pipes to avoid overhangs.
-
-\subsection{Polyjet}
-On polyjet machines, bridging is not possible and stepover is more constrained, because the machine is laying down droplets of liquid that are later UV cured.
-Unsupported overhangs can be approximately $14^{\circ}$ on our Connex printer.
-Polyjet and higher-end FDM machines use a secondary support material that must be manually removed afterwards.
-Some support can be melted out (Projet wax material); other varieties can be dissolved in lye or blasted with high-pressure water.
-Complex pipe geometries can make water blasting impossible and other processes more labor- and time-intensive.
-All black and clear models in Figure \ref{fig:teaser} required support removal post-print: the neon sign and maze were additionally cut into two pieces along the plane of the pipes to give better access to the support.
-
-\subsection{Powder-based and Stereolithography}
-Powder-based and stereolithography (SLA) machines work with a bed of powder or liquid, respectively.
-The material at the surface is fused together via glue droplets (for powder) or a UV laser (for SLA).
-The remaining powder or liquid serves as support for the model.
-These types of printers are perhaps even more suitable for pipe creation than either FDM or multi-jetting printers, as the supporting liquid or loose powder is easier to remove than structural plastic or UV cured material.
-We hope to experiment with them in the future.
+\subsection{Model Orientation and Pipe Geometries}
+The various types of printers permit different sorts of support-free prints.
+FDM allows approximately a $45^\circ$ overhang before support is required.
+For polyjet machines, overhangs can be just $14^\circ$.
+Reorienting a part within the bed can lead to more or less support required.
+MeshMixer already has a built-in tool that reorients models to keep as much surface area as possible within the $45^\circ$ constraint imposed by FDM printers, and alternative pipe cross-section geometries (e.g., teardrop) could alleviate the need for support material to bridge the tops of pipes.
+In the future, pipe routes could be optimized to minimize support material required.
 
 \subsection{Model Assembly}
-For especially complex geometries, models can be ``cut up'' into multiple pieces that can be assembled to avoid support deposition or at least ease material removal; it can also ease insertion of the desired medium.
+For especially complex geometries, models can be ``cut up'' into multiple pieces that can be assembled, avoiding support deposition or easing material removal; this can also ease insertion of the desired medium.
 This is somewhat undesirable as it means more time spent in assembly post-print, and additionally a cut pipe may no longer be fluid-tight.

### 2014-07-01 paper/abstract.tex
paper mods

@@
 3D printers offer extraordinary flexibility for prototyping the shape and mechanical function of objects.
-We investigate how 3D models can be modified to facilitate the creation of \emph{interactive} objects that offer dynamic input and output.
-We introduce a general technique for supporting the rapid prototyping of interactivity by removing interior material from 3D models to form internal pipes.
-We describe this new design space of pipes for interaction design, where variables include openings, path constraints, topologies, and inserted media.
-We then present \systemnamenospace, a tool for routing such pipes through the interior of 3D models, integrated within a 3D modeling program.
+We investigate how 3D models can be modified to facilitate the creation of \emph{interactive} objects offering dynamic input and output.
+We introduce a general technique to support rapidly prototyping interactivity by removing interior material from 3D models to form internal pipes.
+We describe the design space of pipes for interaction design, where variables include openings, path constraints, topologies, and inserted media.
+We then present \systemnamenospace, a tool for routing internal pipes through 3D models, integrated within a 3D modeling program.
 We use two distinct routing algorithms.
-The first has users to define pipes' terminals, and uses path routing and physics-based simulation to minimize pipe bending energy, allowing easy insertion of media post-print.
-The second allows users to supply a desired internal shape to which we fit a pipe route: for this we developed a novel graph-based routing algorithm.
-We present several prototypes created using our tool to show its flexibility and potential.
+The first has users define pipes' terminals and uses path routing and physics-based simulation to minimize pipe bending energy, allowing easy insertion of media post-print.
+The second lets users supply a desired internal shape to which it fits a pipe route: for this we developed a novel graph-based routing algorithm.
+We present prototypes created using our tool showing its flexibility and potential.
 \end{abstract}

### 2014-07-01 paper/designspace.tex
making changes thanks to reviews
paper mods

@@
 \section{The Design Space of Pipes}
-Our review of prior art indicates the potential benifits of using interior pipes for adding interactivity to 3D printed objects, although no general-purpose tools currently exist for this purpose.
+Our review of prior art indicates the potential benefits of using interior pipes for adding interactivity to 3D printed objects, although no general-purpose tools currently exist for this purpose.
 To guide the design of such a tool, we now provide a design space of the types of pipes that a maker may want to utilize.
 We have identified four relevant dimensions of pipe design: \emph{openings}, \emph{path constraints}, \emph{topologies}, and inserted \emph{media}.
-We describe the space (Figure \ref{fig:pipespace}).
+We describe the space in Figure \ref{fig:pipespace}.
 
@@
 
-%\begin{figure}
-%\centering
-%    \includegraphics[width=1.0\columnwidth]{figures/glamour/return.jpg}
-%\caption{A path-constrained open pipe with EL wire as a ``neon'' design.}
-%\label{fig:justneon}
-%\end{figure}
-
 \subsection{Path Constraints}
@@
 Finally, an \emph{unconstrained} pipe does not need a specific path or terminal points.
-A cavity to modify an object's weight is an example of an unconstrained pipe (Figure \ref{fig:openings}c).
+A cavity to modify an object's overall weight is an example of an unconstrained pipe (Figure \ref{fig:openings}c).
 
@@
 A single large particle can be used for display.
-Small particles in variable amounts of fluid can provide variable stiffness haptics via jamming \cite{Follmer-jamming}.
+Small particles agitated by air can generate sound.
 %

### 2014-07-01 paper/discussion.tex
paper mods

@@
 We are also eager to test our tool with users and get feedback from makers who use it in the wild.
-Based on our own experiences, we believe our system could be readily utilized by makers and researchers, and serve as a useful tool for rapdily proptotyping interactive 3D objects.
+Based on our own experiences, we believe our system could be readily utilized by makers and researchers, and serve as a useful tool for rapidly prototyping interactive 3D objects.
+\changes{We plan to release PipeDream as a part of Meshmixer later this year.}

### 2014-07-01 paper/introduction.tex
paper mods

@@
 Recently, human-computer interaction researchers have begun to explore how to prototype interactive objects using additive 3D printing processes.
-Existing approaches, including Printed Optics \cite{Willis-printedoptics} and Sauron \cite{Savage-sauron}, exclusively focus on optical sensing and visual displays --- we seek to grow the vocabulary of supported tangible input and output modalities.
+Existing approaches, including Printed Optics \cite{Willis-printedoptics} and Sauron \cite{Savage-sauron}, exclusively focus on optical sensing and visual displays --- we seek to grow the vocabulary of supported tangible input and output modalities, and we propose using pipes and cavities, allowing designers to integrate secondary materials post-print, for this purpose.
 
-This paper introduces a novel tool that provides path planning and other high-level functionality to support the design of interior pipes and cavities within 3D models.
-These pipes can be filled, post-print, with a variety of media to enable input, display, and tactile feedback.
-This {\em subtractive} approach is complementary to {\em additive} approaches that inject custom print materials, such as conductors, during the actual printing process \cite{Sells-reprap}.
+\changes{This paper introduces a novel tool that provides path planning and other high-level functionality to support the design of interior pipes and cavities within 3D models.
+These pipes can be filled, post-print, with a variety of media to enable input, display, and tactile feedback.}  This {\em subtractive} approach is complementary to {\em additive} approaches that inject custom print materials, such as conductors, during the actual printing process \cite{Sells-reprap}.
 While our approach requires some manual assembly after a print completes, it gives makers the opportunity to cheaply and rapidly prototype a range of interactive objects on 3D printers, including popular single-material fused-deposition modeling machines (colored objects in Figure \ref{fig:teaser}).
+\changes{Additionally, once printers able to deposit conductive materials become commercially available, PipeDream will still support the internal 3D modeling tasks required for directly printing different materials.}
 
@@
 For example, injecting conductive paint can turn pipes into conductors, which can in turn integrate electronic components into prints; electroluminescent (EL) wire can be threaded through pipes to create computer-controlled visual output; or air can be pumped through pipes to produce haptic effects.
-Existing sensing and actuation approaches, such as FlyEye \cite{Wimmer-flyeye}, Swept Frequency Capacitive Sensing \cite{Sato-touche}, and Jamming User Interfaces \cite{Follmer-jamming}, can be expanded to new contexts using pipes: makers utilizing such I/O strategies can locate input and output at arbitrary points on a printed 3D object's surface.
+Existing sensing and actuation approaches, such as FlyEye \cite{Wimmer-flyeye}, swept-frequency capacitive sensing \cite{Sato-touche}, and Jamming User Interfaces \cite{Follmer-jamming}, can be expanded to new contexts using pipes: makers utilizing such I/O strategies can locate input and output at arbitrary points on a printed 3D object's surface.
 We leverage these existing techniques for sensing and actuation while our work's novelty is in \emph{internal modeling} for 3D printing and the \emph{redirection} of I/O to or through arbitrary locations on a device.

### 2014-07-01 paper/relatedwork.tex
paper mods

--- 2014-04-24 paper/relatedwork.tex
@@
 
-Sauron~\cite{Savage-sauron} tracks interactions with printed physical mechanisms by placing a camera inside hollow objects, where all components must be visible to the camera.
-Our technique instead can work with arbitrary, solid models and only carves out pipes for interaction where needed.
+Sauron~\cite{Savage-sauron} tracks interactions with printed physical mechanisms by placing a camera inside hollow objects, where all components must be visible to it.
+Our technique instead can work with arbitrary, solid models and only carves pipes for interaction where needed.
 Midas~\cite{Savage-midas} generates fabrication files for custom capacitive touch sensors.
-Their technique is limited to planar and unfoldable shapes as it routes connections on the surface of objects.
-\systemname can also be used for touch sensing, but routes conductors through the inside of objects.
+Their technique is limited to planar and unfoldable shapes, as it routes connections on the surface of objects.
+\systemname can also be used for touch sensing, but routes conductors \emph{through} objects.
 
@@
 Sarik, et al., \cite{Sarik-tracebrush} use aerosol jetting of conductive material onto printed objects.
-Park, et al., \cite{Park-microchannels} and Majidi, et al., \cite{Majidi-curvature} inject liquid metals.
-Our approach does not deposit such materials automatically, but creates structures inside a model that make manual post-print injection of conductive ink possible.
+Park, et al., \cite{Park-microchannels} and Majidi, et al., \cite{Majidi-curvature} inject liquid metals, \changes{and Hudson \cite{Hudson-bears} provides techniques for embedding conductive thread and other electronics within felted objects.
+}  Our approach does not deposit these materials automatically, but creates structures inside a model that make manual post-print injection of conductive ink possible.
 
 \subsubsection{Interaction for Fabrication vs. Fabrication for Interaction}
-HCI researchers are also developing novel interaction techniques for providing input to digital fabrication machines~\cite{Mueller-constructable,mixfab,Willis-interactive}, and new hybrid hand tools that combine manual processes with digital fabrication~\cite{rivers2012position,zoran2013freed}.
+HCI researchers are also developing novel interaction techniques for providing input to digital fabrication machines~\cite{Mueller-constructable,mixfab,Willis-interactive}, \changes{new applications to extend machine capabilities \cite{Mueller-laserorigami,Hudson-bears}}, and new hybrid hand tools that combine manual processes with digital fabrication~\cite{rivers2012position,zoran2013freed}.
 These projects are complementary as they target different aspects of the design process, though it is conceivable that our pipe design tools be added to AR modeling interfaces like MixFab~\cite{mixfab}.

### 2014-07-01 paper/tubes.tex
paper mods

@@
 The interior of these pipes should have as large a bending radius as possible so that post-print insertion of solid or viscous media is easier.
-To create endpoint-constrained pipes, in which the location of exterior connection points matters, a brush tool is used to specify a pair of points on the mesh's surface.
+To create endpoint-constrained pipes, in which the location of exterior connection points matters, a brush tool is used to specify a pair of points on the mesh's surface (Figure \ref{fig:tool-process-exterior}a).
 The system automatically routes the pipe, and displays a preview.
@@
 Our graph comprises the grid of voxels as nodes, with edges between adjacent nodes; our path cost is based on the Euclidean distance to the selected end point.
-In this formulation, the routed path is naturally constrained to stay within the object boundaries.
+In this formulation, the routed path is naturally constrained to stay within the object boundaries (Figure \ref{fig:tool-process-exterior}b).
 %; for each point it tags the distance from the start (e.g., startPoint = 0, startNeighbors=1, ...).
@@
 a straight configuration, while satisfying the user-selected endpoint position constraints.
-In addition to user constraints, we require that the rod be perpendicular to the surface at its ends to prevent the terminals' shape from deforming.
+In addition to user constraints, we require that the rod be perpendicular to the surface at its ends to prevent the terminals' shape from deforming (Figure \ref{fig:tool-process-exterior}c).
 
@@
     \includegraphics[width=3.4in]{figures/routing-tests.png}
-\caption{A set of example routings produced by our tool.
-Note that because the initial state of the discrete rod is based on an A* routing, it cannot jump around holes in a genus $>0$ mesh (a) even if that would reduce overall bending.
-Pipes successfully route through long, curving paths in knots (b), and multiple pipes placed in a single mesh in a tangled configuration successfully avoid each other (c).}
+\caption{\changes{A set of example routings produced by our tool.
+Note that because the initial state of the discrete rod is based on an A* routing, it cannot ``jump around'' holes in a genus $>0$ mesh (a) to route on the other side, even if this would reduce overall bending.
+Pipes successfully route through long, curving paths in knots (b), and multiple pipes placed in a single mesh in a tangled configuration successfully avoid each other (c).}}
 \label{fig:complex}
@@
     \includegraphics[width=3.4in]{figures/radius.png}
-\caption{Users can interactively change the radius of their pipes using a slider (b).  PipeDream's live preview shows the real size and location of the cut before it is committed to the mesh.}
+\caption{Users can interactively change the radius of their pipes using a slider (b).  PipeDream's live preview shows the real size and location of the cut (a, c) before it is committed to the mesh.}
 \label{fig:radius}
@@
 
-In some cases, such as for the neon sign in Figure \ref{fig:tool-process-interior}, a single path must pass through a set of disconnected path segments (i.e., all letters in the word ``UIST'' ).
+In some cases, such as for the neon route in Figure \ref{fig:tool-process-interior} and resultant sign in Figure \ref{fig:UIST}, a single path must pass through a set of disconnected path segments (i.e., all letters in the word ``UIST'' ).
 \systemname leverages graph theory algorithms to generate this single path, as described below.
@@
 
+\changes{
 \section{Fabrication Techniques}
@@
 Our plane cut option goes towards this, however a more advanced tool for this purpose is outside the scope of this work.
+}

### 2014-07-01 paper/exampleobjects.tex
more paper updates

@@
 When the rabbit inhales through its nose, its abdomen rises, and as it exhales its abdomen falls.
-For this, we used a combination air/vacuum pump (ROB-10398): one terminal creates a vacuum while the other creates positive pressure.
+For this, we used a combination air/vacuum pump (SparkFun ROB-10398): one terminal creates a vacuum while the other creates positive pressure.
 The rabbit model, the well-known Stanford Bunny\footnote{\url{http://en.wikipedia.org/wiki/Stanford_bunny}}, was downloaded from Thingiverse.
@@
 
-\subsection{Speculative Objects and Explorations}
-
-\begin{figure}
-\centering
-    \includegraphics[width=3.4in]{figures/speculative.png}
-\caption{Some possible pipe uses.
-The printed whistle in (a) is a Helmholz resonator: it creates a sound when air is blown through the pipe at the front, because air compresses at the small neck joint between the internal chamber and the pipe.
-(b) gives a comparison of pipe-based resistors, filled with copper paint.
-For a cylindrical wire, resistance is based on its length and cross-sectional area.}
-\label{fig:speculative}
-\end{figure}
-
-\begin{figure}
-\centering
-    \includegraphics[width=3.4in]{figures/gates.png}
-\caption{By attaching a servo to a gate structure, it is possible to direct fluid through that pipe.  The gate not only has binary positions (a) and (b), but can direct fractional flow, as well (c).}
-\label{fig:direct}
-\end{figure}
-
-Our six examples showcase how PipeDream can be used to design objects that support a wide range of input and output modalities.
-In addition to these six complete examples, we also created some smaller explorations of techniques and possibilities using pipes.
-
-Creating aural outputs using printed pipes is one such possibility: the whistle in Figure \ref{fig:speculative}a is a fully printable audio output based on a Helmholz resonator, which can be activated with high pressure air.
-Zoran \cite{Zoran-flute} previously explored printable musical instruments, and it would be interesting to extend PipeDream to aid in the design process of such instruments.
-
-Some passive components required in the design of circuitry could in the future be created using pipes and printed or injected conductive material.
-For example, tubes can be used to create \emph{resistors} in ``printed'' circuitry: a longer or thinner tube has a greater resistance than a shorter or wider one (see Figure \ref{fig:speculative}b).
-
-Liquids can be mixed in specific concentrations (e.g., for color output) using mixing or splitting topologies along with a servo-controlled gate structure as in Figure \ref{fig:direct}.
-This could be used in creating interactive science exhibits about, e.g., logic gates.
-
-Identification of objects is possible using active acoustic sensing \cite{Ono-touchandactivate}: the internal pipes, although not externally visually distinctive, affect the acoustic signature of the object.
-While this is an opportunity for record and recall identification, intentional identity encoding using pipes could also be accomplished, through print weight modification using fully enclosed pipes.
+\changes{Section removed: Speculative Objects and Explorations}

### 2014-07-09 paper/designspace.tex
update figures with new routings

@@
 If a maker wishes the EL wire to exit the rear of the sign to hide the electronics, she can use a fully-constrained path (Figure \ref{fig:teaser}, right).
-Finally, an \emph{unconstrained} pipe does not need a specific path or terminal points.
-A cavity to modify an object's overall weight is an example of an unconstrained pipe (Figure \ref{fig:openings}c).
+Pipes may also be fully \emph{unconstrained}, however in this paper we focus on contexts where designers deliberately place constraints on pipe endpoints or path.
 

### 2014-07-09 paper/introduction.tex
update figures with new routings

@@
 Recently, human-computer interaction researchers have begun to explore how to prototype interactive objects using additive 3D printing processes.
-Existing approaches, including Printed Optics \cite{Willis-printedoptics} and Sauron \cite{Savage-sauron}, exclusively focus on optical sensing and visual displays --- we seek to grow the vocabulary of supported tangible input and output modalities, and we propose using pipes and cavities, allowing designers to integrate secondary materials post-print, for this purpose.
+\changes{Existing approaches, including Printed Optics \cite{Willis-printedoptics} and Sauron \cite{Savage-sauron}, exclusively focus on optical sensing and visual displays --- we seek to grow the vocabulary of supported tangible input and output modalities, and we propose using pipes and cavities, allowing designers to integrate secondary materials post-print, for this purpose.}
 

### 2014-07-09 paper/tubes.tex
update figures with new routings

@@
 \caption{\changes{A set of example routings produced by our tool.
-Note that because the initial state of the discrete rod is based on an A* routing, it cannot ``jump around'' holes in a genus $>0$ mesh (a) to route on the other side, even if this would reduce overall bending.
+Note that there is a tradeoff between bending and length of pipes: our pipes prefer shortness, however pipes could "jump around" holes in genus $> 0$ meshes (a) to reduce bending.
 Pipes successfully route through long, curving paths in knots (b), and multiple pipes placed in a single mesh in a tangled configuration successfully avoid each other (c).}}

### 2014-07-09 paper/introduction.tex
final tweaks for 1AC to check accepted submission

@@
 While our approach requires some manual assembly after a print completes, it gives makers the opportunity to cheaply and rapidly prototype a range of interactive objects on 3D printers, including popular single-material fused-deposition modeling machines (colored objects in Figure \ref{fig:teaser}).
-\changes{Additionally, once printers able to deposit conductive materials become commercially available, PipeDream will still support the internal 3D modeling tasks required for directly printing different materials.}
+\changes{Our software contributions can also provide useful modeling capabilities for future multi-material printers, e.g., printers that that are able to deposit conductive materials directly, without manual assembly.}
 

### 2014-07-09 paper/tubes.tex
final tweaks for 1AC to check accepted submission

@@
 mesh, we compute a discretized distance field, then constrain the
-rod to stay within our offset shell using a simple penalty force.
+rod to stay within our offset shell using a simple penalty force. \changes{This approach can successfully route pipes through complex geometries (Figure \ref{fig:complex}).}
 
@@
 Our voxel grid currently uses a resolution of 128x128x128 voxels, which enables interactive operations, but this parameter can be adjusted for different model requirements.
-The simulation takes about a second for our examples, including those in Figure \ref{fig:complex}.
+The simulation takes about a second for our examples.
 
@@
 \caption{\changes{A set of example routings produced by our tool.
-Note that there is a tradeoff between bending and length of pipes: our pipes prefer shortness, however pipes could "jump around" holes in genus $> 0$ meshes (a) to reduce bending.
+Note that there is a tradeoff between bending and length of pipes: to reduce bending, pipes may be routed through longer paths around holes in genus $> 0$ meshes (a).
 Pipes successfully route through long, curving paths in knots (b), and multiple pipes placed in a single mesh in a tangled configuration successfully avoid each other (c).}}
@@
 Pipes are routed and cut out of the mesh serially.
-This prevents pipe intersections: once a pipe has been cut out of the mesh, it cannot be part of valid routes in any future pipe routing operation.
+\changes{This prevents pipe intersections: once a pipe has been cut out of the mesh, it cannot be part of valid routes in any future pipe routing operation (Figure~\ref{fig:complex}c)}.
 After solving for a path, we add it as a constraint to the system, generating
@@
 This is somewhat undesirable as it means more time spent in assembly post-print, and additionally a cut pipe may no longer be fluid-tight.
-Our plane cut option goes towards this, however a more advanced tool for this purpose is outside the scope of this work.
+%Our plane cut option goes towards this, however a more advanced tool for this purpose is outside the scope of this work.
+Our plane cut option offers a simple way to slice a model; more sophisticated techniques may be necessary for internal paths that are not constrained to a plane.
 }

### 2014-07-11 paper/introduction.tex
change statements about prior art

@@
 Recently, human-computer interaction researchers have begun to explore how to prototype interactive objects using additive 3D printing processes.
-\changes{Existing approaches, including Printed Optics \cite{Willis-printedoptics} and Sauron \cite{Savage-sauron}, exclusively focus on optical sensing and visual displays --- we seek to grow the vocabulary of supported tangible input and output modalities, and we propose using pipes and cavities, allowing designers to integrate secondary materials post-print, for this purpose.}
+\changes{Some existing approaches, including Printed Optics \cite{Willis-printedoptics} and printing teddy bears \cite{Hudson-bears} create one-off designs for interactive devices by using high-end or custom printers which to fabricate materials that are clear or soft.
+Others, like Sauron \cite{Savage-sauron}, offer design tools but focus on visual input and display-based output --- we seek to grow the input and output modalities available to makers.
+We propose using pipes and cavities, allowing integration of secondary materials post-print, for this purpose.}
 

### 2014-07-14 paper/introduction.tex
per chair

--- 2014-07-09 paper/introduction.tex
@@
 
-Recently, human-computer interaction researchers have begun to explore how to prototype interactive objects using additive 3D printing processes.
-\changes{Existing approaches, including Printed Optics \cite{Willis-printedoptics} and Sauron \cite{Savage-sauron}, exclusively focus on optical sensing and visual displays --- we seek to grow the vocabulary of supported tangible input and output modalities, and we propose using pipes and cavities, allowing designers to integrate secondary materials post-print, for this purpose.}
+\changes{Recently, human-computer interaction researchers have begun to explore how to add interaction to 3D printed objects by creating appropriate structures inside those objects.
+They have introduced techniques such as printed light pipes~\cite{Willis-printedoptics} and printed hollow chambers for air pressure sensing~\cite{slyper}.
+These projects introduce point examples of particular techniques.
+In this paper we present a systematic design space of using 3D printed pipes and cavities, and introduce a design tool that enables makers to add interactivity to their models through internal pipes.}
 

### 2014-07-14 paper/introduction.tex
fitting

@@
 
-\changes{Recently, human-computer interaction researchers have begun to explore how to add interaction to 3D printed objects by creating appropriate structures inside those objects.
-They have introduced techniques such as printed light pipes~\cite{Willis-printedoptics} and printed hollow chambers for air pressure sensing~\cite{slyper}.
+\changes{Recently, human-computer interaction researchers have begun to explore adding interaction to 3D printed objects by creating appropriate structures inside those objects.
+They have introduced techniques such as printed light pipes~\cite{Willis-printedoptics} and printed hollow chambers for air pressure sensing~\cite{Slyper-pressure}.
 These projects introduce point examples of particular techniques.
-In this paper we present a systematic design space of using 3D printed pipes and cavities, and introduce a design tool that enables makers to add interactivity to their models through internal pipes.}
+In this paper we present a systematic design space of 3D printed pipes and cavities, and introduce a design tool enabling makers to add interactivity to their models using internal pipes.}
 
